### 2020-05-12 templates/template_lab_report/report_example.tex
initial material

%%
%% This document created by A.Tasora
%%

\documentclass{digitaldynamics}

%%
%% Add additional packages here... --------------------------------------------------
%%

\usepackage{graphicx}
\usepackage{amsmath, amssymb} 
\usepackage[english]{alg}
\usepackage{empheq}
\usepackage{bm}
\usepackage{listings}
\usepackage{nccmath}

% For bibliography: use biber and biblatex. 
% NOTE: your IDE build toolchain should use biber instead of bibtex when compiling.
\usepackage[backend=biber]{biblatex}

% Use \addbibresource{...}  to use .bib databases from our GIT repository if possible.
\addbibresource[location=remote]{"https://raw.githubusercontent.com/DigitalDynamicsLab/bibliography/master/refsFEM.bib"}
\addbibresource[location=remote]{"https://raw.githubusercontent.com/DigitalDynamicsLab/bibliography/master/refsMBS.bib"}

\sloppy

%%
%% Custom commands  --------------------------------------------------------
%%

	 %% vector as boldface   
\def\vect#1{\bm{#1}}
\def\vcovar#1{{\bm{#1}}^*}
\def\vcontr#1{\underline{\bm{#1}}}
\def\matr#1{{#1}}

\def\tensor#1{\bm{#1}}
\def\etensor#1{\underline{\bm{#1}}}
\def\eetensor#1{\underline{\underline{\bm{#1}}}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\title{Example of tech report}

\author{Alessandro Tasora \\ alessandro.tasora@unipr.it}

\maketitle
\thispagestyle{fancy}

\begin{abstract}

This is an example of tech report that uses the digitaldynamics.cls LaTeX class.
Use this as a template to start writing documents in form of articles that will
be stored as internal or external documents of DigitalDynamics Lab.

\end{abstract}

\section{Introduction}

To use this class, create a directory and copy into it the files: digitaldynamics.cls  logo\_digitaldynamics\_red.png  and this .tex file as a starting point. 
You may want to edit digitaldynamics.cls if you want to change the header title.

Example of citing some literature: \cite{cosserat1909}. \cite{anderson00}. Good.

Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

\section{Conclusion}

Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

\printbibliography

\end{document}

### 2020-05-15 2019/CMAME_iga/iga_beams.tex
Added IGA paper in 2019

%%
%% This document created by A.Tasora
%%

\documentclass[review]{elsarticle}

%%
%% Additional packages... --------------------------------------------------
%%

\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{amsmath, amssymb}
\usepackage{amssymb}
\usepackage{psfrag}
\usepackage{siunitx}
\usepackage{booktabs}
%\usepackage[english]{alg}
\usepackage{algpseudocode}
% \usepackage{authblk}
\usepackage{subcaption}
\usepackage{hyperref}
% \usepackage{bm}
% \include{BibFEM.bib}

\sloppy

%% A.TASORA CUSTOM TEX COMMANDS:
%%

\def\avect#1{{\boldsymbol{#1}}}
\def\amatr#1{{\boldsymbol{#1}}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

%%
%% TITLE
%%

\title{A geometrically exact isogeometric beam for large displacements and contacts}

%\author{Alessandro Tasora \\ alessandro.tasora@unipr.it}
\author{Alessandro Tasora}
\ead{alessandro.tasora@unipr.it}
\author{Simone Benatti}
\ead{simone.benatti@studenti.unipr.it}
\author{Dario Mangoni}
\ead{dario.mangoni@unipr.it}
\author{Rinaldo Garziera}
\ead{rinaldo.garziera@unipr.it}
\address{Universit\`a degli Studi di Parma, \\Dpt. of Engineering and Architecture\\Parco Area delle Scienze, 181/A, 43124 Parma, Italy }

% \affil[]{Universit\`a degli Studi di Parma, \\ 
% Dpt. of Engineering and Architecture, Parco Area delle Scienze, 181/A, 43124 Parma, Italy 
% \footnote{E-mail addresses: alessandro.tasora@unipr.it (A. Tasora), \\ simone.benatti@studenti.unipr.it (S. Benatti) dario.mangoni@studenti.unipr.it (D. Mangoni), rinaldo.garziera@unipr.it (R.Garziera).}
% }

% \newcommand{\avect}[1]{{\boldsymbol{#1}}}
% \newcommand{\amatr}[1]{{\boldsymbol{#1}}}

%\thispagestyle{fancy}

\begin{abstract} 
This work discusses an efficient formulation of a geometrically exact three-dimensional beam which can be used in dynamical simulations involving large displacements, collisions and non-linear materials. To this end, we base our model on the shear-flexible Cosserat rod theory and we implement it in the context of Isogeometric Analysis (IGA). 
According to the IGA approach, the centerline of the beam is parameterized using splines; in our work the rotation of the section is parameterized by a spline interpolation of quaternions, and time integration of rotations is performed using the exponential map of quaternions. Aiming at an efficient and robust simulation of contacts, we propose the adoption of a non-smooth dynamics formulation based on differential-variational inequalities. 
The model has been implemented in an open-source physics simulation library that can simulate actuators, finite elements, rigid bodies, constraints, collisions and frictional contacts. 
This beam model has been tested on various benchmarks in order to assess its validity in non-linear static and dynamic analysis; in all cases the model behaved consistently with theoretical results and experimental data. 
\end{abstract}

\begin{keyword}
IGA \sep beam \sep contact \sep multibody \sep non-smooth dynamics
\end{keyword}

\maketitle

%%

%% SECTIONS
%%

\section{Introduction}

Deformable three-dimensional beams withstanding large displacements can be found in many scenarios of practical interest, this is the case of the blades in a helicopter rotor, for instance, or the case of the torsion bar in a car suspension, or the case of a flexible robot arm. In the last decades, similar problems motivated extensive research in the area of fast, robust and reliable computer methods for the simulation of beams, and most of those methods are based on Finite Element (FE) discretizations.

Recently, the Isogeometric Analysis (IGA) computational approach gained increased popularity as it blends the concept of spline interpolation in the context of FE. 
While traditional FE methods discretize the continuum using finite elements that share end nodes, the IGA approach uses a single spline with several nodes. Indeed, nowadays most CAD tools are based on splines of NURBS or B-spline type, hence one of the relevant advantages of IGA is that these geometries can be imported directly in the simulation, whereas FE methods require an intermediate pre-processing step in order to generate a mesh approximation of the original geometry~\cite{HUGHES2005}. Moreover, conventional finite elements using Lagrange polynomials feature $C^0$ continuity at nodal points regardless of their order, while spline functions of order $p$ provide $C^{p-1}$ continuity at all non-multiple knots. This  means that for the same amount of degrees of freedom, IGA shows better robustness and accuracy compared to $C^0$-continuous FE~\cite{COTTRELL2007refinement}.

Although the most obvious application of IGA is in the modeling of structural elements which can map to a line parameterization, such as cables and beams, the same concept can be extended to generic PDEs involving membranes, shells and volumes~\cite{COTTRELL20065257,Benson2010IGAShells}.
Among the vast literature on IGA, we cite in passing the developments on hierarchical refinement~\cite{VUONGSIMEON2011hierarchical}, the generalization to T-splines~\cite{BAZILEVS2010tsplines}, the application to fluid dynamics~\cite{Bazilevs2008fluid} and to contact problems~\cite{Wriggers2012mortar,TEMIZER2014interiorpoint}.

In this paper we use IGA to implement a geometrically exact three-dimensional beam based on the Cosserat rod theory, hence capable of arbitrary large rotations and displacements~\cite{LINN2011cosserat}. Non-linear geometric effects in beams have been studied extensively in computational mechanics and different approaches have been put forward: for example a straightforward method still used nowadays is based on reusing linear finite elements developed for infinitesimal-strain conventional beams of Euler-Bernoulli or Timoshenko type, by updating a local corotational reference frame that can have large displacements and rotations~\cite{Crisfield1990corotational,Felippa2005unified}. In contrast to this, but at the cost of a more sophisticated formulation, the \textit{geometrically exact beam model}, also referred to as Simo-Reissner beam model, draws on the theory of 1D Cosserat continua \cite{cosserat1909} and leads to a general formulation that makes no assumption on the amount of rotations and that allows finite strains, including shear and torsion~\cite{Reissner1973gebt,Antman1974,Simo1985finitestrain}.
%Although some beam theories that consider out-of-plane and in-plane deformations of the cross-sections, we will make the assumption of rigid sections.
The Cosserat rod theory does not require that cross-sections must be orthogonal to the tangent of the centerline, hence it can account for shear effects similarly to the Timoshenko beam theory, of whom it can be considered a generalization to the 3D case~\cite{Timoshenko1921Beams}. In fact, Reissner, Kirchhoff-Love, Timoshenko and Euler-Bernoulli beams can be interpreted as special cases of Cosserat rods.

Many references on IGA-based beams can be found in literature: for instance IGA for Kirchhoff-Love and Euler-Bernoulli beams are discussed in~\cite{GRECO2013kirchhoffbeams} and~\cite{Weeger2013euler}, straight and curved 2D Timoshenko beams are discussed in~\cite{BOUCLIER2012thick}, to name a few. Similarly to FE beams, numerical locking artifacts can affect IGA beams when shear is taken into account. In this context, a collocation approach was used in~\cite{Beiraodaveiga2012IGA,Auricchio2013IGA} to obtain a locking-free spatial IGA beam.

Contact between three-dimensional rods has been dealt in literature by various authors. Among the first results, in \cite{wriggers1997} a frictionless point-wise contact  formulation was developed between pairs of cables with circular section, later extended to the frictional case \cite{wriggers2000}. The problem of contact between beams of rectangular section, that might generate multiple contact points between the edges, has been discussed in \cite{Wriggers2002,litewka2002frictional}. The issue of self-contact between beams has been studied in \cite{chamekh2009,GayNeto2015}, a topic of great importance if one needs to simulate problems such as the tightening of knots, knitting machines or winches.
Most contact models in literature are based on point-based formulations where the position of contact points is obtained by solving local sub-problems of distance minimization between curved geometries \cite{konyukhov2010}. However this approach is near singular when beams intersect with small contact angles: in \cite{meier2016,meier2017} an efficient method has been proposed that can solve this issue.
In \cite{weeger2018frictional} the simulation of frictional contact and self collision is discussed for a class of efficient IGA-based Cosserat rods.

Similarly to~\cite{Weeger2017IGA}, we address the case of a three-dimensional Cosserat rod of high generality, hence accounting for shear, torsion, geometric and material non-linearity. As an alternative to penalty-based beam contact formulations as presented in~\cite{Weeger2017361}, we express it with a formalism that fits well in a time stepper for non-smooth multibody dynamics, using the theory of Differential Variational Inequality (DVI). 
To our knowledge, this is the first time an IGA formulation has been used in the framework of DVI non-smooth dynamics; the benefit being the fact that DVI formulations provide a robust and stable way to simulate contact problems even with large time-steps and many simultaneous contacts.

Early work on non-smooth dynamics can be traced back to the seminal research of Jean-Jacques Moreau on measure-differential inclusions, a special type of DVI \cite{moreau88,Moreau1987}. The non-smooth nature of dynamical problems originates from various phenomena, most often it is a consequence of the Coulomb friction model and impulsive collisions at contact points; instead than regularizing the contact forces -a conventional method that leads to smooth but stiff differential problems- Moreau proposed to embed such set-valued force laws directly in the formulation. This can be done at the cost of assuming that velocities can be discontinuous, hence accelerations are considered as distributions of vector signed measures. Since then, different authors contributed to the field of non-smooth dynamics, see for example \cite{stew98,AniHa02,pfeiffer}. 
The robustness and stability of DVI methods motivated their use in scenarios that involve millions of simultaneous contacts, such as granular dynamics \cite{NegTasLeveraging2011}, or in contexts that require real-time performance and large time steps, such as robotics and virtual reality \cite{bender2014interactive}. Most time stepping schemes for DVI require the solution of a variational inequality (VI) per each time step: early approaches were based on fixed point iterations that both represented a major computational bottleneck and performed badly in presence of finite elements, but recent researches address both issues and stimulate novel applications of DVI methods also to flexible structures \cite{Mazhar2015,heyn2013,MANGONI2018351}.

In order to test our geometrically-exact rod formulation, we implemented it using the C++ language and we embedded it in our CHRONO open source simulation library~\cite{chrono2013}. We used this library to simulate IGA beams in a broader context involving rigid body dynamics, finite elements, motors, actuators and non-smooth frictional contacts: tests presented at the end of the paper show that the geometrically-exact IGA beam represents an efficient, compact and general way to simulate complex problems involving deformable beams.

% citare: 
%~\cite{Auricchio2013IGA} (Locking-free isogeometric collocation methods for spatial Timoshenko rods) CITATO
%~\cite{Weeger2017IGA} (Isogeometric collocation methods for Cosserat rods) CITATO
%~\cite{Timoshenko1921Beams} CITATO
%~\cite{Wriggers2012mortar} (large deformation contact w.mortar and augmented Lagrangian) CITATO
%~\cite{TEMIZER2014interiorpoint} (interior point method for IGA contact} CITATO
%\cite{LINN2011cosserat} (Cosserat rod) CITATO
%\cite{Lang2011CosseractRods} (Cosserat rot, quaternions) CITATO
%\cite{Reissner1973gebt} (large displacement curved beams, GEBT) CITATO
%\cite{Antman1974} (kirchoff, GEBT) CITATO
%\cite{SIMO1985} (GEBT) CITATO
%\cite{GRECO2013kirchhoffbeams} (IGA kirchoff-love beams} CITATO
%\cite{Weeger2013euler} (IGA Euler-bernoulli) CITATO

\section{B-Splines and NURBS}

Since our beam model draws on Basis splines (B-Splines) or Non-Uniform Rational B-Splines (NURBS), basic concepts about these functions are introduced here.

\subsection{B-Splines}

A B-Spline of order \emph{p} is a piecewise polynomial function of degree \emph{p-1} in a parametric variable $\tau \in \mathbb{R}$ (a curvilinear abscissa).

We introduce a set of $n+1$ control points $\avect{x}_i \in \mathbb{R}^3 \quad i = 0...n$ and a set \emph{n+p+1} of non-decreasing breaking points defining a \emph{knot vector} $\avect{T} =(\tau_0, \tau_1... \tau_{n+p})$.

\paragraph{Basis Functions} $N_{i,p}$ is an order \emph{p} and \emph{p-1} degree \emph{Basis Function} on the \emph{i-th} knot of the B-Spline and it is recursively defined as follows:
\begin{equation}
\label{eq:BSplineBasis_start}
N_{i,1}(\tau) =\begin{cases}
    1 & \text{for}\qquad \tau_i \leq \tau \leq\ \tau_{i+1}\\
    0 & \text{otherwise}
\end{cases}
\end{equation}
And, for $p>1$:
\begin{equation}
\label{eq:BSplineBasis_iter}
N_{i,p}(\tau) = \frac{\tau-\tau_i}{\tau_{i+p-1}-\tau_i}N_{i,p-1}(\tau) + \frac{\tau_{i+p}-\tau}  {\tau_{i+p}-\tau_{i+1}}N_{i+1,p-1}(\tau) 
\end{equation}
Basis Function are a partition of unity: it means that $\sum_{i=0}^nN_{i,p}(\tau)=1 \quad \forall \tau \in [\tau_o, \tau_n]$. 
The span of Basis Function increases with the order \emph{p}. \newline
Is useful to remind the derivative of the Basis Function with respect to the parametric variable since it is frequently used:
\begin{equation}
\label{BSplineBasis_derivative}
\mathring{N}_{i,p}(\tau) = \frac{dN_{i,p}(\tau)}{d\tau} = \frac{p-1}{\tau_{i+p-1}-\tau_i}N_{i,p-1}(\tau) - \frac{p-1}{\tau_{i+p}-\tau_{i+1}}N_{i+1,p-1}(\tau) 
\end{equation}
\paragraph{B-Splines formulation} A B-Spline is a linear combination of control points $\avect{x}_i$ and Basis Functions $N_{i,p}(\tau)$
\begin{equation}
\label{eq:BSpline}
\avect{r}(\tau) = \sum_{i=0}^n \avect{x}_i N_{i,p}(\tau) \qquad\qquad n \geq p-1
\end{equation}
Given the number of control points \emph{n+1} and the order of the curve $p$ the number of knots is $n+p+1$, which means there are more knots than control points, so some knots can be coincident on the same control point. When knots are distinct the first $p-1$ derivatives are continuous; when $r$ nodes are coincident, only the first $p-r$ derivatives are continuous. In general, control points do not lie on the curve. When $p$ knots are coincident, the spline passes through the control point with $C^0$ continuity.

\subsection{NURBS}

Non-Uniform Rational B-Splines introduce additional weights ${w_i}>0$, $(i = 1...n)$, so that \emph{rational} basis $R_{i,p}(\tau)$ are used in place of $N_{i,p}(\tau)$:
\begin{equation}
\label{eq:NURBS_Basis}
R_{i,p}(\tau) = \frac{N_{i,p}(\tau)w_i}{\sum_{j=1}^n N_{j,p}(\tau)w_j}
\end{equation}
NURBS have the same properties listed for B-Splines: in particular for $w_i = 1 \; \forall i$, NURBS reduce to B-Splines. In addition, by using proper weights and few coarse control points, NURBS allow the exact (not approximated) representation of conic sections like circles and ellipses; this is a relevant feature because canonical primitives in CAD models are most often built from conical sections.

% THE SECTION ON ROTATIONS AND LIE GROUPS HAS BEEN 
% REMOVED.
% JUST REMOVE THE COMMENT BEFORE \input{} TO GET IT BACK.
%
%\input{"chapter_rotations.tex"}

\section{Kinematics of Cosserat rods}

Our implementation of IGA beams draws on the Cosserat rod theory.
This implies that the rotation of the beam section is independent 
from the centerline position. This differs from the Euler-Bernoulli or
Kirchhoff beam theory which implies that sections remain orthogonal to the centerline,
an assumption that holds only for thin beams, because shear effects
cannot be modelled. 
%On the other hand the Timoshenko theory, at the cost of introducing rotational degrees of freedom per each node, allows also to model shear effects in thick beams.

\subsection{Configuration}

A Cosserat beam is represented by the line of its mass centroids (\emph{centerline}), which is described by a curve parameterized~\cite{Weeger2017IGA} by a curvilinear abscissa $s \in [0,T] $:
%
\begin{equation}
	\avect{r} = \avect{r}(s) \quad \in \mathbb{R}^3
\end{equation}

Section rotations $\amatr{R}$ are parameterized by  the same curvilinear abscissa $s \in [0,T] $ (even though not mandatory~\cite{Auricchio2013IGA} ), assuming that the $X$ axis of the $\amatr{R}$ frame represents the normal of the beam section, and $Y$ and $Z$ 
axes represent the height and width directions, respectively:
%
\begin{equation}
	\amatr{R} = \amatr{R}(s) \quad \in \mathsf{SO}(3)
\end{equation}

That is, at some point $s_a$ along the curve we have independent positions and rotations: 
$\avect{r}(s_a)$, $\amatr{R}(s_a)$.

We remark a first source of complication: $\amatr{R}$ is a 3D rotation matrix, that is a 3x3 orthogonal matrix (hence
the \emph{special orthogonal group} $\mathsf{SO}(3)$ in the terminology of Lie groups). Its 3x3 elements are not independent as $\amatr{R}\amatr{R}^T=\amatr{I}$ must hold: in general it is better to avoid using all nine elements of such matrix and use quaternions or rotation angles to parameterize rotation frames~\cite{Lang2011CosseractRods}.

\subsection{Strains}

Following~\cite{Simo1986finitestrain}, we can express  
$\underline{\avect{\epsilon}}$ and $\underline{\avect{\kappa}}$, hereafter called translational strains and rotational strains, respectively. 
% Simo86 Table 2
The translational strain $\underline{\avect{\epsilon}}$ is:
%
\begin{equation}
	\underline{\avect{\epsilon}} = \amatr{R}^T \avect{r}' - \avect{e_x}
	\label{eq:epsilon}
\end{equation}
%
where we introduced $\avect{e_x} = \{1,0,0\}$ and we introduced the \emph{curve gradient} 
%
\begin{equation}
\avect{r}' = \frac{d\avect{r}}{ds}. 
\end{equation}

Note that for an
undeformed beam we have $\amatr{R}^T\avect{r}' = \{1,0,0\}$, that is we assume that $s$ 
is a uniform arc-length parameterization. Is other words, at the initial state,
the curvilinear length of a curve from $s=a$ to $s=b$, i.e. $L=\int^b_a ||\avect{r}'||ds$, is exactly $b-a$. Usually, 
however, splines are not parameterized with uniform arc-length abscissa, because a 
generic abscissa $\tau$ is used instead. If so, we have  
%
\begin{equation}
\avect{r}' = \frac{d\avect{r}}{d\tau} \frac{d\tau}{ds} = \mathring{\avect{r}} J^{-1}
\end{equation}
%
where a jacobian is defined as $J_{s\tau} = \frac{ds}{d\tau}$, and $\mathring{\avect{r}} =\frac{d\avect{r}}{d\tau}$ is the parametric derivative (something that is quite easy to compute when dealing with splines).

During an initialization phase, when the beam is still undeformed, values of jacobians $J_{s\tau}$ are computed and stored in memory for all the integration points - they will be used later, when computing the internal forces.

The rotational strain  $\underline{\avect{\kappa}}$ is computed as~\cite{Weeger2017IGA}:
%
\begin{equation}
	\tilde{\underline{\avect{\kappa}}} = \amatr{R}^T \amatr{R}'
\end{equation}
%
Here $\amatr{R}' = \frac{d\amatr{R}}{ds}$ , while the tilde operator means:
%
\begin{equation}
	\tilde{\underline{\avect{\kappa}}} = 
	\left[  
	\begin{array}{ccc}
	 0 & -\kappa_z & \kappa_y \\
	 \kappa_z & 0 & -\kappa_x \\
	 -\kappa_y & \kappa_x & 0 
	\end{array}
	\right]
\end{equation}
%
so it is obvious that one can build $\underline{\avect{\kappa}}$ from $\tilde{\underline{\avect{\kappa}}}$ or vice versa.

Finally, in case the beam starts in an initially-curved configuration, initial values $\underline{\avect{\epsilon}}_0$ and $\underline{\avect{\kappa}}_0$ can be computed using the expression above, then one would compute effective strains with:
%
\begin{align}
	\avect{\epsilon} &= \underline{\avect{\epsilon}} - \underline{\avect{\epsilon}}_0 \\
	\avect{\kappa}   &= \underline{\avect{\kappa}} - \underline{\avect{\kappa}}_0 
\end{align}

\subsection{Constitutive model}

Once $\avect{\epsilon}$ and $\avect{\kappa}$ are computed, with some generic constitutive model one can compute their conjugate vectors, i.e. the generalized cut-section forces $\avect{n}$ and cut-section torques $\avect{m}$, henceforth called translational stresses and rotational stresses, respectively.
The most generic constitutive model is expressed by a non-linear function:
%
\begin{align}
  \left\{  
	 \avect{\epsilon},
	 \avect{\kappa}
	\right\} \in \mathbb{R}^6
	\rightarrow
	\left\{  
	 \avect{n},
	 \avect{m} 
	\right\} \in \mathbb{R}^6
\label{eq:mnconstitutive}
\end{align}

Note that $\avect{n}$ and $\avect{m}$ are expressed in the local coordinate system of the section frame: if one
needs the absolute values of such vectors, they can be computed easily, later, as 
% Simo 86 eq 2.3
\begin{align}
\avect{n}_a &= \amatr{R}\avect{n}\\ 
\avect{m}_a &= \amatr{R}\avect{m}
\end{align}

When structural damping is needed, instead of \eqref{eq:mnconstitutive} we use:
%
\begin{align}
  \left\{  
	 \avect{\epsilon},
	 \avect{\kappa},
	 \dot{\avect{\epsilon}},
	 \dot{\avect{\kappa}}
	\right\} \in \mathbb{R}^{12}
	\rightarrow
	\left\{  
	 \avect{n},
	 \avect{m} 
	\right\} \in \mathbb{R}^6
\label{eq:mnconstitutivewdamping}
\end{align}
%
by introducing the %
 strain derivative as 
$\dot{\avect{\epsilon}}$
and the curvature derivative as
$\dot{\avect{\kappa}}$.

See  Appendix A 
%(\ref{sec:constitutive}) 
for more details on the methods for computing generalized strains.

\subsection{Equilibrium}

The strong form for the equilibrium of the Cosserat beam is:
%
\begin{align}
\avect{n}'_a + \underline{\avect{n}}_a &= \avect{0}\\ 
\avect{m}'_a + \avect{r}'_a \times \avect{n}_a + \underline{\avect{m}}_a &= \avect{0}
\end{align}
%
where, if provided, 
$\underline{\avect{n}}_a$ is the external force distributed on the beam, and 
$\underline{\avect{m}}_a$ is the external torque distributed on the beam, both expressed in 
absolute coordinates. The subscript \emph{a} means we are using absolute coordinates for $ \avect{r}'$ too.

\section{Discretization with IGA elements}

\begin{figure}[ht]
\centering
\includegraphics[width=0.75\textwidth]{beam_cosserat.pdf}
\caption{Spline discretization of the beam with independent position and rotation fields.}
\label{figdiscretize}
\end{figure}

The beam is approximated with IGA. 
To this end we assume that the beam is represented as a B-spline whose control points are nodes, each i-th node being a coordinate system $\{\avect{x}_i, \amatr{R}_i\}$ with position $\avect{x}_i \in \mathbb{R}^3$ and rotation $\amatr{R}_i \in \mathsf{SO}(3)$, as shown in Fig.~\ref{figdiscretize}.
For the rotation, however, we choose to parameterize rotation matrices $\amatr{R}$ using unit-length quaternions $\avect{\rho} \in \mathbb{H}_1$, so the configuration of each node is a compact set of 7 scalars: $\{\avect{x}_i, \avect{\rho}_i\}$.
We recall the basic properties of quaternions:
$\avect{\rho}=\rho_{0}+\avect{i}\rho_{1}+\avect{j}\rho_{2}+\avect{k}\rho_{3}$ with
$\avect{i}^2=\avect{j}^2=\avect{k}^2=\avect{i}\avect{j}\avect{k}=-1$, 
often written succinctly
$\avect{\rho}=[\rho_{s},\avect{\rho}_{v}]$ to facilitate the expression of quaternion multiplication:
\begin{align} 
\avect{\tau}\avect{\rho} = [ \tau_{s}\rho_{s} - \avect{\tau}_{v}\cdot\avect{\rho}_{v}, \;
\tau_{s}\avect{\rho}_{v} + \rho_{s}\avect{\tau}_{v} + \avect{\tau}_{v} \times \avect{\rho}_{v} ]
	\label{eq:quatproduct}
\end{align}

If needed, one can always convert quaternions into rotation matrices, as $\amatr{R}_i = \amatr{R}(\avect{\rho}_i)$, using the following property:
\begin{align}
\amatr{R}(\avect{\rho}) = \left[
	\begin{matrix}
	 \rho_0^2 + \rho_1^2 - \rho_2^2 - \rho_3^2  &  2(\rho_1 \rho_2 - \rho_3 q_0)   &  2(\rho_1 \rho_3 + \rho_2 \rho_0)  \cr
	 2(\rho_1 \rho_2 + \rho_3 \rho_0)  &  \rho_0^2 - \rho_1^2 + \rho_2^2 - \rho_3^2  &   2(-\rho_1 \rho_0 + \rho_2 \rho_3) \cr
   2(\rho_1 \rho_3 - \rho_2 \rho_0)  &  2(\rho_1 \rho_0 + \rho_2 \rho_3)  & \rho_0 ^2 -\rho_1^2 -\rho_2^2 + \rho_3^2   
	\end{matrix}
	\right]
	\label{eq:quatmatrix}
\end{align}
%
% and a similar procedure can be used, viceversa, to compute $\avect{\rho}_i = \avect{\rho}(\amatr{R}_i)

Finally we denote the quaternion conjugate as $\avect{\rho}^*$, 
with $\avect{\rho}^*=\rho_{0}-\avect{i}\rho_{1}-\avect{j}\rho_{2}-\avect{k}\rho_{3}$,
such that $\amatr{R}(\avect{\rho}_i^*)\amatr{R}(\avect{\rho}_j) = \amatr{R}(\avect{\rho}_i)^T\amatr{R}(\avect{\rho}_j) $.

\subsection{System state vectors}

The state of the system contains velocities and angular velocities of each frame $\{ \dot{\avect{x}}_i, \avect{\omega}_i \}$. We consider $\avect{\omega}_i$ expressed in frame-local coordinates.

This means that the state of the system, for $n_n$ nodes, is given by the system-level configuration $\avect{q}$ and the system-state velocity $\avect{v}$ as: 
%
\begin{align}
\avect{q} &= \{ \avect{x}_1, \avect{\rho}_1, \avect{x}_2, \avect{\rho}_2, \ldots, \avect{x}_{n_n}, \avect{\rho}_{n_n} \} \\ 
\avect{v} &= \{ \dot{\avect{x}}_1, \avect{\omega}_1, \dot{\avect{x}}_2, \avect{\omega}_2, \ldots, \dot{\avect{x}}_{n_n}, \avect{\omega}_{n_n} \}
\end{align}
%

In both statics and dynamics analysis it happens that one has to update by updating the $\avect{q}$ configuration by applying some computed increment. For instance, in a linear static problem one has $\amatr{K} \delta\avect{q}=\avect{b}$ and $\avect{q}_{new} = \avect{q} + \delta\avect{q}$. Also numerical methods for DAEs and ODEs in dynamics proceed by performing updates on the configurations at each time step. The problem is that, within our state $\avect{q}$, positions can be updated with straight sums as 
$\avect{x}_{i,new} = \avect{x}_i + \delta\avect{x}_i$, but quaternions cannot be updated as easily. In fact doing a straight sum with a $\delta\avect{\rho}_{i} \in \mathbb{H}_1$ as in $\avect{\rho}_{i,new} = \avect{\rho}_{i} + \delta\avect{\rho}_{i}$ could invalidate the unit-length of $\avect{\rho}_{i,new}$. A more rigorous solution is to do incremental updates of quaternions using the exponential map. This requires some basic concepts of Lie algebras.

Recalling concepts of differential geometry, for an element $\amatr{R}$ in Lie group $\mathsf{SO}(3)$ and an element $\delta\amatr{\Theta}$ in the corresponding Lie algebra $\mathfrak{so}(3)$, one has
%
\begin{align}
\amatr{R}      &= \mathrm{exp}(\delta\amatr{\Theta}) \label{Rexp}\\
\delta\amatr{\Theta} &= \mathrm{log}(\amatr{R}) \label{Rlog}
\end{align}
%
One can extract the three dimensional rotation pseudovector $\delta\avect{\theta}$ from $\delta\amatr{\Theta}$ via $\delta\avect{\theta} = \text{axis}(\delta\amatr{\Theta})$, recalling that $\delta\amatr{\Theta} =  \mathrm{skew}(\delta\avect{\theta}) = \delta\tilde{\avect{\theta}}$. 
For our purposes, $\delta\avect{\theta}$ can be considered a (not necessarily infinitesimal) incremental rotation; for example in a time stepper it could be $\delta\avect{\theta} =  \avect{\omega} dt$.

For sake of compactness and performance, we avoid updating $\amatr{R}$ and we rather exploit the fact that $\mathbb{H}_1$ is a double cover of $\mathsf{SO}(3)$. Just like in \eqref{Rexp} and \eqref{Rlog}, an exponential map links $\mathbb{H}_1$, (unit  quaternions), and its Lie algebra $\mathsf{Im}(\mathbb{H})$ of \emph{pure quaternions} $\delta\avect{\rho} = [0,\delta\avect{\theta}]$:
%
\begin{align}
\avect{\rho}      &= \mathrm{exp}(\delta\avect{\rho}) \label{eq:rexp}\\
\delta\avect{\rho} &= \mathrm{log}(\avect{\rho})   \label{eq:rlog}
\end{align}
%
We can define two operators to convert pure quaternions from and to rotation pseudovectors: 
$\delta\avect{\theta}= \text{imag}([0,\delta\avect{\theta}])$,
and
$[0,\delta\avect{\theta}]= \text{pure}(\delta\avect{\theta})$.
The exponential map \eqref{eq:rexp} can be explicitly computed from $\delta\avect{\theta}$ as $\avect{\rho} = \exp (\text{pure}(\delta\avect{\theta}))$ using the following closed-form expression:
%
\begin{align}
\avect{\rho} =
\exp ([0, \delta\avect{\theta}]) = 
\left\{ 
\cos\left(\frac{||\delta\avect{\theta}||}{2}\right), 
\frac{\delta\avect{\theta} }{ ||\delta\avect{\theta}|| } \sin\left( \frac{||\delta\avect{\theta}||}{2}\right) 
\right\} \label{eq:qexponential}
\end{align}

This said, in our code we can work with increments in the following form:
%
\begin{align}
\delta\avect{q} &= \{ \delta\avect{x}_1, \delta\avect{\theta}_1, \delta\avect{x}_2, \delta\avect{\theta}_2, \ldots, \delta\avect{x}_{n_n}, \delta\avect{\theta}_{n_n} \} 
\end{align}
%
and, where one has to perform the incremental update to find $\avect{q}_{new}$, the rotational parts are incremented pre-multiplying the quaternion by an exponential map, as:
%
\begin{align}
\avect{x}_{i, new}    &= \avect{x}_i + \delta\avect{x}_i \label{eq:xinc}
\\ 
\avect{\rho}_{i, new} &= \avect{\rho}^{\delta}_{i} \avect{\rho}_{i}
\label{eq:rinc}
\end{align}
%
where one computes 
$\avect{\rho}^{\delta}_{i} = \exp ([0, \delta\avect{\theta}_{i}]) $ using \eqref{eq:qexponential}, then computes the product $\avect{\rho}^{\delta}_{i} \avect{\rho}_{i}$ using \eqref{eq:quatproduct}.
More succinctly, the incremental update is a map
$\avect{q}_{new}=\Lambda(\avect{q},\delta\avect{q})$.

In the following, for expressing formulas in a easier way, we group all the translational degrees of freedom and all the rotational degrees of freedom in two separate vectors $\avect{q}_x$ and $\avect{q}_{\rho}$, and we do the same for the increments 
$\delta\avect{q}_x$ and $\delta\avect{q}_{\rho}$:
%
\begin{align}
\avect{q}_x &= \{ \delta\avect{x}_1, \delta\avect{x}_2, \ldots, \delta\avect{x}_{n_n} \} \\
\avect{q}_{\rho} &= \{ \avect{\rho}_1, \avect{\rho}_2, \ldots, \avect{\rho}_{n_n} \} \\
\delta\avect{q}_x &= \{ \delta\avect{x}_1, \delta\avect{x}_2, \ldots, \delta\avect{x}_{n_n} \} \\
\delta\avect{q}_{\rho} &= \{ \delta\avect{\theta}_1, \delta\avect{\theta}_2, \ldots, \delta\avect{\theta}_{n_n} \}
\end{align}

\subsection{Interpolation}

In IGA, the role of FEA shape functions is done by Basis Functions of the spline.
For a given knot abscissa $\tau$ along the spline, one can compute interpolated position and gradient using Basis Functions $N_i(\tau)$ and their derivatives $\mathring{N}_i(\tau) = \frac{dN_i(\tau)}{d\tau}$, that is:
%
\begin{align}
\avect{r}(\tau)  &= \sum_i N_i(\tau) \avect{x}_i  \\
\avect{r}'(\tau) &= \frac{d\avect{r}(\tau)}{d\tau} J^{-1}_{s\tau}    \\
                 &= \sum_i \mathring{N}_i(\tau) \avect{x}_i  J^{-1}_{s\tau} 
\end{align}

Once this is computed, $\underline{\avect{\epsilon}}(\tau)$ can be computed with \eqref{eq:epsilon}.

Similarly, one would be tempted to interpolate the rotation at a given abscissa $\tau$ using a similar method, however applying directly $\avect{\rho}(\tau)  = \sum_i N_i(\tau) \avect{\rho}_i$ is not possible because the linear sum of unit quaternions does not represent, in general, a rotation. In fact, the spline interpolation of quaternions is still a debated problem and exact solutions presented in literature, such as \cite{kim1995general,kim1995c}, often lead to complex formulas that add complication and computational overhead. As a simplification we introduce an auxiliary co-rotated system $\underline{\avect{\rho}}$, for instance an average of rotations of the closest nodes, and we interpolate rotations by doing a weighted sum of rotation pseudovectors %$\text{imag}(\log(\underline{\avect{\rho}}^* \avect{\rho}_i))$, 
later recast as a quaternion using the exponential map:
%
\begin{align}
\avect{\rho}(\tau)  &= \underline{\avect{\rho}} \exp \left( \text{pure} \sum_i N_i(\tau) \text{imag}(\log( \underline{\avect{\rho}}^* \avect{\rho}_i)) \right) 
\end{align}
%
Then, the $\amatr{R}$ matrix can be computed as $\amatr{R}(\avect{\rho})$.
%
Also, the curvature $\avect{\kappa}$ is computed as:
%
\begin{align}
%\avect{\kappa}(\tau)  &=  \amatr{R}^T \sum_i J^{-1}_{s\tau} \mathring{N}_i(\tau) \text{imag}(\log(\avect{\rho}_i))  
\avect{\kappa}(\tau)  &=   \sum_i J^{-1}_{s\tau} \mathring{N}_i(\tau) \text{imag}( \log(\avect{\rho}(\tau)^* \avect{\rho}_i))
\end{align}

These allow the computation of local stresses $\avect{m}(\tau)$ and $\avect{n}(\tau)$ according to \eqref{eq:mnconstitutive}.
If a constitutive material includes damping effects, one can compute also
$\dot{\avect{\epsilon}}(\tau)= \amatr{R}^T \sum_i \mathring{N}_i(\tau) \dot{\avect{x}}_i  J^{-1}_{s\tau}$
and 
$\dot{\avect{\kappa}}(\tau) = \amatr{R}^{T} \sum_i \mathring{N}_i(\tau) \amatr{R}_i \avect{\omega}_i  J^{-1}_{s\tau}$ to be used in \eqref{eq:mnconstitutivewdamping}.

\subsection{Gauss quadrature}

As in most FEA frameworks, also in IGA the backbone of the process is the computation of three main ingredients: the vector of internal forces $\avect{f}_{int}$, the tangent stiffness matrix $\amatr{K_t}$, and the mass matrix $\amatr{M}$. Once these are computed, one can solve, for instance, linear elastic problems as $\amatr{K_t}\delta\avect{q}=\avect{f}_{ext}$, non linear elastic problems that iterate on $\amatr{K_t}\delta\avect{q}=\avect{f}_{ext}+\avect{f}_{int}$ up to convergence, explicit dynamic integration as in $\amatr{M}\ddot{\avect{q}}=\avect{f}_{ext}+\avect{f}_{int}$ and so on. 
Therefore, in the following, we present the procedure for computing these terms, starting from the most relevant part, that is the vector of internal forces $\avect{f}_{int}$.

Here we focus on computing $\avect{f}_{int}$ via typical Gauss integration. We remark that an appropriate choice of integration points must be used in order to avoid shear locking phenomena, to this end we used selective reduced integration requiring fewer quadrature points than standard integration techniques. Efficient choice of Gauss points is discussed by various authors, for instance~\cite{Hughes2010301,Hillman2015521,Adam2015732}. In fact, a more recent and efficient approach to IGA consists in using collocation instead of Gauss integration~\cite{Auricchio2013IGA,Marino2017546,weeger2018frictional}.
%We interface this Cosserat rod implementation with our C++ to leverage its contact simulation capabilities. However, it does not offer support for collocation methods.

Also, we use the same idea of FEA of computing internal forces and matrices on a per-element basis, where later all element terms will be assembled/overlapped in global system-level matrices and vectors. In this sense, in our embodiment the j-th element is the j-th span of the B-spline, so for each element we compute 
$\avect{f}_{int,j}$, and later all those per-element internal forces are assembled in a single vector
\footnote{Differently from FEA, in IGA one could compute the system-level vector without passing through intermediate per-element vectors, but here for clarity we develop our formulation on a per-element basis}.

%The potential energy of the j-th span from $s=a$ to $s=b$ is:
%
%\begin{align}
%\Pi = \frac{1}{2} \int_{a}^{b} \left( \avect{\epsilon} \avect{n} + \avect{\kappa}\avect{m} \right) ds
%- \Pi_{ext}
%\end{align}
%
Over the j-th span from $s=s_A$ to $s=s_B$ the weak form of the equilibrium reads:
%
\begin{align}
\delta\Pi = 
 \int_{s_A}^{s_B} \left( \delta\avect{\epsilon}_a \avect{n}_a + \delta\avect{\kappa}_a \avect{m}_a \right) ds 
-\int_{s_A}^{s_B} \left( \delta\avect{x}_a \underline{\avect{n}}_a + \delta\avect{\theta}_a\underline{\avect{m}}_a \right) ds 
\end{align}
%
Recalling \eqref{eq:epsilon}, evaluating its increment and referring it to the global coordinates:
%reissner: 12a/ 12b from 8 a/b
\begin{align}
\delta\avect{\epsilon}_a &= \delta\avect{r}'_a - \delta\avect{\theta}_a \times \avect{r}'_a  
\end{align}

So, at a point at abscissa $\tau$ along the spline, $\delta\avect{\epsilon}$ and $\delta\avect{\kappa}$ can be expressed using interpolation of the B-spline given the following terms:
%
\begin{align}
\delta\avect{r}'_a       &= \sum_i J^{-1}_{s\tau} \mathring{N}_i \delta\avect{x}_i    \\
\delta\avect{\kappa}_a   &= \sum_i J^{-1}_{s\tau} \mathring{N}_i \amatr{R}_i \delta\avect{\theta}_i \\
\delta\avect{\theta}_a   &= \sum_i      {N}_i \amatr{R}_i \delta\avect{\theta}_i
\end{align}

We rewrite the weak equilibrium introducing the vector of internal forces
%
\begin{align}
\delta\avect{q}^T_j \avect{f}_{int,j} -  \delta\avect{q}^T_j \avect{f}_{ext,j} = \avect{0} 
\end{align}

where $\avect{f}_{int,j}$ and $\avect{f}_{ext,j}$ are the internal and external forces acting on the $j$-th node.

Looking at the expressions above, one can write the expression of the contribution from the $i$-th control node to $\avect{f}_{int,j}$ as an integral in $s$ arc-length coordinate:
%
\begin{align}
\avect{f}_{int,j,i} = \int_{s_A}^{s_B} 
 \left[  
	\begin{array}{cc}
	 J^{-1}_{s\tau} \mathring{N}_i  \amatr{I}  &  N_i \amatr{R}_i \tilde{\avect{r}}'\\
	 0                           & J^{-1}_{s\tau} \mathring{N}_i \amatr{R}_i 
	\end{array}
	\right]^T
	\left[  
	\begin{array}{c}
	 \avect{n}_a  \\
	 \avect{m}_a  
	\end{array}
	\right]
	ds
\end{align}

The integral is computed with a sum over $n_{gp}$ Gauss points, so we introduce Gauss point weights $w_k$, jacobians $J_{\tau\zeta} = \frac{d\tau}{d\zeta}$ where $\zeta \in [-1,+1]$ is the coordinate used for Gauss quadrature. Because of change of coordinates, we also have to multiply the integrand by $J_{s\tau}$, so some jacobians will simplify as $J^{-1}_{s\tau} J_{s\tau}=1$. 
Finally one has:
%
\begin{align}
\avect{f}_{int,j,i} = \sum_k^{n_{gp}} 
 w_k J_{\tau\zeta}
 \left[  
	\begin{array}{cc}
	 \mathring{N}_i \amatr{I}   & J_{s\tau} N_i \amatr{R}_i \tilde{\avect{r}}'\\
	 0                     & \mathring{N}_i \amatr{R}_i 
	\end{array}
	\right]^T
	\left[  
	\begin{array}{c}
	 \avect{n}_a  \\
	 \avect{m}_a  
	\end{array}
	\right]
	\label{fint}
\end{align}

Here all the terms of the integrand (namely, $N_i$, $\mathring{N}_i$, $J_{s\tau}$, $\tilde{\avect{r}}'$, $\avect{n}_a$ and $\avect{m}_a$) must be computed for the k-th knot abscissa $\tau_k =  \left(\frac{s_B-s_A}{2} \zeta_k + \frac{s_B+s_A}{2}\right)$ where $\zeta_k$ is one of the $n_{gp}$ tabulated abscissas for Gauss integration, i.e. coupled to the corresponding weight $w_k$. By the way one can see that $J_{\tau\zeta} = \frac{s_B-s_A}{2}$.

We remark that control points will affect neighbouring spans (elements) hence  terms $\avect{f}_{int,j,i}$ from different spans will overlap and must be summed to obtain $\avect{f}_{int,j}$.

The number of nodes influencing a single span increases with the spline order. IGA spline Basis Functions can be considered like shape functions affecting more knot spans, combining each other in the shared segments of support. For this reason an higher spline order leads to more overlapped Basis Functions, while increasing the element order in FE analysis adds nodes within the element but internal nodes do not affect adjacent elements. This leads to a band-diagonal stiffness matrix whose band width increases with the spline order, as opposed to traditional FE elements characterized by a block-diagonal stiffness matrix whose blocks overlap only at the boundary nodes.

\section{Stiffness, mass and damping matrices}

For the time integration of the equations of motion, the mass matrix is needed. If implicit integration schemes are used, also the tangent stiffness and damping matrices are required.

In our implementation we use a lumped mass matrix where each $i$-th IGA control point has an atomic mass $m_i$ and a tensor of inertia $I_i \in \mathbb{R}^{3\times3}$. This has a computational advantage with respect to the explicit quadrature of the beam shape, because the tensor of inertia $I_i=\text{diag}(I_{xx_i},I_{yy_i},I_{zz_i})$ is diagonal and constant as a consequence of the fact that we assumed angular velocities $\avect{\omega}_i$ to be expressed in the local coordinate system. Values of $m_i, I_{xx_i}, I_{yy_i}, I_{zz_i}$ can be computed from the geometry of the section and from the density of the material.

The tangent stiffness matrix and the damping matrix are computed by numerical differentiation of $\avect{f}_{int}$.
%
\begin{align}
\amatr{K_t} 
= -\nabla_{\avect{q}} \avect{f}_{int} 
= - \left[  \frac{\partial\avect{f}_{int}}{\partial\avect{q}} \right], 
\quad
\amatr{R_t} 
= -\nabla_{\avect{v}} \avect{f}_{int} 
= - \left[  \frac{\partial\avect{f}_{int}}{\partial\avect{v}} \right]
	\label{KRmatr}
\end{align}

Since the procedure for computing \eqref{fint} is quite simple, the computational overhead is not excessive with respect to an analytic evaluation of the matrix, yet it has the useful property of being of general validity even when using black-box nonlinear constitutive laws for materials. For a practical implementation, for the sake of high performance, one should exploit the sparsity pattern of $\amatr{K_t}$. In detail, the numerical differentiation operates on a per-span basis, to obtain per-span stiffness matrices $\amatr{K_t}_{j}$ that are summed afterward to obtain the large and sparse $\amatr{K_t}$ system-level matrix.

\section{Contacts}

\begin{figure}[ht]
\centering
\includegraphics[width=0.55\textwidth]{beam_cosserat_collision.pdf}
\caption{Sampling the beam with contact shapes (proxies) for fast collision detection.}
\label{figcontact}
\end{figure}

We developed a framework for simulating contacts between moving parts, either beams and other objects (rigid bodies, other finite elements, etc.).

When targeting complex scenarios with thousands of elements, we experienced that an efficient approach is to sample the rod with simple collision geometries that can be easily handled by fast collision algorithms. For example a rod with a circular section can be approximated with invisible collision spheres $S_i$, evenly spaced along the beam centerline, at given parametric abscissas $\tau_i$, as shown in Fig.~\ref{figcontact}. Those proxies are passed to the collision engine that computes candidate collision points at each time step: to this end we use a Sweep-And-Prune (SAP) broad-phase optimization that allows a near linear-time complexity even in case of many shapes, and a Gilbert-Johnson-Keerthi~(GJK) algorithm for the narrow-phase.

The narrow-phase returns a variable number of candidate contact pairs, each represented by two points $\avect{P}$ and $\avect{Q}$ belonging to two colliding shapes, and a coordinate system $\amatr{R_c}$ whose $x$ axis represent the contact normal. We assume that the contact distance $\Phi=|\avect{P}-\avect{Q}|$ is differentiable. The distance vector, in the contact coordinates, is $\avect{D}=\amatr{R_c}^T(\avect{P}-\avect{Q})$.

In order to write the equation of motion in the following section, we need to express the relative speed of the two points, $\avect{u}=\amatr{R_c}^T(\dot{\avect{P}}-\dot{\avect{Q}})$ as a function of generalized coordinates $\avect{v}$.

Assuming point $\avect{P}$ belongs to a beam, we can recover $\tau_P$, the abscissa at which the colliding shape was sampled.
This allows the immediate 
\footnote{An optional refinement can happen at this point: since the collision shape -a sphere, a capsule...- was only an approximation of the smoothly bent surface of the beam, one can iteratively correct the (discrete-sampled) contact abscissa in order to satisfy $\avect{P}(\tau_P)=\avect{r}(\tau_B)+\amatr{R}(\tau_P)\avect{p}_{c}$ until the position of the contact point in the beam section frame, $\avect{p}_{c}$, has perfectly zero off-section value: $p_{c,x}=0$}
computation of $\amatr{R}(\tau_P)$, the rotation of the section at the contact, and $\avect{p}_{c}$, the position of the contact point in the section coordinate system, as in $\avect{P}(\tau_P)=\avect{r}(\tau_P)+\amatr{R}(\tau_P)\avect{p}_{c}$.

Taking the time derivative, simplifying for $\dot{\avect{p}}_c=0$ and recalling the spline Basis Functions:
%
\begin{subequations}
\begin{align}
\dot{\avect{P}}(\tau_P) &=
 \dot{\avect{r}}(\tau_B) +
 \dot{\amatr{R}}(\tau_P) \avect{p}_c +
 \amatr{R}(\tau_P) \dot{\avect{p}}_c \\
 &=\sum_i{N_i(\tau_P)\dot{\avect{x}}_i } +
  \amatr{R}(\tau_P) \tilde{\avect{\omega}}(\tau_P) \avect{p}_{c} + 0 \\
 &=\sum_i{N_i(\tau_P)\dot{\avect{x}}_i } 
  -\amatr{R}(\tau_P) \tilde{\avect{p}}_{c} \avect{\omega}(\tau_P) \\
 &=\sum_i{N_i(\tau_P)\dot{\avect{x}}_i } -
  \sum_i{N_i(\tau_P)\amatr{R}(\tau_P) \tilde{\avect{p}_{c}} \avect{\omega}_{i}} \label{eq:jaco}
\end{align}
\end{subequations}

At this point it is possible to write the contact speed of the $k$-th contact using generalized coordinates, 
$\avect{u}_k=\nabla_{\avect{v}}^T\amatr{D}_k \avect{v}$,
where the jacobian $\nabla_{\avect{v}}^T  \amatr{D}_k \in \mathbb{R}^{3 \times n_v}$ has a sparse structure with two blocks:
\begin{equation}
\nabla_{\avect{v}}^T\amatr{D_k} =
\left[\dots, \nabla_{\avect{v}}^T\amatr{D_k}_P, \dots, \nabla_{\avect{v}}^T\amatr{D_k}_Q, \dots \right].
\end{equation}
In detail, from \eqref{eq:jaco} one has
%
\begin{multline}
\nabla_{\avect{v}}^T\amatr{D_k}_P = 
 \Big[ 
  N_1(\tau_P) \amatr{R_c}^T,
  -N_1(\tau_P) \amatr{R_c}^T \amatr{R}(\tau_P) \amatr{\tilde{p_{c}}}, \\
  N_2(\tau_P) \amatr{R_c}^T, 
  -N_2(\tau_P) \amatr{R_c}^T \amatr{R}(\tau_P) \amatr{\tilde{p_{c}}}, \\
  \dots,
  N_n(\tau_P) \amatr{R_c}^T,
  -N_n(\tau_P) \amatr{R_c}^T \amatr{R}(\tau_P) \amatr{\tilde{p_{c}}}
 \Big]
 \label{eq:jacodetail}
\end{multline}
%
whereas an equivalent expression holds for $\nabla_{\avect{v}}^T\amatr{D_k}_Q$ if also $\avect{Q}$ belongs to a beam 
\footnote{For instance, if $\avect{Q}$ belongs to a rigid body, the corresponding jacobian block has a simpler expression: 
$\nabla_{\avect{v}}^T\amatr{D_k}_Q=
\left[ 
-\amatr{R_c}^T,  \amatr{R_c}^T \amatr{R_b} \amatr{\tilde{q_{c}}}
\right]$ where $\amatr{R_b}$ is the rotation of the body and $\avect{q}_{c}$ is the position of $\avect{Q}$ in body local coordinates.}.

As a special case, very thin beams have $||\avect{p_{c}}||\approx 0$ hence $-N_n(\tau_P) \amatr{R_c}^T \amatr{R}(\tau_P) \amatr{\tilde{p_{c}}}$ terms could be removed from the jacobian in sake of performance.

Finally, it is possible to group all $n_c$ contact jacobians in a system-level sparse matrix $\nabla_\avect{v}^T \avect{D} \in \mathbb{R}^{3n_c \times n_v}$.

\section{Time stepping and implementation}

In the most general setting, we assume that the system is made by a several IGA beams, as well as other finite elements and rigid bodies, and we assume that all their states are grouped in the system-level vectors  $\avect{q}$ and $\avect{v}$. Also, we assume that the configuration is subject to holonomic-rheonomic constraint equations $\avect{C}(\avect{q},t) = \avect{0}$. Differentiation of constraints leads to 
$\frac{\partial \avect{C}}{\partial \avect{q}} \frac{\partial \avect{q}}{\partial t} + \frac{\partial \avect{C}}{\partial t} = \avect{0}$, 
that is 
$\nabla_\avect{v}^T \avect{C} \avect{v} + \avect{C}_t = \avect{0}$
where
$\nabla_\avect{v}^T \avect{C} = \frac{\partial \avect{C}}{\partial \avect{q}} \Gamma(\avect{q})$ is the constraint jacobian, $\Gamma(\avect{q})$ is a linear operator for 
$\dot{\avect{q}} = \Gamma(\avect{q}) \avect{v}$, 
and $\avect{C}_t$ is a null term in case of scleronomic constraints, but can appear in case of time-dependant constraints (motors, imposed trajectories, etc.).

We perform time integration either with implicit integrators \footnote{HHT \cite{JayNegrut2007} or Newmark \cite{negrutGavrea2005} in our implementation, here not described for compactness.} for smooth Differential-Algebraic-Equations (DAE), or with a custom time stepper for non-smooth DVI.
The latter allows also an efficient treatment of contacts. In fact, if one wants to simulate contacts in the framework of conventional smooth dynamics, in most cases contact constraints would be dealt with regularization, i.e. penalty functions that would require very short time steps. On the other hand, the DVI approach embeds the set-valued non-smooth nature of contact forces directly in the formulation~\cite{negrut2018posing}.
At $l$-th time step we solve:
\begin{align}
    \begin{bmatrix}
    \amatr{H} & \nabla_\avect{v}  \avect{C} & \nabla_\avect{v} \avect{D}\\
    \nabla_\avect{v}^T \avect{C} & 0 & 0 \\
    \nabla_\avect{v}^T \avect{D} & 0 & 0 
    \end{bmatrix}
    \begin{pmatrix}
    \avect{v}^{(l+1)} \\
    \avect{\lambda}_C \\
    \avect{\lambda}_D \\
    \end{pmatrix}
    -
    \begin{pmatrix}
    \amatr{H}   \avect{v}^{(l)} + h(\avect{f}_{int}+\avect{f}_{est})\\
    -\frac{1}{h}\avect{C}(\avect{q},t) - \avect{C}_t\\
    -\frac{1}{h}\avect{\Phi}(\avect{q})
    \end{pmatrix}
    = 
    \begin{pmatrix}
    \avect{0}\\
    \avect{0}\\
    \avect{y}
    \end{pmatrix} \label{eq:dvi}
    \\
    \avect{y} \in \Upsilon^* \bot \avect{\lambda}_D \in \Upsilon  \label{eq:complementarity} 
    \\
    \avect{q}^{(l+1)}=\Lambda(\avect{q}^{(l)},h\avect{v}^{(l+1)}) \label{eq:increment}
\end{align}
%
where $\amatr{H}=\amatr{M}+h^2\amatr{K}_t+h\amatr{R}_t$ is a sparse matrix, $h$ is the time step, contact jacobians are computed as in \eqref{eq:jacodetail}, $\Upsilon = \oplus_i \Upsilon_i$ is the Cartesian product of all Coulomb friction cones generated by contacts, and 
$\Upsilon^* = 
\left \{\avect{\lambda}_D \in \Upsilon^*: \langle  \avect{\lambda}_D, \avect{z} \rangle \geq 0 \quad \forall \avect{z}\in \Upsilon  \right \}
= \oplus_i \Upsilon_i^*$ 
is its dual. Finally, \eqref{eq:increment} is updated with exponential maps as in \eqref{eq:xinc}-\eqref{eq:rinc}.

In absence of contacts, it can be demonstrated that the algorithm \eqref{eq:dvi}-\eqref{eq:increment} simplifies to a backward Euler linearly implicit integrator, of whom it inherits the same properties of first order accuracy in time and L-stability. If higher order accuracy or less numerical damping is needed, recent research demonstrated the possibility of second-order DVI time steppers at the cost of additional complexity in the implementation \cite{bruls2018constraints}.

The problem \eqref{eq:dvi}-\eqref{eq:complementarity} is a Second-Order Cone Complementarity Problem (SO-CCP) with unknowns $\avect{v}^{(l+1)}$, the speed at the end of time step, and $\avect{\lambda}$, the reaction impulses in contacts or bilateral constraints. The convex cone complementarity problem is also a special case of a Variational Inequality (VI), hence the DVI acronym~\cite{StPa08}.

The complementarity constraint \eqref{eq:complementarity} introduces a major numerical difficulty, however the monolithic SO-CCP \eqref{eq:dvi}-\eqref{eq:complementarity} can be solved efficiently using the primal-dual interior point method that we presented in~\cite{MANGONI2018351} and that we implemented in C++.

Although the DVI integrator \eqref{eq:dvi}-\eqref{eq:increment} is stable also when using large time steps, we remark that there are scenarios where the time step must be extremely short anyway, for example because one wants to simulate collisions between very thin beams, a case that requires small steps in order to avoid tunneling and aliasing effects in the computation of contact points; in such cases, the advantage of using the DVI is less evident.

Comparing this non-smooth DVI integrator to conventional smooth explicit integrators, one can see that DVI requires more computational effort per each time step, because of the SO-CCP overhead. On the other hand, explicit integrators require a much larger number of time steps to achieve stability, depending on how stiff is the contact formulation.

Comparing the DVI integrator to conventional smooth implicit integrators embedded with penalty formulations leads to similar conclusions: the SO-CCP solver overhead is still higher than the few linear system solves required in the Newton iterations of implicit steppers, but on the other side the total number of time steps can be smaller. Implicit solvers, in fact, could use large time steps as the DVI, but the highly non-linear nature of the contact forces often impair their performance to the point that they need short time steps anyway.

In general, since the SO-CCP solution dominates the overall computational effort per each time step of the DVI, the efficiency of the DVI method respect to smooth integrators depends mostly on how optimized is the SO-CCP solver. 
In order to maximize the computational performance, the inner loop of our interior-point method leverages the MKL parallel direct solver from Intel. As an alternative, we tested also CVXOPT, a Python library for large-scale cone programming, again based on an interior-point method~\cite{andersen2011interior}.

%In conclusion the total CPU time used to compute a simulation with DVI, hence its overall efficiency, is less -or more- than one used by smooth integrators depending on the following factors that change from case to case: how stiff is the contact penalty (steel-like structures are more likely to be efficiently solved by DVI, rather than soft structures), how fast the SO-CCP solver can compute the solution (objects with odd mass scaling, for instance, are known to require longer CPU times in our SO-CCP solver, and some implementations of SO-CCP solvers converge much faster than others), presence of collisions between very thin objects.

\section{Numerical tests}

% REMOVED SUBSECTION ON LINEAR STATICS (CANTILEVER
% BENCHMARK WITH TABLES) AS IT SEEMS TO MUCH. 
% REMOVE COMMENT IN FRONT OF \input{...} 
% TO ENABLE AGAIN, IF NEEDED. ALESSANDRO.
%
% \input{"chapter_cantilever.tex"}

\subsection{The ring bending test}

This is a classical benchmark where a cantilever beam is subject to a concentrated torque $T$ acting on the tip to obtain a ring bending deformation. Since the length of the neutral axis is constant under pure bending, the initial beam length \emph{L} must be equal to the final circumference, hence the curvature radius must be $\rho = \frac{L}{2\pi}$.
For a linear elastic rod it holds
$\rho = \frac{E I}{T}$, so the torque required to obtain a perfect circle must be 
%
\begin{equation}
T = \frac{2\pi EI}{L} \nonumber
\end{equation}

Because of the large deformations, we performed a non-linear static analysis where the torque value has been increased gradually to avoid divergence. Our simulations converge consistently with the analytical solution, as shown in Fig.~\ref{figringbending}.

\begin{figure}[ht]
\centering
\includegraphics[width=0.65\textwidth]{benchmark_RingBending/bending_progression.pdf}
\caption{Progression of ring bending}
\label{figringbending}
\end{figure}

\subsection{Princeton beam experiment}

A thin beam, depicted in Fig.~\ref{figB1setup}, is subject to large deformations and large rotations because of a tip load at $E$, for different angles $\theta$. A non-linear analysis shows that, for large displacements, a strong twisting action couples to the bending action, hence obtaining out-of-plane displacements even if the load is vertical.

Three loading conditions are tested: $P_1 = 4.448$N, $P_2 = 8.896$N,
and $P_3 = 13.345$N, for $\theta$ ranging in the $[0^\circ,90^\circ]$ interval.
The beam has length $L = 0.508$m, section height $H = 12.77$mm, section thickness $T = 3.2024$mm, Young modulus $E = 71.7$GPa, $\nu = 0.31$, $G = E \frac{(1 + \nu)}{2} = 27.37$GPa.

Because of the geometric nonlinearities, the solver must perform Newton-Raphson steps
before obtaining a zero residual in the equations of static equilibrium.
% For very large nonlinearities, a continuation strategy might help the convergence of the Newton-Raphson solver.

Results in Figs.~\ref{figB1r1}, \ref{figB1u2} and \ref{figB1u3} show a good agreement
between the present IGA formulation and reference data~\cite{BAUCHAU-2014-IMSD}. In detail, there is a good agreement with other geometrically-exact beam formulations (non IGA-based) discussed in~\cite{BAUCHAU-DYMORE} for Dymore and in~\cite{FV-AIAA} for MBDyn, 
as well as an agreement with the experimental results
in~\cite{DOWELL-1975-PRINCETON-1194,DOWELL-1975-PRINCETON-1257}, obtained with a beam made with 7075 aluminium alloy.

\begin{figure}[ht]
 \begin{minipage}[b]{0.48\linewidth}
    \centering
    \includegraphics[width=0.99\textwidth]{benchmark_princeton/setup_princeton.eps}
    \caption{Setup of the benchmark for the Princeton beam experiment.}
    \label{figB1setup}
 \end{minipage}
 \hspace{0.4cm}
 \begin{minipage}[b]{0.48\linewidth}
    \centering
    \includegraphics[width=0.99\textwidth]{benchmark_princeton/benchmark_princeton_r1.pdf}
    \caption{Twist rotation of the beam for the Princeton experiment.}
    \label{figB1r1}
 \end{minipage}
\end{figure}

\begin{figure}[ht]
 \begin{minipage}[b]{0.48\linewidth}
    \centering
    \includegraphics[width=0.99\textwidth]{benchmark_princeton/benchmark_princeton_u2.pdf}
    \caption{Flapwise displacement at the beam tip versus loading angle for three loading conditions.}
    \label{figB1u2}
 \end{minipage}
 \hspace{0.4cm}
 \begin{minipage}[b]{0.48\linewidth}
    \centering
    \includegraphics[width=0.99\textwidth]{benchmark_princeton/benchmark_princeton_u3.pdf}
    \caption{Chordwise displacement at the beam tip versus loading angle for three loading conditions.}
    \label{figB1u3}
 \end{minipage}
\end{figure}

% REMOVED SUBSECTION ON MODDAL ANALYSIS
% IT SEEMS TO MUCH. REMOVE COMMENT IN FRONT
% OF \input{...} TO ENABLE IF NEEDED. 
%\input{"chapter_modal.tex"}

\subsection{Jeffcott rotor}

\begin{figure}%[ht]
 \begin{minipage}[b]{0.4\linewidth}
    \centering
    \includegraphics[width=0.99\textwidth]{benchmark_jeffcott/setup_jeffcott.eps}
    \caption{Setup of the unbalanced rotating shaft benchmark.}
    \label{figJRsetup}
 \end{minipage}
 \hspace{0.4cm}
 \begin{minipage}[b]{0.55\linewidth}
    \centering
	\psfrag{u[m]}{\tiny $\boldsymbol{u}_2$, $\boldsymbol{u}_3$ [m]}
    \includegraphics[width=0.99\textwidth]{benchmark_jeffcott/benchmark_jeffcott.pdf}
    \caption{Mid-point transverse displacement of unbalanced rotating shaft.}
    \label{figJRdisp}
 \end{minipage}
\end{figure}

This benchmark explores the reliability of the numerical method in the nonlinear dynamical analysis
of a flexible system rotating at finite angular velocity.
A rotating unbalanced shaft of length $L$ = \SI{6}{\meter} is integrated in time.
As shown in Fig.~\ref{figJRsetup}, a rigid disk is connected to the shaft at mid-span,
above the reference shaft axis by an offset $d$ = \SI{0.05}{\meter}.
The shaft is made of steel (density $\rho$ = 7800kg/m$^3$,
Young's modulus $E$ = \SI{210}{\giga\pascal}, Poisson's ratio $\nu$ = 0.3).
The cross section is annular ($r_i$ = \SI{0.045}{\meter}, $r_o$ = \SI{0.05}{\meter}).
The mass of the disk is $m_d$ = 70.573kg,
the radius is $r_d$ = 0.24m,
and the thickness is $t_d$ = 0.05m.
The system is subjected to gravity ($g$ = \SI{9.81}{\meter\per\square\second})
directed transversely.
The end $R$ of the shaft is connected to the ground
by a cylindrical joint (displacement along and rotation
about the shaft's axis are permitted).
The end $T$ is supported by a revolute joint;
the relative angular velocity about the shaft axis is prescribed
as a function of time, % $\omega = \Omega(t)$:
\[
\Omega(t) = \left\{
\begin{array}{lr}
	A_1 \omega (1 - \cos(\pi t / T_1))/2 & 0 \le t \le T_1
	\\
	A_1 \omega & T_1 < t \le T_2
	\\
	A_1 \omega + (A_2 - A_1) \omega (1 - \cos(\pi (t - T_2)/(T_3 - T_2)))/2 & T_2 < t \le T_3
	\\
	A_2 \omega & T_3 < t
\end{array}
\right.
\]
with
$A_1$ = 0.8,
$A_2$ = 1.2,
$T_1$ = 0.5s,
$T_2$ = 1s,
$T_3$ = 1.25s, and
$\omega$ = \SI{60}{\radian\per\second}, close to the first natural frequency of the system.
%
The shaft accelerates from zero and passes from sub-critical to super-critical regime;
when passing through the first natural bending frequency of the system,
lateral oscillations occur and significant forces take place, as predicted
by the linear theory of unbalanced rotors.

Results plotted in~Fig.\ref{figJRdisp} for the case of a 3rd order IGA with nine nodes, using a second order implicit time stepping integrator, show a good agreement with reference results obtained with MBDyn.

\subsection{Lateral buckling}

This is a benchmark that tests the beam formulation in the context of a dynamical problem of difficult integration.

In detail, a flat beam is bent in its plane of highest flexural rigidity, up to the point where lateral buckling is instantly triggered. In a quasi-static non-linear analysis,  
results are visible in Fig.~\ref{figB2static}. In the context of dynamics, when buckling occurs, the beam snaps laterally and twists, inducing highly oscillatory motions. The IGA beam discussed in this paper can capture the nonlinear nature of this phenomena.

\begin{figure}%[ht]
 \begin{minipage}[b]{0.48\linewidth}
    \centering
    \includegraphics[width=0.70\textwidth]{benchmark_buckling/setup_buckling.eps}
    \caption{Setup of the benchmark for lateral buckling dynamics.}
    \label{figB2setup}
 \end{minipage}
 \hspace{0.4cm}
 \begin{minipage}[b]{0.48\linewidth}
    \centering
    \includegraphics[width=0.99\textwidth]{benchmark_buckling/benchmark_buckling_disp_static.pdf}
    \caption{Static displacement of the beam along $i_2$, at the mid point.}
    \label{figB2static}
 \end{minipage}
\end{figure}

As depicted in Fig.~\ref{figB2setup}, a $RC$ beam with length $L=1$m and rectangular section  $H=100$mm, $B=10$mm, is clamped at the $R$ end point.
The snapping is caused by a tip load at $C$, generated by mean of a rotating crank $GB$ and a vertical rod $TB$, with a spherical joint in $C$ and a revolute joint in $B$. An initial imperfection is simulated by introducing a small offset $d=0.1$mm in the off-plane direction $i_2$ between the crank and the vertical bar.

The crank has length $L_c=0.05$m and its circular section has diameter $D_r=24$mm, the vertical rod 
has length $L_r=0.25$m and its circular section has diameter $D_r=48$mm.
The rotation of the crank is enforced by a prescribed motion function 
$\phi_c(t)= \pi(1-\cos(\pi t/T_c) )/2$, with $T_c = 0.4$s, 
then after $t>T_c$ it holds $\phi_c(t)=\pi$.

All parts have Young modulus $E = 73$GPa and Poisson ratio $\nu = 0.3$. For the three beams, inertia values $I_{zz}$ and $I_{yy}$ and torsion constants $J$ are computed using formulas available in classical textbooks.

\begin{figure}%[ht]
  \begin{minipage}[b]{0.48\linewidth}
    \centering
    \includegraphics[width=0.99\textwidth]{benchmark_buckling/benchmark_buckling_disp.pdf}
    \caption{Displacement of the beam along $i_2$, at the mid point.}
    \label{figB2disp}
 \end{minipage}
 \hspace{0.4cm}
 \begin{minipage}[b]{0.48\linewidth}
    \centering
    \includegraphics[width=0.99\textwidth]{benchmark_buckling/benchmark_buckling_vel.pdf}
    \caption{Angular velocity of the beam, at the mid point.}
    \label{figB2wvel}
  \end{minipage}
\end{figure}

In our tests the crank and the rod are modeled with 4 nodes each, using first order IGA beams, whereas the $RC$ beam is modeled with 12 nodes, using third order IGA beams.

We performed the dynamical analysis both with a conventional HHT time integrator and with the DVI time stepper of \eqref{eq:dvi}-\eqref{eq:complementarity}.
Comparing the results with the reference data obtained with MBDyn in Figs.~\ref{figB2disp} and \ref{figB2wvel}, a remarkable fact is that the lateral buckling is triggered exactly at the same moment for all the formulations, and the resulting oscillations have the same period. 
However, as expected, the DVI time stepper introduces some numerical damping. In fact, in scenarios that do not involve frictional contacts as in this benchmark, the conic complementarity constraint \eqref{eq:complementarity} disappears and the DVI time stepper boils down to a linearly-implicit first-order scheme, hence it shows the same damping effect of an implicit Euler method~\cite{TasoraAnitescuCMAME10}. When using the HHT second-order implicit time integrator, results are affected by numerical damping to a lower degree. The drawback of the numerical damping in the DVI method can still be accepted when its superior stability and its efficiency in contact problems are attractive, as shown in the following example.

\subsection{Contacts with rigid body}

This benchmark shows that the proposed IGA beam performs well in a dynamical simulation with complex spatial contacts, especially when using the non-smooth formulation embedded in the DVI time stepper. 
A bundle of IGA beams has been fixed between two shafts, one of which is rotating at constant speed. A fixed central rod has been added, so that it will be wrapped by the IGA beams during the simulation, see Fig.~\ref{fig:contact}.

The bundle consists in a set of eight beams, each being an IGA beam of third order with \num{57} nodes, for a total length of~\SI{0.5}{\meter}. The section is circular with a diameter of \SI{0.01}{\meter}, the Young modulus is $E=\SI{0.5}{\giga\pascal}$, the shear modulus $G=\SI{0.35}{\giga\pascal}$, the density is $\rho=\SI{1000}{\kilogram\per\cubic\meter}$. The internal cylinder diameter is~\SI{0.5}{\meter}, while the beams are distributed on a circular pattern of diameter~\SI{0.5}{\meter}. The rotation speed is~\SI{1}{\radian\per\second}.

The interior-point solver used in this simulation is request to achieve a tolerance of~\num{1d-10} over the residuals and complementarity gap.

%Additional information can be found in~Tab.\ref{tab:torque_data}.

%\begin{table}[ht]
%\centering
%    \begin{tabular}{lc}
%    \toprule
%    Parameter & Value \\
%    \midrule
%    Young Modulus & \SI{0.5}{\giga\pascal} \\
%    Shear Modulus & \SI{0.35}{\giga\pascal} \\
%    Density & \SI{1000}{\kilogram\per\cubic\meter} \\
%    \bottomrule
%    \end{tabular}
%\caption{IGA beam parameter for contact benchmark.}
%\label{tab:torque_data}
%\end{table}

\begin{figure}[ht]
    \centering
    \begin{subfigure}[b]{0.475\textwidth}
        \centering
        \includegraphics[width=\textwidth]{benchmark_contact/NSC001deg000}
        \caption{{\small Rotation: \SI{0}{\degree}}}    
        % \label{}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.475\textwidth}  
        \centering 
        \includegraphics[width=\textwidth]{benchmark_contact/NSC001deg120}
        \caption[]{{\small Rotation: \SI{120}{\degree}}}    
        % \label{}
    \end{subfigure}
    \vskip\baselineskip
    \begin{subfigure}[b]{0.475\textwidth}   
        \centering 
        \includegraphics[width=\textwidth]{benchmark_contact/NSC001deg240}
        \caption[]{{\small Rotation: \SI{240}{\degree}}}    
        % \label{}
    \end{subfigure}
    \quad
    \begin{subfigure}[b]{0.475\textwidth}   
        \centering 
        \includegraphics[width=\textwidth]{benchmark_contact/NSC001deg360}
        \caption[]{{\small Rotation: \SI{360}{\degree}}}    
        % \label{}
    \end{subfigure}
    \caption[Contact with rigid body]
    {\small Contact with rigid body (in red, elements that are fixed; in blue, the rotating shaft; in green, the IGA beams).} 
    \label{fig:contact}
\end{figure}

The dynamical analysis has been performed using the DVI time stepper \eqref{eq:dvi}-\eqref{eq:complementarity}. For this non-smooth dynamics problem, frictional contacts are enforced as complementarity constraints that do not require any tuning of penalty. %and the tangency of the contacting surfaces is guaranteed up to micrometric tolerances at each run of the SO-CCP solver. 
Note that when using conventional implicit or explicit integrators for smooth dynamics, high penalty stiffness would be needed to approximate contact between rigid materials without contact compenetration, but in turn this requires very short time steps to avoid numerical instability.

Multiple tests have been run with increasing time steps, from~\SI{1}{\ms} up to~\SI{75}{\ms}. 
This range includes values that are unusually large for this type of analysis; just for reference we report that we ran the same benchmark using the HHT time stepper and penalty contacts, but to avoid divergence the time step could not be larger than $h=\SI{1e-5}{\s}$.

The forces induced by beams on the shafts have been reported in~Fig.\ref{fig:contact_conv} and the relative error respect to a reference solution (computed with much shorter time step) is shown: being always under~\num{3}\% even for the largest time step is a good proof for the robustness of the DVI solver. It is remarkable that the solver was stable even for time steps larger than~\SI{100}{\ms}, but at that point the precision started to be compromised by other factors such as the precision of the collision detection algorithms etc.

\begin{figure}[ht]
    \centering
    \includegraphics[width=0.7\textwidth]{benchmark_contact/ContactConvWithoutNSC100.png}
    \caption{Reaction force on the shaft for different time steps. In the legend, labels like NSC010 means Non-Smooth Contact test with time step $h=$\SI{10}{\ms} }
    \label{fig:contact_conv}
\end{figure}

%The overall trend of the root-mean-square error at increasing time step values is shown in~Fig.\ref{fig:contact_rmse}.

%\begin{figure}[ht]
%    \centering
%    \includegraphics[width=0.8\textwidth]{benchmark_contact/contact_rmse.png}
%    \caption{RMSE of reaction forces on the shaft for different time steps.}
%    \label{fig:contact_rmse}
%\end{figure}

\subsection{Woven mesh}
In order to assess the performance of the DVI method for the case of mutual contact between IGA beams, a woven mesh of IGA beams has been reproduced and compared to literature~\cite{weeger2018frictional}.
The experiment consists in a 7x7 mesh where wires are clamped at one end, free at the other. Each wire is modeled with 64 nodes and a third-order IGA rod with cross-sectional radius of \SI{1}{mm}, Young modulus $E = \num{10e8}$ and Poisson ratio of $\nu = \num{0.5}$, for a total length $L = 4L_w = \SI{0.12}{m}$ where $L_w = \SI{0.03}{m}$ is the wavelength of the curve used to weave the mesh. Because of the non-smooth contact model no additional compliance parameter is required.

A distributed load of \SI{0.1}{N/m} is gradually applied to the beams in the vertical direction, and the DVI method is used to perform a non-linear analysis with a time step of~\SI{10}{ms} using the Interior-Point solver described in the previous example. Results are shown in~Fig.\ref{fig:woven_mesh}. % to reproduce the cited experiment. % but with significant difference that no progressive step has to be made. 
We remark that the time step could be increased up to~\SI{50}{ms} without incurring in divergence. Although the tolerance of the solver has been kept at the very strict threshold of~\num{1d-10} on residuals and complementarity, the computational time never exceeded \SI{0.997}{s/step} on a \SI{2.4}{GHz} Intel i7-4700 processor. %, with the perspective of achieving even faster performance if lower tolerances can be accepted and if future optimizations will be implemented in the code. %Contact points are computed automatically by the collision engine considering a sweep of spheres along the beams.

\begin{figure}[ht]
    \centering
    \begin{subfigure}[b]{0.475\textwidth}
        \centering
        \includegraphics[width=\textwidth]{benchmark_mesh/mesh_load000.png}
        \caption{{\small Load: 0\%}}    
        % \label{}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.475\textwidth}  
        \centering 
        \includegraphics[width=\textwidth]{benchmark_mesh/mesh_load030.png}
        \caption{{\small Load: 30\%}}
        % \label{}
    \end{subfigure}
    \vskip\baselineskip
    \begin{subfigure}[b]{0.475\textwidth}   
        \centering 
        \includegraphics[width=\textwidth]{benchmark_mesh/mesh_load060.png}
        \caption{{\small Load: 60\%}}
        % \label{}
    \end{subfigure}
    \quad
    \begin{subfigure}[b]{0.475\textwidth}   
        \centering 
        \includegraphics[width=\textwidth]{benchmark_mesh/mesh_load100.png}
        \caption{{\small Load: 100\%}}
        % \label{}
    \end{subfigure}
    \caption[Woven mesh]
    {\small Snapshots from the simulation of the woven mesh, for increasing load.} 
    \label{fig:woven_mesh}
\end{figure}

\section{Conclusions}

A geometrically-exact spatial Cosserat rod has been implemented via IGA discretization, using quaternions as rotational coordinates. We implemented the model in a open-source physics library using C++ language and we tested it in various non-linear static and dynamic problems, showing generality, robustness and efficiency of the formulation.

Using a DVI time stepping scheme and a SO-CCP solver, we proved that simulations with non-smooth frictional contacts between IGA beams and moving shapes can be performed with large time steps. The enhanced robustness and stability indicate that the DVI method is a viable alternative to conventional methods based on contact penalty.

%
% ---- Bibliography ----
%
\section*{\refname}
\bibliographystyle{plain}
\bibliography{mybib,BibMBS,BibFEM}

\appendix
\section{Constitutive models for beams}
\label{sec:constitutive}

In the most generic case, \eqref{eq:mnconstitutive} may be a non-linear function, but in the following we present special cases of practical interest.

\subsection{Generic linear elasticity}
\label{sec:generic}

The generic case of constitutive model for the Cosserat rod, with linear elasticity, requires a matrix $\amatr{E} \in \mathbb{R}^{6x6}$, not necessarily sparse. Such matrix can be provided by some detailed 3D FEA analysis of a chunk of beam, in a preprocessing stage.
%
\begin{align}
    \left\{  
    \begin{array}{c}
     \avect{n} \\
     \avect{m} 
    \end{array}
    \right\}
    &= 
    \amatr{E}
    \left\{  
    \begin{array}{c}
     \avect{\epsilon} \\
     \avect{\kappa}
    \end{array}
    \right\}
\end{align}

We remark that, depending on the 36 values used in the 6x6 $\amatr{E}$ matrix,  $\avect{n}$ and $\avect{m}$ might have coupled effects. Figure \ref{fig:section} shows the reference coordinate system of the section.

\begin{figure}[ht]
    \centering
    \includegraphics[width=0.60\textwidth]{section_beam.pdf}
    \caption{Section of the beam. Generic case.}
    \label{fig:section}
\end{figure}

\subsection{Basic diagonal linear elasticity}
\label{sec:basic}

For centered symmetric sections, the previous relation can be simplified. In fact one can express the constitutive relation via a linear mapping where $\avect{n}$ and  $\avect{m}$ effects are uncoupled:
%
\begin{align}
\avect{n} &= \avect{n}(\avect{\epsilon}) \\ 
\avect{m} &= \avect{m}(\avect{\kappa})
\label{eq:mn}
\end{align}
%
In detail, with linear elasticity, a very common case is the linear mapping:
%
\begin{align}
\avect{n} &= \amatr{C} \avect{\epsilon}  \\ 
\avect{m} &= \amatr{D} \avect{\kappa} 
\label{eq:mnlin}
\end{align}
%
where one simply uses the material matrices:
%
\begin{align}
	\amatr{C} = 
	\left[  
	\begin{array}{ccc}
	 E A & 0 & 0 \\
	 0 & G A k_y & 0 \\
	 0 & 0 & G A k_z 
	\end{array}
	\right]
	\quad
	\amatr{D} = 
	\left[  
	\begin{array}{ccc}
	 G J & 0 & 0 \\
	 0 & E I_y & 0 \\
	 0 & 0 & E I_z
	\end{array}
	\right]
\end{align}
%
for given material Young' modulus $E$, shear modulus $G$, area $A$, Timoshenko shear correction factors $k_y,k_z$, torsion constant $J$, and second moments of area $I_y, I_z$ computed in the section reference. Note that the center of axial forces $C_a$ and the shear center $C_s$ are in the origin, by assumption. See Fig.~\ref{fig:section_diagonal}.

\begin{figure}[ht]
    \centering
    \includegraphics[width=0.60\textwidth]{section_beam_diagonal.pdf}
    \caption{Section of the beam. Simplified case (diagonal $\amatr{E}$).}
    \label{fig:section_diagonal}
\end{figure}

\subsection{Advanced section for linear elasticity}
\label{sec:advanced}

A more general constitutive model is the one depicted in Fig.~\ref{fig:section_advanced}, where the $I_y, I_z$ are computed respect to an auxiliary reference $C_a$, center of axial forces, that has displacement $y_a, z_a$ and rotation $\alpha$ respect to the reference center line of the beam~\cite{MASARATI-2014-IJSS} .

Also, in case of non-symmetric sections, it may happen that the shear center $C_s$ does not coincide with $C_a$; if so, one can provide displacements $y_s, z_s$ and rotation $\beta$ of the reference of the shear center $C_s$ respect to the reference center line of the beam.

Usually, for symmetric sections, it tends to $y_s=y_a$, $z_s=z_a$, $\alpha=\beta$. For simple problems like rectangular or circular sections centered on the reference line of the beam, one has $y_s=y_a=0$, $z_s=z_a=0$, $\alpha=\beta=0$.

\begin{figure}[ht]
    \centering
    \includegraphics[width=0.60\textwidth]{section_beam_advanced.pdf}
    \caption{Section of the beam. Advanced case.}
    \label{fig:section_advanced}
\end{figure}

The linear model becomes:
%
\begin{align}
\left\{  
	\begin{array}{c}
	 \avect{n} \\
	 \avect{m} 
	\end{array}
	\right\}
	&= 
	\left[  
	\begin{array}{cccccc}
	 a_{11} & 0 & 0 & 0 & a_{12} & a_{13} \\
	 0      & s_{11} & s_{12} & s_{13} & 0 & 0 \\
	 0      & s_{21} & s_{22} & s_{23} & 0 & 0 \\
	 0      & s_{31} & s_{32} & s_{33} & 0 & 0 \\
	 a_{21} & 0      & 0      & 0      & a_{22} & a_{23} \\
	 a_{31} & 0      & 0      & 0      & a_{32} & a_{33}
	\end{array}
	\right]
	\left\{  
	\begin{array}{c}
	 \avect{\epsilon} \\
	 \avect{\kappa}
	\end{array}
	\right\}
\end{align}

Here the components $a_{ij}=\amatr{A}$ and $s_{ij}=\amatr{S}$ are obtained by rotations $\amatr{R}$ and translations $\amatr{T}$ of the diagonal constitutive matrices $\amatr{A}_{C_a}$ and $\amatr{S}_{C_s}$:
%
\begin{align}
\amatr{A} &= \amatr{T}_{C_a} \amatr{R}_{C_a} \amatr{A}_{C_a}  \amatr{R}_{C_a}^T \amatr{T}_{C_a}^T \\
\amatr{S} &= \amatr{R}_{C_s}^T \amatr{T}_{C_s}^{-1} \amatr{S}_{C_s}  \amatr{T}_{C_s}^{-T} \amatr{R}_{C_s}
\end{align}
%
where
% 
\begin{align}
\amatr{A}_{C_a} = 
	\left[  
	\begin{array}{ccc}
	 E A & 0 & 0 \\
	 0 & E I_y & 0 \\
	 0 & 0 & E I_z 
	\end{array}
	\right]
	\quad
	\amatr{S}_{C_s} = 
	\left[  
	\begin{array}{ccc}
	 G A k_y  & 0 & 0 \\
	 0 & G A k_z & 0 \\
	 0 & 0 & G J
	\end{array}
	\right]
\end{align}

The above model requires the following parameters: Young modulus $E$, shear modulus $G$, area $A$, Timoshenko shear correction factors $k_y,k_z$, torsion constant $J$, and second moments of area $I_y, I_z$, as the diagonal simplified model, plus the  position and rotation of $C_a$ as $y_a, z_a$ and $\alpha$, plus the position and rotation of $C_s$ as $y_s, z_s$ and $\beta$.

% REMOVED SUBSECTION ON MESH-INTEGRATED SECTION, 
% IT SEEMS TO MUCH. REMOVE COMMENT IN FRONT
% OF \input{...} TO ENABLE IF NEEDED. ALESSANDRO.
%\input{"chapter_integrated_section.tex"}

%\subsection{Other}
%
%For the case of elasto-plastic constitutive models, see %later.

% REMOVED SUBSECTION ON BEAM PLASTICITY, 
% IT SEEMS TO MUCH. REMOVE COMMENT IN FRONT
% OF \input{...} TO ENABLE IF NEEDED. ALESSANDRO.
%\input{"chapter_plasticity.tex"}

\end{document}

### 2020-05-15 2019/ICACR_Prague/sig-alternate-sample.tex
Added ICACR

% This is "sig-alternate.tex" V2.1 April 2013
% This file should be compiled with V2.5 of "sig-alternate.cls" May 2012
%
% This example file demonstrates the use of the 'sig-alternate.cls'
% V2.5 LaTeX2e document class file. It is for those submitting
% articles to ACM Conference Proceedings WHO DO NOT WISH TO
% STRICTLY ADHERE TO THE SIGS (PUBS-BOARD-ENDORSED) STYLE.
% The 'sig-alternate.cls' file will produce a similar-looking,
% albeit, 'tighter' paper resulting in, invariably, fewer pages.
%
% ----------------------------------------------------------------------------------------------------------------
% This .tex file (and associated .cls V2.5) produces:
%       1) The Permission Statement
%       2) The Conference (location) Info information
%       3) The Copyright Line with ACM data
%       4) NO page numbers
%
% as against the acm_proc_article-sp.cls file which
% DOES NOT produce 1) thru' 3) above.
%
% Using 'sig-alternate.cls' you have control, however, from within
% the source .tex file, over both the CopyrightYear
% (defaulted to 200X) and the ACM Copyright Data
% (defaulted to X-XXXXX-XX-X/XX/XX).
% e.g.
% \CopyrightYear{2007} will cause 2007 to appear in the copyright line.
% \crdata{0-12345-67-8/90/12} will cause 0-12345-67-8/90/12 to appear in the copyright line.
%
% ---------------------------------------------------------------------------------------------------------------
% This .tex source is an example which *does* use
% the .bib file (from which the .bbl file % is produced).
% REMEMBER HOWEVER: After having produced the .bbl file,
% and prior to final submission, you *NEED* to 'insert'
% your .bbl file into your source .tex file so as to provide
% ONE 'self-contained' source file.
%
% ================= IF YOU HAVE QUESTIONS =======================
% Questions regarding the SIGS styles, SIGS policies and
% procedures, Conferences etc. should be sent to
% Adrienne Griscti (griscti@acm.org)
%
% Technical questions _only_ to
% Gerald Murray (murray@hq.acm.org)
% ===============================================================
%
% For tracking purposes - this is V2.0 - May 2012

%\documentclass{sig-alternate-05-2015}

\documentclass{sig-alternate-05-2015}
\usepackage{textcomp}
\usepackage{biblatex}
\addbibresource{BibDRL.bib}
\addbibresource{BibMBS.bib}
\usepackage{acmcopyright}
\begin{document}

% Copyright
\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\setcopyright{usgov}
%\setcopyright{usgovmixed}
%\setcopyright{cagov}
%\setcopyright{cagovmixed}

% DOI
\doi{10.1145/3365265.3365274}

% ISBN
\isbn{978-1-4503-7288-6/19/10}

%Conference
\conferenceinfo{ICACR 2019,}{\emph{October 11--13, 2019 Prague, Czech Republic}}

\acmPrice{\$15.00}

%
% --- Author Metadata here ---

\CopyrightYear{2019} % Allows default copyright year (20XX) to be over-ridden - IF NEED BE.
%\crdata{0-12345-67-8/90/01}  % Allows default copyright data (0-89791-88-6/97/05) to be over-ridden - IF NEED BE.
% --- End of Author Metadata ---

\title{A Modular Simulation Platform for Training Robots via Deep Reinforcement Learning and Multibody Dynamics}

%\subtitle{[Extended Abstract]
%\titlenote{A full version of this paper is available as
%\textit{Author's Guide to Preparing ACM SIG Proceedings Using
%\LaTeX$2_\epsilon$\ and BibTeX} at
%\texttt{www.acm.org/eaddress.htm}}}

%
% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.

\numberofauthors{6} %  in this sample file, there are a *total*
% of EIGHT authors. SIX appear on the 'first-page' (for formatting
% reasons) and the remaining two appear in the \additionalauthors section.
%
\author{
% You can go ahead and credit any number of authors here,
% e.g. one 'row of three' or two rows (consisting of one row of three
% and a second row of one, two or three).
%
% The command \alignauthor (no curly braces needed) should
% precede each author name, affiliation/snail-mail address and
% e-mail address. Additionally, tag each line of
% affiliation/address with \affaddr, and tag the
% e-mail address with \email.
%
% 1st. author
\alignauthor
Simone Benatti\titlenote{Corresponding Author}\\
       \affaddr{University of Parma Dpt. of Engineering and Architecture}\\
       \affaddr{Parco Area delle Scienze, 181/A}\\
       \affaddr{Parma, Italy}\\
       \email{simone.benatti@\linebreak[0]studenti.unipr.it}
% 2nd. author
\alignauthor
Alessandro Tasora\\
       \affaddr{University of Parma Dpt. of Engineering and Architecture}\\
       \affaddr{Parco Area delle Scienze, 181/A}\\
       \affaddr{Parma, Italy}\\
       \email{alessandro.tasora@unipr.it}
% 3rd. author
\alignauthor 
Dario Fusai\\
       \affaddr{University of Parma Dpt. of Engineering and Architecture}\\
       \affaddr{Parco Area delle Scienze, 181/A}\\
       \affaddr{Parma, Italy}\\
       \email{dario.fusai@unipr.it}
\and  % use '\and' if you need 'another row' of author names
% 4th. author
Dario Mangoni\\
       \affaddr{University of Parma Dpt. of Engineering and Architecture}\\
       \affaddr{Parco Area delle Scienze, 181/A}\\
       \affaddr{Parma, Italy}\\
       \email{dario.mangoni@unipr.it}
}
% There's nothing stopping you putting the seventh, eighth, etc.
% author on the opening page (as the 'third row') but we ask,
% for aesthetic reasons that you place these 'additional authors'
% in the \additional authors block, viz.
\additionalauthors{Additional authors: John Smith (The Th{\o}rv{\"a}ld Group,
email: {\texttt{jsmith@affiliation.org}}) and Julius P.~Kumquat
(The Kumquat Consortium, email: {\texttt{jpkumquat@consortium.net}}).}
\date{30 July 1999}
% Just remember to make sure that the TOTAL number of authors
% is the number that will appear on the first page PLUS the
% number that will appear in the \additionalauthors section.

\maketitle
\begin{abstract}
In this work we focus on the role of Multibody Simulation in creating Reinforcement Learning virtual environments for robotic manipulation, showing a versatile, efficient and open source toolchain to create directly from CAD models.
Using the Chrono::Solidworks plugin we are able to create robotic environments in the 3D CAD software Solidworks\textregistered and later convert them into PyChrono models (PyChrono is an open source Python module for multibody simulation).
In addition, we demonstrate how collision detection can be made more efficient by introducing a limited number of contact primitives instead of performing collision detection and evaluation on complex 3D meshes, still reaching a policy able to avoid unwanted collisions.
We tested this approach on a 6DOF robot Comau Racer3: the robot, together with a 2 fingers gripper (Hand-E by Robotiq) was modelled using Solidworks\textregistered, imported as a PyChrono model and then a NN was trained in simulation to control its motor torques to reach a target position.
To demonstrate the versatility of this toolchain we also repeated the same procedure to model and then train the ABB IRB 120 robotic arm.

\end{abstract}

%
% The code below should be generated by the tool at
% http://dl.acm.org/ccs.cfm
% Please copy and paste the code instead of the example below. 
%
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10003752.10010070.10010071.10010261</concept_id>
<concept_desc>Theory of computation~Reinforcement learning</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10010147.10010371.10010352.10010379</concept_id>
<concept_desc>Computing methodologies~Physical simulation</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10010147.10010257.10010293.10010294</concept_id>
<concept_desc>Computing methodologies~Neural networks</concept_desc>
<concept_significance>300</concept_significance>
</concept>
<concept>
<concept_id>10010520.10010553.10010554.10010556</concept_id>
<concept_desc>Computer systems organization~Robotic control</concept_desc>
<concept_significance>100</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Theory of computation~Reinforcement learning}
\ccsdesc[500]{Computing methodologies~Physical simulation}
\ccsdesc[300]{Computing methodologies~Neural networks}
\ccsdesc[100]{Computer systems organization~Robotic control}
%
% End generated code
%

%
%  Use this command to print the description
%
\printccsdesc

% We no longer use \terms command
%\terms{Theory}

\keywords{Physical Simulation; Multibody Simulation; Reinforcement Learning; Deep Learning; Neural Networks; Robotics; Control}

\section{Introduction}
Since its introduction \cite{Mnih13}, Deep Reinforcement Learning has been applied successfully to a wide range of applications in robotic grasping \cite{Levine2016} \cite{LevineFDA15} and dexterous manipulation \cite{openai2018HandManipulation}.

This being said, the role of simulation so far has been primarily aimed to benchmarking of DRL algorithms in continuous state-action tasks. In this context, model definition through xml files using URDF syntax is totally appropriate, favoring the model sharing, however, this approach lacks of versatility.
Since the next years will probably see an increasing diffusion of industrial applications of DRL for robotic control, there will be a major interest in the ability of creating tailored and efficient training environments. 
For this reason we developed and tested a toolchain which allows to create Multibody Simulation models from CAD and use them to train NN.

In addition , we want to demonstrate that in most cases there is no need to perform collision detection and evaluate contact forces on complex triangular meshes, and a limited number of primitive contact shapes placed in critical points is sufficient for the agent to learn to avoid unwanted collisions. 
To demonstrate these statements, we tested this approach on a Comau Racer3 robot (a small 6DOF robotic arm). The model, created with Solidworks\textregistered and imported in PyChrono \cite{Tasora2015ChronoAO}, was used to train a NN using Proximal Policy Optimization \cite{Schulman2017PPO}
to control the motor torques to reach a random position within a working area.

\section{Physical Model}
\subsection{Importing the model}
From the 3D CAD model we are able to import bodies, centers of gravity positions, links, masses, inertia matrices and shapes by means of our open-source plugin for Solidworks.

The plugin creates a Python script that uses PyChrono for Multibody Simulation. Using Python for simulating physics makes exchanging data between the simulation and the DL algorithm easier.

Only actuators have to be modeled manually, but since the reference frames defining the concentric links can be used to define the motors, the process can be automated. Indeed, once the model was set up for the Racer3 robot, there was no need to further modify the code for the ABB IRB 120 (except for speed and rotation limits and maximum torques).
\subsubsection{Mass properties}
Mass properties deserve a particular attention. Almost any multibody simulator is able to evaluate mass, COG and inertia matrix knowing the body shape and density. This approach assumes uniform density, thus it is correct for uniform bodies, but it can lead do huge errors in mass properties estimation when the body density is highly non uniform, such as a robotic arm, in which the mass is predominantly concentrated in the motor, while the rest of the arm is essentially hollow.
These errors can be a major source of difference between realty and simulation, and thus the policy, trained in simulation, becomes harder to transfer in the real world.

CAD software allows to precisely evaluate mass properties of subsystem, considering the right density and shape of each component, and thus the simulation importing these data will be closer to reality. \footnote{We could not exploit this in the cases shown later since precise mass properties are not usually given, but designers of custom robots interested in NN control could particularly take advantage of this feature}

%\subsubsection{Joints and Motors}

\subsection{Contacts}
In general contact simulation could be performed on triangular meshes approximating the real shape of the bodies, but we will show that in many cases this is not necessary. What we do instead is attaching to the bodies a small number of contact primitives (cylinders, boxes spheres) as shown in figure 1.
\begin{figure}[ht] \label{ConShapes}
\centering
\includegraphics[width=0.35\textwidth]{Figures/CollShapes.JPG}
\caption{Collision Shapes (in Red)}
\end{figure}

Following this approach it is possible to reduce the computational time and the memory required by the training process, since contact detection and contact force evaluation becomes computationally lighter and triangular meshes are not even loaded in RAM since they are no longer required.

\subsubsection{Arm Contact}
Consider that because of body shapes, constraints and maximum allowed joints rotation only certain portion of the robotic arm can collide; moreover we are not interested in evaluating the exact contact but only in detecting the collision to penalize the reward in order to instruct the policy to avoid self collision.
\subsubsection{Fingers Contact}
The contact between the fingers and the box, on the contrary, has to be modelled precisely, but in many cases, such as this one, fingertips can also be modeled with contact primitive, as shown in figure 2 (contact shapes are in red).

\begin{figure} \label{FingerCS}
\centering
\includegraphics[width=0.3\textwidth]{Figures/Finger_CS.JPG}
\caption{Fingers Collision Shapes Closeup}
\end{figure}

\section{Robotic Environment}
\begin{figure} \label{ComauR3}
\centering
\includegraphics[width=0.35\textwidth]{Figures/RobotRender.JPG}
\caption{The Comau- Racer-3 6DOF Robot}
\end{figure}
The virtual environment is composed of a 6DOF robotic arm fixed on a planar surface, on which a small cubic box is rested. A 2 finger robotic hand is coupled to the robotic arm but the finger cannot move since in this example we have not introduced pick and place yet. The agent must learn to control the robot such that its fingertips reach and keep the position of the box center.

The agent must also avoid:
\begin{itemize}
    \item Collisions between robotic arm parts.
    \item Collisions between the robot fingers and the table.
    \item Energy wasting policies
    \item Stuck joints
\end{itemize}

\subsection{States and Actions}
The state $s \in \mathbb{R}^{18}$ is composed of the potions and velocities of the robot joints ($\in \mathbb{R}^{12}$), the target position and the end effector position (both $\in \mathbb{R}^{3}$).

The actions are the 6 motor torques. 
\subsection{Reward Shaping}

The reward is shaped as follows:

\begin{equation} \label{eq:Reward}
    r = \frac{r_d}{d+\epsilon} + S_c + F_c + c_e \left\Vert  T \omega \right\Vert + c_j S_j
\end{equation}
Where:
\begin{itemize}
    \item $r$ and $c$ are reward and cost constant coefficients referred to distance, energy and stuck joints (\textit{d, e, j} subscripts).
    \item $d$ is the distance between the fingers and the objective, while $\epsilon=0.0001$ avoids division by 0.
    \item $F_c$ is the finger contact penalty, equal to:

    $-\left(\left\Vert F_{f1}\right\Vert + \left\Vert F_{f2}\right\Vert \right)$

    where  $F_{f1}$ and $F_{f2}$ are the contact forces on the 2 fingers.
    \item $A_c$ is the arm collision penalty, equals to -1000 whenever any part of the robot (except fingers) collide.
    \item $T$ and $\omega$ are the vectors  motor torques and speed, whose dot product is the vector of motor powers.
    \item $S_j$ is the number of stuck joints (a joint is stuck when its rotation is equal to its maximum rotation).
\end{itemize}
Moreover, if $A_c<0$ (which means that a body of the robotic arm is colliding), the episode is reset so that the agent cannot collect any reward afterwards.

\section{DRL Algorithm}
\subsection{Proximal Policy Optimization}
Policy Gradient Methods are, in general, particularly suitable for tasks involving continuous states and actions (such as robotic tasks), and Proximal Policy Optimization has shown to perform well in robotic control applications \cite{openai2018HandManipulation}.

The objective function for the parameter optimization in the implementation of the penalized version of the PPO algorithm \cite{Heess2017} is:
\begin{align*}
       \sum_{t=0}^\infty\gamma^t\frac{\pi_{\theta}(a_t|s_t)}{\pi_{\theta_{old}}(a_t|s_t)}
       A^{\pi, \gamma}(s_t,a_t)  -\beta_kD_{KL}(\theta_{old}||\theta) \\
        -\eta \max(0, D_{KL}(\theta_{old}||\theta) - 2KL_{targ})^2
\end{align*}
Where 
\begin{itemize}
    \item $\theta$ is the set of NN parameters (\emph{old} is their value at the beginning of the epoch). 
    \item $D_{KL}(\theta_{old}||\theta)$ is the Kullback-Leibler divergence between the policy at the beginning of the optimization epoch and the last policy.
    \item $KL_{targ}$ is the target value of the Kullback-Leibler divergence
    \item $\beta_k$ and $\eta$ are penalty coefficient used to penalize large policy updates.
    \item $A^\pi(s_t,a_t)$ is the Advantage Function
    \item $\gamma$ is the discount factor, used to reduce the relevance of future rewards.
\end{itemize}
We recall the definitions of discounted Value Function $V^{\pi, \gamma}(s_t)$: the expected sum of discounted rewards in the state $s_t$ at time $t$ following the policy $\pi$. 
\begin{equation}\label{eq_ValueFunc}
    V^{\pi, \gamma}(s_t) := E_{\underset{a_{t:\infty}}{s_{t+1:\infty}}}
    \left[\sum_{l=0}^\infty \gamma^l r_{t+l}\right]
\end{equation}

And discounted Advantage function  $A^{\pi, \gamma}(a_t,s_t)$ which is the advantage (in terms of reward) at timestep $t$ in state $s_t$, of taking the action $a_t$ and then following the policy, compared to following the policy.
\begin{equation} \label{eq:AdvFunc}
    A^{\pi, \gamma} = E_{s^{t+1}}\left[r_t+\gamma V^{\pi,\gamma}(s_{t+1})- V^{\pi,\gamma}(s_t)\right]
\end{equation}

To estimate the Advantage Function we trained a NN to estimate $V^{\pi, \gamma}(s_t)$, and then $A^{\pi, \gamma}(a_t,s_t)$ can be evaluated as suggested by Schulman et al. \cite{Schulman15GAE}.

The Kullback-Leibler divergence penalties prevents large parameters update avoiding the divergence of the policy; the KL divergence is indeed a measure of the difference between the two distributions.

PPO is an Actor-Critic algorithm: the Actor is NN actually prescribing the action  to the agent (the Policy), while the Critic in another Neural Network whose purpose is estimating the Critic performance, in our case by estimating the Value Function.

\subsection{NN Architecture}
We used fully connected NN with tanh activation function for both (Actor and Critic).
The policy NN has 18 inputs (the number of states), 6 outputs (the number of actions) and 3 hidden layers of 180, 104 and 60 neurons respectively.

The Critic NN has of course 18 inputs as well, and 1 output (the estimated value function). The 3 hidden layers have 180, 42 and 10 neurons respectively.
\section{Results}
\subsection{Comau Racer3}

\begin{figure}[ht] \label{Objective}
\centering
\includegraphics[width=0.3\textwidth]{Figures/objective.jpg}
\caption{Comau R3 Reaching Target Position }
\end{figure}

After 20000 training episodes the NN is able to control the robotic arm motor torques in order to reach a random position within a working area. The final position can be seen in Figure  4, and the learning progression is reported in Figure 5.
In addition, the control Policy learned to minimize energy consumption and contacts on the gripper fingers and above all, contacts between the robotic arm parts are always avoided. 
Using the simulation render to observe the Policy behaviour it can be noted that visualization shapes never overlap, demonstrating that using a limited number of contact primitives is indeed sufficient to avoid collisions between the real shapes. As can be seen in Figure 6 after the first half of the training, as soon as the Policy begins to undertake the wanted behavior, the episodes always last 3 seconds: since the agent dies if there is a collision between robotic arm parts, this means that they never happen after 10000 episodes of training.

It can also be noted that the policy learns to avoid unwanted collision first, then it learns to bring the end effector in the right position.

\begin{figure}[ht] \label{R3RewPlot}
\centering
\includegraphics[width=0.3\textwidth]{Plots/R3Rew.pdf}
\caption{Comau R3 Sum Of Collected \newline Rewards (10 Batches Mean)}
\end{figure}

\begin{figure}[ht] \label{R3TimePlot}
\centering
\includegraphics[width=0.3\textwidth]{Plots/R3Time.pdf}
\caption{Comau R3 Episode Duration \newline (10 Batches Mean)}
\end{figure}

\subsubsection{Transferring the Policy}
Training an agent on a simple task and then transfer the learning in a more complex environment is a well studied strategy and has also been applied in deep reinforcement learning for continuous control problems \cite{Berseth18}.
%
%\paragraph{Constant to Random Position}

Training the agent to reach an object whose position is fixed and then training the same policy to reach a random position within a working area proved to be a poor strategy, as shown in the plot in Figure 7. The policy performance is bad and the benefit in term of training time is negligible  compared to training with a random object position from scratch as previously shown.

The limited training speedup discouraged us from considering more advanced transfer learning techniques.

\begin{figure}[ht] \label{R3ConstToRand}
\centering
\includegraphics[width=0.3\textwidth]{Plots/R3_C_to_Rand_Rew.pdf}
\caption{Switching from fixed to random position after 20000 episodes. Sum of rewards
(10 B Mean)}
\end{figure}

%\paragraph{Fixed to Moving Objective Box}

\subsection{ABB IRB 120}
Finally, we used the same procedure to train a different robotic arm. We choose the ABB IRB 120 because it is similar in terms of working area and payload to the Comau Racer3, thus they can be both used in the same applications.

Here we fully exploited the versatility of our simulation platform, since except for collision shapes in the CAD and maximum rotations, velocities and torques in the code, we did not have to introduce modifications.

In Figure 8 is reported the learning progression while Figure 9 is a render frame of the simulation of the trained Policy.

\begin{figure}[ht] \label{ABB_Rew}
\centering
\includegraphics[width=0.3\textwidth]{Plots/ABB_Rew.pdf}
\caption{ABB IRB 120 Sum Of Collected \newline Rewards (10 Batches Mean)}
\end{figure}

\begin{figure}[ht] \label{ABB}
\centering
\includegraphics[width=0.3\textwidth]{Figures/ABB.jpg}
\caption{ABB IRB 120 Reaching Target Position}
\end{figure}

%\subsection{Object Following}
\section{Conclusions}
After the results previously shown, we can state that:
\begin{itemize}
    \item Physical model for NN training can be built in CAD software This approach is user-friendly and above all versatile: once the simulation environment is set up minor changes in the model (such as dimensions, positions of mass properties) can be automatically passed to the simulation.
    \item Since all 6DOF robotic arms share the same structure (7 bodies connected by 6 revolute joints), changing the robotic arm model requires minimal modifications.
    Indeed, passing from the Comau Racer3 to the ABB IRB 120 required to modify only the hard coded specifications of the robotic arm.

    The possibility of applying the same algorithm to different robotic arms will be especially welcome when industrial applications of NN-controlled robots will start to spread.
    \item A small number of contact primitives can be effectively used to train the Policy to avoid collisions.

\end{itemize}

Furthermore, we plan to use this approach together with Convolutional Neural Networks fed with  images from simulation render to train visuomotor policies.

%\begin{figure*}
%\centering
%\includegraphics{flies}
%\caption{A sample black and white graphic
%that needs to span two columns of text.}
%\end{figure*}

\section{Acknowledgments}

This research benefits from the HPC (High Performance Computing) facility of the University of Parma, Italy.

\printbibliography

\end{document}

### 2020-05-18 templates/template_phdThesis_ingInf/app1.tex
Added PhD Thesis LaTeX template

\chapter{Esempio di appendice}
\thispagestyle{empty}

\section{Paragrafo 1}
bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla
bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla

\subsection{Sottoparagrafo1}
bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla
bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla

### 2020-05-18 templates/template_phdThesis_ingInf/cap1.tex
Added PhD Thesis LaTeX template

--- 2020-05-18 templates/template_phdThesis_ingInf/app1.tex
@@
-\chapter{Esempio di appendice}
+\chapter{Capitolo 1}
 \thispagestyle{empty}
+
+{                                               %Eventuale citazione
+\flushright\footnotesize
+\textit{Citazione (l1)}\\
+\textit{Citazione (l2)}\\
+\vspace{0.2cm}
+\textrm{-- Autore Citazione}\\
+\vspace{1.0cm}
+}                                               %Fine Eventuale citazione
 
@@
 bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla
+
+\subsection{Sottoparagrafo2}
+bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla
+bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla
+
+\section{Paragrafo 2}
+bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla
+bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla
+
+\subsection{Sottoparagrafo1}
+bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla
+bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla
+
+\subsection{Sottoparagrafo2}
+bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla
+bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla

### 2020-05-18 templates/template_phdThesis_ingInf/introd.tex
Added PhD Thesis LaTeX template

\chapter*{Introduzione}
\markboth{Introduzione}{Introduzione}
\addcontentsline{toc}{chapter}{\numberline{}Introduzione}
\thispagestyle{empty}

Questa è l'introduzione

bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla
bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla bla

### 2020-05-29 courses/computational_mechanics/computational_mechanics.TEX
Added docs for the Computational Mechanics course

%%
%% This document created by A.Tasora
%%

\documentclass{digitaldynamics}

%%
%% Additional packages... --------------------------------------------------
%%

\usepackage{graphicx}
\usepackage{amsmath, amssymb} 
\usepackage[english]{alg}
\usepackage{empheq}
\usepackage{bm}
\usepackage{listings}
\usepackage{nccmath}

\sloppy

\graphicspath{{./pics/}}

%%
%% Custom commands  --------------------------------------------------------
%%
\newtheorem{theorem}{Theorem}
\newtheorem{assumption}{Assumption} 
\newtheorem{proposition}{Proposition} 
\newtheorem{lemma}{Lemma}
\newtheorem{definition}{Definition}
\newtheorem{defn}{Definition}

	 %% vector as boldface   
\def\vect#1{\bm{#1}}
\def\vcovar#1{{\bm{#1}}^*}
\def\vcontr#1{\underline{\bm{#1}}}

\def\tensor#1{\bm{#1}}
\def\etensor#1{\underline{\bm{#1}}}
\def\eetensor#1{\underline{\underline{\bm{#1}}}}

\def\matr#1{{#1}}

\def\mycite#1{\cite{#1}}

\newcommand{\norm}[1]{\left|\left| #1 \right| \right|}

\def\mycorollary{cor}
\def\mytheorem{thm}
\def\myproof{pf}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

%%
%% TITLE
%%

\title{An introduction to \\ Computational mechanics}

\author{Alessandro Tasora \\ alessandro.tasora@unipr.it}

\maketitle
\thispagestyle{fancy}

\begin{abstract}

This document contains some short notes for the seminar on Computational Mechanics
for the PhD students at the University of Parma. It starts covering a review of tensor notation,
then it moves to PDE applications in engineering, such as FEA and CFD, 
highlighting how in many cases they share a common underlying mathematical structure. Then it
presents basic concepts about strong-weak formulations and discretization, and it discusses
numerical methods for solving large scale linear problems. Finally it presents a review
of the most relevant integration methods for time-dependant problems and it discusses the
implementation details of for index-2 ODEs and DAEs that are of particular interest for the mechanical 
engineering community.

Note: this is a first release, it will be extended in future. Please report typos and suggestions for future revisions.

\end{abstract}

%%
%% SECTIONS
%%

\section{Introduction}

Most problems in applied and theoretical mechanics share a common structure, ad require going through three steps:
\begin{itemize}
	\item find a mathematical formulation of the physical problem
	\item find a discretized version of the formulation
	\item implement solvers for the discretized formulation
\end{itemize}

An interesting aspect is that, even if applied to completely different contexts, a problem of fluid dynamics, a problem of heat diffusion or a problem of elasticity might share the same mathematical formulation. Often these formulations are expressed in differential form. This will be discussed in the section on \textit{Classes of problems}.

One of the biggest challenges of computational mechanics (and of computational sciences in general) is how to transform those mathematical formulations into \textit{discretized} problems with a \textit{finite} number of unknowns. This is the rationale of the finite element methods, or the finite volume methods, or the meshless particle methods, for example. This will be discussed in the sections on \textit{Discretization}. Additionally, time-dependant problems might require a time integration process: this is discussed in the \textit{Time integration} section.

Going a step further: once the problem is discretized and turned into a solution procedure, such procedure might embed numerical problems that often fall into a limited number of classes - for instance finite element methods require the solution of a system of linear equations. This brings in some issues: how to efficiently solve a linear system? How to handle very large and sparse matrices? Topics like this are discussed in the section on \textit{Linear solvers} etc.

The following sections represent a primer on computational mechanics, starting from general general mathematical tools such as tensor notation and vector algebra, then showing a taxonomy of mechanical problems, and finally presenting some insight on the numerical methods that can solve computational primitives that often are encountered while solving computational mechanics: sparse large linear systems, performing time integration, etc.

\newpage
\section{Notation, tensors and other mathematical tools}

In the following we list some basic notations that will be used later when dealing with computer methods. 
When possible, we highlight the similarities between the vector/matrix notation and the tensor notation.

\subsection{Symbols}

A cheat sheet with notation that will be used in the following:

\begin{itemize}
 \item $s$ is a scalar
 \item $\vect{v}$ is a generic vector 
 \item $\vcovar{v}$ is a covariant vector (or covariant base vector)
 \item $\vcontr{v}$ is a contravariant vector (or contravariant base vector)
 \item $\matr{A}$ is a matrix
 \item $A_{ij}$ is a (scalar) element of matrix $\matr{A}$
 \item $\tensor{T}$ is a generic tensor
 \item $T^i{}_j{}^k$ is a (scalar) component of a tensor $\tensor{T}$
 \item $\etensor{t}$ is an euclidean tensor, rank 1.
 \item $\eetensor{T}$ is an euclidean tensor, rank 2.
\end{itemize}

\subsection{Index notation}

First, we need to introduce the index notation.

Index notation is used as an alternative to the conventional notation in matrix algebra.
It is used also in tensor formulas, and one might be tempted to think that tensors are just like matrices, i.e. multidimensional
arrays, however tensors are not just like matrices because they convey more information about the underlying coordinate
systems (see contravariance and covariance later), and vice-versa there are matrices that could not be interpreted as tensors.

By the way, index notation (despite more intricate than matrix notation at the first glance) offer more possibilities later when coming to 
formulas for differential geometry.

\begin{itemize}

	\item The \textit{Einstein\footnote{It is remarkable how many bright scientists attended the University of Göttingen.} summation convention} is used: it assumes that repeated indexes 
	are automatically summed, so the $\sum$ symbol can be omitted:
	\begin{align*}
  \vect{v} &= \sum_{i=1}^{3} v_i \vect{e}_i   \quad \Rightarrow \quad   \vect{v} =  v_i \vect{e}_i 
	\end{align*}

	\item With index notation,	one can express, for example, a product of matrix by vector in this way:
	\[
	\vect{v} = \matr{A} \vect{b}   \quad \Rightarrow \quad  v_i = A_{i j} b_j
	\]
	and a product between matrices as:
	\[
	\vect{C} = \matr{A} \matr{B}   \quad \Rightarrow \quad  C_{i j} = A_{i k} B_{k j} 
	\]

	\item The \textit{transpose} of a matrix translates into switching the order of the indexes:
	\[
	C_{i j} = (C^T)_{j i} 
	\]

	\item The \textit{trace} of a matrix translates into:
	\[
	\text{tr} A = A_{i i}
	\]

	\item The \textit{Kronecker delta} is the equivalent of the identity matrix $\matr{I}_{ij}=\delta_{ij}$ in index notation:
	\[
			\delta_{ij} = \begin{cases}
		0 &\text{if } i \neq j,   \\
		1 &\text{if } i=j.   \end{cases}
	\]

	\item With index notation, and with tensors in general, one does not use bold symbols because
	all formulas are effectively speaking about scalars (ex. the $i$-th element of a vector, in formula above, where the fact that $i$ expands into multiple values is implicitly assumed by the \textit{range convention}).

	\item Index notation can be easily translated into computer code: \textit{Einstein convention} and \textit{range convention} lead to \texttt{for} loops, for instance
	\[
	D_{i j} = A_{i k} B_{k l} C_{l j}  \quad \Rightarrow \quad  D_{i j} = \sum_{k=1}^{m} \sum_{l=1}^{m} A_{i k} B_{k l} C_{l j}
	\]
	where the two $\sum$ become two nested \texttt{for} loops with sums (also \textit{reduction} primitives in a parallel computing environment),
	all enclosed in two additional \texttt{for} loops over the $i$ and $j$ indexes for storing results of sums into the $C$ data structure.

	\item The \texttt{for} loops with sums can become \textit{reduction} primitives in a parallel computing environment.

	\item The repeated indexes are called \textit{dummy indexes}.

	\item The non-repeated indexes are called \textit{free indexes}.

	\item In a tensor expression indexes are repeated maximum two times (if dummy indexes).

	\item The \textit{outer product} is also called dyadic product or tensor product. Assuming vectors with basis $\vect{e}_i$, it is:
	\begin{align}
		\vect{a}\otimes\vect{b} &= \vect{a b}^\mathsf{T} = 
		\begin{pmatrix}
			a_1 \\ a_2 \\ a_3
		\end{pmatrix}\begin{pmatrix}
			b_1 & b_2 & b_3
		\end{pmatrix} = \begin{pmatrix}
			a_1 b_1 & a_1 b_2 & a_1 b_3 \\
			a_2 b_1 & a_2 b_2 & a_2 b_3 \\
			a_3 b_1 & a_3 b_2 & a_3 b_3
		\end{pmatrix}. \\
		[\vect{a} \otimes \vect{b}]_{i j} &= a_i b_j
		\\
		\vect{a} \otimes \vect{b} &= a_i b_j \vect{e}_i \otimes \vect{e}_j
	\end{align}

	\item Some properties of dyadic product:
	\begin{align}
    (\mathbf{u} \otimes \mathbf{v})^\mathsf{T} &= (\mathbf{v} \otimes \mathbf{u}) \\
  (\mathbf{v} + \mathbf{w}) \otimes \mathbf{u} &= \mathbf{v} \otimes \mathbf{u} + \mathbf{w} \otimes \mathbf{u} \\
  \mathbf{u} \otimes (\mathbf{v} + \mathbf{w}) &= \mathbf{u} \otimes \mathbf{v} + \mathbf{u} \otimes \mathbf{w} \\
             c (\mathbf{v} \otimes \mathbf{u}) &= (c\mathbf{v}) \otimes \mathbf{u} = \mathbf{v} \otimes (c\mathbf{u})
\end{align}

\end{itemize}

\subsection{Contravariant and covariant transformations}

\begin{itemize}
	\item We introduce a \textit{manifold} $M$ parametrized by $n$ coordinates $\theta_i$.

	\item When changing coordinates from $\theta_i$ to new coordinates $\hat{\theta}_i$, we say that 
	an entity such as a vector $a_i$ changes with a \textit{contravariant transformation rule} if it holds:
	\[
	\hat{a}^i = \frac{\partial \hat{\theta}_i}{\partial \theta_j} a^j
	\]

	\item We say that an entity $a_i$ changes with a \textit{covariant transformation rule} if it holds:
	\[
	\hat{a}_i = \frac{\partial {\theta}_j}{\partial \hat{\theta}_i} a_j
	\]

	\item Indexes denoting covariant components are denoted with \textit{superscripts}. Indexes denoting contravariant components are denoted with \textit{subscripts}. More on this later, when discussing tensors, that can have multiple indexes and mixed contravariant/covariant properties.

	\item Example: the derivative of a scalar function, $a_j=\frac{\partial f(\cdot)}{\partial \theta_j}$, transforms covariantly. Same for base vectors $\vcovar{g}_i = \frac{\partial}{\partial \theta_i}$, tangents to the (maybe curvilinear) grid of coordinates. Differential forms transform contravariantly.

\end{itemize}

\subsection{Coordinate basis}

\begin{itemize}

\item We introduce the \textit{basis}, a set of linearly independent vectors spanning a vector space $V$; usually $V$ is the Euclidean 3D space $\mathbb{R}^3$,

\item We denote a set of \textit{covariant basis vectors} as $\left\{ \vcovar{g}_1, \vcovar{g}_2, \vcovar{g}_3 \right\} = \left\{ \vcovar{g}_i \right\}$, these can be computed as tangents to the coordinate grid: $\vcovar{g}_i = \frac{\partial}{\partial \theta_i}$.

\item Covariant basis vectors transform covariantly as: 
\begin{align*}
	\hat{\vcovar{g}}_i = \frac{\partial {\theta}_j}{\partial \hat{\theta}_i} \vcovar{g}_j
	\label{eq:basiscovariant}
\end{align*}

\item We denote a set of \textit{contravariant basis vectors} as $\left\{ \vcontr{g}^1, \vcontr{g}^2, \vcontr{g}^3 \right\} = \left\{ \vcontr{g}^i \right\}$
such that 
\begin{align}
  \vcontr{g}^i \cdot \vcovar{g}_j = \delta^i_j
\end{align}

\item Contravariant basis vectors transform contravariantly as: 
\begin{align}
	\hat{\vcontr{g}}^i = \frac{\partial \hat{\theta}_i}{\partial \theta_j} \vcontr{g}^j 
	\label{eq:basiscontravariant}
\end{align}

\item Note that, apart from using $\vcontr{g}^i \cdot \vcovar{g}_j = \delta^i_j$, these might be computed also as $\vcontr{g}^i = \nabla_{\vect{r}} \theta_i$ if one could have a function $\vect{\theta}(\vect{r})$ with $\vect{r}$ as manifold point; more often however one has $\vect{r}(\vect{\theta})$. More easily, the contravariant basis, if needed, can be computed just from the covariant basis by passing through the inverse of the metric tensor (see later).

\item A vector is represented as a linear combination of covariant basis, or equivalently, as a linear combination of contravariant basis:
\begin{align*}
	\vect{v} = \; v^i \vcovar{g}_i \; = \; v_i \vcontr{g}^i 
\end{align*}

\item The vector with covariant basis $v^i \vcovar{g}_i$ is called \textit{contravariant vector} because, for representing the same $\vect{v}$ after a change of coordinates, its components $v^i$ must follow the contravariant transformation, hence are \textit{contravariant components}: 
\[
	\hat{v}^i = \frac{\partial \hat{\theta}_i}{\partial \theta_j} v^i
\]
Similarly, the vector with contravariant basis $v_i \vcontr{g}^i$ is called \textit{covariant vector} because its components $v^i$ must follow the contravariant transformation, hence are \textit{covariant components}:
	\[
	\hat{v}_i = \frac{\partial {\theta}_j}{\partial \hat{\theta}_i} v_j
	\]

\item Covariant vectors are also called \textit{covectors}.

\item Covariant and contravariant vectors differ respect to the way they are transformed with a change of basis coordinates.
	For a vector $\vect{v}$ to remain invariant under coordinate changes, the components $v^i$ should transform following a 
	contravariant rule and the components $v_i$ should transform following a covariant rule.

\item Examples. Velocity is a contravariant vector. A gradient is a covariant vector. In fact, say you shrink the basis of a velocity vector by a factor 0.1, you must then scale by a factor of 10 its components to keep it the same. Viceversa, if you shrink the basis by factor 0.1, the gradient scales 0.1 times as well.

\item As a rule of thumb: vectors that contain distance at the numerator in a dimensional expression (ex. velocity = distance / time) transform contravariantly; vectors that have the dimension of distance at the denominator (ex. the gradient) transform covariantly.

\end{itemize}

\subsection{Tensors}

\begin{itemize}

	\item We assume that tensors represent physical entities whose meaning does not change when changing frame of reference, so it
	must come equipped with some type of information about how it is transformed when its basis is changed, just like we did for vectors (which are, indeed, special cases of tensors).

	\item Given basis vectors $\{\vcontr{g}^i\}$ in $\underbar{E}$ and  dual basis vectors $\{\vcovar{g}_{i}\}$ in $E^*$, both of dimension $m$, a $p$-times contravariant and $q$-times covariant \textbf{tensor} $\tensor{T}$ is:
	\begin{align*}
  \tensor{T} = T^{i_1, i_2, ..., i_p}_{j_1, j_2, ..., j_q} 
	\vcovar{g}_{i_1} \otimes \vcovar{g}_{i_2} \otimes ... \otimes \vcovar{g}_{i_p} \otimes 
	\vcontr{g}^{j_1} \otimes \vcontr{g}^{j_2} \otimes ... \otimes \vcontr{g}^{j_p}
	\end{align*}

	\item  The \textit{type} or \textit{valence} of tensor is given by the couple $(p,q)$.

	\item  The \textit{rank} of the tensor is the sum $n=p+q$, also \textit{order} \footnote{Order and rank are often used interchangeably, but some Authors distinguish rank from order where the former does not count repeated indices (for contractions - see later).}.

	\item  The \textit{dimension} of the tensor is $m$, ex. $m=3$ for most problems in three dimensional space, $m=4$ in general relativity, etc.

	\item  The number of \textit{components} is $m^n$.

	\item Tensors with covariant components only, as in $R_{a b}$, are said \textit{covariant} tensors.

	\item Tensors with contravariant components only, as in $R^{a b}$, are said \textit{contravariant} tensors.

	\item Tensors with both covariant and contravariant components, as in $R^a{}_{b c d}$ are said \textit{mixed} contravariant/covariant tensors.

	\item Often the $\vcovar{g}_{i_1} \otimes \vcovar{g}_{i_2} \otimes ... \otimes \vcovar{g}_{i_p} \otimes 
	\vcontr{g}^{j_1} \otimes \vcontr{g}^{j_2} \otimes ... \otimes \vcontr{g}^{j_p}$ dyadic products are omitted and tensors are written
	simply as:
	\[
	   T^{i_1, i_2, ..., i_p}_{j_1, j_2, ..., j_q}
	\]
	We call this the \textit{squeezed} simplified notation. Using this purely indicial notation, an example of tensor could be
	\[
	   T^{ijk}_{lmn}
	\]

	\item Another simplified notation, again purely indicial, uses vertically aligned superscripts and subscripts as, for example,
	\[
	   T^{ij}{}_{lm}{}^{k}{}_{n}
	\]
	In fact we assumed a dyadic product of all covariant bases followed by all contravariant bases, although
	this is arbitrary. For example, as dyadic product is not commutative,  the tensor $T^{ij}_{k} \vcovar{g}_{i} \otimes \vcovar{g}_{j} \otimes \vcontr{g}^{k}$ is not exactly 
	$T^{ij}_{k} \vcovar{g}_{i} \otimes \vcontr{g}^{k} \otimes \vcovar{g}_{j}$, yet if we write them with the squeezed simplified notation we obtain the same symbol $T^{ij}_{k}$, and we loose the information on how we ordered contravariant/covariant bases. 
	For this reason, some authors introduce small spaces in the index super/subscripts to retain the ordering information, so the two examples
		become $T^{ij}{}_{k}$ and $T^{i}{}_{k}{}^{j}$, respectively. For example $R^i{}_j$ is not $R_j{}^i$ unless symmetric, whereas $R^i_j$ would be equivocal.

		In the following we will use simplified notations only in the cases where there is no risk of misunderstanding.

		\item
		\textit{Two point tensors} are tensors where some of the base vectors belong to a coordinate reference A, and some other belong to a coordinate reference B. Often, the distinction is done by using uppercase/lowercase both for the dyads and for the indexes, so for example the former base vectors are $\vcovar{g}_i$ or $\vcontr{g}_j$, and latter are $\vcovar{G}_I$ or $\vcontr{G}_J$. An example is the deformation tensor\footnote{This holds assuming a material point in coordinates of the reference configuration is $\vect{X}=X^I \vcovar{G}_I$, and in current configuration is $\vect{x}=x^i \vcovar{g}_i$, for $d\vect{x}=\tensor{F} d\vect{X}$ or equivalently $dx^i = \frac{\partial x^i}{\partial X^I} dX^I$ .} in nonlinear elasticity:
		\[
		\tensor{F} = F^i{}_I \vcovar{g}_{i} \otimes \vcontr{G}^{I}
		\]

	\item From the point of view of computer implementation, a tensor $\tensor{T}$ can be stored in memory just like a \textit{multi-dimensional array} of
	scalars, ie. 	$T^{i_1, i_2, ..., i_p}_{j_1, j_2, ..., j_q}$, and there is no need to store the basis vectors $\vcontr{g}^i$ and $\vcovar{g}_j$ of the dyadic products. 
	In fact it is implicitly assumed that who write the tensor expression is aware of the basis (whose components can be stored once in a single place, if needed).
	Also there is a binary information per each index (covariant or contravariant?) that can be omitted as it is implicitly assumed that
	who writes the formulas in the computer code is aware of the permitted transformations.

	\item Similarly to vectors, covariant and contravariant components of tensors differ respect to the way they are transformed with a change of basis.
	For a vector $\vect{v}$ to remain invariant under coordinate changes, contravariant components should transform following a 
	contravariant rule and covariant components should transform following a covariant rule. Formally, when changing coordinates $\vect{\theta} \rightarrow \hat{\vect{\theta}}$, one passes from $\tensor{T}$ to $\hat{\tensor{T}}$, hence from 
	$T^{abc}{}_{def} \vcovar{g}_a \otimes \vcovar{g}_b \otimes \vcovar{g}_c \otimes \vcontr{g}^d \otimes \vcontr{g}^e \otimes \vcontr{g}^f$ 
	to  
	$\hat{T}^{abc}{}_{def} \hat{\vcovar{g}}_a \otimes \hat{\vcovar{g}}_b \otimes \hat{\vcovar{g}}_c \otimes \hat{\vcontr{g}}^d \otimes \hat{\vcontr{g}}^e \otimes \hat{\vcontr{g}}^f$
	but for invariance $\tensor{T}=\hat{\tensor{T}}$ to hold, transforming all bases as in \eqref{eq:basiscontravariant} and \eqref{eq:basiscovariant}, one obtains the \textit{transformation rule} of tensor components:
	\begin{align}
  \hat{T}^{abc\ldots}{}_{def\ldots} = {T}^{ijk\ldots}{}_{lmn\ldots} 
			\frac{\partial \hat{\theta}_a}{\partial \theta_i}
			\frac{\partial \hat{\theta}_b}{\partial \theta_j}
			\frac{\partial \hat{\theta}_c}{\partial \theta_k}\ldots
			\frac{\partial \theta_l}{\partial \hat{\theta}_d}
			\frac{\partial \theta_m}{\partial \hat{\theta}_e}
			\frac{\partial \theta_n}{\partial \hat{\theta}_f}\ldots
			\label{eq:trasftensor}
	\end{align}

	\item More succinctly, one can define $\frac{\partial \hat{\theta}_i}{\partial \theta_j} = \beta^i_j$ and $\frac{\partial \theta_i}{\partial \hat{\theta}_j} = \alpha^i_j$, which are basically matrices, like jacobians, and write:
	\begin{align}
  \hat{T}^{abc\ldots}{}_{def\ldots} = {T}^{ijk\ldots}{}_{lmn\ldots} 
			\beta^a_i \beta^b_j \beta^c_k \ldots
			\alpha^l_d \alpha^m_e \alpha^n_f \ldots
			\label{eq:trasftensorab}
	\end{align}

	\item Note that one might obtain $\alpha^i_j$ matrices directly with some simple procedures, without need of passing through $\vect{\theta}$ coordinates and derivatives. For example, a common scenario is a rotation in Cartesian coordinates via a $\mathsf{SO}(3)$ matrix $R$, ie.e a 3x3 array, that would lead to $\alpha^i_j = R^i{}_j$.

	\item One can compute $\beta=\alpha^{-1}$ by matrix inversion with linear algebra, since one can show $\alpha^k_i \beta^j_k = \delta^j_i$. 
	In the example with the rotation matrix, given orthogonality of $R$, one has $R^{-1} = R^{T}$ and  $(R^{T})^i{}_j = R_j{}^i$, so it would be $\beta^i_k = (\alpha^{-1})^i_k = R_j{}^i$.

	\item An example of change of basis with orthogonal transformation (rotation matrix $R$), for a case of a (0,1) tensor like a velocity, corresponding to well known linear algebra $\hat{\vect{v}} = R^{T} \vect{v}$: 
	\begin{align*}
  \hat{v}^{i} = R_j{}^i v^j
	\end{align*}

	\item An example of change of basis with orthogonal transformation (rotation matrix $R$), for a case of a (0,2) tensor like stress tensor, corresponding to well known linear algebra $\hat{\sigma} = R^{T} \vect{\sigma} R$:  
	\begin{align*}
  \hat{\sigma}^{ij} = R_k{}^i R_m{}^j  \sigma^{km}
	\end{align*}

	\item Some examples of tensors.

	\begin{itemize}
		\item $s$ : a scalar is a order 0 tensor,
		\item $v^i$ : velocity is a first order contravariant three dimensional tensor, hence of type $(1,0)$, with $n=3$ components,
		\item $d_i$ : gradients are first order covariant tensors, hence of type $(0,1)$, with $n=3$ components if in 3D space,
		\item $g_{\alpha \beta}$ : the \textit{metric tensor} is a covariant second order tensor, hence of type $(0,2)$
		\item $R_{i j}$ : the Ricci curvature tensor is a covariant second order tensor, of type $(0,2)$
		\item $g^{\alpha \beta}$ : the \textit{inverse metric tensor} is a contravariant second order tensor, of type $(2,0)$
		\item $J^{i j}$ : the \textit{inertia tensor} is a contravariant second order tensor, of type $(2,0)$
		\item $\sigma^{ij}$ : stress tensors (v.Piola-Kirchhoff) are contravariant second order tensors, of type $(2,0)$
		\item $\epsilon_{kl}$ : strain tensors (v.Green-Lagrange) are covariant second order tensors, of type $(0,2)$
		\item ${R^\alpha}_{\beta\gamma\nu}$ : the Riemann\footnote{It is remarkable how many bright scientists attended the University of Göttingen.} curvature tensor is a mixed tensor of type $(1,3)$
		\item ${\delta^i}_j$ : the Kronecker delta tensor is a mixed order tensor, of type $(1,1)$
		\item ${R^i}_j$ : a rotation matrix is a mixed order tensor, of type $(1,1)$
	\end{itemize}

	\item Tensor \textbf{multiplication by a scalar} is associative and commutative, for ex. (omitting dyads):
	\begin{align}
	 a \tensor{T} \; = \; \tensor{T} a  \; = a T^{ij}{}_{k}  
	\label{eq:scalartensor}
	\end{align}

	\item Tensor \textbf{addition and subtraction} is associative and commutative, and can be done only for tensors of same rank and type, for ex. (omitting dyads):
	\begin{align}
	 \tensor{A} + \tensor{B} - \tensor{C} \; = \; {A}^{ij}{}_k + {B}^{ij}{}_k - {C}^{ij}{}_k
	\label{eq:scalartensor}
	\end{align}

	\item Note that tensor addition requires same super/subscripts. Dummy indexes do not count, ex. 
	${A}^{ij}{}_k + {B}^{ij}{}_k + {C}^{mij}{}_{mk}$ is legal too. In fact see contraction, later, for which ${C}^{mij}{}_{mk}={D}^{ij}{}_k$.

	\item An example of tensor expression with addition and scalar multiplication: the Ricci flow (with $g_{ij}$ metric tensor, $R_{ij}$ Ricci curvature):
	\[
	\partial_t g_{ij} + 2 R_{ij} = 0 
	\]

	\item Tensor \textbf{transpose}, defined for 2nd rank tensors, is obtained swapping the order of indexes; this correspond simply to swapping columns and rows in the $T$ matrix of components. Note that also corresponding bases in dyadic products are switched, so the contravariant/covariant property of each index persists:
	\begin{align}
	 ({}^t T)_{ij} = T_{ji} \quad 
	 ({}^t T)^{ij} = T^{ji} \quad
	 ({}^t T)^{i}{}_{j} = T_{j}{}^{i}
	\label{eq:transpose}
	\end{align}

	\item Tensor \textbf{product}, denoted with $\otimes$, takes two vectors of types $(p_a,q_a)$ and $(p_b,q_b)$  
	and returns a tensor of type $(p_a+p_b, q_a+q_b)$, appending dyadic products (and components) as in this example:
	\begin{align}
	 \tensor{A} \otimes \tensor{B} &= \tensor{C} \\
																 &= A^{ij}{}_{k} B^{m}{}_{n} 
																     \vcovar{g}_i \otimes \vcovar{g}_j \otimes \vcontr{g}^k 
																		 \otimes
																		 \vcovar{g}_m \otimes \vcontr{g}^n \\
																 &= C^{ij}{}_{k}{}^{m}{}_{n} 
																		 \vcovar{g}_i \otimes \vcovar{g}_j \otimes \vcontr{g}^k 
																		 \otimes
																		 \vcovar{g}_m \otimes \vcontr{g}^n
	\label{eq:tensorproduct}
	\end{align}
	that is, in simplified indicial notation:
	\begin{align}
						 A^{ij}{}_{k} B^{m}{}_{n} = C^{ij}{}_{k}{}^{m}{}_{n}								 
	\label{eq:tensorproductindicial}
	\end{align}

	\item The tensor product, as in dyadic product seen before, is associative but \textit{not commutative}. Hence, except special cases: 
	\begin{align}
	 \tensor{A} \otimes \tensor{B} \neq \tensor{B} \otimes \tensor{A}
	\label{eq:tensorproduct}
	\end{align}

	\item Note that one can commute the order of terms in the indicial notation of the product without problems, such as in:
	\[
	   A^i{}_j B^k{}_l = B^k{}_l A^i{}_j 
	\]
	and this independence of order of terms holds for all tensor expressions in indicial form. Where is the trick? 
	The reason is that the indicial form deals with simple scalar products, hence commutative, yet the hidden sorcery is that the order of tensor  
	product is preserved \textit{because we do not change the indexes},
	in fact: $\tensor{A} \otimes \tensor{B} = A^i{}_j B^k{}_l = B^k{}_l A^i{}_j = C^{i}{}_j{}^k{}_l$, but 
	$\tensor{B} \otimes \tensor{A} = B^i{}_j A^k{}_l = C^{i}{}_j{}^k{}_l$, note how the same $C^{i}{}_j{}^k{}_l$ term has two different expressions in A and B 
	components depending on the order of multiplication?

	\item This said, by the way, there is no need to prefer a "well ordered" expression $A^i{}_j B^k{}_l = C^{i}{}_j{}^k{}_l$ respect to $B^k{}_l A^i{}_j = C^{i}{}_j{}^k{}_l$, although the former might be more intuitive.

	\item Also the name of indexes are arbitrary, for instance $A^i B^k = C^{ik}$ would work equally well as $A^p B^m = C^{pm}$, and repeated indexes (that will elide because of by Einstein notation, implying contraction - see later) are allowed everywhere as in $A^{iz}{}_{jz} B^k = C^{i}{}_j{}^k$.

  \item Tensor \textbf{contraction} is an operation where two indices in a tensor terms are set the same, hence implying summation over that indices by Einstein convention. The repeated indices and corresponding bases are elided if rewriting the tensor term.
	This means that it takes a tensor $(p,q)$ and returns a tensor of $(p-1,q-1)$ type: 
	\begin{align}
	 \mathrm{contract}_{j,l}(A^{ij}{}_{kl}) = A^{ic}{}_{kc} = A^{i}{}_{k}
	\label{eq:tensorcontraction}
	\end{align}
	Note that choosing the pair of indices to contract is not arbitrary: one must have a "physical" reason to do so.

	\item Contraction can be done only between a covariant and a contravariant index, or viceversa.

	\item In general an index does not repeat more than two times in a tensor term. Expressions like $T^{ii}_{i}$ are illegal.

	\item The \textbf{trace} of a rank-2 matrix in matrix algebra can be written as a contraction using tensors, es:
	\[ 
		\mathrm{tr}(\tensor{A}) = A^i_i
	\]

	\item Tensor \textbf{contracted product}, denoted $\cdot$, is just a shorthand notation for representing a tensor product (that gives an "`inflated"' tensor) followed by a contraction over a pair of indices (that "`deflates"' the tensor), as for example:
	\begin{align}
	 \tensor{C} = \tensor{A} \cdot \tensor{B} = A^{ij}{}_{kl} B^{l}_{m} = C^{ij}{}_{km}
	\label{eq:tensorcontractedproduct}
	\end{align}
	Note that there is no agreed rule on the pair of indices must be contracted, but usually one writes the expressions such that the contracted indices are the last of the first term, and the first index of the second term.

	\item Sometimes a more compact notation is used, where the dot is skipped if the meaning is clear, thus $\tensor{A} = \tensor{B} \cdot \tensor{C} \cdot \tensor{D}$ could be written as 
	\[
	\tensor{A} = \tensor{B} \tensor{C} \tensor{D}
	\]

	\item Tensor contracted product $\cdot$ is not commutative just like the tensor product.

	\item A special case: the contracted product of a covariant $(0,1)$ tensor $\tensor{a} \in \underline{E}$ and a contravariant $(1,0)$ tensor $\vect{b} \in E^*$, where the couple of indices to contract is obvious, returns a scalar, i.e. a tensor of (0,0) type, hence it is equivalent to a scalar product between the two vectors $\vect{a}$ and $\vect{b}$:
	\[
	c = \vect{a} \cdot \vect{b} = a_i b^i
	\]
	Some authors call this a \textbf{tensor dot product}.

	\item A special case: the contracted product of a second order mixed tensor $\tensor{T} \in E^* \otimes \underline{E}$ and a first order tensor $\tensor{v} \in E^*$, is the equivalent of a matrix-by-vector product $\vect{u}=\matr{T}\vect{v}$ with linear algebra:
	\[
	\tensor{u} = \tensor{T} \cdot \vect{v} = T^i{}_j v^j \vcovar{g}_i;
	\]
	Similarly, for the contracted product of a second order mixed tensor $\tensor{T} \in \underbar{E} \otimes E^*$ and a first order tensor $\tensor{w} \in \underline{E}$:
	\[
	\tensor{z} = \tensor{T} \cdot \vect{w} = T_i{}^j w_j \vcontr{g}^i;
	\]

	\item A special case: the contracted product of a second covariant tensor $\tensor{T} \in \underline{E} \otimes \underline{E}$ and two first order tensors $\tensor{u} \in E^*$, $\tensor{v} \in E^*$, that is equivalent to the expression $\vect{u}^T\matr{T}\vect{v}$ with linear algebra:
	\[
	c = \tensor{u} \cdot \tensor{T} \cdot \tensor{v} = T_{ij} u^i v^j
	\]

	\item The previous example also explain that a tensor is also a \textit{multilinear map}: a $p$-contravariant $q$-covariant tensor, 
	$\tensor{T}\in E^*_1 \otimes E^*_2 \otimes \ldots \otimes E^*_p \otimes  \underline{E}_1 \otimes \underline{E}_2 \otimes \ldots \otimes \underline{E}_q$ 
	is a multilinear form on vector space 
	$(\underbar{E})^p \times ({E}^*)^q$, 
	in fact in the above example the tensor $\tensor{T}\in \underline{E} \otimes \underline{E} $ is a multilinear map of type $c=\tensor{T}(\tensor{u},\tensor{v})$ hence a multilinear map $E^* \times E^* \rightarrow \mathbb{R}$.

	\item An example of tensor expression with products and contractions: the Einstein field equation in general relativity, with $R_{\mu \nu}$ Ricci curvature tensor, $R$ Ricci scalar curvature -a scalar given by double contraction $g^{\mu \nu} R_{\mu \nu}$, $g_{\mu \nu}$ is the metric tensor, $\Lambda$ is the cosmological constant, $T_{\mu \nu}$ is the stress-energy tensor ex. depending on mass density, $G$ is Newton gravitation constant, $c$ is speed of light constant:
	\[
	  R_{\mu \nu} - \tfrac{1}{2}R g_{\mu \nu} + \Lambda g_{\mu \nu} = \frac{8 \pi G }{c^4} T_{\mu \nu}
	\]

	\item The \textbf{Kroeneker symbol} $\delta^i_j$ does not perform any transformation on the components, just like in linear algebra with an identity matrix $\matr{A}= \matr{I} \matr{A}$ for $I_{ij}=\delta_{ij}$, but just keep in mind that it changes the name of an index, as in:
	\[
	 a_i \delta^i_j = a_j
	\]

	\item The \textbf{Levi-Civita symbol} $\varepsilon_{a_1 a_2 a_3 \ldots a_n}$ is a pseudotensor defined as
	\begin{align}
	\varepsilon_{a_1 a_2 a_3 \ldots a_n} =
		\begin{cases}
		+1 & \text{if }(a_1 , a_2 , a_3 , \ldots , a_n) \text{ even permutation of } (1,2,3,\dots,n) \\
		-1 & \text{if }(a_1 , a_2 , a_3 , \ldots , a_n) \text{ odd permutation of } (1,2,3,\dots,n) \\
		\;\;\,0 & \text{otherwise}
		\end{cases}
		\label{eq:levicivita}
	\end{align}
	and in the simple two-dimension case, for example, it becomes a 2x2 hemisymmetric matrix $[0,1;-1,0]$. 
	Also note $\varepsilon_{a_1 a_2 a_3 \ldots a_n} = \varepsilon^{a_1 a_2 a_3 \ldots a_n}$.

	\item The \textbf{cross product} of two vectors $\vect{a} \in \mathbf{R}^3$, $\vect{b} \in \mathbf{R}^3$, denoted $\times$ in vector algebra, can be represented using the Levi-Civita symbol and tensor notation - that also generalizes to other dimensions:
	\begin{align}
		\vect{a} \times \vect{b} &= a^j b^k \varepsilon_{i j k} \vcovar{g}_i \\
		[\vect{a} \times \vect{b}]^i &= a^j b^k \varepsilon_{i j k}
	\end{align}

	\item The cross product commutes as
	\[
	\vect{a} \times \vect{b} = - \vect{b} \times \vect{a}
	\]

	\item Just for curiosity:
	\begin{align}
		\vect{a} \cdot (\vect{b} \times \vect{c}) &= a^i b^j c^k \varepsilon_{i j k}
	\end{align}

	\item The Levi-Civita symbol $\varepsilon$ is a \textit{pseudo}tensor because it changes sign after an improper orthogonal transformation of coordinates, ex. a reflection like in a mirror.

	\item The determinant of a second rank tensor can be computed with the Levi-Civita symbol as
	\begin{align}
		\mathrm{det}(\tensor{A}) &= \varepsilon_{i j k \ldots m} A^i_1 A^j_2 A^k_3 \ldots A^m_m
	\end{align}

	\item Tensor \textbf{double contracted product}, denoted $:$, is just a shorthand notation for representing a tensor product followed by two contractions over two pairs of indices, as for example:
	\begin{align}
	 c = \tensor{A} : \tensor{B} = A^{ij} B_{ji} 
	\label{eq:tensordoublecontractedproduct}
	\end{align}
	In most cases this operation is done on two second-rank tensors. 
	Note that most Authors agree to assume the two indexes being ${}^{ij}$ and ${}^{ji}$, that is, swapped, but other Authors assume ${}^{ij}$ and ${}^{ij}$. With out notation the latter would rather $ A^{ij} B_{ij} = \tensor{A} : {}^t\tensor{B}$. For the latter, another symbol could be used:

	\item Tensor \textbf{double contracted product}, denoted $\cdot \cdot$, corresponds to:
	\begin{align}
	 c = \tensor{A} \cdot \cdot \tensor{B} = A^{ij} B_{ij} 
	\label{eq:tensordoublecontractedproductt}
	\end{align}

	\item The \textbf{invariants} of a second order mixed tensor of type (1,1) are scalars computed as
	\begin{align}
	 I_1 &= \mathrm{tr} \tensor{A}  = A^{i}_{i} \\
	 I_2 &= \frac{1}{2} \mathrm{tr} (\tensor{A} \cdot \tensor{A})   = \frac{1}{2}  A^{i}_{j} A^{j}_{i} \\
	 I_3 &= \frac{1}{3} \mathrm{tr} (\tensor{A} \cdot \tensor{A} \cdot \tensor{A})   = \frac{1}{3}  A^{i}_{j} A^{j}_{k}  A^{k}_{i}\\
	 I_4 &= \ldots
	\label{eq:invariants}
	\end{align}

	\item A manifold $( M , \tensor(G) )$ is said a \textit{Riemannian manifold} if it is a differentiable manifold $M$ equipped with an 
	everywhere non-degenerate, smooth, symmetric, positive-definite metric tensor $\tensor(G)$. If the metric tensor is not necessarily 
	positive-definite, it is said \textit{pseudo-Riemannian} manifold: a special case is the Lorentzian manifold in relativity.

	\item The \textbf{metric tensor} is a type (0,2) tensor, a covariant second order tensor defined as
	\begin{align}
	 \tensor{G} = g_{ij} \vcontr{g}^i \otimes \vcontr{g}^j
	\label{eq:metrictensor}
	\end{align}
	where the components are defined as
	\begin{align}
	 g_{ij} = \vcontr{g}^i \cdot \vcontr{g}^j
	\label{eq:metrictensor}
	\end{align}

	\item We can also compute the inverse of the metric tensor as 
	\begin{align}
	 g^{ij} = \vcovar{g}_i \cdot \vcovar{g}_j
	\label{eq:invmetrictensor}
	\end{align}
	and noting that $g_{ij} g^{jk} = \delta_i^j$, this also means that one can compute $g^{ij}$ using linear algebra as a matrix inverse of $g_{ij}$, and vice versa.

	\item Contracted product by the metric tensor $\tensor{G}$ gives an \textbf{index lowering}, ex. getting a covariant tensor from a contravariant vector:
	\begin{align}
	  \tensor {a} = a_i \vcontr{g}^i = g_{ij} b^j \vcontr{g}^i
	\label{eq:lowering}
	\end{align}
	or shortly: $a_i = g_{ij} b^j$.

	\item Contracted product by the inverse metric tensor $\tensor{G}^{-1}$ gives an \textbf{index raising}, ex. getting a contravariant tensor from a covariant vector:
	\begin{align}
	  \tensor {b} = b^i \vcovar{g}_i = g^{ij} a_j \vcovar{g}_i
	\label{eq:raising}
	\end{align}
	or shortly: $b^i = g^{ij} a_j$.

	\item The index lowering/raising via tensor metric is also called \textbf{musical isomorphism}, with notation
	\begin{align}
	  \tensor {a} &= \tensor{b}^\flat \\
		\tensor {b} &= \tensor{a}^\sharp 
	\label{eq:musical}
	\end{align}

	\item The $\mathrm{d}s$ distance on a manifold is defined as
	\begin{align}
	  \mathrm{d}s^2 &= \mathrm{d}\tensor{r} \cdot \tensor{G} \cdot \mathrm{d}\tensor{r} = g_{ij} \mathrm{d}\theta^i \mathrm{d}\theta^j
	\label{eq:distance}
	\end{align}
	also written as a quadratic form with map $G(\mathrm{d}\tensor{r})$ or $\left\langle \mathrm{d}\tensor{r}, \mathrm{d}\tensor{r} \right\rangle_G$.
	For a Cartesian system, with ortonormal basis, it boils down to $\mathrm{d}s^2 = \mathrm{d}r^i \delta_{ij} \mathrm{d}r^i = \sum_i \mathrm{d}r_i^2$.

	\item Note that the metric tensor can be computed as a matrix product of transformation jacobian by its transpose, $G = J^T J$,  
	\[ 
		g_{ij} = \alpha^k_i \alpha^k_j
	\], 
	for example in the case of a surface in 3D space parametrized as $\vect{x}=\vect{x}(\vect{\theta})$ one has $\alpha^i_j = \frac{\partial x_i}{\partial \theta_j}$.

	\item In the special case where the manifold is a surface in 3D space parametrized by $\theta_1, \theta_2$, infinitesimal area is scaled by square root of determinant of the metric tensor as in
	\begin{align}
	  \mathrm{d}A &= \sqrt{\mathrm{det}(g_{ij})} d\theta_1 d\theta_2 \\
		\mathrm{d}A &= \mathrm{det}(\alpha^i_j) d\theta_1 d\theta_2   \quad  \alpha^i_j = \frac{\partial x_i}{\partial \theta_j}
	\label{eq:darea}
	\end{align}
	and this for example allows to compute the area of a surface, given its metric, as $A=\int\int \sqrt{|\mathrm{det}(g_{ij})|} d\theta_1 d\theta_2$. Note that if the local reference is Cartesian, $\mathrm{det}(g_{ij}) = 1$.

	\item In the special case where the manifold is a curved volume in 3D space parametrized by $\theta_1, \theta_2, \theta_3$, infinitesimal volume is scaled by square root of determinant of the metric tensor as in
	\begin{align}
	  \mathrm{d}V &= \sqrt{\mathrm{det}(g_{ij})} d\theta_1 d\theta_2 d\theta_3 \\
		\mathrm{d}V &= \mathrm{det}(\alpha^i_j) d\theta_1 d\theta_2 d\theta_3  \quad  \alpha^i_j = \frac{\partial x_i}{\partial \theta_j}
	\label{eq:dvolume}
	\end{align}

	\item In the special case where the manifold is a surface in 3D space, the metric tensor is also called \textit{first fundamental form} of the surface, $\tensor{I}$, that in an intuitive sense tells how the surface "stretches".

	\item A vector space equipped with an inner product $s = \vect{a} \cdot \vect{b}$ is called an \textit{inner product space}. 
	An inner product $\vect{a} \cdot \vect{a}$ induces a \textit{quadratic form}, if the form is positive we say that the vector space is \textit{Euclidean}.

	\item In a Euclidean space we can define the \textbf{dot product} using the metric tensor as in:
	\[
	   \vect{a} \cdot \vect{b} = g_{ij} a^i b^i 
	\]
	This is sometimes written as a map $G(\vect{a},\vect{b})$.

	\item If a basis is \textit{ortonormal}, it is identical to its dual basis, and $\tensor{G} = \tensor{G}^{-1}$,
	In terms of matrix algebra, $\tensor{G}$ would be a unit matrix, as in $g_{ij} = \delta_{ij}$.

	\item \textit{Cartesian reference systems} have ortonormal basis, thus in a Cartesian reference one could forget the distinction between
	contravariant and covariant components, and super/subindexes could be written all as subscripts. This simplifies the notation quite a bit,
	for example $A^{i}_{j}$ could be written $A_{ij}$, or $C_{ij}{}^{kl}$ could be written $C_{ijkl}$, etc.

	\item The all-subscript notation is often used in computational mechanics where problems are formulated in a Cartesian system. However this simplification
	might not be used, for instance, when using polar coordinates, etc.

\end{itemize}

\subsection{Differential operators}

\begin{itemize}

  \item We introduce the \textbf{Christoffel symbols of the 2nd kind}, denoted as ${\Gamma^k}_{i j}$ and defined as
	\begin{align}
	   \frac{\partial \vcovar{g}_s}{\partial \theta^n} =  {\Gamma^p}_{sn} \vcovar{g}_p  
	\label{eq:christoffel2}
	\end{align}
	intuitively representing how a basis is changing along coordinates of a manifold, expressed in the basis itself.

	\item We introduce the \textbf{Christoffel symbols of the 1st kind}, denoted as ${\Gamma}_{i j k}$ and defined as
	\begin{align}
	   \frac{\partial \vcontr{g}^s}{\partial \theta^n} =  {\Gamma}_{p s n} \vcontr{g}^p 
	\label{eq:christoffel1}
	\end{align}
	intuitively representing how a basis is changing along coordinates of a manifold, expressed along the dual basis.
	One has $\Gamma_{psn} = g_{pd} {\Gamma^d}_{sn}$.

	\item Sometimes Christoffel symbols of the 2nd kind ${\Gamma^p}_{s n}$ are written as $\left\{{}^p_{s n} \right\}$. 
	\item Sometimes Christoffel symbols of the 1st kind ${\Gamma}_{p s n}$ are written as $[s n,p]$.

	\item One can compute Christoffel symbols of 2nd kind, if one knows $\vect{\theta}=\vect{\theta}(\vect{x})$ with $\vect{x}$ coordinates of a Cartesian system, as:
	\[
		{\Gamma^p}_{sn} = \frac{\partial^2 x^m}{\partial \theta^s \partial \theta^n} \frac{\partial \theta^p}{\partial x^n}
	\]
	or directly from the metric tensor as:
	\[
	 {\Gamma}_{cab} = \frac{1}{2} \left(\frac{\partial g_{ca}}{\partial x^b} + \frac{\partial g_{cb}}{\partial x^a} - \frac{\partial g_{ab}}{\partial x^c} \right)
	\]

	\item The \textbf{covariant derivative} of a contravariant vector field $\vect{u} = u^i \vcovar{g}_i$ is 
	$\frac{\partial \vect{u}}{\partial \theta^n} = \frac{\partial u^i}{\partial \theta^n} \vcovar{g}_i + u^i \frac{\vcovar{g}_i}{\partial \theta^n}$, hence:
	\begin{align}
	  \nabla_{j} \vect{u} = \left(\frac{\partial u^k}{\partial \theta^j} + u^i {\Gamma^k}_{ij} \right) \vcovar{g}_k
	\label{eq:covariantderivative}
	\end{align}

	\item The \textbf{covariant derivative} of a covariant vector field $\vect{u} = u_i \vcontr{g}^i$ is 
	\begin{align}
	  \nabla_{j} \vect{u} = \left(\frac{\partial u_k}{\partial \theta^j} - u_i {\Gamma^i}_{k j} \right) \vcontr{g}^k
	\label{eq:covariantderivativecovector}
	\end{align}

	\item Note that the Christoffel symbols in \eqref{eq:covariantderivative} \eqref{eq:covariantderivativecovector} account
	for the fact that the basis might change direction as in curved manifolds. In a straight manifold (ex. in a Cartesian reference) they vanish to zero.

	\item The \textbf{covariant directional derivative} of a contravariant $\vect{u}$ along a direction $\vect{v}$ is:
	\begin{align}
	  \nabla_{\vect{v}} \vect{u} = \left( v^j \frac{\partial u^k}{\partial \theta^j} + v^j u^i {\Gamma^k}_{i j} \right) \vcontr{g}_k
	\label{eq:directionalderivative}
	\end{align}
	and the covariant directional derivative of a covariant $\vect{u}$ along a direction $\vect{v}$ is:
	\begin{align}
	  \nabla_{\vect{v}} \vect{u} &= \left( v^j \frac{\partial u_k}{\partial \theta^j} - v^j u_i {\Gamma^i}_{k j} \right) \vcontr{g}^k
	\label{eq:directionalderivativecovector}
	\end{align}

	\item For a more compact notation, we introduce the following semicolon ";" notation for the (components of) the covariant derivative:
	\begin{align}
	   u^k{}_{;j} &= \left(\frac{\partial u^k}{\partial \theta^j} + u^i {\Gamma^k}_{i j} \right) \\
		 u_k{}_{;j} &= \left(\frac{\partial u_k}{\partial \theta^j} - u_i {\Gamma^i}_{k j} \right)
	\end{align}

	To sum up things: alternative notations for the covariant derivative:
	\begin{align}
	  \nabla_j \vect{u} = \nabla_{\vect{g}_j} \vect{u} = u^i|_j \vcovar{g}_i = u^i{}_{;j} \vcovar{g}_i
	\label{eq:covariantderivative3}
	\end{align}
	Moreover, some Authors call only the component part $u^i{}_{;j}$ as "covariant derivative", not including $\vcovar{g}_i$.

	\item One can see the covariant derivative as an extended version of the partial derivative, the latter often being denoted with a "," in shorthand
	notation instead of semicolon ";": 
	\begin{align}
	  \frac{\partial u^i}{\partial \theta^j} = \partial_j u^i = u^i{}_{,j} 
	\label{eq:partialderivative}
	\end{align}

	\item Using the simplified notation above, one can express the following special cases:

	\begin{itemize}
		\item The covariant derivative of a scalar field is also the partial derivative regardless of the curvature,
		\[
		   \phi_{,j} = \phi_{;j} =  \partial_j \phi  
		\]

		\item The covariant derivative of a contravariant vector field:
		\[
		   {u}^i{}_{;j} = u^i{}_{,j} + u^k {\Gamma^i}_{kj} 
		\]

		\item The covariant derivative of a covariant vector field:
		\[
		   {u}_i{}_{;j} = u_i{}_{,j} -  u_k {\Gamma^k}_{ij}
		\]

		\item The covariant derivative of a contravariant vector field in a Cartesian system simplifies to:
		\[
		   {u}^i{}_{;j} = u^i{}_{,j} 
		\]
		The covariant derivative of a covariant vector field in a Cartesian system simplifies to:
		\[
		   {u}_i{}_{;j} = u_i{}_{,j} 
		\]

		\item In a Cartesian reference, as we have seen before, contravariant and covariant basis are the same so we 
		can take a further simplification and write indifferently contravariant and covariant vector components with
		subscripts anyway, as often happens in computational mechanics where derivatives are written as $a_{ij,k}$, $v_{i,j}$, etc.

	\end{itemize}

	\item The covariant derivative for a generic (p,q) type tensor can be expressed as
	\[
		 \nabla_{\vect{g}_n} \tensor{T} = T^{ij..}{}_{kl.. ;n} \vcovar{g}_i \otimes \vcovar{g}_j \ldots \otimes \vcontr{g}^k \otimes \vcontr{g}^l \ldots
	\]
	and again, introducing the conventional partial derivatives $T^{ij..}{}_{kl.. ,n} = \frac{\partial T^{ij..}{}_{kl..} }{\partial \theta^n}$, one corrects them with the Christoffel symbols as:
	\begin{align}
	T^{ij..}{}_{kl.. ;n} =& T^{ij..}{}_{kl.. ,n} \\% \frac{\partial T^{ij..}{}_{kl..} }{\partial \theta^n} 
	 &+ T^{sj..}{}_{kl..} {\Gamma^i}_{sn} 
	 + T^{is..}{}_{kl..} {\Gamma^j}_{sn} 
	 + \ldots \\
	 &- T^{ij..}{}_{sl..} {\Gamma^s}_{kn} 
	 - T^{ij..}{}_{ks..} {\Gamma^s}_{ln} 
	 - \ldots
	\end{align}

	\item The $\nabla$ is a \textit{connection}, see the \textit{Levi-Civita connection} in Riemannian geometry.

	\item In the special case where the manifold is a surface in 3D space, the \textit{second fundamental form} is the tensor $\tensor{I\!I}$, and this 
	is related to the covariant derivative as 
	$\tensor{I\!I}(\vect{u},\vect{v}) = \left\langle \nabla_{\vect{u}} \vect{v}, \vect{n} \right\rangle$, 
	with $\vect{n}$ normal vector field to the manifold.

	\item The \textbf{gradient} operator is obtained with the covariant derivative.
	Some examples in different notations in the following; note the added tensor product by $\vcontr{g}^{j} \otimes \ldots$ in the dyads.

	\begin{itemize}
		\item The gradient of a scalar is a covariant vector:
		\[
				 \mathrm{grad}(\phi) = \nabla \phi = \phi_{\textcolor{blue}{;i}} \textcolor{blue}{\vcontr{g}^{i}}
		\]

		\item The gradient of a contravariant vector is a rank-2 tensor of mixed type (1,1):
		\[
				 \mathrm{grad}(\vect{v}) = \nabla \vect{v} = v^i{}_{\textcolor{blue}{;j}} \; \textcolor{blue}{\vcontr{g}^{j}} \otimes \vcovar{g}_{i}
		\]

		\item The gradient of a covariant vector is a rank-2 tensor of covariant type (0,2):
		\[
				 \mathrm{grad}(\vect{v}) = \nabla \vect{v} = v_i{}_{\textcolor{blue}{;j}} \; \textcolor{blue}{\vcontr{g}^{j}} \otimes \vcontr{g}^{i}
		\]

		\item The gradient of a n-rank tensor of type (p,q), in general, is a (n+1) rank tensor of type (p,q+1):
		\[
					\mathrm{grad}(\tensor{T}) = \nabla \tensor{T}  = T^{ij..}{}_{kl.. \textcolor{blue}{;n}} \; \textcolor{blue}{\vcontr{g}^n} \otimes \; \vcovar{g}_i \otimes \vcovar{g}_j \ldots \otimes \vcontr{g}^k \otimes \vcontr{g}^l \ldots
		\]
	\end{itemize}

	\item The \textbf{divergence} operator takes a (n) rank tensor and returns a (n-1) rank tensor, 
	basically it is a contraction -usually on the first index- after the covariant derivative.
	Some examples in different notations in the following.

	\begin{itemize}
		\item The divergence of a scalar field is not defined.

		\item The divergence of a contravariant vector field is a scalar: 
		\[
				 \mathrm{div}(\vect{v}) = \nabla \cdot \vect{v} = \phi^i_{\textcolor{blue}{;i}}
		\]

		\item The divergence of a covariant vector field is a scalar - note the metric tensor for index raising to allow contraction: 
		\[
				 \mathrm{div}(\vect{v}) = \nabla \cdot \vect{v} = \textcolor{blue}{g^{ni}} \phi_i{}_{\textcolor{blue}{;n}} 
		\]

		\item The divergence of a tensor field in general - note the covariant basis vector $\vcovar{g}_i$ that disappears from dyads:
		\begin{align}
					\mathrm{div}(\tensor{T}) = \nabla \cdot \tensor{T}  
					&= T^{ij..}{}_{kl.. \textcolor{blue}{;i}} \; \vcovar{g}_j \ldots \otimes \vcontr{g}^k \otimes \vcontr{g}^l \ldots \\
					&= \textcolor{blue}{g^{ni}} T_i{}^{j..}{}_{kl.. \textcolor{blue}{;n}} \; \vcovar{g}_j \ldots \otimes \vcontr{g}^k \otimes \vcontr{g}^l \ldots
		\end{align}
	\end{itemize}

	\item The \textbf{rotor} operator takes a (n) rank tensor and returns a (n) rank tensor, 
	using the Levi-Civita symbol.
	The rotor of a tensor in general is:
		\begin{align}
					\mathrm{curl}(\tensor{T}) = \nabla \times \tensor{T}  
					&= \vcontr{g}^n \times T^{ij..}{}_{kl.. \textcolor{blue}{;n}} \; \vcovar{g}_i \otimes \vcovar{g}_j \ldots \otimes \vcontr{g}^k \otimes \vcontr{g}^l \ldots \\
					&= \textcolor{blue}{\varepsilon_{n i m}} T^{ij..}{}_{kl.. \textcolor{blue}{;l}} \; \textcolor{blue}{g^{nl}} \vcontr{g}^m \otimes \vcovar{g}_i \otimes \vcovar{g}_j \ldots \otimes \vcontr{g}^k \otimes \vcontr{g}^l \ldots
		\end{align}

	Some special cases in the following.

	\begin{itemize}
		\item The rotor of a contravariant vector is a covariant vector: 
		\[
				 \mathrm{curl}(\vect{v}) = \nabla \cdot \vect{v}
				   = \varepsilon_{ijk} v^j{}_{;n} g^{ni} \vcontr{g}^k
		\]

		\item The rotor of a covariant vector is a contravariant vector: 
		\[
				 \mathrm{curl}(\vect{v}) = \nabla \cdot \vect{v}
				   = \varepsilon^{ijk} v_j{}_{;n} \vcovar{g}_k
		\]

	\end{itemize}

	\item The \textbf{Laplacian} operator $\nabla^2$ takes a (n) rank tensor and returns a (n) rank tensor,
	and is defined in different notations 
	\[
		\nabla^2 \tensor{T} \; = \delta \tensor{T} \; = \; \nabla \cdot \nabla \tensor{T} \; = \; \mathrm{div} (\mathrm{grad} \tensor{T})
	\]
	Some examples in the following:

	\begin{itemize}
		\item The Laplacian of a scalar field, hence divergence of a gradient vector, is a scalar:
		\[
				  \nabla^2 \phi = \phi_{\textcolor{blue}{;ii}}
		\]

		\item The Laplacian of a contravariant vector field is a contravariant vector: 
		\[
				 \nabla^2  \vect{v} = v^i{}_{;nl} g^{nl} \vcovar{g}_i
		\]

		\item The Laplacian of a covariant vector field is a covariant vector:
		\[
				 \nabla^2  \vect{v} = v_i{}_{;nl} g^{nl} \vcontr{g}^i
		\]

		\item The Laplacian of a tensor field in general is again a tensor of same type:
		\begin{align}
					\nabla^2 \tensor{T} = \nabla \cdot \nabla \tensor{T}  
					&= T^{ij..}{}_{kl.. \textcolor{blue}{;nl}} \; \textcolor{blue}{g^{nl}} \vcovar{g}_i \otimes \vcovar{g}_j \ldots \otimes \vcontr{g}^k \otimes \vcontr{g}^l \ldots 
		\end{align}
		where the components of the double covariant derivative are represented by the complex expression
		\begin{align}
		T^{ij..}{}_{kl.. \textcolor{blue}{;nl}} =& T^{ij..}{}_{kl.. ,nl}  \\
	 &+ T^{sj..}{}_{kl..;l} {\Gamma^i}_{sn} + T^{sj..}{}_{kl..} \frac{\partial{\Gamma^i}_{sn}}{\partial \theta^l} \\
	 &+ T^{is..}{}_{kl..;l} {\Gamma^j}_{sn} + T^{is..}{}_{kl..} \frac{\partial{\Gamma^j}_{sn}}{\partial \theta^l} \\
	 &+ \ldots 
		\end{align}
		Hovever the expression above simplifies in case of Cartesian coordinate systems as:
		\[
		T^{ij..}{}_{kl.. \textcolor{blue}{;nl}}  = T^{ij..}{}_{kl.. ,nl}  = \frac{\partial^2 T^{ij..}{}_{kl}}{\partial \theta^n \partial \theta^l}
		\]
	\end{itemize}

\end{itemize}

\subsection{Special case: Cartesian coordinates}

Many problems in computational mechanics are expressed in Cartesian coordinates, for example finite element problems, multibody problems, etc. Under this assumption, many of the
expressions of the previous sections will simplify, in fact in a Cartesian coordinate system one has ortonormal base vectors, hence: 
covariant and contravariant bases are cohincident,
the $g_{ij}$ and $g^{ij}$ metric tensors can be eliminated from formulas (or just them in place enough to raise/lower indices for formal correctness), Christoffel symbols can be omitted in
derivatives.

In this simplified setting, one has the following expressions (where we use subscripts regardless if components are covariant or contravariant):

\begin{table}[]
\centering
\caption{Table of simplified formulas for Cartesian coordinate system}
\label{tab:my-table}
\begin{tabular}{lll}
\hline\hline
& Tensor notation     &    Indicial notation  \\ 
\hline\hline
\scriptsize inner product & $\vect{u} \cdot \vect{v} = u_i v_i$     &    $\vect{u} \cdot \vect{v} = u_i v_i$  \\ 
\scriptsize outer product & $\vect{u} \otimes \vect{v} = u_i v_j  \vect{g}_i \otimes \vect{g}_j$  &   $[\vect{u} \otimes \vect{v}]_{ij} = u_i v_j$   \\
\scriptsize cross product & $\vect{u} \times \vect{v} = u_i v_j \varepsilon_{ijk} \vect{g}_k$  &   $[\vect{u} \times \vect{v}_{k}]_k = u_i v_j \varepsilon_{ijk}$   \\
\hline
\scriptsize matrix product     & $\tensor{A} \vect{v} = A_{ij} v_j \vect{g}_i$  &   $[\tensor{A} \vect{v}]_i = A_{ij} v_j $   \\
\scriptsize matrix product     & $\tensor{A}^T \vect{v} = A_{ji} v_j \vect{g}_i$  &   $[\tensor{A}^T \vect{v}]_i = A_{ji} v_j $   \\
\scriptsize matrix product     & $\vect{u}^T \tensor{A}^T \vect{v} = A_{ji} u_i v_j$    &   $\vect{u}^T \tensor{A}^T \vect{v} = A_{ji} u_i v_j $    \\
\scriptsize matrix product     & $\tensor{A}\cdot\tensor{B} = A_{ij} B_{jk} \vect{g}_i \otimes \vect{g}_j$  &  $[\tensor{A}\cdot\tensor{B}]_{ij} = A_{ij} B_{jk}$ \\
\scriptsize matrix trace       & $\mathrm{tr}(\tensor{A}) = A_{ii}$ &  $\mathrm{tr}\tensor{A} = A_{ii}$ \\
\scriptsize matrix determinant & $\mathrm{det}(\tensor{A}) = \varepsilon_{ijk} A_{i1} A_{j2} A_{j3}$  &  $\mathrm{det}(\tensor{A}) = \varepsilon_{ijk} A_{i1} A_{j2} A_{j3}$ \\
\hline
\scriptsize $\mathrm{grad}(\phi)$          & $ \nabla \phi = \phi_{,i}\;\vect{g}_i$ &  $[\nabla \phi]_i = \phi_{,i}$ \\
\scriptsize $\mathrm{grad}(\vect{v})$      & $ \nabla \vect{v} = v_{i,j}\;\vect{g}_i\otimes\vect{g}_j$ &  $[\nabla \vect{v}]_{ij} = v_{i,j}$ \\
\scriptsize $\mathrm{grad}(\tensor{A})$    & $ \nabla \tensor{A} = A_{ij,k}\;\vect{g}_i\otimes\vect{g}_j\otimes\vect{g}_k$   &  $[\nabla \tensor{A}]_{ijk} = A_{ij,k}$ \\
\scriptsize $\mathrm{div}(\vect{v})$       & $ \nabla \cdot \vect{v} = v_{i,i}$ &  $ \nabla \cdot \vect{v} = v_{i,i}$ \\
\scriptsize $\mathrm{div}(\tensor{A})$     & $ \nabla \cdot \tensor{A} = A_{ij,j} \;\vect{g}_i$ &  $\nabla \cdot \tensor{A} = A_{ij,j}$ \\
\scriptsize $\mathrm{curl}(\vect{v})$      & $ \nabla \times \vect{v} = v_{i,j} \varepsilon_{jik} \vect{g}_k$ &  $  [\nabla \times \vect{v}]_k = v_{i,j} \varepsilon_{jik}$ \\
\scriptsize $\mathrm{curl}(\tensor{A})$    & $ \nabla \times \tensor{A} = A_{ij,k} \varepsilon_{kjl} \;\vect{g}_i \otimes \vect{g}_l$ &  $[\nabla \times \tensor{A}]_{il} = A_{ij,k} \varepsilon_{kjl} \;\vect{g}_i \otimes \vect{g}_l$ \\
\scriptsize $\mathrm{div} (\mathrm{grad} (\phi)) $     & $ \nabla \cdot \nabla \phi = \nabla^2 \phi = \phi_{,jj}$ &  $ \nabla \cdot \nabla \phi = \phi_{,ii}$ \\
\scriptsize $\mathrm{div} (\mathrm{grad} (\vect{v})) $ & $ \nabla \cdot \nabla \vect{v} = \nabla^2 \vect{v}  = v_{j,ii}  \vect{g}_j$ &  $ [\nabla \cdot \vect{v}]_j = v_{j,ii}$ \\
 &  \\ 
\hline\hline
\end{tabular}
\end{table}

\subsection{Properties}

Here is a list of important properties of vector calculus, as often met in physical problems in Cartesian coordinates with vectors $\mathbf{v},\mathbf{u},$ etc. and scalars $\psi, \phi,$ etc.

\textbf{Gradient properties}

\begin{align}
\nabla(\psi+\phi) &=\nabla\psi+\nabla\phi \\
\nabla(\psi \phi) &= \phi\nabla \psi  + \psi \nabla \phi \\
\nabla(\psi \mathbf{v} ) &= \nabla \psi \otimes \mathbf{v} + \psi \nabla \mathbf{v} \\
\nabla(\mathbf{v} \cdot \mathbf{u}) &= (\mathbf{v} \cdot \nabla)\mathbf{u} + (\mathbf{u} \cdot \nabla)\mathbf{v} + \mathbf{v} \times (\nabla \times \mathbf{u})
 + \mathbf{u} \times (\nabla \times \mathbf{v}) 
\end{align}

\textbf{Divergence properties}

\begin{align}
\nabla\cdot(\mathbf{v}+\mathbf{u}) &= \nabla\cdot\mathbf{v}+\nabla\cdot\mathbf{u} \\
\nabla\cdot\left(\psi\mathbf{v}\right) &= \psi\nabla\cdot\mathbf{v}+\mathbf{v}\cdot\nabla \psi \quad \leftarrow \textrm{see use in weak formulations} \\ 
\nabla\cdot\left(\mathbf{v}\times\mathbf{u}\right) &= (\nabla\times\mathbf{v})\cdot \mathbf{u}-(\nabla\times\mathbf{u})\cdot \mathbf{v}  \\
\nabla\cdot(\nabla \times \mathbf{v}) &= 0
\end{align}

\textbf{Curl properties}

\begin{align}
\nabla\times(\mathbf{v}+\mathbf{u}) &=\nabla\times\mathbf{v}+\nabla\times\mathbf{u} \\
\nabla\times\left(\psi\mathbf{v}\right) &=\psi\,(\nabla\times\mathbf{v})+\nabla\psi\times\mathbf{v} \\
\nabla\times\left(\psi\nabla\phi\right) &= \nabla \psi \times \nabla \phi \\
\nabla\times\left(\mathbf{v}\times\mathbf{u}\right) &= \mathbf{v}\left(\nabla\cdot\mathbf{u}\right)-\mathbf{u} \left( \nabla\cdot\mathbf{v}\right)+\left(\mathbf{u}\cdot\nabla\right)\mathbf{v}- \left(\mathbf{v}\cdot\nabla\right)\mathbf{u} \\
\nabla\times(\nabla\psi) &= \mathbf{0}
\end{align}

\textbf{Laplacian properties}

\begin{align}
\nabla^2\psi &= \nabla \cdot (\nabla\psi)\\ 
\nabla^2\mathbf{v} &= \nabla\left(\nabla \cdot \mathbf{v}\right) - \nabla \times \left(\nabla \times \mathbf{v}\right) \\
\nabla \cdot (\phi\nabla\psi) &= \phi\nabla^2\psi + \nabla\phi \cdot \nabla\psi \\
\psi\nabla^2\phi - \phi\nabla^2\psi &= \nabla \cdot \left(\psi\nabla\phi - \phi\nabla\psi\right) \\
\nabla^2(\phi\psi) &= \phi\nabla^2\psi + 2(\nabla\phi) \cdot(\nabla\psi) + \left(\nabla^2\phi\right)\psi \\
\nabla^2(\psi\mathbf{v}) &= \mathbf{v}\nabla^2\psi + 2(\nabla\psi \cdot \nabla)\mathbf{v} + \psi\nabla^2\mathbf{v} \\
\nabla^2(\mathbf{v} \cdot \mathbf{u}) &= \mathbf{v} \cdot \nabla^2\mathbf{u} - \mathbf{u} \cdot \nabla^2\!\mathbf{v} + 2\nabla \cdot ((\mathbf{u} \cdot \nabla)\mathbf{v} + \mathbf{u} \times (\nabla \times \mathbf{v})) 
\end{align}

\subsection{Integral theorems}

In the following we present some formulas that are often used in computational mechanics, ex. to transform surface integrals into volume integrals etc. 
We show that using the tensor notation one can generalize the divergence theorem (aka Ostrogradsky–Gauss\footnote{It is remarkable how many bright scientists attended the University of Göttingen.} theorem) and the Stokes theorem (aka Stokes-Cartan theorem).

\begin{itemize}

\item The \textbf{divergence theorem} states that, given a continuously differentiable vector field 
$\vect{v}$ on a compact volume $V$, with piecewise smooth boundary 
$S = \partial V$ with normal $\vect{n}$, it holds:
\begin{align}
 \int_V \nabla \cdot \vect{v} dV = \int_{S} \vect{v} \cdot \vect{n} \, dS
\end{align}
We can express this with tensor notation and generalize it:
\begin{itemize}

	\item The divergence theorem for vector fields:
	\begin{align}
		\int_V v_{i,i} dV = \int_{S} v_i n_i \, dS
	 \end{align}

	\item The divergence theorem for rank-two tensor fields:
	\begin{align}
		\int_V T_{ij,j} dV = \int_{S} T_{ij} n_j \, dS
	 \end{align}

\end{itemize}

\item The \textbf{Stokes theorem} states that, given a continuously differentiable vector field $\vect{v}$ on a surface $S$, with piecewise smooth contour boundary $C = \partial S$ with tangent $\vect{r}$, it holds:
\begin{align}
 \int_S (\nabla \times \vect{v}) \cdot \vect{n} \, dS = \int_{C} \vect{v} \cdot d\vect{r}
\end{align}
We can express this with tensor notation and generalize it:
\begin{itemize}

	\item The Stokes theorem for vector fields:
	\begin{align}
		\int_S \varepsilon_{ijk} v_{k,j} n_i \, dS = \int_{C} v_i dx_i
	 \end{align}

	\item The Stokes theorem for rank-two tensor fields:
	\begin{align}
		\int_S \varepsilon_{ijk} T_{kl,j} n_i \, dS = \int_{C} T_{il} dx_i
	 \end{align}

\end{itemize}

\end{itemize}

\newpage
\section{Classes of problems}

Thie following is a short overview of important special cases of PDEs.

\subsection{Partial differential equations}

Given an unknown function $u(\vect{x},t)$, depending on independent variables $\vect{x}$ where one can have time $t$ among these, in the most 
general case a \textbf{Partial Differential Equation (PDE)} is a (possibly nonlinear)
equation that include partial derivatives of $u$ as in:
\[
f \left (x_1, \ldots x_n, u, \frac{\partial u}{\partial x_1}, \ldots \frac{\partial u}{\partial x_n}, \frac{\partial^2 u}{\partial x_1 \partial x_1}, \ldots \frac{\partial^2 u}{\partial x_1 \partial x_n}; \ldots \right) = 0 
\]
If $f$ depends only linearly to the partial derivatives, this becomes a \textit{linear PDE}, as in the Laplace equation, Poisson equation, etc., and it is a case of big interest.

The \textit{degree} of the PDE is the highest derivative degree, usually is 2 in many cases of practical interest.

In most cases, PDEs are paired with \textit{boundary conditions}. Assume a boundary $\partial \Omega$ of the $\Omega$ domain, and assume that the boundary can be split in disjoint parts $\Gamma_1, \Gamma_2, ...$ with $\Gamma_i \in \partial\Omega$, $\partial\Omega = \cup_i \Gamma_i$, one can impose one or multiple (mixed) boundary conditions of the type:

\begin{itemize}
	\item The \textbf{Dirichlet\footnote{It is remarkable how many bright scientists attended the University of Göttingen.} boundary condition} imposes the value $b$ of the function on a part of the boundary:
	\begin{align}
	u(\vect{x}) = b(\vect{x})  \quad \forall \vect{x} \in \Gamma_i
	\label{dirichlet}
	\end{align}

	\item The \textbf{Neumann \footnote{It is remarkable how many bright scientists attended the University of Göttingen.} boundary condition} imposes the value $d$ of the derivative of the function on a part of the boundary:
	\begin{align}
	\frac{\partial{u(\vect{x})}}{\partial \vect{n}} = d(\vect{x}) \quad \forall \vect{x} \in \Gamma_i
	\label{dirichlet}
	\end{align}

	\item The \textbf{Robin boundary condition} imposes the a linear combination of the function and its derivative on a part of the boundary:
	\begin{align}
	a_1 u(\vect{x})  + a_2 \frac{\partial{u(\vect{x})}}{\partial \vect{n}} = r(\vect{x}) \quad \forall \vect{x} \in \Gamma_i
	\label{robin}
	\end{align}

	\item The \textbf{Cauchy boundary condition} imposes at the same time the value of the function and its derivative on a part of the boundary:
	\begin{align}
	\left\{
	\begin{aligned}
		u(\vect{x}) &= b(\vect{x}) \\
		\frac{\partial{u(\vect{x})}}{\partial \vect{n}} &= d(\vect{x})
	\end{aligned}
	\right.
	\quad \forall \vect{x} \in \Gamma_i
	\label{cauchy}
	\end{align}

\end{itemize}

In the following we will see how few classes of PDE can cover different physical problems under similar mathematical structures.

\subsection{Poisson equation}

The \textbf{Poisson equation} is a PDE of the following type:
\begin{align}
\nabla^2 u = f
\end{align}
where $u$ is the unknown function $u(\vect{x})$ (usually scalar, but also tensor-valued) and $f$ is a given term (scalar, but also tensor-valued) over a manifold,
and it is often paired with Dirichlet and Neumann boundary conditions.

In the following we list some physical problems that share the same mathematical structure of the Poisson equation, 
all assumed in a 3D domain $\Omega$, with $\vect{x} \in \Omega \subset \mathbb{R}^3$:

\begin{itemize}
	\item The \textbf{steady state heat equation}:
	\[
	k\nabla^2 T = -q
	\]
	where $T(\vect{x})$ is the unknown temperature field in [K] units, 
	$k$ is the thermal conductivity [W/(m K)], 
	$q$ is the heat-flux [W/m${}^3$] of sources in $\Omega$ if any - for example heat from uranium fission. 
	\begin{itemize}
		\item a Dirichlet boundary condition $T=T_0, \; \forall \vect{x}\in\Gamma_i$ means imposing a temperature on a boundary surface $\Gamma_i$, 
		\item a Neumann   boundary condition $-k\partial{T}/\partial{\vect{n}}=q_0 \; \forall \vect{x}\in\Gamma_i$ means imposing a heat flux $q_0$ [W/m${}^2$] through a boundary zone $\Gamma_i$, remembering $\vect{q}=-k \nabla T$.
	\end{itemize}

	\item The \textbf{electrostatics equation}:
	\[
	\epsilon \nabla^2 \phi = -\rho
	\]
	where $\phi(\vect{x})$ is the unknown scalar electric potential field in [V] units, 
	$\epsilon$ is the permittivity [F/m] of the material,
	$\rho$ is the total volume charge density of sources in $\Omega$, in [C/m$^3$], if any. 
	\begin{itemize}
		\item a Dirichlet boundary condition $\phi=\phi_0, \; \forall \vect{x}\in\Gamma_i$ means imposing an electric potential [V] on a boundary surface $\Gamma_i$, 
		\item a Neumann   boundary condition $\partial{\phi}/\partial{\vect{n}}=-E_0 \; \forall \vect{x}\in\Gamma_i$ means imposing an electric field [V/m] $E_0$ through a boundary zone $\Gamma_i$, remembering that electric field in [V/m] is $\vect{E}=-\nabla \phi$.
	\end{itemize}

	\item The \textbf{magnetostatics equation}:
	\[
	\nabla^2 \vect{A} = -\mu \vect{J}
	\]
	where $\vect{A}(\vect{x})$ is the unknown vector magnetic potential field in [V s / m] units, 
	$\mu$ is the permeability [H/m] of the material,
	$\vect{J}$ is the volume current density, a vector in [A / m${}^2$] units. 
	\begin{itemize}
		\item a Dirichlet boundary condition $\vect{A}=\vect{A}_0, \; \forall \vect{x}\in\Gamma_i$ means imposing a vector magnetic potential [A] on a boundary surface $\Gamma_i$, 
		%\item a Neumann   boundary condition $\partial{\vect{A}}/\partial{\vect{n}}=-A_0 \; \forall \vect{x}\in\Gamma_i$ means imposing a magnetic field intensity [A/m] $H_0$ through a boundary zone $\Gamma_i$, remembering that magnetic field intensity in [A/m] is $\vect{H}=-\nabla \psi$.
		\item Once the vector magnetic potential is computed, one can get the magnetic field intensity in [A/m] as $\vect{H} = \frac{1}{\mu} \nabla \times \vect{A}$ and magnetic flux density as $\vect{B} = \nabla \times \vect{A}$.
	\end{itemize}

	\begin{figure}
		\centering
			\includegraphics[width=5cm]{pics/es_magnetostatics.jpg}
			\includegraphics[width=5cm]{pics/es_electrostatics.jpg}
		\caption{Examples of finite element solution to electrostatics/magnetostatics problems (from FEMM and COMSOL documentation)}
		\label{fig:es_electrostatics}
	\end{figure}

	%\item The \textbf{magnetostatics equation}:
	%\[
	%\mu \nabla^2 \psi = -\xi
	%\]
	%where $\psi(\vect{x})$ is the unknown scalar magnetic potential field in [A] units, 
	%$\mu$ is the permeability [H/m] of the material,
	%$\xi$ is the volume charge density in $\Omega$, divergence of magnetization $\xi = \nabla \cdot \vect{M}$, if any. 
	%\begin{itemize}
		%\item a Dirichlet boundary condition $\phi=\phi_0, \; \forall \vect{x}\in\Gamma_i$ means imposing a magnetic potential [A] on a boundary surface $\Gamma_i$, 
		%\item a Neumann   boundary condition $\partial{\psi}/\partial{\vect{n}}=-H_0 \; \forall \vect{x}\in\Gamma_i$ means imposing a magnetic field intensity [A/m] $H_0$ through a boundary zone $\Gamma_i$, remembering that magnetic field intensity in [A/m] is $\vect{H}=-\nabla \psi$.
	%\end{itemize}

	\item The \textbf{Newtonian gravitational field equation}:
	\[
	\nabla^2 \phi = 4 \pi \rho G
	\]
	where $\phi(\vect{x})$ is the unknown scalar gravitational potential field, 
	$\rho$ is the density of the material,
	$G$ is the gravitational constant. 
	\begin{itemize}
		\item a Dirichlet boundary condition $\phi=\phi_0, \; \forall \vect{x}\in\Gamma_i$ means imposing a gravitational potential on a boundary surface $\Gamma_i$, 
		\item a Neumann   boundary condition $\partial{\phi}/\partial{\vect{n}}=g_0 \; \forall \vect{x}\in\Gamma_i$ means imposing a gravitational field $g_0$ perpendicular to a boundary zone $\Gamma_i$, remembering that $\vect{g}= -\nabla \phi$.
	\end{itemize}

	\item The \textbf{Reynolds equation on lubrication of thin films}:
	\[
	\nabla \cdot (-\vect{q}) =  \rho W
	\]
	where $\vect{q} \in \mathbb{R}^2$ is a mass flow rate 
	$\vect{q}=\rho \left\{ 
	\frac{Uh}{2}-\frac{h^3}{12\eta}\frac{\partial P}{\partial x},
	\frac{Vh}{2}-\frac{h^3}{12\eta}\frac{\partial P}{\partial y}
	\right\}$ 
	that stems from the original Reynolds equation that is often written in  in the more conventional form 
	\[
	 \frac{\partial}{\partial x} \left( \frac{\rho h^3}{12 \eta}\frac{\partial P}{\partial x}\right) 
  +\frac{\partial}{\partial y} \left( \frac{\rho h^3}{12 \eta}\frac{\partial P}{\partial y}\right) 
	= \frac{\partial}{\partial x}\left( \frac{\rho U h}{2}\right)
	+ \frac{\partial}{\partial y}\left( \frac{\rho V h}{2}\right)
	+ \rho W
	\] 
	for $P$ as pressure, $\rho$ as density, $U,V,W$ as (tangential,tangential,vertical) relative surface velocities, $h$ film thickness, $\eta$ viscosity. Boundary conditions often constrain the pressure on the boundary.

	\item The Laplace equation is a Poison equation without the source term

	\item The thin elastic membrane on rigid support is $K \nabla^2 z = -f$ with $z$ vertical displacement and $f$ is vertical force.
\end{itemize}

\subsection{Diffusion equations}

\textbf{Diffusion equations} are parabolic PDEs of the following type:
\begin{align}
\frac{\partial u}{\partial t}  = \mathbf{\nabla} \cdot (D \mathbf{\nabla} u) + R
\label{eq:diffusion}
\end{align}
where $u$ is the unknown $u(\vect{x},t)$, often scalar-valued but might also be tensor-valued,
$D$ is a diffusivity coefficient, 
$\vect{V}$ is the velocity, 
$R$ is a source (sink) term expressing loads per units of volume.

In the following we list some physical problems that share the same mathematical structure of the diffusion equation, 
all assumed in a 3D domain $\Omega$, with $\vect{x} \in \Omega \subset \mathbb{R}^3$:

\begin{itemize}

	\item The transient \textbf{heat equation}:
	\begin{align}
	\rho c_p \frac{\partial T}{\partial t}   - k\nabla^2 T =  q
	\label{eq:transientheat}
	\end{align}
	where $T(\vect{x})$ is the unknown temperature field in [K] units, 
	$\rho$ is the material density [kg/m${}^3$], 
	$c_p$ is the material specific heat capacity [J/(K kg)], 
	$k$ is the thermal conductivity [W/(m K)], 
	$q$ is the heat-flux [W/m${}^3$] of sources in $\Omega$ if any - for example heat from uranium fission. 
	Note that with no $q$ source, and by introducing the thermal diffusivity $\alpha=k/(\rho c_p)$, in [m${}^2$/s], one has the more conventional heat equation $\frac{\partial T}{\partial t} - \alpha \nabla^2 T=0$, and here one recognize that $\alpha$ is the $D$ is a diffusivity coefficient in \eqref{eq:diffusion}.
	Also note that assuming steady state, $\frac{\partial T}{\partial t} = 0$, hence the Poisson equation for steady state heat $k\nabla^2 T = -q$ is a special case of this.
	Boundary and initial conditions:
	\begin{itemize}
		\item a Dirichlet boundary condition $T=T_i, \; \forall \vect{x}\in\Gamma_i$ means imposing a temperature on a boundary surface $\Gamma_i$, 
		\item a Neumann   boundary condition $-k\partial{T}/\partial{\vect{n}}=q_0 \; \forall \vect{x}\in\Gamma_i$ means imposing a heat flux $q_0$ [W/m${}^2$] through a boundary zone $\Gamma_i$, remembering $\vect{q}=-k \nabla T$.
		\item a initial condition $T=T_0$ at time $t=0$.
	\end{itemize}

	\item the \textbf{second Fick equation}, governing the transport of mass through diffusive means (dilution in liquids, gases, asdorption in porous media, etc.)
	\[
	\frac{\partial \phi}{\partial t}   - D \nabla^2 \phi =  0
	\]
	where $\phi(\vect{x})$ is the unknown concentration of a substance (ex. in [mol/m${}^3$]), $D$ is the diffusion coefficient, in [m${}^2$/s].

\end{itemize}

\subsection{Convection-diffusion equations}

\textbf{Convection-diffusion equations}, sometimes advection-diffusion depending on the context, are PDEs of the following type:
\begin{align}
\frac{\partial u}{\partial t}  = \mathbf{\nabla} \cdot (D \mathbf{\nabla} u) - \mathbf{\nabla} \cdot (\vect{v} u) + R
\end{align}
where $u$ is the unknown $u(\vect{x},t)$, often scalar-valued but might also be tensor-valued,
$D$ is a diffusivity coefficient, 
$\vect{v}$ is the velocity, 
$R$ is a source (sink) term expressing loads per units of volume.

In the following we list some physical problems that share the same mathematical structure of the diffusion equation, 
all assumed in a 3D domain $\Omega$, with $\vect{x} \in \Omega \subset \mathbb{R}^3$:

\begin{itemize}

\item the \textbf{Smoluchowski equation} of drift-diffusion equation, 
\[
\frac{\partial \phi}{\partial t}  = \nabla \cdot (D \nabla \phi) - \nabla \cdot \left( \zeta^{-1} \vect{F} \phi \right) + R
\]
similar to the second Fick equation governing the transport of mass through diffusive means (dilution in liquids, gases, asdorption in porous media, etc.) but also adding a force field $\vect{F}$ as in electrophoresis and assuming a viscous drag so that velocity is $\vect{v}=\zeta^{-1} \vect{F}$. Again, $\phi(\vect{x})$ is the unknown concentration of a substance (ex. in [mol/m${}^3$]), $D$ is the diffusion coefficient, in [m${}^2$/s]. 
This is also similar to the \textbf{Fokker-Plank equation} if one considers the probability density function of velocities of particles in brownian motion instead of concentrations.

\item the incompressible Navier-Stokes equation is closely related to a convection-diffusion problem as:
\[
\frac{\partial \vect{M}}{\partial t} = \frac{\mu}{\rho} \nabla^2 \vect{M} -\vect{v} \cdot \nabla \vect{M} + (\vect{f}-\nabla P)
\]
where $\vect{M} = \rho \vect{v}$ is fluid momentum per unit of volume, $\mu$ is viscosity, $P$ is pressure, $\vect{f}$ is a load per unit of volume, for instance gravity.

\end{itemize}

\subsection{Linear elasticity equations}

\begin{figure}
		\centering
			\includegraphics[width=12cm]{pics/es_elasticity.jpg}
		\caption{Example of finite element solution to a linear elasticity problem (from GETFem++ documentation)}
		\label{fig:es_elasticity}
	\end{figure}

A basic representation of \textbf{linear elasticity} is given by the following PDE, embedding an equilibrium equation, a strain-displacement equation, and a constitutive equation: 
\begin{align}
	\left\{
	\begin{aligned}
		 & \nabla \cdot \tensor{\sigma} + \tensor{b} = \vect{0} \\
		 & \tensor{\epsilon} = \frac{1}{2} \left( \nabla \vect{u} + \nabla \vect{u}^T \right) \\
		 & \tensor{\sigma} = \tensor{C} : \tensor{\epsilon}
	\end{aligned}
	\right.
\end{align}
where 
$\vect{u}$ is the unknown vector displacement field,
$\vect{b}$ is an applied per-unit-of-volume load,
$\tensor{\epsilon}$ is the strain rank-2 tensor,
$\tensor{\sigma}$ is the stress rank-2 tensor,
$\tensor{C}$ is the stiffness rank-4 tensor. %, of whose $3^4=81$ components only two are independent in a isotropic material, ex. Poisson ratio and Young modulus.

For isotropic elasticity, where $\tensor{C}$ can be made dependant on just two parameters and 
$\tensor{\sigma}=2\mu \tensor{\epsilon} +\lambda \; \mathrm{tr}(\tensor{\epsilon})\tensor{I}$,
one can also write the more compact expression as \textbf{Navier-Lamé equation}:
\[
\mu \nabla^2 \vect{u} + (\lambda + \mu) \nabla ( \nabla \cdot \vect{u}) + \tensor{b} = \vect{0}
\]
with the Lamé constants 
$\lambda = \frac{E\nu}{(1+\nu)(1-2\nu)}$ and 
$\mu = G = \frac{E}{2(1+\nu)}$.
This is again a PDE.

\begin{itemize}
		\item a Dirichlet boundary condition $\vect{u}=\vect{u}_0, \; \forall \vect{x}\in\Gamma_i$ means imposing a known displacement (hence a known position) on a boundary surface $\Gamma_i$, 
		\item a Neumann boundary condition $\tensor{\sigma} \cdot \vect{n} = \vect{t}_0 \; \forall \vect{x}\in\Gamma_i$ means imposing a known traction $\vect{t}_0$ on the boundary surface $\Gamma_i$.
	\end{itemize}

\subsection{Navier-Stokes equations}

The \textbf{compressible Navier-Stokes equation} describes the motion of fluids under very general assumptions:
\begin{align}
		 \rho \frac{\partial \vect{v}}{\partial t} + \rho \,\vect{v} \cdot \nabla \vect{v} - \mu \, \nabla^2 \vect{v} + \nabla \bar{p}  - \tfrac13 \mu \, \nabla (\nabla\cdot\vect{v}) - \vect{f} = \vect{0}
\label{eq:navierstokes}
\end{align}
It stems from the following assumptions:
\begin{itemize}
	\item the Cauchy momentum equation $\rho \frac{D\vect{v}}{Dt} = \nabla \cdot \tensor{\sigma} + \vect{f}$, with material derivative $D\vect{v}/Dt = \partial \vect{v}/\partial t + \rho \vect{v} \cdot \nabla \vect{v}$,
	\item the rate of strain is $\tensor{\epsilon}=\frac{1}{2}(\nabla\vect{v}+\nabla{\vect{v}}^T)$,
	\item the linear stress constitutive equation is a isotropic function of rate of strain as 
	$\tensor{\sigma} = \zeta (\nabla\cdot\vect{v}) \mathbf I + \mu \left(\nabla\vect{v} + ( \nabla\vect{v} )^{T} - \tfrac23 (\nabla\cdot\vect{v}) \tensor{I}\right)$ 
	\item the continuity equation for balance of mass: $\frac{\partial \rho}{\partial t} = \nabla \cdot (\rho\vect{v})$
\end{itemize}
where $\vect{v}$ is velocity, $\mu$ is dynamic viscosity, $\lambda$ is bulk viscosity, $\zeta$ is the "dilational" second viscosity as in $\zeta \equiv \lambda + \tfrac23 \mu$, pressure is $\bar{p}= p - \zeta \lambda \cdot \vect{v}$, external loads per unit of volume are $\vect{f}$, in many cases $\vect{f}=\rho\vect{g}$ where $\vect{g}$ is gravity acceleration field.

\begin{itemize}

\item The \textbf{incompressible Navier-Stokes equation} is a special case of the Navier-Stokes equations. Except for particular problems at high Mach numbers, in fact, the compressibility of the fluid can be neglected. Again with the assumption of isotropic Newtonian fluids, and recovering the continuity condition as $\nabla\cdot\vect{v}=0$
\begin{align}
	\left\{
	\begin{aligned}
		  \rho\frac{\partial\vect{v}}{dt} + \rho \,\vect{v} \cdot \nabla \vect{v} - \mu \,\nabla^2 \vect{v}  +\boldsymbol{\nabla}p - \vect{f} &= \vect{0} \\
			\nabla\cdot\vect{v}&= 0
	\end{aligned}
	\right.
	\label{eq:stokes}
\end{align}

\item The \textbf{steady state Navier-Stokes equation} is a special case of the Navier-Stokes equations with no dependence on time. 
With the simplification of Newtonian fluid and incompressible flow one has:
\begin{align}
	\left\{
	\begin{aligned}
		  \rho \,\vect{v} \cdot \nabla \vect{v} - \mu \,\nabla^2 \vect{v}  +\boldsymbol{\nabla}p - \vect{f} &= \vect{0} \\
			\nabla\cdot\vect{v}&= 0
	\end{aligned}
	\right.
	\label{eq:stokes}
\end{align}

\item The \textbf{Stokes equations} is a special case of the steady state Navier-Stokes equations where the advective terms are negligible respect to the viscous forces, as in slow creeping fluids with low Reynolds numbers $Re \rightarrow 0$. With simplification of Newtonian fluid and incompressible flow one has:
\begin{align}
	\left\{
	\begin{aligned}
		  -\mu \,\nabla^2 \vect{v} +\boldsymbol{\nabla}p - \vect{f} &= \vect{0} \\
			\nabla\cdot\vect{v}&= 0
	\end{aligned}
	\right.
	\label{eq:stokes}
\end{align}

\item The \textbf{inviscid Euler equations} is a special case of the Navier-Stokes equations where the diffusive terms are negligible, as in high Reynolds numbers $Re \rightarrow \infty$. With simplification of Newtonian fluid and incompressible flow one has:
\begin{align}
	\left\{
	\begin{aligned}
		  \rho\frac{\partial\vect{v}}{dt} + \rho \vect{v} \cdot \nabla \vect{v} + \boldsymbol{\nabla}p - \vect{f} &= \vect{0} \\
			\nabla\cdot\vect{v}&= 0
	\end{aligned}
	\right.
	\label{eq:stokes}
\end{align}

\end{itemize}

\begin{figure}
		\centering
			\includegraphics[height=4cm]{pics/es_cfd.jpg}
			\includegraphics[height=4cm]{pics/es_cfd2.jpg}
		\caption{Examples of computational fluid dynamics, left: with finite elements (Autodesk CFD simulator), right: with finite volume method (from ANSYS Fluent documentation)}
		\label{fig:es_cfd}
	\end{figure}

\newpage
\section{Discretization}

There are different ways to cast the previous infinite dimensional problems into finite dimensional problems that can be attacked with computational tools. Among the different strategies we mention the finite element method, the finite different method, the finite volume method, etc. In the following we will focus on the finite element Galerkin method.

\subsection{Function spaces}

The following is a primer on basic concepts of functional analysis, this is useful because literature on FEA often cite concepts like Sobolev spaces etc.

\begin{itemize}
	\item A \textbf{vector space}, over a scalar field $K \in \mathbb{R}$, is a set $V$ equipped with 
	addition $+: V \times V \rightarrow V$, commutative and associative, 
	and scalar multiplication $\cdot: F \times V \rightarrow V$, distributive and associative.

	\item An \textbf{inner product space} is a vector space equipped with an inner product, $\left\langle\cdot,\cdot\right\rangle : V \times V \rightarrow F$.

	\item A \textbf{normed vector space} is a vector space equipped with a norm, $\| \cdot \| : V \rightarrow \mathbb{R}$.

	\item A $L^p$-norm is a norm defined as $\| \vect{x} \|_p = \left( |x_1|^p + |x_2|^p + \dotsb + |x_n|^p \right) ^{1/p}$, the Euclidean $L^2$ norm being the most common.

	\item A space $V$ is \textbf{complete} if all Cauchy sequences converge to an element in $V$.

	\item A \textbf{Banach} space is a complete normed vector space.

	\item A \textbf{Hilbert}\footnote{It is remarkable how many bright scientists attended the University of Göttingen.} space is a complete normed inner product space.

	\item A \textbf{function space} is a topological space whose points are functions. Some examples will follow.

	\begin{itemize}
		\item $B$ is the space of \textit{bounded} functions

		\item $C$ is the space of \textit{continuous} functions

		\item $C_c$ is the space of \textit{continuous} functions with compact support

		\item $C^r$ is the space of functions that are \textit{continuous} up to the first $r$ \textit{derivatives}

		\item $C^\infty$ is the space of \textit{smooth functions}, ex. $f=\sin(x)$.

		\item $L^p$ is the \textbf{Lebesgue space} of measurable functions with finite norm $\| f \|_{L^p}=\left( \int_\Omega |f|^p d\mu \right)^{1/p}$, these functions are not necessarily continuous.
		\[
		L^p(\Omega) = \left\{ f \; \left| \; \int_\Omega |f|^p d\mu < \infty \right. \right\}
		\]

		\item Lebesgue $L^p$ space is a Banach space with norm 
		$\| f \|_{L^p}=\left( \int_\Omega |f|^p d\mu \right)^{1/p}$.

		\item Lebesgue $L^2$ space is a Hilbert space with inner product 
		$\left\langle u , v \right\rangle_{L^2} = \int_\Omega u v d\mu$.

		\item Let $f \in L^2(\Omega)$, the \textit{weak derivative} $\partial^\alpha f$ of order $\alpha$ is a function in $L^2(\Omega)$ that satisfies:
		\[
		\int_\Omega \partial^\alpha f \psi d\mu =  (-1)^{|\alpha|} \int_\Omega f \partial^{|\alpha|} \psi d\mu, \quad \forall \psi 
		\in C^{\infty}_c
		\]
		It does not require the function to be differentiable, it is enough that it is integrable.

		\item  $W^{m,p}$ is the \textbf{Sobolev space} is the subspace of $L^p(\Omega)$ functions which possess up to order-$m$ weak derivatives $\partial^\alpha$, with $\alpha \leq m$.
		\[
		W^{m,p}(\Omega) = \left\{ f \in L^p(\Omega) \; | \; \partial^\alpha f \in L^p(\Omega), 1\leq  \alpha \leq m \right\}
		\]

		\item  $H^m$ is the \textbf{Hilbert-Sobolev space} more in detail is a $W^{m,2}$ space:
		\[
		H^m(\Omega) = \left\{ f \in L^2(\Omega) \; | \; \partial^\alpha f \in L^2(\Omega), 1\leq  \alpha \leq m \right\}
		\]

		\item The norm in $H^m$ is defined as
		\[
		\| f \|_{H^m} = \left( \sum_{\alpha=0}^m \int_\Omega |\partial^\alpha f|^2 d\mu \right)^{1/2} \; = \; 
							    \left( \sum_{\alpha=0}^m \int_\Omega ||\partial^\alpha f||^2_{L^2} d\mu \right)^{1/2}
		\]

		\item The importance of Sobolev space is that solutions of PDEs are often searched in $H^m$ space (so called \textit{weak}, or integral solutions) rather than in the $C^k$ space (as in \textit{strong} solutions for order-k PDEs), thus relaxing the continuity requirement.

		\item We denote with $H^m_0$ the subspace of $H^m$ with $m-1$ derivatives equal to zero at the boundary.

		\item In detail $H^1_0$ is the subspace of $H^1$ functions that are zero at the boundary.

		\item In detail $H^1_g$ is the subspace of $H^1$ functions that have a prescribed value $g$ at the boundary.

		\item Functions in $H^m$ are required to be in $C^{m-1}$.

		\item Example: the step function is in $L^2$ but not in $H^1$ (as derivative contains Dirac deltas), the hat function is in $L^2$ and in $H^1$, meanwhile it is in $C^0$.

	\end{itemize}

\end{itemize}

\subsection{The weak formulation}

The weak formulation reduces the regularity assumptions for the solution of the PDE: whereas the solution of the original PDE must be in $C^2$, the weak solution must be only in $H^1_0$, a requirement that makes the problem much more practical. This is the positive side effect of using an integral (variational) form of the equations.

We introduce the weak formulation starting from the case of a generic Poisson equation (a PDE like the steady state heat equation or the electrostatic problem already discussed previously) with unknown scalar function $u$, paired with Dirichlet and Neumann boundary conditions:

\begin{align}
\boxed{
	\left\{
	\begin{aligned}
			& k\nabla^2 u + f = 0	& \forall \vect{x}\in\Omega \\ 
			& u=u_D, \; 					& \forall \vect{x}\in\Gamma_D \\
			& k\partial{u}/\partial{\vect{n}}=q_N \; & \forall \vect{x}\in\Gamma_N
	\end{aligned}
	\right.
	\label{eq:strongpoisson}
	}
\end{align}

A solution to the strong form above should satisfy $u \in C^2(\Omega)$, that is, at least it should be twice differentiable over the entire domain $\Omega$.

We will rewrite it as a weak formulation, that requires less stringent requirement on the functional class of the solution -letting it to be approximated by a piecewise polynomial $u^h \approx u$ at the end. In fact we will assume a $u \in H^1_{u_D}(\Omega)$, that is we assume the solution to be a function with just the first derivative being square integrable, and with prescribed value $u_D$ at the boundary in order to satisfy the Dirichlet boundary condition.

We also introduce a \textit{test function}, namely $v \in H^1_0(\Omega)$, assumed null at the $\Gamma_D$ boundary. 
For compactness we introduce functional spaces $\mathcal{U}$ and $\mathcal{U}_0$ as:
\begin{align}
	u \in \mathcal{U}   &=& \left\{  u \in H^1(\Omega) \; : \; u=u_D \forall \vect{x} \in \Gamma_D \right\} \\
	v \in \mathcal{U}_0 &=& \left\{  v \in H^1(\Omega) \; : \; v=0   \forall \vect{x} \in \Gamma_D \right\}
\end{align}

We start multiplying the terms of the PDE by the test function $v$ and we integrate over the domain, obtaining:
\begin{align}
	\int_\Omega v \, k\nabla^2 u  \, d\Omega + \int_\Omega  v \, f \, d\Omega = 0
	\label{eq:weakpoisson01}
\end{align}

The first term can be integrated by parts (or just applying the Leibnitz product rule, that leads here to the known property of divergence operator: $\nabla\cdot\left(v\mathbf{A}\right)= v\nabla\cdot\mathbf{A}+\mathbf{A}\cdot\nabla v$ where our $\nabla^2 u$ is in fact $\nabla\cdot\mathbf{A}$ for $\mathbf{A}=\nabla u$) thus obtaining:
\begin{align}
	\int_\Omega k \nabla \cdot (v \nabla u)  \,d\Omega - \int_\Omega k \nabla v \cdot \nabla u \,d\Omega    + \int_\Omega  v \, f  \,d\Omega = 0
	\label{eq:weakpoisson02}
\end{align}

Then, we use the Gauss divergence theorem to turn the first term into a boundary integral:
\[
\int_\Omega k \nabla \cdot (v \nabla u)  \,d\Omega = \int_{\Gamma} k v \nabla u \cdot \vect{n} \,d\Gamma
\]
where on the Dirichlet boundaries $\Gamma_D$ it simplifies to zero because $v=0$ on $\Gamma_D$ by construction, whereas on the 
boundaries $\Gamma_N$ it assumes the value prescribed by Neumann condition as $ k \partial{u}/\partial{\vect{n}} = k \nabla u \cdot \vect{n} = q_N$:
\begin{align}
\int_{\Gamma_D} k v \nabla u \cdot \vect{n} \,d\Gamma_D &= 0 \\
\int_{\Gamma_N} k v \nabla u \cdot \vect{n} \,d\Gamma_N &= \int_{\Gamma_N} v \, q_N \,d\Gamma_N
\end{align}

Substitute the above in \eqref{eq:weakpoisson02} and see that the weak formulation simplifies to:
\begin{align}
  \int_\Omega k \nabla v \cdot \nabla u \,d\Omega = \int_\Omega  v \, f  \,d\Omega  + \int_{\Gamma_N} v \, q_N \,d\Gamma_N
	\label{eq:weakpoisson}
\end{align}

We note in passing that going to the weak formulation of the Poisson equation, the original $\nabla^2$ term disappeared, 
and now it does not require anything more than the first derivative of $u$ to be integrable 
(and same for the test function $v$).

Also note that if no Neumann conditions are required it, the weak formulation simplifies even more: 
\[
\int_\Omega k \nabla v \cdot \nabla u \,d\Omega = \int_\Omega  v \, f \,d\Omega 
\]

The weak formulation \eqref{eq:weakpoisson} can be rewritten with a more compact notation, introducing a
bilinear operator 
$\mathcal{B}: \mathcal{U} \times \mathcal{U}_0 \rightarrow \mathbb{R}$ 
and a linear operator 
$\mathcal{L}:  \mathcal{U}_0 \rightarrow \mathbb{R}$ 
that, in our Poisson case above, are:
\begin{align}
  \mathcal{B}(u,v) &= \int_\Omega k \nabla v \cdot \nabla u \,d\Omega      \label{eq:bilinearop}\\
	\mathcal{L}(v)   &= \int_\Omega v \, f \,d\Omega    + \int_{\Gamma_N} v \, q_N \,d\Gamma_N         \label{eq:linearop}
\end{align}

This said, the weak form can be written as
\begin{align}
\boxed{
\mathcal{B}(u,v) = \mathcal{L}(v)
}
\label{eq:weakbl}
\end{align}
In the following we see that most of the PDE problems seen in the previous sections can be expressed with a sum of bilinear and linear terms like the above (but the expressions of the bilinear and linear operators can be different in the sense that the integrands could have different expressions).

We remark in passing that, for symmetrical $\mathcal{B}(u,v)$, \eqref{eq:weakbl} is equivalent to a variational problem $u = \mathrm{argmin}_{u}I[u]$ with the functional $I[u]=\frac{1}{2}\mathcal{B}(u,u) - \mathcal{L}(u)$ representing some form of energy, and whose stationary condition is $\mathcal{B}(u,\partial u) - \mathcal{L}(\partial u) = 0$; note how in elasticity the $\partial u$ can be considered a virtual displacement, and this turns into the principle of virtual work.

\subsection{Discretization}

Now, one can introduce an approximation of $u$ and $v$, to pass from an infinite-dimensional problem to a finite-dimensional problem that depends on a countable set of variables, hence more computer-friendly. 
We assume a discretized approximate solution $u^h \approx u$, with $u^h \in \mathcal{U}^h \subset \mathcal{U}$. 
Here $h$ represents the discretization size, and $u^h \rightarrow u$ as $h\downarrow 0$.
Similarly, $v^h \approx v$, with $v^h \in \mathcal{U}_0^h \subset \mathcal{U}_0$.

The finite-dimensional function spaces $\mathcal{U}^h, \mathcal{U}_0^h$ are built using $n$ \textit{shape functions}, here named $N_i(\vect{x})$, as in:
\begin{align}
u^h(\vect{x}) &= \sum_i^n \hat{u}_i N_i(\vect{x}) \\
v^h(\vect{x}) &= \sum_j^n \hat{v}_j N_j(\vect{x})
\end{align}

Thank to this discretization, the unknown is not a function $u$ anymore, but it is rather a vector with a finite amount of $\hat{u}_i$ unknowns. 
We also remark that one can split the entire domain $\Omega$ in multiple finite elements, each with its $\hat{u}_i$ unknowns, leading to a finer 
discretization and to a larger amount of $\hat{u}_i$ unknowns.

Shape functions can be linear, quadratic, etc., depending on the finite element type.

This variational setting is attributed to W.Ritz\footnote{It is remarkable how many bright scientists attended the University of Göttingen.}
and B.Gal{\"e}rkin\footnote{The pronounciation of {\"e} in russian sounds like \textit{yo}, so one should pronounce \textit{Galyorkin}, I do not why everyone spells it Gal\textit{e}rkin.}, hence the name \textbf{Galerkin} method. More in detail, if the same approximation space is used for both $u$ and $v$, this is sometimes referred as \textbf{Bubnov-Galerkin} method, and if a different approximation space is chosen for $v$, this is called \textbf{Petrov-Galerkin} method. 
The modern mathematical setting and terminology for the finite element method is credited to R.Courant\footnote{It is remarkable how many bright scientist attended the University of Göttingen.}.

Using the discretization, the weak form can be written:
\begin{align}
\mathcal{B}(u^h,v^h) &= \mathcal{L}(v^h) \\
\mathcal{B}\left(\sum_i^n u_i N_i(\vect{x}),\sum_j^n v_j N_j(\vect{x})\right) &= \mathcal{L}\left(\sum_j^n v_j N_j(\vect{x})\right)
\end{align}

Exploiting the fact that $\mathcal{B}$ is bilinear and $\mathcal{L}$ is linear, one can take the $\hat{u}_i$ and $\hat{v}_i$ weights outside integrals, leading to:
\begin{align}
\sum_j^n \hat{v}_j \sum_i^n \hat{u}_i \mathcal{B}\left(N_i(\vect{x}),N_j(\vect{x})\right) &= \sum_j^n \hat{v}_j \mathcal{L}\left( N_j(\vect{x})\right)
\end{align}

One can simplify the $\hat{v}_j$ terms and obtain a very simple form of the discretized weak form where one can see that the bilinear and linear operators act only on the shape functions:
\begin{align}
\sum_i^n \hat{u}_i \mathcal{B}\left(N_i(\vect{x}),N_j(\vect{x})\right) &= \mathcal{L}\left( N_j(\vect{x})\right)
\end{align}

In sake of further compactness one can group all $\hat{u}_i$ in the vector of unknowns $\hat{\vect{u}}$, introduce a matrix $\matr{K} \in \mathbb{R}^{n\times n}$ and a vector $\vect{b} \in \mathbb{R}^{n}$, and obtain a linear system: 
\begin{align}
\boxed{
\begin{aligned}
\matr{K} \hat{\vect{u}}  &= \vect{b} & \\[2mm]
 K_{ij}   &=  \mathcal{B}\left(N_i,N_j \right) \\
 b_j &=  \mathcal{L}\left( N_j )\right)
\end{aligned}
}
\end{align}
This $\matr{K} \vect{u} = \vect{b}$ linear system gives the solution $\vect{u}$ to the discretized PDE problem. Of course, the finer the discretization, the larger the size of $\hat{\vect{u}}$, hence the larger the matrix $\matr{K}$, and the more computationally intensive is the solution. To this end, one might consider using special types of solvers, such as Krylov solvers, that exploit the fact that in many cases the $\matr{K}$ matrix is sparse and contains many structural zeros: please go to the section on linear solvers for a primer on this topic.

\subsection{Quadrature}

One can see that, for computing the elements of $\matr{K}$, in our example of the Poisson equation, one needs to compute 
\[
K_{ij}  = \mathcal{B}\left(N_i,N_j \right) = \int_\Omega k \nabla N_i(\vect{x}) \cdot \nabla N_j(\vect{x}) d\Omega
\]
just like for the elements of $\vect{b}$ one computes
\[
b_j  = \mathcal{L}\left( N_j )\right) = \int_\Omega N_i(\vect{x}) \, f \,d\Omega + \int_{\Gamma_N} N_i(\vect{x}) \, q_N \,d\Gamma_N
\]
This amounts to performing some type of integration over a domain (the finite element volume, for instance) per each term. This can be
done via analytical methods only in limited cases, otherwise using Newton-Cotes quadrature, or Gauss-Legendre quadrature, or other approximations. 
We discuss the Gauss quadrature briefly, ie. an approximation of the integral via a weighted sum of values sampled in some Gauss points $\vect{x}_g$ scattered in the volume domain.

In general, Gauss integration over $n{GP}$ Gauss points with corresponding $w_g$ weights at $\vect{x}_g$ abscyssae points is done with a sum, for example:
\begin{align*}
\int_\Omega k \nabla N_i(\vect{x}) \cdot \nabla N_j(\vect{x}) d\Omega &= \sum_{g=1..n_{GP}} w_g \nabla N_i(\vect{x}_g) \cdot \nabla N_j(\vect{x}_g)
\end{align*}
and the same for other bilinear/linear terms.

The lower the degree of shape functions, the smaller the amount of Gauss point samples that are needed to correctly compute the quadrature; for example, a tetrahedral finite element with its four linear shape functions requires only a single Gauss point in the middle, etc.

Note, however, that most often the $w_g$ and values are tabulated assuming generic abscyssae $\zeta_g$ ranging in the $[-1,+1]$ interval, hence one has to express $\vect{x}=\vect{x}(\vect{\zeta})$ (as happens in the \textit{isoparametric} finite element approach, where $\vect{\zeta}$ for example contains the three parametric coordinates along the direction of a brick each ranging in $[-1,+1]$ and shape functions are functions of these parametric coordinates) and simply scale the result using jacobians $\partial\vect{x}/\partial\vect{\zeta}$ as in
\[
\int_a^b f(x) dx = \int_{-1}^{+1} J_{\zeta} f(\zeta) d\zeta \approx J_{\zeta} \sum_{g=1..n_{GP}} w_g f(\zeta_g)
\]
that in our case would lead to:
\begin{align*}
\int_\Omega k \nabla N_i(\vect{x}) \cdot \nabla N_j(\vect{x}) d\Omega \approx J_{\vect{\zeta}} \sum_{g=1..n_{GP}} w_g \nabla N_i(\vect{\zeta}_g) \cdot \nabla N_j(\vect{\zeta}_g)
\end{align*}

The reader interested in quadrature should look also at the following connected problems: \textit{hourglass modes} (zero-energy spurious modes), selective and \textit{reduced integration}, \textit{shear locking} (ex.in Timoshenko beams) and how to get rid of it using ad-hoc integration.

\subsection{Constraints}

Apart from the already mentioned treatment of Dirichlet and Neumann conditions, a straightforward and powerful way to apply generic constraints to nodes of a FE mesh -especially in a context of mechanical problems where nodes represent points of a deformable material- is to couple the discretized PDE with algebraic constraints. A (possibly linear) constraint equation is a function that gives zero residual when satisfied:
\[
\vect{C}(\hat{\vect{u}})=\vect{0}
\]
For example, in the context of 3D continuum elasticity, if one requires that the displacement $\hat{\vect{u}}_k$ of the $k$-th node is equal to some $\overline{\vect{u}}$ value, one would write:
\[
\vect{C}(\hat{\vect{u}}) = \hat{\vect{u}}_k - \overline{\vect{u}} 
\]
Another example, again with a node in a 3D continuum: if one requires the node to be able to move on the yz plane of a coordinate system with rotation matrix $\matr{R}$, while precluding the x displacement relative to such coordinate system, one would write:
\[
\vect{C}(\hat{\vect{u}}) = [1, 0, 0] \matr{R}^T \hat{\vect{u}}_k  
\]

One can add multiple constraints in this way, obtaining a large $\vect{C}(\hat{\vect{u}})$ vector of $m$ scalar constraints, $\vect{C}:\mathbb{R}^n \rightarrow \mathbb{R}^m$. 
The constraint jacobian is a matrix $\matr{C}_u \in \mathbb{R}^{m \times n}$:
\[
\matr{C}_u = \frac{\partial \vect{C}(\hat{\vect{u}})}{\partial \hat{\vect{u}}}
\]
Not all $m$ constraints act on all the $n$ variables of the system-wide $\hat{\vect{u}}$ vector, so the jacobian is often a very sparse matrix.

By linearizing the constraint equation, one has
\[
\vect{C}(\hat{\vect{u}}) \approx \matr{C}_u (\hat{\vect{u}}-\hat{\vect{u}}_0) + \vect{C}(\hat{\vect{u}}_0)
\]
where here we pose $\hat{\vect{u}}_0 = \vect{0}$, thus the constraint equations leads to a linear algebra equation 
$\matr{C}_u \hat{\vect{u}} + \vect{C}(\hat{\vect{u}}_0) = \vect{0}$.

Roughly speaking, the transpose of the jacobian matrix multiplied by a unknown multiplier $\vect{\lambda}$ gives a reaction force in global coordinates, where the unknown multiplier is the reaction in the coordinates of the constraint, therefore the final linear system to be solved becomes:

\begin{align}
\boxed{
\begin{bmatrix}
	\matr{K} & \matr{C}_u^T \\
	\matr{C}_u  & \matr{0}
\end{bmatrix}
\begin{Bmatrix}
	\hat{\vect{u}} \\
	\vect{\lambda}
\end{Bmatrix}
=
\begin{Bmatrix}
	\vect{b} \\
	-\vect{C}(\vect{u}_0)
\end{Bmatrix}
}
\end{align}

The structure of the linear system above, featuring a square matrix $\matr{K}$ bordered to the right and below by constraint jacobians $\matr{C}_u$ and additional unknown multipliers $\vect{\lambda}$ (the constraint reaction forces in this case) like in a Karush Kuhn Tucker (KKT) matrix, is very frequent in other problems involving constraints - not only in structural mechanics. Solving the KKT matrices efficiently is a key to a robust and fast computational tool.

The question is: we added constraints using the KKT matrix by enforcing constraints at level of an already discretized problem, but is it possible to recover the same KKT machinery also if the constraints are defined in the previous step, i.e. when the formulation is still infinite-dimensional? The answer is yes, but the variational procedure is left to the user.

\subsection{Weak formulations for other classes of problems}

So far, we have seen how to translate a Poisson problem (stationary heat problem, magnetostatics, Reynolds lubrication etc.) into a corresponding weak form, and from the weak form we obtained a discretized problem to be solved via a linear system. One can see that other weak formulations can be derived for other classes of problems. For sake of compactness here we present just a few.

\begin{itemize}
	\item Weak form of the Navier-Lamé (linar elasticity) equation:
	\begin{align}
	\int_\Omega \tensor{\sigma}(\vect{u}) : \tensor{\epsilon}(\vect{v}) \,d\Omega = 
	 \int_\Omega \vect{f} \cdot \vect{v} \,d\Omega + 
	 \int_\Gamma \tensor{\sigma}(\vect{u}) \vect{n} \cdot \vect{v} \,d\Gamma   
	\end{align}
	This can be again expressed as $\mathcal{B}(\vect{u}^h,\vect{v}^h) = \mathcal{L}(\vect{v}^h)$, and therefore also to a linear system 
	\[
	\matr{K}\hat{\vect{u}}=\vect{b}
	\]
	once discretized (but look how the expression of $\mathcal{B}$ changes respect to the Poisson problem).

	\item Weak form of the steady-state incompressible Navier-Stokes equation:
	\begin{align}
	\left\{
	\begin{aligned}
   \int_{\Omega} \mu \nabla \boldsymbol u \cdot \nabla \boldsymbol v  + \int_{\Omega}\rho (\boldsymbol u \cdot \nabla) \boldsymbol u \cdot \boldsymbol v  
	- \int_{\Omega} p \nabla \cdot \boldsymbol v &= \int_{\Omega}\boldsymbol f \cdot \boldsymbol v +  \int_{\Gamma_N} \boldsymbol h \cdot \boldsymbol v\\
	\int_{\Omega} q \nabla \cdot \boldsymbol u &= 0 
	\end{aligned}
	\right.
	\end{align}
	assuming a Dirichlet condition $\tensor{\sigma}(\boldsymbol u, p) \boldsymbol{n} = \boldsymbol h$ in $\Gamma_N$.
	Note that this required an additional test function $q$, for the incompressibility constraint, and note that the discretization will lead to a KKT matrix, 
	%
\begin{align*}
\begin{bmatrix}
	\matr{K} & \matr{C}_u^T \\
	\matr{C}_u  & \matr{0}
\end{bmatrix}
\begin{Bmatrix}
	\hat{\vect{u}} \\
	\hat{\vect{p}}
\end{Bmatrix}
=
\begin{Bmatrix}
	\vect{b} \\
	\vect{0}
\end{Bmatrix}
\end{align*}
%
	where the lower-left and upper-right submatrices $\matr{C}_u$ are indeed jacobians of the incompressibility constraint. The $K$ upper left submatrix will depend on the viscosity and density. The unknowns of the discretized KKT system are indeed the vector of discrete velocities $\hat{\vect{u}}$, and the vector of discrete pressures $\hat{\vect{p}}$ as lagrangian multipliers of the constraints.\footnote{Finite elements for velocity and pressure must be to be well matched, otherwise the CFD discretized problem has no proper solution, for instance the shape function polynomials of pressure are often one degree lower respect to the degree of polynomials of velocity.}

	Note that if adding also the time dependent term, $\displaystyle \int_{\Omega}\rho \dfrac{\partial \boldsymbol u}{\partial t}\cdot \boldsymbol v$, one has the weak form of the full incompressible Navier-Stokes equation.

\end{itemize}

\newpage
\section{Numerical methods for linear systems}

We assume a square matrix $\matr{A}  \in \mathbb{R}^{n \times n}$ and a known vector $\vect{b} \in \mathbb{R}^n$. 
The problem we want to solve is
\[
\matr{A} \vect{x} = \vect{b}
\]
with $\vect{x}$ unknowns.

This is a problem often encountered in the numerical methods for computational mechanics.

Potential difficulties stem from the following facts:

\begin{itemize}
	\item the $\matr{A}$ matrix might have a large size,
	\item the $\matr{A}$ matrix might be rank-deficient or badly conditioned,
	\item in some context the $\matr{A}$ is never explicitly assembled, because it is known in a factored form or as a sparse data structure, and in an optimal scenario the solver should be able to exploit this data storage type without building the full matrix.
\end{itemize}

There are different classes of solvers that can be used, each with benefits or drawbacks. Among the most relevant classes:
\begin{itemize}
	\item Direct methods
	\item Stationary iterations (fixed point iterations)
	\item Krylov methods
	\item Multigrid methods
\end{itemize}
In the following we briefly list the main solver types belonging to these classes.

\subsection{Direct methods}

Direct methods provide the exact solution to the linear problem, at least up to the floating-point precision of the machine, in a limited number of steps.
Their drawback is that, except some special cases (ex. matrices with banded structure) the number of steps grows superlinearly with the number of unknowns, up to $O(n^3)$ complexity in the worst case for full matrices.

The \textbf{Gauss elimination} is one of the oldest direct schemes. It reduces the matrix in a form where the lower left triangle is filled with zeros, so a quick backward elimination procedure can compute $\vect{x}$. 
%This is the pseudocode for computing the matrix with lower left zeros:
%{
%\footnotesize
%\begin{lstlisting}
 %h := 1 
 %k := 1 
 %while h <= m and k <= n
     %/* Find the k-th pivot: */
     %i_max := [[argmax]] (i = h ... m, abs(A[i, k]))
     %if A[i_max, k] = 0
         %k := k+1
     %else
          %swap rows(h, i_max)
          %for i = h + 1 ... m:
                 %f := A[i, k] / A[h, k]
                 %A[i, k] := 0
                 %for j = k + 1 ... n:
                      %A[i, j] := A[i, j] - A[h, j] * f
          %h := h + 1
          %k := k + 1
%\end{lstlisting}
%}
Note that the algorithm requires a pivoting strategy in case it encounters a zero pivot (a zero on the diagonal) because it would cause a division by zero. Pivoting can still fail if the matrix is singular, in detail it fails $p$ steps before reaching the last line if $p$ is the number of zero singular values.

The \textbf{LU decomposition} decomposes the matrix into $\matr{A}=\matr{L} \matr{U}$ with upper right triangular matrix $\matr{U}$ ad lower left triangular matrix $\matr{L}$. This done, the system is easily solved via backward substitutions. This method is quite similar to the Gauss substitution in terms of storage, complexity, CPU time.

The \textbf{Cholesky decomposition} decomposes the matrix into $\matr{A}=\matr{L} \matr{L}^T$ with lower left triangular matrix $\matr{L}$, this can be seen as a special type of LU decomposition for symmetric positive definite $\matr{A}$ matrices. Exploiting symmetry, it is a twice as faster as LU.

The \textbf{LDLt decomposition} decomposes the matrix into $\matr{A}=\matr{L} \matr{D} \matr{L}^T$ with lower left triangular matrix $\matr{L}$ and diagonal $\matr{D}$, and it is similar to the Cholesky decomposition. It avoids square roots of Cholesky algorithm, and the matrix can be also negative definite (leading to negatives values in D diagonal, in fact). Pivoting can be used to make the algorithm more stable if pivots are too small, but to avoid breaking symmetry it cannot use generic pivoting, and it must use a diagonal pivoting.

\subsubsection{Parallelization and HPC}

In general, it is difficult to implement direct methods on\textit{ parallel computing architectures} and on High Performance Computing (HPC) at large. There are some noticeable examples of direct parallel solvers, though, and among these we mention PARDISO from the MKL Intel libarary, and MUMPS. Both are able to exploit multicore CPUs to a good extent.

\subsection{Stationary iterations}

This is the simplest class of iterative methods. Stationary methods, also known as fixed point methods, reach the solution by iterative refinements.

Benefits: 
their implementation is very simple, 
they do not need to store the full matrix (they are sparse-matrix friendly),
in some flavors they are robust in case of singular matrices,
their implementation is most often based on a single computational primitive: (sparse) matrix by vector. 
Also, they can be easily modified to become \textit{projected} fixed point iterations to solve also nonlinear complementarity problems (ex obstacle problems).

Drawbacks: 
their convergence is very slow (in some case they can stall and converge to desired tolerance in a too large number of iterations, hence eliding the benefit of the simple implementation). Especially in the case of PDEs arising from FEA, and more markedly FEA with structural elements such as shells, the A matrix might be badly conditioned (bad spectral radius) and this has a huge negative effect on the speed of convergence, up to rendering those methods unusable.

Here is a list of the most famous fixed point iterations for solving $\matr{A} \vect{x} = \vect{b}$

\subsubsection{Jacobi method}

The \textbf{Jacobi method} assumes a splitting $\matr{A}=\matr{D}+ \matr{R}$, and obtain the solution by iterating 
\[
\vect{x}^{(k+1)} = D^{-1} (\vect{b} - R \vect{x}^{(k)})
\]
or, in element-wise notation, by iterating 
\[
x^{(k+1)}_i  = \frac{1}{a_{ii}} \left(b_i -\sum_{j\ne i}a_{ij}x^{(k)}_j\right),\quad i=1,2,\ldots,n.
\]
up to $k \leq k_{max iters}$ or up to reaching tolerance.
A sufficient -but not necessary- condition for the method to converge is that $\matr{A}$ is strictly diagonally dominant. 
One can see that the Jacobi method is intuitively like solving each linear equation separately for one variable, assuming the other variable fixed, and repeating the process in $k$ loop.

\subsubsection{Gauss-Jacobi method}

The \textbf{Gauss-Jacobi method} assumes a splitting $\matr{A}=\matr{L}+ \matr{U}$, where $L$ contains the diagonal and $U$ doesn't, and obtain the solution by iterating 
\[
\vect{x}^{(k+1)} = L^{-1} (\vect{b} - U \vect{x}^{(k)})
\]
There is no need to invert $L$ explicitly, neither to build $L$ or $U$ explicitly, in fact one can exploit the triangular nature of $L$ so that $L^{-1}\ldots$ becomes, componentwise, a sequential backward substitution:
\[
x^{(k+1)}_i  = \frac{1}{a_{ii}} \left(b_i - \sum_{j=1}^{i-1}a_{ij}x^{(k+1)}_j - \sum_{j=i+1}^{n}a_{ij}x^{(k)}_j \right),\quad i=1,2,\dots,n.
\]
From an intuitive point of view, this method is similar to the Jacobi method, i.e solves each linear equation separately for one variable, sequentially in each iteration $k$, but instead of keeping the other variable fixed as they were known at the last $k-1$ iteration, here it reuses instantly the variables as soon as updated.

\subsubsection{SOR successive over-relaxation method}

The \textbf{SOR successive over-relaxation method} assumes a splitting $\matr{A}=\matr{L}+ \matr{D} + \matr{U}$ with $\matr{D}$ diagonal, then performs 
\[
\mathbf{x}_{(k+1)} = (D+\omega L)^{-1} \big(\omega \mathbf{b} - [\omega U + (\omega-1) D ] \mathbf{x}_{(k)}\big)=L_w \mathbf{x}_{(k)}+\mathbf{c},
\]
Again, this can be written sequentially to avoid matrix inverse, and we get the sequential form:
\[
x_{(k+1)}{}_i  = (1-\omega)x_{(k)}{}_i + \frac{\omega}{a_{ii}} \left(b_i - \sum_{j<i} a_{ij}x_{(k+1)}{}_j - \sum_{j>i} a_{ij}x_{(k)}{}_j \right),\quad i=1,2,\ldots,n. 
\]
It is similar to the Gauss-Jacobi method, with the exception that it introduces the relaxation parameter $\omega$. For $\omega=1$ it is exactly Gauss-Jacobi. For higher $\omega$ it can converge faster - but this depends on the spectral radius of the matrix, too high $\omega$ will lead to divergence.

\subsubsection{Parallelization and HPC}

Note that Gauss-Jacobi parallelize very easily (just assign lines of the matrix to different threads and perform the updates: these are independent and they just need a synchronization at each $k$ iteration).

On the other side, SOR and Gauss-Jacobi updates imply cross-dependency of variables, hence cannot be parallelized as in Jacobi otherwise one can run into race conditions.

To avoid this issue, one must develop \textit{red-black partitioning} methods.

\subsection{Krylov methods}

Krylov methods, also called Krylov subspace methods, are based on the construction of a basis of successive matrix powers times the initial residual. A Krylov subspace is:
\[
\mathcal{K}_r(A,\vect{b}) = \operatorname{span} \, \{ \vect{b}, A\vect{b}, A^2\vect{b}, \ldots, A^{r-1}\vect{b} \}. 
\]
Building the subspace draws on \textit{matrix by vector} multiplications, and we will see that this computational primitive is the basic building block of all Krylov methods.

Krylov methods find the $\mathcal{K}_r(A,\vect{b})$ space, and in theory Krylov methods should converge to a maximum number of iterations $n$ with $n$ being the amount of unknowns. In many practical situations, however, they might converge to acceptable precision much faster than this worst case scenario. On the other hand, finite precision of floating point algebra means that the upper limit $n$ could be even broken, and this could happen in unlucky cases with badly conditioned matrices, for example.

For very large scale problems, Krylov methods are competitive respect to direct methods, and the fact that they can work seamlessly with sparse matrices make them a good choice for many PDE solvers.

Parallelization of Krylov methods can be performed at various levels (ex on GPU via CUDA or OpenCL, on multicore using OpenMP, or even on multiple nodes using MPI for extremely large problems) and in general it boils down to parallelizing the matrix x vector operation. Other computational primitives that might be parallelized are the inner products and the vector sums and scaling.

In the following we comment some of the most used Krylov methods, leaving the discussion on others (QMR, TFQMR, QMRCGSTAB, CGS, Barzilai-Borwein, etc) do specialized textbooks.

\subsubsection{Steepest descend}

This is a very simple method that is not used in practice, but useful for didactical purposes It requires the matrix $\matr{A}$ to be 
\begin{itemize}
\item positive definite,
\item symmetric.
\end{itemize}
The method requires iterating through these steps:
\begin{align}
	\left| \quad
	\begin{aligned}
			& \text{repeat} \\
		  & \qquad \vect{r}_{(k)} = \vect{b} - \matr{A}\vect{x}_{(k)} \\
			& \qquad \hbox{if } ||\mathbf{r}_{(k)}|| < \epsilon \quad \text{exit loop} \\
			& \qquad \alpha_{(k)} = \frac{\vect{r}_{(k)}^T\vect{r}_{(k)}}{\vect{r}_{(k)}^T \matr{A} \vect{r}_{(k)}} \\
			& \qquad \vect{x}_{(k+1)} = \vect{x}_{(k)} + \alpha_{(k)} \vect{r}_{(k)}
	\end{aligned}
	\right.
	\label{eq:stokes}
\end{align}
Note that the computational bottleneck is the computational primitive matrix x vector in $\matr{A} \vect{r}^{(k)}$.

The matrix can be sparse and there is no need of fill-in elements as the sparsity pattern is not changed (this is a positive feature of all Krylov methods).

The issue of this method is that it converges slowly, as it can be intuitively interpreted by remembering that it corresponds to the gradient descend algorithm for finding the minimum of a function $f(\vect{x})= 1/2 \vect{x}^T\matr{A}\vect{x} - \vect{b}^T \vect{x} + c$, that can get stuck in a zigzag pattern if the "valley" is skewed and squeezed in a diagonal direction.

\subsubsection{Conjugate gradient}

This method greatly improves the performance of the steepest descend gradient method. It requires the matrix $\matr{A}$ to be 
\begin{itemize}
\item positive definite,
\item symmetric.
\end{itemize}

The pseudocode is:
\begin{align}
\left| \quad
	\begin{aligned}
& \mathbf{r}_0 := \mathbf{b} - \matr{A} \mathbf{x}_0 \\
& \mathbf{p}_0 := \mathbf{r}_0 \\
%& k := 0 \\
& \text{repeat} \\
& \qquad \alpha_{(k)} := \frac{\mathbf{r}_{(k)}^\mathsf{T} \mathbf{r}_{(k)}}{\mathbf{p}_{(k)}^\mathsf{T} \mathbf{A p}_{(k)}}  \\
& \qquad \mathbf{x}_{(k+1)} := \mathbf{x}_{(k)} + \alpha_{(k)} \mathbf{p}_{(k)} \\
& \qquad \mathbf{r}_{(k+1)} := \mathbf{r}_{(k)} - \alpha_{(k)} \matr{A} \mathbf{p}_{(k)} \\
& \qquad \hbox{if } ||\mathbf{r}_{(k+1)}|| < \epsilon \quad \text{exit loop} \\
& \qquad \beta_{(k)} := \frac{\mathbf{r}_{(k+1)}^\mathsf{T} \mathbf{r}_{(k+1)}}{\mathbf{r}_{(k)}^\mathsf{T} \mathbf{r}_{(k)}} \\
& \qquad \mathbf{p}_{(k+1)} := \mathbf{r}_{(k+1)} + \beta_{(k)} \mathbf{p}_{(k)} \\
%& \qquad k := k + 1 \\
\end{aligned}
	\right.
\end{align}

Also this method requires only one CPU-intensive operation per each iteration, namely the matrix x vector computational primitive.

Note that also the convergence of the conjugate gradient method can be negatively affected by ill-conditioned matrices. In such circumstances, one should revert to using some type of \textbf{preconditioner}, that is like an approximate solver of other specie that is interleaved to the Krylov iteration. There is no space here to discuss preconditioners more in detail and how to rewrite the CG iteration in order to accommodate a preconditioner, suffices to say that there are many options (none being optimal in all scenarios) and we name a few: diagonal scaling, ILU (Incomplete LU factorization), block ILU, multigrid, stationary iterations, spectral preconditioners, etc.

\subsubsection{MINRES}

The Minimum Residual (MINRES) method is similar to the Conjugate Gradient solver but it can work also with singular matrices, in fact it corresponds to finding a minimum norm residual solution to the system of linear equations.
It requires the matrix $\matr{A}$ to be 
\begin{itemize}
\item symmetric.
\end{itemize}

Just like all other iterative methods, MINRES is usually paired with a preconditioning method in order to improve convergence.

The MINRES method is very simple to implement, it does not require substantial additions to the CG method, and shows about the same convergence speed in many practical scenarios.

\subsubsection{GMRES}

The Generalized Minimum Residual (GMRES) method is a more complex Krylov solver that fixes the limitation of the Conjugate Gradient solver, in fact it can work also with non-symmetric matrices. This comes at a cost: the code is more complex (not shown here for conciseness) and it requires the storage of more auxiliary vectors.
It requires the matrix $\matr{A}$ to be 
\begin{itemize}
\item positive definite.
\end{itemize}

Just like all other iterative methods, GMRES is usually paired with a preconditioning method in order to improve convergence.

The GMRES method is sometimes restarted after a user-defined number of iterations. Also, anti-stagnation procedures must be put in place in this practical implementation.

\subsubsection{BiCGSTAB}

The Biconjugate gradients stabilized method (BiCGSTAB)  is a refined version of the biconjugate gradient method (BiCG) that exhibits a better convergence. The implementation is simplier than GMRES, but it may exhibit easier stagnation and non monotone irregular convergence.
It requires the matrix $\matr{A}$ to be 
\begin{itemize}
\item positive definite.
\end{itemize}

Note that the BiCGSTAB method requires two matrix x vector operations per each iteration. The BiCG method requires one multiplication by an inverse of $A$, that makes it not practical especially when using sparse matrices.

Although the convergence is irregular and non monotone, it is one of the best among Krylov methods.

\subsection{Multigrid methods}

(To do...)

\begin{itemize}
	\item iterate in 'V' or 'W' cycles until tolerance reached,
  \item solve the problem on different scales, from fine to coarse, then back,
	\item the solver for each level is a very simple stationary method, often one or two steps of Gauss-Seidell might suffice,
	\item complex implementation,
	\item very good performance also on badly conditioned matrices,
  \item can be used as stand-alone or can be used as preconditioners to other methods (ex Krylov methods).
\end{itemize}

We distinguish between \textit{geometric multigrid}	- requiring structured meshes (rectangular grids)
and \textit{algebraic multigrid} - requiring a complex implementation, especially of the smoothing operator.

\begin{figure}
		\centering
			\includegraphics[height=5cm]{pics/es_multigrid.jpg}
		\caption{The V cycle in a multigrid method.}
		\label{fig:es_multigrid}
	\end{figure}

\newpage
\section{Time integration}

Among the PDEs, there are some problems that depend on time. Studying the time evolution of such systems, starting from given 
initial conditions, is a numerical problem that can be solved via time integration.

There are many different time integration 
methods (aka time steppers) which fit better some cases or others.

Here we cannot present all of them, but we briefly 
discuss relevant aspects: the interested reader can find additional details in literature.

\subsection{Classes of time-dependent problems}

In this section we introduce terminology and classifications for time-dependent problems.
Among these, the ODEs and DAEs represent by far the most relevant, and we urge the reader to 
understand their meaning. Remaining classes presented here, namely DIs, DVIs, MDIs are used
only in the more rare cases of problems subject to non-smooth set-valued laws, such as in
case of multibody systems with hard frictional contacts or simulation of analog electronic 
circuits with diods etc.

\subsubsection{ODEs}

Given a generic state $\vect{x}$ and time $t$, an \textit{Ordinary Differential Equation (ODE)} is a system of equations as:
%
\begin{align}
  \frac{d\vect{\vect{x}}}{dt}  = \vect{f} \left( \vect{x}, t\right)
\end{align}
%
together with prescribed initial conditions $\vect{x}(t_0)=\vect{x}_0$ 
\footnote{The expression above is the \textit{explicit form} of the ODE. One could express ODEs also in \textit{implicit form} as 
\[
\vect{F} \left( \frac{d\vect{\vect{x}}}{dt}, \vect{x}, t\right) = \vect{0}
\].
}.

An example: the transient heat equation \eqref{eq:transientheat} is an ODE with unknown $x$ as the temperature $T$, 
in fact it can be written as 
\[
\frac{\partial T}{\partial t} = \frac{1}{\rho c_p}\left(k\nabla^2 T + q\right)
\]
Solving such ODE means that one can obtain the time evolution of the temperature $T=T(\vect{x},t)$ for each point $\vect{x} \in \Omega$. Initial temperature
$\vect{T}(t_0)=\vect{T}_0$ must be prescribed as initial conditions.

The expression above is an ODE of \textit{index one}. Generic \textit{index-n} ODEs are in the form
%
\begin{align}
  \frac{d^n\vect{\vect{x}}}{dt^n}  = \vect{f} 
	\left( 
	\frac{d^{(n-1)}\vect{\vect{x}}}{dt^{(n-1)}},
	\ldots
	\frac{d^2\vect{\vect{x}}}{dt^2},
	\frac{d\vect{\vect{x}}}{dt},
	\vect{x}, t\right)
\end{align}

Its interesting to note that all index-n ODE can be converted into an index-one ODE by introducing a larger state with all the derivatives up to $n-1$ as in  $\vect{s}=\left\{\frac{d^{(n-1)}\vect{\vect{x}}}{dt^{(n-1)}}, \ldots, \frac{d^2\vect{\vect{x}}}{dt^2}, \frac{d\vect{\vect{x}}}{dt}, \vect{x} \right\}^T$ and rewriting it as:
%
\begin{align}
\frac{d\vect{\vect{s}}}{dt}  = 
\left\{
\begin{array}[pos]{c}
\frac{d^n\vect{\vect{x}}}{dt^n} \\
\ldots \\
\frac{d^2\vect{\vect{x}}}{dt^2} \\
\frac{d\vect{\vect{x}}}{dt}
\end{array}
\right\}
=
\left\{
\begin{array}[pos]{c}
\vect{f} 
	\left( 
	\frac{d^{(n-1)}\vect{\vect{x}}}{dt^{(n-1)}},
	\ldots
	\frac{d^2\vect{\vect{x}}}{dt^2},
	\frac{d\vect{\vect{x}}}{dt},
	\vect{x}, t\right) \\
\ldots \\
\frac{d^2\vect{\vect{x}}}{dt^2} \\
\frac{d\vect{\vect{x}}}{dt}
\end{array}
\right\};
\quad
\frac{d\vect{\vect{s}}}{dt}  = \vect{f}_s \left( \vect{s}, t\right)
\end{align}

An example is the index-2 ODE of unconstrained multibody dynamics: 
%
\begin{align}
  \matr{M} \frac{d\vect{\vect{v}}}{dt}  = \vect{f} \left( \vect{v},\vect{x}, t \right)  
	\label{eq:odemechanics}
\end{align}
%
that can be converted into a first-order ODE $\frac{d\vect{\vect{s}}}{dt}  = \vect{f}_s \left( \vect{s}, t \right)$ with $\vect{s} = \left\{ \vect{v}, \vect{x} \right\}$ and $\frac{d\vect{\vect{s}}}{dt}  = \left\{ \vect{a}, \vect{v} \right\} = \left\{ \matr{M}^{-1}\vect{f}(\vect{v},\vect{x}, t), \vect{v} \right\} = \vect{f}_s \left( \vect{s}, t \right)$.

This means that integration methods designed for first-order ODEs can be applied also to second-order ODEs without much work. However, this said, it is better to develop custom integrators for second-order ODEs when possible. %; this is the case of implicit integrators in \Chrono, for example.

The \textit{Picard-Lindelh\"{o}f theorem} states that for $\vect{f}\left( \vect{x}, t\right)$ uniformly Lipschitz continuous in $\vect{x}$ and continuous in $t$, there is an unique solution $\vect{x}(t)$ to the ODE for a given initial value. See also Cauchy–Lipschitz theorem.

For infinite-dimensional problems like the transient heat equation and similars, in most cases the ODE is expressed 
at the discretized level (look the section on weak formulation and discretization) 
so that one has a finite dimensional problem with a countable set of components in the unknown vector.

\subsubsection{DAEs}

Given a generic state $\vect{x}$, and time $t$, a \textit{Differential Algebraic Equation (DAE)} is a system of equations as:
%
\begin{align}
  \frac{d\vect{\vect{x}}}{dt}  &= \vect{f} \left( \vect{x}, \vect{y}, t\right) \\
	\vect{g} \left( \vect{x}, \vect{y}, t\right) &= \vect{0}
\end{align}
%
where $\vect{g}(.)$ is a set of constraint equations, also called \textit{geometric} or \textit{algebraic constraints}, and where $\vect{y}$ are the \textit{algebraic variables}.

The expression above is the \textit{semi-explicit form} of the DAE. One could express DAEs also in \textit{implicit form} as 
%
\begin{align}
\vect{F} \left( \frac{d\vect{\vect{x}}}{dt}, \vect{x}, \vect{y}, t\right) = \vect{0}
\end{align}
%
where the jacobian matrix $\frac{\partial \vect{F}}{\partial \vect{y}}$ is singular; in fact if it was singular, one could cast it as a simple ODE $\frac{d\vect{\vect{x}}}{dt}  = \vect{f} \left(\vect{x}, t\right)$.

The DAE \textit{initial conditions} cannot be provided as arbitrarily as in ODEs, because they must be \textit{consistent}, i.e. they must satisfy $\vect{F} \left( \frac{d\vect{\vect{x}}(t_0)}{dt}, \vect{x}(t_0), \vect{y}(t_0), t_0 \right) = \vect{0}$.

The \textit{differentiation index} of a DAE, or simply the \textit{index} of a DAE is the number of the differentiations that are required on the constraint equations of the semi-explicit form in order to obtain a pure ODE, after substitutions. An ODE is a DAE of index 0.

After index-reduction to an ODE, one might be tempted to perform the time integration of such ODE with classical integration schemes of ODEs. This might work for some steps but later some error tends to build up in the algebraic equations, giving a \textit{drift-off} phenomena in constraints. This because constraints would be satisfied only at the n-th derivative level (ex. acceleration, and not position) in the reduced ODE. To solve such drifting problem there are different methods. A rough approach is to introduce Baumgarte stabilization terms, like spring dampers that attracts the solution back on the manifold of the original, non differentiated constraints. Another approach is to satisfy constraints in implicit integrations methods specially crafted for DAEs, using Newton-Raphson solvers, and this is the case of HHT, generalized-$\alpha$ and other DAE-oriented implicit solvers.

An example: the incompressible Navier-Stokes equations are a DAE\footnote{Some authors call it a PDAE
to underline the (P)artial derivative differential algebraic equation nature, but it is only a matter of
terminology.} of index 2, where the constraints $\vect{g}(\vect{x},t) = \vect{0}$ are given
by the incompressibility constraint $\nabla \cdot \vect{u}=0$.

An example: the index-2 DAE of constrained multibody dynamics
%
\begin{align}
  \matr{M} \frac{d\vect{\vect{v}}}{dt} - \matr{C}_q^T \vect{\lambda}  &= \vect{f} \left( \vect{q},\vect{v}, t \right)   \\
	\vect{C}(\vect{q},\vect{v},t) &=\vect{0} \\
	\dot{\vect{q}} &= \Gamma(\vect{q}) \vect{v}
	\label{eq:daemechanics}
\end{align}
%
with constraint reactions $\vect{\lambda}$, configuration $\vect{q}$, velocities $\vect{v}$, external and gyroscopic forces $\vect{f}$. 
Usually $\vect{v}=\dot{\vect{q}}$, but the complication of rotations in 3D might lead to the heterogeneous choice of using, say, quaternions in $\vect{q}$ and
angular velocities in $\vect{v}$, and this would require the linear operator $\Gamma$ for $\dot{\vect{q}} = \Gamma(\vect{q}) \vect{v}$.

\subsubsection{DIs}

A Filippov \textbf{Differential Inclusion} (DI) can be interpreted as an ODE 
for problems with discontinuous $\vect{f}(\vect{x},t)$:
\begin{equation}
\label{eq:FilippovDI}
\frac{d\vect{x}}{dt} \in \mathcal{F}\vect{f}(\vect{x},t) 
\quad
 \mathcal{F}f(\vect{x},t) = \bigcap_{\eta > 0} \bigcap_{N:\lambda_0(N)=0}
  \bar{\mathrm{co}} \vect{f}(\vect{x+\eta B_1}\setminus N,t)
\end{equation}
where $\lambda_0(E)$ is a Lebesgue measure on $E$, $B_1$ is a origin-centered 
unit ball, and $\vect{f}$ is discontinuous in $\vect{x}$.

More in general, a \textit{Differential Inclusion} (DI) is a problem
\begin{equation}
\label{eq:DI}
\frac{d\vect{x}}{dt} \in \mathcal{F}(\vect{x},t) 
\end{equation}
where the set-valued function $\mathcal{F}(\vect{x},t)$ is
closed, bounded and convex and is upper semi-continuous, or
equivalently, $\mathcal{F}(\vect{x},t)$ has closed graph.

\subsubsection{DVIs}

We introduce \textbf{Differential Variational Inequality} (DVI) problems as:
\begin{subequations}
\label{eq:DVI}
\begin{align}
\frac{d\vect{x}}{dt} &= \vect{f}(\vect{x},\vect{u},t)  \\
\vect{u} &\in \mathsf{SOL}(\vect{F},\mathcal{K})
\end{align}
\end{subequations}
where $\mathsf{SOL}(\vect{F},\mathcal{K})$ is the (set of) solution 
to the VI$(\vect{F},\mathcal{K})$.
This is a special type of DI, too.
One can see that a DVI includes DAE as a special case: with $n$ bilateral
algebraic constraints one takes $\vect{F}$ as the vector of algebraic constraint residuals,
and uses $\mathcal{K}=\mathbb{R}^n$ so that $\vect{F}=\vect{0}$ everywhere by definition VI.

\subsubsection{MDIs}

A \textbf{Measure Differential Inclusion} (MDI) is a
generalization of DI (DVI) that also accomodates impulsive events. 
For second order problems as in mechanics, with 
$\vect{v}(t)=d\vect{q}/dt$
it reads:
\begin{equation}
\label{eq:MDI}
\frac{d\vect{v}}{dt} \in \mathcal{K}(\vect{q},t) 
\end{equation}
where $\vect{v}$ is a function of bounded variation and
$\mathcal{K}(\vect{q},t)$ is a set-valued function with
closed graph and closed convex values. 
The \textit{strong} definition of solution (Moreau) follows the singular
measure decomposition of $d\vect{v}=\vect{\nu}$ into 
$\vect{\nu}_s+\vect{h}\lambda_0$,
with respect to the singular part $\vect{\nu}_s$ 
and Lebesgue measure
$\lambda_0$ for continuous $\vect{h}(t) \in L^1(a,b)$: then the strong definition of solution is: 
$\vect{h}(t) \in \mathcal{K}(t)$ almost all $t$, and 
$d\vect{\nu}_s/|\vect{\nu}_s|(t) \in \mathcal{K}(t)_\infty$, 
i.e. the Radon-Nikodym derivative fits the horizon cone.
The \textit{weak} definition of solution (Stewart) is:
\begin{equation}
\label{eq:MDI_weak}
\frac{\int \phi(t) d\vect{\nu}(dt)}{ \int \phi(t) dt} 
 \in \bar{\mathrm{co}} \bigcup_{\tau:\phi(\tau)\neq 0} \mathcal{K}(\tau) 
\end{equation}
for any continuous $\phi(t):\mathbb{R}\rightarrow\mathbb{R}^+$ with compact support.

\subsection{Time integration numerical methods}

The goal is to provide methods that, given a state $\vect{x}_n$ at time $t_{n}$, can proceed to the next state at time $\vect{x}_{n+1}$ at time $t_{n+1}$ by advancing with a time step $h$, with $h=t_{n}-t_{n+1}$.

Focusing on ODEs only, for the moment, we can introduce some taxonomies of methods.

\begin{itemize}

\item
We can distinguish between \textbf{multistep} and \textbf{Runge\footnote{It is remarkable how many bright scientist attended the University of Göttingen.}-Kutta} methods.
The former depends on a history of previous timesteps in order to extrapolate a new one at the end of the new step, whereas the latter starts from the last timestep information then computes temporary intermediate $\vect{f}_j \left( \vect{x}_j, t\right)$ values after it, in order to compute the value at the end of the step.

\item
We can distinguish between \textbf{explicit} and \textbf{implicit} methods, the former using formulas that do not require any information on the value of state at the end of the step, whereas the latter uses (nonlinear) formulas that depend on the value of state at the end of the step, thus leading to more intricate solution schemes that rely on fixed point iterations or Newton iterations or in general iterative schemes that reach the solution at the end of the time step by iterating multiple times over a nonlinear set of equations. 
A notable feature of implicit methods is that they can solve problems with large stiffness (steep integrals with sudden rapid oscillations, so to speak) without the need of short time steps, whereas explicit methods require much shorter time steps to avoid divergence - see L-stability etc.

\item
We can distinguish between \textbf{fixed} time step or \textbf{variable} time step methods depend on the fact that, in the former case the time step $h$ is assumed constant, whereas in the second case the method can embed some smart heuristics to change the time step $h$ depending on its needs, for example making it shorter when approaching steepy stiff integrals, and making it larger when the time integration can proceed faster with lower CPU overhead. The latter case often involve the ability of un-doing and \textit{repeating the time step} with shorter $h$ when convergence or precision is not satisfactory. 
Think at this property as a simple add-on feature: almost all fixed step methods can be made of variable time step type by adding the rewinding capability to the original algorithms and adding some heuristics for shrinking/enlarging the time step.

\item
We mention that someone distinguishes between \textbf{stiff} and \textbf{non-stiff} methods, meaning the capability of handling problems with high stiffness without incurring in divergence. As this property is specific of implicit methods, often one can consider the class of stiff methods to be synonym of implicit methods. In a broader interpretation, one can use the term stiff also for explicit methods that, once embedded with some variable time step scheme, can attack (moderately) stiff problems by reducing automatically the time step up to avoid divergence.

\item
We also mention the not so frequent definition of \textbf{predictor-corrector} schemes. Rather than being a pure mathematical quality, this is an elaborate trick to help the convergence of sub-iterations of some implicit methods by warmstarting them with the help of another (often cheaper) integrator. In fact, in many predictor-corrector schemes, an explicit method is used to roughly estimate the new state working a bit like an extrapolator, and then an implicit integrator uses this estimate as a starting point for the refinement of the solution by running few Neton Raphson (or fixed point) iterations.

\end{itemize}

The qualities above are orthogonal, in the sense that a specific integration method could be, for example the backward Euler method is of Runge-Kutta type, implicit, usually with fixed time step, and stiff.

We present some relevant integration schemes in the following.

\subsubsection{Explicit Runge-Kutta methods}

These methods are based on multiple evaluation of the derivative at intermediate time steps, thus subdividing the $h$ time step into invisible temporary sub-steps.
The count of substeps is called \textit{stages}. The general formula for an \textbf{explicit Runge Kutta} method of $s$ stages is:
\begin{fleqn}
\begin{align}
\left[
\begin{aligned}
\vect{x}_{n+1} &= \vect{x}_n + h \sum_{i=1}^s b_i \vect{k}_i  \\
t_{n+1} &= t_n + h &
\end{aligned}
\right.
\end{align}
\end{fleqn}
with 
\begin{align}
 \vect{k}_1 & = \vect{f}(t_n, \vect{x}_n), \\
 \vect{k}_2 & = \vect{f}(t_n+c_2h, \vect{x}_n+h(a_{21}\vect{k}_1)), \\
 \vect{k}_3 & = \vect{f}(t_n+c_3h, \vect{x}_n+h(a_{31}\vect{k}_1+a_{32}\vect{k}_2)), \\
     & \ \ \vdots \\
 \vect{k}_s & = \vect{f}(t_n+c_sh, \vect{x}_n+h(a_{s1}\vect{k}_1+a_{s2}\vect{k}_2+\cdots+a_{s,s-1}\vect{k}_{s-1})).
\end{align}
where coefficients are given by a \textit{Butcher tableau}, satisfying $\sum_{j=1}^{i-1} a_{ij} = c_i \text{ for } i=2, \ldots, s.$

Among the explicit Runge-kutta methods, the most popular is the 4th order one, with $c_1=0, c_2=0.5, c_3=0.5, c_4=1, b_1=1/6, b_2=1/3, b_3=1/3, b_4=1/6$. 
It has a local truncation error on the order of $O ( h^5 )$.

One can estimate the local error by performing a 5th order RK alongside the 4th order one, but to avoid
computing 4+5=9 functions $f()$, special methods such as the Runge-Kutta Fehlberg 4/5th order have been developed, that for the 5th order reuse 4 evaluations of $f()$ of the 4th order plus just one additional $f()$, that is used only for estimating the local error. A similar idea is used for Runge-Kutta 2/3 methods. These are known as ODE45 and ODE23 in Matlab or other packages.

Knowing an estimate of the local truncation error is useful for two reasons: for diagnostic during postprocessing, and/or to tell the integrator that the $h$ step is too large to give acceptable precision and therefore $h$ must be decreased or even re-done, if the integrator is of \textit{variable} time step type.

\subsubsection{Explicit Euler}

The \textbf{explicit Euler} method is a special type of explicit Runge-Kutta method, the simpliest one by the way. It boils down to:
\begin{fleqn}
\begin{align}
\left[
\begin{aligned}
\vect{x}_{n+1} &= \vect{x}_n + h \vect{f}(t_n, \vect{x}_n) \\
   t_{n+1} &= t_n + h 
\end{aligned}
\right.
\end{align}
\end{fleqn}

In terms of precision, it is a first-order method, thus it has a local truncation error on the order of $O ( h^2 )$ and its precision is a bit disappointing (even a simple parabolic ballistic trajectory will be approximated with growing error accumulation).

Moreover, large time steps in stiff problems will easily bring this method to divergence.

The positive note, however, is that the explicit Euler method is very cheap in the computational sense, because it just requires one evaluation of the $\vect{f}(t_n, \vect{x}_n)$ term at each time step, given the current state $\vect{x}_n$.  Therefore, one can afford making very short time steps to circumvent its limitation on precision and to avoid the risk of divergence.

\subsubsection{Implicit Runge-Kutta methods}

The \textbf{implicit Runge-Kutta} methods are defined as:
\begin{fleqn}
\begin{align}
\left[
\begin{aligned}
\vect{x}_{n+1} &= \vect{x}_n + h \sum_{i=1}^s b_i\vect{k}_i \\
t_{n+1} &= t_n + h 
\end{aligned}
\right.
\end{align}
\end{fleqn}
%
with terms that depend on $\vect{x}_{n+1}$ as in
\begin{align}
\vect{k}_i = \vect{f}\left( t_n + c_i h,\ \vect{x}_{n} + h \sum_{j=1}^s a_{ij} \vect{k}_j \right), \quad i = 1, \ldots, s.
\end{align}

\subsubsection{Implicit Euler}

A very famous sub-case of implicit Runge-Kutta schemes is the \textbf{implicit Euler} (also known as backward Euler) method.
It is expressed as:
\begin{fleqn}
\begin{align}
\left[
\begin{aligned}
\vect{x}_{n + 1} &= \vect{x}_n + h f(t_n + h,\ \vect{x}_{n + 1}) \\
t_{n+1} &= t_n + h 
\end{aligned}
\right.
\end{align}
\end{fleqn}

Note here a typical complication of all implicit methods: a system of (possibly nonlinear) algebraic equations has 
to be solved to compute $\vect{x}_{n + 1}$, and this must happen at each time step.

\subsubsection{Adams-Bashforth methods}

This is an example from the class of multistep methods. As for all multisteps methods, it requires an history of previous time steps.

Recycling the information of previous time steps will help the precision with low computational cost - consider that Runge-Kutta methods achieve accuracy at the cost of computing intermediate time steps that are discarded. However it also leads to some issues, one being the fact that they require some tricks in starting them from scratch, when previous time history is not given.

The Adams-Bashforth methods of order $s$ are explicit linear multistep methods given by the general formula:
\begin{fleqn}
\begin{align}
\left[
\begin{aligned}
\sum_{j=0}^s a_j \vect{x}_{n+j} &= h\sum_{j=0}^s b_j \vect{f}(t_{n+j},\vect{y}_{n+j}) \\
t_{n+1} &= t_n + h 
\end{aligned}
\right.
\end{align}
\end{fleqn}
where one has $a_s=-1$, $a_j=0$ for $j<s$, and $b_j$ coefficients are given by a Lagrange polynomial interpolation hence $b_{s-j-1} = \frac{(-1)^j}{j!(s-j-1)!} \int_0^1 \prod_{i=0 \atop i\ne j}^{s-1} (u+i) \,du, \qquad \text{for } j=0,\ldots,s-1$; for example the 4-th order Adams-Bashforth boils down to 
\begin{fleqn}
\begin{align}
\left[
\begin{aligned}
\vect{x}_{n+4} &= \vect{x}_{n+3} + h\left( \frac{55}{24} \vect{f}(t_{n+3}, y_{n+3}) - \frac{59}{24} \vect{f}(t_{n+2}, y_{n+2}) + \frac{37}{24} \vect{f}(t_{n+1}, y_{n+1}) - \frac{9}{24} \vect{f}(t_n, y_n) \right) \\
t_{n+1} &= t_n + h 
\end{aligned}
\right.
\end{align}
\end{fleqn}
and one can see that it can reuse three previous $\vect{f}$ informations each time it advances one $h$ step.

Adams-Bashforth methods with $s$ steps have local truncation error on the order of $O ( h^{(s+1)} )$.

\subsubsection{Adams-Moulton methods}

This is another class of linear multistep methods, similar to Adams-Bashforth, but this time it is an implicit integration.
For example, the 4-th order Adams-Moulton method boils down to:
\begin{fleqn}
\begin{align}
\left[
\begin{aligned}
\vect{x}_{n+3} &= \vect{x}_{n+2} + h \left( \frac{9}{24} \vect{f}(t_{n+3},\vect{x}_{n+3}) + \frac{19}{24} \vect{f}(t_{n+2},\vect{x}_{n+2}) - \frac{5}{24} \vect{f}(t_{n+1},\vect{x}_{n+1}) + \frac{1}{24} \vect{f}(t_n,\vect{x}_n) \right) \\
t_{n+1} &= t_n + h 
\end{aligned}
\right.
\end{align}
\end{fleqn}
and one can see that the $\vect{f}(t_{n+3},\vect{x}_{n+3})$ terms depends on the state at the end of the time step $\vect{x}_{n+3}$, making it an implicit method (and consequently requiring some complicate machinery like Newton-Raphson iterations to solve for $\vect{x}_{n+3}$).

Adams-Moulton methods with $s$ steps have order $s+1$ and local truncation error of order $O ( h^{(s+2)} )$.

\subsubsection{Trapezoidal method}

The trapezoidal integrator is in implicit integrator, and it can be seen both as a special case of Adams-Moulton integrator with one step, or an implicit Runge-Kutta of 2nd order. After the impicit Euler, it is perhaps the simplest implicit integration scheme.
The scheme is:
\begin{fleqn}
\begin{align}
\left[
\begin{aligned}
\vect{x}_{n+1} &= \vect{x}_n + \tfrac12 h \left( \vect{f}(t_n,\vect{x}_n) + \vect{f}(t_{n+1},\vect{x}_{n+1} \right) \\
t_{n+1} &= t_n + h 
\end{aligned}
\right.
\end{align}
\end{fleqn}
This method has order 2, local truncation error of order $O ( h^{(s+2)} )$. The method unstable even if it is unconditionally stable.

Among the implicit methods, it stands out because it does not introduce numerical damping, but this property is less usable than it seems in practical scenarios with strong nonlinearities that can create divergence (the unconditional stability holds in the theoretical ideal case of linear stiffness) so a bit of numerical damping in implicit integrators would be welcome anyway.

\subsubsection{BDF methods}

These are implicit linear multistep methods, but are a bit simplier and more efficient than the Adams-Moulton methods. The general formula for an order-$s$ BDF scheme is:
\begin{fleqn}
\begin{align}
\left[
\begin{aligned}
\sum_{k=0}^s a_k \vect{x}_{n+k} &= h \beta \vect{f}(t_{n+s}, \vect{x}_{n+s}) \\
t_{n+1} &= t_n + h 
\end{aligned}
\right.
\end{align}
\end{fleqn}
with properly chosen coefficients $\beta$ and $a_k$.

The implicit Euler is also the special case of BDF with order one. We mention here the coefficients for other orders:
\begin{itemize}
	\item second order BDF: 
	\[
	\vect{x}_{n+2} - \tfrac43 \vect{x}_{n+1} + \tfrac13 \vect{x}_n = \tfrac23 h \vect{f}(t_{n+2}, \vect{x}_{n+2})
	\]
	\item third order BDF: 
	\[
	\vect{x}_{n+3} - \tfrac{18}{11} \vect{x}_{n+2} + \tfrac9{11} \vect{x}_{n+1} - \tfrac2{11} \vect{x}_n = \tfrac6{11} h \vect{f}(t_{n+3}, \vect{x}_{n+3})
	\]
	\item fourth order BDF: 
	\[
	\vect{x}_{n+4} - \tfrac{48}{25} \vect{x}_{n+3} + \tfrac{36}{25} \vect{x}_{n+2} - \tfrac{16}{25} \vect{x}_{n+1} + \tfrac{3}{25} \vect{x}_n = \tfrac{12}{25} h \vect{f}(t_{n+4}, \vect{x}_{n+4})
	\]
\end{itemize}

Note that linear multistep methods with an order greater than 2 cannot be A-stable (see stability).

\subsection{Numerical implementation of methods}

In the previous section we presented some of the most relevant integration schemes, but in a general setting that assumes an index-1 ODE, without telling how to perform the Newton-Raphson substeps for the implicit methods, and telling nothing about DAEs.

In a practical setting, however, one should better exploit the structure of the problem at hand. For example, in mechanics, it is true that one can turn an order-2 ODE (with accelerations) into an index-1 ODE grouping positions and speeds in a single state $\vect{x}=(\vect{q},\vect{v})$, however it is better to express the method directly at the level of index-2 ODE with separate equations for speeds and accelerations. This also makes easier to recover the jacobian matrices for the implicit methods that require Newton-Raphson iterations.

Moreover, what can be done in case of DAEs, that is when algebraic constraints are added to ODEs?

In this section we show, for a limited set of integrators, how to implement them at the algorithmical level, also showing some schemes for DAEs.

\subsubsection{Trapezoidal, for index-2 ODE}

For simplicity, let's start with the ODE case, i.e. also a DAE where no constraints are present.

For an index-2 ODE as used in dynamics in \label{eq:odemechanics} (btw the DAE case will be dealt later), we use state 
$\vect{x}=\left\{ \vect{q},\vect{v} \right\}$, $\dot{\vect{x}}=\left\{ \vect{v},\vect{a} \right\}$ so we have:

\begin{empheq}[left=\empheqlbrack]{align}
\vect{q}^{l+1}-\vect{q}^l - \left( \frac{\vect{v}^{l+1}+\vect{v}^l}{2} \right) h = 0
\label{eq:trapzoidal_q}
\\
\vect{v}^{l+1}-\vect{v}^l - \left( \frac{\vect{a}^{l+1}+\vect{a}^l}{2} \right) h = 0
\label{eq:trapzoidal_v} \\
t^{l+1} = t^l + h
\end{empheq}

In general one has computable $\vect{a}^{l}$ and $\vect{a}^{l+1}$, where in sake of compactness we 
write
$\vect{a}^{l}$ for $\vect{a}(\vect{q}^{l},\vect{v}^{l},t^{l})$, as well as 
$\vect{a}^{l+1}$ for $\vect{a}(\vect{q}^{l+1},\vect{v}^{l+1},t^{l+1})$.
Given that in a mechanical context one splits $\vect{f}=\matr{M} \vect{a}$, 
one can also write the following (that fits in ODE too) where for brevity we write 
$\vect{f}^{l}$ for $\vect{f}(\vect{q}^{l},\vect{v}^{l},t^{l})$, as well as 
$\vect{f}^{l+1}$ for $\vect{f}(\vect{q}^{l+1},\vect{v}^{l+1},t^{l+1})$, :

\begin{align}
\vect{q}^{l+1}-\vect{q}^l - \left( \frac{\vect{v}^{l+1}+\vect{v}^l}{2} \right) h = 0
\label{eq:trapzoidal_q2}
\\
\left( \vect{v}^{l+1}-\vect{v}^l \right) \left( \matr{M}^{l+1}+\matr{M}^{l} \right) - h \vect{f}^{l+1} -h \vect{f}^l = 0
\label{eq:trapzoidal_v2}
\end{align}

Let's call $\vect{G}=\left\{ \vect{G}_q, \vect{G}_v \right\}$ the residual of the system above. 
To satisfy $\vect{G}=0$, one can use a Newton-Raphson iteration (the subscribt means the iteration number) and solve 
for the proper $\vect{q}^{l+1},\vect{v}^{l+1}$:

\begin{align}
\vect{G}(\vect{q}^{l+1},\vect{v}^{l+1}) = 0 \\
\vect{G}(\vect{q}^{l+1}_n,\vect{v}^{l+1}_n) + 
\left[ 
\begin{array}[pos]{cc}
	\frac{\partial \vect{G}_q}{\partial \vect{q}} & \frac{\partial \vect{G}_q}{\partial \vect{v}}  \\
	\frac{\partial \vect{G}_v}{\partial \vect{q}} & \frac{\partial \vect{G}_v}{\partial \vect{v}} 
\end{array}
 \right]_n^{l+1} 
\left\{ 
\begin{array}[pos]{c}
	\vect{q}^{l+1}_{n+1} - \vect{q}^{l+1}_{n} \\ 
	\vect{v}^{l+1}_{n+1} - \vect{v}^{l+1}_{n}
\end{array}
 \right\}
 = 0  
\end{align}
%
also using deltas as a more compact notation for corrections:
\begin{align}
\vect{G}(\vect{q}^{l+1}_n,\vect{v}^{l+1}_n) + \left[ 
\begin{array}[pos]{cc}
	\frac{\partial \vect{G}_q}{\partial \vect{q}} & \frac{\partial \vect{G}_q}{\partial \vect{v}}  \\
	\frac{\partial \vect{G}_v}{\partial \vect{q}} & \frac{\partial \vect{G}_v}{\partial \vect{v}} 
\end{array}
 \right]_n^{l+1}
\left\{  
\begin{array}[pos]{c}
	\Delta\vect{q}^{l+1} \\ 
	\Delta\vect{v}^{l+1}
\end{array}
 \right\} 
 = 0  
\label{eq:trapzoidal_qv}
\end{align}
%

or, in a even more compact notation for jacobian $\matr{J}$, it boils down to a Newton-Raphson step of the type 
\begin{align}
\matr{J}
\left\{ 
\begin{array}[pos]{c}
	\Delta\vect{q}^{l+1} \\ 
	\Delta\vect{v}^{l+1}
\end{array}
 \right\} 
 = -\vect{G}(\vect{q}^{l+1}_n,\vect{v}^{l+1}_n)
\end{align}

Now, note that a straight implementation of the above NR step is not efficient since $J$ would be highly sparse and unsymmetric. 
One can unroll the equations, and from Eq.\ref{eq:trapzoidal_q} one gets
$\Delta\vect{q}^{l+1}=\frac{h}{2}\Delta\vect{v}^{l+1}$, that allows 
	\footnote{The fact that $\Delta\vect{q}^{l+1}=\frac{h}{2}\Delta\vect{v}^{l+1}$ is used in various literature, 
	but it is true only close to the solution, when 
	the residual $\vect{G}_q$ is close to zero, 
	otherwise it would be 
	$\Delta\vect{q}^{l+1}=\frac{h}{2}\Delta\vect{v}^{l+1}-\vect{G}_q = \frac{h}{2}\Delta\vect{v}^{l+1} -\vect{q}^{l+1}+\vect{q}^l + \left( \frac{\vect{v}^{l+1}+\vect{v}^l}{2} \right) h$. However this might overly complicate the NR step that follows. } 
to rewrite as:

\begin{align}
\matr{J} = 
\left[ 
\begin{array}[pos]{cc}
	\frac{\partial \vect{G}_q}{\partial \vect{q}} & \frac{\partial \vect{G}_q}{\partial \vect{v}}  \\
	\frac{\partial \vect{G}_v}{\partial \vect{q}} & \frac{\partial \vect{G}_v}{\partial \vect{v}} 
\end{array}
 \right]_n^{l+1}
 = 
\left[ 
\begin{array}[pos]{cc}
	\matr{I} & -\frac{h}{2} \matr{I}  \\
	-h \nabla_q \vect{f}^{l+1}  & (M^{l+1}+ M^l) -h \nabla_v \vect{f}^{l+1}
\end{array}
 \right] 
\end{align}

Now one can put that jacobian in Eq.\ref{eq:trapzoidal_qv}, develop the equations and arrange them so that one can write the step Eq.\ref{eq:trapzoidal_qv} as three substeps to be repeated in sequence for each Newton iteration, until $\Delta\vect{q}^{l+1}$ and $\Delta\vect{v}^{l+1}$ go to zero. 
In such steps we define $\left( \frac{\matr{M}^{l+1}+\matr{M}^l}{2} \right) = \hat{\matr{M}}$, that is often the constant mass $\matr{M}$ in many cases. 
One gets the Newton step procedure:

\begin{empheq}[box=\fbox]{align}
\left[ \hat{\matr{M}} - \frac{h^2}{4} \nabla_q \vect{f}^{l+1} - \frac{h}{2} \nabla_v \vect{f}^{l+1} \right] \Delta\vect{v}^{l+1} &=
\left( \vect{v}^{l}-\vect{v}^{l+1} \right) \hat{\matr{M}} + \frac{h}{2} \vect{f}^{l} + \frac{h}{2} \vect{f}^{l+1} 
\label{eq:trapzoidal_qv1}
\\
\vect{v}^{l+1}_{n+1} &= \vect{v}^{l+1}_{n} + \Delta\vect{v}^{l+1}
\label{eq:trapzoidal_qv2}
\\
\vect{q}^{l+1} &= \vect{q}^{l} + \left( \frac{\vect{v}^{l+1}_{n+1} + \vect{v}^{l}}{2} \right) h
\label{eq:trapzoidal_qv3}
\end{empheq}
%
where only the 1st step Eq.\ref{eq:trapzoidal_qv1} is the computationally intensive step. But the matrix is hermitian, at least.

\begin{itemize}

	\item Note that the residual in Eq.\ref{eq:trapzoidal_qv1} is the same residual of Eq.\ref{eq:trapzoidal_v2}, except everything is divided by a factor 2 for convenience.

	\item Note that if the exact expression $\Delta\vect{q}^{l+1}=\frac{h}{2}\Delta\vect{v}^{l+1}-\vect{G}_q$ is used, the term $-\vect{G}_q$ must modify the residual in Eq.\ref{eq:trapzoidal_qv1} by adding $... + \frac{h}{2} \nabla_q \vect{f}^{l+1} \vect{G}_q$ that is $... + \frac{h}{2} \nabla_q \vect{f}^{l+1} \left( \vect{q}^{l+1}-\vect{q}^l - \left( \frac{\vect{v}^{l+1}+\vect{v}^l}{2} \right) h \right)$.

	\item Note that $\nabla_q \vect{f}$ is $-\matr{K}$, the stiffness matrix in structural elments, and $\nabla_v \vect{f}$ is $-\matr{R}$, the damping matrix.

\end{itemize}

\subsubsection{Euler implicit, for index-2 ODE}

The Euler implicit method (backward Euler), for a 2nd order ODE, is:

\begin{empheq}[left=\empheqlbrack]{align}
\vect{q}^{l+1}-\vect{q}^l - \vect{v}^{l+1} h = 0
\label{eq:euler_q}
\\
\vect{v}^{l+1}-\vect{v}^l - \vect{a}^{l+1} h = 0
\label{eq:euler_v} 
\\
t^{l+1}= t^l + h
\end{empheq}

Again, as with the trapezoidal, this can be solved with a Newton-Raphson iteration.
Let's call $\vect{G}=\left\{ \vect{G}_q, \vect{G}_v \right\}$ the residual of the system above. 
%
\begin{align}
\matr{J}
\left\{ 
\begin{array}[pos]{c}
	\Delta\vect{q}^{l+1} \\ 
	\Delta\vect{v}^{l+1}
\end{array}
 \right\} 
 = -\vect{G}(\vect{q}^{l+1}_n,\vect{v}^{l+1}_n)
\end{align}
%
where the jacobian, in this case, is:
%
\begin{align}
\matr{J} = 
\left[ 
\begin{array}[pos]{cc}
	\frac{\partial \vect{G}_q}{\partial \vect{q}} & \frac{\partial \vect{G}_q}{\partial \vect{v}}  \\
	\frac{\partial \vect{G}_v}{\partial \vect{q}} & \frac{\partial \vect{G}_v}{\partial \vect{v}} 
\end{array}
 \right]_n^{l+1}
 = 
\left[ 
\begin{array}[pos]{cc}
	\matr{I} & -h\matr{I}  \\
	-h \nabla_q \vect{f}^{l+1}  & \hat{\matr{M}} -h \nabla_v \vect{f}^{l+1}
\end{array}
 \right] 
\end{align}

Also in this case we defined $\left( \frac{\matr{M}^{l+1}+\matr{M}^l}{2} \right) = \hat{\matr{M}}$, that is often the constant mass $\matr{M}$ in many cases. 
By developing the expression above and by setting $\Delta\vect{q}^{l+1}=h \Delta\vect{v}^{l+1}$  one gets the Newton step procedure:

\begin{empheq}[box=\fbox]{align}
\left[ \hat{\matr{M}} - h^2 \nabla_q \vect{f}^{l+1} - h \nabla_v \vect{f}^{l+1} \right] \Delta\vect{v}^{l+1} &=
\left( \vect{v}^{l}-\vect{v}^{l+1} \right) \hat{\matr{M}} + h \vect{f}^{l+1} 
\label{eq:euler_qv1}
\\
\vect{v}^{l+1}_{n+1} &= \vect{v}^{l+1}_{n} + \Delta\vect{v}^{l+1}
\label{eq:euler_qv2}
\\
\vect{q}^{l+1} &= \vect{q}^{l} + h \vect{v}^{l+1}_{n+1} 
\label{eq:euler_qv3}
\end{empheq}

Some remarks here.

\begin{itemize}

	\item 
	In Eq.\ref{eq:euler_qv3} one could use $\Delta\vect{q}^{l+1}=h \Delta\vect{v}^{l+1}$ and rather write 
	$\vect{q}^{l+1}_{n+1} = \vect{q}^{l+1}_{n} + h \Delta\vect{v}^{l+1}_{n+1}$, but this is questionable.

	\item 
	The single first step of this NR process is exactly the Euler semi-implicit DVI in Chrono::Engine
	(not considering the constraints and contacts, here) because 
	one can express it with unknowns $\vect{v}^{l+1}_{n+1}$ rather than deltas, so it can be also:
%
\begin{align}
\left[ \hat{\matr{M}} - h^2 \nabla_q \vect{f}^{l+1} - h \nabla_v \vect{f}^{l+1} \right] \vect{v}^{l+1}_{n+1} &= 
\left[ \hat{\matr{M}} - h^2 \nabla_q \vect{f}^{l+1} - h \nabla_v \vect{f}^{l+1} \right] \vect{v}^{l+1}_{n} + \nonumber \\
& + \left( \vect{v}^{l}-\vect{v}^{l+1} \right) \hat{\matr{M}} + h \vect{f}^{l+1} 
\label{eq:euler_qv1b}
\\
\vect{q}^{l+1} &= \vect{q}^{l} + \vect{v}^{l+1}_{n+1} h
\label{eq:euler_qv2b}
\end{align}
%
and then, if at the first step we set $\vect{v}^{l+1}_0 = \vect{v}^{l}$, and assuming $\vect{f}^{l+1} \approx \vect{f}^{l}$, we simply get:
%
\begin{align}
\left[ \hat{\matr{M}} - h^2 \nabla_q \vect{f}^{l+1} - h \nabla_v \vect{f}^{l+1} \right] \vect{v}^{l+1} &= 
\left[ \hat{\matr{M}} - h^2 \nabla_q \vect{f}^{l+1} - h \nabla_v \vect{f}^{l+1} \right] \vect{v}^{l} + h \vect{f}^{l} 
\label{eq:euler_qv1c}
\\
\vect{q}^{l+1} &= \vect{q}^{l} + \vect{v}^{l+1}_{n+1} h
\label{eq:euler_qv2c}
\end{align}
%
(just by removing the stiffness and damping matrices $\nabla_q \vect{f}^{l+1}$, $\nabla_v \vect{f}^{l+1}$ if there are no finite elements, and by adding jacobians to satisfy bilateral constraints, one turns Eq.\ref{eq:euler_qv1c} into the same problem solved by the DVI timestepper of Chrono::Engine if without unilateral constraints)

\end{itemize}

\subsubsection{Euler implicit, for DAE}

Let's consider an index-2 DAE as in constrainted multibody mechanics, see \eqref{eq:daemechanics}.

There are different options for the DAE solution, for instance use constraints in DAE implicit form $F()$ to solve with Newton-Raphson, or do not solve them monolithically with DAE but just do projections on the manifold at each Newton iteration, etc. 
Here we enforce constraints (at the end of the time step) using a Newton-Raphson iteration, together with the Euler first order approximation used in the constraint-less case.

Add constraints $\vect{C}(\vect{q},t)=\vect{0}$. 
For brevity, denote 
\begin{align}
\matr{C}_q &= \nabla_q \vect{C}^T =  \frac{\partial \vect{C}}{\partial \vect{q}} \\
\matr{C}_t &= \nabla_t \vect{C}^T =  \frac{\partial \vect{C}}{\partial t}
\end{align}
Also, in sake of compactness we write
$\vect{C}^{l}$ for $\vect{C}(\vect{q}^{l},t^{l})$, as well as 
$\vect{C}^{l+1}$ for $\vect{C}(\vect{q}^{l+1},t^{l+1})$.

The Euler implicit method (backward Euler), for DAE with constraints satisfied at the end of timestep, is:

\begin{empheq}[left=\empheqlbrack]{align}
\vect{q}^{l+1}-\vect{q}^l - \vect{v}^{l+1} h = 0
\label{eq:eulerDAE_q}
\\
\hat{\matr{M}}(\vect{v}^{l+1}-\vect{v}^l) - h \vect{f}^{l+1} - h \matr{C}_q^T \vect{\lambda}^{l+1} = 0
\label{eq:eulerDAE_v}
\\
\vect{C}(\vect{q}^{l+1},t^{l+1}) = \vect{0}
\label{eq:eulerDAE_c}
\\
t^{l+1} = t^l + h
\end{empheq}

Let's call $\vect{G}=\left\{ \vect{G}_q, \vect{G}_v, \vect{G}_c \right\}$ the residual of the system above.
Solving with a Newton-Raphson iteration: 
%
\begin{align}
\left[ 
\begin{array}[pos]{ccc}
\matr{I} & -h\matr{I} & 0 \\
	-h \nabla_q \vect{f}^{l+1}  & \hat{\matr{M}} -h \nabla_v \vect{f}^{l+1} & - h \matr {C}_q^T  \\
	\matr {C}_q & 0 & 0
\end{array}
 \right]
\left\{ 
\begin{array}[pos]{c}
	\Delta\vect{q}^{l+1} \\ 
	\Delta\vect{v}^{l+1} \\ 
	\Delta\vect{\lambda}^{l+1}
\end{array}
 \right\} 
 = -\vect{G}
\end{align}
%

By developing the expression above and by setting $\Delta\vect{q}^{l+1}=\frac{h}{2}\Delta\vect{v}^{l+1}$ (see footnote in previous page) 
one gets the 'unrolled' Newton step procedure, where the hard part is the 1st step, the classical saddle-point problem:

\begin{empheq}[box=\fbox]{align}
\left[
\begin{array}[pos]{cc}
\left[ \hat{\matr{M}} - h^2 \nabla_q \vect{f}^{l+1} - h \nabla_v \vect{f}^{l+1} \right] & \matr{C}_q^T \\ 
\matr{C}_q & 0
\end{array}
\right] 
&
\left\{
\begin{array}[pos]{c}
\Delta\vect{v}^{l+1} \\
-h \Delta\vect{\lambda}^{l+1}
\end{array}
\right\}
 = \nonumber \\
 \left\{
\begin{array}[pos]{c}
\left( \vect{v}^{l}-\vect{v}^{l+1} \right) \hat{\matr{M}} + h \vect{f}^{l+1} + h \matr{C}_q^T \vect{\lambda}^{l+1} \\
\frac{\vect{C}^{l+1}}{h}
\end{array}
\right\} &
\label{eq:eulerDAE_qv1}
\\
\vect{v}^{l+1}_{n+1} &= \vect{v}^{l+1}_{n} + \Delta\vect{v}^{l+1}
\label{eq:eulerDAE_qv2}
\\
\vect{\lambda}^{l+1}_{n+1} &= \vect{\lambda}^{l+1}_{n} + \Delta\vect{\lambda}^{l+1}
\label{eq:eulerDAE_qv3}
\\
\vect{q}^{l+1} &= \vect{q}^{l} + h \vect{v}^{l+1}_{n+1} 
\label{eq:eulerDAE_qv4}
\end{empheq}

\subsubsection{Trapezoidal, for DAE}

A possibility is to add to Eqs. \ref{eq:trapzoidal_q2} and \ref{eq:trapzoidal_v2} the following constraint that requires positions to be satisfied at the end of the step:
\[
\vect{C}(\vect{q}^{l+1},t^{l+1})=\vect{0}
\] 
and add reaction forces $\vect{\lambda}$ too, as $\vect{f} + \nabla_v \vect{C} \vect{\lambda} =\matr{M}\vect{a} $.
This leads to the following trapezoidal rule for DAE:

\begin{empheq}[left=\empheqlbrack]{align}
\vect{q}^{l+1}-\vect{q}^l - \left( \frac{\vect{v}^{l+1}+\vect{v}^l}{2} \right) h = 0
\label{eq:trapzoidal_q3}
\\
\left( \vect{v}^{l+1}-\vect{v}^l \right) \left( \matr{M}^{l+1}+\matr{M}^{l} \right) - h \vect{f}^{l+1} -h \vect{f}^l  
- h \nabla_v \vect{C}^{l} \vect{\lambda}^{l} - h \nabla_v \vect{C}^{l+1} \vect{\lambda}^{l+1}= 0
\label{eq:trapzoidal_v3}
\\
\vect{C}(\vect{q}^{l+1},t^{l+1})=\vect{0}
\label{eq:trapzoidal_c3}
\\
t^{l+1} = t^l +h
\end{empheq}

This done, one can express a Newton-Raphson process to compute the unknowns $\vect{v}^{l+1}$, $\vect{v}^{l+1}$, $\vect{\lambda}^{l+1}$ that give a zero residual $\vect{G}$ for the three equations above:
%
\begin{align}
\left[ 
\begin{array}[pos]{ccc}
\matr{I} & -\frac{h}{2}\matr{I} & 0 \\
	-\frac{h}{2} \nabla_q \vect{f}^{l+1}  & \hat{\matr{M}} -\frac{h}{2} \nabla_v \vect{f}^{l+1} &  -\frac{h}{2} \matr {C}_q^{l+1,T}  \\
	\matr {C}_q^{l+1} & 0 & 0
\end{array}
 \right]
\left\{ 
\begin{array}[pos]{c}
	\Delta\vect{q}^{l+1} \\ 
	\Delta\vect{v}^{l+1} \\ 
	\Delta\vect{\lambda}^{l+1}
\end{array}
 \right\} 
 = -\vect{G}
\end{align}

This can be unrolled to work with a more friendly linear system with hermitian matrix and two following uncoupled steps:

\begin{empheq}[box=\fbox]{align}
&
\left[
\begin{array}[pos]{cc}
\left[ \hat{\matr{M}} - \frac{h^2}{4} \nabla_q \vect{f}^{l+1} - \frac{h}{2} \nabla_v \vect{f}^{l+1} \right] & \matr {C}_q^{l+1,T} \\ 
\matr {C}_q^{l+1} & 0
\end{array}
\right] 
\left\{
\begin{array}[pos]{c}
\Delta\vect{v}^{l+1} \\
-\frac{h}{2} \Delta\vect{\lambda}^{l+1}
\end{array}
\right\}
 = \nonumber \\
& 
 \left\{
\begin{array}[pos]{c}
\left( \vect{v}^{l}-\vect{v}^{l+1} \right) \hat{\matr{M}} 
+ \frac{h}{2}\left( \vect{f}^{l} + \vect{f}^{l+1} + \matr{C}_q^{l,T} \vect{\lambda}^{l} + \matr{C}_q^{l+1,T} \vect{\lambda}^{l+1} \right) \\
\frac{1}{h}\vect{C}^{l+1}
\end{array}
\right\} 
\label{eq:trapezDAE_qv1}
\\ &\quad
\vect{v}^{l+1}_{n+1} = \vect{v}^{l+1}_{n} + \Delta\vect{v}^{l+1}
\label{eq:trapexDAE_qv2}
\\ &\quad
\vect{\lambda}^{l+1}_{n+1} = \vect{\lambda}^{l+1}_{n} + \Delta\vect{\lambda}^{l+1}
\label{eq:trapezDAE_qv3}
\\ &\quad
\vect{q}^{l+1} = \vect{q}^{l} + \frac{h}{2} \left( \vect{v}^{l} + \vect{v}^{l+1}_{n+1} \right)
\label{eq:trapezDAE_qv4}
\end{empheq}

\subsubsection{Newmark, for ODE and DAE}

The Newmark family of second-order integrators is very popular in the FEM community.

One has

\begin{empheq}[left=\empheqlbrack]{align}
\vect{q}^{l+1}-\vect{q}^l - h \vect{v}^l - \frac{h^2}{2} \left[ (1-2\beta) \vect{a}^l + 2\beta \vect{a}^{l+1} \right]  = 0
\label{eq:newmark_q}
\\
\vect{v}^{l+1}-\vect{v}^l - h \left[(1-\gamma) \vect{a}^l + \gamma \vect{a}^{l+1} \right]= 0
\label{eq:newmark_v}
\\
\matr{M} \vect{a}^{l+1} + (\matr{C}_q^T\vect{\lambda}-\vect{f})^{l+1}  = 0
\label{eq:newmark_a}
\\
t^{l+1} = t^l + h
\end{empheq}

Different values of the $\gamma$ and $beta$ parameters provide different behaviors.
\begin{itemize}
	\item The $\gamma$ parameter usually ranges in the $[1/2, 1]$ interval. 
	\item For $\gamma = 1/2$, no numerical damping.
	\item For $\gamma > 1/2$, numerical damping increases.
	\item The $\beta$ parameter usually ranges in the $[0, 1]$ interval.
	\item For $\beta = 1/4, \gamma = 1/2$ one gets the constant acceleration method.
	\item For $\beta = 1/6, \gamma = 1/2$ one gets the linear acceleration method.
	\item The method is second order accurate only for $\gamma = 1/2$.
\end{itemize}

For this type of method, considering constraints being enforced in the DAE, the Newton iteration is represented again by a linear problem, this time in accelerations:

\begin{empheq}[box=\fbox]{align}
&
\left[
\begin{array}[pos]{cc}
 \matr{H} & \matr{\overline{C}}_q^T \\ 
\matr{\overline{C}}_q & 0
\end{array}
\right] 
\left\{
\begin{array}[pos]{c}
\Delta\vect{a}^{l+1} \\
\Delta\vect{\lambda}^{l+1}
\end{array}
\right\}
 = \nonumber \\
& 
 \left\{
\begin{array}[pos]{c}
\matr{M}\vect{a}^{l+1}
+ (\matr{C}_q^T\vect{\lambda}-\vect{f})^{l+1} 
\\
\frac{1}{\beta h^2}\vect{C}^{l+1}
\end{array}
\right\} 
\label{eq:newmark_qv1}
\\ &\quad
\vect{a}^{l+1}_{n+1} = \vect{a}^{l+1}_{n} + \Delta\vect{a}^{l+1}
\label{eq:newmark_qv2}
\\ &\quad
\vect{\lambda}^{l+1}_{n+1} = \vect{\lambda}^{l+1}_{n} + \Delta\vect{\lambda}^{l+1}
\label{eq:newmark_qv3}
\\ &\quad
\vect{v}^{l+1} = \vect{v}^l + h \left[(1-\gamma) \vect{a}^l + \gamma \vect{a}^{l+1} \right]
\label{eq:newmark_qv4}
\\ &\quad
\vect{q}^{l+1} = \vect{q}^l + h \vect{v}^l + \frac{h^2}{2} \left[ (1-2\beta) \vect{a}^l + 2\beta \vect{a}^{l+1} \right] 
\label{eq:newmark_qv5}
\end{empheq}

where 
\[
\matr{H} = 
\left[
\matr{M}
- \gamma h  \nabla_v \vect{f}^{l+1}
- \beta h^2 \nabla_q \vect{f}^{l+1}
+ \beta h^2 \left[ (\matr{M}\vect{a})_q + (\matr{C_q}^T\vect{\lambda})_q \right]
\right]
\]

\subsubsection{HHT, for DAE}

The HHT integrator is a generalization of the Newmark family of second-order integrators, and provides a control on the numerical dissipation yet retaining a second order accuracy as the trapezoidal method.

One has

\begin{empheq}[left=\empheqlbrack]{align}
\vect{q}^{l+1}-\vect{q}^l - h \vect{v}^l - \frac{h^2}{2} \left[ (1-2\beta) \vect{a}^l + 2\beta \vect{a}^{l+1} \right]  = 0
\label{eq:hht_q}
\\
\vect{v}^{l+1}-\vect{v}^l - h \left[(1-\gamma) \vect{a}^l + \gamma \vect{a}^{l+1} \right]= 0
\label{eq:hht_v}
\\
\matr{M} \vect{a}^{l+1}+ (1+\alpha)(\matr{C}_q^T\vect{\lambda}-\vect{f})^{l+1} -  \alpha(\matr{C}_q^T\vect{\lambda}-\vect{f})^{l}  = 0
\label{eq:hht_a}
\\
t^{l+1} = t^l +h
\end{empheq}

The HHT method provides the A-stability and order provided that $\alpha \in [-\frac{1}{3},0]$ and 
%
\begin{align}
\gamma = \frac{1-2\alpha}{2} \quad \beta=\frac{(1-\alpha)^2}{4}
\end{align}

The closer to 0 is $\alpha$, the less damping has the method, where for $\alpha=0$ one has precisely the trapezoidal method.

Looking at the work in \cite{negrutHHT07}, one sees that the Newton iteration is represented again by a linear problem with unknown accelerations and constraint forces:

\begin{empheq}[box=\fbox]{align}
&
\left[
\begin{array}[pos]{cc}
 \matr{H} & \matr{\overline{C}}_q^T \\ 
\matr{\overline{C}}_q & 0
\end{array}
\right] 
\left\{
\begin{array}[pos]{c}
\Delta\vect{a}^{l+1} \\
\Delta\vect{\lambda}^{l+1}
\end{array}
\right\}
 = \nonumber \\
& 
 \left\{
\begin{array}[pos]{c}
\frac{1}{1+\alpha}(\matr{M}\vect{a}^{l+1})
+ (\matr{C}_q^T\vect{\lambda}-\vect{f})^{l+1} 
- \frac{\alpha}{1+\alpha}(\matr{C}_q^T\vect{\lambda}-\vect{f})^{l}
\\
\frac{1}{\beta h^2}\vect{C}^{l+1}
\end{array}
\right\} 
\label{eq:hht_qv1}
\\ &\quad
\vect{a}^{l+1}_{n+1} = \vect{a}^{l+1}_{n} + \Delta\vect{a}^{l+1}
\label{eq:hht_qv2}
\\ &\quad
\vect{\lambda}^{l+1}_{n+1} = \vect{\lambda}^{l+1}_{n} + \Delta\vect{\lambda}^{l+1}
\label{eq:hht_qv3}
\\ &\quad
\vect{v}^{l+1} = \vect{v}^l + h \left[(1-\gamma) \vect{a}^l + \gamma \vect{a}^{l+1} \right]
\label{eq:hht_qv4}
\\ &\quad
\vect{q}^{l+1} = \vect{q}^l + h \vect{v}^l + \frac{h^2}{2} \left[ (1-2\beta) \vect{a}^l + 2\beta \vect{a}^{l+1} \right] 
\label{eq:hht_qv5}
\end{empheq}

where 
\[
\matr{H} = 
\left[
\matr{M}
- \gamma h  \nabla_v \vect{f}^{l+1}
- \beta h^2 \nabla_q \vect{f}^{l+1}
+ \beta h^2 \left[ (\matr{M}\vect{a})_q + (\matr{C_q}^T\vect{\lambda})_q \right]
\right]
\]

Note the term 
$\beta h^2 \left[ (\matr{M}\vect{a})_q + (\matr{C_q}^T\vect{\lambda})_q \right]$, 
could be omitted for faster performance, at the risk of lower Newton convergence.

\subsubsection{Generalized-$\alpha$, for DAE}

The generalized-$\alpha$ integrator is an evolution of the Newmark and HHT methods, where a single parameter $\rho_\infty$ can be used to define the numerical damping. Just like the HHT integrator, it is a second-order implicit integrator that provides a control on the numerical dissipation yet retaining a second order accuracy as the trapezoidal method.

We recall the second barrier of the Dahlquist theorem: there are no explicit A-stable and linear multistep methods, and the implicit ones have order of convergence at most 2. The trapezoidal rule has the smallest error constant amongst the A-stable linear multistep methods of order 2. The HHT and the 
generalized-$\alpha$ methods are among the few timesteppers of practical interest that feature the second order of convergence, and that can solve constrained DAEs. In fact the trapezoidal method is confined mostly to ODEs since it gives oscillatory unstable reactions in constraints if used in DAEs.

In generalized-$\alpha$ one has

\begin{empheq}[left=\empheqlbrack]{align}
\vect{q}^{l+1}-\vect{q}^l - h \vect{v}^l - \frac{h^2}{2} \left[ (1-2\beta) \vect{a}^l + 2\beta \vect{a}^{l+1} \right]  = 0
\label{eq:alpha_q}
\\
\vect{v}^{l+1}-\vect{v}^l - h \left[(1-\gamma) \vect{a}^l + \gamma \vect{a}^{l+1} \right]= 0
\label{eq:alpha_v}
\\
(1-\alpha_m)\vect{a}^{l+1} + \alpha_m\vect{a}^{l} = (1-\alpha_f)\vect{a*}^{l+1} + \alpha_f\vect{a*}^{l}
\label{eq:alpha_aa}
\\
\matr{M} \vect{a*}^{l+1}+ (\matr{C}_q^T\vect{\lambda}-\vect{f})^{l+1} = 0
\label{eq:alpha_a}
\\
t^{l+1} = t^l + h
\end{empheq}

The coefficients in the generalized-$\alpha$ method are automatically set as following, 
once the spectral value $\rho_\infty$ is set in the $[0,1]$ range:

\begin{align}
\alpha_m = \frac{2 \rho_\infty - 1}{1+ \rho_\infty} 
\\
\alpha_f = \frac{\rho_\infty}{1+ \rho_\infty} 
\\
\beta = \frac{1}{4} \left(\gamma + \frac{1}{2} \right)^2
\\
\gamma =\frac{1}{2} + \alpha_f - \alpha_m
\end{align}

Note that for $\rho_\infty = 0$ one has the maximum asymptotic dissipation, 
and the method introduce the maximum numerical damping (i.e. whatever frequency is 
damped in a single step $h$).

One can see that the HHT method is like the generalized-$\alpha$ with $\alpha_m=0$, $\alpha_f=\alpha$. Indeed they behave in a similar way, and they also are the only implicit methods of practical interest that feature second-order accuracy and that can solve constrained DAEs.

On the other side, the limit of $\rho_\infty = 1$ has no numerical dissipation,
just like in the trapezoidal method. Differently from the trapezoidal method, though, this 
integrator behaves well with constrains and does not lead to oscillatory reactions in constraints.

Of course it is always convenient to introduce some artificial numerical dissipation even if the system, at the physical level, is not dissipative at all. This because, even with simplectic integrators, numerical issues might lead to increasing hamiltonian after many oscillations - something that in the long run is more likely to cause divergence and bad artifacts.

Usually, an intermediate $\rho_\infty \in [0,1]$ value is used. This helps the user to discard unnecessary high frequency oscillations that are of little interest, ant that would just hamper the performance of the solver. Setting a proper value of $\rho_\infty$ in general, non-linear cases, is often an heuristic process. One can start with a near zero value and raise it by repeating the same  test simulation, until unnecessary high frequency oscillations start to appear.

The Newton iteration is represented again by a linear problem. As with the HHT case, the iteration can be expressed at the acceleration level, at the velocity level, at the configuration (i.e. position) level. Here we report the iteration with position (increments) as unknowns:

\begin{empheq}[box=\fbox]{align}
&
\left[
\begin{array}[pos]{cc}
 \matr{H} & \matr{\overline{C}}_q^T \\ 
\matr{\overline{C}}_q & 0
\end{array}
\right] 
\left\{
\begin{array}[pos]{c}
\Delta\vect{q}^{l+1} \\
\Delta\vect{\lambda}^{l+1}
\end{array}
\right\}
 = \nonumber \\
& 
 \left\{
\begin{array}[pos]{c}
\vect{r}^q
\\
\vect{r}^{\lambda}
\end{array}
\right\} 
\label{eq:alpha_qv1}
\\ &\quad
\vect{q}^{l+1}_{n+1} = \vect{q}^{l+1}_{n} + \Delta\vect{q}^{l+1}
\label{eq:alpha_qv2}
\\ &\quad
\vect{v}^{l+1}_{n+1} = \vect{v}^{l+1}_{n} + \gamma' \Delta\vect{q}^{l+1}
\label{eq:alpha_qv3}
\\ &\quad
\vect{a*}^{l+1}_{n+1} = \vect{a*}^{l+1}_{n} + \beta' \Delta\vect{q}^{l+1}
\label{eq:alpha_qv4}
\\ &\quad
\vect{\lambda}^{l+1}_{n+1} = \vect{\lambda}^{l+1}_{n} + \Delta\vect{\lambda}^{l+1}
\label{eq:alpha_qv5}
\end{empheq}

where 
\[
\matr{H} = 
\left[
      \nabla_q \left[ \matr{M}\vect{a}^{l+1} -\vect{f}^{l+1} + \matr{C}_q^T\vect{\lambda}^{l+1} \right]
- \gamma'   \nabla_v \vect{f}^{l+1}
+ \beta'  \matr{M}\vect{a} -
\right]
\]
and
\[
\beta' = \frac{1-\alpha_m}{h^2 \beta(1-\alpha_f)}
\]
\[
\gamma' = \frac{\gamma}{h \beta}
\].

After the iteration has converged, one updates
\[
\vect{a}^{l+1} = \vect{a}^{l} + \frac{1-\alpha_f}{1-\alpha_m} \vect{a*}^{l+1}
\]
and similarly, before starting the iteration, $\vect{q}^{l+1}_{n}$, $\vect{v}^{l+1}_{n}$, $\vect{a}^{l+1}_{n}$ are initialized with proper formulas.

Note that the term 
$\nabla_q \left[ \matr{M}\vect{a}^{l+1} -\vect{f}^{l+1} + \matr{C}_q^T\vect{\lambda}^{l+1} \right]$ 
could be simplified, approximating to $- \nabla_q \vect{f}^{l+1} $ for higher performance, at the risk of a worse convergence in the Newton loop.

Note that, although the method is unconditionally stable, this theoretical result
holds for linearly stiff problems. This means that the method might still diverge if
too large $h$ is used in real cases that feature marked nonlinear geometric behaviour or highly nonlinear material properties.

Also, the Newton iteration is not guaranteed to converge in highly nonlinear cases.
For instance, when dealing with buckling and near-bifurcation situations, some
continuation or globalization strategies in the Newton computation are required, otherwise
smaller timesteps must be used anyway.

\subsection{Abstracting a common solver architecture}

Looking at equations in boxed frames for all the DAE time-steppers, one can see that all methods share a common feature: at each time step there is one (or more) linear system to solve that has a saddle-point structure, usually (but not always) with unknowns about velocities and reactions, always with such structure:
%
\begin{align}
\left[
\begin{array}[pos]{cc}
\matr{H}  & \matr{D}^t \\ 
\matr{D} & 0
\end{array}
\right] 
&
\left\{
\begin{array}[pos]{c}
\vect{x} \\
\vect{\lambda}
\end{array}
\right\}
 = 
 \left\{
\begin{array}[pos]{c}
\vect{a} \\
\vect{b}
\end{array}
\right\} &
\label{eq:structure}
\end{align}

All implicit integrators therefore will interface to the physical system via two procedures: one that can provide the $\matr{H}$ jacobians (usually from mass, stiffness and damping matrices), and one that can provide the $\vect{a}$ terms (usually from internal+external forces). If constraints are used too, also $\matr{D}$ constraint jacobians and $\vect{b}$ terms should be fetched.

\subsection{Accuracy, convergence, stability}

In general, shorter time steps allow integration schemes to achieve a better \textit{accuracy}. Also, short time steps are
used to avoid \textit{divergence}, that is the behavior that amplifies errors leading to solutions that explode to infinite values after one or 
few diverging oscillations, where divergence is more likely to happen if dealing with stiff equations. Some methods are less prone 
to divergence even if dealing with stiff equations, that is, they exploit better qualities of \textit{stability}, and 
there are methods that are \textit{unconditionally stable}, that is, their stability is guaranteed regardless of the integration time step. 
How these concepts expressed in a formal way?

We refer the reader for more details on this subject, here we just recall basic terminology and concepts.

\begin{itemize}

	\item
	The \textbf{global error} is the error accumulated after $n$ integration steps, that is, the error at the end of a sequence of time steps 
	as the difference between the exact value $x(t_{n})$ and the computed value $x_{n}$.
	\[
	\epsilon_n = x(t_{n}) - x_{n}
	\]
	Note that in most cases there is no analytical solution $x(t)$ even for moderately complex
	ODEs and DAEs, hence the exact error value cannot be computed. In lucky scenarios, maybe one can just estimate upper bounds.

	\item 
	The \textbf{local truncation error} of a method is the error committed over a single step when moving to the approximated $x_{n+1}$, respect to the exact $x(t_{n+1})$ 
	\[
	\tau_{n+1} = \left(x(t_{n+1}) - x(t_n) \right) - \left(x_{n+1}  - x_{n} \right)
	\]
	Note that for all methods presented so far, one can express $x_{n+1}$ as a function $x_{n+1} = L(x_{n},x_{n-1},x_{n-2},\ldots,h)$
	and therefore 
	\[
	\tau_{n+1} =  L(x_{n},x_{n-1},x_{n-2},\ldots,h)  -  x(t_{n+1})
	\]

	\item 
	A method is said \textbf{consistent} if 
	\[
	\lim_{h \rightarrow 0} \frac{L(x_{n},x_{n-1},x_{n-2},\ldots,h)}{h} = 0
	\]

	\item
	A method has \textbf{order} $p$ if
	\[
	\tau_{n+1} = O(h^{p+1}) \quad \mathrm{as} \quad h\rightarrow 0
	\]

	\item 
	A method is said \textbf{convergent} if $x_{n}$ approaches the true solution $x(t_{n})$ as the time step is refined to zero: $\lim_{h \rightarrow 0} x_{n} = x(t_{n})$, or equivalently
	\[
	\lim_{h \rightarrow 0} \max_{0<n<N} \| \epsilon_n \| = 0
	\]

	\item 
	A method is said to be \textbf{stable} if there exists an integration time step $h_0>0$ so that for any $h\in[0,h_0]$, 
	a finite variation of the state at time $t_n$ leads only a non-increasing variation of the state at a successive time.
	Consistency and convergence are necessary and sufficient conditions for stability. See also \textit{zero-stability}.

	\item
	A system is said to be \textbf{stiff} if stability requirements, rather than those of accuracy, constrain the step length to be small. 
	Stiff systems, in a mechanical intuitive sense, are those that exhibit sudden changes of values in the state vector, as in jumps, sudden
	oscillations, quick decays.

	The big issue of stiff systems is that, in common engineering scenarios, those quick phenomena are often interleaved by calm and smooth time intervals, 
	so if one picks a short time step to be stable across the quick phenomena, it also wastes computational time during the smooth events where a long time step 
	woudl have guaranteed stability anyways. Variable-time-stepping schemes can fix this to some extent, but not always; a more radical solution is to use 
	implicit methods that exhibits \textit{absolute} stability -see later- and that, for this reason, can overcome stiff phenomena even with large time steps.

	\item 
	In order to study the stability of methods, one can restrict the attention to a special case of ODE, that is the \textit{linearly stiff} differential equation
	\[
	\dot{x}= \lambda x  \quad \mathrm{for} \; \lambda \in \mathbb{C}_- \, , \; x_0 \in \mathbb{R}
	\]
	for whom, by the way, one already knows the analytical solution $x(t)=x_0 e^{\lambda t}$. This can be stable only if $\mathrm{Re} \lambda < 0$ otherwise $x(t)$ will grow to $\infty$ anyway, also in the analytical solution regardless of the numerical scheme, that is why we restrict $\lambda$ to be a complex but excluding the positive real plane: $\lambda \in \mathbb{C}_-$, thus decaying from $x_0$ and converging as $\lim_{t\rightarrow 0} x(t)=0$.
	Finally, one can intuitively see that the larger $|\lambda|$, the faster the decay, thus the stiffer the equation.

	\begin{figure}[ht]
		\centering
			\includegraphics[width=0.48\textwidth]{stability_euler.png}
			\includegraphics[width=0.48\textwidth]{stability_ieuler.png}
		\caption{Stability region of explicit Euler (left) and stability region of implicit Euler (right)}
		\label{fig:stability_euler}
	\end{figure}
	\begin{figure}[ht]
		\centering
			\includegraphics[width=0.48\textwidth]{stability_trapezoidal.png}
			\includegraphics[width=0.48\textwidth]{stability_adams.png}
		\caption{Stability region of trapezoidal method (left) and stability region of Adams-Bashforth (right)}
		\label{fig:stability_euler}
	\end{figure}
	\begin{figure}[ht]
		\centering
			\includegraphics[width=0.48\textwidth]{stability_runge.png}
			\includegraphics[width=0.48\textwidth]{stability_bdf.png}
		\caption{Stability region of explicit Runge-Kutta (left) and stability region of implicit BDF (right)}
		\label{fig:stability_euler}
	\end{figure}

	\item 
	If we can express an integration method, applied to the linearly stiff test equation above, as a product $x_{n+1} = \Phi(h \lambda) x_{n}$, the term $\Phi(h \lambda)$ is called \textbf{stability function}, and one can see that the series converges $\lim_{n \rightarrow \infty} x_{n} = 0$ only for $|\Phi(h \lambda)| < 1$. 
	For the values $h \lambda$ where it holds, the method is called \textbf{linearly stable}.

	\item 
	For a given method, one can plot the region(s) where the linearly stable property holds, as a function of the $h \lambda$ value in the complex plane,
	obtaining areas shown in Fig.\ref{fig:stability_euler}. Those region where the stability function is $|\Phi(h \lambda)| < 1$, is called \textbf{stability region} $\mathcal{R}_A$of the method
	\[
	\mathcal{R}_A = \left\{ h \lambda \, : \,  |\Phi(h \lambda)| < 1 \right\}
	\]

	\item
	All methods of practical interest contain the origin $0,0$ of the plane, that is called \textbf{zero-stability}. Note that, for bounded stability regions like circles etc., the larger the $h$ value, or the larger the $\lambda$ value, the farther we go from the origin and the easier we go out of the stability region. This explains why the need of using short $h$ if meeting large stiffness $\lambda$.

	\item 
	Methods whose stability region covers at least the entire half plane of negative real values, are called \textit{absolutely stable}, hence the property called \textbf{A-stability}. 
	\[
	\mathrm{A-stability} \; \Leftrightarrow \; \mathcal{R}_A \supseteq \left\{ h \lambda \, : \,  \mathrm{Re} \{h \lambda\} < 0 \right\}
	\]
	Those method converges to zero with the test function regardless of how much stiffness $\lambda$ or how large is $h$. Example: the implicit Euler, the trapezoidal method. Note: not all implicit methods are A-stable, for example BDF of order greater than 2 cover a large portion of the left half plane, but not entirely.

	\item 
	Theorem (the \textbf{second Dahlquist barrier}). An A-stable multistep method must be of order $r\leq 2$. 
	\footnote{This means that, despite the efforts, no one would ever invent a method that has absolute stability and integration order greater than 2. If one needs an
	high order because he is concerned with extreme precision, as it happens in celestial mechanics when studying trajectories, probably he better give up with 
	the requirement of absolute stability (luckily enough, in those cases there are no problems of stiff integration).}

	\item 
	A more relaxed requirement it the \textbf{A($\alpha$)-stability}, where the stability region must include a cone in the left semiplane, with aperture $2\alpha$:
	\[
	\mathrm{A}\alpha\mathrm{-stability} \; \Leftrightarrow \; \mathcal{R}_A \supseteq \left\{ h \lambda \, : \,  -\alpha < \pi - \mathrm{arg}(h \lambda) < \alpha \right\}
	\]

	\item 
	A more stringent requirement is the \textbf{L-stability}, that requires the method to be A-stable plus $\lim_{ h \lambda \rightarrow -\infty} |\Phi(h \lambda)| = 0 $. For instance, the trapezoidal method is A-stable but not L-stable because its $|\Phi(h \lambda)| \rightarrow 1$ when $\lambda \rightarrow -\infty$\footnote{In practical scenarios, this can be seen as the known issue of the trapezoidal method: it dampens decays and harmonics like other implicit methods, but the damping is less and less noticeable for increasing frequencies.}. The implicit Euler is L-stable.

	\item
	From the stronger to the more relaxed property:
	\[
	\mathrm{L-stability}  \Rightarrow \mathrm{A-stability}  \Rightarrow \mathrm{A}\alpha\mathrm{-stability} \rightarrow \mathrm{A}_0\mathrm{-stability}  
	\]
	From a practical perspective, methods that stay on the left of this scale are more and more robust (in the computer science sense, that is, they cope better with unknown and unpredictable systems). For instance, the implicit Euler, although a first order and not so accurate integrator, is often used in VR, videogames and real-time applications because of its L-stability, that make it able to handle sudden stiff situations without exploding into divergence.

	\item
	Remember that the stability classes introduced so far are relative to a linearly-stiff test function, but in real scenarios the right hand side can be highly non-linear, therefore it is not guaranteed that even a L-stable method will work all the times. For example, divergence or breakdown of the method could happen for many reasons such as missed convergence in Newton-Raphson iterations, resonance and aliasing with rheonomic terms, finite floating point precision, etc. For instance, if simulating contacts with stiff penalty functions, also L-stable methods could diverge or give unusable results. 
	This said, anyway, L-stable and A-stable method are much less likely to have problems than others.

	\item
	Despite the superior stability properties, implicit methods have some drawbacks. First, they introduce some numerical damping, that is more and more visible the larger the $h$ time step. For example, the implicit Euler for DAEs presented in these pages will approach a static analysis for $h\rightarrow \infty$ - look at the formulas. This can be a positive side effect (this artificial numerical damping can be welcome to make the simulation more stable and robust), but if not needed, few implicit methods allow to adjust it, one notable case is the generalized-$\alpha$ method.

	\item 
	Another drawback of implicit methods: they require a complex implementation leading to at least 3-4 Newton-Raphson iterations at each time step (hence the need of solving 3-4 linear systems per time step). Explicit integrators on the other hand do not need any linear system to be solved, so they require much less CPU time per each time step. Therefore, implicit integrators pay back if you use large time steps thus making good use of their stability properties. Otherwise, if you must use extremely short time steps anyway, for example for geometric reasons such as when studying high-speed collisions with thin objects, maybe that explicit integrators would be more efficient (as they would be stable anyway with such short time steps).

\end{itemize}

%%%%%%%%%Bibliography%%%%%%%%
\bibliographystyle{plain}

%\bibliography{../../../Papers/BibFiles/LaTeXrelated/BibMBS}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

### 2020-05-29 courses/computational_mechanics/stabilityplots.m

@@ -1,0 +1,201 @@
+close all
+clear all
+
+C = 'color'; c = {'b','r','g','m','y','c'};
+x = [0 0]; y = [-8 8]; K = 'k'; LW = 'linewidth'; FS = 'fontsize';
+
+%%%%%%%%%%%%%%%%%
+
+figure;
+hold on;
+
+plot(y,x,K,LW,1, 'HandleVisibility','off'), plot(x,y,K, 'HandleVisibility','off')
+
+t = chebfun('t',[0 2*pi]);
+z = exp(1i*t); r = z-1;
+
+s = 1; 
+P = feval(r./s, linspace(0, 2*pi, 100))';
+shp = polyshape(real(P), imag(P));
+plot(shp, 'EdgeColor','green', 'FaceColor','green','FaceAlpha',0.1, 'DisplayName', 'Euler');
+
+
+axis([-3 1 -2 2]), axis square, grid on
+xlabel('Re(h\lambda)');
+ylabel('Im(h\lambda)');
+
+title('Explicit Euler')
+saveas(gcf,'stability_euler.png');
+
+%%%%%%%%%%%%%%%%%
+
+figure;
+hold on;
+
+plot(y,x,K,LW,1, 'HandleVisibility','off'), plot(x,y,K, 'HandleVisibility','off')
+
+d = 1-1./z; r = 0;
+%for i = 1:4
+%  r = r+(d.^i)/i;
+%  plot(r,C,c{i},LW,2)
+%end
+
+r = r+(d.^1)/1;
+P = feval(r, linspace(0, 2*pi, 100))';
+shp = polyshape(real(P), imag(P));
+shi = addboundary(shp,[-10,-10,10,10],[-10,+10,+10,-10]);
+plot(shi, 'EdgeColor','green', 'FaceColor','green','FaceAlpha',0.1, 'DisplayName', 'implicit Euler');
+
+
+axis([-1 3 -2 2]), axis square, grid on
+xlabel('Re(h\lambda)');
+ylabel('Im(h\lambda)');
+
+title('Implicit Euler')
+saveas(gcf,'stability_ieuler.png');
+
+
+
+%%%%%%%%%%%%%%%%%
+
+figure;
+hold on;
+
+plot(y,x,K,LW,1, 'HandleVisibility','off'), plot(x,y,K, 'HandleVisibility','off')
+
+shi = polyshape([-10,-10,0,0],[-10,+10,+10,-10]);
[…]

### 2020-06-08 courses/computational_mechanics/computational_mechanics.TEX
Updated notes on computational mechanics

@@
 \usepackage{nccmath}
+
+% For bibliography: use biber and biblatex. 
+% NOTE: your IDE build toolchain should use biber instead of bibtex when compiling.
+\usepackage[backend=biber]{biblatex}
+
+% Use \addbibresource{...}  to use .bib databases from our GIT repository if possible.
+% ..using relative path in this GIT:
+\addbibresource{"../../bibliography/refsFEM.bib"}
+\addbibresource{"../../bibliography/refsMBS.bib"}
 
@@
 When possible, we highlight the similarities between the vector/matrix notation and the tensor notation.
+The reader is invited to look for additional details on the topic in textbooks such as \cite{Hackbusch2012,Truesdell2004}.
 
@@
 
+\subsection{Exercises}
+
+{
+\small
+
+\begin{enumerate}
+
+	\item Write a software procedure to compute the 3x3 stress tensor from a given 3x3 strain tensor and a given fourth-order $\tensor{C}$ elastic tensor:
+	\[
+	\tensor{\sigma} = \tensor{C} : \tensor{\epsilon}
+	\]
+	You can make assumptions like material isotropy to justify that you do not need to pass all 81 components of $\tensor{C}$.
+
+	\item Write $\tensor{\sigma} = \tensor{C} : \tensor{\epsilon}$ using the Voigt notation and conventional linear algebra, where one can exploit the symmetry of $\tensor{\epsilon}$ and $\tensor{\sigma}$ to rewrite stress and strains as simple column vectors:
+	$\tensor{\sigma} = \{ \sigma_{xx}, \sigma_{yy}, \sigma_{zz}, \sigma_{xy}, \sigma_{zx}, \sigma_{yz} \}^T 
+	                  = \{ \sigma_x, \sigma_y, \sigma_z, \tau_{xy}, \tau_{zx}, \tau_{yz} \}^T$,
+	$\tensor{\epsilon} = \{ \epsilon_{xx}, \epsilon_{yy}, \epsilon_{zz}, 2\epsilon_{xy}, 2\epsilon_{zx}, 2\epsilon_{yz} \}^T  
+	                   = \{ \epsilon_x, \epsilon_y, \epsilon_z, \gamma_{xy}, \gamma_{zx}, \gamma_{yz} \}^T $ 
+	and $\tensor{C}$ as a 6x6 matrix.
+	Note the use of the engineering strain definition $\epsilon_{xy} = \frac{1}{2}\gamma_{xy}$.
+	Demonstrate that one needs at most 21 independent components to define the symmetric $\tensor{C}$ matrix.
+
+\end{enumerate}
+
+}
+
 \newpage
@@
 Thie following is a short overview of important special cases of PDEs. 
+The interested reader can find more details on PDEs in the area of mechanics on specialized textbooks such as \cite{cosserat1909,Truesdell2004}.
 
@@
 
+\subsection{Exercises}
+
+{
+\small
+
+\begin{enumerate}
+
+	\item Write the strong form of the straight Euler-Bernoulli beam theory in linear elasticity. Extend it to the case of dynamics, adding inertial forces caused by a linear mass density of the beam.
+
+	\item Write the strong form of the straight Timoshenko beam theory, where shear is taken into account. Extend it to the case of dynamics, adding inertial forces.
+
+	\item Write the strong form of the geometrically-exact beam theory (Cosserat rods).  Extend it to the case of dynamics, adding inertial forces.
+
+	\item Write the strong form of the Kirchhoff-Love plate, where one assumes that segments normal to the mid-surface remain straight and normal to the mid-surface after deformation.
+Add the contribution of inertial forces.
+	Write the strain-displacement relations to compute $\epsilon_{\alpha\beta}$ from the normal displacement field $u$, where $\alpha$ and $\beta$ can be either direction 1 or 2 of the plate, and show that $\epsilon_{\alpha 3}$ and $\epsilon_{33}$ are always null.
+
+	\item Write the strong form of the Reissner-Mindlin plate, that extends the Kirchhoff-Love plate theory by taking into account also the effect of shear. Add the contribution of inertial forces.
+
+\end{enumerate}
+
+}
+
 \newpage
@@
 In the following we will focus on the finite element Galerkin method.
+
+Deeper insights on the finite element method can be found in the books of the authors that pioneered it, such as \cite{zienkiewiczFEA1977,tjrhughes87,bathe1982,wriggers2006} to name a few.
 
@@
 More in detail, if the same approximation space is used for both $u$ and $v$, this is sometimes referred as \textbf{Bubnov-Galerkin} method, and if a different approximation space is chosen for $v$, this is called \textbf{Petrov-Galerkin} method.
-The modern mathematical setting and terminology for the finite element method is credited to R.Courant\footnote{It is remarkable how many bright scientist attended the University of Göttingen.}.
+The modern mathematical setting and terminology for the finite element method is credited to R.Courant\footnote{It is remarkable how many bright scientists attended the University of Göttingen.}.
 
@@
 
+\subsection{Exercises}
+
+{
+\small
+
+\begin{enumerate}
+
+	\item Implement the finite element for the Timoshenko beam theory in a 2D plane using Matlab or similar tools.
+Assume a single beam aligned to the X axis, with N nodes each having $y$ displacement and $\theta$ rotation as degrees of freedom (hence no bar compression effects in the horizontal direction and no inclined beams for simplicity).
+Hint: the main task is implementing the procedure to compute the stiffness matrix.
+As a benchmark, simulate a cantilever with 100 finite elements: constrain the left end and apply a vertical force to the right end, and compare the end displacement to the displacement fo the analytical solution.
+Hint: use a single integration point in Gauss quadrature.
+
+	\item Using the finite element above, perform the cantilever benchmark with increasing number of nodes up to n=256, starting from n=2 and each time doubling them.
+Plot the relative error using a logarithmic plot, and see how the result converges to the analytical solution with an almost straight line in the plot.
+Repeat this using two Gauss quadrature points per element, and see how the convergence gets worse and worse when the thickness/length slenderness ratio becomes smaller (the shear lock numerical phenomena are more evident in thin, long beams).
+
+	\item Implement the finite element for a 2D triangle, assuming linear displacements interpolated from the three corner displacements, and use it to study linear elasticity in 2D membranes.
+Hint: use triangle coordinates as auxiliary parametric coordinates, this will simplify shape functions.
+Hint: this will require a single integration point in the center, while for non-homogeneous materials (or if you want to implement quadratic triangles, with more than 3 nodes each) you can use more Gauss points, but the distribution of those points and their weights for the quadrature must be retrieved from special Gauss tables aimed at triangle quadratures.
+
+\end{enumerate}
+
+}
+
 \newpage
@@
 In the following we briefly list the main solver types belonging to these classes.
+
+For additional details on numerical methods for large scale linear systems, we suggest the interested reader to look at these textbooks: \cite{Saad2003,greenbaum1997}.
+Other references for the more specific topic of multigrid methods can be, for instance, \cite{brandt2011multigrid,trottenberg2000multigrid}.
 
@@
 
+\subsection{Exercises}
+
+{
+\small
+
+\begin{enumerate}
+
+	\item Implement the LDLt decomposition using Matlab or similar software tools. Try it with a singular $A$ matrix: what causes the breakup of the method?
+
+	\item Implement the MINRES Krylov method using Matlab or similar software tools. Test it with randomly generated symmetric matrices. Generate a nearly singular matrix and plot the convergence.
+
+	\item Implement the preconditioned CG Krylov method using Matlab or similar software tools.
+Test it on random symmetric matrices using a diagonal preconditioner and later using an ILU preconditioner (you can reuse the ILU code in Matlab to avoid implementing the ILU) and show which is converging faster.
+
+	\item Use a Krylov solver from the Matlab toolkit to solve randomly generated \textit{sparse} linear systems (or better a linear system generated from a FEA problem).
+Add the ILU preconditioner and see how the convergence improves.
+Force the ILU to exploit symmetry if any.
+Plot the convergence of the preconditioned Krylov method for decreasing ILU tolerance to fills-in creation, up to the case where ILU is forced to create no fills-in at all.
+
+\end{enumerate}
+
+}
+
 \newpage
@@
 discuss relevant aspects: the interested reader can find additional details in literature.
+
+Classical textbooks about ODE time integration are \cite{hawa93,hawa96}. Further details on DAE time stepping schemes can be found in \cite{aspe98}.
+The quite advanced topic of DVIs and MDIs can be approached using sources like \cite{pastew03dvi,pfeifferGlocker1996,acary2008numerical}.
 
@@
 \item
-We can distinguish between \textbf{multistep} and \textbf{Runge\footnote{It is remarkable how many bright scientist attended the University of Göttingen.}-Kutta} methods.
+We can distinguish between \textbf{multistep} and \textbf{Runge\footnote{It is remarkable how many bright scientists attended the University of Göttingen.}-Kutta} methods.
 The former depends on a history of previous timesteps in order to extrapolate a new one at the end of the new step, whereas the latter starts from the last timestep information then computes temporary intermediate $\vect{f}_j \left( \vect{x}_j, t\right)$ values after it, in order to compute the value at the end of the step.
@@
 
-The Adams-Bashforth methods of order $s$ are explicit linear multistep methods given by the general formula:
+The Adams-Bashforth methods of order $s$ are explicit methods given by the general linear multistep formula:
 \begin{fleqn}
@@
 \begin{aligned}
-\sum_{j=0}^s a_j \vect{x}_{n+j} &= h\sum_{j=0}^s b_j \vect{f}(t_{n+j},\vect{y}_{n+j}) \\
+\sum_{j=0}^s a_j \vect{x}_{n+j} &= h\sum_{j=0}^s b_j \vect{f}(t_{n+j},\vect{x}_{n+j}) \\
 t_{n+1} &= t_n + h 
@@
 \end{fleqn}
-where one has $a_s=-1$, $a_j=0$ for $j<s$, and $b_j$ coefficients are given by a Lagrange polynomial interpolation hence $b_{s-j-1} = \frac{(-1)^j}{j!(s-j-1)!} \int_0^1 \prod_{i=0 \atop i\ne j}^{s-1} (u+i) \,du, \qquad \text{for } j=0,\ldots,s-1$; for example the 4-th order Adams-Bashforth boils down to
+where one has $a_s=-1$, $a_j=0$ for $j<s$, and $b_j$ coefficients are given by a Lagrange polynomial interpolation hence $b_{s-j-1} = \frac{(-1)^j}{j!(s-j-1)!} \int_0^1 \prod_{i=0 \atop i\ne j}^{s-1} (u+i) \,du, \qquad \text{for } j=0,\ldots,s-1$, since the method is explicit $b_s=0$; for example the 4-th order Adams-Bashforth boils down to
 \begin{fleqn}
@@
 \begin{aligned}
-\vect{x}_{n+4} &= \vect{x}_{n+3} + h\left( \frac{55}{24} \vect{f}(t_{n+3}, y_{n+3}) - \frac{59}{24} \vect{f}(t_{n+2}, y_{n+2}) + \frac{37}{24} \vect{f}(t_{n+1}, y_{n+1}) - \frac{9}{24} \vect{f}(t_n, y_n) \right) \\
+\vect{x}_{n+4} &= \vect{x}_{n+3} + h\left( \frac{55}{24} \vect{f}(t_{n+3}, \vect{x}_{n+3}) - \frac{59}{24} \vect{f}(t_{n+2}, \vect{x}_{n+2}) + \frac{37}{24} \vect{f}(t_{n+1}, \vect{x}_{n+1}) - \frac{9}{24} \vect{f}(t_n, \vect{x}_n) \right) \\
 t_{n+1} &= t_n + h 
@@
 
+\subsection{Exercises}
+
+{
+\small
+
+\begin{enumerate}
+
+	\item Implement the Runge-Kutta 4th order scheme using Matlab or similar softare tools, and use it to simulate the time integral of a horizontal linear spring-dashpot system attached to a mass, where the parameters are: stiffness $k=1000$Nm, damping  $c=30$Nm/s, mass $m=2$kg.
+The initial conditions are: $x=0.2$m, $v=0.5$m/s.
+Test it with increasing time steps $h$ starting from $h=0.002$s and see how precision and stability is affected.
+
+	\item Implement an implicit Euler scheme using Matlab or similar softare tools, and use it to simulate the time integral of a horizontal non-linear spring-dashpot system attached to a mass, where the parameters are: stiffness $k=1000$Nm when $x>0$, $k=100000$Nm when $x\leq0$, damping  $c=30$Nm/s, mass $m=2$kg.
+The initial conditions are: $x=0.2$m, $v=0.5$m/s.
+Test it with increasing time steps $h$ starting from $h=0.002$s and see how precision and stability is affected.
+How does the explicit Euler scheme behaves for the same time steps?
+
+	\item A pendulum of length $L=1$m has a disc connected at the lower end, with mass $M=2$kg and inertia $J=0.1$kgm$^2$.
+The gravitational field is $g=-9.8$ms$^{-2}$ Write the equations of motion as an ODE, with a single generalized coordinate $\theta$ expressing the rotation, and simulate it with an explicit Euler scheme using Matlab or similar tools, from initial conditions $\theta=0.1$rad, $\dot{\theta}=0$rad/s.
+
+	\item For the pendulum of the previous exercise, write the equations of motion as an index-2 DAE, using body coordinates $x,y,\theta$ and unknown reactions in the revolute joint, $\lambda_x, \lambda_y$.
+How to compute consistent initial values $x_0,y_0,\theta_0, \lambda_{x0},\lambda_{y0}$ from the initial conditions $\theta=0.1$rad, $\dot{\theta}=0$rad/s?
+
+	\item For the DAE of the previous exercise, implement a solver using the \textit{implicit Euler for DAE} method presented in this paper, and simulate it.
+Compare the results to the same pendulum expressed as an ODE and simulated with an explicit Euler scheme.
+How are the computational times of the DAE solver compared to the ODE solver in this case?
+
+\end{enumerate}
+
+}
+
+\printbibliography
+
 %%%%%%%%%Bibliography%%%%%%%%
-\bibliographystyle{plain}
+%\bibliographystyle{plain}
 

### 2020-08-06 2021/NODYCON_nonsmooth_ADMM/Nodycon2021_LateX_Template.tex
Add NODYCON 2021 abstract for the ADMM paper

\documentclass[11pt]{nodycon2021}
\usepackage{epsfig,amsmath,amsfonts}

%%
\let\OLDthebibliography\thebibliography
\renewcommand\thebibliography[1]{
  \OLDthebibliography{#1}
  \setlength{\parskip}{0pt}
  \setlength{\itemsep}{0pt plus 0.3ex}
}
\usepackage[official]{eurosym}
%%%

\title{Template for NODYCON 2021 One-page abstract}

\author{\textbf{First Author}$^\ast$, Second Author$^\ast$ and Third Author $^{\ast\ast}$}

\address{$^\ast$Department of Mechanical and Aerospace Engineering, Florida Institute of Technology, Melbourne, FL, USA\\	 $^{\ast\ast}$Institution, City, State, Country}

\abstract{The following guidelines have been prepared for authors of papers to be presented at the \textit{Second International Nonlinear Dynamics Conference NODYCON 2021} originally planned to be held at Sapienza University of Rome, in Rome (Italy), February 16-19, 2021, and now hosted as a virtual online conference. These guidelines provide the rules for the preparation of abstracts (maximum 1 page). Authors are invited to follow these guidelines for inclusion of their abstracts in the digital Book of Abstracts.}
\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

The full scope of the conference which reflects the rich spectrum of topics covered by NONLINEAR DYNAMICS can be found in the Call for Papers available on the conference website https://nodycon.org/. The one-page abstracts to be submitted online using the Conference portal    will be included in the Conference Book of Abstracts. The deadline for submitting the one-page Abstract is Friday, July 10, 2020.  Authors will have the option of submitting full-length papers for publication consideration as Springer Conference Proceedings. The papers will be made available to the following indexing services: Conference Proceedings Citation Index (CPCI), part of Clarivate Analytics’ Web of Science, EI Engineering Index (Compendex and Inspec databases), Google Scholar, MathSciNet, Scopus, Zentralblatt MATH. A short list of papers will be invited to a Special Issue of NONLINEAR DYNAMICS. To accelerate the reviewing process, referees will be invited from those who will attend the conference. The Springer Conference Proceedings papers will be possibly made available during NODYCON 2021. The website will be open for full-length papers submission until September 11, 2020.  The papers must be written in English and should present material that is novel and  unpublished at the time of the Conference. Prospective presenters are asked to submit their abstracts through the Conference website. The electronic submission of the one-page Abstract is required in PDF format only. The template for the manuscript preparation according to the conference guidelines can be downloaded both in Word Office and LaTeX formats. The layout of the one-page Abstract has to follow the style of this document, starting with the title, followed by the author(s) names and their affiliation(s). \textbf{The name of the author presenting the paper should be highlighted in boldface}.

\begin{figure}[h!]
\centering
\epsfig{file=BannerNodycon2021.eps,scale=0.75}
  \caption{Welcome to the Second International Nonlinear Dynamics Conference NODYCON 2021.}
\label{fig:Fig1}
\end{figure}

\section{Electronic submission of the one-page Abstract (PDF format)}
The one-page  Abstract should be organized in non-numbered sections  by adopting the format used in these guidelines. In particular, the headings have to be centred in bold lower-case letters (12 pt). The text should be single-spaced using 11 pt font. The typing area of the A4 page should be 170 x 267 mm, leaving 20 mm left and right margins, and 15 mm top and bottom margins. The total length of the  Abstract, including one figure and the references, should be one page. The text should be written using Times New Roman font. The name of the author presenting the paper should be highlighted in boldface. The title should appear 15 mm below the top edge of the page. The manuscript has to be organized as follows: i) a brief, descriptive \textit{Abstract} of approximately 150 words; ii) a short \textit{Introduction} to the problem formulation containing also the motivation and the purpose of the study together with a brief summary of previous works on the topic; iii) \textit{Results and discussion} containing the core  findings and highlights  of the work. Finally, the \textit{References} list should be collocated at the end of the page.

\begin{thebibliography}{99}

\bibitem{Cooler} Cooler A. S. (1999) Binary Flow Systems.  {\em J. Fluid Mech}  {\bf 999}:999-996.

\bibitem{Icer}	Icer D.F., Adams J.A. (1977) Mathematical Elements for Computer Simulation. McGraw Hill, NY.

\bibitem{NODYCON}	NODYCON 2021, website \textbf{https://nodycon.org/}.

\end{thebibliography}
 \end{document}

### 2020-08-06 2021/NODYCON_nonsmooth_ADMM/abstract_Nodycon_Tasora.tex
Add NODYCON 2021 abstract for the ADMM paper

\documentclass[11pt]{nodycon2021}
\usepackage{epsfig,amsmath,amsfonts}

%%
\let\OLDthebibliography\thebibliography
\renewcommand\thebibliography[1]{
  \OLDthebibliography{#1}
  \setlength{\parskip}{0pt}
  \setlength{\itemsep}{0pt plus 0.3ex}
}
\usepackage[official]{eurosym}
%%%

\title{Solving Non-smooth Dynamic Problems using \\ the 
Alternating Direction Method of Multipliers }

\author{\textbf{Alessandro Tasora}$^\ast$, Dario Mangoni$^\ast$ and Simone Benatti$^{\ast\ast}$}

\address{$^\ast$Department of Engineering and Architecture, University of Parma, ITALY}

\abstract{We propose to use the Alternating Direction Method of Multipliers (ADMM) for solving the variational inequality that arises in non-smooth contact problems in multibody simulations. The ADMM method is based on simple computational primitives, it scales well with the problem size, it can handle problems that mix finite elements and rigid bodies, it offers superior convergence properties.}

\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

Non-smooth dynamical problems, such as those involving contact and friction, can be solved by means of special time stepping methods that offer superior robustness and stability at the cost of solving a variational inequality per each time step.
% \cite{acary2008numerical}. 
Under mild assumptions such variational inequality can be stated as a convex cone complementarity problem, that can be also cast as an optimization problem, namely a quadratic program with convex constraints \cite{anitescuTasora2008}. During the last years, different numerical methods have been proposed to solve such problem: fixed point iterations such as projected Jacobi, first-order spectral methods such as SPG, Interior Point (IP) methods and others. Recently there has been a revival of Alternating Direction Method of Multipliers (ADMM) and similar operator-splitting methods for solving this class of problems \cite{Goldstein2014}. %  [2014 Alternating Direction Optimization Methods]. 
Even if the theoretical convergence of such methods is worse than the one of IP methods, recent developments showed that in practical scenarios they offer superior speed, scalability and robustness 
\cite{Cannon2019}. % [COSMO  conic operator splitting method for large convex problems].
In 
\cite{Stellato2020} % [OSQP: an operator splitting solver for quadratic programs].
a specialized ADMM method for quadratic problems with conic constraints has been presented: on the top of this method we develop a solver that can exploit the nature of non-smooth dynamical problems. In our embodiment, unknowns are primal variables (velocity measures) and dual variables (impulses in contact points and joints), cone constraints stem from the Coulomb friction law, and the system matrix includes terms from the mass matrices and from the tangent stiffness/damping matrices (hence it is sparse, but not necessarily diagonal).

\begin{figure}[h!]
\centering
\includegraphics[width=0.50\textwidth]{t8_convergence.pdf}
\includegraphics[width=0.35\textwidth, trim=0cm -3cm 0 3cm]{t8_snapshot.png}
\caption{Convergence of the ADMMM method within one time step of the wrecking ball benchmark (600 bricks in four walls): residual in frictional constraint   violation compared to fixed point Jacobi iterations and to first-order SPG methods.}
\end{figure}

\section{Results and discussion}

Our ADMM method requires few computational primitives: basically a projection of dual variables on conic sets, a backward solve of a linear system, and a forward solve. The latter is a computational bottleneck, but it can be performed only once per run, as the matrix does not change often during the iterations. 
A good estimation of the ADMM step size proved to be fundamental in achieving good convergence: using some heuristics we obtained an efficient auto-tuning algorithm. 
We noted that ADMM can be successfully applied to problems that exhibit temporal coherence because, unlike IP methods, it supports warm-starting.

\bibliographystyle{unsrt}
\bibliography{../../bibliography/refsMBS,../../bibliography/refsOPT}

 \end{document}

### 2020-08-25 2020/CoRL_2020/main.tex
Added BookIta template, Eccomas DRL paper, GVSETS AV paper, CoRL (under review)

\documentclass{article}

\usepackage{corl_2020} % Use this for the initial submission.
%\usepackage[final]{corl_2020} % Uncomment for the camera-ready ``final'' version.

\usepackage{graphicx}
\usepackage{url}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{mathtools}
\usepackage{caption}
\usepackage{subcaption}
%\usepackage{xcolor}
\usepackage{wrapfig}
\usepackage{float}

\title{End-to-end learned autonomous off-road navigation using a physics-based simulation framework}

% Reinforcement learning using end-to-end networks for off-road autonomous navigation

% Reinforcement learned for off-road autonomous navigation using physics-based simulation

% The \author macro works with any number of authors. There are two
% commands used to separate the names and addresses of multiple
% authors: \And and \AND.
%
% Using \And between authors leaves it to LaTeX to determine where to
% break the lines. Using \AND forces a line break at that point. So,
% if LaTeX puts 3 of 4 authors names on the first line, and the last
% on the second line, try using \AND instead of \And before the third
% author name.

% NOTE: authors will be visible only in the camera-ready (ie, when using the option 'final'). 
% 	For the initial submission the authors will be anonymized.

\author{
  Asher Elmquist \\
  Department of Mechanical Engineering \\
  University of Wisconsin-Madison \\
  Madison WI, USA \\
  amelmquist@wisc.edu \\
  \And
  Simone Benatti \\
  Dipartimento di Ingegneria ed Architettura \\
  Università di Parma \\
  Parma, Italy \\
  simone.benatti@studenti.unipr.it \\
  \And
  Aaron Young \\
  Department of Mechanical Engineering \\
  University of Wisconsin-Madison \\
  Madison WI, USA \\
  aryoung5@wisc.edu \\
  \And
  Alessandro Tasora \\
  Dipartimento di Ingegneria ed Architettura \\
  Università di Parma \\
  Parma, Italy \\
  alessandro.tasora@unipr.it \\
  \And
  Radu Serban \\
  Department of Mechanical Engineering \\
  University of Wisconsin-Madison \\
  Madison WI, USA \\
  serban@wisc.edu \\
  \And
  Dan Negrut  \\
  Department of Mechanical Engineering \\
  University of Wisconsin-Madison \\
  Madison WI, USA \\
  negrut@wisc.edu \\
}

\include{macros}

\begin{document}
\maketitle

%===============================================================================

\begin{abstract}
This contribution discusses an end-to-end reinforcement learning approach for off-road autonomous vehicle (AV) navigation. Both learning and testing are done in simulation. The contribution highlights: ($i$) the strategy employed for end-to-end learning; and ($ii$) the simulation infrastructure used to learn and test the control policy. For ($i$), we use sensor data fusion (camera, GPS, and IMU). For ($ii$), we developed a physics-based, open source integrated simulation environment that supports vehicle dynamics, deformable/undeformable terrains, and virtual sensing. We use a Gator off-road vehicle to demonstrate how a policy learned on flat undeformable terrain performs when used in hilly conditions while navigating around a course of randomly placed obstacles. The hilly terrain covers a swath of 80$ \times $80 meters and the soil can everywhere be either undeformable, or deformable hard (silt-like), or deformable soft (snow-like). The results reported herein can be reproduced with models and data available in a public repository \cite{CoRLsupportData2020}. Movies associated with the tests run are available online  \cite{CoRLpaperVideos2020}.
%The policy is learned on rigid-terrain, with uneven terrain before being deployed in an environment with deformable terrain, and additional terrain obstacles to test and understand robustness. This study flexes an open-source software platform that supports robot and autonomous vehicle simulation by providing support for vehicle dynamics, deformable terrain, and virtual sensors. The learned policy is tested for robustness in the environment to understand the policy's ability to successfully and consistently navigate a complex off-road environment and test the sensitivity of the policy to the training environment.

\end{abstract}

% Two or three meaningful keywords should be added here
\keywords{Simulation, Reinforcement Learning, Off-road Autonomous Vehicles, Deformable Terrain}

%===============================================================================

\input{sections/introduction}
\input{sections/related}
\input{sections/environment}
\input{sections/learning}
\input{sections/demonstration}
\input{sections/results}
\input{sections/conclusion}

%===============================================================================

% The maximum paper length is 8 pages excluding references and acknowledgements, and 10 pages including references and acknowledgements

\clearpage
% The acknowledgments are automatically included only in the final version of the paper.
\acknowledgments{This work was carried out in part with support from the Safety Research using Simulation (SAFER-SIM) program. SAFER-SIM is funded by a grant from the U.S. Department of Transportation’s University Transportation Centers Program (69A3551747131). The U.S. Government assumes no liability for the ideas expressed in this document or the use thereof.}

%===============================================================================

\bibliography{refs/newREFS,refs/refsSensors,refs/refsAutonomousVehicles,refs/refsChronoSpecific,refs/refsSBELspecific,refs/refsMBS,refs/refsTerramech,refs/refsRobotics,refs/refsStatsML}

\end{document}

### 2020-08-25 2020/CoRL_2020/sections/alternatives.tex
Added BookIta template, Eccomas DRL paper, GVSETS AV paper, CoRL (under review)

%----------------------------------
The big players in the ``V'' field: MSC.ADAMS of MSC.Software \cite{ADAMSsoftware,adamsManual}, SimPack of Dassault Syst\'emes \cite{SimpackSoftware}, RecurDyn of FunctionBay \cite{functionBay}, and Siemens' Virtual.Lab \cite{lmsVirtualLab}, do not have any product that covers the A or C axes. Limited support for virtual worlds along the ``E'' exists, yet it is aligned with manufacturing applications for automation purposes. These softwares are expensive and not open source.

Simulation tools are available for virtual testing and validation of Advanced Driver-Assistance Systems (ADAS).  Most of them span the A and E axes well. Connectivity is seldom considered, and the support along the V-axis relies on rudimentary vehicle models or third party solutions.  With the notable exception of Gazebo~\cite{gazebo}, these are mostly commercial, closed-source solutions. As such, they cannot be easily extended to test new vehicle models, tinker with virtual sensing, or augment traffic scenarios to include, for instance, heterogeneous agents (pedestrians and riders).

%----------------------------------
Gazebo \cite{gazebo,gazebo-VRC} (collaboration letter included) is an Apache 2.0 license, open-source ``3D dynamic simulator with the ability to accurately and efficiently simulate populations of robots in complex indoor and outdoor environments. While similar to game engines, Gazebo offers physics simulation at a much higher degree of fidelity, a suite of sensors, and interfaces for both users and programs'' \cite{gazebo, gazebo-VRC}. While Gazebo has been successfully used for DARPA's Virtual Robotics Challenge~\cite{gazebo-VRC}, its V-axis relies on physics engines not suited, or at least not equipped, to model ground vehicles at the level of sophistication and fidelity required by a CAVE. Lastly, for reasons discussed in section \S\ref{ss:preamble-A-axis}, its sensing prowess is lacking.

%----------------------------------
The Virtual Autonomous Navigation Environment (VANE), developed by the US Army Engineer Research and Development Center is a ``high-fidelity, physics-based software that simulates terrain and environmental conditions.'' It is intended to support the design, development, and testing of autonomy algorithms for intelligent ground vehicles \cite{VANE-2012}. It consists of a set of independent models for sensors, vehicle dynamics, and terramechanics. The VANE uses high-fidelity environment geometry and ray-tracing to simulate GPS, LiDAR, and camera sensors. Inertial sensors are simulated using the dynamics engine. Closed-source and unavailable to outside researchers, VANE also relies on a physics engine, the Open Dynamics Engine (ODE)~\cite{ODE-website, Drumwright-2010} originating in the video gaming industry. Current VANE development efforts, unrelated to this group, aim at interfacing to {\CHRONO}::Vehicle.

%----------------------------------
The Autonomous Navigation and Virtual Environment Laboratory (ANVEL) is a real time simulation software desktop application developed by Quantum Signals that ``enables the analysis and testing of autonomous vehicle dynamics, terrain/obstacle interaction, sensors, algorithms, and operation in a virtual world'' ~\cite{ANVEL-website}. ANVEL serves as a real-time, desktop interface for the VANE. As of June 2013, ANVEL is available to qualified academic institutions. Like VANE, ANVEL relies on the game physics engine ODE.

%----------------------------------
PreScan~\cite{PreScan-website}, developed by TASS International, is a simulation platform for advanced driver assistance systems and uses a MATLAB/Simulink interface. It can be used for model-based controller design, software-in-the-loop, and hardware-in-the-loop (HIL) systems. PreScan can operate in open-loop and closed-loop, offline and online modes. It is an open software platform which has flexible interfaces to link to third party vehicle dynamics commercial solutions such as CarSIM and dSPACE ASM, and third party HIL simulators/hardware, e.g., ETAS, dSPACE, and Vector. It has no support along the C and A axes.

%----------------------------------
Pro-SiVIC~\cite{Hiblot-2010, Belbachir-2015}, developed by the French Institute of Science, Transport Technology and Network (IFSTTAR) and commercialized by CIVITEC (a spin-off of IFSTTAR), provides very good support along the A and E axes, with no support for the C and V axes. It is widely used for Advanced Driver Assistance Systems (ADAS), as it provides a family of sensors in a virtual environment.

%----------------------------------
Finally, there are gaming engines, e.g., \cite{unityGaming,unrealEngine}, which may or may not be open source. Using the nomenclature adopted in this document, they support avatar vehicles. There is no support for sensing or connectivity insofar simulating DSRC is concerned. Some game engines provide support along the E-axis, yet there is little concern with the interplay between sensing and the environment. Along the V-axis, the underlying dynamics is of little concern as long as the motion looks realistic.

### 2020-08-25 2020/CoRL_2020/sections/conclusion.tex
Added BookIta template, Eccomas DRL paper, GVSETS AV paper, CoRL (under review)

%!TEX root = ../root.tex

\section{Conclusion and Future Work}
\label{sec:conclusions}
This contribution has two main themes. The first pertains a simulation platform, called Chrono, that supports AV off-road mobility studies. The second focuses on end-to-end learning as enabled by the Chrono environment, which anchors both the learning and testing phases. The end-to-end policy is used in flat and hilly landscapes, with deformable terrain that can be hard (silt-like) or soft (snow-like). We noted the following: policies learned on flat terrain are insufficient for navigating hilly scenarios; policies learned on rigid terrain transfer quite well to deformable terrain when the terrain is flat; the hillier the landscape, the harder it is to navigate it (Fig.\ref{fig:successRates}); the more obstacles are randomly placed on the course, the less likely it is for the policy to see the vehicle through; the control policy led to trip trajectories that came rather close to those associated with PSO, a third party trajectory planning tool (Fig.\ref{fig:pso_length_compare}); there is a sizable number of head-on collisions that point to room for improvement in the derived policy (Fig.\ref{fig:rigid_failure_metric_histogram}). Note that the testing conditions, in terms of average number of obstacles per unit area, were more harsh than the learning conditions.

Looking ahead, we plan to pursue several research thrusts and simulation platform development avenues. One direction is to investigate more complex control stacks that would combine the end-to-end with more traditional strategies such as model predictive control. The current work should be expanded to understand how tracked vehicles perform under similar conditions given that their traction and turning radius differ significantly from their wheel counterparts. Work is underway for faster SCM simulation that will allow straight training on deformable soil. Not analyzed in this contribution is the steering control input to the vehicle, which can often be noisy based on the output of the NN. Finally, we plan to investigate approaches that enhance the chance of simulation-derived policies transferring effectively to the real world.

%I reworked this section since it was a single sentence. I also dropped the reference to the noisy steering data since we never analyzed it. I think its good to mention it (I don't want to sweep anything under the rug), but I don't think we should put the results out there unless we can do it justice by analyzing and understanding it. - Asher

%more complex AV control stacks, that in addition to end-to-end policies bring into discussion model predictive control strategies;

%the case of tracked vehicles, which are quite different than wheeled vehicles in that they can turn in place at zero translational velocity;

%a faster SCM simulation strategy, which will allow training on deformable terrains;

%how we can smooth out the steering wheel input produced by the policy -- it is far from what a human would do in that is aggressive in relation to left/right swings (plots not shown but available in supplemental material \cite{CoRLsupportData2020});

%and, approaches that improve the chances of simulation-derived policies to transfer to the real world.

### 2020-08-25 2020/CoRL_2020/sections/demonstration.tex
Added BookIta template, Eccomas DRL paper, GVSETS AV paper, CoRL (under review)

%!TEX root = ../main.tex
\section{Simulation Experiments}
\label{sec:demoTechnology}
To demonstrate and analyze the capabilities of the control policy, we used an 80$ \times $80 m patch of terrain. The reduced scale of the test environment allowed for rapid evaluation and inclusion of tests using deformable terrain based on the Soil Contact Model (SCM) \cite{ChronoSCM2019}. For all tests, the vehicle is placed at world location (-35,35); to be successful, it must navigate safely to within a 10 m radius of world location (35,-35). This setup was a matter of convenience and is not a limitation of the policy. Since training used a larger patch of 120$ \times $120 m, it is important to note that parameters such as number of obstacles and height of terrain cannot be compared directly with testing; this was done on purpose. For the terrain patch, and equivalent number of obstacles to the 50 used in training is approximately 22. Additionally, the maximum height difference of 10 m in training is equivalent to a height of approximately 7.25 m on the testing terrain. Furthermore, in testing, to prevent the vehicle from successfully navigating purely along the diagonal straight toward the destination, five obstacles were placed randomly near the diagonal to force non-trivial trajectories. Overall, the tests were conducted with a higher-complexity environment that used in training. They were designed to probe the robustness of the algorithm and understand to what extent the policy could be used on never-before-seen terrain. We looked at ($i$) increased levels of hilliness; ($ii$) increased number of obstacles; and ($iii$) alterations of the soil including rigid, hard-deformable (silt-like), and soft-deformable (snow-like).

Figure \ref{fig:scm_soft_hilly_deformed} shows a snapshot of one scenario: the soft deformable terrain used for testing robustness; a capture of the image from the camera sensor used by the NN; and the position of the vehicle along its current trajectory with the obstacles and height map overlaid for context (height map: black is valleys, white is peaks).

\begin{figure}[h]
    \centering
    \includegraphics[height=.1095\paperheight]{images/demonstration/scm_soft_hilly_montage_final.png}
    \caption{Snapshot of a scenario used for testing. Left: third-person perspective of the vehicle; center: image from the camera's view at the resolution used in training 80$ \times $45 pixels; right: the current progress of the vehicle amount the hills and obstacles corresponding to the images on left. }
    \label{fig:scm_soft_hilly_deformed}
\end{figure}

Figure \ref{fig:example_paths} shows an example set of paths navigated by the Gator in the test environment. In Fig. \ref{fig:rigid_flat_3_example12}, the vehicle avoids a sparse environment, limited to 20 obstacles. By increasing the obstacles to 50, as shown in Fig. \ref{fig:rigid_flat_9_example22}, the complexity of the test environment forces the policy into a correspondingly complex path. An example failure is illustrated in Fig. \ref{fig:rigid_flat_9_failure_example0} where the vehicle does not reach the destination owing to a collision. Any scenario where the vehicle collides with an obstacle is deemed a failure, regardless of how directly the vehicle collides with the obstacle. The last example, shown in Fig. \ref{fig:rigid_height_map_example13}, demonstrates the vehicle navigating a hilly terrain, created from a programmatically generated random height map.

\begin{figure}[h]
    \captionsetup{justification=centering}
    \centering
    \begin{subfigure}{0.24\textwidth}
        \includegraphics[height=.12\paperheight]{images/demonstration/rigid_flat_3_example12.png}
        \caption{}
        \label{fig:rigid_flat_3_example12}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.24\textwidth}
        \includegraphics[height=.12\paperheight]{images/demonstration/rigid_flat_9_example22.png}
        \caption{}
        \label{fig:rigid_flat_9_example22}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.24\textwidth}
        \includegraphics[height=.12\paperheight]{images/demonstration/rigid_flat_9_failure_example0.png}
        \caption{}
        \label{fig:rigid_flat_9_failure_example0}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.24\textwidth}
        \includegraphics[height=.12\paperheight]{images/demonstration/rigid_height_map_example13.png}
        \caption{}
        \label{fig:rigid_height_map_example13}
    \end{subfigure}
    \caption{Example test scenarios: (a) 20 obstacles, (b) 50 obstacles, (c) failure with 50 obstacles, (d) hilly terrain with 50 obstacles.}
    \label{fig:example_paths}
\end{figure}

In an attempt to assess the practicality of the vehicle's chosen path, a Particle Swarm Optimization (PSO) algorithm \cite{PSOObstacleAvoidance} with global knowledge of the environment was used to generate reference trajectories in the environment; they were used in post-processing only, for comparison purposes. Figure \ref{fig:rigid_flat_pso_example3} shows a comparison between the trained vehicle's path and a trajectory generated using the PSO path planner. In this example, 40 obstacles are present on rigid, flat terrain. %What would have been undesirable to see was that the trained vehicle skirted the edges of the domain or assumed unreasonable trajectories. This turned out to not be the case.

\begin{figure}[h]
    \captionsetup{justification=centering}
    \centering
    \begin{subfigure}{0.329\textwidth}
        \centering
        \includegraphics[height=.155\paperheight]{images/demonstration/rigid_flat_pso_example3.png}
        \caption{Example comparison between the Gator's path and PSO path.}
        \label{fig:rigid_flat_pso_example3}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.329\textwidth}
        \centering
        \includegraphics[height=.155\paperheight]{images/demonstration/rigid_40_failure_100.png}
        \caption{Example collision where the directness of the impacts was 100\%.}
        \label{fig:rigid_40_failure_100}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.329\textwidth}
        \centering
        \includegraphics[height=.155\paperheight]{images/demonstration/rigid_40_failure_34.png}
        \caption{Example collision where the directness of the impacts was 34\%.}
        \label{fig:rigid_40_failure_34}
    \end{subfigure}
    \caption{Example paths showing comparison with global path planner and two levels of collision directness.}
\end{figure}

### 2020-08-25 2020/CoRL_2020/sections/environment.tex
Added BookIta template, Eccomas DRL paper, GVSETS AV paper, CoRL (under review)

%!TEX root = ../main.tex

\section{Chrono Simulation Environment}
\label{sec:simEnv}
The physics-based simulator used in conjunction with this work is called Chrono. It is actively developed, it is open source, and it is released under a permissive BSD3 license for unfettered use/change/distribution \cite{projectChronoGithub}. A full description of the simulation platform falls outside the scope of this document; for an overview, see \cite{chronoOverview2016}. Chrono provides support for multibody dynamics (multi-core), nonlinear finite element analysis (multi-core), fluid solid interaction (GPU), granular dynamics (GPU), terramechanics (multi-core/GPU/MPI), sensing (GPU), and the simulation of large collections of AVs running in one joint scenario (MPI). The hardware support is as follows: multi-core CPUs, GPU computing via CUDA, and distributed memory (clusters/supercomputers) via the Message Passing Interface (MPI) standard. The four Chrono components relevant herein are: Chrono::Engine, Chrono::Vehicle, terramechanics, and Chrono::Sensor. Chrono::Engine is the solver that advances the simulation in time. Chrono::Vehicle provides support for rapidly setting up and analyzing vehicles (tracked or wheeled) via a library of templates for vehicle components (suspensions, steering systems, tires, powertrains/diverlines, tracks, etc.) \cite{ChronoVehicle2019}. The terramechanics support comes in three flavors: expeditious and semi-empirical approaches \cite{alessandroSCM2019}, via continuum representations, and using the discrete element method (fully resolved granular terrain).

Sensing support in Chrono is provided as an additional module that sits on top of Chrono::Engine to provide measurement data from inside the simulation and virtual environment. Currently, there is support for RGB cameras, lidar, GPS, and IMU \cite{TR-2020-07}. The sensor module's purpose is to provide realistic data for training and testing autonomous controls. For GPS and IMU, ground truth data, queried from Chrono::Engine is augmented to introduce noise commonly found on accelerometers and gyroscopes \cite{shah2018airsim} as well as GPS receivers. For camera and lidar, the visual environment is ray traced using custom GPU kernels that model the acquisition process of the specific sensor. The ray-traced data is then augmented to introduce noise and distortion to model the true sensor output. All sensors are parameterized by their update frequencies, noise characteristics, and lag.

Our lidar model augments ground-truth data with noise (based on the measurements of range, intensity, and angular precision) to produce the final point cloud. The lidar leverages ray-tracing to create a point cloud based on the visual scene. This, in combination with supersampling for beam divergence, allows Chrono::Sensor to generate high-fidelity point clouds of complex environments. The beam discretization model extends that proposed in \cite{goodin2018enabling} to allow a user-defined number of rays per lidar beam. By incorporating beam divergence, we can model multiple return modes and encountered objects. In addition to beam divergence, the ray-tracing method allows the temporal sampling of a scanning lidar to be based on modern motion blurring techniques resulting in realistic and continuous distortions that are not possible with large time steps in video gaming collision detection systems employed by other learning environments, e.g. \cite{shah2018airsim,carlaAVsim2017}.

The implemented camera simulator introduces lens and image sensors models to improve the realism of the data. The camera is parameterized based on the frequency, resolution, field of view, exposure time, and lag. Based on the exposure time, motion blur that accounts for object and camera movement is introduced. The camera lens model draws on work from \cite{tang2017precision} to allow for wide angle lenses. The noise model is based on modified version of the EMVA standard \cite{standard20101288}, which introduces intensity-dependent noise based on the image sensor characteristics. Additional components of the image signal processor (ISP) are in development since the ISP introduces additional sensing artifacts such as compression, demosaicking, and color correction.

The Python API of Chrono, known as \emph{PyChrono}, provides access to the vast majority of Chrono API from Python, including Chrono::Vehicle and Chrono::Sensor. This allows for a simulation to be directly interfaced to the Python API of popular ML frameworks. By using the SWIG wrapper \cite{Beazley96SWIG} to directly interface with the C++ binaries, minimal overhead is introduced when running a simulation from Python. As an example, large data from sensor simulations (such as RGB images or lidar) are cast to NumPy arrays without instantiating new memory by means of SWIG typemaps.

### 2020-08-25 2020/CoRL_2020/sections/introduction.tex
Added BookIta template, Eccomas DRL paper, GVSETS AV paper, CoRL (under review)

%!TEX root = ../main.tex

\section{Introduction}

We use physics-based simulation to produce end-to-end policies for controlling AV navigation directly from raw sensory data. The AVs operate in hilly, off-road conditions with randomly placed obstacles (rocks) obstructing safe navigation. Training is based on a curriculum learning approach; the complexity of the environment is increased as the policy converges. The training is exclusively done with undeformable terrain since these simulations run faster than real time \cite{negrutGVSETS2020}. In the current implementation, deformable terrain simulation is approximately 100$ \times $ slower than the undeformable counterpart. As such, learning on deformable terrain is expensive. The control policy derived is tested on deformable soils that have different textures and soil deformation attributes. The deformable soils are of two categories: deformable but hard (silt-like), and deformable but soft (snow-like). The end-to-end approach to navigation is certainly not new, see, for instance \cite{end2endNVIDIA2016,AminiRKR19}. However, to the best of our knowledge, ($a$) this is the first example of off-road navigation (driving control + reaching a goal) with reinforcement learning; and, ($b$) the simulation environment developed is the first \textit{open-source}, physics-based platform that brings together tracked/wheeled vehicle dynamics, sensor, and terrain simulation. %While the authors acknowledge that it is commonly known that simulated sensor data, specifically rendered camera images, do not directly translate well to reality, other contributions have focused on methods for bridging this gap, and this is not the focus of this paper.

Our goal is to demonstrate that off-road mobility of AVs can rely on simulation for the development of control policies; and, to report on tests that assess these policies' effectiveness. We do not comment on the transferability of the derived control policies \cite{sim2realGapEssex1995}. Rather than addressing this important issue, for which we actually do not have an answer yet, this contribution focuses on the learning and the platform that enabled it, including probing the robustness of the policy in simulation. Thus, in \S\ref{sec:relatedWork} we provide an overview of similar ongoing efforts in the simulation-in-robotics area. Section \S\ref{sec:simEnv} provides an outline of the simulation environment developed by this group. Section \S\ref{sec:end2endLearning} presents the end-to-end learning approached for off-road AV mobility. Simulation results are described in \S\ref{sec:demoTechnology}. The last section contains concluding remarks and directions of future work.

### 2020-08-25 2020/CoRL_2020/sections/learning.tex
Added BookIta template, Eccomas DRL paper, GVSETS AV paper, CoRL (under review)

%!TEX root = ../main.tex

\section{End-to-end learning approach}
\label{sec:end2endLearning}

The control policy employed in this work is end-to-end: the NN takes as inputs the raw sensor data and directly outputs the control values (steering and throttle in this case). In addition, the policy is trained from scratch. The objective for the navigation algorithm is to control a John Deere Gator to reach a target destination given by GPS coordinates. The algorithm uses a GPS and IMU that provide the neural network with the current vehicle location and orientation. The vehicle is also equipped with a low-resolution RGB camera, which the network includes as input in order to allow for avoiding obstacles (rocks of various shapes, sizes, and textures).

For training, the environment consists of a 120x120 m swath of terrain on which obstacles are placed randomly. The vehicle initial position is picked randomly on a 80 m diameter circle while the goal is placed on the opposite side of the same circle; in other words, given $\alpha$ the angle of the vehicle initial position, the angle of the goal will be $\alpha + \beta$ with $\beta \in [\frac{\pi}{2} , \frac{3\pi}{2}]$.
The vehicle must navigate to within 10 m of the destination and the reward is proportional to the vehicle's approaching speed. The episode is terminated with a reward penalty if the vehicle hits an obstacle, goes outside the terrain boundary, or the timeout is reached, while it is terminated with a reward bonus if it reaches the goal.
The observation consists of a two-element tuple: an 80x45 pixel RGB image and a five-element array containing the components (x,y) of the distance from the goal (in the vehicle frame, based on GPS measurements), the vehicle orientation (compass angle), the heading with respect to the goal, and the vehicle speed. 
Based on these inputs, the policy controls the steering (-1 to 1) and the throttle/brake value (-1 to 1, with negatives implying braking). 
To train the NN, we adopted the constrained version of the Proximal Policy Optimization (PPO) algorithm \cite{Schulman2017PPO}, using two separated NN for the actor and the critic. PPO is known as one of the best performing DRL algorithms for continuous control \cite{openai2018HandManipulation}.
The NN model inputs come from several sensors: GPS, IMU, and RGB camera. Through PyTorch \cite{paszke2017PyTorch} the NN model was implemented as follows. A five-element array is fed to a fully connected layer, while the RGB image is processed by a CNN as in \cite{Mnih13}. The output of the CNN and the single fully connected hidden layer are concatenated and then processed through three fully-connected hidden layers, as shown in Fig. \ref{fig:NN_arch}. All layers use the ReLU activation function.

\begin{figure}[h]
    \centering
    \includegraphics[width=.95\linewidth]{images/CORL_paper_arch.pdf}
    \caption{Actor neural network architecture}
    \label{fig:NN_arch}
\end{figure}

%local vs global (from GPS)

Given the GPS coordinates of the vehicle and the goal, along with the orientation of the vehicle from the IMU, it is straightforward (for a human) to evaluate the distance between the vehicle and the objective. This being said, there are several ways to pass this information to the NN as an input: directly feeding the GPS coordinates, the coordinate difference, the relative distance in a frame oriented along the cardinal points or the distance in the local frame of the vehicle. The last is obviously the most meaningful information but we might want to use coarser inputs and leave to the optimization to infer the correlation. This approach, however, would be inefficient as shown in Fig. \ref{fig:plotLocGlob}, where is clearly shown the benefit, in terms of convergence, of simply rotating the position of the goal with respect to the vehicle in the vehicle reference frame (called \emph{local}).

\begin{wrapfigure}{r}{0.50\textwidth}
    \vspace{-25pt}
    \begin{center}
        \includegraphics[width=.90\linewidth]{images/rewards_localglobal.pdf}
        \caption{On flat terrain without obstacles, the algorithm converges immediately when feeding the position of the goal rotated in the vehicle reference frame. When the position is given in the global frame, training does not converge.}
        \label{fig:plotLocGlob}
    \end{center}
    \vspace{-15pt}
\end{wrapfigure}We adopted a curriculum learning approach \cite{BengioCurriculumLearning09}, progressively increasing the complexity of the task, as shown in Fig. \ref{fig:rew_progr}. The first part of the training was performed on flat terrain with a random number of obstacles (from 0 to 30). After $\sim$ 200 policy updates, when the convergence is reached, the obstacle number is fixed at 30, and after a drop off the convergence is reached again quickly. When, after 376 policy updates, the flat terrain is replaced with hilly terrain (while keeping the same number of obstacles), the agent struggles and many updates are required in order to converge again. In the third and last stage of the training, the obstacle count was increased to 50 and the terrain texture was randomized. Curriculum learning was deemed necessary when convergence could not be directly reached from scratch on hilly terrain. Moving through multiple training approaches, we found that: ($i$) Irregularities in terrain height cause policies trained exclusively on flat terrains to perform poorly; ($ii$) This problem can by solved by undergoing further training on hilly terrain; and, ($iii$) The curriculum learning approach (see Fig.~\ref{fig:rew_progr}) was instrumental in eventually handling the complex tasks; i.e., hills with many random obstacles.

% distributed training, batch and minibatch size, Adam with HP, 
Training relied on the ADAM algorithm \cite{Kingma14ADAM} with a learning rate of 1e-4. The training set at each update included 6000 tuples (timesteps), fed by 1000 element mini-batches to the optimizer, which operated eight epochs per update.
Since Chrono is compatible with OpenAI gym \cite{Brockman16Gym} environments, the dataset was collected by running six parallel episodes leveraging OpenAI baselines \cite{baselines} multiprocessing tools.
% MB size comment?
\begin{figure}[h]
    \centering
    \includegraphics[width=.75\linewidth]{images/CORLrewards.pdf}
    \caption{Reward progression. Vertical lines represent the changes introduced to make the environment more challenging. In order of occurrence, the dotted lines mark: fix obstacle count at 30; change from flat to hilly terrain; and increase of obstacle count to 50. }
    \label{fig:rew_progr}
\end{figure}

### 2020-08-25 2020/CoRL_2020/sections/related.tex
Added BookIta template, Eccomas DRL paper, GVSETS AV paper, CoRL (under review)

%!TEX root = ../main.tex

\section{Related Work}
\label{sec:relatedWork}

%This section provides a summary of the state of the art in four areas: simulation engines (\S\ref{subsec:dynamicsEngines}); sensing simulation (\S\ref{subsec:sensingSimulation}); techniques used for learning (\S\ref{subsec:learningTechniques}); and software environments for learning automation (\S\ref{subsec:learningTechniques}).

This section provides a summary of the state of the art in simulation environments (\S\ref{subsec:simEnvironments}), and learning techniques for autonomous navigation (\S\ref{subsec:learningTechniques}). The discussion of simulation environments is restricted to those commonly used for training reinforcement learning algorithms.

\subsection{Simulation Environments for Reinforcement Learning}
\label{subsec:simEnvironments}

Gazebo \cite{koenig2004design}, one of the most broadly used simulators in robotics, has been used for reinforcement learning by leveraging the open-source nature and tight integration with ROS. Gazebo exposes an environment that wraps multiple dynamics engines and sensors. However, it lacks specific support for vehicle modeling and deformable terrain for off-road scenarios.

Widely used for ``in-door robotics'' reinforcement learning, MuJoCo \cite{todorovMujoco2012} is a dynamics engine that supports URDF-based modeling. MuJoCo is not open-source and it does not support high-fidelity vehicle dynamics with deformable terrain interaction which can dominate many off-road maneuvers. The sensing support is also limited due to the simplicity of the noise models that can be applied to the sensors, most of which are restricted to dynamics-based interoceptive sensors.

An open-source alternative to MuJoCo is PyBullet \cite{matas2018simtoreal}, which exposes many of the same features but also lacks sophisticated sensing support. PyBullet provides an interface to generate the specific sensor data desired by the user. The strengths of PyBullet are in rigid-body dynamics and ease of use, making it a convenient choice for simpler RL applications, but limiting for off-road navigation.

In a more vehicle-centric approach, CARLA \cite{carlaAVsim2017}, AirSim \cite{shah2018airsim}, and Torcs \cite{torcsRacingSimulation2020} all seek to allow vehicle simulation for training and testing of control algorithms. While these are vehicle-focused, they cannot perform off-road simulation and have limited sensor fidelity. Torcs, which is originally based on a racing game, provides limited support for sensing, allowing access to camera and simplified lidar, with dynamic information available directly from the physics engine. CARLA and AirSim are designed for on-road applications and support an array of sensors. The sensors include basic distortions and noise. Due to limited geometric fidelity and time-resolution of collision-based ray-casting, the sensor data is typically overly clean or has obvious discontinuities or modeling artifacts. None of these on-road simulation environments support complex off-road navigation.

%%%%%%%%%%%%%%%%%%%%%%%%%% SIMONE
\subsection{Learning techniques}
\label{subsec:learningTechniques}

Deep Reinforcement Learning (DRL) techniques have shown outstanding level of performance since their very introduction \cite{Mnih13}. Subsequently, DRL has been used in vision-based robotic manipulation tasks. Robots controlled by DRL-trained neural networks (NN) have been shown to solve complex tasks in unstructured environments with \cite{zhu2018reinforcement} or without \cite{LevineFDA15} the use of imitation learning. 
End-to-end RL approaches have also been successfully applied to on-road autonomous driving. One of the major challenges in this area is the gap between RGB images generated by simulators and real world camera images, that can cause the autonomous driving policies trained in simulation to perform poorly in the real world. This issues has been addressed in various ways, such as using synthesized realistic images \cite{YurongALDriving17} or tools to generate images directly from real-world sampling \cite{Amini2020RLDriving}.

Sensor fusion with DRL techniques have shown promising results in controlling small indoor robots with camera + LiDAR \cite{BohezVCVSD17,PatelCKK2017}. RL in conjunction with imitation learning have been used for off-road driving when the vehicle learned to run laps quickly  \cite{Pan2017OffRoadAV}. However, to the best of our knowledge, there has been no demonstration of an end-to-end, off-road driving policy capable of reaching a target position while avoiding randomly placed obstacles on deformable soil and hilly terrain.

### 2020-08-25 2020/CoRL_2020/sections/results.tex
Added BookIta template, Eccomas DRL paper, GVSETS AV paper, CoRL (under review)

%%!TEX root = ../main.tex
%
%
%\section{Results}

To quantify the end-to-end learned navigation, the primary metric is success rate. This is simply a measure of the vehicle's ability to reach the destination without colliding with any obstacles. For this metric, any collision, regardless of severity, is considered a failure and the simulation is terminated. In addition to the primary metric, the length of the path is analyzed relative to the PSO path. Since the PSO path would generally be able to find a shorter path than the trained algorithm, which only has local information, the comparison will be used to show that the path taken by the trained algorithm is reasonable, and not simply a path that avoids obstacles yet produces bizarre trajectories.

When testing the effect of hilliness on navigation, the maximum height of the random height map was increased from 0 m (flat) to 12 m in increments of 2 m. At each level, 200 simulations on random height maps generated using simplex noise were performed to understand the success rate of the algorithm with respect to the terrain change. The results are shown in Fig. \ref{fig:rigid_height_success}. As expected, the trained network's ability to safely navigate the environment decreases with increased hilliness. Additionally, as the number of obstacles is increased, the task becomes more difficult. Tests conducted with 20 obstacles represent the closest point to the training environment.

In addition to the increasing height tests, experiments were conducted with hilliness of 0 m and 6 m with three terrain models: undeformable (rigid), hard-deformable, and soft-deformable. Since no deformable terrain was experienced in training, this is also a measure of policy robustness with respect to the terrain. These results, for varying terrain, heights, and obstacles are shown in Fig. \ref{fig:obstacles_hilly_and_flat_success}. While the policy can maintain relatively high success on deformable soil when the terrain is flat, the combination of deformable and hilly terrain proves difficult for the policy with success rates falling below 50\%.

\begin{figure}
    \captionsetup{justification=centering}
    \centering
    \begin{subfigure}{0.36\textwidth}
        \includegraphics[height=.25\paperheight]{images/demonstration/rigid_height_success_rate.png}
        \caption{Success rate in \% vs hilliness on rigid terrain.}
        \label{fig:rigid_height_success}
    \end{subfigure}%
    \hfill
    \begin{subfigure}{0.44\textwidth}
        \includegraphics[height=.25\paperheight]{images/demonstration/terrain_obstacles_flat_and_height_success_rate.png}
        \caption{Success rate in \% vs hilly and flat terrains with varying stiffness.}
        \label{fig:obstacles_hilly_and_flat_success}
    \end{subfigure}
    \caption{Success rate when increasing complexity of the terrain.}
    \label{fig:successRates}
\end{figure}

The second metric, which compares the length of the chosen path to the length of a path generated by the PSO, is discussed for a single environment setup. This is shown in Fig. \ref{fig:pso_length_compare} for flat, rigid terrain with 50 obstacles, which is more than double the obstacle density used in training. The length difference is computed as $length(NN path) - length(PSO path)$. These results show that the mode of the distribution is around -5 m, meaning the NN path is most often 5 m longer than the PSO path. While a direct comparison is not feasible since the PSO takes into account global knowledge and does not make any claims about optimality, the path taken by the NN appears to be moderately close in length to a global planner. This means that the policy is appropriately weighting the directness of the path as expected. Only successful paths were included in the calculation of this metric.

To further understand and analyze the success rate of the algorithm, the full set of collisions were evaluated based on the directness of collision. The directness of collision was computed by measuring the overlap of the projection of the vehicle and obstacle onto a plane perpendicular to the vehicle heading. This metric quantifies scraping collisions near 0\% and direct collisions near 100\%. This percentage can be interpreted as the percent of the frontal area of the vehicle that collided with the obstacle. While this cannot directly assign severity to the collision, it can hint at the type of collisions experienced by the policy. The distribution of results in Fig. \ref{fig:rigid_failure_metric_histogram} show that while the mode is near 100\%, there is also a significant portion of the collisions that have low directness. %This suggests that further analysis and adjustments could be made to the training to understand if increased penalty to force the vehicle away from obstacles could improve the success rate further.

\begin{figure}
    \captionsetup{justification=centering}
    \centering
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[height=.22\paperheight]{images/demonstration/rigid_9_pso_metric_histogram.png}
        \caption{Difference in path length between the pso trajectory and vehicle path for flat rigid terrain with 50 obstacles.}
        \label{fig:pso_length_compare}
    \end{subfigure}%
    \hfill
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[height=.22\paperheight]{images/demonstration/rigid_failure_metric_histogram.png}
        \caption{Failure severity on flat rigid terrain for all obstacle configurations.}
        \label{fig:rigid_failure_metric_histogram}
    \end{subfigure}
    \caption{Additional metrics analyzing path length and collision directness.}
\end{figure}

### 2020-08-25 2020/GVSETS_2020/GVSETS.tex
Added BookIta template, Eccomas DRL paper, GVSETS AV paper, CoRL (under review)

\documentclass[12pt,twocolumn]{article}

% Page dimensions
\setlength{\hoffset}{-0.4in}
\setlength{\voffset}{-0.5in}
%\setlength{\headsep}{-0.2in}
\setlength{\oddsidemargin}{0in}
\setlength{\textwidth}{7.4in}
\setlength{\textheight}{8.5in}
\setlength{\columnsep}{0.3in}

% Packages
\usepackage{fancyhdr}
\usepackage{abstract}
\usepackage{amsmath,amssymb} 
\usepackage{subcaption} 
\usepackage[usenames,dvipsnames]{color}
\usepackage{placeins}
\usepackage{hyperref}
\usepackage{sectsty}
\usepackage{enumitem}
\usepackage{todonotes}
\usepackage{lastpage}

\hypersetup{
	colorlinks=false,
	pdfborder={0 0 0},
}

%\usepackage[inline]{showlabels}
%\showlabels[\color{blue}]{cite}
%\showlabels[\color{blue}]{ref}

%% ============================================================

% Macros
%\newcommand{\CHRONO}{{\sffamily{{Chrono}}}}
%\newcommand{\ChronoFEA}{{\sffamily{Chrono}}::FEA}
%\newcommand{\ChronoVehicle}{{\sffamily{Chrono}}::Vehicle}
%\newcommand{\ChronoFSI}{{\sffamily{Chrono}}::FSI}
%\newcommand{\ChronoGranular}{{\sffamily{Chrono}}::Granular}
%\newcommand{\ChronoParallel}{{\sffamily{Chrono}}::Parallel}
%\newcommand{\ChronoDistributed}{{\sffamily{Chrono}}::Distributed}
%\newcommand{\ChronoOpenGL}{{\sffamily{Chrono}}::OpenGL}

\newcommand{\chrono}{Chrono}
\newcommand{\chronomod}[1]{\chrono{}::#1}
\newcommand{\synchrono}{Syn\chrono{}}
\newcommand{\gymchrono}{Gym\chrono{}}
\newcommand{\pychrono}{Py\chrono{}}

\newcommand{\gps}{\textsc{gps}}
\newcommand{\imu}{\textsc{imu}}
\newcommand{\mpi}{\textsc{mpi}}
\newcommand{\lidar}{lidar}
\newcommand{\cpu}{\textsc{cpu}}
\newcommand{\dsrc}{\textsc{dsrc}}
\newcommand{\gpu}{\textsc{gpu}}
\newcommand{\ros}{\textsc{ros}}
\newcommand{\tcp}{\textsc{tcp}}
\newcommand{\fea}{\textsc{fea}}
\newcommand{\api}{\textsc{api}}
\newcommand{\scm}{\textsc{scm}}
\newcommand{\optix}{OptiX}

%% ============================================================

% Styles

\definecolor{my-gray}{gray}{0.4}

% First page header
\fancypagestyle{firststyle} {
	\lhead{}
	\rhead{
	\footnotesize
	\textbf{2020 NDIA GROUND VEHICLE SYSTEMS ENGINEERING AND TECHNOLOGY SYMPOSIUM}\\
	\textbf{\sc Autonomy, Artificial Intelligence \& Robotics (AAIR) Technical Session}\\
	\textbf{\sc August 11-13, 2020 -- Novi, Michigan}	
	}
	\chead{}
	%
	\lfoot{}
	\cfoot{}
	\cfoot{\footnotesize DISTRIBUTION STATEMENT A. Approved for public release; distribution unlimited. OPSEC\#864}
	\rfoot{}
}

% All other pages (header and footer)
\pagestyle{fancy}
\fancyhf{}
\rhead{
	\color{my-gray}
	\footnotesize Proceedings of the 2020 Ground Vehicle Systems Engineering and Technology Symposium (GVSETS)
	}
\cfoot{
	\color{my-gray}
	\footnotesize
	{\FooterTitle}\\
	Page \thepage~of~\pageref{LastPage}
}

\renewcommand{\headrulewidth}{0pt}

% Font and styles for sections
\allsectionsfont{\fontsize{12}{15}\selectfont}

\renewcommand{\abstractname}{ABSTRACT}
\renewcommand{\refname}{REFERENCES}

%% ============================================================

\input{macros}
\newcommand{\gonogo}{{\normalfont\textsc{go/no-go}}\xspace}
\newcommand{\go}{{\normalfont\textsc{go}}\xspace}
\newcommand{\nogo}{{\normalfont\textsc{no-go}}\xspace}
%\newcommand{\chrono}{{\textsl{Chrono}}\xspace}

\newcommand{\Reg}{\textsuperscript{\tiny\textregistered}}

\newcommand{\papertitle}{Enabling Artificial Intelligence Studies in Off-Road Mobility Through Physics-Based Simulation of Multi-Agent Scenarios}

%% ============================================================

\title{\bf\large \MakeUppercase{\papertitle{}}}

\newcommand{\FooterTitle}{\papertitle{}, Negrut, et al.}

\author{	
	{\bf D. Negrut, R. Serban, A. Elmquist, J. Taves, A. Young}\\
	University of Wisconsin -- Madison\\
	Madison, WI
	\and
	{\bf A. Tasora, S. Benatti}\\
	Dipartimento di Ingegneria ed Architettura \\
	University of Parma, Italy
}

\newcommand{\MyAbstract}{\textit{
\noindent 
We describe a simulation environment that enables the design and testing of control policies for off-road mobility of autonomous agents. The environment is demonstrated in conjunction with the design and assessment of a reinforcement learning policy that uses sensor fusion and inter-agent communication to enable the movement of mixed convoys of conventional and autonomous vehicles. Policies learned on rigid terrain are shown to transfer to hard (silt-like) and soft (snow-like) deformable terrains. The enabling simulation environment, which is Chrono-centric, is used as follows: the training occurs in the GymChrono learning environment using PyChrono, the Python interface to Chrono. The GymChrono-generated policy is subsequently deployed for testing in SynChrono, a scalable, cluster-deployable multi-agent testing infrastructure that uses MPI. The Chrono::Sensor module simulates sensing channels used in the learning and inference processes. The software stack described is open source. Relevant movies: \cite{simsPaperGVSETS2020}.
} 
}

%% ============================================================

\begin{document}
\date{}

% ------------------------------------------

\twocolumn[
\maketitle             % full width title
\thispagestyle{firststyle}
\begin{onecolabstract} % ditto abstract
\MyAbstract
\end{onecolabstract}
\vspace{0.1in}
]

\newpage

%% ============================================================

\section{INTRODUCTION}
\label{s:introduction}
Computer simulation has been extensively used in the design and analysis of various automation aspects tied to on-road mobility, see, for instance \cite{carlaAVsim2017}. A similar statement cannot be made for off-road mobility owing to a smaller market and a set of stiff challenges brought along by the task at hand. However, a predictive simulation platform for off-road mobility analysis of autonomous agents (AAs) is very desirable since it can accelerate the engineering design cycle, reduce costs, perform more thorough testing, and produce more performant and safer designs. Simulation is not a silver bullet as it has its limitations, first of all the issue of simulation-to-reality transfer \cite{sim2realGapEssex1995}, which pertains to the failure of control policies derived in simulation to work well in the real world. Furthermore, models are difficult to set up and calibrate, the validation process can be tedious and time consuming, and open source simulation tools that are both predictive and expeditious are not readily available. This contribution addresses the third point. It describes a simulation environment whose stated purpose is to allow the practitioner to gain insights into the operation of AAs (robots and autonomous wheeled or tracked vehicles) in off-road conditions with an eye towards: improving mechanical designs of AAs; and, producing and testing control policies that govern the operations of the AAs. These are topical goals for the Army, see, for instance, \cite{jayakumarAutonomous2011,jayAbhiMarco2015,jayGorsich2018,jayGVSETS2019}, or, for a historical perspective on early contributions, \cite{frederickTARDECrobotics2014}.

There are several ongoing efforts that seek to address the AA simulation issue. In robotics, Gazebo~\cite{gazebo,koenig2004design} is a widely used 3D multi-robot simulator with dynamics.  It is not a simulation engine per se, but a \textit{platform} that exposes several engines: ODE~\cite{ode2015}, Bullet~\cite{bulletPhysicsEngine2020}, DART~\cite{dart2018}, and Simbody~\cite{sherman2011simbody}. Unlike Gazebo, which is open source, CoppeliaSim (formerly V-REP)~\cite{V-REP2013} is a commercial multi-robot simulation solution that also exposes a set of simulation engines: MuJoCo~\cite{todorovMujoco2012}, Vortex Dynamics~\cite{cmLabs2020}, Bullet, and Newton Dynamics~\cite{newtonDynamics2020}. ROAMS~\cite{abhiROAMS2004} and ANVEL~\cite{ANVEL-website} are two other simulation engines for off-road AAs. The former is used for mission planning by NASA and draws on an in-house dynamics engine~\cite{DARTS}; the latter relies on the ODE simulation engine and has been used in the past for off-road military applications~\cite{ANVELdescription2012} in combination with a sensor simulation package~\cite{VANE-2012}. MAVS is an off-road AA simulation environment that is currently under active development \cite{mavsCarruth2018}. It provides an in-house developed, sophisticated sensor simulation module \cite{goodin2018enabling,goodin2019predicting}, has a ROS bridge, and uses Chrono as its dynamics engine. USARSim is an AA simulation platform, not under active development, that draws on a game engine (Unreal Engine~\cite{unrealEngine}), a choice with pluses (scalability, ability to create complex worlds) and minuses (the simulation engine is designed for plausibility rather than accuracy). For autonomous vehicle (AV) simulation, Carla~\cite{carlaAVsim2017} and AirSim~\cite{shah2018airsim} are two often used open-source simulators, the former for on-road AV driving scenarios simulation, the latter originally designed for drones but now including support for on-road traffic of AVs as well. Carla and AirSim rely on Unreal Engine but several other engines are used for AA simulation, e.g. Unity~\cite{unityGaming} and TORCS~\cite{torcsRacingSimulation2020}. For a survey of other solutions for on-road mobility the reader is referred to~\cite{surveyAVsimulationBerger2019,reviewPerception2019}.

The AA off-road mobility simulation platform discussed here is Chrono-centric~\cite{projectChronoWebSite,chronoOverview2016}. In its purpose, it is similar to the ANVEL-VANE environment as it seeks to simulate robots and wheeled/tracked vehicles operating in off-road conditions. Compared to the ANVEL-VANE solution, the Chrono environment is different in several respects: it is open source and available for unfettered use under a BSD3 license; it uses its own multi-physics engine; it is scalable and deployable on supercomputers, clusters, or multi-core architectures owing to its reliance on the MPI message passing standard for parallel computing~\cite{MPI-Forum-1994}; and is under active development. Chrono is an ecosystem of modules and toolkits. It has support for rigid and flexible body dynamics (Chrono::Engine), fluid-solid interaction (Chrono::FSI), and granular dynamics (Chrono::Parallel and Chrono::Granular) applications. It has Python bindings, support for sensor simulation in Chrono::Sensor, an API for ROS~\cite{ROS-2009} bridging, as well as facilities for: rapid vehicle modeling via parameterized templates~\cite{ChronoVehicle2019}; control policy design with GymChrono; and scalable control policy testing with SynChrono. Chrono relies on GPU computing for fluid-solid interaction and select granular dynamics simulations, multi-core for most of the other modules, and MPI-enabled parallel computing for co-simulation when handling large terramechanics applications or collections of AAs. Real-time simulation is not one of Chrono's priorities. Although for vehicle-on-rigid-terrain simulation it provides faster than real-time performance, there are numerous applications that lead to long run times in Chrono, e.g., deformable terrain mobility, nonlinear flexible body dynamics, fording scenarios, etc.

This contribution highlights the Chrono components that support the design and testing of control policies through simulation: PyChrono, GymChrono, Chrono::Sensor, and SynChrono. To show these components at work, a Reinforcement Learning (RL) approach is used herein to produce a control policy. There is nothing special about the RL approach; other techniques to design control policies could be used equally well, a point touched upon in more detail in Section \S\ref{sec:ML}. Section \S\ref{sec:simInfrastructure} describes the Chrono infrastructure that facilitates artificial intelligence studies in off-road, multi-agent mobility scenarios. Section \S\ref{sec:demoTechnology} covers simulation experiments that highlight two aspects: the scalability of the SynChrono testing environment, and the process of designing the RL control policy along with an evaluation of the policy's robustness. Concluding remarks and directions of future work round off the contribution.

\section{DERIVING CONTROL POLICIES THROUGH SIMULATION}
\label{sec:ML}
Derived using an accurate simulation framework, control algorithms have been shown to bridge the sim-to-reality gap successfully~\cite{VISTASimMITRL, VirtualToRealRL}. The use of vehicles with Level 1 and Level 2 autonomy has grown considerably~\cite{J3016_201401, SAEHardwareInTheLoop} and the automotive industry is making major strides in the transition to Levels 3 and 4 autonomy~\cite{WayveRLSimToReal, TeslaSummon}. The use of simulation for on-road AVs is an area of intense research and development as this technology is seen as an important catalyst of the aforementioned transition. One active area of research is focused on sampling-based methods. These approaches generate many candidate trajectories a vehicle can follow, selecting and executing the controls associated with the best candidate~\cite{UrbanAVBezierMPC, RRTRealTime}. Graph search methods are commonly associated with each trajectory. The approach is real-time challenged, since achieving robust results requires a high number of samples to be analyzed~\cite{TrajectoryPlanningMIQP}. Algorithms such as Dijkstra's, A-Star (A*), or the Rapidly-exploring Random Tree-Star (RRT*), sample the state space either deterministically or stochastically~\cite{RRTRealTime}. Depending on the complexity of the traffic scenario, these algorithms can prove computationally costly and provide sub-optimal results.

Model Predictive Control (MPC) is another common AV control approach~\cite{Borrelli2005MPCbasedAT}. Using a dynamic model of a vehicle, the MPC algorithm computes trajectories over the state space and determines an optimal trajectory using gradient-descending optimization techniques~\cite{TrajectoryPlanningMIQP, Liniger2017OptimizationBasedAR}. A limited time horizon is employed to reduce unneeded computation for times too far out into the future. In comparison to sampling algorithms, the MPC approaches show improved performance that is tied to the use of the gradient fields in the optimization problem that comes into play~\cite{Borrelli2005MPCbasedAT}.

The accuracy of the simulation platform plays a critical role both for MPC as well as sampling-based controllers. To adequately validate and subsequently verify a controller such as MPC, the simulation must be of high enough definition to carry over successfully to reality~\cite{AVTestingMethods}. When using the more pedestrian PID controller solutions, for which gains must be carefully selected, an inaccurate simulation platform could yield a poor design that leads to undesired consequences when deployed on a real vehicle.

The design of a robust controller that performs adequately in complex environments using the aforementioned strategies has proven difficult when aiming for a generalized policy~\cite{Kuutti2019ASO}. An emerging approach that has gained momentum in recent years is Machine Learning (ML) based \cite{surveyRLrobotics2013}. ML has shown promise in producing efficient and robust models that generalize well in a variety of situations. The three pillars of ML include supervised learning, unsupervised learning, and reinforcement learning. In the AA problem, deep reinforcement learning (DRL) has been very successful, as it displays the ability to learn and respond in complex scenarios without the need for preprocessed or labeled data~\cite{VirtualToRealRL}.

Since its introduction~\cite{Mnih13}, DRL has proven successful in robotics applications~\cite{Levine2016,inHandManipulation2020}. At its core, DRL is an iterative learning process in which an \emph{agent} interacts with an \emph{environment}; at each iteration the agent collects an \emph{observation} (or \emph{state}), then performs an \emph{action} based on the previous observation and gets a \emph{reward} which is a measure of its performance. The goal of RL is to find a policy that maximizes the sum of the collected reward.

RL allows for complex control policies viable in unstructured and stochastic environments; it is model-free in that it does not require a model that predicts environment transients; and can learn from scratch. RL's major flaw is its need for a massive amount of training data to infer a robust policy. The role of simulation is to produce this collection of samples. \emph{Policy Gradient Algorithms} are a subset of RL algorithms whose goal is to directly learn an optimal stochastic policy $\pi_\theta(a|s)$, where $s$, $a$, and $\theta$ are the state, action, and a set of learnable parameters, respectively. If $\pi$ is a Neural Network (NN), $\theta$ are the weights and biases of the NN, the state is the NN input, and the action is its output. Proximal Policy Optimization (PPO)~\cite{Schulman2017PPO} is one of the most widely used algorithms for continuous state and action environments and will be the algorithm of choice in this contribution. PPO is a policy gradient algorithm whose goal is to optimize a stochastic policy. It is also an actor-critic method since another NN is trained and used to estimate the Value Function~\cite{SuttonBarto1998RL} employed to estimate the Advantage Function~\cite{Kakade02approximatelyoptimal}. The Advantage Function is used in the objective function to be maximized.

\section{SIMULATION INFRASTRUCTURE}
\label{sec:simInfrastructure}
The purpose of the simulation environment described is twofold. First, it is used to produce the data needed to design a control policy. Second, it is used for testing purposes. To this end, it exposes the control policy produced in a model-based or model-free approach to a battery of tests to gauge its correctness and robustness. This section outlines the components of this Chrono-centric simulation environment. The emphasis is placed on four components: Chrono::Sensor, PyChrono, GymChrono and SynChrono, that anchor the AA design and testing process and have not been discussed elsewhere in detail. More established Chrono components or functionality will be touched upon in passing; more details are provided in  \cite{chronoOverview2016,ChronoVehicle2019,ChronoSCM2019}.

\paragraph{Chrono.} Under active development for over two decades, Chrono~\cite{chronoOverview2016} is a multi-physics simulation engine distributed as open-source under a permissive BSD license.  Its core module, Chrono::Engine, provides support for rigid multibody dynamics, nonlinear finite element analysis, and frictional contact dynamics. Chrono is modular, with optional modules providing support for additional classes of physics simulation (e.g., fluid-solid interaction or large-scale granular dynamics), for modeling and simulation of specialized mechanical systems (e.g., ground vehicles), for interfaces to external solvers (e.g., sparse direct linear solvers), or for dedicated parallel algorithms targeting different computing architectures (multi-core, distributed, and GPU) for large-scale simulations.

Written almost entirely in C++, Chrono is middleware and therefore supports customized solutions that involve user code and potentially third-party software. The software is portable and can be built on different platforms, under different operating systems, and using various compilers. Chrono is actively developed, has a continuous integration process, an active user forum, and is managed through GitHub~\cite{projectChronoGithub}. It's latest release is 5.0.1, available as of March 2020.  Chrono is used by academic, industrial, and government research and development groups and projects.

\paragraph{Chrono::Vehicle.} Chrono::Vehicle~\cite{ChronoVehicle2019} is a specialized Chrono module that makes available a collection of templates (fully parameterized models) for various topologies of both wheeled and tracked vehicle subsystems. It provides facilities for modeling rigid, deformable, and granular terrain; support for closed-loop and interactive driver models; and run-time and off-line visualization of simulation results. Chrono::Vehicle leverages and works in tandem with other Chrono modules for run-time visualization or finite element, granular dynamics, and parallel computing support.

Chrono::Vehicle provides vehicle \textit{subsystem} templates for tires, suspensions, steering mechanisms, drivelines, sprockets, track shoes, etc.; templates for external systems such as powertrains, drivers, and terrain models; and additional utility classes and functions for vehicle visualization, monitoring, and collection of simulation results.  As a middleware library, Chrono::Vehicle requires the user to provide C++ classes for a concrete instantiation of a particular template.  An optional Chrono library provides complete sets of template instantiations for several concrete ground vehicles, both wheeled and tracked, which can serve as examples for developing more customized vehicle models. An alternative mechanism for defining concrete instantiation of vehicle system and subsystem templates is based on input specification files in the JSON format \cite{json}.  For additional flexibility and to allow integration of third-party software, Chrono::Vehicle is designed to permit either monolithic simulations or co-simulation where the vehicle, powertrain, tires, driver, and terrain/soil interaction can be simulated independently.

\begin{figure}
    \centering
    \includegraphics[width=0.8\columnwidth]{Figs/HMMWV-Granular.png}
    \caption{{\small Chrono::Vehicle HMMWV with flexible tires navigating granular terrain demonstrating vehicle dynamics, flexible body dynamics, and parallel computing support in Chrono \cite{antonioVehicleTireGranMatSim2017}.}}   
    \label{fig:hmmwvgranular}
\end{figure}

Chrono::Vehicle provides several classes of terrain and soil models, of different fidelity and computational complexity, ranging from rigid, to semi-empirical Bekker-Wong type models, to complex physics-based models based on either a granular or finite-element based soil representation.  For simple terramechanics simulations, Chrono::Vehicle implements a customized Soil Contact Model (SCM), based on Bekker theory, with extensions to allow non-structured triangular grids, adaptive mesh refinement, and incorporation of bulldozing effects~\cite{ChronoSCM2019}.  Second, Chrono provides an FEA continuum soil model based on multiplicative plasticity theory with Drucker-Prager failure criterion and  specialized brick elements. Finally, leveraging Chrono support for large-scale granular dynamics and for multi-core, GPU, and distributed parallel computing, off-road vehicle simulations can be conducted using fully-resolved, granular dynamics-based complex terramechanics, using a Discrete Element Method approach, see Fig.~\ref{fig:hmmwvgranular} \cite{antonioVehicleTireGranMatSim2017, serbanCosimIJVP2019}.

\paragraph{Chrono::Sensor.} Chrono::Sensor provides sensor simulation support for software-in-the-loop testing. Cameras, lidar, GPS, and IMU sensors can be placed within a Chrono simulation to generate synthetic data based on user-defined sensor parameters and attributes of the virtual world hosting the AA simulation experiment. The goal of the module is to allow realistic data generation based on sensor characteristics such as noise, distortion, and lag. Sensors can be attached to objects within the simulation and configured to match corresponding real sensors. For modeling convenience, sensors can be defined through a JSON file \cite{json}. Additionally, custom sensors and post-processing filtering can be implemented, leveraging the existing render framework or physics interface.

For interoceptive sensing (GPS and IMU), the module utilizes the internally computed physical quantities from the Chrono system and can augment this ground truth with drift, Gaussian noise, lag, and filtering characteristics from finite collection time. For exteroceptive sensors that provide information about scene characteristics (camera and lidar), Chrono::Sensor leverages hardware accelerated ray tracing through the OptiX library~\cite{optixNVIDIA} and implements physically based rendering techniques. The ray tracing approach allows for the physical reconstruction of the light-based data acquisition process and thus controls the attributes of the synthetically generated sensor data. For camera, lens models and post-processing noise augmentation are supported, with an interface to extend or implement custom models. For lidar, the framework expands on work from~\cite{goodin2018enabling} to provide a beam divergence model that supports multiple modes of lidar return and reduced intensity during partial beam reflectance. The camera and lidar can also be parameterized by update rate, time over which to collect data, and lag.

Chrono::Sensor is currently in development with planned expansion of sensor support and additional distortion model implementations. All sensors and capabilities are written in C++, but can also be accessed from Python through the PyChrono interface. The entire module can be run headless without the requirement of a render context, allowing for ease of deployment in machine learning applications on remote servers (in the cloud).

\paragraph{PyChrono.} While the main Chrono API is expressed in C++, in recent years a concerted effort was dedicated to providing automatically-generated Python wrappers for much of the Chrono functionality. The purpose was twofold: to provide a lower-entry point to Chrono simulations for users less familiar with C++; and, provide a bridge to various machine learning platforms, e.g. TensorFlow \cite{tensorflow}, PyTorch \cite{paszke2017PyTorch}, Theano \cite{theano2016}, and CAFFE \cite{caffe2014}.

The Python wrapping relies on using automated technology provided by SWIG \cite{swig2003}, which generates the interface between Python user-code and the underlying Chrono C++ libraries. Presently, a large set of Chrono functionality is exposed to Python users, including the core multibody and FEA module, the interface to CAD systems (like SolidWorks), run-time visualization with Irrlicht, etc. In particular, full support is available for both modeling, simulation, and visualization of wheeled ground vehicles and use of the Chrono::Vehicle existing wheeled vehicle models, as well as using the sensor models provided by Chrono::Sensor. PyChrono for Python 3 can be built from sources on any of the supported platforms (Linux, Windows, MacOS).  Alternatively, pre-built conda PyChrono packages are available on the project's Anaconda page~\cite{pyChronoCondaWebSite} (note that Chrono::Sensor is not available yet via the conda PyChrono packages).

\paragraph{GymChrono.} GymChrono is an extension of OpenAI Gym~\cite{Brockman16Gym}. It exposes a set of environments providing continuous control tasks for physics and sensor simulation run by the Chrono backend. These environments inherit from OpenAI Gym classes. As such, they can be used out of the box with any algorithm or DRL framework made for gym environments. They can also draw on gym's environment parallelization for learning acceleration.

\paragraph{SynChrono.} SynChrono is a software component that uses Chrono to implement a distributed-memory execution model when simulating scenarios that include multiple AAs. By leveraging the Message Passing Interface (MPI) standard~\cite{mpi-3.0}, SynChrono can manage multiple instances of Chrono running together \textit{one} mobility analysis on a supercomputer, cluster, or multi-core setup thus supporting the scalable and distributed simulation of multiple agents (robots, tracked vehicles, wheeled vehicles, etc.) The paradigm embraced is that of running the dynamics of one AA as one MPI rank, with the ranks/AAs communicating through MPI messages to maintain space and time coherent state for all agents participating in the study. As an example, if there are two agents, SynChrono makes it possible to synchronously run the two agents on two different compute nodes in a supercomputer. By the same token, if there are 50 agents and 50 compute nodes in a cluster, SynChrono provides the infrastructure to keep the 50 agents operating in a coherent (timewise and spacewise) virtual world. Although each agent represents a Chrono simulation, SynChrono makes it possible for ruts generated by agent 27 to be picked up on a camera sensor on agent 31 if the camera points to agent 27. The time coherence aspect prevents some agents from racing into the future while other agents lag behind in the past. The global synchronization mechanism in SynChrono ensures that all 50 agents march forward in simulation time in a coherent fashion so that mutual interaction (a vehicle crossing the ruts of a different one, a vehicle sensing another vehicle, etc.) happens as in real life.

The structure of SynChrono's MPI framework is shown in Fig.~\ref{fig:mpischematic}. SynChrono manages multiple AAs as multiple processes via as many MPI ranks. Each AA runs in its own SynChrono process (an MPI rank) and interfaces with its dedicated control stack for software-in-the-loop control. The control stack is fed synthetic data generated by Chrono::Sensor and acts upon the environment through Chrono::Vehicle control inputs (throttle, steering, braking). The control algorithm for each agent is also configurable and can vary from complex algorithms that fuse sensor feeds/data streams, to controls based on empirical models, and on to pre-recorded driver inputs from a human or human-driven control in scenarios that are simple enough to allow real-time simulation. SynChrono supports human-in-the-loop experiments as well.

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{Figs/syn_cars_diagram.png}
	\caption{{\small Schematic of the SynChrono framework. Dynamics simulations are done in separate Chrono systems and the outcome of the dynamics simulation is synchronized between ranks using MPI.}}   
	\label{fig:mpischematic}
\end{figure}

Each SynChrono process is responsible for the dynamics of a single agent. At a slow frequency (relative to the simulation time-step), all SynChrono processes communicate via the SynChrono daemon to exchange state information. State information is intended to be brief, sufficient to enable a SynChrono process to reconstruct a ``ghost'' version of outside agents in its own world for visualization and sensing purposes. In an example where each agent is a vehicle, the state information consists of the vehicle location and orientation along with pose information for each wheel. This information is packaged via the FlatBuffers serialization library \cite{flatbuffers}.

One limitation of the implementation is that currently two agents running as two SynChrono processes cannot crash with each other or participate in an operation that couples their dynamics, e.g. lifting together a heavy object. Such a scenario should be run in Chrono, since the coupling is too tight to be handled by SynChrono. This will make the simulation longer to run since more agents will have to be handled within one Chrono process. However, if the agent coupling happens via sensing or the virtual world, e.g., one agent sensing another one, or one vehicle crossing over and being jolted by the ruts left by a different agent, then SynChrono can be relied on, thus ensuring scalability. The goal of SynChrono is to have virtually constant scaling up to a high threshold where the data passing overhead becomes the simulation bottleneck.

\paragraph{Interface to an external controller.} For testing of control algorithms that are intended to be easily transferred to real-life vehicles or robots, the simulation platform provides an external control interface that is exposed in SynChrono. An agent in SynChrono can send messages (i.e. sensor data packets) to the external autonomous controls framework which can then send a message back (i.e. control inputs). The control stack is independent of the SynChrono platform (e.g. a bridge has been developed for ROS \cite{ROS-2009}), and can be tested with inputs replicating those from reality, such as sensor and/or V2X communication data.

%% ===========================================================
\section{TECHNOLOGY DEMONSTRATION}
\label{sec:demoTechnology}
All simulation scenarios considered in this section use a Chrono::Vehicle HMMWV model. Chrono::Vehicle was benchmarked as part of the Next Generation-NRMM (NG-NRMM) exercise \cite{NATObenchmark2018}; Chrono::Vehicle-specific benchmark findings are detailed in \cite{TR2017-05NRMM-wheeled,TR2016-08NRMM-tracked}. Note that: the NG-NRMM benchmarking effort involved Chrono::Vehicle but no the particular model used in this study; and, all results reported herein were obtained using a simulation time step of $\Delta t=2 \times 10^{-3}$ seconds, both for rigid and deformable terrain. This time step information is relevant when discussing real-time performance and scalability aspects.

\subsection{SynChrono scaling analysis}\label{sec:syn-scaling-analysis} 
SynChrono uses $N+1$ processes executing on a cluster/supercomputer to simulate the dynamics of $N$ agents handled as $N$ independent Chrono simulations. Each Chrono simulation is a process; the extra process is involved in maintaining the time and space coherence for the virtual world shared by the $N$ AAs. The SynChrono daemon executes as the extra process and it coordinates via MPI the dynamics of the $N$ agents. The numerical experiments described here answer the following questions: ($i$) How does the time to complete a simulation change as $N$ increases? ($ii$) How fast is SynChrono in mobility studies on rigid terrain? ($iii$) How fast is it when an SCM deformable terrain model is considered in the simulation scenario? The Linux cluster used exposed 15 nodes. Thus, one set of experiments was conducted with up to $N=14$ AAs, using rigid, SCM hard (silt-like), and SCM soft (snow-like) terrain, with one agent per compute node. However, the experiment also included scenarios with up to 56 AAs. In this case, four AAs ran on the same compute node. On the upside, this allowed more AAs to participate in the benchmark; on the downside, the simulation times went up since the hardware allocation per AA went down by a factor of four.

The handling of a virtual world that has SCM terrain is challenging since each of the $N$ vehicles changes the terrain at the same time and these changes must be space and time coherent. This is facilitated and managed by the SynChrono daemon. The key component of the SCM terrain is the deformation of each vertex in the underlying SCM mesh. All other terrain properties can be computed based on the height of each vertex alone~\cite{ChronoSCM2019}. At each simulation step, a SynChrono rank that is associated with an agent moving on SCM terrain collects a list of the deformed vertices that the agent produced during the course of that time step. Mesh deformation data may not be sent at every simulation time-step, so this collection of mesh changes is in general persistent across simulation time steps. Once an agent reaches a SynChrono synchronization point, the cumulative mesh deformations produced by one agent are sent via the daemon to every other agent's world. Each agent then applies the deformations to their own copy of the SCM mesh and resets their collection of new mesh deformations. This means that two agents should not come close enough that they deform the same vertices during the same synchronization period, but this is just as restrictive as SynChrono's assumption that any two agents will not interact by crashing. While Chrono's SCM implementation allows for meshes that auto-refine, this is currently not supported in SynChrono due to the significant increase in algorithmic complexity for relatively little gain in performance. The main concern for scalability is message size, since for synchronizing SCM terrain there is about 15 times as much data to send for each agent relative to the rigid terrain case.

The scenario discussed herein is that of many vehicles crossing perpendicularly on a rectangular patch of SCM terrain, see Fig.~\ref{fig:scalingenv} and online movies \cite{simsPaperGVSETS2020}. In this setup, one can easily scale up the number of vehicles and verify that the SCM terrain deformation is properly synchronized across multiple ranks. The scaling metric used was the Real Time Fraction (RTF), representing the amount of wall-time required to finish a simulation divided by the amount of time simulated. Running in real-time corresponds to a factor of 1, while slower than real-time corresponds to factors larger than one. The tests were run on the Euler computing cluster at the University of Wisconsin-Madison. Each node has an eight core, Broadwell-generation Intel processor; inter-node communication is facilitated via a Gigabit Ethernet interconnect.

In a first sub-experiment, we assign each SynChrono process (which runs one vehicle) to distinct cluster nodes. The scaling analysis results are presented in Fig.~\ref{fig:scmscaling}. The plot shows on the $y$-axis the $ log_{10} $ of the RTF. Thus, any dot below the horizontal $x$-axis is associated with an experiment that ran faster than real time. Simulations on the rigid terrain ran with an RTF of 0.75 and simulations on either type of SCM terrain (hard or soft) ran with an RTF of approximately 110. The RTF value of 110 is independent of the of SCM soil parameters (i.e. soft vs. hard), but is highly dependent on the processor performance, MPI implementation, compiler, and compilation optimization. These results confirm weak scaling behavior: no significant performance penalty is observed for including additional agents in the simulation when each SynChrono process is assigned to a different compute node.

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{Figs/Syn-SCM-setup.png}
    \caption{Environment used for SynChrono scaling analysis for agents operating on SCM terrain. Two lines of vehicles move across a rectangular patch, crossing orthogonally and making ruts in the SCM soil. Simulations were run on both soft and hard SCM terrain, as well as on rigid terrain.}
    \label{fig:scalingenv}
\end{figure}

Using the same scenario as above, scaling analyses were also conducted for a larger number of agents on rigid terrain to ensure that the same weak scaling continued to hold. To this end, two other sub-experiments were conducted: two SynChrono processes were deployed per node; and, four SynChrono processes were deployed per node. Given the number of nodes made available on the cluster (15), for two-per-node, SynChrono handled 28 AAs; for four-per-node, it managed a joint simulation of 56 AAs. Timing results are shown in Fig.~\ref{fig:scmscalingrigid}, which shows that 56 AAs run in slightly more than 1.8$ \times $ real time. Given a supercomputer or larger cluster, SynChrono opens the possibility of simulating swarms of AAs.
%Due to peculiarities in the network configuration, not all tests could be run with the same MPI implementation which had the side-effect of demonstrating that simulation performance was in general very dependent on the MPI implementation.

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{Figs/syn_SCM_scaling.png}
    \caption{\synchrono{} scaling analysis for rigid, SCM hard, and SCM soft terrains. Logarithmic scale used.}   
    \label{fig:scmscaling}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{Figs/syn_SCM_many.png}
    \caption{\synchrono{} scaling analysis on rigid terrain for a large number of agents.}   
    \label{fig:scmscalingrigid}
\end{figure}

%%\FloatBarrier

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%  LEARNING EXAMPLE  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\subsection{Learning to drive in a convoy}
\label{subsec:learningDemo}

%{\SBELcomment{We should make it clear what the vehicles know, sense, communicate, etc. I don't think we do justice to this at this point.}}
%

Chrono helps with two tasks: learning a control policy, and testing a policy that was designed in Chrono or brought from outside. In this example, PyChrono and GymChrono are used to design a policy. SynChrono is subsequently used to test it. The RL-based learning is done on rigid terrain using a nondescript texture. The goal is to enable a vehicle to move as part of a convoy. To test it, the policy is deployed on vehicles that are part of a four-vehicle convoy driving on rigid or SCM deformable terrain. Up to three of the convoy vehicles use this policy while driving in a platoon. Thus, the possible scenarios are: three lead vehicles and one following vehicle (3L+1F), two lead and two followers (2L+2F), and one lead and three followers (1L+3F). The \textit{lead} vehicles are programmed to follow a path defined by way-points; for all purposes, these can be considered human driven. A follower vehicle is autonomous and uses the learned policy to follow the vehicle in front of it. In doing so, it should not crash in the vehicle ahead of it, and avoid hitting obstacles in the vicinity of the path. To this end, it relis on a camera sensor, location acquired through GPS sensing, and compass heading. Note that communication allows a vehicle to find out the GPS location and velocity of the vehicle in front. Given that four vehicles are involved in this platooning experiment, SynChrono is subsequently used to test the policy to reduce simulation times. Indeed, this validation could be run in Chrono::Engine but would roughly require four times longer to complete it. The salient points of this experiment are as follows: although vehicles are run in different SynChrono processes, there is time and space coherence between them to the point where the vehicles sense each other; the learning occurs using rigid terrain with nondescript texture but the policy is tested on deformable terrain that is white (snow-like); this is an end-to-end policy that uses sensor data fusion to control both the steering and acceleration/braking of the vehicle.

\paragraph{Designing a policy.} The policy was obtained through training using a custom implementation of the PPO reinforcement learning algorithm leveraging PyTorch~\cite{paszke2017PyTorch} as the Deep Learning framework. The agent is a HMMWV vehicle modeled in Chrono::Vehicle. The goal of the training process is to develop a control policy that enables an agent (in this case the HMMWV) to drive in a convoy. For training, to increase the randomness of the path and thus the robustness of the control, two different path types were used on a 90x90 meters area. The first is S-shaped, starting from a corner and finishing in the opposite corner; the second is C-shaped, starting and finishing on the same side of the driving area. To further increase the randomness, these paths can be flipped along the $y$-axis (left-right) and rotated about the vertical $z$-axis to obtain 16 different possible paths. The starting point is picked randomly within the first half of the path as shown in Fig.~\ref{fig:paths}. Eight obstacles placed near the path are randomly selected from various rock, tree, and bush assets.

%\todo[inline]{SB: Two options, pick your favorite.}
%\begin{figure}
%    \centering
%    \includegraphics[width=\columnwidth]{Figs/all_paths.pdf}
%    \caption{All possible paths. The dashed lines represent the segment of the path in which the simulation episode can start}   
%    \label{fig:paths_all}
%\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{Figs/base_paths.pdf} 
    \caption{The double S and C paths used during training. Each one of these is randomly flipped and rotated, resulting in 16 different possible paths. The dashed lines represent the segment of the path in which the simulation episode can start.}
    \label{fig:paths}
\end{figure}

In order for the agent to accomplish its task, the vehicle must be aware of its surroundings. To that end, the HMMWV used two sensors simulated in Chrono::Sensor: a GPS sensor, and an RGB camera placed on the front bumper. The camera, which updates at 30 Hz, has a resolution of 80 $\times$ 45 pixels. This level of resolution suffices, since detailed features that can be extracted from higher resolution images are not needed by the control policy. Furthermore, given that the dataset contains one image per interaction, images of too high resolution can quickly deplete the available GPU memory due to the increased memory footprint of the NN update process. As such, large images that are contained in observations are typically avoided.

As with any other RL environment, an observation and a reward is provided to the ML algorithm at each timestep. Subsequently, the agent must perform an action prescribed by the ML algorithm in order to maximize the reward collected. The action is a two element array: the first is a steering value; the second element is a combined throttling and braking value. The choice of collapsing throttle and brake control into the same action was taken to avoid simultaneous braking and throttling and because throttling and braking both directly control the vehicle's acceleration.

The agent is rewarded only when it successfully stays behind the leader, meaning a reward is provided when the angle between the heading of the leader and the follower is in the range $[-\pi/4, \pi/4]$. When this condition is met, the agent receives the maximum reward if it keeps an optimal distance (within a prescribed tolerance) from the lead vehicle. In other words, the follower is rewarded when it stays in a sector of an annulus centered at the leader. When the vehicle is outside the desired area, the reward decreases hyperbolically. 
\begin{figure*}
    \centering
    \includegraphics[angle=90,width=0.6\linewidth]{Figs/MS_NN.png}
    \caption{{\small Sensor Fusion NN architecture.}}   
    \label{fig:NNarch}
\end{figure*}

The learning draws on information from several sensors, whose output is organized into two tuples. The first element is a 80 $\times$ 45 $\times$ 3 RGB image. The second is a vector of four values: the latitude and longitude difference between the leader and the follower; the heading according to the compass; and the speed at which the follower is approaching the leader. This multi-sensor observation required the NN architecture to incorporate an input composed of a 3D and a 1D tensor. The image is processed in a Convolutional Neural Network (CNN) as in~\cite{Mnih13}. Its output is then concatenated with the output of the one Fully Connected (FC) hidden layer Deep Neural Network (DNN) which takes the 1D Tensor as input. Their concatenated output is then processed by three FC hidden layers. The architecture of the model is shown in Fig.~\ref{fig:NNarch}.

RL-based training requires a very large number of iterations. Three decisions helped speed up the learning process: ($i$) the lead vehicles were not simulated, but only rendered at the correct location and orientation (as this has no bearing for sensing purposes); ($ii$) a reduced-order model of the HMMWV vehicle was used in the first stage of training, with the more computationally demanding full vehicle model substituted during the training process (see Fig.~\ref{fig:Rewards}) to further refine the NN parameters; and ($iii$) the learning process was accelerated using the OpenAI~\cite{Brockman16Gym} Baselines tool for environment parallelization, thus allowing several simulations running simultaneously to speed up the collection of the dataset samples.

%{\SBELcomment{NOTE: In the caption for Fig.~\ref{fig:Rewards} we talk about things that we don't touch upon in the text. Caption is hard to understand (except switching to full vehicle -- we do touch on this in the text).}} {\SBELcomment{SB: I got rid of some unnecessary details}}

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{Figs/rewards.pdf}
	\caption{Plot of the moving average of the sum of collected rewards with respect to the policy updates (updated every 1500 interactions). The vertical dashed line represents the switch from the reduced to the full HMMWV model.}
	\label{fig:Rewards}
\end{figure}

\paragraph{Testing of learned policy.}  The AA control policy derived in PyChrono and GymChrono was tested in SynChrono for various convoy setups while operating on three terrain types. The platoons were 3L+1F, 2L+2F, and 1L+3F. The terrains were rigid, SCM hard (silt-like), and SCM soft (snow-like). This leads to a set of nine platooning scenarios. In the following discussion, the leader and follower vehicles are numbered starting from the head of the convoy; for example, the order of the vehicles in a 1L+3F configuration is: Leader, Follower 1, Follower 2, Follower 3.

For each platooning scenario, data recorded from simulations included position, velocity, and acceleration for each of the four vehicles. In addition, a high definition camera sensor was attached to the last vehicle in the convoy in order to visualize the simulation. Still frames captured with this camera are shown in Fig.~\ref{fig:simscreenshots} and full-length videos of representative simulations are available online~\cite{simsPaperGVSETS2020}. Different ground textures and colors were used to further differentiate between the three terrain types (rigid, SCM-Hard, and SCM-Soft).
%
\begin{figure*}
    \centering
    \begin{subfigure}{0.33\textwidth}
        \includegraphics[width=\linewidth]{Figs/Demonstration/rigid_1i_3f.png}
        \caption{} \label{fig:rigid13}
    \end{subfigure}%
    \begin{subfigure}{0.33\textwidth}
        \includegraphics[width=\linewidth]{Figs/Demonstration/hard_1i_3f.png}
        \caption{} \label{fig:hard13}
    \end{subfigure}%
    \begin{subfigure}{0.33\textwidth}
        \includegraphics[width=\linewidth]{Figs/Demonstration/soft_1i_3f.png}
        \caption{} \label{fig:soft13}
    \end{subfigure}%
    \caption{Still frames from attached third person camera: (a) rigid terrain; (b) SCM-Hard terrain; (c) SCM-Soft terrain.}   
    \label{fig:simscreenshots}
\end{figure*}

Table~\ref{tab:vehiclepositions} shows top-down views of the convoy trajectories for each platooning scenario, with solid and dashed lines representing leader and follower trajectories, respectively.
As these images indicate, different simulation configurations lead to different reactions of the follower vehicles. The training process described in \S\ref{subsec:learningDemo} rewards a follower vehicle for being within a certain angle and distance of the leader; as a result, this allows for deviations between the paths of follower and leader vehicles as well as deviations in the speed at which leader and follower vehicles negotiate a certain path segment.

%
\begin{table*}
    \centering
    \caption{2D Positions of each vehicle in each simulation configuration}
    \label{tab:vehiclepositions}
    \begin{tabular} { cccc }
         & \hspace*{.75cm}3L$ + $1F& \hspace*{.75cm}2L$ + $2F& \hspace*{.75cm}1L$ + $3F \\
        %\hline 
        %\\ 
        {\rotatebox[origin=c]{90}{Rigid}}&
        \parbox[c]{2.02in}{\includegraphics[height=1.8in]{Figs/Demonstration/3f_1i_rigid_positions.png}}&
        \parbox[c]{2.02in}{\includegraphics[height=1.8in]{Figs/Demonstration/2f_2i_rigid_positions.png}}&
        \parbox[c]{2.02in}{\includegraphics[height=1.8in]{Figs/Demonstration/1f_3i_rigid_positions.png}}\\
        %\newline \\
        %\hline 
        %\\ 
        {\rotatebox[origin=c]{90}{SCM-Hard}}&
        \parbox[c]{2.02in}{\includegraphics[height=1.8in]{Figs/Demonstration/3f_1i_hard_positions.png}}&
        \parbox[c]{2.02in}{\includegraphics[height=1.8in]{Figs/Demonstration/2f_2i_hard_positions.png}}&
        \parbox[c]{2.02in}{\includegraphics[height=1.8in]{Figs/Demonstration/1f_3i_hard_positions.png}}\\
        %\newline \\
        %\hline 
        %\\
        {\rotatebox[origin=c]{90}{SCM-Soft}}&
        \parbox[c]{2.02in}{\includegraphics[height=1.8in]{Figs/Demonstration/3f_1i_soft_positions.png}}&
        \parbox[c]{2.02in}{\includegraphics[height=1.8in]{Figs/Demonstration/2f_2i_soft_positions.png}}&
        \parbox[c]{2.02in}{\includegraphics[height=1.8in]{Figs/Demonstration/1f_3i_soft_positions.png}}
        %\newline \\ 
        %\hline \\
    \end{tabular}
\end{table*}

In an effort to quantify the performance and robustness of the derived platooning policy, we present next results from a statistical analysis using ensemble convoy simulations. This study includes results from 128 independent simulations for each one of the three terrain types mentioned above. In order to allow relative comparisons among different terrain types as well as between follower positions in the convoy, all simulations used the 1L+3F convoy configuration and each one of the three sets of 128 simulation used the same set of randomly-generated trajectories. The performance metrics used in this analysis measure the path and speed deviation of a follower vehicle from that of the convoy's leader.

%%%  Path planning ----------------------------

The 128 distinct scenarios were created in a 100 m $\times$ 100 m swath of flat terrain. A script generated 30 randomly placed circular obstacles with sizes uniformly distributed between 5 m and 7 m. The obstacle positions were drawn from a uniform distribution with an additional restriction to allow overlap of no more than half their radius~\cite{MATLABBubbleBath}. Finally, to increase the complexity of the resulting paths and reduce the likelihood of a straight path, \textit{each} configuration included four additional fixed obstacles with radius of 8 m placed equidistant along the diagonal from start to end.

A path was designed to get from start to finish while avoiding the obstacles. 
The path start and end positions were fixed to the south-west and north-east corners, respectively. A sample obstacle field is shown on the left in Fig.~\ref{fig:synchronodemoexample}.
A Particle Swarm Optimization (PSO) algorithm~\cite{PSOObstacleAvoidance} was used to generate a shortest distance path connecting the start and end locations, as shown in the middle image of Fig.~\ref{fig:synchronodemoexample}. The path generation was constrained to produce trajectories that remained in the domain and did not intersect obstacles.

Each of the 128 resulting paths shown in Fig.~\ref{fig:synchronoallpaths} was then used as the prescribed trajectory for the path-follower PID-based lateral controller implemented on the leader vehicle, in conjunction with a PID-based longitudinal controller that prescribes a target speed linearly increasing in time (corresponding to a constant acceleration of 0.47 m/s$^2$).

To produce paths that are feasible for simulated HMMWV vehicles, the generation of the corresponding environment setup in SynChrono involved scaling obstacle meshes (rocks, trees, bushes) such that their bounding sphere allows 2 m of separation from the generated path centerline; in other words, in each of the 128 environments, the path prescribed for the leader vehicle ensures a minimum width of 4 m. An overhead view of the resulting SynChrono simulation environment is shown on the right in Figure \ref{fig:synchronodemoexample}.

%The quantitative analysis was carried out systematically as follows: 128 unique environments were created with 34 total obstacles in each, a \textit{Particle Swarm Optimization} (PSO) based algorithm calculated a possible trajectory through the environment from designated start and end points \cite{PSOObstacleAvoidance}, and then a simulation was run for each environment. In the simulation tests, the leader vehicles utilize PID-based lateral and longitudinal controllers to follow the predetermined path with a prescribed target speed (with the latter linearly increasing in time) and the follower vehicles utilize the trained network described previously.

%To produce paths feasible for a simulated HMMWV vehicle to follow, careful consideration was taken when generating each environment, its obstacles and the corresponding path through it. First, obstacles with radii of 5m to 7m were uniformly distributed across a 92 by 92 block of terrain centered at [0,0]. A check was made to ensure the obstacles only overlapped at a magnitude less than or equal to their radius \cite{MATLABBubbleBath}. Four obstacles with radii of 8m were also placed equidistant along the diagonal from start to end. This was done to increase the difficulty of the computed paths, eliminating the likelihood of straighter paths. Figure \ref{fig:obstaclegeneration} shows an example result from this obstacle generation. The positions and radii of the generated obstacles were then fed into the PSO algorithm which calculated a possible trajectory through the environment. During the computation of the candidate path, a cubic spline interpolant was calculated to verify certain conditions weren't being violated: the path didn't lead through any obstacles and that it stayed within the 92 by 92 block of terrain. The candidate path generated is shown in Figure \ref{fig:psopathgeneration}. This progression was then performed 128 times, where the obstacle positions and the candidate path represented as a bezier curve were stored to be read by \synchrono{} later. Figure \ref{fig:synchronoallpaths} shows all 128 computed paths and demonstrates the overall spread of which the candidate paths span the terrain.

%Once all 128 environments with obstacles and paths were generated, the policy was then tested with the 1L+3F configuration on the three terrain types: rigid, hard SCM, soft SCM. As a result, 384 total simulations were performed using the Euler computing cluster at the University of Wisconsin-Madison. For each simulation, the corresponding environmental factors were loaded in: obstacles and path to follow. As seen in Figure \ref{fig:psopathgeneration}, the generated path is extended close to the edge of some obstacles. When loaded into \chrono{}, the obstacle meshes were scaled so that their new radii allowed for 2m of separation from the prescribed path to the mesh. This was done to provided adequate space for the leader to drive through the environment without hitting an obstacle asset. An overhead view of the resulting simulation environment in \synchrono{} is shown in Figure \ref{fig:synchronodemoexample}.

% \begin{figure}
%     \centering
%     \includegraphics[width=\columnwidth]{Figs/StatisticalAnalysis/obst_gen_68.png}
%     \caption{Sample obstacle field.}   
%     \label{fig:obstaclegeneration}
% \end{figure}

% \begin{figure}
%     \centering
%     \includegraphics[width=\columnwidth]{Figs/StatisticalAnalysis/path_gen_68.png}
%     \caption{Sample PSO-based path planning.}   
%     \label{fig:psopathgeneration}
% \end{figure}

% \begin{figure}
%     \centering
%     \includegraphics[width=\columnwidth]{Figs/StatisticalAnalysis/path_gen_rigid_68.png}
%     \caption{Overhead view of the corresponding SynChrono environment setup.}   
%     \label{fig:synchronodemoexample}
% \end{figure}

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{Figs/StatisticalAnalysis/path_gen_progression.png}
    \caption{Sample obstacle field, PSO-based path planning, and the corresponding SynChrono environment setup.}   
    \label{fig:synchronodemoexample}
\end{figure*}

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{Figs/StatisticalAnalysis/all_possible_paths.png}
    \caption{Ensemble of the 128 paths used in the statistical analysis.}   
    \label{fig:synchronoallpaths}
\end{figure}

%%%  Metrics ----------------------------

In order to perform a statistical analysis of the performance of the platooning policy, we define a set of six performance metrics that measure the deviations of a follower vehicle from that of the convoy leader and encode both lateral path deviation and deviations in the vehicle speed at a given location along the leader's path.  These metrics are defined in such a way as to allow comparisons between the performance of followers at different positions in the convoy, as well as across the three different terrain types considered here.

To eliminate differences due to the fact that vehicles in a convoy are inherently staggered along the path, the evaluation of the performance metrics is based on a common path segment.  Figure~\ref{fig:metrics_path} illustrates this process for the same trajectory used as an example before. The sample results used in this description are results of a simulation on rigid terrain and consider the last vehicle in the convoy (Follower 3).  The start clip point is defined as the point on the follower path closest to the leader's initial location. Similarly, the end clip point is defined as the point on the leader's path closest to the follower's final location.  The resulting segment on the leader's path is then sampled at intervals of equal arc-length and the closest point on the follower's path to each such sampled point is identified. The distance between corresponding points on the leader and follower paths are then used to define a {\em follower path deviation} as function of distance traveled along the leader's path (see Fig.~\ref{fig:metrics_path_deviation}).  This allows us to define the first two performance metrics: $m_{p}^{avg}$, the average follower path deviation, and $m_{p}^{max}$, the maximum path deviation.

Next, we compare the vehicle speeds at corresponding points on the follower and leader paths as shown in the left plot of Fig.~\ref{fig:metrics_speed_deviation} from which we derive the absolute and relative speed errors (the latter being the speed difference scaled by the leader's speed at that location).  The underlying assumption in these definitions is that a follower's speed should match as closely as possible the speed of the leader vehicle at the same location on the path (and not at the same point in time); this is also why, when calculating the subsequent speed deviation metrics, we discard the values corresponding to the first 10 m of travel (to allow the vehicles to accelerate from rest to the desired convoy speed).  With these we define two more pairs of performance metrics: $m_{as}^{avg}$, $m_{as}^{max}$, for the average and maximum absolute speed deviation between leader and follower, and $m_{rs}^{avg}$, $m_{rs}^{max}$, for the average and maximum relative speed deviation of the follower.

\begin{figure}
    \centering
    \includegraphics[width=0.9 \columnwidth]{Figs/Metrics/path.png}
    \caption{Path deviation metrics calculation (rigid terrain, Follower 3). The segment of the leader path used in calculations for this particular scenario (rigid terrain) was $L = 158.2 \text{ m}$.}
    \label{fig:metrics_path}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.9\columnwidth]{Figs/Metrics/path_deviation.png}
    \caption{Deviation in path between Leader and Follower 3. The lateral deviation metrics for this particular scenario (rigid terrain) were $m_{p}^{avg} = 0.473 \text{ m}$, $m_{p}^{max} = 2.484 \text{ m}$.}   
    \label{fig:metrics_path_deviation}
\end{figure}

\begin{figure*}
    \centering
    \includegraphics[width=0.9\textwidth]{Figs/Metrics/speed_deviation.png}
    \caption{Deviation in speed between Leader and Follower 3.  The vehicle speeds are evaluated at the same location along the leader's path. The speed deviation metrics for this particular scenario (rigid terrain) where $m_{as}^{avg} = 0.505 \text{ m/s}$, $m_{as}^{max} = 2.154 \text { m/s}$ and $m_{rs}^{avg} = 12.2 \%$, $m_{rs}^{max} = 69.2 \%$.}   
    \label{fig:metrics_speed_deviation}
\end{figure*}

%%%  Statistical analysis ----------------------------

The six metrics defined above were evaluated for each of the three follower vehicles in each scenario in the three sets of 128 environments on rigid, SCM-Hard, and SCM-Soft terrain, respectively. The resulting statistics are presented as box-and-whisker diagrams in Figs.~\ref{fig:metricpath}, \ref{fig:metricspeed}, and \ref{fig:metricrelspeed}, providing measures of the variability in the three statistical populations without any assumption on their underlying statistical distribution (which is unknown, due to the manner in which the sample trajectories were constructed).  For each metric, on each terrain type and for each of the three follower vehicles in the 1L+3F configuration, these standard box plots provide information on their mean, second and third quartiles, as well as minimum and maximum values.

We assume that a perfect control policy for a follower vehicle would result in a convoy in which each follower vehicle runs exactly in the tracks of the vehicle preceding it and achieves the exact same speed at any given location.  Given their definition, this ideal case corresponds to zero values for each of the six performance metrics.

The results of the statistical analysis presented herein confirm the intuition that, in a 1L+3F configuration, a less than perfect control policy will lead to worse performance for the trailing follower vehicles and better performance on harder surfaces (especially taking into account that the training was performed exclusively on rigid terrain).  However, as the results for both path and speed deviation show, this is not the case when comparing performance on SCM-Hard and SCM-Soft terrains, with the latter showing consistent lower metrics values (i.e., better performance in terms of maintaining position in the convoy).

The explanation for this behavior is likely a combination of several factors.  First, even though the target speed profile for the leader vehicle was set identically for all three terrain types, the increased terrain resistance in the SCM-Hard and SCM-Soft cases resulted in the leader vehicle being unable to continuously increase its speed to the specified value; on both deformable terrain types, the vehicles were unable to shift in the higher gears and their speed limited to lower levels than on rigid terrain (an effect more pronounced on SCM-Soft terrain than on SCM-Hard). The ensuing overall lower convoy speed results in driving scenarios where the control policy can adapt better.  
%
Second, as seen in Fig.~\ref{fig:metrics_speed_deviation} and typical of all simulations conducted as part of this analysis, current deficiencies in the RL-based control policy result in relatively jerky motion of the follower vehicles and noisy speed profiles.  These spurious accelerations and decelerations are inhibited on the SCM-Soft due to the increased motion resistance.
%
Finally, the largest deviations in a follower's path (see Fig.~\ref{fig:metrics_path}) always occur at tight turns where the leader vehicle must go around an obstacle.  In these situations, the tendency of the control policy is to ``cut corners'' and thus direct the vehicle to increase steering input.  However, these control steering inputs are more difficult to follow in a deformable terrain soft enough to result in deep ruts, thus resulting in the follower vehicles more closely matching the leader's vehicle path around obstacles.

While the path planning procedure and the path-following PID-based control policy implemented for the leader vehicles ensures that a leader vehicle always avoids obstacles, this is not the case for the RL-based control policy implemented for the follower vehicles, which occasionally are unable to avoid an obstacle (in few situations, a follower vehicle, especially one in position 2 or 3, may end up going on the opposite side of an obstacle).  This behavior has multiple compounding causes, including the particular reward system used in the current training as well as configurations where perception of the leader vehicle is obstructed by an obstacle or the leader vehicle is out of the camera sensor's field of view while negotiating a tighter turn.
%
The paths of all vehicles on each terrain type are shown in Fig.~\ref{fig:pathsranksall}. The path information therein is used to gauge the robustness of the control policy in terms of obstacle avoidance. Figure~\ref{fig:obstacleshit} provides the cumulative statistics in terms of number of obstacles hit, over all ensemble simulations, for all three terrain types and for each of the three follower vehicles.  These results show the same relative performance trends observed before, with the trailing follower on SCM-Hard terrain exhibiting worst performance.

%Using the data collected during each of the 384 simulations, the aforementioned metric calculations were performed and their results recorded for subsequent analysis, of which is presented here.

%To gauge the overall performance of the policy, where the primary goal is that the follower drives in the tracks of its leader, the integrated error metric is most insightful. A better control policy would produce a smaller integrated error metric and when Figure \ref{fig:metricpathavg} is examined, it can be seen the SCM hard simulations produces disproportionally larger values. The comparative performance between rigid and SCM soft isn't as evident; however, the tests performed on rigid do behave better than SCM soft for the followers at depths 1 and 2, but approximately the same for follower 3. Furthermore, the spread for rigid is far less than that for SCM hard and SCM soft for follower 1 and 2. For follower 3, the spread between the rigid and SCM soft experiments is approximately equal, where as the SCM hard is quite larger. A possible explanation for the performance drop experienced by the tests on SCM hard could be attributed to the friction associated with a deformable terrain (i.e. there is more friction at the wheels, so the policy has difficulty reacting to this new stimuli). This hypothesis is contradicted by the SCM soft tests; however, a suitable explanation could be the fact the SCM soft is \textit{considerably} softer than the harder parameters and produces very deep ruts; hard enough, in fact, that it could mean the vehicle has more difficulty escaping the ruts of it's leader. These ideas are also evident in Figure \ref{fig:pathsranksall} and Figure \ref{fig:obstacleshit}, where they show the path taken and the number of obstacles hit by each vehicle, respectively.

%Similar to the integrated path error, the maximum deviation is also insightful when attempting to analyze how the policy performs. Most prominent for paths where tight corners are present, significant deviations from the leaders path could result in the follower clipping obstacles or failing to see its leader all together. As shown in Figure \ref{fig:metricpathmax}, rigid is consistently better than both the SCM experiments. One simple explanation for this discrepancy is the fact that vehicles drive faster on the rigid terrain, meaning the foresight to make a turn must be greater and vehicle ends up cutting corners less. In addition, the policy was trained on rigid terrain, which could represent a fault of this approach. Moreover, the SCM soft tests typically have maximum deviations less than that of the SCM hard tests. This can be explained again by the fact that the snow-like soft SCM parameters result in the vehicle staying in the tracks of its leader more often.

%As alluded to before, the speed of the SCM tests are far less than that for the rigid terrain trials. The friction associated with the deformable terrain is too great for the vehicle to reach its shift points. This results in decreased overall speeds for the vehicles simulated on soft and hard SCM. The soft SCM is also much softer than the hard SCM terrain, so its speed is also frequently smaller. This would mean that the integrated error between the speed of the follower and the leader would be less because of the lower speeds, seen in Figure \ref{fig:metricspeedavg}. This trend is also present in the maximum velocity deviation from the leader, shown in Figure \ref{fig:metricspeedmax}. Due to the simple fact that the SCM simulations run at lower speeds, the maximum speed deviation will be less.

% \begin{figure}
%     \centering
%     \includegraphics[width=\columnwidth]{Figs/StatisticalAnalysis/metric_path_avg.png}
%     \caption{Integrated error of the position between the followers and leaders path's.}   
%     \label{fig:metricpathavg}
% \end{figure}

% \begin{figure}
%     \centering
%     \includegraphics[width=\columnwidth]{Figs/StatisticalAnalysis/metric_path_max.png}
%     \caption{Maximum deviation metric between the followers and leaders path's.}   
%     \label{fig:metricpathmax}
% \end{figure}

% \begin{figure}
%     \centering
%     \includegraphics[width=\columnwidth]{Figs/StatisticalAnalysis/metric_speed_avg.png}
%     \caption{Integrated error between the followers and leaders velocity profile's.}   
%     \label{fig:metricspeedavg}
% \end{figure}

\begin{figure*}
    \centering
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\columnwidth]{Figs/StatisticalAnalysis/metric_path_avg.png}
        \caption{} \label{fig:metricpathavg}
    \end{subfigure}
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\columnwidth]{Figs/StatisticalAnalysis/metric_path_max.png}
        \caption{} \label{fig:metricpathmax}
    \end{subfigure}%
    \caption{Statistics of average and maximum follower path deviations.}  
    \label{fig:metricpath}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\columnwidth]{Figs/StatisticalAnalysis/metric_speed_avg.png}
        \caption{} \label{fig:metricspeedavg}
    \end{subfigure}
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\columnwidth]{Figs/StatisticalAnalysis/metric_speed_max.png}
        \caption{} \label{fig:metricspeedmax}
    \end{subfigure}%
    \caption{Statistics of average and maximum follower absolute speed deviations.}  
    \label{fig:metricspeed}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\columnwidth]{Figs/StatisticalAnalysis/metric_rel_speed_avg.png}
        \caption{} \label{fig:metricrelspeedavg}
    \end{subfigure}
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\columnwidth]{Figs/StatisticalAnalysis/metric_rel_speed_max.png}
        \caption{} \label{fig:metricrelspeedmax}
    \end{subfigure}%
    \caption{Statistics of average and maximum follower relative speed deviations.}  
    \label{fig:metricrelspeed}
\end{figure*}

% \begin{figure}
%     \centering
% \end{figure}

% \begin{figure}
%     \centering
%     \includegraphics[width=\columnwidth]{Figs/StatisticalAnalysis/metric_rel_speed_max.png}
%     \caption{Maximum error between the followers and leaders velocity profiles, scaled by the speed of the leader at each position along its path.}   
%     \label{fig:metricrelspeedmax}
% \end{figure}

\begin{figure*}
    \centering
    \begin{subfigure}{0.33\textwidth}
        \includegraphics[width=\linewidth]{Figs/StatisticalAnalysis/1L_3F_rigid_128_pos_v5.png}
        \caption{} \label{fig:rigid13pos}
    \end{subfigure}%
    \begin{subfigure}{0.33\textwidth}
        \includegraphics[width=\linewidth]{Figs/StatisticalAnalysis/1L_3F_hard_128_pos_v5.png}
        \caption{} \label{fig:hard13pos}
    \end{subfigure}%
    \begin{subfigure}{0.33\textwidth}
        \includegraphics[width=\linewidth]{Figs/StatisticalAnalysis/1L_3F_soft_128_pos_v5.png}
        \caption{} \label{fig:soft13pos}
    \end{subfigure}%
    \caption{Paths of every vehicle on each terrain type: (a) Rigid; (b) SCM-Hard; (c) SCM-Soft.}   
    \label{fig:pathsranksall}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{0.33\textwidth}
        \includegraphics[width=\linewidth]{Figs/StatisticalAnalysis/rigid_histogram.png}
        \caption{} \label{fig:rigidhistogram}
    \end{subfigure}%
    \begin{subfigure}{0.33\textwidth}
        \includegraphics[width=\linewidth]{Figs/StatisticalAnalysis/hard_histogram.png}
        \caption{} \label{fig:hardhistogram}
    \end{subfigure}%
    \begin{subfigure}{0.33\textwidth}
        \includegraphics[width=\linewidth]{Figs/StatisticalAnalysis/soft_histogram.png}
        \caption{} \label{fig:softhistogram}
    \end{subfigure}%
    \caption{Number of obstacles hit by each vehicle on the three terrain types: (a) Rigid; (b) SCM-Hard; (c) SCM-Soft.}   
    \label{fig:obstacleshit}
\end{figure*}

%\FloatBarrier

%% ============================================================ 
%% ============================================================ 
\section{CONCLUSIONS. FUTURE WORK}\label{s:conclusion}
This contribution discussed a Chrono-centric simulation platform designed to facilitate the design and testing of control policies for AAs operating in off-road conditions. The platform draws on a physics-based simulation engine; has templates for wheeled and tracked vehicles; enforces space and time coherence; allows for human-in-the-loop scenarios; provides sensor simulation capabilities; has a bridge to ROS; can simulate mobility on fully resolved, continuum, or SCM representations of the terrain; is open source; and is cluster deployable to support multi-AA mobility studies. This software framework is used here to design an RL-based control policy that allows AAs to follow in a convoy formation. The learning took place on rigid terrain but was demonstrated to work when deployed on AAs that operate on deformable SCM soils. The virtual environments used in testing differed in textures and colors from the ones used in the training, thus demonstrating robustness of the inferred policy that relies on inputs from an RGB camera sensor. Unsurprisingly, the fewer AAs in the platoon, the tighter it managed to follow a prescribed path. Looking ahead, we plan to speed up the SCM implementation; augment the sensing simulation support; improve scalability; and use this simulation infrastructure to derive new control policies for off-road AA mobility.

%% ============================================================

\section*{Acknowledgments}
The Euler supercomputer used in this work contains hardware procured through the Army Research Office DURIP instrumentation grant W911NF1810476. Support for terramechanics research is provided through US Army Research Office project W911NF1910431. Ongoing support for core {\chrono} development is provided by National Science Foundation project CISE1835674. Ongoing support for SynChrono development is provided by National Science Foundation project CPS1739869. Support for the development of \chronomod{Vehicle} was provided by U.S. Army GVSC grant W56HZV-08-C-0236. Support for the development of \chronomod{Sensor} and {\synchrono} has been provided by the SAFER-SIM program, which is funded through a grant from the U.S. Department of Transportation's University Transportation Centers Program (69A3551747131).

%% ============================================================
\bibliographystyle{unsrt}
\bibliography{Bib/refsAutonomousVehicles,Bib/refsMachineLearning,Bib/refsChronoSpecific,Bib/refsCompSci,Bib/refsDEM,Bib/refsFSI,Bib/refsGraphics,Bib/refsMBS,Bib/refsRobotics,Bib/refsSBELspecific,Bib/refsSensors,Bib/refsTerramech,Bib/refsDRL}

%% ============================================================
%\input{oldMaterial}
%% ============================================================

\end{document}

### 2020-08-25 2020/GVSETS_2020/GVSETSintermediate.tex
Added BookIta template, Eccomas DRL paper, GVSETS AV paper, CoRL (under review)

--- 2020-08-25 2020/GVSETS_2020/GVSETS.tex
@@
 \noindent 
-We describe a simulation environment that enables the design and testing of control policies for off-road mobility of autonomous agents.
-The environment is demonstrated in conjunction with the design and assessment of a reinforcement learning policy that uses sensor fusion and inter-agent communication to enable the movement of mixed convoys of conventional and autonomous vehicles.
+We describe a simulation environment that enables the designing and testing of control policies for off-road mobility of autonomous agents.
+Herein, the environment is used in conjunction with the design and demonstration of a reinforcement learning generated policy that uses sensor fusion and inter-agent communication to enable the movement of mixed convoys of conventional and autonomous vehicles.
 Policies learned on rigid terrain are shown to transfer to hard (silt-like) and soft (snow-like) deformable terrains.
@@
 Simulation is not a silver bullet as it has its limitations, first of all the issue of simulation-to-reality transfer \cite{sim2realGapEssex1995}, which pertains to the failure of control policies derived in simulation to work well in the real world.
-Furthermore, models are difficult to set up and calibrate, the validation process can be tedious and time consuming, and open source simulation tools that are both predictive and expeditious are not readily available.
+Furthermore, models are difficult to set up and calibrate, the validation process can be tedious and time consuming, and simulation tools that are both predictive and expeditious are not available for all types of analysis.
 This contribution addresses the third point.
-It describes a simulation environment whose stated purpose is to allow the practitioner to gain insights into the operation of AAs (robots and autonomous wheeled or tracked vehicles) in off-road conditions with an eye towards: improving mechanical designs of AAs; and, producing and testing control policies that govern the operations of the AAs.
-These are topical goals for the Army, see, for instance, \cite{jayakumarAutonomous2011,jayAbhiMarco2015,jayGorsich2018,jayGVSETS2019}, or, for a historical perspective on early contributions, \cite{frederickTARDECrobotics2014}.
+It describes a simulation environment whose stated purpose is to allow the practitioner to gain insights into the operation of AAs (robots and autonomous wheeled or tracked vehicles) in off-road conditions.
+If such a simulation platform is validated, it can be used to two ends: produce mechanical design of AAs; and generate and test control policies that govern the operations of the AAs.
 
-There are several ongoing efforts that seek to address the AA simulation issue.
+There are many other ongoing efforts that seek to address the AA simulation issue.
 In robotics, Gazebo~\cite{gazebo,koenig2004design} is a widely used 3D multi-robot simulator with dynamics.
@@
 Unity~\cite{unityGaming} and TORCS~\cite{torcsRacingSimulation2020}.
-For a survey of other solutions for on-road mobility the reader is referred to~\cite{surveyAVsimulationBerger2019,reviewPerception2019}.
+For a survey of other solutions for on-road mobility the user is referred to~\cite{surveyAVsimulationBerger2019}.
 
@@
 In its purpose, it is similar to the ANVEL-VANE environment as it seeks to simulate robots and wheeled/tracked vehicles operating in off-road conditions.
-Compared to the ANVEL-VANE solution, the Chrono environment is different in several respects: it is open source and available for unfettered use under a BSD3 license; it uses its own multi-physics engine; it is scalable and deployable on supercomputers, clusters, or multi-core architectures owing to its reliance on the MPI message passing standard for parallel computing~\cite{MPI-Forum-1994}; and is under active development.
+Compared to the ANVEL-VANE solution, the Chrono environment is different in several respects: it is open source and available for unfettered use under a BSD3 license, it uses its own multi-physics engine, it is scalable and deployable on supercomputers, clusters, or multi-core architectures owing to its reliance on the MPI message passing standard for parallel computing~\cite{MPI-Forum-1994}, and is under active development.
 Chrono is an ecosystem of modules and toolkits.
 It has support for rigid and flexible body dynamics (Chrono::Engine), fluid-solid interaction (Chrono::FSI), and granular dynamics (Chrono::Parallel and Chrono::Granular) applications.
-It has Python bindings, support for sensor simulation in Chrono::Sensor, an API for ROS~\cite{ROS-2009} bridging, as well as facilities for: rapid vehicle modeling via parameterized templates~\cite{ChronoVehicle2019}; control policy design with GymChrono; and scalable control policy testing with SynChrono.
-Chrono relies on GPU computing for fluid-solid interaction and select granular dynamics simulations, multi-core for most of the other modules, and MPI-enabled parallel computing for co-simulation when handling large terramechanics applications or collections of AAs.
+It has Python bindings, support for sensor simulation in Chrono::Sensor, an API for interaction with ROS~\cite{ROS-2009}, as well as facilities for: rapid vehicle modeling via parameterized templates~\cite{ChronoVehicle2019}; control policy design with GymChrono; and scalable control policy testing with SynChrono.
+Chrono relies on GPU computing for fluid-solid interaction and some granular dynamics simulations, multi-core for most of the other modules, and MPI-enabled parallel computing for co-simulation when handling large terramechanics applications or collections of AAs.
 Real-time simulation is not one of Chrono's priorities.
-Although for vehicle-on-rigid-terrain simulation it provides faster than real-time performance, there are numerous applications that lead to long run times in Chrono, e.g., deformable terrain mobility, nonlinear flexible body dynamics, fording scenarios, etc.
+Although as shown in this contribution, for simple robotics or vehicle-on-rigid-road simulation it provides faster than real-time performance, there are numerous applications that lead to long run times, e.g., deformable terrain mobility, nonlinear flexible body dynamics, fording scenarios, etc.
 
@@
 \label{sec:ML}
-Derived using an accurate simulation framework, control algorithms have been shown to bridge the sim-to-reality gap successfully~\cite{VISTASimMITRL, VirtualToRealRL}.
-The use of vehicles with Level 1 and Level 2 autonomy has grown considerably~\cite{J3016_201401, SAEHardwareInTheLoop} and the automotive industry is making major strides in the transition to Levels 3 and 4 autonomy~\cite{WayveRLSimToReal, TeslaSummon}.
-The use of simulation for on-road AVs is an area of intense research and development as this technology is seen as an important catalyst of the aforementioned transition.
-One active area of research is focused on sampling-based methods.
+Derived using an accurate and robust simulation framework, controls algorithms have shown to carry over to real-life successfully~\cite{VISTASimMITRL, VirtualToRealRL}.
+The use of vehicles with Level 1 and Level 2 autonomy has grown considerably~\cite{J3016_201401, SAEHardwareInTheLoop} and the automotive industry is making major strides towards producing Level 3 and Level 4 vehicles~\cite{WayveRLSimToReal, TeslaSummon}.
+Owing to advantages already described, the use of simulation in the on-road AV sector is an area of intense research and development.
+A common solution to the AV control problem relies on sampling-based methods.
 These approaches generate many candidate trajectories a vehicle can follow, selecting and executing the controls associated with the best candidate~\cite{UrbanAVBezierMPC, RRTRealTime}.
+Consequently, they require a discrete state space.
 Graph search methods are commonly associated with each trajectory.
-The approach is real-time challenged, since achieving robust results requires a high number of samples to be analyzed~\cite{TrajectoryPlanningMIQP}.
+In general, this approach is real-time challenged, as achieving robust results requires a high number of samples to be analyzed~\cite{TrajectoryPlanningMIQP}.
 Algorithms such as Dijkstra's, A-Star (A*), or the Rapidly-exploring Random Tree-Star (RRT*), sample the state space either deterministically or stochastically~\cite{RRTRealTime}.
@@
 
-Model Predictive Control (MPC) is another common AV control approach~\cite{Borrelli2005MPCbasedAT}.
+Model Predictive Control (MPC) is another common AV control approach in which a continuous state space is used~\cite{Borrelli2005MPCbasedAT}.
 Using a dynamic model of a vehicle, the MPC algorithm computes trajectories over the state space and determines an optimal trajectory using gradient-descending optimization techniques~\cite{TrajectoryPlanningMIQP, Liniger2017OptimizationBasedAR}.
@@
 
-The accuracy of the simulation platform plays a critical role both for MPC as well as sampling-based controllers.
+The accuracy of the simulation platform plays a critical role for MPC as well as sampling-based controllers.
 To adequately validate and subsequently verify a controller such as MPC, the simulation must be of high enough definition to carry over successfully to reality~\cite{AVTestingMethods}.
-When using the more pedestrian PID controller solutions, for which gains must be carefully selected, an inaccurate simulation platform could yield a poor design that leads to undesired consequences when deployed on a real vehicle.
+Likewise, for PID controller solutions, for which gains must be carefully selected, an inaccurate simulation platform could yield a poor design that leads to undesired consequences when deployed on a real vehicle.
 
@@
 
-Since its introduction~\cite{Mnih13}, DRL has proven successful in robotics applications~\cite{Levine2016,inHandManipulation2020}.
+Since its introduction~\cite{Mnih13}, DRL has proven successful in continuous control applications of robots~\cite{Levine2016,inHandManipulation2020}.
 At its core, DRL is an iterative learning process in which an \emph{agent} interacts with an \emph{environment}; at each iteration the agent collects an \emph{observation} (or \emph{state}), then performs an \emph{action} based on the previous observation and gets a \emph{reward} which is a measure of its performance.
-The goal of RL is to find a policy that maximizes the sum of the collected reward.
+The goal of RL is finding a policy that maximizes the sum of the collected reward.
+In a broad sense, a policy prescribes an action given a state.
 
-RL allows for complex control policies viable in unstructured and stochastic environments; it is model-free in that it does not require a model that predicts environment transients; and can learn from scratch.
+RL allows for complex control policies viable in unstructured and stochastic environments; it is model-free in that it does not require a model that predicts the environment transition; and can learn from scratch.
 RL's major flaw is its need for a massive amount of training data to infer a robust policy.
@@
 If $\pi$ is a Neural Network (NN), $\theta$ are the weights and biases of the NN, the state is the NN input, and the action is its output.
-Proximal Policy Optimization (PPO)~\cite{Schulman2017PPO} is one of the most widely used algorithms for continuous state and action environments and will be the algorithm of choice in this contribution.
+Proximal Policy Optimization (PPO)~\cite{Schulman2017PPO} is one of the most widely used algorithms for continuous state and action environments and is the algorithm of choice for this demonstration.
 PPO is a policy gradient algorithm whose goal is to optimize a stochastic policy.
@@
 First, it is used to produce the data needed to design a control policy.
-Second, it is used for testing purposes.
-To this end, it exposes the control policy produced in a model-based or model-free approach to a battery of tests to gauge its correctness and robustness.
+Second, it is used for testing purposes; i.e., exposes the control policy that was produced in a model-based or model-free approach to a battery of tests to gauge its correctness and robustness.
 This section outlines the components of this Chrono-centric simulation environment.
 The emphasis is placed on four components: Chrono::Sensor, PyChrono, GymChrono and SynChrono, that anchor the AA design and testing process and have not been discussed elsewhere in detail.
-More established Chrono components or functionality will be touched upon in passing; more details are provided in  \cite{chronoOverview2016,ChronoVehicle2019,ChronoSCM2019}.
+Other Chrono components or functionality will be touched upon in passing since they have been described elsewhere \cite{chronoOverview2016,ChronoVehicle2019,ChronoSCM2019}.
 
@@
 The software is portable and can be built on different platforms, under different operating systems, and using various compilers.
-Chrono is actively developed, has a continuous integration process, an active user forum, and is managed through GitHub~\cite{projectChronoGithub}.
-It's latest release is 5.0.1, available as of March 2020.
+Chrono is actively developed, has a continuous integration process, an active user forum, and is managed through GitHub~\cite{projectChronoGithub}; it's latest release is 5.0.1, available as of March 2020.
 Chrono is used by academic, industrial, and government research and development groups and projects.
 
-\paragraph{Chrono::Vehicle.} Chrono::Vehicle~\cite{ChronoVehicle2019} is a specialized Chrono module that makes available a collection of templates (fully parameterized models) for various topologies of both wheeled and tracked vehicle subsystems.
-It provides facilities for modeling rigid, deformable, and granular terrain; support for closed-loop and interactive driver models; and run-time and off-line visualization of simulation results.
+\paragraph{Chrono::Vehicle.} Chrono::Vehicle~\cite{ChronoVehicle2019} is a specialized Chrono module which provides a collection of templates (parameterized models) for various topologies of both wheeled and tracked vehicle subsystems.
+It also provides support for modeling of rigid, deformable, and granular terrain, support for closed-loop and interactive driver models, and run-time and off-line visualization of simulation results.
 Chrono::Vehicle leverages and works in tandem with other Chrono modules for run-time visualization or finite element, granular dynamics, and parallel computing support.
 
-Chrono::Vehicle provides vehicle \textit{subsystem} templates for tires, suspensions, steering mechanisms, drivelines, sprockets, track shoes, etc.; templates for external systems such as powertrains, drivers, and terrain models; and additional utility classes and functions for vehicle visualization, monitoring, and collection of simulation results.
+Chrono::Vehicle provides a comprehensive set of vehicle subsystem templates (for tires, suspensions, steering mechanisms, drivelines, sprockets, track shoes, etc.), templates for external systems (powertrains, drivers, terrain models), and additional utility classes and functions for vehicle visualization, monitoring, and collection of simulation results.
 As a middleware library, Chrono::Vehicle requires the user to provide C++ classes for a concrete instantiation of a particular template.
 An optional Chrono library provides complete sets of template instantiations for several concrete ground vehicles, both wheeled and tracked, which can serve as examples for developing more customized vehicle models.
-An alternative mechanism for defining concrete instantiation of vehicle system and subsystem templates is based on input specification files in the JSON format \cite{json}.
+An alternative mechanism for defining concrete instantiation of vehicle system and subsystem templates is based on input specification files in the JSON format.
 For additional flexibility and to allow integration of third-party software, Chrono::Vehicle is designed to permit either monolithic simulations or co-simulation where the vehicle, powertrain, tires, driver, and terrain/soil interaction can be simulated independently.
@@
 Chrono::Vehicle provides several classes of terrain and soil models, of different fidelity and computational complexity, ranging from rigid, to semi-empirical Bekker-Wong type models, to complex physics-based models based on either a granular or finite-element based soil representation.
-For simple terramechanics simulations, Chrono::Vehicle implements a customized Soil Contact Model (SCM), based on Bekker theory, with extensions to allow non-structured triangular grids, adaptive mesh refinement, and incorporation of bulldozing effects~\cite{ChronoSCM2019}.
+For simple terramechanics simulations, Chrono::Vehicle provides a customized implementation of the Soil Contact Model (SCM), based on Bekker theory, with extensions to allow non-structured triangular grids, adaptive mesh refinement, and incorporation of bulldozing effects~\cite{ChronoSCM2019}.
 Second, Chrono provides an FEA continuum soil model based on multiplicative plasticity theory with Drucker-Prager failure criterion and  specialized brick elements.
@@
 
-\paragraph{Chrono::Sensor.} Chrono::Sensor provides sensor simulation support for software-in-the-loop testing.
-Cameras, lidar, GPS, and IMU sensors can be placed within a Chrono simulation to generate synthetic data based on user-defined sensor parameters and attributes of the virtual world hosting the AA simulation experiment.
+\paragraph{Chrono::Sensor.} Chrono::Sensor is a specialized module of Chrono that enables the modeling and simulation of sensors for simulating robots and autonomous vehicles for software-in-the-loop testing.
+This module allows for cameras, lidar, GPS, and IMU sensors to be placed within a Chrono simulation to generate synthetic data based on user-defined sensor parameters.
 The goal of the module is to allow realistic data generation based on sensor characteristics such as noise, distortion, and lag.
 Sensors can be attached to objects within the simulation and configured to match corresponding real sensors.
-For modeling convenience, sensors can be defined through a JSON file \cite{json}.
-Additionally, custom sensors and post-processing filtering can be implemented, leveraging the existing render framework or physics interface.
+For ease of modeling, sensors can be specified through a JSON file format.
+Additionally, custom sensors and post-processing filtering can be implemented leveraging the existing render framework or physics interface.
 
-For interoceptive sensing (GPS and IMU), the module utilizes the internally computed physical quantities from the Chrono system and can augment this ground truth with drift, Gaussian noise, lag, and filtering characteristics from finite collection time.
+For interoceptive sensing (GPS and IMU), the module utilizes the internally computed physical quantities from the Chrono system and can augment this ground truth with drift, Gaussian noise, lag, and collection time.
 For exteroceptive sensors that provide information about scene characteristics (camera and lidar), Chrono::Sensor leverages hardware accelerated ray tracing through the OptiX library~\cite{optixNVIDIA} and implements physically based rendering techniques.
@@
 For camera, lens models and post-processing noise augmentation are supported, with an interface to extend or implement custom models.
-For lidar, the framework expands on work from~\cite{goodin2018enabling} to provide a beam divergence model that supports multiple modes of lidar return and reduced intensity during partial beam reflectance.
-The camera and lidar can also be parameterized by update rate, time over which to collect data, and lag.
+For lidar, the framework expands on work from~\cite{goodin2018enabling} to provide a beam divergence model that can support multiple modes of lidar return and reduced intensity during partial beam reflectance.
 
@@
 All sensors and capabilities are written in C++, but can also be accessed from Python through the PyChrono interface.
-The entire module can be run headless without the requirement of a render context, allowing for ease of deployment in machine learning applications on remote servers (in the cloud).
+The entire module can be run headless without the requirement of a render context, allowing for ease of deployment in machine learning applications on remote servers.
 
-\paragraph{PyChrono.} While the main Chrono API is expressed in C++, in recent years a concerted effort was dedicated to providing automatically-generated Python wrappers for much of the Chrono functionality.
-The purpose was twofold: to provide a lower-entry point to Chrono simulations for users less familiar with C++; and, provide a bridge to various machine learning platforms, e.g.
-TensorFlow \cite{tensorflow}, PyTorch \cite{paszke2017PyTorch}, Theano \cite{theano2016}, and CAFFE \cite{caffe2014}.
+\paragraph{PyChrono.} While the main Chrono API is C++, in recent years a concerted effort was dedicated to providing automatically-generated Python wrappers for much of the Chrono functionality.
+The main rationale for this was providing a lower-entry point to modeling and simulation with Chrono for users less familiar with C++ but also, as importantly, providing an interface to various machine learning platforms, such as TensorFlow \cite{tensorflow}, PyTorch \cite{paszke2017PyTorch}, Theano \cite{theano2016}, and CAFFE \cite{caffe2014}.
 
-The Python wrapping relies on using automated technology provided by SWIG \cite{swig2003}, which generates the interface between Python user-code and the underlying Chrono C++ libraries.
-Presently, a large set of Chrono functionality is exposed to Python users, including the core multibody and FEA module, the interface to CAD systems (like SolidWorks), run-time visualization with Irrlicht, etc.
+The Python wrapping relies on using automated technology provided by SWIG, which generates the interface between Python user-code and the underlying Chrono C++ libraries.
+As of today, a large set of Chrono functionality is exposed to Python users, including the core multibody and FEA module, the interface to CAD systems (like SolidWorks), run-time visualization with Irrlicht, etc.
 In particular, full support is available for both modeling, simulation, and visualization of wheeled ground vehicles and use of the Chrono::Vehicle existing wheeled vehicle models, as well as using the sensor models provided by Chrono::Sensor.
@@
 These environments inherit from OpenAI Gym classes.
-As such, they can be used out of the box with any algorithm or DRL framework made for gym environments.
-They can also draw on gym's environment parallelization for learning acceleration.
+As such, they can be used out of the box with any algorithm or DRL framework made for gym environments; and, they can draw on gym's environment parallelization for learning acceleration.
 
 \paragraph{SynChrono.} SynChrono is a software component that uses Chrono to implement a distributed-memory execution model when simulating scenarios that include multiple AAs.
-By leveraging the Message Passing Interface (MPI) standard~\cite{mpi-3.0}, SynChrono can manage multiple instances of Chrono running together \textit{one} mobility analysis on a supercomputer, cluster, or multi-core setup thus supporting the scalable and distributed simulation of multiple agents (robots, tracked vehicles, wheeled vehicles, etc.) The paradigm embraced is that of running the dynamics of one AA as one MPI rank, with the ranks/AAs communicating through MPI messages to maintain space and time coherent state for all agents participating in the study.
+It leverages the Message Passing Interface (MPI) standard~\cite{mpi-3.0}, which allows the toolset to have multiple instances of Chrono run simultaneously on a supercomputer, cluster, or multi-core setup to allow for the distributed simulation of multiple agents (robots, tracked vehicles, wheeled vehicles, etc.) The paradigm embraced is that of running one Chrono agent simulation as one MPI rank, with multiple ranks communicating through MPI messages to maintain a space and time coherent solution for all agents participating in the study.
 As an example, if there are two agents, SynChrono makes it possible to synchronously run the two agents on two different compute nodes in a supercomputer.
-By the same token, if there are 50 agents and 50 compute nodes in a cluster, SynChrono provides the infrastructure to keep the 50 agents operating in a coherent (timewise and spacewise) virtual world.
+By the same token, if there are 50 agents and 50 compute nodes in a cluster, SynChrono provides the compute infrastructure to keep the 50 agents operating in a coherent (timewise and spacewise) virtual world.
 Although each agent represents a Chrono simulation, SynChrono makes it possible for ruts generated by agent 27 to be picked up on a camera sensor on agent 31 if the camera points to agent 27.
@@
 The structure of SynChrono's MPI framework is shown in Fig.~\ref{fig:mpischematic}.
-SynChrono manages multiple AAs as multiple processes via as many MPI ranks.
-Each AA runs in its own SynChrono process (an MPI rank) and interfaces with its dedicated control stack for software-in-the-loop control.
+Each agent runs in its own MPI rank and interfaces with its dedicated control stack for software-in-the-loop control.
 The control stack is fed synthetic data generated by Chrono::Sensor and acts upon the environment through Chrono::Vehicle control inputs (throttle, steering, braking).
-The control algorithm for each agent is also configurable and can vary from complex algorithms that fuse sensor feeds/data streams, to controls based on empirical models, and on to pre-recorded driver inputs from a human or human-driven control in scenarios that are simple enough to allow real-time simulation.
-SynChrono supports human-in-the-loop experiments as well.
+This ``one agent-to-one Chrono system'' mapping ensures scalability; a monolithic Chrono system could not deliver this, a point demonstrated in section \S\ref{sec:demoTechnology}.
+The control algorithm for each agent is also configurable and can vary from complex algorithms that fuse a variety of sensor data streams, to controls based on empirical models, and on to pre-recorded driver inputs from a human or human-driven control in scenarios that are simple enough to allow real-time simulation (SynChrono supports human-in-the-loop experiments).
 
@@
 	\centering
-	\includegraphics[width=\columnwidth]{Figs/syn_cars_diagram.png}
+	\includegraphics[width=\columnwidth]{Figs/MPI-schematic.png}
 	\caption{{\small Schematic of the SynChrono framework.
-Dynamics simulations are done in separate Chrono systems and the outcome of the dynamics simulation is synchronized between ranks using MPI.}}
+Decisions are passed to the Chrono system for dynamics simulation and the outcome of the dynamics simulation is synchronized between ranks using MPI.}}
 	\label{fig:mpischematic}
@@
 
-Each SynChrono process is responsible for the dynamics of a single agent.
-At a slow frequency (relative to the simulation time-step), all SynChrono processes communicate via the SynChrono daemon to exchange state information.
-State information is intended to be brief, sufficient to enable a SynChrono process to reconstruct a ``ghost'' version of outside agents in its own world for visualization and sensing purposes.
+Each MPI rank is responsible for the dynamics of a single agent and simulates those dynamics as if that agent were the only agent in the world.
+At a slow frequency (relative to the simulation time-step), all ranks communicate with a central controller to exchange state information.
+State information is intended to be brief, communicating the bare minimum of information to enable other ranks to reconstruct a visual-only version of the agent in their own world.
 In an example where each agent is a vehicle, the state information consists of the vehicle location and orientation along with pose information for each wheel.
 This information is packaged via the FlatBuffers serialization library \cite{flatbuffers}.
+This allows a rank receiving this state information to construct a visual copy in their world for visualization or sensing purposes.
+SynChrono can add additional agents to the simulation at almost no extra cost, the only overhead being a higher volume of messages moving on the network and through the central controller.
 
-One limitation of the implementation is that currently two agents running as two SynChrono processes cannot crash with each other or participate in an operation that couples their dynamics, e.g.
+One limitation of the implementation is that currently two agents running as two different Chrono processes (ranks) cannot crash with each other or participate in an operation that couples their dynamics, e.g.
 lifting together a heavy object.
 Such a scenario should be run in Chrono, since the coupling is too tight to be handled by SynChrono.
-This will make the simulation longer to run since more agents will have to be handled within one Chrono process.
-However, if the agent coupling happens via sensing or the virtual world, e.g., one agent sensing another one, or one vehicle crossing over and being jolted by the ruts left by a different agent, then SynChrono can be relied on, thus ensuring scalability.
-The goal of SynChrono is to have virtually constant scaling up to a high threshold where the data passing overhead becomes the simulation bottleneck.
+This will make the simulation longer to run since more agents will have to be handled within one rank (Chrono process).
+However, if the agent coupling happens via sensing or the virtual world, e.g., one agent sensing another one, one vehicle crossing over and being jolted by the ruts left by a different agent, then SynChrono can be relied on, thus ensuring scalability.
+The goal of SynChrono is to have virtually constant scaling up to a high threshold where high message volume dominates over the dynamics computation.
 
@@
 control inputs).
-The control stack is independent of the SynChrono platform (e.g.
+The control stack is independent of the SynChrono platform (i.e.
 a bridge has been developed for ROS \cite{ROS-2009}), and can be tested with inputs replicating those from reality, such as sensor and/or V2X communication data.
@@
 \label{sec:demoTechnology}
-All simulation scenarios considered in this section use a Chrono::Vehicle HMMWV model.
-Chrono::Vehicle was benchmarked as part of the Next Generation-NRMM (NG-NRMM) exercise \cite{NATObenchmark2018}; Chrono::Vehicle-specific benchmark findings are detailed in \cite{TR2017-05NRMM-wheeled,TR2016-08NRMM-tracked}.
-Note that: the NG-NRMM benchmarking effort involved Chrono::Vehicle but no the particular model used in this study; and, all results reported herein were obtained using a simulation time step of $\Delta t=2 \times 10^{-3}$ seconds, both for rigid and deformable terrain.
-This time step information is relevant when discussing real-time performance and scalability aspects.
 
 \subsection{SynChrono scaling analysis}\label{sec:syn-scaling-analysis} 
-SynChrono uses $N+1$ processes executing on a cluster/supercomputer to simulate the dynamics of $N$ agents handled as $N$ independent Chrono simulations.
-Each Chrono simulation is a process; the extra process is involved in maintaining the time and space coherence for the virtual world shared by the $N$ AAs.
-The SynChrono daemon executes as the extra process and it coordinates via MPI the dynamics of the $N$ agents.
+SynChrono is an environment that uses $N+1$ processes executing on a cluster/supercomputer to simulate the dynamics of $N$ agents handled as $N$ independent Chrono simulation.
+Each Chrono simulation is a process; the extra compute node is involved in maintaining the time and space coherence for the virtual world shared by the $N$ agents.
+The so-called SynChrono daemon executes as the extra process and it manages through MPI the dynamics of the $N$ agents.
 The numerical experiments described here answer the following questions: ($i$) How does the time to complete a simulation change as $N$ increases?
@@
 ($iii$) How fast is it when an SCM deformable terrain model is considered in the simulation scenario?
-The Linux cluster used exposed 15 nodes.
-Thus, one set of experiments was conducted with up to $N=14$ AAs, using rigid, SCM hard (silt-like), and SCM soft (snow-like) terrain, with one agent per compute node.
-However, the experiment also included scenarios with up to 56 AAs.
-In this case, four AAs ran on the same compute node.
-On the upside, this allowed more AAs to participate in the benchmark; on the downside, the simulation times went up since the hardware allocation per AA went down by a factor of four.
+These experiments were conducted with up to $N=14$ agents, using rigid, SCM hard (silt-type), and SCM soft (snow-like) terrain.
 
@@
 This means that two agents should not come close enough that they deform the same vertices during the same synchronization period, but this is just as restrictive as SynChrono's assumption that any two agents will not interact by crashing.
-While Chrono's SCM implementation allows for meshes that auto-refine, this is currently not supported in SynChrono due to the significant increase in algorithmic complexity for relatively little gain in performance.
+While Chrono's SCM implementation allows for meshes that auto-refine, this is currently not supported in SynChrono due to the significant increase in algorithmic complexity for little gain in performance.
 The main concern for scalability is message size, since for synchronizing SCM terrain there is about 15 times as much data to send for each agent relative to the rigid terrain case.
 
-The scenario discussed herein is that of many vehicles crossing perpendicularly on a rectangular patch of SCM terrain, see Fig.~\ref{fig:scalingenv} and online movies \cite{simsPaperGVSETS2020}.
-In this setup, one can easily scale up the number of vehicles and verify that the SCM terrain deformation is properly synchronized across multiple ranks.
-The scaling metric used was the Real Time Fraction (RTF), representing the amount of wall-time required to finish a simulation divided by the amount of time simulated.
-Running in real-time corresponds to a factor of 1, while slower than real-time corresponds to factors larger than one.
-The tests were run on the Euler computing cluster at the University of Wisconsin-Madison.
-Each node has an eight core, Broadwell-generation Intel processor; inter-node communication is facilitated via a Gigabit Ethernet interconnect.
-
-In a first sub-experiment, we assign each SynChrono process (which runs one vehicle) to distinct cluster nodes.
-The scaling analysis results are presented in Fig.~\ref{fig:scmscaling}.
-The plot shows on the $y$-axis the $ log_{10} $ of the RTF.
-Thus, any dot below the horizontal $x$-axis is associated with an experiment that ran faster than real time.
-Simulations on the rigid terrain ran with an RTF of 0.75 and simulations on either type of SCM terrain (hard or soft) ran with an RTF of approximately 110.
-The RTF value of 110 is independent of the of SCM soil parameters (i.e.
-soft vs.
-hard), but is highly dependent on the processor performance, MPI implementation, compiler, and compilation optimization.
-These results confirm weak scaling behavior: no significant performance penalty is observed for including additional agents in the simulation when each SynChrono process is assigned to a different compute node.
+The scenario discussed herein is that of many vehicles crossing perpendicularly on a rectangular patch of SCM terrain, shown in Fig.~\ref{fig:scalingenv}.
+This scenario allows for easily scaling up the number of vehicles and verifying that the SCM terrain deformation is properly synchronized across multiple ranks.
 
@@
 
-Using the same scenario as above, scaling analyses were also conducted for a larger number of agents on rigid terrain to ensure that the same weak scaling continued to hold.
-To this end, two other sub-experiments were conducted: two SynChrono processes were deployed per node; and, four SynChrono processes were deployed per node.
-Given the number of nodes made available on the cluster (15), for two-per-node, SynChrono handled 28 AAs; for four-per-node, it managed a joint simulation of 56 AAs.
-Timing results are shown in Fig.~\ref{fig:scmscalingrigid}, which shows that 56 AAs run in slightly more than 1.8$ \times $ real time.
-Given a supercomputer or larger cluster, SynChrono opens the possibility of simulating swarms of AAs.
-%Due to peculiarities in the network configuration, not all tests could be run with the same MPI implementation which had the side-effect of demonstrating that simulation performance was in general very dependent on the MPI implementation.
+The scaling metric used was the Real Time Fraction (RTF), representing the amount of wall-time required to finish a simulation divided by the amount of time simulated.
+Running in real-time corresponds to a factor of 1, while slower than real-time corresponds to factors larger than one.
+The tests were run on the Euler computing cluster at the University of Wisconsin-Madison.
+Each node has an eight core, Broadwell-generation Intel processor; inter-node communication is facilitated via a Gigabit Ethernet interconnect.
+While in practice ranks would sit on as few nodes as possible to better share data, we assigned each rank to distinct cluster nodes to focus on the communication overhead for the worst-case scenario.
+The scaling analysis results are presented in Fig.~\ref{fig:scmscaling}.
+The plot shows on the $y$-axis the $ log_{10} $ of the RTF.
+Thus, any dot below the horizontal $x$-axis is associated with an experiment that ran faster than real time.
+Simulations on the rigid terrain ran with an RTF of 0.2 and simulations on either type of SCM terrain (hard or soft) ran with an RTF of approximately 80.
+The RTF value of 80 is independent of the of SCM soil parameters (i.e.
+soft vs.
+hard), but is highly dependent on the processor performance, compiler, and compilation optimization.
+These results confirm weak scaling behavior: no significant performance penalty is observed for including additional agents in the simulation, even when each MPI rank is assigned to a different compute node.
 
@@
     \label{fig:scmscaling}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\columnwidth]{Figs/syn_SCM_many.png}
-    \caption{\synchrono{} scaling analysis on rigid terrain for a large number of agents.}   
-    \label{fig:scmscalingrigid}
 \end{figure}
@@
 \label{subsec:learningDemo}
-
-%{\SBELcomment{We should make it clear what the vehicles know, sense, communicate, etc. I don't think we do justice to this at this point.}}
-%
-
-Chrono helps with two tasks: learning a control policy, and testing a policy that was designed in Chrono or brought from outside.
+Chrono can help with two tasks: learning a control policy, and testing a policy that was designed in Chrono or brought from outside.
 In this example, PyChrono and GymChrono are used to design a policy.
 SynChrono is subsequently used to test it.
-The RL-based learning is done on rigid terrain using a nondescript texture.
-The goal is to enable a vehicle to move as part of a convoy.
-To test it, the policy is deployed on vehicles that are part of a four-vehicle convoy driving on rigid or SCM deformable terrain.
+The RL-based learning is done on rigid terrain and its goal is to enable a vehicle to move as part of a convoy.
+To test it, the policy is deployed on vehicles that are part of a four-vehicle convoy driving on rigid of SCM deformable terrain.
 Up to three of the convoy vehicles use this policy while driving in a platoon.
 Thus, the possible scenarios are: three lead vehicles and one following vehicle (3L+1F), two lead and two followers (2L+2F), and one lead and three followers (1L+3F).
-The \textit{lead} vehicles are programmed to follow a path defined by way-points; for all purposes, these can be considered human driven.
+The lead vehicles are programmed to follow a path defined by way-points; for all purposes, these can be considered human driven.
 A follower vehicle is autonomous and uses the learned policy to follow the vehicle in front of it.
 In doing so, it should not crash in the vehicle ahead of it, and avoid hitting obstacles in the vicinity of the path.
-To this end, it relis on a camera sensor, location acquired through GPS sensing, and compass heading.
-Note that communication allows a vehicle to find out the GPS location and velocity of the vehicle in front.
-Given that four vehicles are involved in this platooning experiment, SynChrono is subsequently used to test the policy to reduce simulation times.
-Indeed, this validation could be run in Chrono::Engine but would roughly require four times longer to complete it.
-The salient points of this experiment are as follows: although vehicles are run in different SynChrono processes, there is time and space coherence between them to the point where the vehicles sense each other; the learning occurs using rigid terrain with nondescript texture but the policy is tested on deformable terrain that is white (snow-like); this is an end-to-end policy that uses sensor data fusion to control both the steering and acceleration/braking of the vehicle.
 
@@
 Subsequently, the agent must perform an action prescribed by the ML algorithm in order to maximize the reward collected.
-The action is a two element array: the first is a steering value; the second element is a combined throttling and braking value.
+The action is a 2-dimensional array: the first element is a steering value; the second is a combined throttling and braking value.
 The choice of collapsing throttle and brake control into the same action was taken to avoid simultaneous braking and throttling and because throttling and braking both directly control the vehicle's acceleration.
@@
 The terrains were rigid, SCM hard (silt-like), and SCM soft (snow-like).
-This leads to a set of nine platooning scenarios.
-In the following discussion, the leader and follower vehicles are numbered starting from the head of the convoy; for example, the order of the vehicles in a 1L+3F configuration is: Leader, Follower 1, Follower 2, Follower 3.
+This leads to a set of nine \textit{platooning scenarios} in which the AA policy was tested.
 
-For each platooning scenario, data recorded from simulations included position, velocity, and acceleration for each of the four vehicles.
-In addition, a high definition camera sensor was attached to the last vehicle in the convoy in order to visualize the simulation.
-Still frames captured with this camera are shown in Fig.~\ref{fig:simscreenshots} and full-length videos of representative simulations are available online~\cite{simsPaperGVSETS2020}.
-Different ground textures and colors were used to further differentiate between the three terrain types (rigid, SCM-Hard, and SCM-Soft).
-%
 \begin{figure*}
@@
     \end{subfigure}%
-    \caption{Still frames from attached third person camera: (a) rigid terrain; (b) SCM-Hard terrain; (c) SCM-Soft terrain.}   
+    \caption{Still frames from attached third person camera: (a) rigid terrain; (b) hard SCM terrain; (c) soft SCM terrain.}   
     \label{fig:simscreenshots}
@@
 
-Table~\ref{tab:vehiclepositions} shows top-down views of the convoy trajectories for each platooning scenario, with solid and dashed lines representing leader and follower trajectories, respectively.
-As these images indicate, different simulation configurations lead to different reactions of the follower vehicles.
-The training process described in \S\ref{subsec:learningDemo} rewards a follower vehicle for being within a certain angle and distance of the leader; as a result, this allows for deviations between the paths of follower and leader vehicles as well as deviations in the speed at which leader and follower vehicles negotiate a certain path segment.
+For each platooning scenario, data recorded to evaluate the performance of the convoy included position, velocity, and acceleration for each of the four vehicles.
+In addition, a high definition camera sensor was attached to the last vehicle in the convoy in order to visualize the simulation.
+Still frames captured with this camera are shown in Fig.~\ref{fig:simscreenshots}.
+Full length videos of these experiments are available online \cite{simsPaperGVSETS2020}.
+Different ground textures were used to further differentiate between the three terrain types.
+Dirt, mud and snow textures were each used for the rigid, hard SCM and soft SCM terrains, respectively.
+In Table~\ref{tab:vehiclepositions}, top-down views of the convoy trajectories are shown for each platooning scenario.
+Each leader is represented as a solid line and each follower is seen as a dashed line.
 
-%
 \begin{table*}
@@
         %\\ 
-        {\rotatebox[origin=c]{90}{Rigid}}&
+        {\rotatebox[origin=c]{90}{Rigid Terrain}}&
         \parbox[c]{2.02in}{\includegraphics[height=1.8in]{Figs/Demonstration/3f_1i_rigid_positions.png}}&
@@
         %\\ 
-        {\rotatebox[origin=c]{90}{SCM-Hard}}&
+        {\rotatebox[origin=c]{90}{Hard SCM}}&
         \parbox[c]{2.02in}{\includegraphics[height=1.8in]{Figs/Demonstration/3f_1i_hard_positions.png}}&
@@
         %\\
-        {\rotatebox[origin=c]{90}{SCM-Soft}}&
+        {\rotatebox[origin=c]{90}{Soft SCM}}&
         \parbox[c]{2.02in}{\includegraphics[height=1.8in]{Figs/Demonstration/3f_1i_soft_positions.png}}&
@@
 
-In an effort to quantify the performance and robustness of the derived platooning policy, we present next results from a statistical analysis using ensemble convoy simulations.
-This study includes results from 128 independent simulations for each one of the three terrain types mentioned above.
-In order to allow relative comparisons among different terrain types as well as between follower positions in the convoy, all simulations used the 1L+3F convoy configuration and each one of the three sets of 128 simulation used the same set of randomly-generated trajectories.
-The performance metrics used in this analysis measure the path and speed deviation of a follower vehicle from that of the convoy's leader.
-
-%%%  Path planning ----------------------------
-
-The 128 distinct scenarios were created in a 100 m $\times$ 100 m swath of flat terrain.
-A script generated 30 randomly placed circular obstacles with sizes uniformly distributed between 5 m and 7 m.
-The obstacle positions were drawn from a uniform distribution with an additional restriction to allow overlap of no more than half their radius~\cite{MATLABBubbleBath}.
-Finally, to increase the complexity of the resulting paths and reduce the likelihood of a straight path, \textit{each} configuration included four additional fixed obstacles with radius of 8 m placed equidistant along the diagonal from start to end.
-
-A path was designed to get from start to finish while avoiding the obstacles. 
-The path start and end positions were fixed to the south-west and north-east corners, respectively. A sample obstacle field is shown on the left in Fig.~\ref{fig:synchronodemoexample}.
-A Particle Swarm Optimization (PSO) algorithm~\cite{PSOObstacleAvoidance} was used to generate a shortest distance path connecting the start and end locations, as shown in the middle image of Fig.~\ref{fig:synchronodemoexample}.
-The path generation was constrained to produce trajectories that remained in the domain and did not intersect obstacles.
-
-Each of the 128 resulting paths shown in Fig.~\ref{fig:synchronoallpaths} was then used as the prescribed trajectory for the path-follower PID-based lateral controller implemented on the leader vehicle, in conjunction with a PID-based longitudinal controller that prescribes a target speed linearly increasing in time (corresponding to a constant acceleration of 0.47 m/s$^2$).
-
-To produce paths that are feasible for simulated HMMWV vehicles, the generation of the corresponding environment setup in SynChrono involved scaling obstacle meshes (rocks, trees, bushes) such that their bounding sphere allows 2 m of separation from the generated path centerline; in other words, in each of the 128 environments, the path prescribed for the leader vehicle ensures a minimum width of 4 m.
-An overhead view of the resulting SynChrono simulation environment is shown on the right in Figure \ref{fig:synchronodemoexample}.
-
-%The quantitative analysis was carried out systematically as follows: 128 unique environments were created with 34 total obstacles in each, a \textit{Particle Swarm Optimization} (PSO) based algorithm calculated a possible trajectory through the environment from designated start and end points \cite{PSOObstacleAvoidance}, and then a simulation was run for each environment.
-In the simulation tests, the leader vehicles utilize PID-based lateral and longitudinal controllers to follow the predetermined path with a prescribed target speed (with the latter linearly increasing in time) and the follower vehicles utilize the trained network described previously.
-
-%To produce paths feasible for a simulated HMMWV vehicle to follow, careful consideration was taken when generating each environment, its obstacles and the corresponding path through it.
-First, obstacles with radii of 5m to 7m were uniformly distributed across a 92 by 92 block of terrain centered at [0,0].
-A check was made to ensure the obstacles only overlapped at a magnitude less than or equal to their radius \cite{MATLABBubbleBath}.
-Four obstacles with radii of 8m were also placed equidistant along the diagonal from start to end.
-This was done to increase the difficulty of the computed paths, eliminating the likelihood of straighter paths.
-Figure \ref{fig:obstaclegeneration} shows an example result from this obstacle generation.
-The positions and radii of the generated obstacles were then fed into the PSO algorithm which calculated a possible trajectory through the environment.
-During the computation of the candidate path, a cubic spline interpolant was calculated to verify certain conditions weren't being violated: the path didn't lead through any obstacles and that it stayed within the 92 by 92 block of terrain.
-The candidate path generated is shown in Figure \ref{fig:psopathgeneration}.
-This progression was then performed 128 times, where the obstacle positions and the candidate path represented as a bezier curve were stored to be read by \synchrono{} later.
-Figure \ref{fig:synchronoallpaths} shows all 128 computed paths and demonstrates the overall spread of which the candidate paths span the terrain.
-
-%Once all 128 environments with obstacles and paths were generated, the policy was then tested with the 1L+3F configuration on the three terrain types: rigid, hard SCM, soft SCM.
-As a result, 384 total simulations were performed using the Euler computing cluster at the University of Wisconsin-Madison.
-For each simulation, the corresponding environmental factors were loaded in: obstacles and path to follow.
-As seen in Figure \ref{fig:psopathgeneration}, the generated path is extended close to the edge of some obstacles.
-When loaded into \chrono{}, the obstacle meshes were scaled so that their new radii allowed for 2m of separation from the prescribed path to the mesh.
-This was done to provided adequate space for the leader to drive through the environment without hitting an obstacle asset.
-An overhead view of the resulting simulation environment in \synchrono{} is shown in Figure \ref{fig:synchronodemoexample}.
-
-% \begin{figure}
-%     \centering
-%     \includegraphics[width=\columnwidth]{Figs/StatisticalAnalysis/obst_gen_68.png}
-%     \caption{Sample obstacle field.}   
-%     \label{fig:obstaclegeneration}
-% \end{figure}
-
-% \begin{figure}
-%     \centering
-%     \includegraphics[width=\columnwidth]{Figs/StatisticalAnalysis/path_gen_68.png}
-%     \caption{Sample PSO-based path planning.}   
-%     \label{fig:psopathgeneration}
-% \end{figure}
-
-% \begin{figure}
-%     \centering
-%     \includegraphics[width=\columnwidth]{Figs/StatisticalAnalysis/path_gen_rigid_68.png}
-%     \caption{Overhead view of the corresponding SynChrono environment setup.}   
-%     \label{fig:synchronodemoexample}
-% \end{figure}
-
-\begin{figure*}
-    \centering
-    \includegraphics[width=\textwidth]{Figs/StatisticalAnalysis/path_gen_progression.png}
-    \caption{Sample obstacle field, PSO-based path planning, and the corresponding SynChrono environment setup.}   
-    \label{fig:synchronodemoexample}
-\end{figure*}
+As seen in Table~\ref{tab:vehiclepositions}, each simulation configuration resulted in a different reaction of the follower vehicles.
+As indicated in \S\ref{subsec:learningDemo}, the training process rewards the follower for being within a certain angle \emph{and} distance of the leader.
+As a result, this provided some room for the vehicle to deviate from the prescribed path.
+This can be seen clearly in the three leader vehicle configuration, where the follower diverges from its leaders' path at a small margin.
+In the 2L+2F and 1L+3F platooning scenarios, this effect is magnified, as for each additional follower, a similar effect is observed; hence, the last follower in the one leader configuration tests is approximately three times further from the prescribed path around corners compared to the 3L+1F configuration.
+On occasion, this artifact led to the last vehicle clipping obstacles positioned on the inside of these corners, which points to a weakness of the control policy design.
 
@@
 	\centering
-    \includegraphics[width=\columnwidth]{Figs/StatisticalAnalysis/all_possible_paths.png}
-    \caption{Ensemble of the 128 paths used in the statistical analysis.}   
-    \label{fig:synchronoallpaths}
+	\includegraphics[width=1\columnwidth]{Figs/Demonstration/error_vs_agent.png}
+	\caption{{\small Performance metric vs.
+the number of leader vehicles.
+The performance metric is the integrated error between the leader's and follower's paths divided by the length of the leader's path.}}
+	\label{fig:errorvsagent}
 \end{figure}
-
-%%%  Metrics ----------------------------
-
-In order to perform a statistical analysis of the performance of the platooning policy, we define a set of six performance metrics that measure the deviations of a follower vehicle from that of the convoy leader and encode both lateral path deviation and deviations in the vehicle speed at a given location along the leader's path.
-These metrics are defined in such a way as to allow comparisons between the performance of followers at different positions in the convoy, as well as across the three different terrain types considered here.
-
-To eliminate differences due to the fact that vehicles in a convoy are inherently staggered along the path, the evaluation of the performance metrics is based on a common path segment.
-Figure~\ref{fig:metrics_path} illustrates this process for the same trajectory used as an example before.
-The sample results used in this description are results of a simulation on rigid terrain and consider the last vehicle in the convoy (Follower 3).
-The start clip point is defined as the point on the follower path closest to the leader's initial location.
-Similarly, the end clip point is defined as the point on the leader's path closest to the follower's final location.
-The resulting segment on the leader's path is then sampled at intervals of equal arc-length and the closest point on the follower's path to each such sampled point is identified.
-The distance between corresponding points on the leader and follower paths are then used to define a {\em follower path deviation} as function of distance traveled along the leader's path (see Fig.~\ref{fig:metrics_path_deviation}).
-This allows us to define the first two performance metrics: $m_{p}^{avg}$, the average follower path deviation, and $m_{p}^{max}$, the maximum path deviation.
-
-Next, we compare the vehicle speeds at corresponding points on the follower and leader paths as shown in the left plot of Fig.~\ref{fig:metrics_speed_deviation} from which we derive the absolute and relative speed errors (the latter being the speed difference scaled by the leader's speed at that location).
-The underlying assumption in these definitions is that a follower's speed should match as closely as possible the speed of the leader vehicle at the same location on the path (and not at the same point in time); this is also why, when calculating the subsequent speed deviation metrics, we discard the values corresponding to the first 10 m of travel (to allow the vehicles to accelerate from rest to the desired convoy speed).
-With these we define two more pairs of performance metrics: $m_{as}^{avg}$, $m_{as}^{max}$, for the average and maximum absolute speed deviation between leader and follower, and $m_{rs}^{avg}$, $m_{rs}^{max}$, for the average and maximum relative speed deviation of the follower.
+As the terrain becomes softer, the followers tend to deviate more from the leader's path.
+This trend is noticeable in the images of Table~\ref{tab:vehiclepositions}.
+We quantified the drift by calculating a performance metric defined as the area between the paths of the leader vehicle and of the last following vehicle in the convoy divided by the length of the leader's path.
+A value of zero would indicate a perfect convoy maneuver in which followers run in the tracks of the lead vehicle.
+The results presented in Fig.~\ref{fig:errorvsagent} indicate that the convoy is best maintained on rigid terrain and when the platoon contains a single follower.
+As expected, the worst performance under this metric is observed in snow-like conditions for a platoon with three followers.
 
@@
     \centering
-    \includegraphics[width=0.9 \columnwidth]{Figs/Metrics/path.png}
-    \caption{Path deviation metrics calculation (rigid terrain, Follower 3).
-The segment of the leader path used in calculations for this particular scenario (rigid terrain) was $L = 158.2 \text{ m}$.}
-    \label{fig:metrics_path}
+    \includegraphics[width=\columnwidth]{Figs/Demonstration/3f_1i_rigid_velocities.png}
+    \caption{{\small Speed of each vehicle in the 3L$ + $1F configuration; rigid terrain.}}
+    \label{fig:rigid31vel}
 \end{figure}
+%
+\begin{figure}
+    \includegraphics[width=\columnwidth]{Figs/Demonstration/3f_1i_soft_velocities.png}
+    \caption{{\small Speed of each vehicle in the 3L$ + $1F configuration; soft SCM terrain.}}
+    \label{fig:soft31vel} 
+\end{figure}
+For each simulation test, the leader vehicles utilize PID-based lateral and longitudinal controllers to follow the predetermined path with a prescribed target speed (with the latter linearly increasing in time).
+The speed profiles of the four vehicles in the 3L+1F configuration on rigid and soft SCM terrain are presented in Figs.~\ref{fig:rigid31vel} and \ref{fig:soft31vel}, respectively.
+In these simulations, the three leader vehicles use identical parameters for their path-following and adaptive cruise-control PID controllers.
+While the speed of the leader vehicles can increase continuously throughout the simulation length when driven over rigid terrain, due to the increased motion resistance on the soft SCM soil the speed levels off approximately 40 seconds into the simulation.
 
-\begin{figure}
-    \centering
-    \includegraphics[width=0.9\columnwidth]{Figs/Metrics/path_deviation.png}
-    \caption{Deviation in path between Leader and Follower 3.
-The lateral deviation metrics for this particular scenario (rigid terrain) were $m_{p}^{avg} = 0.473 \text{ m}$, $m_{p}^{max} = 2.484 \text{ m}$.}
-    \label{fig:metrics_path_deviation}
-\end{figure}
-
-\begin{figure*}
-    \centering
-    \includegraphics[width=0.9\textwidth]{Figs/Metrics/speed_deviation.png}
-    \caption{Deviation in speed between Leader and Follower 3.
-The vehicle speeds are evaluated at the same location along the leader's path.
-The speed deviation metrics for this particular scenario (rigid terrain) where $m_{as}^{avg} = 0.505 \text{ m/s}$, $m_{as}^{max} = 2.154 \text { m/s}$ and $m_{rs}^{avg} = 12.2 \%$, $m_{rs}^{max} = 69.2 \%$.}
-    \label{fig:metrics_speed_deviation}
-\end{figure*}
-
-%%%  Statistical analysis ----------------------------
-
-The six metrics defined above were evaluated for each of the three follower vehicles in each scenario in the three sets of 128 environments on rigid, SCM-Hard, and SCM-Soft terrain, respectively.
-The resulting statistics are presented as box-and-whisker diagrams in Figs.~\ref{fig:metricpath}, \ref{fig:metricspeed}, and \ref{fig:metricrelspeed}, providing measures of the variability in the three statistical populations without any assumption on their underlying statistical distribution (which is unknown, due to the manner in which the sample trajectories were constructed).
-For each metric, on each terrain type and for each of the three follower vehicles in the 1L+3F configuration, these standard box plots provide information on their mean, second and third quartiles, as well as minimum and maximum values.
-
-We assume that a perfect control policy for a follower vehicle would result in a convoy in which each follower vehicle runs exactly in the tracks of the vehicle preceding it and achieves the exact same speed at any given location.
-Given their definition, this ideal case corresponds to zero values for each of the six performance metrics.
-
-The results of the statistical analysis presented herein confirm the intuition that, in a 1L+3F configuration, a less than perfect control policy will lead to worse performance for the trailing follower vehicles and better performance on harder surfaces (especially taking into account that the training was performed exclusively on rigid terrain).
-However, as the results for both path and speed deviation show, this is not the case when comparing performance on SCM-Hard and SCM-Soft terrains, with the latter showing consistent lower metrics values (i.e., better performance in terms of maintaining position in the convoy).
-
-The explanation for this behavior is likely a combination of several factors.
-First, even though the target speed profile for the leader vehicle was set identically for all three terrain types, the increased terrain resistance in the SCM-Hard and SCM-Soft cases resulted in the leader vehicle being unable to continuously increase its speed to the specified value; on both deformable terrain types, the vehicles were unable to shift in the higher gears and their speed limited to lower levels than on rigid terrain (an effect more pronounced on SCM-Soft terrain than on SCM-Hard).
-The ensuing overall lower convoy speed results in driving scenarios where the control policy can adapt better.
-%
-Second, as seen in Fig.~\ref{fig:metrics_speed_deviation} and typical of all simulations conducted as part of this analysis, current deficiencies in the RL-based control policy result in relatively jerky motion of the follower vehicles and noisy speed profiles.
-These spurious accelerations and decelerations are inhibited on the SCM-Soft due to the increased motion resistance.
-%
-Finally, the largest deviations in a follower's path (see Fig.~\ref{fig:metrics_path}) always occur at tight turns where the leader vehicle must go around an obstacle.
-In these situations, the tendency of the control policy is to ``cut corners'' and thus direct the vehicle to increase steering input.
-However, these control steering inputs are more difficult to follow in a deformable terrain soft enough to result in deep ruts, thus resulting in the follower vehicles more closely matching the leader's vehicle path around obstacles.
-
-While the path planning procedure and the path-following PID-based control policy implemented for the leader vehicles ensures that a leader vehicle always avoids obstacles, this is not the case for the RL-based control policy implemented for the follower vehicles, which occasionally are unable to avoid an obstacle (in few situations, a follower vehicle, especially one in position 2 or 3, may end up going on the opposite side of an obstacle).
-This behavior has multiple compounding causes, including the particular reward system used in the current training as well as configurations where perception of the leader vehicle is obstructed by an obstacle or the leader vehicle is out of the camera sensor's field of view while negotiating a tighter turn.
-%
-The paths of all vehicles on each terrain type are shown in Fig.~\ref{fig:pathsranksall}.
-The path information therein is used to gauge the robustness of the control policy in terms of obstacle avoidance.
-Figure~\ref{fig:obstacleshit} provides the cumulative statistics in terms of number of obstacles hit, over all ensemble simulations, for all three terrain types and for each of the three follower vehicles.
-These results show the same relative performance trends observed before, with the trailing follower on SCM-Hard terrain exhibiting worst performance.
-
-%Using the data collected during each of the 384 simulations, the aforementioned metric calculations were performed and their results recorded for subsequent analysis, of which is presented here.
-
-%To gauge the overall performance of the policy, where the primary goal is that the follower drives in the tracks of its leader, the integrated error metric is most insightful.
-A better control policy would produce a smaller integrated error metric and when Figure \ref{fig:metricpathavg} is examined, it can be seen the SCM hard simulations produces disproportionally larger values.
-The comparative performance between rigid and SCM soft isn't as evident; however, the tests performed on rigid do behave better than SCM soft for the followers at depths 1 and 2, but approximately the same for follower 3.
-Furthermore, the spread for rigid is far less than that for SCM hard and SCM soft for follower 1 and 2.
-For follower 3, the spread between the rigid and SCM soft experiments is approximately equal, where as the SCM hard is quite larger.
-A possible explanation for the performance drop experienced by the tests on SCM hard could be attributed to the friction associated with a deformable terrain (i.e.
-there is more friction at the wheels, so the policy has difficulty reacting to this new stimuli).
-This hypothesis is contradicted by the SCM soft tests; however, a suitable explanation could be the fact the SCM soft is \textit{considerably} softer than the harder parameters and produces very deep ruts; hard enough, in fact, that it could mean the vehicle has more difficulty escaping the ruts of it's leader.
-These ideas are also evident in Figure \ref{fig:pathsranksall} and Figure \ref{fig:obstacleshit}, where they show the path taken and the number of obstacles hit by each vehicle, respectively.
-
-%Similar to the integrated path error, the maximum deviation is also insightful when attempting to analyze how the policy performs.
-Most prominent for paths where tight corners are present, significant deviations from the leaders path could result in the follower clipping obstacles or failing to see its leader all together.
-As shown in Figure \ref{fig:metricpathmax}, rigid is consistently better than both the SCM experiments.
-One simple explanation for this discrepancy is the fact that vehicles drive faster on the rigid terrain, meaning the foresight to make a turn must be greater and vehicle ends up cutting corners less.
-In addition, the policy was trained on rigid terrain, which could represent a fault of this approach.
-Moreover, the SCM soft tests typically have maximum deviations less than that of the SCM hard tests.
-This can be explained again by the fact that the snow-like soft SCM parameters result in the vehicle staying in the tracks of its leader more often.
-
-%As alluded to before, the speed of the SCM tests are far less than that for the rigid terrain trials.
-The friction associated with the deformable terrain is too great for the vehicle to reach its shift points.
-This results in decreased overall speeds for the vehicles simulated on soft and hard SCM.
-The soft SCM is also much softer than the hard SCM terrain, so its speed is also frequently smaller.
-This would mean that the integrated error between the speed of the follower and the leader would be less because of the lower speeds, seen in Figure \ref{fig:metricspeedavg}.
-This trend is also present in the maximum velocity deviation from the leader, shown in Figure \ref{fig:metricspeedmax}.
-Due to the simple fact that the SCM simulations run at lower speeds, the maximum speed deviation will be less.
-
-% \begin{figure}
-%     \centering
-%     \includegraphics[width=\columnwidth]{Figs/StatisticalAnalysis/metric_path_avg.png}
-%     \caption{Integrated error of the position between the followers and leaders path's.}   
-%     \label{fig:metricpathavg}
-% \end{figure}
-
-% \begin{figure}
-%     \centering
-%     \includegraphics[width=\columnwidth]{Figs/StatisticalAnalysis/metric_path_max.png}
-%     \caption{Maximum deviation metric between the followers and leaders path's.}   
-%     \label{fig:metricpathmax}
-% \end{figure}
-
-% \begin{figure}
-%     \centering
-%     \includegraphics[width=\columnwidth]{Figs/StatisticalAnalysis/metric_speed_avg.png}
-%     \caption{Integrated error between the followers and leaders velocity profile's.}   
-%     \label{fig:metricspeedavg}
-% \end{figure}
-
-\begin{figure*}
-    \centering
-    \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\columnwidth]{Figs/StatisticalAnalysis/metric_path_avg.png}
-        \caption{} \label{fig:metricpathavg}
-    \end{subfigure}
-    \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\columnwidth]{Figs/StatisticalAnalysis/metric_path_max.png}
-        \caption{} \label{fig:metricpathmax}
-    \end{subfigure}%
-    \caption{Statistics of average and maximum follower path deviations.}  
-    \label{fig:metricpath}
-\end{figure*}
-
-\begin{figure*}
-    \centering
-    \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\columnwidth]{Figs/StatisticalAnalysis/metric_speed_avg.png}
-        \caption{} \label{fig:metricspeedavg}
-    \end{subfigure}
-    \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\columnwidth]{Figs/StatisticalAnalysis/metric_speed_max.png}
-        \caption{} \label{fig:metricspeedmax}
-    \end{subfigure}%
-    \caption{Statistics of average and maximum follower absolute speed deviations.}  
-    \label{fig:metricspeed}
-\end{figure*}
-
-\begin{figure*}
-    \centering
-    \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\columnwidth]{Figs/StatisticalAnalysis/metric_rel_speed_avg.png}
-        \caption{} \label{fig:metricrelspeedavg}
-    \end{subfigure}
-    \begin{subfigure}{0.45\textwidth}
-        \includegraphics[width=\columnwidth]{Figs/StatisticalAnalysis/metric_rel_speed_max.png}
-        \caption{} \label{fig:metricrelspeedmax}
-    \end{subfigure}%
-    \caption{Statistics of average and maximum follower relative speed deviations.}  
-    \label{fig:metricrelspeed}
-\end{figure*}
-
-% \begin{figure}
-%     \centering
-% \end{figure}
-
-% \begin{figure}
-%     \centering
-%     \includegraphics[width=\columnwidth]{Figs/StatisticalAnalysis/metric_rel_speed_max.png}
-%     \caption{Maximum error between the followers and leaders velocity profiles, scaled by the speed of the leader at each position along its path.}   
-%     \label{fig:metricrelspeedmax}
-% \end{figure}
-
-\begin{figure*}
-    \centering
-    \begin{subfigure}{0.33\textwidth}
-        \includegraphics[width=\linewidth]{Figs/StatisticalAnalysis/1L_3F_rigid_128_pos_v5.png}
-        \caption{} \label{fig:rigid13pos}
-    \end{subfigure}%
-    \begin{subfigure}{0.33\textwidth}
-        \includegraphics[width=\linewidth]{Figs/StatisticalAnalysis/1L_3F_hard_128_pos_v5.png}
-        \caption{} \label{fig:hard13pos}
-    \end{subfigure}%
-    \begin{subfigure}{0.33\textwidth}
-        \includegraphics[width=\linewidth]{Figs/StatisticalAnalysis/1L_3F_soft_128_pos_v5.png}
-        \caption{} \label{fig:soft13pos}
-    \end{subfigure}%
-    \caption{Paths of every vehicle on each terrain type: (a) Rigid; (b) SCM-Hard; (c) SCM-Soft.}   
-    \label{fig:pathsranksall}
-\end{figure*}
-
-\begin{figure*}
-    \centering
-    \begin{subfigure}{0.33\textwidth}
-        \includegraphics[width=\linewidth]{Figs/StatisticalAnalysis/rigid_histogram.png}
-        \caption{} \label{fig:rigidhistogram}
-    \end{subfigure}%
-    \begin{subfigure}{0.33\textwidth}
-        \includegraphics[width=\linewidth]{Figs/StatisticalAnalysis/hard_histogram.png}
-        \caption{} \label{fig:hardhistogram}
-    \end{subfigure}%
-    \begin{subfigure}{0.33\textwidth}
-        \includegraphics[width=\linewidth]{Figs/StatisticalAnalysis/soft_histogram.png}
-        \caption{} \label{fig:softhistogram}
-    \end{subfigure}%
-    \caption{Number of obstacles hit by each vehicle on the three terrain types: (a) Rigid; (b) SCM-Hard; (c) SCM-Soft.}   
-    \label{fig:obstacleshit}
-\end{figure*}
+As seen in these plots, the speed profiles of the follower vehicles are not smooth; even though they generally follow the same overall rate of increase as that prescribed for the leader vehicles, their motion is jerky with successive acceleration and braking.
+This can also be seen in the associated videos \cite{simsPaperGVSETS2020}.
+Improvements to the RL-based policy to mitigate this behavior are currently being pursued and will be included in the final version of the manuscript.
 
@@
 This contribution discussed a Chrono-centric simulation platform designed to facilitate the design and testing of control policies for AAs operating in off-road conditions.
-The platform draws on a physics-based simulation engine; has templates for wheeled and tracked vehicles; enforces space and time coherence; allows for human-in-the-loop scenarios; provides sensor simulation capabilities; has a bridge to ROS; can simulate mobility on fully resolved, continuum, or SCM representations of the terrain; is open source; and is cluster deployable to support multi-AA mobility studies.
-This software framework is used here to design an RL-based control policy that allows AAs to follow in a convoy formation.
+This simulation platform targets deployment on computing clusters and displays weak scaling.
+The platform draws on a physics-based simulation engine, has templates for wheeled and tracked vehicles, enforces space and time coherence, allows for human-in-the-loop scenarios, provides sensor simulation capabilities, has a bridge to ROS, can simulate mobility on fully resolved, continuum, or SCM representations of the terrain, and is open source.
+This software framework is used here to design an RL-based control policy that allows an AA to drive in a convoy formation.
+Once the learning is completed, one, two, or three AAs were deployed in a four-vehicle convoy that drove on paths different from those used in training.
 The learning took place on rigid terrain but was demonstrated to work when deployed on AAs that operate on deformable SCM soils.
-The virtual environments used in testing differed in textures and colors from the ones used in the training, thus demonstrating robustness of the inferred policy that relies on inputs from an RGB camera sensor.
+The virtual environments used in testing differed in textures and colors from the ones used in the training, thus demonstrating robustness of the inferred policy which relies on inputs from an RGB camera sensor.
 Unsurprisingly, the fewer AAs in the platoon, the tighter it managed to follow a prescribed path.
-Looking ahead, we plan to speed up the SCM implementation; augment the sensing simulation support; improve scalability; and use this simulation infrastructure to derive new control policies for off-road AA mobility.
+The final version of the paper will augment the present work in two respects: it will contain a statistical analysis that quantifies the effectiveness of the RL-derived control policy and will discuss an improved RL-based policy that smooths out the driving style of the AAs, which is currently jerky.
 

### 2020-08-25 2020/GVSETS_2020/oldMaterial.tex
Added BookIta template, Eccomas DRL paper, GVSETS AV paper, CoRL (under review)

%% ============================================================

\clearpage
From this point on is old stuff
\section{FRAMEWORK}\label{s:framework}

%% ============================================================

\subsection{\chronomod{Sensor}}\label{s:ChSensor}

\chrono{} and \synchrono{} and \chronomod{Vehicle} \chronomod{Sensor} is a Project \chrono{} submodule for sensing simulation. Supported sensors include camera, lidar, \gps{}, and \imu{}. Exteroceptive sensing (camera, \lidar{}) leverages headless, off-screen ray-tracing support via \optix{}. \chronomod{Sensor} seeks to generate realistic data that is equivalent to data from its physical counterpart. To this end, \chronomod{Sensor} supports modeling and simulation of the noise and distortion observed on the data streams of physical sensors. Examples of \chronomod{Sensor} used to simulate a scaled autonomous vehicle navigating a closed course are shown below.

\begin{figure}
	\centering
	\includegraphics[width=0.8\columnwidth]{Figs/AV-third-person.png}
	\caption{{\small Third person perspective of scaled autonomous vehicle navigating a course using a simulated camera and \lidar{}.}}   
	\label{fig:avthirdperson}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=0.8\columnwidth]{Figs/AV-lidar.png}
	\caption{{\small Lidar data showing the course from the perspective of the virtual vehicle. The data is used in the control stack to detect the barriers and determine the desired path.}}   
	\label{fig:avlidar}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=0.8\columnwidth]{Figs/AV-greyscale.png}
	\caption{{\small Image from a simulated grayscale camera mounted to the front of the scaled autonomous vehicle.}}   
	\label{fig:avgreyscale}
\end{figure}

%%============================================================

\subsection{\pychrono{} and \gymchrono{}}\label{s:pyandgym}

\pychrono{} is a package that provides Python bindings for the \chrono{} \api{}. Using an interface compiler (SWIG), the vast majority of the \chrono{} \api{} (multibody dynamics, Vehicle, Sensor, \fea{}, etc.) can be directly accessed from Python. Beyond ease of use, the Python wrapping allows interfacing \chrono{} and ML frameworks for training neural networks. \gymchrono{} provides a set of Robotics and Autonomous Driving environments leveraging \chrono{} for physics simulation and OpenAI gym support for Reinforcement Learning baselines and tools for scalable training.

%%============================================================

\subsection{\synchrono{}}\label{s:synchrono}

\synchrono{} is a software component built on top of \chrono{} that allows for a distributed-memory execution of simulation scenarios that include tens of agents. It builds off the Message Passing Interface (\mpi{}), which allows the toolset to have multiple instances of \chrono{} run simultaneously on a supercomputer/cluster/multi-core setup to allow for the distributed simulation of multiple agents (robots, tracked vehicles, wheeled vehicles, etc.) The paradigm embraced is that of running one \chrono{} agent simulation as one \mpi{} rank, with multiple ranks, say N, communicating through messages to maintain a space and time coherent solution for the N agents participating in the study. This multi-agent form of simulation distribution is possible because of the largely decoupled nature of mobility scenarios. Each agent runs in its own rank and interfaces with its dedicated control stack for software-in-the-loop control. The control stack is fed synthetic data generated by \chronomod{Sensor} and acts upon the environment through \chronomod{Vehicle} control input (throttle, steering, braking, etc.) This “one agent-to-one \chrono{} system” mapping ensures scalability over a monolithic \chrono{} system as shown in the center picture below. The control algorithm or “brain” for each agent is also configurable and can vary from complex algorithms that collect a variety of sensor data, to controls based on empirical models, to human-driven control in scenarios that are simple enough to allow real-time simulation.

\begin{figure}
	\centering
	\includegraphics[width=0.8\columnwidth]{Figs/MPI-schematic.png}
	\caption{{\small Schematic of the \synchrono{} framework. Decisions are passed to the \chrono{} system for dynamics simulation and the outcome of the dynamics simulation is synchronized between ranks using \mpi{}.}}   
	\label{fig:mpischematicold}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=0.8\columnwidth]{Figs/Syn-Chrono-Scaling.png}
	\caption{{\small Scaling comparison between \chrono{} and \synchrono{}. By running each simulation in parallel \synchrono{} handles scenarios with large numbers of vehicles much faster and closer to real-time than \chrono{}.}}   
	\label{fig:synchronoscalingold}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=0.8\columnwidth]{Figs/Syn-Platoon.png}
	\caption{{\small Image of a collection of agents in a shared, coherent virtual world.}}   
	\label{fig:synplatoonold}
\end{figure}

\subsection{ROS \chrono{} Control Interface}\label{s:roscontrolinterface}

To allow for the seamless testing of control algorithms that are intended to be easily transferred to real vehicles or robots, the toolset provides a \ros{} control interface that is exposed in \synchrono{}. Using \tcp{}, messages from \synchrono{} (i.e. sensor data packets), are sent into a \ros{} node on the same system running the vehicle control stack. This node also listens to the \ros{} control messages that are produced and sends them back to \synchrono{} for input to the \chrono{} agents. With this \ros{} interface, the tested control stack is independent of the \synchrono{} platform (i.e. can be run on a different operating system or architecture) and can be tested with inputs replicating those from reality, such as sensor and/or V2X communication data.

\clearpage

%% ============================================================
\section{Demonstration}\label{s:demonstration}
\todo[inline]{Asher's details on the proposed demonstration: the goal of the demonstration is to show a multi-agent simulation of vehicles driven with policies learned using RL in PyChrono. To accomplish this we will train a single agent to participate in a follow-the-leader convoy through a narrow path. It will be trained on rigid terrain until an optimal policy is found. We can then refine the policy on SCM terrain. The trained policy will be deployed for all vehicles in SynChrono. We can remove the need for sensing the SCM terrain if we using a simulated vehicle-to-vehicle message from the lead vehicle. I think it would be more interesting to include a camera as input and remove the need to know the position of the lead vehicle. Final decision on that TBD. The study will then be on how that policy is deployed in SynChrono and how it responds to perturbations in the form of ruts in the soil.}

\todo[inline]{DN - Training for convoy vehicle: sensor (GPS+Camera) fusion + rigid terrain + ghost vehicle(s). Training for lead vehicle: path (perhaps way points) + sensor fusion + obstacles}

\todo[inline]{DN: SynChrono scaling analysis on SCM with 1, 2, 4, 8, etc. vehicles; half going NS, half EW. SCM deformable terrain; deformable terrain shown in simulation and also picked up by sensor.}

%% ============================================================

\begin{figure}
    \centering
    \includegraphics[width=0.8\columnwidth]{Figs/M113Fording.png}
    \caption{{\small M113 fording maneuver demonstrating \chrono{} and \chronomod{Vehicle} capabilities.}}   
    \label{fig:m113fording}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.8\columnwidth]{Figs/HMMWV-Granular.png}
    \caption{{\small HMMWV with flexible tires navigating granular terrain demonstrating vehicle dynamics, flexible body dynamics, and parallel computing support in \chrono{}.}}   
    \label{fig:hmmwvgranular}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.8\columnwidth]{Figs/GymChrono-Pilars.png}
    \caption{{\small \gymchrono{}-enabled learning with simulated sensor data provided by \chronomod{Sensor} gradually allows the vehicle to understand how to avoid random obstacles placed in its path}}   
    \label{fig:gymchronopillars}
\end{figure}

This contribution outlines the current state of a BSD3 open-source simulation toolset whose goal is to assist practitioners in endowing autonomous agents with artificial intelligence; i.e., producing control strategies that enable heterogeneous teams of robots and wheeled/tracked vehicles to operate in off-road conditions in a coordinated fashion towards accomplishing a shared task. The toolset draws on: \chrono{} , which provides multi-physics support for agent modeling; \chronomod{Sensor}, which produces synthetic data for simulating sensing; \pychrono{} and \gymchrono{}, which provide support for machine learning; \synchrono{}, a software component for software/hardware/human in the loop studies that include multiple agents; and the \ros{}-\chrono{} Control Interface, which allows the toolset to converse with the software libraries/utilities provided by \ros{} . The reinforcement learning (RL) support provided through \gymchrono{} is demonstrated in conjunction with an RL experiment in which a HMMWV-type vehicle learns how to avoid road obstacles while moving in a convoy.

\section*{Acknowledgments}
This work was supported by a U.S. Army TARDEC project. \chrono{} development was supported in part by U.S. Army TARDEC Rapid Innovation Fund grant No. W911NF-13-R-0011, Topic No. 6a, ``Maneuverability Prediction''.

%% *******************************************************************
%% Examples for including single- and double-column figures and tables
%% *******************************************************************

%\begin{figure}
%	\centering
%	\includegraphics[width=0.8\columnwidth]{Figs/hmmwv_1.png}
%	\caption{{\small A figure spanning a single column. The caption should wrap nicely here.}}   
%	\label{fig:one}
%\end{figure}

%\begin{figure*}
%	\centering
%	\includegraphics[width=0.75\textwidth]{Figs/hmmwv_1.png}
%	\caption{{\small A figure spanning the entire paper width}}    
%	\label{fig:two}
%\end{figure*}

% \begin{table}
% \begin{center}
% 	\begin{tabular}{||c |c | c||} 
% 		\hline
% 		Test  & Terrain  & Controller Vehicle \\
% 		Number &  Type & Model\\ [0.5ex] 	
% 		\hline\hline
% 		1 & Rigid & 2-DOF \\ 
% 		\hline
% 		2 & Rigid & 14-DOF \\
% 		\hline
% 		3 & Granular & 2-DOF \\
% 		\hline
% 		4 & Granular & 14-DOF \\
% 		\hline
% 	\end{tabular}
% \end{center}
% \caption{A table spanning a single column. The caption should wrap nicely here.}
% \label{t:one}
% \end{table}

% \begin{table*}
% 		\centering
% \begin{tabular}{ ||p{6cm}|p{1.8cm}|p{1.8cm}|p{1.8cm}|p{1.8cm}||  }
% 		\hline
% 		Test Number & 1 & 2 & 3 & 4\\
% 		\hline
% 		Controller Model & 2-DOF & 14-DOF & 2-DOF & 14-DOF\\
% 		\hline
% 		Terrain & Rigid & Rigid & Granular & Granular\\
% 		\hline
% 		Time to Target (s)  & 26.67 & 26.15 & 28.32 & 28.03\\ 
% 		\hline
% 		Minimum Obstacle Distance (m) & 0.897 & 5.462 & 3.491 & 4.721\\
% 		\hline
% 		Controller Effort & 0.0340 & 0.0340 & 0.0340 & 0.0306\\
% 		\hline
% 		Max. Lateral Acceleration (m/s$^{2}$)& 2.78 & 1.57 & 2.47 & 2.33 \\
% 		\hline
% 		Avg. Lateral Acceleration (m/s$^{2}$) &0.54 & 0.51 & 0.55 & 0.46\\
% 		\hline
% \end{tabular}
% \caption{A table spanning both columns.}
% \label{t:two}
% \end{table*}

### 2020-08-25 templates/Book_Ita/Contenuto/chapter_1.tex
Added BookIta template, Eccomas DRL paper, GVSETS AV paper, CoRL (under review)

%%%%%%%%%%%%%%%%%%%%% chapter.tex %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% esempio di capitolo
%
% Usare questo  file come template per il vostro documento.
%
%%%%%%%%%%%%%%%%%%%%%%%% Springer-Verlag %%%%%%%%%%%%%%%%%%%%%%%%%%

\chapter{Rotazioni}
\label{Rotazioni} % Fornire sempre un'unica label
% usare \chaptermark{}
% per alterare o aggiustare l'intestazione del capitolo nella running head

Trasformazioni di un vettore utilizzando angoli di rotazione o quaternioni.
\section{Matrice di rotazione}
Con la matrice di rotazione $[_1^0A]$ si intende la matrice che trasforma le coordinate del vettore r espresse nel riferimento ruotato 1 ${^1r}$ , nel sistema di riferimento 0. \newline In generale sarà presente anche una traslazione tra i due riferimenti, ragion per cui compare un termine di traslazione d:
\begin{equation}
\label{eq:1.1}
\left\{^0r\right\} = \left[_1^0A\right]\left\{^1r\right\} + \left\{^0d\right\}
\end{equation}

\begin{figure}[h!]
\centering
\includegraphics[height=6cm]{rotation_1.png}
\caption{Trasformazione delle coordinate di un punto}
\label{fig:1.1}
\end{figure}
La matrice di rotazione $\in \mathbb{R}^{3x3}$ per cui è definita da 9 elementi, tuttavia per calcolarla bastano 3 o 4 parametri a seconda che si utilizzino gli angoli di rotazione o i quaternioni.
\section{Angoli di Eulero}
La matrice relativa alla rotazione attorno ad un asse del riferimento può essere ricavata con calcoli trigonometrici. Poiché una qualsiasi rotazione può essere espressa come una successione di 3 rotazioni attorno a 3 assi distinti è possibile risalire alla coordinata del punto nel riferimento 3 arbitrariamente ruotato con un prodotto di matrici:
\begin{equation}
\label{eq:1.2}
\left\{^0r\right\} = \left[_1^0A\right]\left[_2^1A\right]\left[_3^2A\right]\left\{^3r\right\}
\end{equation}
La scelta della sequenza di assi di rotazione non è univoca. Per questo si sono diffuse differenti convenzioni. La più diffusa è quella degli angoli di Eulero:
\begin{figure}[h!]
\centering
\includegraphics[height=10cm]{euler_angles.pdf}
\caption{Angoli di Eulero}
\label{fig:1.2}
\end{figure}
L'utilizzo della matrice di rotazione fornisce formule dirette per il calcolo di posizioni, velocità ed accelerazioni nel riferimento fisso a patto di conoscere angoli e velocità angolari dei sistemi di riferimento solidali ai corpi del sistema. Emergono tuttavia due limiti a questo approccio: \begin{itemize}
\item L'ordine di applicazione degli angoli di rotazione dà luogo a configurazioni finali differenti (d'altronde il prodotto tra matrici non è commutativo). L'influenza dell'ordine di applicazione degli angoli ha portato diversi enti ad adottare diverse convenzioni.
\item Il limite maggiore di questo approccio tuttavia è l'emergere di singolarità nella soluzione del problema inverso; ovvero qualora si vogliano trovare gli angoli a partire dalla matrice, problema molto comune nelle applicazioni robotiche. Formalmente, esprimendo con x', y' e z' i versori lungo gli assi del riferimento mobile (ovviamente rispetto a quello fisso), si devono eguagliare elemento per elemento le matrici
\[ \begin{bmatrix}   \\ x'& y'&z' \\ \text{ } \end{bmatrix} = \begin{bmatrix}
a_{11} & a_{12} & a_{13} \\ a_{21} & a_{22} & a_{23} \\ a_{31} & a_{32} & a_{33}\end{bmatrix} \]
Con: \qquad $a_{33} = cos\theta,\qquad a_{32} = -sin\theta cos\phi \qquad$ ecc... \newline
È evidente che in corrispondenza di certi angoli si riscontrano singolarità ad è impossibile risolvere il sistema di 9 equazioni. Questo ha anche un riscontro pratico detto blocco cardanico o \textit{gimbal lock}: una sospensione cardanica con 3 cerniere con lo scopo di isolare un oggetto dalle rotazioni del telaio in certe configurazioni non permette più rotazioni arbitrarie, di fatto bloccando alcune rotazioni.  
\end{itemize}

\section{Quaternioni}
Per questa ragione si utilizzano 4 parametri, descrivendo le rotazioni tramite quaternioni. Si evitano così le singolarità nella cinematica inversa, l'utilizzo di funzioni trigonometriche per la soluzione del sistema ed il problema dell'arbitrarietà della sequenza di rotazioni.
In Generale un quaternione è un elemento scrivibile come $ \overline{q} = a + bi + cj+ ck$ con $i^2=j^2=k^2 = ijk -1$. \newline Il quaternione descrive una rotazione di un angolo $\theta$ attorno ad un versore u come in figura \ref{fig:1.3}.

La rotazione di un angolo $\theta$ attorno ad un versore di componenti ${a, b, c}$ è quindi decritta dal quaternione
\begin{equation} 
    \overline{q} = cos\frac{\theta}{2} + (ai + bj + ck)sin\frac{\theta}{2}
\end{equation}{}

Tramite l'utilizzo dei quaternioni si può esprimere:
\begin{itemize}
\item La cinematica diretta \begin{equation}
\label{eq:1.3}
(0, v') = \overline{q}(0,v)\overline{q}^*\end{equation}
Con (0,v) vettore di 4 elementi e $\overline{q}^*$ coniugato di $\overline{q}$
\item La cinematica inversa: \begin{equation}
\label{eq:1.4}
(0, v) = \overline{q}^*(0,v')\overline{q}\end{equation}
\item Rotazioni successive :
\begin{equation} 
    \overline{q} = \overline{q}_2 \overline{q}_1 
\end{equation}{}
\item La matrice di rotazione A:
\begin{equation}\left[A(q)\right] = \left[F(q)_+\right]\left[F(q)_-\right]^T \label{eq:1.5}\end{equation}
Con 
\begin{equation} \label{eq:1.6} \left[F(q)_+\right]\ = \begin{bmatrix}
+q_1 & +q_0 & -q_3 & +q_2 \\ +q_2 & +q_3 & +q_0 & -q_1 \\ +q_3 & -q_2 & +q_1 &+q_0\end{bmatrix}
\end{equation}
\begin{equation} \label{eq:1.7} \left[F(q)_-\right]\ = \begin{bmatrix}
+q1 & +q_0 & +q_3 & -q_2 \\ +q_2 & -q_3 & +q_0 & +q_1 \\ +q_3 & +q_2 & -q_1 &+q_0\end{bmatrix}
\end{equation}
\item Velocità in funzione di $ \dot{\overline{q}}$ e viceversa
\begin{equation} \label{eq:1.8}
\dot{\overline{q}} = \frac{1}{2}\left[F(q^*)_-\right]\overrightarrow{\omega}_0 = \frac{1}{2}\left[F(q^*)_+\right]\overrightarrow{\omega}_l\end{equation}
\begin{equation} \label{eq:1.9}
\overrightarrow{\omega}_0 = 2\left[F(q^*)_-\right]\dot{\overline{q}}\end{equation}
\begin{equation} \label{eq:1.10}
\overrightarrow{\omega}_l = 2\left[F(q^*)+-\right]\dot{\overline{q}}\end{equation}
\item Accelerazione angolare in funzione di $\ddot{\overline{q}}$ e viceversa
\end{itemize}
\begin{figure}[h!]
\centering
\includegraphics[height=5cm]{quaternion_def.png}
\caption{Trasformazione delle coordinate di un punto}
\label{fig:1.3}
\end{figure}

%

### 2020-08-25 templates/Book_Ita/Contenuto/chapter_2.tex
Added BookIta template, Eccomas DRL paper, GVSETS AV paper, CoRL (under review)

\chapter{Formula di Lagrange}
\label{Formula di Lagrange} 
%\textbf{Nota:} Per facilitare la lettura in seguito si userà il grassetto per le quantità vettoriali mentre l'overline indicherà che le coordinate in questione sono relative. \newline
\section{Lavoro Virtuale}
Un passaggio essenziale nella formulazione lagrangiana delle equazioni della dinamica del sistema multibody è il calcolo delle forze generalizzate associate con le forze generalizzate del sistema.
In questa sezione le forze generalizzate sono introdotte applicando il principio dei lavori virtuali nei casi statico e dinamico. Nello svolgimento in seguito si considera un sistema di particelle.
Assumendo che un corpo rigido non è altro che un grande numero di particelle queste espressioni saranno estese ai corpi rigidi con poche modifiche.
\paragraph{Equilibrio Statico}
Si consideri un sistema composto da $n_p$ particelle. Se il sistema di particelle è in equilibrio significa che 
\begin{equation}
\label{eq:equilibrium_virtualwork}
\mathbf{ \sum_{i=1}^{n_p} F^i \delta r^i }=  0
\end{equation}
Dove F ed r sono la forza agente e lo spostamento virtuale relativi ad un particella i. La forze sono di 2 tipi: forze esterne e forze di vincolo: 
$F^i_e +F^i_c = F^i$ ed applicando questa suddivisione all'equazione \ref{eq:equilibrium_virtualwork} si ottiene:
\[\sum_{i=1}^{n_p}F^i\delta r^i = 0 \Rightarrow \sum_{i=1}^{n_p}F_e^i\delta r^i + \sum_{i=1}^{n_p}F_c^i\delta r^i = 0 \Rightarrow \delta W_e + \delta W_c = 0 \]
Lo spostamento di ogni particella è funzione delle coordinate del sistema $q= (q_1, q_2,\quad ... \quad q_n)$.
\[r_i = r_i(q_1, q_2,\quad ... \quad q_n) \Rightarrow \delta r_i = 
\frac{\partial r_i}{\partial q_1}\delta q_1 + \frac{\partial r_i}{\partial q_2}\delta q_2 + \frac{\partial r_i}{\partial q_3}\delta q_3 + .... + \frac{\partial r_i}{\partial q_n}\delta q_n \]
Si consideri il caso in cui i vincoli non compiono lavoro (\emph{workless constraints}), ad esempio quando i vincoli non hanno attrito (o la forza di attrito viene inserita fra le forze esterne). In questo caso il lavoro virtuale è dovuto alle sole forze esterne: \begin{equation}
\delta W = \delta W _e = 
\sum_{i=1}^{n_p}\left( F_e^i \sum_{j=1}^n \frac{\partial r_i}{\partial q_j}\delta q_j \right) = 0
\end{equation}
Si definisce $Q_j$ come: \begin{equation}
Q_j = \sum_{i=1}^n F_e^i \frac{\partial r_i}{\partial q_j} = \sum_{i=1}^n \left(F_e^i\right)^T r^i_{q_j} \qquad \left( r^i_{q_j}=\frac{\partial r_i}{\partial q_j} \right)
\end{equation}
Per cui si ottiene: \begin{equation}
\delta W = \delta W_e = \sum_{j=1}^n Q_j\delta q_j = Q^T \delta q = 0
\end{equation}
Con \begin{equation}
Q = \left[Q_1, Q_2 , .... , Q_n\right]^T
\end{equation}
Vettore delle forze generalizzate.

\paragraph{Equilibrio dinamico, Principio di D'Alembert}
La seconda legge di Newton afferma che: \begin{equation}
\label{eq:newton_law}
F^i = \dot{P}^i \qquad o \qquad F^i - \dot{P}^i = 0
\end{equation}
Con F risultante delle forze e P quantità di moto agenti sulla generica particella i. 
Da \ref{eq:newton_law} esplicitando il lavoro virtuale di un sistema di particelle in equilibrio dinamico si 
ottiene il \emph{principio di D'Alembert}
\begin{gather} 
\label{eq:dalembert_principle}
\sum_{i=1}^{n_p} (F_e^i + F_c^i -\dot{P}^i )\delta r^i = \sum_{i=1}^{n_p} (F_e^i -\dot{P}^i )\delta r^i + \sum_{i=1}^{n_p} F_c^i\delta r^i = 0 \nonumber \\ \underset{workless constraints}{\Rightarrow} \sum_{i=1}^{n_p} (F_e^i -\dot{P}^i )\delta r^i = 0 \nonumber \\ \Rightarrow \sum_{i=1}^{n_p} (F_e^i -\dot{P}^i )\sum_{j=1}^n \frac{\partial r_i}{\partial q_j} \delta q_j = 0
\end{gather}

Ridefinendo $Q_j$ nel caso dinamico comprendendo il contributo delle forze d'inerzia si ottiene:
\begin{equation}
Q_j = \sum_{i=1}^{n_p} \left(F_e^i-\dot{P}^i\right)\frac{\partial r^i}{\partial q_j}
\end{equation}
Da cui:
\begin{equation}
\sum_{j=1}^n \sum_{i=1}^{n_p} \left(F_e^i-\dot{P}^i\right)\frac{\partial r^i}{\partial q_j}\delta q_j = \sum_{j=1}^n Q_j \delta q_j = Q^T = 0
\end{equation}
\section{Dinamica Lagrangiana per particelle}
Si introduce qui la formulazione lagrangiana della dinamica di un sistema di particelle che verrà in seguito estesa
ai corpi rigidi.
\subsection{Lavoro Virtuale su un sistema di particelle}
\paragraph{Spostamento di una particella \emph{i}}
Lo spostamento di una particella \emph{r} è funzione del sistema di coordinate \emph{q} e del tempo \emph{t}:
\begin{equation}
r^i = r^i\left(q_1, q_2, q_3 ... ,q_n, t\right)
\end{equation}
Da cui si ottiene, derivando a catena:
\begin{align}
\label{eq:particle_disp_time_der}
\dot{r}^i = \frac{D r^i}{D t} & = \frac{\partial r^i}{\partial q_1}\dot{q_1} + \frac{\partial r^i}{\partial q_2}\dot{q_2}
+ ... + \frac{\partial r^i}{\partial q_n}\dot{q_n} + \frac{\partial r^i}{\partial t} \nonumber \\
& = \sum_{j=1}^n \frac{\partial r^i}{\partial q_j}\dot{q_j} + \frac{\partial r^i}{\partial t}
\end{align}
Si osservi che la sommatoria è dovuta alla variazione delle coordinate della particella nel riferimento, mentre il secondo termine è la variazione di $r^i$ mantenendo $q_j$ fissato, per cui potremmo vedere i due termini come una velocità relativa ed una velocità di trascinamento di un punto rispetto ad un sistema di riferimento non inerziale. \newline
Lo spostamento virtuale si può esprimere in funzione delle coordinate $q_j$ come:
\begin{equation}
\label{eq:particle_virtual_disp}
\delta r^i = \sum_{j=1}^n\frac{\partial r^i}{\partial q_j}\delta q_j
\end{equation}
\paragraph{Lavoro su una particella \textit{i}}
Dalla formula \ref{eq:particle_virtual_disp} per lo spostamento virtuale della particella si ottiene il lavoro virtuale della forza agente sulla particella 1\emph{i}:
\begin{equation}
\label{eq:virt_work_particle}
{F^i}^T \delta r^i = \sum_{j=1}^n {F^i}^T \frac{\partial r^i}{\partial q_j}\delta q_j
\end{equation}
Estendendo l'equazione \ref{eq:virt_work_particle} al sistema di particelle:
\begin{equation}
\label{eq:virt_work_particlesyst}
\sum_{i=1}^{n_p} {F^i}^T \delta r^i = \sum_{i=1}^{n_p}\sum_{j=1}^n {F^i}^T \frac{\partial r^i}{\partial q_j}\delta q_j = \sum_{j=1}^n Q_j \delta q_j
\end{equation}
Dove $Q_j$ è chiamato le \emph{componente delle forze generalizzate associato alla coordinata $q_j$} ed è:
\begin{equation}
Q_j = \sum_{i=1}^{n_p} {F^i}^T \frac{\partial r^i}{\partial q_j}
\end{equation}
\paragraph{Lavoro virtuale della forza di inerzia}
Il lavoro virtuale di tutte le forze d'inerzia del sistema può essere scritto come somma dei lavori virtuali delle forze di inerzia agenti di ogni \emph{i}-esima particella di massa $m^i$:
\begin{equation}
\label{eq:inertia_vwork_particle}
\delta W_i = \sum_{i=1}^{n_p}m^i\ddot{r}^i\cdot \frac{\partial r^i}{\partial q_j}\delta q_j 
\end{equation}
Un' utile identità:
\[\sum_{i=1}^{n_p}\frac{d}{dt}\left(m^i\dot{r}^i\frac{\partial r^i}{\partial q_j}\right)
 = \sum_{i=1}^{n_p} \left(m^i\ddot{r}^i\frac{\partial r^i}{\partial q_j}\right) +  \sum_{i=1}^{n_p} \frac{d}{dt} \left(m^i\dot{r}^i\frac{d}{dt}\frac{\partial r^i}{\partial q_j}\right) \]
Da cui:
\begin{equation}
\label{eq:part_inertia_work_identity}
\sum_{i=1}^{n_p}\left(m^i\ddot{r}^i\frac{\partial r^i}{\partial q_j}\right) 
 = \sum_{i=1}^{n_p}\left[ \frac{d}{dt}\left(m^i\dot{r}^i\frac{\partial r^i}{\partial q_j}\right) - m^i\dot{r}^i\frac{d}{dt}\left(\frac{\partial r^i}{\partial q_j}\right) \right]
\end{equation}
Si ottiene grazie all'equazione \ref{eq:particle_disp_time_der}
\begin{equation}
\label{eq:drdot_dq}
\frac{d}{dt}\left(\frac{\partial r^i}{\partial q_j}\right) = \sum_{k=1}^n \frac{\partial ^2 r^i}{\partial q_j\partial q_k}\delta \dot{q}_k + \frac{\partial ^2r^i}{\partial q_j\partial t} = \frac{\partial \dot{r}^i}{\partial q_j}
\end{equation}
Inoltre, sempre a partire dalla \ref{eq:particle_disp_time_der} e considerando la derivata parziale di $\dot{r}^i$ rispetto ad $q_j$:
\begin{equation}
\label{eq:drdot_dqdot}
\frac{\partial\dot{r}^i}{\partial \dot{q}_j} = \frac{\partial r^i}{\partial q_j}
\end{equation}
Dall'identità \ref{eq:part_inertia_work_identity} si ottiene:
\begin{align}
& \sum_{i=1}^{n_p} \frac{d}{dt} \left(m^i\ddot{r}^i\frac{\partial r^i}{\partial q_j}\right) 
 = \sum_{i=1}^{n_p}\left[ \frac{d}{dt}\left(m^i\dot{r}^i\frac{\partial r^i}{\partial q_j}\right) - m^i\dot{r}^i\frac{d}{dt}\left(\frac{\partial r^i}{\partial q_j}\right) \right]  \nonumber \\
&= \sum_{i=1}^{n_p}\left\{\frac{d}{dt}\left[\frac{\partial}{\partial \dot{q}_j}\left(\frac{1}{2}m^i\dot{r^i}^T\dot{r}^i\right)\right] -\frac{\partial}{\partial q_j}\left(\frac{1}{2}m^i\dot{r^i}^T\dot{r}^i\right)\right\}
\end{align}
\subsection{Equazione di Lagrange}
Denotando con $T^i$ l'energia cinetica della \emph{i}-esima particella:
\begin{equation}
\label{eq:part_kinenergy}
T^i = \frac{1}{2}m^i\dot{r^i}^T\dot{r}^i
\end{equation}
E con \emph{T} l'energia totale del sistema:
\[T = \sum_{i=1}^{n_p}t^i\]
Si ottiene:
\begin{align}
\label{eq:part_inertia_j}
\sum_{i=1}^{n_p} \left(m^i\ddot{r}^i\frac{\partial r^i}{\partial q_j}\right) 
&= \sum_{i=1}^{n_p}\left\{\frac{d}{dt}\left[\frac{\partial}{\partial\dot{q}_j}(T^i)\right] -\frac{\partial T^i}{\partial q_j} \right\} \\
&= \frac{d}{dt}\left(\frac{\partial T}{\partial\dot{q}_j}\right) - \frac{\partial T}{\partial q_j} 
\end{align}
Sostituendo \ref{eq:part_inertia_j} in \ref{eq:inertia_vwork_particle} ed applicandola al principio di D'Alembert (\ref{eq:dalembert_principle}
si perviene alla \emph{Formula di Lagrange} :
\begin{equation}
\label{eq:lagrange_notind}
\sum_j\left[ \frac{d}{dt}\left( \frac{\partial T}{\partial\dot{q}_j}\right) -\frac{\partial T}{\partial q_j} -Q_j \right] = 0
\end{equation}
Se le coordinate $q_j$ sono linearmente indipendenti si ha:
\begin{equation}
\label{eq:lagrange}
\frac{d}{dt}\left( \frac{\partial T}{\partial\dot{q}_j}\right) -\frac{\partial T}{\partial q_j} -Q_j = 0 \qquad j = 1, 2, ... ,n
\end{equation}
Si può riscrivere la \ref{eq:lagrange_notind} (e conseguentemente la \ref{eq:lagrange}) in forma matriciale:
\begin{equation} \label{eq:lagrange_matrix}
\left[ \frac{d}{dt} \left(\frac{\partial T}{\partial \dot{q}} \right) - \frac{\partial T}{\partial q} - Q_e^T \right] \delta q = 0
\end{equation}
%

### 2020-08-25 templates/Book_Ita/Contenuto/chapter_3.tex
Added BookIta template, Eccomas DRL paper, GVSETS AV paper, CoRL (under review)

%%%%%%%%%%%%%%%%%%%%% chapter.tex %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% esempio di capitolo
%
% Usare questo  file come template per il vostro documento.
%
%%%%%%%%%%%%%%%%%%%%%%%% Springer-Verlag %%%%%%%%%%%%%%%%%%%%%%%%%%

\chapter{Dinamica dei corpi rigidi}
\label{dinamica_corporigido}
\section{Dinamica vincolata}
\subsection{Matrice Jacobiana dei vincoli $C_q$} \label{sec:Jabobian}
\paragraph{Vincoli}
Per bloccare alcuni gradi di libertà di un sistema e permetterne altri vengono inseriti vincoli (\emph{constraints}) nel sistema. A seconda di quanti e quali gradi di libertà vengono bloccati esistono vari tipi di vincoli, uno dei più comuni è la cerniera (\emph{revolute joint}) illustrata in figura \ref{fig:2.1}.
\begin{figure}[ht]
\centering
\includegraphics[height=7cm]{revolute.pdf}
\caption{Cerniera tra i corpi i e j}
\label{fig:2.1}
\end{figure}
In questo caso i corpi i e j hanno un solo grado di libertà relativo, ovvero la rotazione (si noti che siamo nel caso di moto piano).
In un sistema multibody le coordinate possono essere messe in relazione grazie ai vincoli, si potrà quindi scegliere quali coordinate sono dipendenti e quali indipendenti. 
Ogni equazione vincolare può eliminare una coordinata (che può essere quindi scritta in funzione di un'altra)
Perciò, un sistema di  $n$ coordinate ed $n_c$ equazioni vincolari ha $n-n_c$ coordinate indipendenti. 
Le coordinate indipendenti sono dette anche Gradi Di Libertà (GDL) del sistema.

\paragraph{Matrice Jacobiana $C_q$} Per poter considerare la presenza dei vincoli nella soluzione sarà necessario trovare relazioni tra i gradi di libertà dovuti ai vincoli, le quali andranno poi organizzate in un sistema di equazioni, per definire infine una matrice dei vincoli $C(q,t)$. \newline Le condizioni imposte da ciascun vincolo danno luogo ad un sistema di equazioni del tipo:
\begin{equation} \label{eq:Cqt}
C(q,t) = 0
\end{equation}
La relazione che definisce una cerniera nel caso di moto piano come mostrato in figura \ref{fig:2.1} ad esempio è data imponendo che la posizione del punto P considerato solidale al corpo i sia uguale alla posizione di P considerato solidale a j:
\[\overline{R}_i + A_i\overline{u}_i -\overline{R}_j - A_j\overline{u}_j = 0\]
Dove con R si indica la posizione assoluta delle origini dei riferimenti solidali ad i e j, con A la matrice di rotazione, con u la posizione di P e con i pedici i e j si indica che l'entità è relativa a quel corpo.
Ogni vincolo dà luogo ad numero di equazioni pari al suo grado di vincolo (ossia il numero di gradi di libertà bloccati dal vincolo). Poiché il sistema dovrà essere risolto numericamente con l'algoritmo di Newton-Raphson si rende necessario derivare rispetto agli spostamenti le equazioni dei vincoli ed organizzare i coefficienti in una matrice. La matrice così ottenuta sarà la matrice Jacobiana dei vincoli o sempicemente Jacobiano:
\begin{equation}
\label{eq:jacobian}
C_q =  \begin{bmatrix}
\frac{\partial C_1}{\partial q_1} & \frac{\partial C_1}{\partial q_2} & \qquad ... \qquad & \frac{\partial C_1}{\partial q_n} \\\
\frac{\partial C_2}{\partial q_1} & \frac{\partial C_2}{\partial q_2} & \qquad ... \qquad & \frac{\partial C_2}{\partial q_n} \\ . \\. \\ 
\frac{\partial C_{n_c}}{\partial q_1} & \frac{\partial C_{n_c}}{\partial q_2} & \qquad ... \qquad & \frac{\partial C_{n_c}}{\partial q_n}
\end{bmatrix} 
\end{equation}
Sviluppando l'equazione \ref{eq:Cqt} si può quindi scrivere:
\begin{equation} \label{jacobian_expansion}
C_q \delta q = 0
\end{equation}
Prendendo ad esempio il sistema con 2 soli corpi in figura [\ref{fig:2.1}] si ottiene:
\[ (C_1 , C_2)^T = \overline{R}_i + A_i\overline{u}_i -\overline{R}_j - A_j\overline{u}_j \]
Il sistema ha 6 coordinate e 2 equazioni vincolari date dall'unica cerniera. Le derivate rispetto alle posizioni sono banali, per quanto riguarda le rotazioni:
\[\frac{\partial C_1}{\partial \theta_i } = \frac{\partial A_i}{\partial \theta_i } \overline{u}_i  = A_i'\overline{u}_i\]
Dove A' indica la derivata della matrice di rotazione rispetto all'angolo che la definisce. Indicando un $u_{i,1}, u_{i,2}, u_{j,1} u_{j,2}$ le componenti nel primo e secondo asse dei vettori $u_i$ ed $u_j$, si può scrivere:
\begin{equation} C_q \delta q = 
\begin{bmatrix}
1 \quad & 0\quad & -u_{i,1}sin\theta_i-u_{i,2}cos\theta_i \quad& -1 \quad& 0 \quad& -u_{j,1}sin\theta_j-u_{j,2}cos\theta_j \\
0 \quad& 1 \quad& u_{i,1}cos\theta_i-u_{i,2}sin\theta_i \quad& 0 \quad& -1 \quad& u_{j,1}cos\theta_j-u_{j,2}sin\theta_j
\end{bmatrix}
\begin{pmatrix}
\delta R_{i,1} \\ \delta R_{i,2} \\ \delta \theta_i \\ \delta R_{j,1} \\ \delta R_{j,2} \\ \delta \theta_j 
\end{pmatrix} = 0
\end{equation}

\subsection{Moltiplicatori Lagrangiani}
Nella sezione  \ref{sec:Jabobian} è stato introdotto il calcolo dello Jacobiano e il concetto di coordinate dipendenti ed indipendenti. Per la soluzione numerica è conveniente considerare le reazioni vincolari come segue. \newline
\paragraph{Singolo Vincolo}
il jacobiano del vincolo $C_q$ tra i corpi \emph{i} e \emph{j} viene spezzato in 2 matrici:
\begin{equation} \label{eq:jacobian_composition}
C_q = \left[ C_{q^i} C_{q^j}\right]
\end{equation}
Si definisce quindi un sistema di forze \emph{equipollente} agente nelle origini dei sistemi di riferimento dei corpi di i e j. $\overline{\lambda}$ è il vettore delle forze e dei momenti esercitati dal vincolo:
\begin{equation} \label{eq:lambda}
\overline{\lambda} = - \begin{bmatrix} \overline{F} \\ \overline{M}
\end{bmatrix} \end{equation}
Le forze di reazione agenti sui corpi i e j sono:
\begin{equation}
\overline{F}^i = -\overline{\lambda} = 
\begin{bmatrix} \overline{F} \\ \overline{M} \end{bmatrix} \qquad ; 
\qquad \overline{F}^J = \overline{\lambda} = -\begin{bmatrix} \overline{F} \\ \overline{M} \end{bmatrix}
\end{equation}
Mentre, considerate nelle origini come sistema equipollente si ottiene:
\begin{equation}
Q^i_c = \begin{bmatrix} \overline{F} \\ \overline{M} + A^i\overline{u}^i_p x F \end{bmatrix} \qquad ; 
\end{equation}
Dove $\overline{u}^i_p$ è il vettore posizione di P (ove è applicato il vincolo) nelle coordinate di i e $A^i$ matrice di rotazione di i.
Da \cite{shabana94} si ha che:
\[A^i\overline{u}^i_p x F = {\overline{u}^i_p}^T {A_\theta^i}^T  \]
Si osserva quindi che le reazioni generalizzate possono essere espresse come:
\begin{equation}
\label{eq:lagrangemulti_to_reaction}
Q_c^i = -C^T_{q^i} \lambda
\end{equation}
\paragraph{Più vincoli}
Si può estendere la procedura illustrata al caso di più corpi collegati da più vincoli. Se sul corpo i ci sono $n_i$ vincoli le equazioni che li descrivono vengono espresse come:
\begin{equation} \label{eq:multiplejointeq} \begin{cases}
C_1(q,t) = 0 \\ C_2(q,t) = 0 \\ . \\ .\\ . \\ C_{n_i}(q,t) = 0
\end{cases}
\end{equation}
Ogni equazione vettoriale in \ref{eq:multiplejointeq} contiene tante equazioni quanti sono i \emph{gradi di vincolo} (gradi di libertà eliminati dal vincolo) del vincolo corrispondente. Dalla \ref{eq:lagrangemulti_to_reaction} si può estendere:
\begin{equation}\begin{cases}
Q_1^i = -\left( C_1 \right)^T_{q^i} \lambda_1 \\ Q_2^i = -\left( C_2 \right)^T_{q^i} \lambda_2 \\ . \\
Q_{ni}^i = -\left( C_{ni} \right)^T_{q^i} \lambda_{ni}
\end{cases} \end{equation}
Indicando con $Q^i_c$ la risultante delle reazioni vincolari sul corpo \emph{i} si può scrivere:
\begin{equation}\label{eq:reactions_result_body}
Q^i_c = -\left[ \left( C_1 \right)^T_{q^i} \quad \left( C_2 \right)^T_{q^i} \qquad \cdot\cdot\cdot 
\qquad \left( C_{ni} \right)^T_{q^i} \right] \begin{bmatrix}
\lambda_1 \\ \lambda2 \\ \cdot \\ \cdot \\ \cdot \\ \lambda_{n_i} 
\end{bmatrix} \end{equation}
Estendendo il calcolo a tutti gli $n_b$ corpi del sistema si calcolano le reazioni generalizzate dell'intero sistema $Q_c$. Essendo:
\[Q_c^i = -C^T_{q^i}\lambda\]
Si otterrà:
\begin{equation} \label{eq:sys_ge_reactions}
Q_c = \left[{Q_c^1}^T \qquad {Q_c^2}^T \quad \cdot \quad \cdot \quad \cdot {Q_c^{n_b}}^T\right]^T = \begin{bmatrix}Q_c^1 = -C^T_{q^1}\lambda \\ Q_c^2 = -C^T_{q^2}\lambda
\\ \cdot \\ \cdot \\ \cdot \\ Q_c^{n_b} = -C^T_{q^{n_b}}\lambda \end{bmatrix}
\end{equation}
Estendendo la composizione dello jacobiano in \ref{eq:jacobian_composition} si può quindi scrivere:
\begin{equation} \label{eq:reactions_result_sys}
Q_c = - \left[ C_{q^1} \qquad C_{q^2} \quad \cdot\quad\cdot\quad\cdot\quad Q_{q^{n_b}} \right]^T \lambda = -C_q^T\lambda
\end{equation}
\section{Equazioni della dinamica vincolata}
Per estendere quanto visto riguardo ai sistemi di particelle  nel cap. \ref{Formula di Lagrange} a sistemi di corpi rigidi sono necessarie alcune considerazioni:
\paragraph{Cinematica del corpo rigido}
La posizione di un corpo si esprime per mezzo di una traslazione dell'origine ed una rotazione:
\begin{equation} \label{eq:body_point_pos}
r^i = R^i + A^i\overline{u}^i
\end{equation}
Dove r è la posizione assoluta del punto P appartenente al corpo i, R è la posizione dell'origine del riferimento solidale ad i, A è la matrice di rotazione del corpo i e $\overline{u}$ è la posizione relativa di P. La velocità si ottiene derivando rispetto al tempo:
\begin{equation} \label{eq:body_point_vel}
\dot{r}^i = \dot{R}^i + \dot{A}^i\overline{u}^i = \dot{R}^i - A^i \tilde{\overline{u}}^i\overline{G}\dot{\theta}^i
\end{equation}
Con $\theta = \overline{q}$ quaternione unitario o vettore dei parametri di Eulero (indifferentemente), ed $\overline{G} = 2F(q^*)_-$ dell'equazione \ref{eq:1.9}. $\tilde{\overline{u}}$ è la matrice emisimmetrica ottenuta dalla posizione relativa $\overline{u}$ del generico punto P:
\begin{eqnarray} \tilde{\overline{u}}  = \begin{bmatrix}
0 & -x_3^i & x_2^i \\ x_3^i & 0 & -x_1^i \\ -x_2^i & x_1^i & 0 \end{bmatrix}\end{eqnarray}
Lo spostamento può essere dunque scritto in forma partizionata:
\begin{equation} \label{eq:body_point_vel_partform}
\dot{r}^i = \left[ I \qquad - A^i\tilde{\overline{u}}^i\overline{G}^i\right] \begin{bmatrix}\dot{R}^i \\ \dot{\theta}^i \end{bmatrix}  \end{equation}
\subsection{Matrice di massa}
Estendendo ad un corpo la definizione di energia cinetica data nell'equazione \ref{eq:part_kinenergy} si ottiene:
\begin{align} \label{eq:body_kinenergy}
T^i &= \frac{1}{2}\int_{V^i}\rho^i\dot{r}^{i^T} \dot{r}^idV^i 
\\ &= \frac{1}{2}\int_{V^i}\rho^i [\dot{R}^{i^T}\quad \dot{\theta}^{i^T}] \begin{bmatrix}I \\ - \overline{G}^{i^T}\tilde{\overline{u}^{i^T}A^{i^T}} \end{bmatrix} [I \quad -A^i\tilde{\overline{u}}^i\overline{G}^i] \begin{bmatrix} \dot{R}^i \\ \dot{\theta}^i \end{bmatrix}  dV^i \nonumber \\ &= \frac{1}{2}[\dot{R}^{i^T}\quad \dot{\theta}^{i^T}]\left\{\int_{V^i}\rho^i \begin{bmatrix}
I & -A^i\tilde{\overline{u}}^i\overline{G}^i \\ \quad sym_b \quad & \overline{G}^{i^T}\tilde{\overline{u}}^{i^T}\tilde{\overline{u}}^i\overline{G}^i
\end{bmatrix}  dV^i\right\} \begin{bmatrix} \dot{R}^i \\ \dot{\theta}^i \end{bmatrix} \nonumber
\\ &= \frac{1}{2}\dot{q}_r^{i^T}M^i\dot{q}_r^i
\end{align}
Con $sym_b$ si intende che il blocco è simmetrico.
$M^i$ è la \emph{matrice di massa} del corpo rigido i di volume $V^i$:
\begin{align}
M^i &= \int_{V^i}\rho^i \begin{bmatrix}
I & -A^i\tilde{\overline{u}}^i\overline{G}^i \\ \quad sym_b \quad & \overline{G}^{i^T}\tilde{\overline{u}}^{i^T}\tilde{\overline{u}}^i\overline{G}^i \end{bmatrix}  dV^i \\
&= \begin{bmatrix} m^i_{RR} & m^i_{R\theta} \\ \quad sym_b\quad &m^i_{\theta\theta}  \end{bmatrix} \end{align}
Si verifica facilmente che $M^i_{RR} = m^iI^{3x3}$ con $m_i$ massa dell'i-esimo corpo.
La matrice $M_{RR}$ associata alla traslazione del corpo è quindi una matrice diagonale costante.
La matrice associata all'interazione di rotazione e traslazione $m_{R\theta}$, ricordando che $a^i$ e $\overline{G}^i$ non dipendono dallo spazio si può esprimere come:
\begin{equation}
m^i_{R\theta} = - \int_{V^i}\rho^iA^i\tilde{\overline{u}}^i\overline{G}^idV^*i = -A^i \left[ \int_{V^i}\rho^i\tilde{\overline{u}}^i dV^i \right]  \overline{G}^i = -A^i\tilde{\overline{U}}^i\overline{G}^i
\end{equation}
Si noti che, qualora l'origine del riferimento coincida con il baricentro del corpo, $\tilde{\overline{U}}^i$ si annulla per la definizione di baricentro. 
Procedendo similmente per $m^i_{\theta\theta}$ si ottiene: \begin{align}
m^i_{\theta\theta} &= \overline{G}^{i^T}I_{\theta\theta}\overline{G}^i \\
I_{\theta\theta} &= \int_{V^i} \rho^i\tilde{\overline{u}}^{i^T} \tilde{\overline{u}}^i dV^i \\
I_{\theta\theta}&= \begin{bmatrix}i_{11} & i_{12} & i_{13} \\ \quad & i_{22} & i_{23} \\ sym & \quad & i_{33}\end{bmatrix}
\end{align}
Con $sym$ si nitende che la matrice è simmetrica
Con \begin{align} \label{eq:inertia_moment}
i_{jj} &= \int_{V^i} \rho^i[(x_k^i)^2+(x_l^i)^2]dV^i \qquad j\neq k\neq l\neq j ; \quad j,k,l = 1,2,3 \\
\label{eq:inertia_prod} i_{jk} &= \int_{V^i} \rho^ix_j^ix_k^i dV^i \qquad \qquad\qquad j\neq k \qquad j,k = 1,2,3
\end{align}
Gli elementi diagonali della matrice $I_{\theta\theta}$ descritti nell'equazione \ref{eq:inertia_moment} sono detti momenti d'inerzia, mentre i termini non diagonali di \ref{eq:inertia_prod} sono chiamati prodotti d'inerzia. Questi ultimi si annullano qualora il riferimento solidale al corpo sia \emph{principale di inerzia}.
L'energia cinetica di un corpo rigido può essere così scomposta:
\begin{align} \nonumber
T^i_{RR} &= \frac{1}{2}\dot{R}^{i^T}m^i_{RR}\dot{R}^i; \qquad T^i_{r\theta} = \dot{R}^{i^T}m^i_{r\theta}\dot{\theta}^i, \\ \nonumber
T^i_{\theta\theta} &= \frac{1}{2}\dot{\theta}^{i^T}m^i_{\theta\theta}\dot{\theta}^i =  \frac{1}{2} \overline{\omega}^{i^T}\overline{I}^i_{\theta\theta}\overline{\omega}^i
\end{align}
\subsection{Equazioni della dinamica dei corpi rigidi vincolati}
\paragraph{Equazioni della Dinamica}
L'equazione di Lagrange \ref{eq:lagrange_matrix} per i corpi rigidi assume la forma:
\begin{align} \label{eq:lagrange_body} \nonumber
\frac{d}{dt}\left( \frac{\partial T^i}{\partial\dot{q}}\right) -\frac{\partial T^i}{\partial q} -Q^i_e = 0 \\ \nonumber
\frac{d}{dt}\frac{\partial}{\partial \dot{q}}\left( \frac{1}{2}\dot{q}^{i^T}M^i\dot{q}^i \right) -\frac{\partial T^i}{\partial q} -Q^i_e = 0 \\ \nonumber
\frac{d}{dt}\left( M^i\dot{q}^i \right) -\frac{\partial T^i}{\partial q} -Q^i_e = 0 \\ 
 M^i\ddot{q}^i +\dot{M}^i\dot{q}^i -\frac{\partial T^i}{\partial q} -Q^i_e = 0 
\end{align}
Si consideri:
\begin{equation} \label{eq:Qv}
Q_v^i = -\dot{M}^i\dot{q}^i+\frac{\partial T}{\partial q}
\end{equation}
In verità l'espressione si riduce a \cite{shabana2005} : % pag 149 PDF 
\begin{align} \nonumber
Q_v^i = -\dot{M}^i\dot{q}^i+\left(\frac{\partial T}{\partial q}\right)^T =  [0_3^T \qquad -2\overline{\omega}^{i^T} \overline{I}^i_{\theta\theta}\dot{\overline{G}}^i ]
\end{align}
Infine, osservando che le equazioni sopra valgono per ogni corpo \emph{i} si ottiene in forma matriciale:
\begin{equation} \label{eq:dyn_sys_1stline}
M\ddot{q}+C_q^T\lambda = Q_e + Q_v
\end{equation}
con: 
\[M = \begin{bmatrix}
M^1 & \quad & \quad & \quad \\ \quad & M^2 & \quad & 0 \\ 0 & \quad & \ddots & \quad \\ \quad & \quad & \quad & M^{n_b}\end{bmatrix}\]
\[C_q^T = \begin{bmatrix} C^T_{q^1} \\ C^T_{q^2} \\ \vdots \\ C^T_{q^{n_b}} \end{bmatrix}; \qquad
Q_e = \begin{bmatrix} Q_e^1 \\\ Q_e^2 \\ \vdots \\ Q_e^{n_b} \end{bmatrix}; \qquad
Q_v = \begin{bmatrix} Q_v^1 \\\ Q_v^2 \\ \vdots \\ Q_v^{n_b} \end{bmatrix} \]
\textit{Si noti l'aggiunta di un termine vincolare e l'utilizzo del caso con coordinate indipendenti. Si veda il paragrafo seguente per approfondimenti.} \newline
Per considerare le equazioni di vincolo si deriva 2 volte rispetto al tempo l'equazione \ref{eq:Cqt} ottenendo:
\[C_q\ddot{q} + C_{tt} + (C_q\dot{q})_q\dot{q} + C_{qt}\dot{q}\]
Dove i pedici \emph{t} e \emph{q} indicano la derivazione parziale rispetto a tempo e posizione.
Considerando:
\begin{equation} \label{eq:Qd}
Q_d = - C_{tt} - (C_q\dot{q})_q\dot{q} - C_{qt}\dot{q}
\end{equation}
Da cui;
\begin{equation} \label{eq:dyn_sys_2ndline}
C_q\ddot{q} = Q_d
\end{equation}
Le equazioni \ref{eq:dyn_sys_1stline} e \ref{eq:dyn_sys_2ndline} danno luogo al seguente sistema:
\begin{equation} \label{eq:dyn_sys}
\begin{bmatrix} M & C_q^T \\ C_q & 0 \end{bmatrix}
\begin{bmatrix} \ddot{q}\\ \lambda \end{bmatrix} = 
\begin{bmatrix}
Q_e + Q_v \\ Q_d
\end{bmatrix}
\end{equation}
Questo sistema lineare può essere risolto per il vettore delle accelerazioni $\ddot{q}$ e dei moltiplicatori lagrangiani $\lambda$. A partire dalle condizioni iniziali è possibile integrare numericamente le accelerazioni per ottenere velocità e posizioni.
\paragraph{Formulazione Aumentata}
\textbf{Nota:} nella trattazione che porta all'equazione \ref{eq:dyn_sys_1stline} si è eliminato il termine $\delta q$ considerando le coordinate indipendenti. Nel caso di sistema vincolato tuttavia alcuni gradi di libertà saranno dipendenti a causa della presenza dei vincoli. La \emph{formulazione aumentata} \cite{shabana94} permette di trattare indipendentemente le coordinate di sistemi vincolati. \newline
Si considerino le equazioni \ref{eq:lagrange_body} e \ref{jacobian_expansion} ($Q_v$ si consideri compreso nel termine $Q_e$):
\begin{align} \nonumber \ \delta q^T\left[ M\ddot{q} - Q_e \right] = 0 \\ \nonumber
C_q \delta q = 0
\end{align}
Si deduce che  \[ \lambda^T C_q \delta q = 0 \] e quindi 
\[\delta q^T\left[ M\ddot{q} - Q_e +C_q^T\lambda \right] = 0 \]
Il vettore delle posizioni q viene partizionato in coordinate dipendenti ed indipendenti, e si ottiene:
\begin{equation}  [\delta q_i^T \quad \delta q_d^T] \left\{ \begin{bmatrix} M_{ii} & M_{id} \\ M_{di} & M_{dd}\end{bmatrix}
\begin{bmatrix}\ddot{q}_i \\ \ddot{q}_d \end{bmatrix}
- \begin{bmatrix} Q_{ei} \\ Q_{ed} \end{bmatrix}+ \begin{bmatrix}
C_{qi}^T \\ C_{qd}^T\end{bmatrix} \lambda
\right\}   = 0 \end{equation}
Ne consegue che:
\begin{align} \label{eq_aug_for_a}
\delta q_i^T \left[ M_{ii}\ddot{q}_i + M_{id}\ddot{q}_d - Q_{e_i} + C^T_{q_i}\lambda \right] = 0
\\ \label{eq_aug_for_b}
\delta q_d^T \left[ M_{di}\ddot{q}_i + M_{dd}\ddot{q}_d - Q_{e_d} + C^T_{q_d}\lambda \right] = 0
\end{align}
Le coordinate dipendenti possono sempre essere scelte in modo che $C_q$ sia non singolare in quanto le equazioni di vincolo sono linearmente indipendenti. Il vettore dei moltiplicatori lagrangiani sarà quindi la soluzione (unica) del sistema di equazioni \begin{equation}
C^T_{q_d} \lambda = Q_{e_d} - M_{id}\ddot{q}_i - M_{dd}\ddot{q}_d
\end{equation}
Ragion per cui i coefficienti di $\delta q$ nell' equazione \ref{eq_aug_for_b} sono tutti uguali a 0. Nell'equazione \ref{eq_aug_for_a} invece, poiché i $q_i$ sono indipendenti: \begin{equation}
M_{ii} \ddot{q}_i + M_{id}\ddot{q}_d +C^T_{q_i} \lambda = Q_{e_i}
\end{equation}
Combinando le 2 equazioni sopra si ottiene:
\[ \begin{bmatrix}  M_{ii} & M_{id} \\ M_{di} & M_{dd} \end{bmatrix}
\begin{bmatrix}\ddot{q}_i \\ \ddot{q}_d \end{bmatrix} +
\begin{bmatrix} C_{qi}^T \\ C_{qd}^T\end{bmatrix} \lambda = 
\begin{bmatrix} Q_{ei} \\ Q_{ed} \end{bmatrix}
\]
Da cui la forma generale: \begin{equation}
M\ddot{q} + C^T_q = Q_e	\end{equation}
Dando per noto il vettore delle forze esterne $Q_e$, le incognite sono $\ddot{q}$ e $\lambda$. Il numero di incognite è pari al numero di gradi di libertà indipendenti \emph{n} più il totale dei gradi di vincolo $n_c$. per questo viene aggiunta l'equazione \ref{eq:dyn_sys_2ndline}.

### 2020-08-25 templates/Book_Ita/Contenuto/chapter_4.tex
Added BookIta template, Eccomas DRL paper, GVSETS AV paper, CoRL (under review)

\chapter{Definizione e Proprietà dei metodi} \label{chap:cauchy}
\section{Definizione del problema di Cauchy}
Risolvere problemi di dinamica vincolata richiede di risolvere un sistema di equazioni differenziali-algebriche (DAE, \emph{Differential-Algebraic Equations}). La soluzione delle equazioni differenziali-algebriche passa per la soluzione delle equazioni differenziali ordinarie (ODE, \emph{Ordinary Differential Equation})
Il problema di Cauchy ai valori iniziali consiste un problema della forma:\newline
trovare $y(t)\in C^1(I)$ t.c. :
\begin{equation}  \label{eq:cauchy_def}  \begin{cases}
y'(t) = f(t, y(t)) \qquad  t\in I \\
y(t_0) = y_0 
\end{cases}
\end{equation}
La seconda equazione del sistema, ovvero il valore assunto dalla soluzione $y(t)$ in un dato istante $t_0$, è la condizione necessaria per l'unicità della soluzione, infatti Un’equazione differenziale ordinaria ammette in generale infinite soluzioni. Per fissarne una è necessario imporre una condizione che prescriva il valore assunto dalla soluzione in un punto dell'intervallo di integrazione.
La funzione \emph{f(t, y(t))} è assegnata e continua in entrambe le variabili nella striscia $ S = I \times (- \infty , + \infty)$. \newline
\textbf{Osservazione} se f continua in t, integrando \ref{eq:cauchy_def} si ottiene:
\begin{equation} \label{eq:cauchy_integral_equivalence}
y(t) - y_0 = \int^t_{t_0} y'(\tau)d\tau = \int^t_{t_0}f(\tau,y(\tau)) d\tau \end{equation}
L'equivalenza tra problema di Cauchy ed equazione integrala sarà sfruttata in seguito. \newline
\textbf{Nota:} in questa prima parte su tratteranno equazioni differenziali scalari, in seguito verrà esteso quanto riportato al caso di equazioni vettoriali.
%\paragraph{Esistenza ed unicità}
%\paragraph{Stabilità del problema}
\section{Metodi ad un passo} 
Nel seguito si introdurranno metodi numerici per la soluzione del problema di Cauchy. La simbologia utilizzata è la seguente: \begin{itemize}
    \item  $T \in \mathbb{R}_0^+ $ : Intervallo di integrazione da cui $ I = (t_0, t_0+T)$
    \item $h>0$ : ampiezza dei sottointervalli (detta \emph{passo di discretizzazione} o \emph{timestep})  
    \item $  t_n = t_0+nh; \quad n = 0,1,2... N_h$ la successione di nodi che discretizza I. Ovviamente $t_{N_h} \leq t_0+T$. \newline
\end{itemize}{}
\begin{definition} [Metodo single step] \label{def:single_step} 
Un metodo numerico per l'approssimazione del problema \ref{eq:cauchy_def}si definisce \emph{ad un passo} o \emph{single step} se $u_{n+1}$ dipende solo da $u_n$. In caso contrario si parla di metodi a più passi o \emph{multistep}.\end{definition}
\begin{definition} [Metodi impliciti ed espliciti]
Un metodo è detto \emph{esplicito} se $u_{n+1}$ non dipende da valori successivi a $t_n$. In caso contrario, ossia se $u_{n+1}$ dipende implicitamente da sé stesso tramite $f$ il metodo è \emph{implicito} \end{definition}
\paragraph{Metodo di Eulero in avanti (Eulero esplicito, Forward Euler)}
\begin{equation} \label{eq:Eul_Forw}
    u_{n+1} = u_n + h f_n
\end{equation}
\paragraph{Metodo di Eulero all'indietro (Eulero implicito, Backward Euler)}
\begin{equation}
    u_{n+1} = u_n + h f_{n+1}
\end{equation}
In entrambi i metodi di Eulero la derivata prima di $y$ con il rapporto incrementale ed entrambe la approssimazioni alle differenze finite di y sono accurate al primo ordine.
\paragraph{Metodo di Crank-Nicholson (del trapezio)}
\begin{equation}
    u_{n+1} = u_n + \frac{h}{2} \left[f_n+f_{n+1}\right]
\end{equation}
\paragraph{Metodo di Heun}
\begin{equation}
    u_{n+1} = u_n + \frac{h}{2} \left[f_n+f(t_{n+1},u_n,hf_n)\right]
\end{equation}

% \subsubsection{Metodi di Runge-Kutta}
% I metodi Runge-Kutta (RK) per aumentare l'accuratezza si avvalgono di molteplici valutazioni funzionali ad ogni passo. Questa strategia sacrifica la linearità e complica la valutazione dell'errore locale.
% In generale un metodo RK può essere scritto come 
% \begin{equation} \label{eq:RK_method_generalA}
%     u_{n+1} = u_n + hF(t_n, u_n, h; f), \qquad n\geq 0
% \end{equation}
% Dove $F$ è la \emph{funzione di incremento}:
% \begin{align} \label{eq:RK_method_generalB}
%     &F(t_n, u_n, h; f) = \sum_{i=1}^s b_i K_i \\  \label{eq:RK_method_generalC}
%     &K_i = f\left(t_n+c_i h, u_n + h \sum_{j=1}^s a_{ij}K_j\right), \qquad i = 1,2... s
% \end{align}
% In \ref{eq:RK_method_generalB} e \ref{eq:RK_method_generalC} \emph{s} indica il numero di stadi del metodo, mentre i coefficienti $\{a_{ij}\} \{c_i\} \{b_i\}$ definiscono completamente un metodo RK. 
% Per praticità è uso comune raccogliere i coefficienti nella \emph{matrice di Butcher}:
% \begin{equation}
% \begin{array}{c|cccc}
% c_1 & a_{11} & a_{12} & \hdots & a_{1s} \\
% c_2 & a_{21} & a_{22} & \hdots & a_{2s} \\
% \vdots & \vdots & \quad & \ddots & \vdots \\
% c_s & a_{s1} & a_{s2} & \hdots & a_{ss} \\  \hline
% \quad & b_1 & b_2 & \hdots & b_s \\
% \end{array}
% \end{equation}
% Appare evidente dalla \ref{eq:RK_method_generalC} che $K_i$ può essere calcolato esplicitamente solo se $a_{ij} = 0 \forall j \geq i$, e si parlerà in questo caso di \emph{schemi espliciti}. Alternativamente si parla di schemi \emph{impliciti e semi-impliciti} ma questi ultimi non verranno trattati.

\section{Metodi numerici e proprietà}

Nel seguito si indicherà con $u_j$ la soluzione approssimata nel nodo $t_j$ della soluzione esatta $y(t_j)$ o, brevemente, $y_j$. \newline
Allo stesso modo $f(t_j, u_j) = f_j$. Infine, poiché all'istante iniziale si dispone della soluzione esatta, $u_0 = y_0$.

Seconda lo definizione \ref{def:single_step} un metodo \textit{single step} può essere definito in generale come:
\begin{equation}\label{eq:Single_Step}
    u_{n+1} = u_n + \Phi(t_n , h, u_n , u_{n+1} ; f)
\end{equation}
la funzione $\Phi$ è detta \emph{funzione di incremento}, e definisce il valore assunto dall'approssimazione in $t_{n+1}$.
\paragraph{Errore di troncamento}
\begin{itemize}
\item Errore di troncamento locale: \newline

\begin{align} \label{eq:local_trunc_err} \nonumber
u_{n+1} &= u_n + h\Phi(t_n, u_n,f_n;h), \qquad 0\leq n \leq N_h - 1 , \qquad u_0 = y_0 \\
\nonumber y_{n+1} &= y_n + h\Phi(t_n, y_n,f_n;h) +\epsilon_{n+1} , \\ \Rightarrow
\epsilon_{n+1} &= h \tau_{n+1}(h) 
\end{align}
Dove $\epsilon_{+1}$ e $\tau_{n+1}$ sono il \emph{residuo} e l'\emph{errore di troncamento locale} nel nodo $t_{n+1}$.
\item Errore di troncamento globale \newline 
Si definisce \emph{errore di troncamento globale}  la quantità:
\begin{equation} \label{eq:glob_trunc_err}
\tau(h) = \underset{0 \leq n \leq N_{h-1}}{max} \left|\tau_{n+1}(h) \right|
\end{equation}
\end{itemize}
\paragraph{Consistenza}
Un metodo numerico ad un passo è completamente caratterizzato dalla funzione di incremento $\Phi$. Inoltre: 
\begin{equation} \label{eq:incr_limit}
\underset{h\rightarrow 0}{lim} (\Phi(t_n,y_nf(t_n,y_n);h)=f(t_n, y_n),\quad \forall t_n \geq t_0
\end{equation}
Da \ref{eq:incr_limit} e considerando che $y_{n+1}-y_n = hy'(t_n) + \mathcal{O}(h^2)$ si ha che
\begin{equation} \label{eq:consistenza}
\underset{h\rightarrow 0}{lim }(\tau(h)) = 0
\end{equation}
Questa proprietà esprime la consistenza del metodo numerico con il problema di Cauchy.

\paragraph{Zero stabilità}
Un metodo numerico è \emph{zero-stabile} se: \[\exists \quad
h_0, C, \epsilon_0 >0 \quad|\quad \forall h\in (0, h_0], \quad \epsilon \in (0, \epsilon_0], \quad 0\leq n \leq N_h \Rightarrow \] 
\begin{equation} \label{eq:zerostab} \left|z_n^{(h)}-u_n^{(h)} \right|\leq C\epsilon \end{equation}
Con \[\begin{cases}
z_{n+1}^{(h)}&=z_n^{(h)} + h\left[ \Phi (t_n, z_n^{(h)}, f(t_n, z_n^{(h)});h) + \delta_{n+1} \right] \quad n=0..N_{h-1}
\\ z_0^{(h)}&= y_0 + \delta_0 
\end{cases} \]\[ \begin{cases}
u_{n+1}^{(h)}&=u_n^{(h)} + h\Phi (t_n, u_n^{(h)}, f(t_n, u_n^{(h)});h) \quad n=0..N_{h-1}
\\ u_0^{(h)}&= y_0 + \delta_0  \end{cases} \]
La zero stabilità richiede che valga \ref{eq:zerostab} in un intervallo limitato $\forall h\leq h_0$. Questa proprietà è relativa al metodo, non al problema. 
\begin{theorem}[Zero Stabilità] \label{teo:zerostab}
Si consideri un metodo numerico ad un passo per la soluzione del problema di Cauchy. Se la funzione di incremento $\Phi(t_n, u_n,f_n;h)$ è lipschitziana di costante $\Lambda$ rispetto al secondo argomento, uniformemente rispetto ad h e a $t_j \in [t_o, t_0 +T]$ allora il metodo è zero-stabile.  \cite{Quarteroni} \end{theorem}
\paragraph{Convergenza}
Un metodo è \emph{convergente} se: \begin{equation} \label{eq:def_convergenza}
\forall \quad n = 0,1,2....,N_h \qquad \left| u_n - y_n \right| \leq C(h)\end{equation}
$C(h)$ è un infinitesimo di $h$. 
Il metodo di dice \emph{convergente di ordine p} se $C(h) = \mathcal{O}(h^p)$ cioè se $\exists C>0 t.c. C(h) \leq Ch^p \forall h>0$ con $p$ massimo.
\begin{theorem}[Convergenza] \label{teo:converg}
Nelle ipotesi del teorema \ref{teo:zerostab}, per il lemma di Gronwall: \begin{equation}
\left|y_n - u_n\right| \leq \left(\left|y_0-u_0\right| + nh\tau(h)\right)e^{nh\tau}, \qquad 1\leq n\leq 
N_h \end{equation} 
Se vale l'ipotesi di consistenza \ref{eq:consistenza} ed $|y_0-u_0| \rightarrow 0$ per $h\rightarrow 0$ il metodo è convergente.Se $|y_0-u_0| = \mathcal{0}(h^p)$ ed il metodo ha ordine p, allora converge con ordine p \end{theorem}
\textbf{Lemma di Gronwall} \newline
Siano $k_n, \phi_n$ successioni (la prima non-negativa) tali che: \begin{equation}\begin{dcases}
\phi_0 \leq g_0 \\ \phi_n \leq g_0 + \sum_{s=0}^{n-1}p_s + \sum_{s=0}^{n-1}k_s\phi_s \qquad n\geq 1
\end{dcases} \end{equation}
Se $g_0, p_n \geq 0 \forall n \geq 0$ allora:
\begin{equation}
\phi_n \leq \left( g_0 + \sum_{s=0}^{n-1}p_s  \right)
exp \left(\sum_{s=0}^{n-1}k_s \right)
\end{equation}
Si osserva quindi che un metodo (single step) consistente e zero-stabile è necessariamente convergente. Questa proprietà è chiamata \emph{teorema di Lax-Richtmyer o di equivalenza} ed è di estrema rilevanza nell'analisi numerica.

\paragraph{Assoluta stabilità}
L'assoluta stabilità è una proprietà che riguarda il comportamento asintotico del metodo. Un metodo si dice assolutamente stabile se, con $h$ fissato, $u_n$ è limitata per $t_n\rightarrow +\infty$.

Si definisce un \emph{problema modello}
\begin{equation}\label{eq:model_cauchyprob}
    \begin{cases} y'(t) = \lambda y(t) \qquad t>0 \\ y(0) = 1 \end{cases} \end{equation}
La soluzione di \ref{eq:model_cauchyprob} è $y = e^{ \lambda t}$, inoltre se $\mathcal{R}(\lambda)<0$, $u_n\rightarrow 0$
% \subsubsection{Proprietà dei metodi di Runge-Kutta}
% \paragraph{Consistenza}
% Si definisce l'errore di troncamento locale a partire dalla formula del residuo:
% \begin{equation}
%     h \tau_{n+1}(h) = y_{n+1} - y_n -hF(t_n, y_n, h; f)
% \end{equation}
% Si verifica che il metodo è consistente, ossia $\tau(h) = max_n|\tau_n(h)| \rightarrow 0$ per $h\rightarrow 0$, se e solo se:
% \begin{equation}
%     \sum_{i=1}^s b_i = 1
% \end{equation}
% \paragraph{Convergenza}
% Trattandosi di un metodo single step, per il teorema \ref{teo:converg} la consistenza implica la convergenza, pertanto avrà le medesime condizioni.

\subsection{Proprietà dei metodi}
\begin{definition} [Assoluta Stabilità]
Un metodo è assolutamente stabile se l'approssimazione della soluzione di \ref{eq:model_cauchyprob}
\begin{equation} \label{eq:abs_stab}
    |u_n| \rightarrow 0 \qquad \text{per} \qquad t_n \rightarrow +\infty
\end{equation}
\end{definition}

Si definisce \emph{problema di Cauchy lineare} o \emph{problema modello}
\begin{equation}  \label{eq:cauchy_def}  \begin{cases}
y'(t) = \lambda y(t) \qquad  t > 0 \\
y(t_0) = 1 
\end{cases}
\end{equation}

Dq questa definizione si può inoltre definire \emph{regione di assoluta stabilità} del metodo il sottoinsieme di $\mathbb{C}$ 
\begin{equation}
    \mathbb{A} = \{ z=h\lambda \in \mathbb{C} \qquad \text{t.c. \ref{eq:abs_stab} è soddisfatta} \}
\end{equation}
\begin{definition}[Metodi A-stabili e condizionatamente stabili]
Un metodo è detto \emph{A-stabile} se $\mathcal{A} \cap \mathbb{C}^- = \mathbb{C}^-$ ovvero se \ref{eq:abs_stab} è soddisfatta $\forall\lambda\in \mathbb{C}^-$ incondizionatamente rispetto ad $h$, mentre qualora siano richiesta condizioni su h sono condizionatamente stabili. 
I metodi $\theta$ stabili, invece, sono metodi Multistep per i quali A contiene la regione angolare definita dagli $z \in \mathcal{C}$ tali che $-\theta < \pi - arg(z) < \theta$ con $\theta \in  (0, \pi/2)$.
\end{definition}
\begin{theorem}[Prima Barriera di Dahlquist]
Non esistono metodi multistep lineari zero-stabili a q-passi con ordine maggiore di q + 1 se q è dispari, q + 2 se q è pari.
\end{theorem}
\begin{theorem}[Seconda Barriera di Dahlquist]\label{theo:DahlquistII}
Un metodo multistep lineare esplicito non può essere né A-stabile, né $\theta$-stabile. Inoltre, non esistono metodi multistep lineari A-stabili di ordine superiore a due. Per ogni
$\theta \in (0, \pi/2)$ esiste almeno un metodo multistep lineare $\theta$-stabile a q passi di
ordine q solo per q = 3 e q = 4.
\end{theorem}
Le barriere di Dahlquist limitano allo stesso modo i metodi ad un passo, considerabili come casi di metodi multistep con $n=1$.

\section{Sistemi di equazioni}
Il sistema di ODE del primo ordine:
\begin{equation}
 \avect{y}' = \amatr{F}(t, \avect{y})
\end{equation}
Con $\avect{y}\in \mathbb{R}^n$ vettore soluzione e $\amatr{F}:\mathbb{R} \times \mathbb{R}^n \rightarrow \mathbb{R}^n$. La soluzione dipende sempre dal valore iniziale che in questo caso è un vettore di $n$ elementi $\avect{y}_0 \in \mathbb{R}^n$.
\paragraph{Lipschitzianità ed esistenza della soluzione}
Se $\amatr{F} è Lipschitziana $ ossia se , essendo continua su $D = [t_0,T]\times \mathbb{R}^n$, con $t_o, T$ finiti esiste $L$ tale che:
\begin{equation}
    \norm{\amatr{F}(t,\avect{y}) - \amatr{F}(t,\overline{\avect{y}})} \leq L \norm{\avect{y}- \overline{\avect{y}}}
\end{equation}
per ogni $(t, \avect{y}), (t, \overline{\avect{y}}) \in D$; si dimostra che esiste ed è unica la soluzione $\avect{y}$  continua in $D$ del problema \ref{eq:cauchy_def}.

Per quanto riguarda le soluzioni numeriche del detto problema i metodi visti (e quelli riportati in seguito) sono estendibili al caso di sistemi di equazioni.

### 2020-08-25 templates/Book_Ita/Contenuto/chapter_5.tex
Added BookIta template, Eccomas DRL paper, GVSETS AV paper, CoRL (under review)

\chapter{Metodi numerici specifici}
Verranno di seguito illustrati metodi numerici per la soluzione del problema di Cauchy espressamente ideati per l'applicazione a simulazioni multibody.

Dal momento che d'ora in avanti si farà specifico riferimento alla dinamica di sistemi meccanici si riprenderà la notazione utilizzata nella parte \ref{part:LagrDyn}. In luogo dei vettori $\avect{y}$ ed $\avect{y}'$ si avranno $\{\avect{q} , \avect{v}\}$ e $\{\avect{v} , \avect{a}\}$; rispettivamente posizioni e velocità e velocità ed accelerazioni del sistema. Sempre riprendendo la notazione si avrà \begin{itemize}
    \item $\amatr{C} = \amatr{C}(\avect{q},t)$ vettore delle equazioni vincolari
    \item $\amatr{C_q} =\frac{\partial\avect{C}}{\partial q}$
    \item $\amatr{M}$ matrice di massa
    \item $\avect{\lambda}$ vettore di moltiplicatori lagrangiani (reazioni vincolari)
    \item $\avect{f}$ forze agenti sul sistema
    \item $l$ indica lo step di integrazione
    \item Con $h \in \mathbb{R}$ si indica sempre l'ampiezza del \emph{timestep}.

\end{itemize}

Riguardo le equazioni della dinamica dei sistemi verranno utilizzate la prima equazione della formulazione lagrangiana \ref{eq:dyn_sys_1stline} e l'equazione delle condizioni di vincolo \ref{eq:Cqt}. Dato lo stato del sistema è possibile trovare accelerazioni e reazioni vincolari grazie a queste relazioni.
A proposito dell'equazione della dinamica dei sistemi vincolati \ref{eq:dyn_sys_1stline} essa può essere riscritta come: \[\amatr{M}\avect{a} + \amatr{C}_q^T\avect{\lambda} = \avect{f} \] Dove il vettore delle forze applicate $f$ comprende sia le forze esterne $Q_e$ che il termine cosiddetto giroscopico $Q_v$. Si consideri inoltre che, a seconda della letteratura di riferimento, l'equazione \ref{eq:lambda} che definisce $\lambda$ può avere o no il segno meno. Nella implementazione seguente si è scelta la seconda opzione, ragion per cui si scriverà d'ora in avanti:
\begin{equation}\label{eq:dynamics}
    \amatr{M}\avect{a}-\amatr{C}_q^T\avect{\lambda}-\avect{f} = 0
\end{equation}

Inoltre nel corso del capitolo verrà menzionato un fenomeno chiamato smorzamento numerico (\textit{numerical damping}), un errore numerico che si manifesta come la presenza di uno smorzamento fittizio. Si fornisce una trattazione più dettagliata del fenomeno in sez. \ref{sec_numdamp}.
 \section{Newmark Method}
 Questo metodo nasce per studiare la risposta dinamica di strutture e solidi modellati con gli elementi finiti e per questa ragione è assai diffuso nella comunità FEM. %La formulazione si basa sul Teorema di Lagrange, ovvero sull'esistenza di un $\avect{a}_{\beta}\quad t.c. \quad \avect{v}^{l+1} = \avect{v}^l h\avect{a}_{\beta}$ con $\avect{a}^l \leq \avect{a}_{\beta} \leq\avect{a}^{l+1}$

 \begin{align} \label{eq:NewmarkDef}
 \amatr{M}\avect{a}^{l+1}-(\amatr{C}_q^t\avect{\lambda}+\avect{f})^{l+1}&=0 \\
 \avect{v}^{l+1}- \avect{v}^l -h \left[(1-\gamma)\avect{a}^l+\gamma \avect{a}^{l+1}\right]&=0 \\
\avect{q}^{l+1}-\avect{q}^l - h \avect{v}^l-\frac{h^2}{2}\left[(1-2\beta)\avect{a}^l+2\beta\avect{a}^{l+1}\right]&=0
 \end{align}
 %È un metodo del second'ordine, implicito ed incondizionatamente stabile per determinati valori dei coefficienti che lo caratterizzano. 
 Caratteristiche del metodo \cite{newmark59}:
\begin{itemize}
    \item si basa sul Teorema di Lagrange, ovvero sull'esistenza di un $\avect{a}_{\beta}\quad $tale che $ \quad \avect{v}^{l+1} = \avect{v}^l + h\avect{a}_{\beta}$ con $\avect{a}^l \leq \avect{a}_{\beta} \leq\avect{a}^{l+1}$
    \item $\gamma$ è normalmente compreso tra 1/2 ed 1
    \item $\gamma$ è solitamente posto uguale ad 1/2, unico valore per il quale si ottiene ordine di convergenza quadratico. 
    \item $\beta$ è compreso tra 0 ed 1.
    \item per $\beta =1/4, \gamma= 1/2$ si ha il metodo dell'accelerazione costante.
    \item per $\beta =1/6, \gamma= 1/2$ si ha il metodo dell'accelerazione lineare.
    \item Per $\gamma \geq 1/2$ e $\beta \geq  \frac{(\gamma + \frac{1}{2})^2}{4}$ il metodo è incondizionatamente stabile.
    \item Il metodo dell'accelerazione costante è incondizionatamente stabile.
    \item Lo smorzamento numerico è nullo per $\gamma = \frac{1}{2}$ mentre si manifesta per valori maggiori, crescendo  con $\gamma$

\end{itemize}
 \section{HHT Method}
 Il metodo Hilber-Hughes-Taylor è una generalizzazione del metodo di Newmark che aggiunge un ulteriore parametro $\alpha$ \cite{Negrut2006}. 
 \begin{align}
\amatr{M}\avect{a}^{l+1}-(1+\alpha)(\amatr{C}_q^t\avect{\lambda}+\avect{f})^{l+1}+\alpha(\amatr{C}_q^T\avect{\lambda}+f)^l&=0 \\
\avect{v}^{l+1}-\avect{v}^l-h[(1-\gamma)\avect{a}^l+\gamma\avect{a}^{l+1}] &=0 \\
\avect{q}^{l+1}-\avect{q}^l-h\avect{v}^l-\frac{h^2}{2}[(1-2\beta)\avect{a}^l+2\beta\avect{a}^{l+1}] &=0
 \end{align}
 Caratteristiche del metodo \cite{Hilber1977} \cite{Hughes1983}:
\begin{itemize}
    \item Il parametro $\alpha$ permette di controllare la dissipazione numerica. Per $\alpha = 0$ il metodo coincide con Newmark, al decrescere del parametro vengono smorzate le frequenze più alte.
    \item Basandosi sull'esperienza la letteratura suggerisce di porre $0\leq \alpha \leq1/3$ , $\beta = (1-\alpha)^2/4 $ e $\gamma = 1/2 - \alpha$.  
    \item A-stabile (per i valori dei parametri riportati sopra)
    \item Ordine di convergenza 2 (per i valori dei parametri riportati sopra)
\end{itemize}
\section{Generalized-$\alpha$ Method}
\paragraph{Confronto fra i metodi in termini di convergenza e stabilità}
Dalla seconda barriera di Dahlquist (Teorema \ref{theo:DahlquistII}) sappiamo che non esistono metodi multistep lineari A-stabili di ordine superiore a due. Il metodo dei trapezi utilizzato per DAE (\emph{Differetial-Algebraic Equations}) dà luogo a reazioni vincolari oscillanti instabili se utilizzato per sistemi vincolati. Al contrario i metodi HHT e generalized-$\alpha$, introdotto nel seguito, sono tra i pochi integratori di ordine 2 che possono risolvere DAE vincolate.

Il metodo generalized-$\alpha$ \cite{generalizedAlphaHulbert93} \cite{JayNegrut2008} è un'evoluzione del metodo HHT, nel quale tutti i parametri vengono settati a partire da un solo parametro $\rho_\infty$. È un metodo del second'ordine come accennato in precedenza, e permette il controllo della dissipazione numerica.

Si definisce come segue:

\begin{align} \label{eq:Gen-Alpha1}
\amatr{M}\avect{a_*}^{l+1}-(\amatr{C}_q^t\avect{\lambda}+f)^{l+1}&=0 \\ \label{eq:Gen-Alpha2}
(1-\alpha_m)\avect{a}^{l+1}+\alpha_m\avect{a}^l = (1-\alpha_f)\avect{a_*}^{l+1} + \alpha_f\avect{a_*}^l&=0 \\
\avect{v}^{l+1}-\avect{v}^l-h[(1-\gamma)\avect{a}^l+\gamma\avect{a}^{l+1}] &=0 \\
\avect{q}^{l+1}-\avect{q}^l-h\avect{v}^l-\frac{h}{2}[(1-2\beta)\avect{a}^l+2\beta\avect{a}^{l+1}] &=0
\end{align}

Come anticipato tutti i parametri elencati vengono settati a partire da un unico parametro $\rho_\infty \in [0,1]$:
\begin{align}
    \alpha_m &= \frac{2\rho_\infty-1}{1+\rho_\infty} \\
    \alpha_f &= \frac{\rho_\infty}{1+\rho_\infty}   \\
    \beta &= \frac{1}{4}\left(\gamma+\frac{1}{2} \right)^2 \\
    \gamma &= \frac{1}{2}+\alpha_f - \alpha_m
\end{align}

Si noti che: \begin{itemize}
    \item Per $\rho_\infty=0$ si ha massima dissipazione asintotica ed il metodo ha smorzamento massimo.
    \item Il metodo HHT è come il metodo generalized-$\alpha$ con $\alpha_m =0$ e $\alpha_f = \alpha$. Effettivamente i metodi hanno proprietà simili. Per verificarlo si sostituisca $\avect{a_*}$ calcolato in $l$ ed $l+1$ secondo
    \ref{eq:Gen-Alpha1} in  \ref{eq:Gen-Alpha2} ($\amatr{M}$ è sempre invertibile).
    \item Per $\rho_\infty=1$ d'altra parte non c'è dissipazione numerica, come nel metodo dei trapezi. A differenza di questo, tuttavia, il metodo generalized-$\alpha$ si comporta bene in presenza di vincoli. 
\end{itemize}
Nella pratica uno smorzamento numerico è quasi sempre inserito anche se il sistema fisico non è dissipativo, onde evitare divergenze in simulazioni prolungate. Di solito si utilizza un valore intermedio di $\rho_\infty$, che aiuta a scartare frequenze di oscillazione elevate di scarso interesse. Solitamente il valore di $\rho_\infty$ viene trovato euristicamente, partendo da valori prossimi a 0 ed aumentandolo gradualmente fino alla comparsa di armoniche a frequenza elevata.

### 2020-08-25 templates/Book_Ita/Contenuto/chapter_6.tex
Added BookIta template, Eccomas DRL paper, GVSETS AV paper, CoRL (under review)

\chapter{Applicazione dei metodi alle simulazioni multibody}
In questo capitolo si vedrà come applicare i metodi numerici visti nel capitolo \ref{chap:cauchy} alle simulazioni multibody.

Si noti che nella definizione del problema di Cauchy (eq. \ref{eq:cauchy_def}) la funzione $f(t, y(t))$ viene considerata assegnata. Questo non accade in una simulazione, dal momento che lo stato del sistema non è noto a priori. \newline
Si consideri il sistema \ref{eq:dyn_sys}: la matrice dei coefficienti ed il vettore dei termini noti dipendono dallo stato futuro, e non è pertanto possibile calcolare direttamente $\avect{\Ddot{q}}$ e $\avect{\lambda}$ nel timestep futuro $t+1$ come richiesto dai metodi impliciti. \newline A questi algoritmi andrà pertanto abbinato il metodo di Newton per i sistemi.

\section{Cenni al metodo di Newton-Raphson o Metodo delle Tangenti}
\subsection{Metodo di Newton per la soluzione di equazioni non lineari}
Il metodo di Newton \cite{Quarteroni} può essere utilizzato per approssimare numericamente gli zeri di una funzione di una variabile reale, ossia data \(f:\mathbb{R} \rightarrow \mathbb{R}\) si cerca $\alpha\in \mathbb{C}\quad t.c.\quad f(\alpha)=0 $ \newline
Se $f \in C^1([a,b])\quad con \quad [a,b]\subset\mathbb{R} \quad t.c. \quad f(a)f(b) < 0 \quad e \quad f'(x) \neq 0 \forall x \in [a,b]$ si può procedere iterativamente al calcolo della radice una volta assegnato il valore iniziale $x^{0}$:
\begin{equation} \label{eq:Newton_1d}
    x^{(k+1)} = x^{(k)} - \frac{f\left(x^{(k)}\right)}{f'\left(x^{(k)}\right)}
\end{equation}
Sebbene il metodo richieda 2 valutazioni funzionali controbilancia l'onere computazionale con un ordine di convergenza 2 (quando la radice è semplice).
\begin{figure}[h!]
\centering
\includegraphics[height=8cm]{newtonRaphsonMethod.png}
\caption{Rappresentazione grafica per i passi del metodo di Newton}
 \label{fig:Newton_1d}
\end{figure}
\subsection{Metodo di Newton per la soluzione di sistemi}
Il metodo di Newton può essere esteso al caso di sistemi di equazioni. 
Il metodo può essere esteso al caso di sistemi di equazioni. Indicando con l'apice n l' n-esimo passaggio, con $\amatr{F}$ il sistema di equazioni e con $\amatr{J}$ il jacobiano di quest'ultimo si procede dato $\avect{x}{(0)}$ come segue:
\begin{align} \label{eq:NewtonRaphsonA}
    \amatr{J}\left( \avect{x}^{(n)} \right) \avect{\delta x}^{(n)} = - \amatr{F}\left( \avect{x}^{(n)} \right) \\
    \label{eq:NewtonRaphsonB}  \avect{x}^{(n+1)} = \avect{x}^{(n)} + \avect{\delta x}^{(n)} 
\end{align}
L'onere computazionale di questo metodo consiste nel risolvere ad ogni passaggio un sistema lineare di matrice $\amatr{J}$.

\section{Smorzamento Numerico} \label{sec_numdamp}
Lo smorzamento numerico (\textit{numerical damping}) indica un errore numerico, che si manifesta come la comparsa di uno smorzamento fittizio, tipica dei metodi impliciti ed, in taluni casi, semi-impliciti. Questo fenomeno è ben osservabile in un sistema massa-molla senza smorzatore, le cui oscillazioni dovrebbero mantenersi ad ampiezza costante, mentre diminuiscono nel tempo. Questo fenomeno si accentua al diradarsi del \textit{timestep}.
\subsection{Esempio numerico: smorzamento numerico in un sistema massa-molla implementato in Matlab}
Per osservare quanto affermato si riporta un esempio numerico del sistema massa molla. Con la solita notazione possiamo scrivere :\[\begin{cases}\frac{v^{l+1}-v^l}{h}M = f^{l+1}\\ q^{l+1}= q^l+v^{l+1}h \end{cases}\]
Nel caso del sistema massa-molla si avrà : \[f^l = Mg-Kq^l \]O, in presenza di smorzatore \[ f^l = Mg-Kq^l-Cv^l\]
Da cui:  \[ \left(v^{l+1}-v^l\right)M = h\left(  Mg-Kq^{l+1} -Cv^{l+1} \right)  \]
Sostituendo il valore di $q^{+1}$ si ottiene:
\[ \left(v^{l+1}-v^l\right)M = h\left(  Mg-K(q^l+v^{l+1}h) -Cv^{l+1} \right)  \]
Si può quindi scrivere:
\[ v^{l+1}\left(M+Kh^2+Ch\right) = Mv^l + h\left(  Mg-Kq^l \right)  \]
Possiamo quindi esprimere $v^{l+1}$
\[ v^{l+1} = \frac{ Mv^l + Mgh -Khq^l }{\left(M+Kh^2+Ch\right)}  \]
A questo punto si procede ad implementare in Matlab un algoritmo che risolve numericamente il sistema con \textit{timestep} crescente così da poter osservare l'andamento della smorzamento numerico. Si riporta la porzione di codice più significativa, ovvero il ciclo \textit{while} che calcola velocità e posizione nell'istante successivo e le memorizza.

\noindent\rule{\textwidth}{1pt}\newline  $ 
    while \  t<= Tmax \\
        vl1 = (m*v+m*g*dt-k*x*dt)/(m+k*dt^2+c*dt); \\
        xl1 = x + dt*vl1; \\
        t = t+ dt; \\
        v = vl1; \\
        x = xl1; \\
        j = j+1; \\
        v_{vett}(j) = vl1; \\
        x_{vett}(j) = xl1; \\
        t_{vett}(j) = t; \\
    end$ \newline
\noindent\rule{\textwidth}{1pt}
Di seguito sono riportati i grafici di spostamento e velocità della simulazione del sistema massa-molla che utilizza il loop riportato. Sebbene sia presente un termine di smorzamento esso è in realtà nullo.

\begin{figure}[ht]
\centering
\includegraphics[height=5cm]{Figure/position.pdf}
\caption{Posizione della massa nel sistema massa-molla in funzione del tempo al variare del timestep}
 \label{fig:Newton_1d}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[height=5cm]{Figure/speed.pdf}
\caption{Velocità della massa nel sistema massa-molla in funzione del tempo al variare del timestep}
 \label{fig:Newton_1d}
\end{figure}

Ciononostante si nota una progressiva diminuzione dell'ampiezza delle oscillazioni, più accentuata all'aumentare del \textit{timestep}, mentre in condizioni di smorzamento nullo la soluzione analitica prevederebbe oscillazioni ad ampiezza costante.

Lo smorzamento fittizio (in quanto derivato da un errore numerico e senza alcun riscontro fisico) osservabile è la manifestazione dello smorzamento numerico in questione. Infatti l'unica differenza tra le soluzioni è il progressivo aumento del \textit{timestep}.

\section{Algoritmi di soluzione per metodi impliciti}
\paragraph{Eulero implicito}
Di seguito si illustra l'applicazione del metodo di integrazione di Eulero implicito (Backward Euler) ad un sistema multibody in presenza di vicoli.
Si riprende l'equazione \ref{eq:dyn_sys}. Come accennato in precedenza non è possibile conoscere direttamente il valore delle forze nello stato $l+1$. Per pervenire ad una soluzione è necessario includere l'integrazione numerica all'interno di un'iterazione di Newton-Raphson.

Per un sistema multibody il metodo di Eulero implicito può essere scritto come il seguente sistema che chiameremo \textbf{G}
\begin{align} \label{eq:eulforw_Gq}
    \avect{q}^{l+1}-\avect{q}^l - \avect{v}^{l+1}h &= 0 \\ \label{eq:eulforw_Gv}
    \amatr{M}(\avect{v}^{l+1} -\avect{v}) - h\avect{f}^{l+1} - h\amatr{C}_q^T \avect{\lambda}^{l+1} &= 0 \\ \label{eq:eulforw_Glam}
    \amatr{C}(\avect{q}^{l+1} , t^{l+1}) &= 0
\end{align}
Per procedere alla soluzione iterativa con Newton come anticipato si deve esprimere lo Jacobiano del sistema \textbf{G}:
\begin{equation} \label{eq:eulforwjacob} \amatr{J} =
    \begin{bmatrix} I & -hI   & 0 \\
    -h\nabla_q\avect{f}^{l+1} & \amatr{M}-h\nabla_v\avect{f}^{l+1} &  -h\amatr{C}_q^T \\
    \amatr{C}_q               &  0                                 &    0  \end{bmatrix}
\end{equation}
Utilizzando il procedimento illustrato in \ref{eq:NewtonRaphsonA} si scrive:
\begin{equation} \label{eq:eulforw_iterstep}
    \amatr{J} \left\{\begin{array}{c}\Delta\avect{q}^{l+1} \\ \Delta\avect{v}^{l+1} 
    \\ \Delta\avect{\lambda}^{l+1} \end{array} \right\} = -\amatr{G} \end{equation}
Dalla prima riga dell'equazione \ref{eq:eulforw_iterstep}:
\[ \Delta\avect{q}^{l+1} = h\Delta\avect{v}^{l+1} -\avect{q}^{+1} +\avect{q}^l + h\avect{v}^{l+1} \]
E, differenziando l'equazione \ref{eq:eulforw_Gq} si ottiene 
\begin{equation}
    \Delta \avect{q}^{l+1} = h \Delta \avect{v}^{l+1}
\end{equation}

Questa approssimazione è vera solo in prossimità della soluzione ma viene comunemente usata per evitare la complicazione che proviene dal considerare tutti i termini del sistema.
Grazie a questa relazione possiamo eliminare la dipendenza di $\avect{q}^{l+1}$ da $\Delta \avect{q}^{l+1}$, infatti la prima equazione di \ref{eq:eulforw_iterstep} si semplifica divenendo identica alla \ref{eq:eulforw_Gq} e il sistema \ref{eq:eulforw_iterstep} diventa:

\begin{equation}
        \begin{bmatrix} I & -hI   & 0 \\
    -h\nabla_q\avect{f}^{l+1} & \amatr{M}-h\nabla_v\avect{f}^{l+1} &  -h\amatr{C}_q^T \\
    \amatr{C}_q               &  0                                 &    0  \end{bmatrix}
    \left\{\begin{array}{c}h\Delta\avect{v}^{l+1} \\ \Delta\avect{v}^{l+1} 
    \\ \Delta\avect{\lambda}^{l+1} \end{array} \right\}
    = -\amatr{G}
\end{equation}
La seconda e terza equazione possono essere riscritte come segue, la prima è riportata in fondo e vengono aggiunti i passaggi del procedimento di Newton Raphson \ref{eq:NewtonRaphsonB} per $v$ d $\lambda$.
\[
    \begin{bmatrix}\left[ M-h^2\nabla_q\avect{f}^{l+1}-h\nabla_v\avect{f}^{l+1} \right] & \amatr{C}_q^T \\\amatr{C}_q &0 \end{bmatrix} \left\{\begin{array}{c} \Delta\avect{v}^{l+1}  \\ 
    -h\Delta\avect{\lambda}^{l+1} \end{array} \right\} = \] \begin{equation} \label{eq:eulfor_NRiter1} =
    \left\{\begin{array}{c} (\avect{v}^l-\avect{v}^{l+1})\amatr{M}+h\avect{f}^{l+1}+h\amatr{C}_q^T \avect{\lambda}^{l+1}\\ -\frac{\amatr{C^{l+1}}}{h} \end{array} \right\}
\end{equation}
\begin{equation}
        \avect{v}_{n+1}^{l+1} = \avect{v}_n^{l+1} + \Delta\avect{v}^{l+1}
\end{equation}
\begin{equation}
        \avect{\lambda}_{n+1}^{l+1} = \avect{\lambda}_n^{l+1} + \Delta\avect{\lambda}^{l+1}
\end{equation}
\begin{equation} \label{eq:eulfor_NRiter4}
        \avect{q}^{l+1} = \avect{q}^l + h\avect{v}_{n+1}^{l+1}
\end{equation}
Si noti che \begin{itemize}
    \item L'indice dell'iterazione è passato al pedice in quanto l'apice è già utilizzato per indicare il timestep.
    \item $\avect{q}$ non dipende da $\Delta\avect{q}$ ma viene calcolato direttamente dato il valore aggiornato di $\avect{v}^{l+1}$ come espresso in \ref{eq:eulfor_NRiter4}
    \item Nella seconda equazione del sistema \ref{eq:eulfor_NRiter1} sono stati divisi entrambi i membri per $h$
\end{itemize}    
Questi passaggi sono implementati nel solver di Chrono e verranno analizzati nel seguente capitolo i dettagli implementativi.

\paragraph{Metodo dei trapezi }
Il procedimento è analogo a quanto visto per il metodo di Eulero, e la stessa cosa può essere detta per i metodi illustrati in seguito. Le differenze risiedono ovviamente nella definizione degli step di integrazione dando luogo a differenze nel sistema da implementare. Nello specifico il nuovo sistema \textbf{G} diventa:
\begin{align} \label{eq:trap_Gq}
    \avect{q}^{l+1}-\avect{q}^l - \frac{\avect{v}^{l+1}-\avect{v}^l}{2} h &= 0 \\ \label{eq:trap_Gv}
    \amatr{M}(\avect{v}^{l+1} -\avect{v}) - \frac{h}{2}(\avect{f}^{l+1} + \avect{f}^l )
    - \frac{h}{2}({\amatr{C}_q^{l+1}}^T \avect{\lambda}^{l+1} + {\amatr{C}_q^l}^T \avect{\lambda}^l) &= 0 \\ \label{eq:trap_Glam} \amatr{C}(\avect{q}^{l+1} , t^{l+1}) &= 0
\end{align}
Mentre in questo caso si avrà:
\begin{equation}
    \Delta \avect{q}^{l+1} = \frac{h}{2} \Delta \avect{v}^{l+1}
\end{equation}

Seguendo i passaggi effettuati per il metodo di Eulero:

\begin{align} \nonumber
\amatr{J} \left\{\begin{array}{c}\Delta\avect{q}^{l+1} \\ \Delta\avect{v}^{l+1}  
        \\ \Delta\avect{\lambda}^{l+1} \end{array} \right\} = -\amatr{G}
\\
\begin{bmatrix} I & -\frac{h}{2}I   & 0 \\
-\frac{h}{2}\nabla_q\avect{f}^{l+1} & \amatr{M}-\frac{h}{2}\nabla_v\avect{f}^{l+1} &-\frac{h}{2}\amatr{C}_q^T \\
    \amatr{C}_q^{l+1}              &  0                                 &    0  \end{bmatrix}
    \left\{\begin{array}{c}\frac{h}{2}\Delta\avect{v}^{l+1} \\ \Delta\avect{v}^{l+1} 
    \\ \Delta\avect{\lambda}^{l+1} \end{array} \right\}
    = - \amatr{G}
\end{align}
Da cui, sviluppando i calcoli ed aggiungendo gli incrementi di $\avect{v}$ e $\avect{\lambda}$:

\[
    \begin{bmatrix}\left[ M-\frac{h^2}{4}\nabla_q\avect{f}^{l+1}-\frac{h}{2}\nabla_v\avect{f}^{l+1} \right] 
    & {\amatr{C}_q^{l+1}}^T \\ \amatr{C}_q^{l+1} & 0 \end{bmatrix} 
    \left\{\begin{array}{c} \Delta\avect{v}^{l+1} \\ -\frac{h}{2}\Delta\avect{\lambda}^{l+1} \end{array}\right\}=\] 
    \begin{equation} \label{eq:eulfor_NRiter1} =
    \left\{\begin{array}{c} (\avect{v}^l-\avect{v}^{l+1})\amatr{M}+\frac{h}{2}(\avect{f}^{l+1} + \avect{f}^l {\amatr{C}_q^{l+1}}^T \avect{\lambda}^{l+1} + {\amatr{C}_q^l}^T \avect{\lambda}^l)\\ -\frac{\amatr{C^{l+1}}}{h} \end{array} \right\}
\end{equation}
\begin{equation}
        \avect{v}_{n+1}^{l+1} = \avect{v}_n^{l+1} + \Delta\avect{v}^{l+1}
\end{equation}
\begin{equation}
        \avect{\lambda}_{n+1}^{l+1} = \avect{\lambda}_n^{l+1} + \Delta\avect{\lambda}^{l+1}
\end{equation}
\begin{equation} \label{eq:eulfor_NRiter4}
        \avect{q}^{l+1} = \avect{q}^l + \frac{h}{2}(\avect{v}^l+\avect{v}_{n+1}^{l+1})
\end{equation}

\paragraph{Newmark}

Procedendo similmente a quanto visto nei casi precedenti si può scrivere \cite{negrutGavrea2005}:
\begin{align} \nonumber
    \begin{bmatrix}
    \amatr{H} & \overline{\amatr{C}_q}^T \\ \overline{\amatr{C}_q} & 0 
    \end{bmatrix}
    \begin{Bmatrix}
    \Delta \avect{a}^{l+1} \\ \Delta \avect{\lambda}^{l+1}
    \end{Bmatrix}  = \\
    \begin{Bmatrix}
        \amatr{M} \avect{a}^{l+1} - (\amatr{C}_q^T \avect{\lambda}+\avect{f})^{l+1} \\
        -\frac{1}{\beta h^2}\avect{C}^{l+1}
    \end{Bmatrix} \\
    \avect{a}_{n+1}^{l+1} = \avect{a}_n^{l+1} + \Delta a^{l+1} \\
    \avect{\lambda}_{n+1}^{l+1} = \avect{\lambda}_n^{l+1} + \Delta \lambda^{l+1} \\
    \avect{v}^{l+1} = \avect{v}^l +h \left[ (1-\gamma)\avect{a}^l + \gamma \avect{a}^{l+1} \right] \\
    \avect{q}^{l+1} = \avect{q}^l +h\avect{v}^l+\frac{h^2}{2}[(1-2\beta)\avect{a}^l+2\beta\avect{a}^{l+1}]
\end{align}
Dove $\amatr{H}$  è una matrice che raccoglie vari termini:
\begin{equation}
    \amatr{H} = \left[ \amatr{M}-\gamma h \nabla_v\avect{f}^{l+1} -\beta h^2 \nabla_q\avect{f}^{l+1} +
    \beta h^2 \left[ (\amatr{M} \avect{a})_q + (\amatr{C}_q^T \avect{\lambda})_q \right] \right]
\end{equation}

\paragraph{HHT} 
Per il metodo HHT \cite{Wang15HHT} \cite{JayNegrut2007}, invece, si ottiene:

\begin{align} \nonumber
    \begin{bmatrix}
    \amatr{H} & \overline{\amatr{C}_q}^T \\ \overline{\amatr{C}_q} & 0 
    \end{bmatrix}
    \begin{Bmatrix}
    \Delta \avect{a}^{l+1} \\ \Delta \avect{\lambda}^{l+1}
    \end{Bmatrix}  = \\
    \begin{Bmatrix}
        \frac{1}{1+\alpha} \amatr{M} \avect{a}^{l+1} - (\amatr{C}_q^T \avect{\lambda}+\avect{f})^{l+1} 
        +\frac{\alpha}{1+\alpha}(\amatr{C}_q^T \avect{\lambda}+\avect{f})^l \\
        -\frac{1}{\beta h^2}\avect{C}^{l+1}
    \end{Bmatrix} \\
    \avect{a}_{n+1}^{l+1} = \avect{a}_n^{l+1} + \Delta a^{l+1} \\
    \avect{\lambda}_{n+1}^{l+1} = \avect{\lambda}_n^{l+1} + \Delta \lambda^{l+1} \\
    \avect{v}^{l+1} = \avect{v}^l +h \left[ (1-\gamma)\avect{a}^l + \gamma \avect{a}^{l+1} \right] \\
    \avect{q}^{l+1} = \avect{q}^l +h\avect{v}^l+\frac{h^2}{2}[(1-2\beta)\avect{a}^l+2\beta\avect{a}^{l+1}]
\end{align}
Dove $\amatr{H}$  è una matrice che raccoglie vari termini:
\begin{equation}
    \amatr{H} = \left[ \frac{ \amatr{M}}{1+\alpha}-\gamma h \nabla_v\avect{f}^{l+1} -\beta h^2 \nabla_q\avect{f}^{l+1} +
    \beta h^2 \left[ (\amatr{M} \avect{a})_q + (\amatr{C}_q^T \avect{\lambda})_q \right] \right]
\end{equation}
Il termine \( \beta h^2 \left[ (\amatr{M} \avect{a})_q + (\amatr{C}_q^T \avect{\lambda})_q \right] \) può essere omesso per semplicità, rallentando tuttavia la convergenza dell'algoritmo di Newton.

### 2020-08-25 templates/Book_Ita/Contenuto/chapter_7.tex
Added BookIta template, Eccomas DRL paper, GVSETS AV paper, CoRL (under review)

\chapter{Implementazione dei Timestepper in Chrono}

% importante parlare di warm start (valori iniziali di Newton Raphson)

%  gli Rold sono la parte dei residui (sistema G) che non cambia, relativo cioè allo stato precedente. Per questo vengono caricati fuori dal ciclo, sarebbe inefficiente ogni volta fornire gli stessi valori, In Newmark non ho termini relativi a old nel sistema (ma davvero?) pertanto non c'è questa fase

%StateSolveCorrection serve per la soluzione di NR in schemi impliciti. è al termine di solveA. Di fatto risolve il sistema lineare dati residui e coefficienti

%StateSolveA trova l'accelerazione in schemi espliciti 
% da linea 60 a 75 calcolo numerico derivata SECONDA (in stato e tempo) di Q. 
% D2Q = [ Q(s+1) + Q(s-1) - 2Q(s) ] /Dt^2
%per calcolare i vari Q sposta lo stato di un delta e valuta il Q
% load constraint applicato + volta somma gli argomenti
% vedi sistema pag

\section{Risolvere il sistema lineare}
Che si tratti di metodi espliciti o impliciti (e si debba iterare con Newton-Raphson), è sempre necessario passare per la risoluzione di un sistema lineare.
Nel primo caso si risolverà, in un singolo passaggio, il sistema \ref{eq:dyn_sys}; nel secondo caso si dovrà risolvere ad ogni iterazione un sistema del tipo \ref{eq:eulforw_iterstep}.
Nel caso specifico la funzione \emph{StateSolveCorrection}, riportata in seguito, è ottimizzata per risolvere  un sistema come \ref{eq:eulforw_iterstep}, che effettivamente deve essere risolto più volte per ogni timestep (una volta per ogni iterazione di Newton-Raphson). Si noti che:
\begin{itemize}
    \item Alcuni termini nel vettore dei termini noti (residui) sono riferiti dall'istante temporale $l$, per cui non variano durante le iterazioni. Ha quindi senso fornirli una sola volta ad ogni timestep.
    \item La funzione invoca  il solutore di sistemi lineari scelto ed ottiene la soluzione. Esistono svariati algoritmi di soluzione (e ne sono implementati diversi in Chrono), tuttavia un'analisi più approfondita esula dagli scopi di questo testo.
    \item La medesima funzione può essere utilizzata per gli schemi espliciti che utilizzano il sistema \ref{eq:dyn_sys}.
\end{itemize}

Si riporta il codice (C++) della funzione \textit{StateSolveCorrection}. Le porzioni relative alla diagnostica sono state omesse.

\noindent\rule{\textwidth}{1pt}\newline
\begin{verbatim}
bool ChSystem::StateSolveCorrection(// result: computed Dv
                                    ChStateDelta& Dv,
                                    // result: computed lagrangian multipliers
                                    ChVectorDynamic<>& L,         
                                    // the R residual
                                    const ChVectorDynamic<>& R,
                                    // the Qc residual
                                    const ChVectorDynamic<>& Qc,  
                                    // the factor in c_a*M
                                    const double c_a,      
                                    // the factor in c_v*dF/dv
                                    const double c_v, 
                                    // the factor in c_x*dF/dv
                                    const double c_x,   
                                    // current state, x part
                                    const ChState& x,             
                                    // current state, v part
                                    const ChStateDelta& v,        
                                    // current time T
                                    const double T,               
                                    // scatters state to the system
                                    bool force_state_scatter,     
                                    // calls the solver's Setup() function
                                    bool force_setup              
                                    ) {
    CH_PROFILE( "StateSolveCorrection");

    if (force_state_scatter)
        StateScatter(x, v, T);

    // R and Qc vectors  --> solver sparse solver structures  
    //(also sets L and Dv to warmstart)
    IntToDescriptor(0, Dv, R, 0, L, Qc);

    // If the solver's Setup() must be called or if the solver's Solve() requires it,
    // fill the sparse system structures with information in G and Cq.
    if (force_setup || GetSolver()->SolveRequiresMatrix()) {
        timer_jacobian.start();

        // Cq  matrix
        ConstraintsLoadJacobians();

        // G matrix: M, K, R components
        if (c_a || c_v || c_x)
            KRMmatricesLoad(-c_x, -c_v, c_a);

        // For ChVariable objects without a ChKblock, just use the 'a' coefficient
        descriptor->SetMassFactor(c_a);

        timer_jacobian.stop();
    }

    // If indicated, first perform a solver setup.
    // Return 'false' if the setup phase fails.
    if (force_setup) {
        timer_setup.start();
        bool success = GetSolver()->Setup(*descriptor);
        timer_setup.stop();
        setupcount++;
        if (!success)
            return false;
    }

    // Solve the problem
    // The solution is scattered in the provided system descriptor
    timer_solver.start();
    GetSolver()->Solve(*descriptor);
    timer_solver.stop();

    // Dv and L vectors  <-- sparse solver structures
    IntFromDescriptor(0, Dv, 0, L);

    solvecount++;

    return true;
}

\end{verbatim}
\noindent\rule{\textwidth}{1pt} \newline

Argomenti della funzione:
\begin{itemize}
    \item In Dv ed L verranno salvati la variazione di velocità e di reazione vincolare calcolati
    \item In R e Qc vengono utilizzati per passare al Solver il vettore de i residui $\avect{G}$ in \ref{eq:eulforw_iterstep}. In R si trovano i termini di  \ref{eq:eulforw_Gq} e \ref{eq:eulforw_Gv} mentre Qc corrisponde a  \ref{eq:eulforw_Glam}.
    \item c\_a, c\_v e c\_x sono i coefficienti della matrice $\amatr{J}$ in \ref{eq:eulforw_iterstep}.
    \item Con x, v e T si passa al Solver lo stato (posizione e velocità) ed il tempo corrente.
    \item Infine 2 flag: la prima forza l'aggiornamento dello stato del sistema, mentre la seconda il setup del solver.
\end{itemize}

Si veda quindi in cosa consiste la funzione:
\begin{itemize}
    \item Se specificato, aggiorna lo stato
    \item Passa le reference residui e risultati al descrittore del solver 
    \item Passa al descrittore del solver la matrice dei coefficienti e dei vincoli
    \item esegue il setup del solver
    \item Il Solver risolve il sistema. Si noti che essendo il descrittore l'argomento della funzione Solve il Solver ha già sia i dati che le allocazioni di memoria dei risultati.
    \item dopo aver aggiornato il descrittore e aumentato il contatore, la funzione restituisce \textit{true}, indicando la riuscita dell'operazione.
\end{itemize}
L'algoritmo di soluzione dipende dalla tipologia di Solver. La funzione GetSolver trova il Solver (che viene specificato nello script della simulazione) e gli fornisce i dati necessari a calcolare la soluzione. Come già accennato una descrizione più estesa di ciò che avviene nella funzione Solve esulerebbe dagli scopi di questo testo.

Se invece si utilizza questa funzione per un sistema del tipo \ref{eq:dyn_sys} è sufficiente adottare semplici accorgimenti:
\begin{itemize}
    \item Dal momento che avrò come incognita $\delta\avect{v}/h$ invece di $\delta\avect{v}$ si dovrà semplicemente utilizzare il risultato coerentemente. Per il Solver, invece, non c'è alcuna differenza.
    \item Per caricare i termini noti del primo sistema inserisco F come un residuo con coefficiente 1, e questo è l'unico termine.
    \item Nella matrice dei coefficienti (argomenti si \textit{StateSolveCorrection)} il coefficiente di M sarà 1 (F=Ma), i restanti coefficienti saranno 0.
    \item Si  $Q_d$ verrà calcolato numericamente (in seguito si vedrà come)
\end{itemize}

% \section{Implementazione del metodo di Eulero esplicito}
% \subsection{Sistema lineare per metodi espliciti}
% Come accennato in precedenza per risolvere il sistema \ref{eq:dyn_sys} si utilizza sempre \emph{StateSolveCorrection}, e per utilizzarla ottenendo direttamente il sistema si introduce la funzione \emph{StateSolveA}, ossia una funzione che risolve la dinamica con l'accelerazione come incognita.

% \noindent\rule{\textwidth}{1pt} \newline
% \begin{verbatim}
%     bool ChIntegrableIIorder::StateSolveA(
%                                       // result: computed a for a=dv/dt
%                                       ChStateDelta& Dvdt,       
%                                       // result: computed lagrangian 
%                                       ChVectorDynamic<>& L,     
%                                       // current state, x
%                                       const ChState& x,         
%                                       // current state, v
%                                       const ChStateDelta& v,    
%                                       // current time T
%                                       const double T,           
%                                       // timestep (if needed)
%                                       const double dt,          
%                                       // scatters state to the system
%                                       bool force_state_scatter  
%                                       ) {
%     if (force_state_scatter)
%         StateScatter(x, v, T);

%     ChVectorDynamic<> R(GetNcoords_v());
%     ChVectorDynamic<> Qc(GetNconstr());
%     const double Delta = 1e-6;

%     LoadResidual_F(R, 1.0);

%     LoadConstraint_C(Qc, -2.0 / (Delta * Delta));

%     // numerical differentiation to get the Qc term in constraints
%     ChStateDelta dx(v);
%     dx *= Delta;
%     ChState xdx(x.GetRows(), this);

%     StateIncrement(xdx, x, dx);
%     StateScatter(xdx, v, T + Delta);
%     LoadConstraint_C(Qc, 1.0 / (Delta * Delta));

%     StateIncrement(xdx, x, -dx);
%     StateScatter(xdx, v, T - Delta);
%     LoadConstraint_C(Qc, 1.0 / (Delta * Delta));

%     StateScatter(x, v, T);  // back to original state

%     bool success = StateSolveCorrection(Dvdt, L, R, Qc, 
%                         1.0, 0, 0, x, v, T, false, true);

%     return success;
% }
% \end{verbatim}
% \noindent\rule{\textwidth}{1pt} \newline

% Argomenti della funzione:
% \begin{itemize}
%     \item In Dvtd ed L verranno salvati la variazione di velocità e di reazione vincolare calcolati.
%     A differenza del caso precedente specifichiamo che in Dvdt verrà calcolata un'accelerazione (più precisamente un impulso ${\Delta\avect{v}}/{\Delta t}$), non più una differenza di velocità.
%      \ref{eq:eulforw_iterstep}.
%     \item Con x, v e T si passa al Solver lo stato (posizione e velocità) ed il tempo corrente.
%     \item Infine 2 flag: la prima forza l'aggiornamento dello stato del sistema, mentre la seconda il setup del solver.
% \end{itemize}
% Rispetto alla funzione riportata nel paragrafo precedente, mancano:
% \begin{itemize}
%     \item R e Qc: infatti i termini noti sono sempre i vettori $\avect{F}$ e $\avect{Q_d}$
%     \item c\_a, c\_v e c\_x : poiché il coefficiente dell'accelerazione è sempre 1 ed i rimanenti sono sempre 0.
% \end{itemize}

% Ciò che viene svolto dalla funzione è piuttosto semplice da intuire se si tiene a mente quanto visto in precedenza. 
% \begin{itemize}
%     \item Viene creato un vettore per i termini noti R, il cui unico membro è la forza $\avect{F}$ con coefficiente 1 come accennato in precedenza.
%     \item Più complesso (e interessante) è il calcolo di $\avect{Q}_d$ dall'equazione \label{eq:Qd}
% \end{itemize}

% \subsection{Eulero esplicito}
% Viene riportata l'implementazione della funzione \emph{Advance} del metodo di Eulero esplicito. Essa è utilizzata per calcolare lo stato futuro dato lo stato attuale, ossia per far avanzare la simulazione di un timestep.
% \begin{verbatim}
%     void ChTimestepperEulerExpl::Advance(const double dt) {
%     // setup main vectors
%     GetIntegrable()->StateSetup(Y, dYdt);

%     // setup auxiliary vectors
%     L.Reset(this->GetIntegrable()->GetNconstr());

%     GetIntegrable()->StateGather(Y, T);  // state <- system

%     GetIntegrable()->StateSolve(dYdt, L, Y, T, dt, false);  // dY/dt = f(Y,T)

%     // Euler formula!
%     //   y_new= y + dy/dt * dt

%     Y = Y + dYdt * dt;  //  also: GetIntegrable().StateIncrement(y_new, y, Dy);

%     T += dt;

%     GetIntegrable()->StateScatter(Y, T);            // state -> system
%     GetIntegrable()->StateScatterDerivative(dYdt);  // -> system auxiliary data
%     GetIntegrable()->StateScatterReactions(L);      // -> system auxiliary data
% }
% \end{verbatim}
% In dYdt ed L vengono salvati i risultati di StateSolve che trova le accelerazioni del sistema e le reazioni vincolari. Lo stato viene aggiornato secondo la formula \ref{eq:Eul_Forw} e stati, accelerazioni e reazioni vengono propagati al sistema tramite le varie funzioni \emph{scatter}.
% \section{Implementazione dei metodi impliciti}
% Si riportano le funzioni di avanzamento di alcuni dei metodi di Eulero Implicito  e Newmark. Si riportano solo questi due esempi in quanto le implementazioni dei metodi sono in realtà simili tra loro.

\subsection{Eulero implicito}
La funzione \emph{Advance} della classe \emph{ChTimestepperEulerImplicit} riceve un solo argomento, cioè l'ampiezza del timestep \emph{dt} (chiamata $h$ nelle formule), in quanto può accedere a tutte le grandezze del sistema, che fa chiamando la funzione \emph{StateGather}, dopo aver resettato tutti i vettori che verranno utilizzati in seguito.
Si noti ora come come avviene l'inizializzazione dell'algoritmo di Newton fornendo un valore iniziale: la posizione viene inizializzata applicando uno step del metodo di Eulero esplicito mentre la velocità viene posta uguale a quella dell'istante di partenza. In generale per la continuità di entrambe il valore al timestep $l$ è un buon punto di partenza per il calcolo dei valori nell'istante successivo, data la continuità di posizione e velocità ed essendo $h$ piccolo. Lo stesso può essere detto per la posizione calcolata secondo il metodo di Eulero esplicito. Quest'ultimo non è consigliabile applicarlo come initial guess per la velocità in quanto questo  schema non trova direttamente le accelerazioni istantanee. Dopo aver azzerato i contatori del solutore si entra in un ciclo \emph{for} che: 
\begin{itemize}
    \item Propaga nel sistema l'ultimo stato calcolato (il warm start nel caso della prima iterazione).
    \item Resetta i vettori dei residui.
    \item Fornisce i valori dei residui aggiornati secondo la formula \ref{eq:eulfor_NRiter1} e riportata nel commento.
    \item Interrompe l'iterazione se entrambi i residui sono al di sotto della tolleranza prestabilita.
    \item Risolve il sistema lineare applicando i coefficienti della matrice in \ref{eq:eulfor_NRiter1}.
    \item Aggiorna stato, contatori e reazioni vincolari.
\end{itemize}
Al termine del ciclo viene aggiornata l'accelerazione, lo stato e le reazioni e vengono diffuse nel sistema.
\begin{verbatim}
    void ChTimestepperEulerImplicit::Advance(const double dt) {
    // downcast
    ChIntegrableIIorder* mintegrable = (ChIntegrableIIorder*)this->integrable;

    // setup main vectors
    mintegrable->StateSetup(X, V, A);

    // setup auxiliary vectors
    Dv.Reset(mintegrable->GetNcoords_v(), GetIntegrable());
    Dl.Reset(mintegrable->GetNconstr());
    Xnew.Reset(mintegrable->GetNcoords_x(), mintegrable);
    Vnew.Reset(mintegrable->GetNcoords_v(), mintegrable);
    R.Reset(mintegrable->GetNcoords_v());
    Qc.Reset(mintegrable->GetNconstr());
    L.Reset(mintegrable->GetNconstr());

    mintegrable->StateGather(X, V, T);  // state <- system

    // Extrapolate a prediction as warm start

    Xnew = X + V * dt;
    Vnew = V;  //+ A()*dt;

    // use Newton Raphson iteration to solve implicit Euler for v_new
    //
    // [ M-dt*dF/dv-dt^2*dF/dx   Cq'][ Dv   ] = [M*(v_old-v_new)+dt*f+dt*Cq'*l]
    // [ Cq                      0  ][-dt*Dl] = [-C/dt]

    numiters = 0;
    numsetups = 0;
    numsolves = 0;

    for (int i = 0; i < this->GetMaxiters(); ++i) {
        mintegrable->StateScatter(Xnew, Vnew, T + dt);  // state -> system
        R.Reset();
        Qc.Reset();
        mintegrable->LoadResidual_F(R, dt);
        mintegrable->LoadResidual_Mv(R, (V - Vnew), 1.0);
        mintegrable->LoadResidual_CqL(R, L, dt);
        mintegrable->LoadConstraint_C(Qc, 1.0 / dt, Qc_do_clamp, Qc_clamping);

        if ((R.NormInf() < abstolS) && (Qc.NormInf() < abstolL))
            break;

        mintegrable->StateSolveCorrection(
            Dv, Dl, R, Qc,
            1.0,                 
            -dt,                 
            -dt * dt,            
            Xnew, Vnew, T + dt,  
            false,               
            true                 
            );

        numiters++;
        numsetups++;
        numsolves++;
        //it is not -(1.0/dt) because StateSolveCorrection flips sign of Dl
        Dl *= (1.0 / dt);  
        L += Dl;

        Vnew += Dv;

        Xnew = X + Vnew * dt;
    }

    mintegrable->StateScatterAcceleration((Vnew - V) * (1 / dt));

    X = Xnew;
    V = Vnew;
    T += dt;

    mintegrable->StateScatter(X, V, T);      
    mintegrable->StateScatterReactions(L);   
}
\end{verbatim}

\subsection{Metodo di Newmark}
\paragraph{Setup dei parametri}
\begin{verbatim}
    void ChTimestepperNewmark::SetGammaBeta(double mgamma, double mbeta) {
    gamma = mgamma;
    if (gamma < 0.5)
        gamma = 0.5;
    if (gamma > 1)
        gamma = 1;
    beta = mbeta;
    if (beta < 0)
        beta = 0;
    if (beta > 1)
        beta = 1;
}
\end{verbatim}

\paragraph{Avanzamento}

\begin{verbatim}
// Performs a step of Newmark constrained implicit for II order DAE systems
void ChTimestepperNewmark::Advance(const double dt) {
    // downcast
    ChIntegrableIIorder* mintegrable = (ChIntegrableIIorder*)this->integrable;

    // setup main vectors
    mintegrable->StateSetup(X, V, A);

    // setup auxiliary vectors
    Da.Reset(mintegrable->GetNcoords_a(), GetIntegrable());
    Dl.Reset(mintegrable->GetNconstr());
    Xnew.Reset(mintegrable->GetNcoords_x(), mintegrable);
    Vnew.Reset(mintegrable->GetNcoords_v(), mintegrable);
    Anew.Reset(mintegrable->GetNcoords_a(), mintegrable);
    R.Reset(mintegrable->GetNcoords_v());
    Rold.Reset(mintegrable->GetNcoords_v());
    Qc.Reset(mintegrable->GetNconstr());
    L.Reset(mintegrable->GetNconstr());

    mintegrable->StateGather(X, V, T);  // state <- system
    mintegrable->StateGatherAcceleration(A);

    // extrapolate a prediction as a warm start

    Vnew = V;
    Xnew = X + Vnew * dt;

    // use Newton Raphson iteration to solve implicit Newmark for a_new

    //
    // [ M - dt*gamma*dF/dv - dt^2*beta*dF/dx    Cq' ] [ Da   ] = [ -M*(a_new) + f_new + Cq*l_new ]
    // [ Cq                                      0   ] [ Dl   ] = [ -1/(beta*dt^2)*C              ]

    numiters = 0;
    numsetups = 0;
    numsolves = 0;

    for (int i = 0; i < this->GetMaxiters(); ++i) {
        mintegrable->StateScatter(Xnew, Vnew, T + dt);  // state -> system

        R.Reset(mintegrable->GetNcoords_v());
        Qc.Reset(mintegrable->GetNconstr());
        mintegrable->LoadResidual_F(R, 1.0);                                                    //  f_new
        mintegrable->LoadResidual_CqL(R, L, 1.0);                                               //   Cq'*l_new
        mintegrable->LoadResidual_Mv(R, Anew, -1.0);                                            //  - M*a_new
        mintegrable->LoadConstraint_C(Qc, (1.0 / (beta * dt * dt)), Qc_do_clamp, Qc_clamping);  //  - 1/(beta*dt^2)*C

        if (verbose)
            GetLog() << " Newmark iteration=" << i << "  |R|=" << R.NormTwo() << "  |Qc|=" << Qc.NormTwo() << "\n";

        if ((R.NormInf() < abstolS) && (Qc.NormInf() < abstolL))
            break;

        mintegrable->StateSolveCorrection(
            Da, Dl, R, Qc,
            1.0,                 // factor for  M
            -dt * gamma,         // factor for  dF/dv
            -dt * dt * beta,     // factor for  dF/dx
            Xnew, Vnew, T + dt,  // not used here (scatter = false)
            false,               // do not StateScatter update to Xnew Vnew T+dt before computing correction
            true                 // force a call to the solver's Setup() function
            );

        numiters++;
        numsetups++;
        numsolves++;
        // Note it is not -= Dl because StateSolveCorrection flips sign of Dl
        L += Dl;  
        Anew += Da;

        Xnew = X + V * dt + A * (dt * dt * (0.5 - beta)) + Anew * (dt * dt * beta);

        Vnew = V + A * (dt * (1.0 - gamma)) + Anew * (dt * gamma);
    }

    X = Xnew;
    V = Vnew;
    A = Anew;
    T += dt;

    mintegrable->StateScatter(X, V, T);        // state -> system
    mintegrable->StateScatterAcceleration(A);  // -> system auxiliary data
    mintegrable->StateScatterReactions(L);     // -> system auxiliary data
}
\end{verbatim}

### 2020-08-25 templates/Book_Ita/Contenuto/chapter_8.tex
Added BookIta template, Eccomas DRL paper, GVSETS AV paper, CoRL (under review)

\chapter{Risultati e confronto}
In questo capitolo si confrontano i risultati di simulazioni eseguite con differenti schemi si risoluzione.
\section{Soluzione Analitica del Pendolo Semplice}
Al fine di stimare l'entità dell'errore dei metodi proposti è necessario confrontare i risultati ottenuti con una soluzione analitica. Un tipico problema di meccanica vincolata per il quale esiste una soluzione analitica è il pendolo semplice. Si prende in considerazione il caso generale, ossia \textbf{non} si assume $sin\theta \sim \theta$

Seguendo la soluzione proposta in \cite{Davis} per conoscere l'angolo $\theta$ e la sua derivata rispetto al tempo, la velocità angolare $\dot{\theta}$ si procede come segue:
\begin{figure}[h!]
\centering
\includegraphics[height=5cm]{Figure/pend.jpg}
\caption{Schema di un pendolo semplice}
 \label{fig:Pend}
\end{figure}
Si consideri innanzitutto il principio della conservazione dell'energia, che in questo caso consta di 2 termini, uno cinetico ed uno gravitazionale. Pertanto, dato un angolo di partenza $\theta_0$ con velocità iniziale nulla si, si può scrivere:
\begin{equation}
    mgl\left(1-cos\theta_0\right) = mgl\left(1-cos\theta \right)+\frac{1}{2}I\dot{\theta}^2
\end{equation}
Da cui:
\begin{equation}
    \dot{\theta} = \sqrt{\frac{2mgl}{I}} \sqrt{cos\theta - cos\theta_0}
\end{equation}
\begin{equation}
    dt =  \sqrt{\frac{I}{2mgl}} \frac{d\theta}{\sqrt{cos\theta - cos\theta_0}} 
\end{equation}
Per integrare il termine in $\theta$ si introduce la sostituzione:
\begin{equation} \label{eq:pend_sub}
    cos\theta = 1-2k^2sin^2\Phi \quad ; \qquad k = sin\frac{\theta_0}{2}
\end{equation}

Si osserva che che:
\begin{align} \nonumber
cos\theta - cos\theta_0 &=  1-2k^2sin^2\Phi - cos\theta_0  \\ \nonumber
&= 1-2k^2(1-cos^2\Phi) - cos\theta_0 \\ \nonumber
&= 1-2k^2+2k^2cos^2\Phi-cos\theta_0 \\ \nonumber
\left( k = sin\frac{\theta_0}{2}, cos\theta = 1-2sin^2\frac{\theta}{2}\right) \rightarrow &=cos\theta_0+ 2k^2cos^2\Phi - cos\theta_0 \\
cos\theta - cos\theta_0 &= 2k^2cos^2\Phi
\end{align}

\begin{align}
    sin\theta &= \sqrt{1-2cos^2\theta} \\ \nonumber
    &= \sqrt{1-(1-2k^2sin^2\Phi} \\ \nonumber
    &= \sqrt{1-(1+4k^4sin^4\Phi-4k^2sin^2\Phi)} \\ \nonumber
    &= \sqrt{4k^2sin^2\Phi - 4k^4sin^4\Phi} \\ \nonumber
    &= \sqrt{4k^2sin^2\Phi(1-k^2sin^2\Phi} \\
    sin\theta &= 2ksin\Phi\sqrt{(1-k^2sin^2\Phi)}  \
\end{align}

\begin{equation}
    sin\theta d\theta = 2k^2sin\Phi cos\Phi d\Phi
\end{equation}

Utilizzando queste relazioni partendo da \ref{eq:pend_sub} si può ottenere: 
\[\sqrt{\frac{I}{2mgl}} \frac{d\theta}{\sqrt{cos\theta - cos\theta_0}} = \sqrt{\frac{I}{2mgl}} 
\frac{sin\theta d\theta}{sin\theta\sqrt{cos\theta - cos\theta_0}} = \]
\[
= \sqrt{\frac{I}{2mgl}} \frac{4k^2sin\Phi cos\Phi d\Phi}{2ksin\Phi\sqrt{(1-k^2sin^2\Phi)} \sqrt{2k^2cos^2\Phi} } 
\]
Ed infine:
\begin{equation} \label{eq:pend_diff}
    dt = \sqrt{\frac{I}{mgl}} \frac{d\Phi}{\sqrt{1-k^2sin^\Phi}} 
\end{equation}
Integrando la \ref{eq:pend_diff} si ottiene un integrale ellittico incompleto del primo tipo:
\begin{equation} \label{eq:pend_int}
    t = \sqrt{\frac{I}{mgl}}  \int_0^\Phi \frac{d\Phi}{\sqrt{1-k^2sin^\Phi}}
\end{equation}
Ricordando che la funzione ellittica di Jacobi è definita come l'inverso dell'integrale ellittico del primo tipo, ossia dato:
\begin{equation}
    u =  \int_0^\Phi \frac{d\Phi}{\sqrt{1-k^2sin^\Phi}}
\end{equation}

\begin{equation} \label{eq:JacobiSN}
    sn(u|k) = sin(\Phi) \qquad cn(u|k) = cos(\Phi) \qquad dn(u|k) = \sqrt{1-k^2sin^\Phi}
\end{equation}

Per cui, utilizzando la prima relazione di \ref{eq:JacobiSN} \newline e ricordando che \( cos\theta = 1-2sin^2\left(\frac{\theta}{2}\right) \)
\begin{equation}
    sn\left( \sqrt{\frac{mgl}{I}t} \quad | \quad k \right) = sin(\Phi) = \frac{1}{k}sin\left(\frac{\theta}{2}\right)
\end{equation}
Infine si ottiene:
\begin{align}
    \theta &= 2 arcsin \left[ k \quad sn\left( \sqrt{\frac{mgl}{I}t} \quad | \quad k \right)   \right] 
    \\ \nonumber  k &= sin\frac{\theta_0}{2}
\end{align}

\subsection{Relative Percent Difference}
%Prossimità dello zero che dipende da sfiga 'punti vicini porta esempio trapz
Nella stima dell'accuratezza di un metodo si dovrà tenere conto di due fattori: innanzitutto ogni metodo è ovviamente esatto nelle condizioni iniziali, ad andrà via via allontanandosi dalla soluzione esatta. Per questa ragione si è considerata la media dell'errore su due oscillazioni complete per avere un dato più significativo dell'accuratezza del metodo.

Bisogna inoltre tener presene che entrambi i valori (analitico e numerico) possono annullarsi. Questo può portare a sovrastimare l'errore relativo, come appare evidente in figura \ref{fig:HHT_relerr}: la soluzione numerica è praticamente sempre sovrapposta a quella analitica, ma in prossimità di x=0 l'errore relativo diventa estremamente elevato. È quindi consigliabile non utilizzare l'errore relativo. Si potrebbe utilizzare la formula della differenza relativa che utilizza il valor medio per scalare il risultato.

\begin{equation}
    rd = \frac{(|v_{an}| - |v_{approx}|)}  { \frac{|v_{an}| + |v_{approx}|}{2}  }
\end{equation}

Dove $v_an$ e $v_approx$ sono rispettivamente il valore analitico e l'approssimazione numerica che si vuole confrontare.

Questo accorgimento limita il sovracitato problema come visibile in figura \ref{fig:HHT_relerr}, tuttavia quando uno dei due valori si annulla essa vale 2 \textbf{indipendentemente} dalla bontà dell'approssimazione. Per questa ragione si è scelto di non utilizzare nessuna scalatura dell'errore. Si utilizza semplicemente la media dei valori assoluti dell'errore \ref{eq:mean_err} per valutare la convergenza dell'algoritmo.
\begin{equation}
    \label{eq:mean_err}
    e_{mean} = \sum_{i=0}^{i=N} |v_{i, an} - v_{i, approx}|
\end{equation}

\begin{figure}[h!]
\centering
\includegraphics[height=7cm]{Figure/HHT_relerr.pdf}
\caption{Metodo HHT timestep = 5E-4}
 \label{fig:HHT_relerr}
\end{figure}

\section{Verifica Numerica della convergenza}
% andamento posizione all'infittire di h (poco), 
%andamento somma errori con h (importante perchè mostra convergenza
%parametri (se presenti)
In questa sezione vengono confrontati i risultati numerici al variare dell'ampiezza del timestep con la soluzione analitica. Si verifica che al ridursi del timestep l'errore medio \ref{eq:mean_err} tenda a zero.

Inoltre si confrontano le posizioni della punta del pendolo per visualizzare graficamente il risultato dell'integrazione numerica.
\subsection{Eulero Implicito}
Presenta il maggior smorzamento numerico, inoltre anche per timestep relativamente brevi si discosta (0.01s) si discosta visibilmente dalla soluzione esatta. 
Al tendere del timestep a $10^{-4}$, tuttavia, l'errore tende rapidamente a 0.

\begin{figure}[h!]
\centering
\includegraphics[height=7cm]{Figure/Convergenza/EulImp_err.pdf}
\caption{Convergenza del metodo di Eulero implicito}
 \label{fig:EulImp_err}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[height=7cm]{Figure/Convergenza/EulImp_pos.pdf}
\caption{Grafico posizione metodo di Eulero implicito}
 \label{fig:EulImp_pos}
\end{figure}
\FloatBarrier
\subsection{Metodo dei Trapezi}
Rispetto al metodo di Eulero implicito lo smorzamento numerico è minore e per timestep pari a 0.01s la soluzione è relativamente vicina a quella esatta.
\begin{figure}[h!]
\centering
\includegraphics[height=7cm]{Figure/Convergenza/Trapz_err.pdf}
\caption{Convergenza del metodo dei trapezi}
 \label{fig:Trapz_err}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[height=7cm]{Figure/Convergenza/Trapz_pos.pdf}
\caption{Grafico posizione del metodo dei trapezi}
 \label{fig:Trapz_pos}
\end{figure}
\FloatBarrier
\subsection{Metodo di Newmark}
Il metodo di Newmark presenta un'elevata instabilità per timestep elevati, al punto che per $ts > 0.05$ la soluzione diverge. Poiché l'errore medio di questi casi schiaccerebbe il grafico, si riportano i risultati per $ts < 0.05$.
Nonostante il problema elencato questo metodo presenta un'ottima approssimazione e come si nota dalla figura \ref{fig:Trapz_pos} già per $ts = 0.025$ l'approssimazione si sovrappone alla soluzione esatta.
\begin{figure}[h!]
\centering
\includegraphics[height=7cm]{Figure/Convergenza/Newm.pdf}
\caption{Convergenza del metodo di Newmark}
 \label{fig:Newm}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[height=7cm]{Figure/Convergenza/Newm_pos.pdf}
\caption{Grafico posizione del metodo di Newmark}
 \label{fig:Newm_pos}
\end{figure}
\FloatBarrier
\paragraph{Tuning dei Parametri}
Il metodo di Newmark presenta ordine 2 solo per $\gamma = 1/2$, per il quale inoltre lo smorzamento numerico è nullo. Per questa ragione ci si limita ad indagare gli effetti della variazione di $\beta$, tenendo $\gamma = 1/2$. Si ricorda che la condizione di assoluta stabilità è $\beta \geq \frac{(\gamma + \frac{1}{2})^2}{4}$, per cui per $\gamma = 1/2$ il metodo è assolutamente stabile per $\beta \geq 1/4$. Utilizzando la accelerazione lineare ($\beta = 1/6$) si esce dalla assoluta stabilità ed infatti la simulazione diverge. Variando $\beta$ nella regione di assoluta stabilità si ottiene invece che se esso è leggermente maggiore di quanto prescritto dalla condizione (0.27 invece di 0.25, come visibile in \ref{fig:NewmComp}) ne beneficia la stabilità per timestep più elevati senza inficiare la bontà dell'approssimazione per timestep ridotti.

\begin{figure}[h!]
\centering
\includegraphics[height=7cm]{Figure/Convergenza/NewmComp.pdf}
\caption{Convergenza del metodo HHT}
 \label{fig:NewmComp}
\end{figure}

\FloatBarrier
\subsection{Metodo HHT}

\begin{figure}[h!]
\centering
\includegraphics[height=7cm]{Figure/Convergenza/HHT_err.pdf}
\caption{Convergenza del metodo HHT}
 \label{fig:HHT_err}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[height=7cm]{Figure/Convergenza/HHT_pos.pdf}
\caption{Grafico posizione del metodo HHT}
 \label{fig:HHT_pos}
\end{figure}
\FloatBarrier

\paragraph{Tuning dei Parametri}
Come accennato in precedenza ponendo $\alpha=0$ ed ottenendo $\beta$ e $\gamma$ a partire da $\alpha$ come suggerito si ottiene il metodo di Newmark con accelerazione costante, e presenta gli stessi problemi di stabilità per timestep più lunghi \ref{fig:HHTComp}. Aumentando $\alpha$ cresce lo smorzamento numerico, ma al contempo il metodo diventa stabile. Si nota che (come anche noto dalla letteratura) il miglior valore di $\alpha$ è il più piccolo valore che permette la convergenza come visibile in \ref{fig:HHTComp2}.
\begin{figure}[h!]
\centering
\includegraphics[height=7cm]{Figure/Convergenza/HHTComp.pdf}
\caption{Errore del metodo HHT al variare di $\alpha$}
 \label{fig:HHTComp}
\end{figure}
\begin{figure}[h!]
\centering
\includegraphics[height=7cm]{Figure/Convergenza/HHTComp2.pdf}
\caption{Errore del metodo HHT al variare di $\alpha$}
 \label{fig:HHTComp2}
\end{figure}
\FloatBarrier
\section{Confronto fra i metodi}
I metodi ad hoc offrono performance migliori, osservando i grafici in figura \ref{fig:ErrComp} ed il particolare per timestep ridotti \ref{fig:ErrComp_zoom}.

Si nota che per timestep relativamente ampi (anche 0.1) il metodo HHT giunge a buone approssimazioni, ed in particolare per $ts \sim 0.01$, ovvero i tipici valori del timestep usati nella simulazioni multibody, l'approssimazione di HHT e Newmark è molto buona e ridotta rispetto ai metodi generici.
\begin{figure}[h!]
\centering
\includegraphics[height=7cm]{Figure/Convergenza/ErrComp.pdf}
\caption{Confronto Errore Medio Metodi Integrazione}
 \label{fig:ErrComp}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[height=7cm]{Figure/Convergenza/ErrComp_zoom.pdf}
\caption{Zoom Confronto Errore Medio Metodi Integrazione}
 \label{fig:ErrComp_zoom}
\end{figure}

### 2020-08-25 templates/Book_Ita/Contenuto/referenc_it.tex
Added BookIta template, Eccomas DRL paper, GVSETS AV paper, CoRL (under review)

%%%%%%%%%%%%%%%%%%%%%%%% referenc_it.tex %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Esempio di referenze
%
%
% Usare questo file come template per il vostro documento.
%
%%%%%%%%%%%%%%%%%%%%%%%% Springer-Verlag %%%%%%%%%%%%%%%%%%%%%%%%%%

%
% Utenti BibTeX: usare
% \bibliographystyle{}
% \bibliography{}
%
% Non-utenti BibTeX: usare
\begin{thebibliography}{[KLR73]}
%
% ed usare \bibitem per creare referenze.
%
% Usare la sintassi ed il markup seguenti per le vostre referenze.
%
% Monografie
\bibitem[KLR73]{monograph} Kagan, A.M., Linnik, Y.V., Rao, C.R.:
Characterization Problems in Mathematical Statistics. Wiley, New York (1973)

% Contributed Works
\bibitem[Mey89]{contribution} Meyer, P.A.: A short presentation of
stochastic calculus. In: Emery, M. (ed) Stochastic Calculus in
Manifolds. Springer, Berlin Heidelberg New York (1989)

% Journal
\bibitem[MR97]{journal} Miller, B.M., Runggaldier, W.J.: Kalman
filtering for linear systems with coefficients driven by a hidden Markov
jump process. Syst. Control Lett., \textbf{31}, 93--102 (1997)

% Tesi
\bibitem[Ros77]{thesis} Ross, D.W.: Lysosomes and storage diseases. MA
Thesis, Columbia University, New York (1977)

\end{thebibliography}

### 2020-09-17 2021/NODYCON_nonsmooth_ADMM/author.tex
Stub of paper for ADMM

%%%%%%%%%%%%%%%% based on Springer template for conference proceedings %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass{svproc}

% to typeset URLs, URIs, and DOIs
\usepackage{url}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{csquotes}

\usepackage{hyperref}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{cleveref}
\usepackage{bm}
\usepackage{mathtools}
\usepackage{siunitx}
\usepackage{subfig}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{tabularx}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{array}
\usepackage{booktabs}
\usepackage{multicol}
\usepackage{xcolor}
\usepackage[title]{appendix}
\usepackage{empheq}

\def\UrlFont{\rmfamily}

\newcommand{\vect}[1]{\bm{#1}}
\newcommand{\mat}[1]{#1}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%5

\begin{document}
\mainmatter              % start of a contribution
%
\title{Solving Non-smooth Dynamic Problems using \\ the 
Alternating Direction Method of Multipliers}
%
\titlerunning{Non-smooth dynamic problem}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Alessandro Tasora \and Dario Mangoni \and Simone Benatti}
%
\authorrunning{Alessandro Tasora et al.} % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
\tocauthor{Alessandro Tasora, Dario Mangoni, and Simone Benatti}
%
\institute{University of Parma, Department of Engineering and Architecture \\ V.le delle Scienze 181/A, 40121 ITALY\\
\email{alessandro.tasora@unipr.it},\\ WWW home page:
\texttt{http://digitaldynamicslab.unipr.it}
}

\maketitle              % typeset the title of the contribution

\begin{abstract}
We propose to use the Alternating Direction Method of Multipliers (ADMM) for solving the variational inequality that arises in non-smooth contact problems in multibody simulations. The ADMM method is based on simple computational primitives and offers good convergence properties especially for scenarios where loose tolerance on the precision can be accepted, thus making it attractive also for real-time applications. The method scales well with the problem size and require a small number of tuning parameters. Differently from other methods known in literature, such as fixed point iterations, it can easily handle problems that mix finite elements and rigid bodies, and it is only marginally affected by odd mass ratios or ill-posed problems. Another attractive feature is that it can be easily warm-started.
\keywords{ADMM, non-smooth dynamics, contact}
\end{abstract}
%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}

Following the seminal works by Moreau 
\cite{mor88,Jean1992} 
most formulations for non-smooth dynamical problems are based on Measure Differential Inclusions: just like differential inclusions they allow set-valued force laws (such as the Coulomb-Amontons dry friction model) but also generalize to the case where velocity is assumed to be a possibly discontinuous function of bounded variation in order to allow impulsive events.

These problems can be solved by means of special time stepping methods that offer superior robustness and stability at the cost of solving a complementarity problem, or more in general a Variational Inequality (VI), per each time step
 \cite{acary2008numerical}. 
In this context, unknowns to be solved are velocity measures and reaction impulses at contact points and at joints: in cases of many parts with lot of frictional contacts, the large dimension of the VI could lead to a bottleneck in the time stepping process. This stimulated lot of research on efficient numerical methods in the last three decades.

One of the former approaches, presented in 
\cite{StTr95}  % LCP solved via Lemke
, was based on solving a Linear Complementarity Problem (LCP) per each time step. LCPs are sub cases of VIs, for whom a direct method (Lemke algorithm) does exist. However, direct LCP solvers offer exact solution at the cost of very expensive pivoting sub iterations that scale badly with increasing number of unknowns, and for these reason they are not much used nowadays.

On the other hand, approximate but efficient methods based on fixed-point iterations became popular in the area of real-time simulators and robotics \cite{Bender2014}.
In most cases they are based on stationary methods like Gauss-Seidell or Jacobi iterations interleaved with projections on friction cones. They can be used to solve the LCP or the Cone Complementarity Problem (CCP), another special case of the VI in which the multibody contact problem can be formulated.
Attempts were made in order to parallelize them and to increase their convergence, such as over-relaxation, Krasnoselskii-Mann smoothing and warm starting, 
\cite{massSplittingRichard2012,TasoraAnitescuCMAME10} % P-GS iteration
but in general they perform poorly when there are odd mass ratios, and when articulated mechanisms such as robots are added to the scenarios - in those cases convergence often stalls, and if the iteration is prematurely truncated, reaction impulses are badly estimated and mechanisms fall apart or bend, and objects might interpenetrate.

Another option is to solve the CCP as an optimization problem, using first-order optimization methods such as 
the Nesterov accelerated projected gradient descend  
\cite{hammadTOG2015}
or the Barzilai-Borwein spectral projected gradient
\cite{heynIJNME2013}. These methods are based on a simple projection operator,  matrix-by-vector multiplications and inner products. Their convergence is better than for fixed points iterations, however they share with them the following issue: that they often require the building of a Delassus operator, a matrix that is easy to compute in efficient (factored and sparse) format if diagonal masses are used because these can be easily inverted, but hard to handle if finite elements are added to the problem, because they introduce stiffness and damping matrices whose inverse would be needed as well.

A further class of solvers is represented by non-smooth newton methods, such as the one presented in 
\cite{Macklin2019} % Non-smooth Newton Method
, that assume a generic Nonlinear Complemetarity Problem (NCP), again a sub case of a VI, and represent the NCP complementarity constraints using non-smooth functions like the Fisher-Burmeister function. A generalized non-smooth Newton method can be used to find the zero of the functions, at the cost of solving a linear system per each iteration.

Under mild assumptions, the VI can be stated as a convex CCP, that can be also cast as an optimization problem, namely a Quadratic Program (QP) with convex constraints \cite{anitescuTasora2008}. For such a class of optimization problems, one can use Interior Point Methods (IPMs), a class of solvers that share some similarities with the above mentioned non-smooth Newton methods - for example the solution of a saddle-point linear system is required at each iteration. IPMs offer the best theoretical convergence. However, their implementation is quite intricate, and despite the encouraging theoretical properties, in practice they do not scale well for large problems. Also, there are no reliable ways to warm-start them
\cite{Mangoni2018}
.

Recently there has been a revival of Alternating Direction Method of Multipliers (ADMM) and similar operator-splitting methods for solving constrained optimization problems
\cite{Goldstein2014}. %  [2014 Alternating Direction Optimization Methods]. 
Even if the theoretical convergence of such methods is worse than the one of IPMs, recent developments showed that in practical scenarios they offer superior speed, scalability and robustness 
\cite{Cannon2019}. % [COSMO  conic operator splitting method for large convex problems].
An attractive property of ADMM is that it requires the solution of a linear system per each iteration, but unlike IPMs, such linear system most often remains unchanged during the iterations, so a factorization can be reused multiple times with a great benefit in terms of speed.

In 
\cite{Stellato2020} % [OSQP: an operator splitting solver for quadratic programs].
a specialized ADMM method for quadratic problems with conic constraints has been presented: on the top of this method we develop a solver that can exploit the nature of non-smooth dynamical problems. In our embodiment, unknowns are primal variables (velocity measures) and dual variables (impulses in contact points and joints), cone constraints stem from the Coulomb friction law, and the system matrix includes terms from the mass matrices and from the tangent stiffness/damping matrices (hence it is sparse, but not necessarily diagonal).

This paper will present a model for the non-smooth dynamics, it will show how to cast it as a QP with convex conic constraint, it will present the ADMM method to solve it, it will discuss practical implementation details and then it will show some benchmarks.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%5

\section{The non-smooth multibody model}

From \cite{TasoraAnitescuCMAME10} one can see that the non-smooth DVI after regularization and convexification turns into this: at each time step one has to solve the following (mixed) Cone Complementarity Problem:

\begin{subequations}
	\begin{empheq}[left=\empheqlbrace]{align}
    M \vect{v} - \vect{k} - D_\epsilon \bar{\vect{\gamma}}_\epsilon &= 0 \\
    D_{\epsilon}^T \vect{v}  + \vect{b}_\epsilon &= \vect{u}_\epsilon\\
    -\Upsilon^{\circ} \ni \vect{u}_\epsilon  \quad \bot &\quad  \bar{\vect{\gamma}}_\epsilon \in \Upsilon
	\end{empheq}
	\label{eq:ChronoMCCP}
\end{subequations}

where 
\begin{itemize}
    \item $M$ is a positive definite, block diagonal matrix containing masses and inertia tensors of the bodies
    \item $D_\epsilon$ is a sparse matrix, the transpose jacobian of all constraints
    \item $\vect{k}$ is a vector containing terms proportional to applied forces, among other things 
    \item $\vect{b}$ is a vector containing constraint stabilization terms, among other things
    \item $\Upsilon$ is the Cartesian product of all cones of admissible constraint forces, $\Upsilon = \bigtimes_i \Upsilon_i$
    \item if a frictional contact is added, $\Upsilon_i \subset \mathbb{R}^3$ is a second order Lorentz cone with aperture proportional to friction coefficient
    \item if a bilateral constraint is added, $\Upsilon_i = \mathbb{R}$ and $\Upsilon^\circ_i = \{0\}$
    \item if a unilateral constraint is added, $\Upsilon_i = \mathbb{R}^+$ and $\Upsilon^\circ_i = \mathbb{R}^-$
    \item $\Upsilon^{\circ}$ is the polar cone, opposite of the dual cone, i.e. $\Upsilon^{*} = -\Upsilon^\circ$.
\end{itemize}

For more details on the model we refer to 
\cite{negrutSerbanTasoraJCND2017}. % Posing Multibody Dynamics with Friction and Contact as a Differential Complementarity Problem

...

[TO REWRITE THE FOLLOWING]

By introducing the Schur complement 
\[
N=D_{\epsilon}^T M^{-1} D_{\epsilon}
\]
and the vector
\[
\vect{r}_\epsilon = D_{\epsilon}^T M^{-1} \vect{k} + \vect{b}_\epsilon
\]
such that
\[
\vect{u}_\epsilon = N \bar{\vect{\gamma}}_\epsilon + \vect{r}_\epsilon \, , 
\]
one can also write the problem as the following CCP:
\begin{align}
    -\Upsilon^{\circ} \ni  N \bar{\vect{\gamma}}_\epsilon + \vect{r}_\epsilon 
    \quad \bot \quad  
    \bar{\vect{\gamma}}_\epsilon \in \Upsilon
	\label{eq:ChronoCCP}
\end{align}

This CCP corresponds, again, to a first order optimality condition of a convex quadratic program, but this time of the type
\begin{subequations}
	\begin{empheq}[box=\fbox]{align}
	\text{min} \quad & \frac{1}{2} \bar{\vect{\gamma}}_\epsilon^T N \bar{\vect{\gamma}}_\epsilon + \vect{r}^T_\epsilon \bar{\vect{\gamma}}_\epsilon \\
	\text{s.t.} \quad & \bar{\vect{\gamma}}_\epsilon \in \Upsilon
	\end{empheq}
	\label{eq:ChronoCCP_min}
\end{subequations}
%
and in fact, the CCP \eqref{eq:ChronoCCP} can be written in the more conventional language of the KKT optimality conditions introducing $\vect{y}$ multipliers, for $\vect{y}=-\vect{u}_\epsilon$:
\begin{subequations}
	\begin{align}
    N \bar{\vect{\gamma}}_\epsilon + \vect{r}_\epsilon + I \vect{y} &= 0 \\
    \bar{\vect{\gamma}}_\epsilon &= \vect{z} \\
    \Upsilon \ni \vect{z}  \quad \bot &\quad \vect{y} \in \Upsilon^\circ % \mathcal{N}(\Upsilon)
	\end{align}
	\label{eq:ChronoCCP_kkt}
\end{subequations}

Also in this case, efficient methods like ADMM can be used to solve the convex program \eqref{eq:ChronoCCP_min}. Differently from \eqref{eq:ChronoMCCP_min} in this case we would iterate to solve for unknown $\bar{\vect{\gamma}}_\epsilon$, and then compute $\vect{v} = M^{-1}( \vect{k} + \bar{D}_\epsilon \bar{\vect{\gamma}}_\epsilon)$ with a quick postprocessing step. We call this approach \textbf{IDEA "A"}.

In this approach, the ADMM method requires a fast way to perform a projection 
$\Pi_{\Upsilon}$ onto the friction cones $\Upsilon$.

\subsection{Generalization: multi-flexible problems and non-conic force functions}

In order to generalize the problem \eqref{eq:ChronoMCCP} we relax the assumption of $M$ being diagonal, and instead of $M$ we introduce a symmetric positive definite matrix $H$, not necessarily sparse, that contains linear combinations of stiffness and damping matrices, added to the mass matrix. Stiffness and damping matrices are needed when dealing with implicit integration of stiff forces as in FEA and compliant parts, in general they are block-sparse, anyway less sparse than the original $M$ matrix, thus rendering the computation of $H^{-1}$ much less friendly than computing $M^{-1}$.

As a second generalization, we assume that $\bar{\vect{\gamma}}_\epsilon$ is not constrained in a friction cone $\Upsilon$, but rather in a generic convex set $\mathcal{C}$. That is, we assume $\bar{\vect{\gamma}}_\epsilon \in \mathcal{C}$ and at the same time we assume an associated \footnote{Not all dissipative constitutive models are associated, for example in some computational plasticity models the plastic flow might deviate respect to the normal to the yeld surface, and the Coulmb contact model itself is not associated -in fact the contact velocity $\vect{u}$ can range into a wider cone than the polar of the friction cone, but our DVI model compensates fur such effect. We will assume associated models heretofore, and we do not deal with non-associated models to avoid additional complexity for the moment.} flow $\vect{u}_\epsilon \in - \mathcal{N}_{C}(\bar{\vect{\gamma}}_\epsilon)$.

Using generic $\mathcal{C}$ allows, for an example, a friction cone capped with some upper limit (ex to simulate plastic contacts that cannot withstand excessive pressure) or a friction cone shifted a bit downward (ex to simulate cohesion in contacts) etc.

Under this more general setting, the time stepping of the multibody model becomes:

\begin{subequations}
	\begin{empheq}[left=\empheqlbrace]{align}
    H \vect{v} - \vect{k} - {D}_\epsilon \bar{\vect{\gamma}}_\epsilon &= 0 \label{eq:ChronoVI_primal}\\
    {D}_{\epsilon}^T \vect{v}  + \vect{b}_\epsilon &= \vect{u}_\epsilon\\
    \bar{\vect{\gamma}}_\epsilon \in \mathcal{C} : \quad  \vect{u}_\epsilon &\in - \mathcal{N}_{C}(\bar{\vect{\gamma}}_\epsilon)  
	\end{empheq}
	\label{eq:ChronoVI}
\end{subequations}

This is a Variational Inequality (VI), and it contains the CCP of \eqref{eq:ChronoCCP_kkt} as a special case. In fact one can introduce a Schur matrix 
\[
N=D_{\epsilon}^T H^{-1} D_{\epsilon}
\] similarly to what presented in  IDEA "A", and rewrite it as:
\begin{subequations}
	\begin{align}
    N \bar{\vect{\gamma}}_\epsilon + \vect{r}_\epsilon - \vect{u}_\epsilon &= 0 \\
    \bar{\vect{\gamma}}_\epsilon \in \mathcal{C} : \quad  -\vect{u}_\epsilon &\in \mathcal{N}_{C}(\bar{\vect{\gamma}}_\epsilon) 
	\end{align}
	\label{eq:ChronoVI_schur}
\end{subequations}
Also, using the same symbols introduced in OSQP, with $\vect{y}=-\vect{u}_\epsilon$, one gets
\begin{subequations}
	\begin{align}
    N \bar{\vect{\gamma}}_\epsilon + \vect{r}_\epsilon + I \vect{y} &= 0 \\
    \bar{\vect{\gamma}}_\epsilon &= \vect{z} \\
    \vect{z} \in \mathcal{C} : \quad  \vect{y} &\in \mathcal{N}_{C}(\bar{\vect{\gamma}}_\epsilon) 
	\end{align}
	\label{eq:ChronoVI_schur_kkt}
\end{subequations}
where one can easily see that this is a VI$(\vect{y}(\bar{\vect{\gamma}}_\epsilon), \mathcal{C})$ and that this is also the first order optimality condition of a convex quadratic program:
\begin{subequations}
	\begin{empheq}[box=\fbox]{align}
	\text{min} \quad & \frac{1}{2} \bar{\vect{\gamma}}_\epsilon^T N \bar{\vect{\gamma}}_\epsilon + \vect{r}^T_\epsilon \bar{\vect{\gamma}}_\epsilon \\
	\text{s.t.} \quad & \bar{\vect{\gamma}}_\epsilon \in \mathcal{C}
	\end{empheq}
	\label{eq:ChronoVI2_min}
\end{subequations}

Again, it is possible to apply ADMM algorithms to solve the problem.

\section{The ADMM solver}

...

[TO REWRITE THE FOLLOWING]

The ADMM starts by stating a penalized objective function, that includes all the constraints by means of indicator functions:
\begin{subequations}
	\begin{align}
	\text{min} \quad & \frac{1}{2} \vect{v}^T M \vect{v} - \vect{k}^T\vect{v} + \mathcal{I}_{\bar{D}_{\epsilon}^T \vect{v} + \vect{b}_\epsilon  = \vect{u}_\epsilon} (\tilde{\vect{v}}, \tilde{\vect{u}}_\epsilon) +
	\mathcal{I}_{\mathcal{K}}(\vect{u}_\epsilon)\\
	\text{s.t.} \quad & (\tilde{\vect{v}}, \tilde{\vect{u}}_\epsilon) = (\vect{v}, \vect{u}_\epsilon)
	\end{align}
\end{subequations}
where $\mathcal{I}_{\bar{D}_{\epsilon}^T \vect{v} + \vect{b}_\epsilon  = \vect{u}_\epsilon}$ and $\mathcal{I}_{\mathcal{K}}$ are infinite if the constraint is violated and zero if not. This introduces a really strong barrier, yet non-smooth.

The associated augmented lagragian is then used to build a equality constrained QP problem that is in the form:
\begin{subequations}
	\begin{align}
	\text{min} \quad & \frac{1}{2} \tilde{\vect{v}}^T M \tilde{\vect{v}} - \vect{k}^T\tilde{\vect{v}} + \frac{\sigma}{2}\norm{\tilde{\vect{v}}-\vect{v}^k}^2 + \frac{\rho}{2}\norm{\tilde{\vect{u}_\epsilon}-\vect{u}_\epsilon^k + \frac{\bar{\vect{\gamma}}_\epsilon}{\rho}}^2\\
	\text{s.t.} \quad & \bar{D}_{\epsilon}^T \tilde{\vect{v}} + \vect{b}_\epsilon  = \vect{u}_\epsilon
	\end{align}
\end{subequations}
whose KKT conditions can be found as:
\begin{equation}
    \begin{bmatrix}
        M + \sigma I_n & \bar{D}_{\epsilon} \\
        \bar{D}_{\epsilon}^T & -\rho^{-1} I_m
    \end{bmatrix}
    \begin{bmatrix}
        \tilde{\vect{v}}^{k+1} \\
        \vect{\nu}^{k+1}
    \end{bmatrix}
    =
    \begin{bmatrix}
        \vect{k} + \sigma \vect{v}^k \\
        \vect{b}_\epsilon - \vect{u}_\epsilon^k + \rho^{-1}\bar{\vect{\gamma}}_\epsilon^k
    \end{bmatrix}
\end{equation}

then we can update variables as
\begin{subequations}
	\begin{align}
	\tilde{\vect{u}}_\epsilon^{k+1} &= \vect{u}_\epsilon^k - \rho^{-1}(\vect{\nu}^{k+1} + \bar{\vect{\gamma}}_\epsilon^k)\\
	\vect{v}^{k+1} &= \alpha \tilde{\vect{v}}^{k+1} + (1-\alpha) \vect{v}^k\\
	\vect{u}_\epsilon^{k+1} &= \Pi_\mathcal{K}( \alpha \tilde{\vect{u}_\epsilon}^{k+1} + (1-\alpha) \vect{u}_\epsilon^k + \rho^{-1} \bar{\vect{\gamma}}_\epsilon^k ) \\
	\bar{\vect{\gamma}}_\epsilon^{k+1} &= \bar{\vect{\gamma}}_\epsilon^k + \rho (  \alpha \tilde{\vect{u}_\epsilon}^{k+1} + (1-\alpha) \vect{u}_\epsilon^k - \vect{u}_\epsilon^{k+1} )
	\end{align}
\end{subequations}

...

\section{Results}

...

\begin{figure}[h!]
\centering
\includegraphics[width=0.50\textwidth]{t8_convergence.pdf}
\includegraphics[width=0.35\textwidth, trim=0cm -3cm 0 3cm]{t8_snapshot.png}
\caption{Convergence of the ADMMM method within one time step of the wrecking ball benchmark (600 bricks in four walls): residual in frictional constraint   violation compared to fixed point Jacobi iterations and to first-order SPG methods.}
\end{figure}

\section{Conclusion}

Our ADMM method requires few computational primitives: basically a projection of dual variables on conic sets, a backward solve of a linear system, and a forward solve. The latter is a computational bottleneck, but it can be performed only once per run, as the matrix does not change often during the iterations. 
A good estimation of the ADMM step size proved to be fundamental in achieving good convergence: using some heuristics we obtained an efficient auto-tuning algorithm. 
We noted that ADMM can be successfully applied to problems that exhibit temporal coherence because, unlike IP methods, it supports warm-starting.

\bibliographystyle{spmpsci}
\bibliography{../../bibliography/refsMBS,../../bibliography/refsOPT}

\end{document}

### 2020-09-18 2021/NODYCON_nonsmooth_ADMM/author.tex
Added text

@@
 \cite{mor88,Jean1992} 
-most formulations for non-smooth dynamical problems are based on Measure Differential Inclusions: just like differential inclusions they allow set-valued force laws (such as the Coulomb-Amontons dry friction model) but also generalize to the case where velocity is assumed to be a possibly discontinuous function of bounded variation in order to allow impulsive events.
+most formulations for non-smooth dynamical problems are based on Measure Differential Inclusions (MDI): just like differential inclusions they allow set-valued force laws (such as the Coulomb-Amontons dry friction model) but also generalize to the case where velocity is assumed to be a possibly discontinuous function of bounded variation in order to allow impulsive events.
 
@@
 
-From \cite{TasoraAnitescuCMAME10} one can see that the non-smooth DVI after regularization and convexification turns into this: at each time step one has to solve the following (mixed) Cone Complementarity Problem:
+Different time stepping schemes have been proposed for MDIs, here we refer to the one discussed \cite{TasoraAnitescuCMAME10} without lack of generality.
+After regularization and convexification and discretization, the MDIs leads to a major numerical bottleneck to be solved at each time step: a (mixed) CCP with unknowns $\vect{v}$ and $\vect{\gamma}_\epsilon$:
 
@@
 	\begin{empheq}[left=\empheqlbrace]{align}
-    M \vect{v} - \vect{k} - D_\epsilon \bar{\vect{\gamma}}_\epsilon &= 0 \\
+    M \vect{v} - \vect{k} - D_\epsilon \vect{\gamma}_\epsilon &= 0 \\
     D_{\epsilon}^T \vect{v}  + \vect{b}_\epsilon &= \vect{u}_\epsilon\\
-    -\Upsilon^{\circ} \ni \vect{u}_\epsilon  \quad \bot &\quad  \bar{\vect{\gamma}}_\epsilon \in \Upsilon
+    -\Upsilon^{\circ} \ni \vect{u}_\epsilon  \quad \bot &\quad  \vect{\gamma}_\epsilon \in \Upsilon
 	\end{empheq}
@@
 \begin{itemize}
+		\item unknown $\vect{v}$ is the speed at the end of the time step, 
+		\item unknown $\vect{\gamma}_\epsilon$ is the reaction in contacts and bilateral joints, a vector-signed measure with Lebesgue decomposition in atomic parts (impacts) and continuous parts (continuous reactions),
     \item $M$ is a positive definite, block diagonal matrix containing masses and inertia tensors of the bodies
@@
 
-...
-
-[TO REWRITE THE FOLLOWING]
-
-By introducing the Schur complement 
+Introducing the Schur complement 
 \[
@@
 \[
-\vect{u}_\epsilon = N \bar{\vect{\gamma}}_\epsilon + \vect{r}_\epsilon \, , 
+\vect{u}_\epsilon = N \vect{\gamma}_\epsilon + \vect{r}_\epsilon \, , 
 \]
@@
 \begin{align}
-    -\Upsilon^{\circ} \ni  N \bar{\vect{\gamma}}_\epsilon + \vect{r}_\epsilon 
+    -\Upsilon^{\circ} \ni  N \vect{\gamma}_\epsilon + \vect{r}_\epsilon 
     \quad \bot \quad  
-    \bar{\vect{\gamma}}_\epsilon \in \Upsilon
+    \vect{\gamma}_\epsilon \in \Upsilon
 	\label{eq:ChronoCCP}
@@
 
-This CCP corresponds, again, to a first order optimality condition of a convex quadratic program, but this time of the type
+This CCP corresponds exactly to a first order optimality condition of a convex quadratic program:
 \begin{subequations}
 	\begin{empheq}[box=\fbox]{align}
-	\text{min} \quad & \frac{1}{2} \bar{\vect{\gamma}}_\epsilon^T N \bar{\vect{\gamma}}_\epsilon + \vect{r}^T_\epsilon \bar{\vect{\gamma}}_\epsilon \\
-	\text{s.t.} \quad & \bar{\vect{\gamma}}_\epsilon \in \Upsilon
+	\text{min} \quad & \frac{1}{2} \vect{\gamma}_\epsilon^T N \vect{\gamma}_\epsilon + \vect{r}^T_\epsilon \vect{\gamma}_\epsilon \\
+	\text{s.t.} \quad & \vect{\gamma}_\epsilon \in \Upsilon
 	\end{empheq}
@@
 %
-and in fact, the CCP \eqref{eq:ChronoCCP} can be written in the more conventional language of the KKT optimality conditions introducing $\vect{y}$ multipliers, for $\vect{y}=-\vect{u}_\epsilon$:
+and in fact, the CCP \eqref{eq:ChronoCCP} can be written in the more conventional language of the KKT optimality conditions on \textit{dual variables} $\vect{y}$ multipliers, for $\vect{y}=-\vect{u}_\epsilon$ and \textit{primal variables} $\vect{\gamma}_\epsilon$:
 \begin{subequations}
 	\begin{align}
-    N \bar{\vect{\gamma}}_\epsilon + \vect{r}_\epsilon + I \vect{y} &= 0 \\
-    \bar{\vect{\gamma}}_\epsilon &= \vect{z} \\
-    \Upsilon \ni \vect{z}  \quad \bot &\quad \vect{y} \in \Upsilon^\circ % \mathcal{N}(\Upsilon)
+    N \vect{\gamma}_\epsilon + \vect{r}_\epsilon + I \vect{y} &= \vect{0} \\
+    \vect{\gamma}_\epsilon &= \vect{z} \\
+    \Upsilon \ni \vect{z}  \quad \bot &\quad \vect{y} \in \Upsilon^\circ  
 	\end{align}
@@
 \end{subequations}
+After the convex program \eqref{eq:ChronoCCP_min} is solved by ADMM, one can compute $\vect{v} = M^{-1}( \vect{k} + \bar{D}_\epsilon \vect{\gamma}_\epsilon)$ with a quick postprocessing step.
 
-Also in this case, efficient methods like ADMM can be used to solve the convex program \eqref{eq:ChronoCCP_min}.
-Differently from \eqref{eq:ChronoMCCP_min} in this case we would iterate to solve for unknown $\bar{\vect{\gamma}}_\epsilon$, and then compute $\vect{v} = M^{-1}( \vect{k} + \bar{D}_\epsilon \bar{\vect{\gamma}}_\epsilon)$ with a quick postprocessing step.
-We call this approach \textbf{IDEA "A"}.
-
-In this approach, the ADMM method requires a fast way to perform a projection 
-$\Pi_{\Upsilon}$ onto the friction cones $\Upsilon$.
-
-\subsection{Generalization: multi-flexible problems and non-conic force functions}
-
-In order to generalize the problem \eqref{eq:ChronoMCCP} we relax the assumption of $M$ being diagonal, and instead of $M$ we introduce a symmetric positive definite matrix $H$, not necessarily sparse, that contains linear combinations of stiffness and damping matrices, added to the mass matrix.
-Stiffness and damping matrices are needed when dealing with implicit integration of stiff forces as in FEA and compliant parts, in general they are block-sparse, anyway less sparse than the original $M$ matrix, thus rendering the computation of $H^{-1}$ much less friendly than computing $M^{-1}$.
-
-As a second generalization, we assume that $\bar{\vect{\gamma}}_\epsilon$ is not constrained in a friction cone $\Upsilon$, but rather in a generic convex set $\mathcal{C}$.
-That is, we assume $\bar{\vect{\gamma}}_\epsilon \in \mathcal{C}$ and at the same time we assume an associated \footnote{Not all dissipative constitutive models are associated, for example in some computational plasticity models the plastic flow might deviate respect to the normal to the yeld surface, and the Coulmb contact model itself is not associated -in fact the contact velocity $\vect{u}$ can range into a wider cone than the polar of the friction cone, but our DVI model compensates fur such effect.
-We will assume associated models heretofore, and we do not deal with non-associated models to avoid additional complexity for the moment.} flow $\vect{u}_\epsilon \in - \mathcal{N}_{C}(\bar{\vect{\gamma}}_\epsilon)$.
-
-Using generic $\mathcal{C}$ allows, for an example, a friction cone capped with some upper limit (ex to simulate plastic contacts that cannot withstand excessive pressure) or a friction cone shifted a bit downward (ex to simulate cohesion in contacts) etc.
-
-Under this more general setting, the time stepping of the multibody model becomes:
-
-\begin{subequations}
-	\begin{empheq}[left=\empheqlbrace]{align}
-    H \vect{v} - \vect{k} - {D}_\epsilon \bar{\vect{\gamma}}_\epsilon &= 0 \label{eq:ChronoVI_primal}\\
-    {D}_{\epsilon}^T \vect{v}  + \vect{b}_\epsilon &= \vect{u}_\epsilon\\
-    \bar{\vect{\gamma}}_\epsilon \in \mathcal{C} : \quad  \vect{u}_\epsilon &\in - \mathcal{N}_{C}(\bar{\vect{\gamma}}_\epsilon)  
-	\end{empheq}
-	\label{eq:ChronoVI}
-\end{subequations}
-
-This is a Variational Inequality (VI), and it contains the CCP of \eqref{eq:ChronoCCP_kkt} as a special case. In fact one can introduce a Schur matrix 
-\[
-N=D_{\epsilon}^T H^{-1} D_{\epsilon}
-\] similarly to what presented in  IDEA "A", and rewrite it as:
+The ADMM method in \cite{Stellato2020} can be used to solve problems in the form
 \begin{subequations}
 	\begin{align}
-    N \bar{\vect{\gamma}}_\epsilon + \vect{r}_\epsilon - \vect{u}_\epsilon &= 0 \\
-    \bar{\vect{\gamma}}_\epsilon \in \mathcal{C} : \quad  -\vect{u}_\epsilon &\in \mathcal{N}_{C}(\bar{\vect{\gamma}}_\epsilon) 
+    P \vect{x} + \vect{q} + A^T \vect{y} &= \vect{0} \\
+    A \vect{x} - \vect{b} &= \vect{z} \\
+    \mathcal{C} \ni \vect{z}  \quad \bot &\quad \vect{y} \in \mathcal{N}_{C}(\vect{z})
 	\end{align}
-	\label{eq:ChronoVI_schur}
+	\label{eq:OQSP_kkt}
 \end{subequations}
-Also, using the same symbols introduced in OSQP, with $\vect{y}=-\vect{u}_\epsilon$, one gets
-\begin{subequations}
-	\begin{align}
-    N \bar{\vect{\gamma}}_\epsilon + \vect{r}_\epsilon + I \vect{y} &= 0 \\
-    \bar{\vect{\gamma}}_\epsilon &= \vect{z} \\
-    \vect{z} \in \mathcal{C} : \quad  \vect{y} &\in \mathcal{N}_{C}(\bar{\vect{\gamma}}_\epsilon) 
-	\end{align}
-	\label{eq:ChronoVI_schur_kkt}
-\end{subequations}
-where one can easily see that this is a VI$(\vect{y}(\bar{\vect{\gamma}}_\epsilon), \mathcal{C})$ and that this is also the first order optimality condition of a convex quadratic program:
-\begin{subequations}
-	\begin{empheq}[box=\fbox]{align}
-	\text{min} \quad & \frac{1}{2} \bar{\vect{\gamma}}_\epsilon^T N \bar{\vect{\gamma}}_\epsilon + \vect{r}^T_\epsilon \bar{\vect{\gamma}}_\epsilon \\
-	\text{s.t.} \quad & \bar{\vect{\gamma}}_\epsilon \in \mathcal{C}
-	\end{empheq}
-	\label{eq:ChronoVI2_min}
-\end{subequations}
+and one can see that \eqref{eq:ChronoCCP_kkt} is a special case of \eqref{eq:OQSP_kkt} where $A=I$, $P=N$, $\mathcal{C}=\Upsilon$, $\vect{q}=\vect{r}$, $\vect{x}=\vect{\gamma}_\epsilon$, $\vect{b}=\vect{0}$, where some optimizations can take place because of the structure of our problem.
 
-Again, it is possible to apply ADMM algorithms to solve the problem.
+The ADMM method requires a fast way to perform a projection $\Pi_{\Upsilon}$ onto the friction cones $\Upsilon$ and a fast way to solve a linear problem with the matrix $A$.
+
+We remark that the ADMM method just makes the assumption of $\Upsilon$ being convex, so the problem can be generalized to $\vect{\gamma}_\epsilon \in \mathcal{C}$ where $\mathcal{C}$ is a generic convex set and at the same time we assume an associated \footnote{Not all dissipative constitutive models are associated, for example in some computational plasticity models the plastic flow might deviate respect to the normal to the yeld surface, and the Coulmb contact model itself is not associated -in fact the contact velocity $\vect{u}$ can range into a wider cone than the polar of the friction cone, but our DVI model compensates fur such effect.
+We will assume associated models heretofore, and we do not deal with non-associated models to avoid additional complexity for the moment.} flow $\vect{u}_\epsilon \in - \mathcal{N}_{C}(\vect{\gamma}_\epsilon)$.
+For instance, $\mathcal{C}$ could be a capped friction cone to represent plasticization of contacts, or a cone translated downward to represent possible adhesion in contact up to a threshold, or the Von Mises yield region if $\vect{\gamma}_\epsilon$ represent stresses in finite elments undergoing plasticization.
+
+Also, in sake of highest generality, in presence of finite elements one might need to use tangent stiffness matrices $K$ and damping matrices $D$ to accommodate an implicit integration scheme for stiff elements, and this means that the Schur complement would be computed as $N=D_{\epsilon}^T H^{-1} D_{\epsilon}$.
+Here $H$, taking the place of the original $M$ matrix, is rather a linear combination of $M$, $K$, $D$ (depending on the integration scheme) and in general is not diagonal anymore.
+This means that one cannot easily compute the Schur complement $N$, even in factored form, and this has some consequences on the implementation of the ADMM, that in fact we will rewrite in an unrolled form.
 
@@
 
-...
-
-[TO REWRITE THE FOLLOWING]
-
-The ADMM starts by stating a penalized objective function, that includes all the constraints by means of indicator functions:
+For variables $(\vect{x},\vect{z}) \in \mathbb{R}^m \times \mathbb{R}^n$, the ADMM methods solve constrained optimization problems with separable structure
 \begin{subequations}
 	\begin{align}
-	\text{min} \quad & \frac{1}{2} \vect{v}^T M \vect{v} - \vect{k}^T\vect{v} + \mathcal{I}_{\bar{D}_{\epsilon}^T \vect{v} + \vect{b}_\epsilon  = \vect{u}_\epsilon} (\tilde{\vect{v}}, \tilde{\vect{u}}_\epsilon) +
-	\mathcal{I}_{\mathcal{K}}(\vect{u}_\epsilon)\\
-	\text{s.t.} \quad & (\tilde{\vect{v}}, \tilde{\vect{u}}_\epsilon) = (\vect{v}, \vect{u}_\epsilon)
+    \text{min} \quad  & f(\vect{x}) + g(\vect{z}) \\
+	  \text{s.t.} \quad &  A \vect{x} + B \vect{z} - \vect{b} = \vect{0}
 	\end{align}
+	\label{eq:admm}
 \end{subequations}
-where $\mathcal{I}_{\bar{D}_{\epsilon}^T \vect{v} + \vect{b}_\epsilon  = \vect{u}_\epsilon}$ and $\mathcal{I}_{\mathcal{K}}$ are infinite if the constraint is violated and zero if not.
-This introduces a really strong barrier, yet non-smooth.
+by introducing an augmented Lagrangian
+\begin{align}
+\mathcal{L}_{\rho}(\vect{x},\vect{z},\vect{y})= f(\vect{x}) + g(\vect{z}) + \vect{y}^T (A \vect{x} + B \vect{z} - \vect{b}) + \frac{\rho}{2} \norm{A \vect{x} + B \vect{z} - \vect{b}}
+\end{align}
+and by iterating over two minimization problems as in the following loop:
+\begin{align}
+ \vect{x}^{k+1} &\in \text{argmin} \mathcal{L}_{\rho}(\vect{x},\vect{z}^k,\vect{y}^k) \\
+ \vect{z}^{k+1} &\in \text{argmin} \mathcal{L}_{\rho}(\vect{x}^{k+1},\vect{z},\vect{y}^k) \\
+ \vect{y}^{k+1} &= \vect{y}^{k} + \rho ( A \vect{x}^{k+1} + B \vect{z}^{k+1} - \vect{b} )
+ \label{eq:admm_iters}
+\end{align}
+Although the convergence is fast in the first iterations, it tends to deteriorate later as in fixed-point iterations; however in practical scenarios where loose tolerances can be accepted, the method proves to be very efficient and robust.
 
-The associated augmented lagragian is then used to build a equality constrained QP problem that is in the form:
+In order to rewrite \eqref{eq:ChronoCCP_min} as a sum of two functions as in \eqref{eq:admm}, we introduce the auxiliary variables 
+$(\tilde{\vect{\gamma}}_\epsilon,\tilde{\vect{z}})$, 
+we state the $\vect{\gamma}_\epsilon \in \Upsilon$ constraint by adding a non-smooth penalty function given by the indicator function 
+$\mathcal{I}_\Upsilon (\vect{\gamma}_\epsilon)$ 
+where $\mathcal{I}_\Upsilon(\vect{\gamma}_\epsilon)=0$ for $\vect{\gamma}_\epsilon \in \Upsilon$ and $=\infty$ otherwise, another indicator function 
+$\mathcal{I}_{\vect{\gamma}_\epsilon = \vect{z}}(\vect{\gamma}_\epsilon,\vect{z})$,
+and finally we have
 \begin{subequations}
 	\begin{align}
-	\text{min} \quad & \frac{1}{2} \tilde{\vect{v}}^T M \tilde{\vect{v}} - \vect{k}^T\tilde{\vect{v}} + \frac{\sigma}{2}\norm{\tilde{\vect{v}}-\vect{v}^k}^2 + \frac{\rho}{2}\norm{\tilde{\vect{u}_\epsilon}-\vect{u}_\epsilon^k + \frac{\bar{\vect{\gamma}}_\epsilon}{\rho}}^2\\
-	\text{s.t.} \quad & \bar{D}_{\epsilon}^T \tilde{\vect{v}} + \vect{b}_\epsilon  = \vect{u}_\epsilon
+    \text{min} \quad & \frac{1}{2} \tilde{\vect{\gamma}}_\epsilon^T N \tilde{\vect{\gamma}}_\epsilon + \vect{r}^T_\epsilon \tilde{\vect{\gamma}}_\epsilon  +  \mathcal{I}_{\tilde{\vect{\gamma}}_\epsilon = \tilde{\vect{z}}}(\tilde{\vect{\gamma}}_\epsilon,\tilde{\vect{z}}) +  \mathcal{I}_\Upsilon(\vect{\gamma}_\epsilon)  \\
+	  \text{s.t.} \quad & (\tilde{\vect{\gamma}}_\epsilon,\tilde{\vect{z}})  =  (\vect{\gamma}_\epsilon,\vect{z})
 	\end{align}
+	\label{eq:admm_mod}
 \end{subequations}
-whose KKT conditions can be found as:
-\begin{equation}
-    \begin{bmatrix}
-        M + \sigma I_n & \bar{D}_{\epsilon} \\
-        \bar{D}_{\epsilon}^T & -\rho^{-1} I_m
-    \end{bmatrix}
-    \begin{bmatrix}
-        \tilde{\vect{v}}^{k+1} \\
-        \vect{\nu}^{k+1}
-    \end{bmatrix}
-    =
-    \begin{bmatrix}
-        \vect{k} + \sigma \vect{v}^k \\
-        \vect{b}_\epsilon - \vect{u}_\epsilon^k + \rho^{-1}\bar{\vect{\gamma}}_\epsilon^k
-    \end{bmatrix}
-\end{equation}
 
-then we can update variables as
+We can write the augmented Lagrangian introducing two step size parameters $\sigma$ and $\rho$ and two vector of multipliers $\vect{w}$ and $\vect{y}$, obtaining:
+\begin{align}
+\mathcal{L}_{\sigma,\rho} \left((\tilde{\vect{\gamma}}_\epsilon,\tilde{\vect{z}}),(\vect{\gamma}_\epsilon,\vect{z}),(\vect{y},\vect{w})\right) &= 
+\frac{1}{2} \tilde{\vect{\gamma}}_\epsilon^T N \tilde{\vect{\gamma}}_\epsilon + \vect{r}^T_\epsilon \tilde{\vect{\gamma}}_\epsilon  
++  \mathcal{I}_{\tilde{\vect{\gamma}}_\epsilon = \tilde{\vect{z}}}(\tilde{\vect{\gamma}}_\epsilon,\tilde{\vect{z}}) 
++  \mathcal{I}_\Upsilon(\vect{\gamma}_\epsilon) \nonumber \\
++ \vect{w}^T (\tilde{\vect{\gamma}}_\epsilon - \vect{\gamma}_\epsilon) 
+&+ \frac{\sigma}{2} \norm{\tilde{\vect{\gamma}}_\epsilon - \vect{\gamma}_\epsilon} 
++ \vect{y}^T (\tilde{\vect{z}} - \vect{z}) 
++ \frac{\rho}{2} \norm{\tilde{\vect{z}} - \vect{z}}
+\end{align}
+that is also, using the property $\vect{y}^T\vect{r} + \frac{\rho}{2}\norm{\vect{r}} = \frac{\rho}{2}\norm{\vect{r}+\frac{1}{\rho} \vect{y}} - \frac{1}{2\rho} \norm{y}$, the following
+\begin{align}
+\mathcal{L}_{\sigma,\rho} \left((\tilde{\vect{\gamma}}_\epsilon,\tilde{\vect{z}}),(\vect{\gamma}_\epsilon,\vect{z}),(\vect{y},\vect{w})\right) &= 
+\frac{1}{2} \tilde{\vect{\gamma}}_\epsilon^T N \tilde{\vect{\gamma}}_\epsilon + \vect{r}^T_\epsilon \tilde{\vect{\gamma}}_\epsilon  
++  \mathcal{I}_{\tilde{\vect{\gamma}}_\epsilon = \tilde{\vect{z}}}(\tilde{\vect{\gamma}}_\epsilon,\tilde{\vect{z}}) 
++  \mathcal{I}_\Upsilon(\vect{\gamma}_\epsilon) \nonumber \\
+&+ \frac{\sigma}{2} \norm{\tilde{\vect{\gamma}}_\epsilon - \vect{\gamma}_\epsilon + \frac{1}{\sigma} \vect{w}} 
+- \frac{1}{2\sigma} \norm{w}
++ \frac{\rho}{2} \norm{\tilde{\vect{z}} - \vect{z} + \frac{1}{\rho} \vect{y}}
+- \frac{1}{2\rho} \norm{y}
+\end{align}
+
+The first step of ADMM requires to compute 
+\[
+(\tilde{\vect{\gamma}}_\epsilon^{k+1},\tilde{\vect{z}}^{k+1}) = \text{argmin} \mathcal{L}_{\sigma,\rho} \left((\tilde{\vect{\gamma}}_\epsilon,\tilde{\vect{z}}),(\vect{\gamma}_\epsilon^k,\vect{z}^k),(\vect{y}^k,\vect{w}^k)\right)
+\]. 
+This can be simplified remembering that $\vect{w}^k$ will always be zero, that some terms are constant and can be left out, and that the $\mathcal{I}_{\tilde{\vect{\gamma}}_\epsilon = \tilde{\vect{z}}}(\tilde{\vect{\gamma}}_\epsilon,\tilde{\vect{z}})$ term can be elided to turn finally the unconstrained $\text{argmin}$ into the equality constrained optimization problem:
 \begin{subequations}
 	\begin{align}
-	\tilde{\vect{u}}_\epsilon^{k+1} &= \vect{u}_\epsilon^k - \rho^{-1}(\vect{\nu}^{k+1} + \bar{\vect{\gamma}}_\epsilon^k)\\
-	\vect{v}^{k+1} &= \alpha \tilde{\vect{v}}^{k+1} + (1-\alpha) \vect{v}^k\\
-	\vect{u}_\epsilon^{k+1} &= \Pi_\mathcal{K}( \alpha \tilde{\vect{u}_\epsilon}^{k+1} + (1-\alpha) \vect{u}_\epsilon^k + \rho^{-1} \bar{\vect{\gamma}}_\epsilon^k ) \\
-	\bar{\vect{\gamma}}_\epsilon^{k+1} &= \bar{\vect{\gamma}}_\epsilon^k + \rho (  \alpha \tilde{\vect{u}_\epsilon}^{k+1} + (1-\alpha) \vect{u}_\epsilon^k - \vect{u}_\epsilon^{k+1} )
+    \text{min} \quad  &  \frac{1}{2} \tilde{\vect{\gamma}}_\epsilon^T N \tilde{\vect{\gamma}}_\epsilon + \vect{r}^T_\epsilon \tilde{\vect{\gamma}}_\epsilon  
++ \frac{\sigma}{2} \norm{\tilde{\vect{\gamma}}_\epsilon - \vect{\gamma}_\epsilon^k}
++ \frac{\rho}{2} \norm{\tilde{\vect{z}} - \vect{z}^k + \frac{1}{\rho} \vect{y}^k}
+	  \text{s.t.} \quad &  \tilde{\vect{\gamma}} = \tilde{\vect{z}}
 	\end{align}
+	\label{eq:admm_1a}
+\end{subequations}
+%
+For such problem, the KKT conditions lead to the following linear problem:
+\begin{subequations}
+	\begin{align}
+    \left[ N + (\sigma+\rho) I \right] \tilde{\vect{\gamma}}_\epsilon = (\sigma+\rho) \vect{\gamma}_\epsilon^k - \vect{r} - \vect{y}^k
+	\end{align}
+	\label{eq:admm_1b}
+\end{subequations}
+%
+The linear system \eqref{eq:admm_1b} can be solved as it is, but in our case it would be better to exploit the fact that the Schur matrix $N$ is a product $D_\epsilon^T H^{-1} D_\epsilon$.
+We would like to avoid computing $H^{-1}$, even storing a precomputed inverse for all iterations would be unpractical because often dense (except when $H$ is a diagonal mass matrix $M$).
+So we propose to replace \eqref{eq:admm_1b} with the equivalent saddle point problem:
+\begin{subequations}
+	\begin{align}
+    \begin{bmatrix}
+		 H   & D_\epsilon \\
+		 D^T & (\sigma+\rho) I
+		\end{bmatrix}
+		\begin{Bmatrix}
+		 \vect{v}    \\
+		 -\tilde{\vect{\gamma}}_\epsilon^{k+1} 
+		\end{Bmatrix}
+		=
+		\begin{Bmatrix}
+		 \vect{k} \\
+		 \vect{b}_\epsilon + (\sigma+\rho) \vect{\gamma}_\epsilon^k-\vect{y}^k 
+		\end{Bmatrix}
+	\end{align}
+	\label{eq:admm_1c}
+\end{subequations}
+that, even if it introduces an auxiliary variable $\vect{v}$, is very sparse and does not require computing any $H^{-1}$.
+
+A remarkable side effect of this approach is that its auxiliary variable $\vect{v}$ is also the velocity term in the original MDI 
+\eqref{eq:ChronoMCCP}, so one does not need to compute it as $\vect{v}=H^{-1}(\vect{k} + D_\epsilon \vect{\gamma}_\epsilon )$ after the iteration converged because it is a byproduct of the linear solver.
+
+The second step of ADMM is a minimization problem too, but it can be rephrased in terms of a projection on the separable set $\Upsilon$ of the friction cones, note that this is an operation that requires a very inexpensive and easily parallelizable operation.
+
+Using these results, and adding a Krasnoselskii-Mann relaxation factor $\alpha \in(0,2]$, one obain the final algorithm:
+%
+\begin{subequations}
+	\begin{align}
+    \begin{bmatrix}
+		 H   & D_\epsilon \\
+		 D^T & (\sigma+\rho) I
+		\end{bmatrix}
+		\begin{Bmatrix}
+		 \vect{v}    \\
+		 -\tilde{\vect{\gamma}}_\epsilon^{k+1}
+		\end{Bmatrix}
+		&=
+		\begin{Bmatrix}
+		 \vect{k} \\
+		 \vect{b}_\epsilon + (\sigma+\rho) \vect{\gamma}_\epsilon^k-\vect{y}^k 
+		\end{Bmatrix}
+		\\
+		\vect{\gamma}^{k+1} &=\tilde{\vect{\gamma}}_\epsilon^{k+1} + (1-\alpha) \vect{\gamma}_\epsilon^k \\
+		\vect{z}^{k+1} &= \Pi_\Upsilon \left( \alpha \vect{\gamma}_\epsilon^{k+1} + (1-\alpha) \vect{z}^k  - \frac{1}{\rho} \vect{y}^k \right) \\
+		\vect{y}^{k+1} &= \vect{y}^k + \rho \left( \alpha \vect{\gamma}_\epsilon^{k+1} + (1-\alpha) \vect{z}^k - \vect{z}^{k+1} \right) \\
+		\vect{r}_{\text{prim}} &= \vect{\gamma}_\epsilon^{k+1}-\vect{z}^{k+1} \\
+		\vect{r}_{\text{dual}} &= N \vect{\gamma}_\epsilon^{k+1}+\vect{r}-\vect{y}^{k+1} 
+	\end{align}
+	\label{eq:admm_final}
 \end{subequations}
 
-...
+The iteration is terminated when $\vect{r}_{\text{prim}}$ and $\vect{r}_{\text{dual}}$ fall under prescribed tolerances.
+We remark that setting proper values for $\rho$ is of fundamental importance in order to obtain a good speed of convergence.
+In most cases one can use heuristics and adjust $\rho$ only one or two times during the iterations, hence allowing to reuse the factorization of the linear system.
+Instead, $\sigma$ acts similarly to a regularization term, and can be left to low values anyway.
 

### 2020-09-19 2021/NODYCON_nonsmooth_ADMM/author.tex
Almost finl version - check equations 16 (especially the signs) and steps in equations 17.x

@@
 a specialized ADMM method for quadratic problems with conic constraints has been presented: on the top of this method we develop a solver that can exploit the nature of non-smooth dynamical problems.
-In our embodiment, unknowns are primal variables (velocity measures) and dual variables (impulses in contact points and joints), cone constraints stem from the Coulomb friction law, and the system matrix includes terms from the mass matrices and from the tangent stiffness/damping matrices (hence it is sparse, but not necessarily diagonal).
 
-This paper will present a model for the non-smooth dynamics, it will show how to cast it as a QP with convex conic constraint, it will present the ADMM method to solve it, it will discuss practical implementation details and then it will show some benchmarks.
+% TO ADD IN EXTENDED VERSION
+%In our embodiment, unknowns are primal variables (velocity measures) and dual variables (impulses in contact points and joints), cone constraints stem from the Coulomb friction law, and the system matrix includes terms from the mass matrices and from the tangent stiffness/damping matrices (hence it is sparse, but not necessarily diagonal).
+%This paper will present a model for the non-smooth dynamics, it will show how to cast it as a QP with convex conic constraint, it will present the ADMM method to solve it, it will discuss practical implementation details and then it will show some benchmarks.
 
@@
 
-The ADMM method requires a fast way to perform a projection $\Pi_{\Upsilon}$ onto the friction cones $\Upsilon$ and a fast way to solve a linear problem with the matrix $A$.
+% TO ADD IN EXTENDED VERSION
+%The ADMM method requires a fast way to perform a projection $\Pi_{\Upsilon}$ onto the friction cones $\Upsilon$ and a fast way to solve a linear problem with the matrix $A$.
 
-We remark that the ADMM method just makes the assumption of $\Upsilon$ being convex, so the problem can be generalized to $\vect{\gamma}_\epsilon \in \mathcal{C}$ where $\mathcal{C}$ is a generic convex set and at the same time we assume an associated \footnote{Not all dissipative constitutive models are associated, for example in some computational plasticity models the plastic flow might deviate respect to the normal to the yeld surface, and the Coulmb contact model itself is not associated -in fact the contact velocity $\vect{u}$ can range into a wider cone than the polar of the friction cone, but our DVI model compensates fur such effect.
-We will assume associated models heretofore, and we do not deal with non-associated models to avoid additional complexity for the moment.} flow $\vect{u}_\epsilon \in - \mathcal{N}_{C}(\vect{\gamma}_\epsilon)$.
+We remark that the ADMM method just makes the assumption of $\Upsilon$ being convex, so the problem can be generalized to $\vect{\gamma}_\epsilon \in \mathcal{C}$ where $\mathcal{C}$ is a generic convex set and at the same time we assume an associated
+% TO ADD IN EXTENDED VERSION
+%\footnote{Not all dissipative constitutive models are associated, for example in some computational plasticity models the plastic flow might deviate respect to the normal to the yeld surface, and the Coulmb contact model itself is not associated -in fact the contact velocity $\vect{u}$ can range into a wider cone than the polar of the friction cone, but our DVI model compensates fur such effect.
+We will assume associated models heretofore, and we do not deal with non-associated models to avoid additional complexity for the moment.}
+flow $\vect{u}_\epsilon \in - \mathcal{N}_{C}(\vect{\gamma}_\epsilon)$.
 For instance, $\mathcal{C}$ could be a capped friction cone to represent plasticization of contacts, or a cone translated downward to represent possible adhesion in contact up to a threshold, or the Von Mises yield region if $\vect{\gamma}_\epsilon$ represent stresses in finite elments undergoing plasticization.
@@
 Here $H$, taking the place of the original $M$ matrix, is rather a linear combination of $M$, $K$, $D$ (depending on the integration scheme) and in general is not diagonal anymore.
-This means that one cannot easily compute the Schur complement $N$, even in factored form, and this has some consequences on the implementation of the ADMM, that in fact we will rewrite in an unrolled form.
+
+% TO ADD IN EXTENDED VERSION
+%This means that one cannot easily compute the Schur complement $N$, even in factored form, and this has some consequences on the implementation of the ADMM, that in fact we will rewrite in an unrolled form.
 
@@
 +  \mathcal{I}_\Upsilon(\vect{\gamma}_\epsilon) \nonumber \\
-&+ \frac{\sigma}{2} \norm{\tilde{\vect{\gamma}}_\epsilon - \vect{\gamma}_\epsilon + \frac{1}{\sigma} \vect{w}} 
-- \frac{1}{2\sigma} \norm{w}
++ \frac{\sigma}{2} \norm{\tilde{\vect{\gamma}}_\epsilon - \vect{\gamma}_\epsilon + \frac{1}{\sigma} \vect{w}} 
+&- \frac{1}{2\sigma} \norm{w}
 + \frac{\rho}{2} \norm{\tilde{\vect{z}} - \vect{z} + \frac{1}{\rho} \vect{y}}
@@
 (\tilde{\vect{\gamma}}_\epsilon^{k+1},\tilde{\vect{z}}^{k+1}) = \text{argmin} \mathcal{L}_{\sigma,\rho} \left((\tilde{\vect{\gamma}}_\epsilon,\tilde{\vect{z}}),(\vect{\gamma}_\epsilon^k,\vect{z}^k),(\vect{y}^k,\vect{w}^k)\right)
-\]. 
-This can be simplified remembering that $\vect{w}^k$ will always be zero, that some terms are constant and can be left out, and that the $\mathcal{I}_{\tilde{\vect{\gamma}}_\epsilon = \tilde{\vect{z}}}(\tilde{\vect{\gamma}}_\epsilon,\tilde{\vect{z}})$ term can be elided to turn finally the unconstrained $\text{argmin}$ into the equality constrained optimization problem:
+\] 
+This can be simplified remembering that $\vect{w}^k$ will always be zero, that some terms are constant and can be left out, and that the $\mathcal{I}_{\tilde{\vect{\gamma}}_\epsilon = \tilde{\vect{z}}}(\tilde{\vect{\gamma}}_\epsilon,\tilde{\vect{z}})$ term can be elided to turn finally the unconstrained $\text{argmin}\mathcal{L}_{\sigma,\rho}$ into the equality constrained optimization problem:
 \begin{subequations}
@@
 + \frac{\sigma}{2} \norm{\tilde{\vect{\gamma}}_\epsilon - \vect{\gamma}_\epsilon^k}
-+ \frac{\rho}{2} \norm{\tilde{\vect{z}} - \vect{z}^k + \frac{1}{\rho} \vect{y}^k}
++ \frac{\rho}{2} \norm{\tilde{\vect{z}} - \vect{z}^k + \frac{1}{\rho} \vect{y}^k} \\
 	  \text{s.t.} \quad &  \tilde{\vect{\gamma}} = \tilde{\vect{z}}
@@
 For such problem, the KKT conditions lead to the following linear problem:
-\begin{subequations}
 \begin{align}
     \left[ N + (\sigma+\rho) I \right] \tilde{\vect{\gamma}}_\epsilon = (\sigma+\rho) \vect{\gamma}_\epsilon^k - \vect{r} - \vect{y}^k
+		\label{eq:admm_1b}
 \end{align}
-	\label{eq:admm_1b}
-\end{subequations}
 %
@@
 
-The second step of ADMM is a minimization problem too, but it can be rephrased in terms of a projection on the separable set $\Upsilon$ of the friction cones, note that this is an operation that requires a very inexpensive and easily parallelizable operation.
+The second step of ADMM is a minimization problem too, but it can be rephrased in terms of a projection on the separable set $\Upsilon$ of the friction cones, note that this requires an inexpensive and parallelizable operation.
 
-Using these results, and adding a Krasnoselskii-Mann relaxation factor $\alpha \in(0,2]$, one obain the final algorithm:
+Using these results, and adding a Krasnoselskii-Mann relaxation factor $\alpha \in(0,2]$, one obtains the final algorithm:
 %
@@
 		\\
-		\vect{\gamma}^{k+1} &=\tilde{\vect{\gamma}}_\epsilon^{k+1} + (1-\alpha) \vect{\gamma}_\epsilon^k \\
+		\vect{\gamma}^{k+1}_\epsilon &=\tilde{\vect{\gamma}}_\epsilon^{k+1} + (1-\alpha) \vect{\gamma}_\epsilon^k \\
 		\vect{z}^{k+1} &= \Pi_\Upsilon \left( \alpha \vect{\gamma}_\epsilon^{k+1} + (1-\alpha) \vect{z}^k  - \frac{1}{\rho} \vect{y}^k \right) \\
@@
 In most cases one can use heuristics and adjust $\rho$ only one or two times during the iterations, hence allowing to reuse the factorization of the linear system.
-Instead, $\sigma$ acts similarly to a regularization term, and can be left to low values anyway.
-
-\section{Results}
-
-...
+Also, $\sigma$ acts similarly to a regularization term, and can be left to low values anyway.
 
@@
 \caption{Convergence of the ADMMM method within one time step of the wrecking ball benchmark (600 bricks in four walls): residual in frictional constraint violation compared to fixed point Jacobi iterations and to first-order SPG methods.}
+\label{fig:t8}
 \end{figure}
 
-\section{Conclusion}
+\section{Results and conclusion}
+
+We performed benchmarks involving multi-body systems with contacts between multiple parts, showing that the performance of the ADMM method is capable of handling problems that would converge too slowly using conventional projected fixed point methods (see Fig.\ref{fig:t8}) or even first-order SPG spectral methods \cite{heynIJNME2013}.
 
@@
 A good estimation of the ADMM step size proved to be fundamental in achieving good convergence: using some heuristics we obtained an efficient auto-tuning algorithm. 
-We noted that ADMM can be successfully applied to problems that exhibit temporal coherence because, unlike IP methods, it supports warm-starting.
+We noted that ADMM can be successfully applied to problems that exhibit temporal coherence because, unlike IPMs, it supports warm-starting.
 

### 2020-10-21 2021/ADMM_journal/author.tex
Add stub of article for a journal, on ADMM topic - better and more detailed version of the Nodycon2021 paper

--- 2020-09-19 2021/NODYCON_nonsmooth_ADMM/author.tex
@@
-%%%%%%%%%%%%%%%% based on Springer template for conference proceedings %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%
+%% This document created by A.Tasora
+%%
 
-\documentclass{svproc}
+\documentclass[1p]{elsarticle}
 
@@
 \usepackage{url}
-\usepackage[T1]{fontenc}
+%\usepackage[T1]{fontenc}
 \usepackage[utf8]{inputenc}
 \usepackage[english]{babel}
-\usepackage{csquotes}
-
 \usepackage{hyperref}
@@
 \usepackage{tabularx}
-\usepackage{algorithm}
-\usepackage{algpseudocode}
+%\usepackage{algorithm}
+%\usepackage{algpseudocode}
+\usepackage[ruled,vlined, linesnumbered]{algorithm2e}
 \usepackage{array}
@@
 
+%% A.TASORA CUSTOM TEX COMMANDS:
+%%
+
+\def\avect#1{{\boldsymbol{#1}}}
+\def\amatr#1{{\boldsymbol{#1}}}
+
 \newcommand{\vect}[1]{\bm{#1}}
@@
 \begin{document}
-\mainmatter              % start of a contribution
-%
-\title{Solving Non-smooth Dynamic Problems using \\ the 
-Alternating Direction Method of Multipliers}
-%
-\titlerunning{Non-smooth dynamic problem}  % abbreviated title (for running head)
-%                                     also used for the TOC unless
-%                                     \toctitle is used
-%
-\author{Alessandro Tasora \and Dario Mangoni \and Simone Benatti}
-%
-\authorrunning{Alessandro Tasora et al.} % abbreviated author list (for running head)
-%
-%%%% list of authors for the TOC (use if author list has to be modified)
-\tocauthor{Alessandro Tasora, Dario Mangoni, and Simone Benatti}
-%
-\institute{University of Parma, Department of Engineering and Architecture \\ V.le delle Scienze 181/A, 40121 ITALY\\
-\email{alessandro.tasora@unipr.it},\\ WWW home page:
-\texttt{http://digitaldynamicslab.unipr.it}
-}
+
+%%
+%% TITLE
+%%
+
+\title{Solving non smooth dynamics using ADMM}
+
+%\author{Alessandro Tasora \\ alessandro.tasora@unipr.it}
+\author{Alessandro Tasora}
+\ead{alessandro.tasora@unipr.it}
+\author{Simone Benatti}
+\ead{simone.benatti@studenti.unipr.it}
+\author{Dario Mangoni}
+\ead{dario.mangoni@unipr.it}
+\address{Universit\`a degli Studi di Parma, \\Dpt. of Engineering and Architecture\\Parco Area delle Scienze, 181/A, 43124 Parma, Italy }
+
+% \affil[]{Universit\`a degli Studi di Parma, \\ 
+% Dpt. of Engineering and Architecture, Parco Area delle Scienze, 181/A, 43124 Parma, Italy 
+% \footnote{E-mail addresses: alessandro.tasora@unipr.it (A.
+Tasora), \\ simone.benatti@studenti.unipr.it (S.
+Benatti) dario.mangoni@studenti.unipr.it (D.
+Mangoni), rinaldo.garziera@unipr.it (R.Garziera).}
+% }
+
+\begin{abstract} 
+This work discusses an efficient formulation of a geometrically exact three-dimensional beam which can be used in dynamical simulations involving large displacements, collisions and non-linear materials.
+To this end, we base our model on the shear-flexible Cosserat rod theory and we implement it in the context of Isogeometric Analysis (IGA).
+According to the IGA approach, the centerline of the beam is parameterized using splines; in our work the rotation of the section is parameterized by a spline interpolation of quaternions, and time integration of rotations is performed using the exponential map of quaternions.
+Aiming at an efficient and robust simulation of contacts, we propose the adoption of a non-smooth dynamics formulation based on differential-variational inequalities.
+The model has been implemented in an open-source physics simulation library that can simulate actuators, finite elements, rigid bodies, constraints, collisions and frictional contacts. 
+This beam model has been tested on various benchmarks in order to assess its validity in non-linear static and dynamic analysis; in all cases the model behaved consistently with theoretical results and experimental data.
+\end{abstract}
+
+\begin{keyword}
+ADMM \sep non-smooth dynamics \sep contact
+\end{keyword}
 
@@
 Another attractive feature is that it can be easily warm-started.
-\keywords{ADMM, non-smooth dynamics, contact}
 \end{abstract}
 %
+\begin{keyword}
+ADMM \sep non-smooth dynamics \sep contact
+\end{keyword}
 
@@
 
-Recently there has been a revival of Alternating Direction Method of Multipliers (ADMM) and similar operator-splitting methods for solving constrained optimization problems
-\cite{Goldstein2014}. %  [2014 Alternating Direction Optimization Methods]. 
-Even if the theoretical convergence of such methods is worse than the one of IPMs, recent developments showed that in practical scenarios they offer superior speed, scalability and robustness 
+The Alternating Direction Method of Multipliers (ADMM) has been proposed in the 1970s 
+\cite{Glowinski75} % [ ADMM invention ]
+\cite{Gabay1976} % [ ADMM invention ]
+as a practical way to solve constrained optimization problems by iterating over the minimization of two (not necessarily differentiable) objectives.
+ADMM introduces auxiliary variables and cast the original problem as the minimization of a separable objective subject to a linear constraint between the original variables and the auxiliary variables.
+With a proper choice of auxiliary variables, the minimization of the separable function is obtained iterating on two distinct minimization steps of lower complexity.
+The method iterates updating primal variables, auxiliary variables and dual variables similarly to a fixed point iteration: as such, they compares less favorably to IPMs if one is interested in convergence to high precision, however recent developments showed that in practical scenarios they offer superior speed, scalability and robustness, moreover their structure map well into parallel computing architectures
+ \cite{Boyd2011}. % [ ADMM general theory and development ].
+These positive properties triggered a recent revival of ADMM and similar operator-splitting methods even as alternatives to IPMs, especially in distributed convex optimization with large-scale problems where moderate precision is acceptable
 \cite{Cannon2019}. % [COSMO  conic operator splitting method for large convex problems].
-An attractive property of ADMM is that it requires the solution of a linear system per each iteration, but unlike IPMs, such linear system most often remains unchanged during the iterations, so a factorization can be reused multiple times with a great benefit in terms of speed.
+
+ADMM methods have been used in many fields, for instance computer graphics 
+ \cite{Zhang2019}
+, computational fluid dynamics
+ \cite{Gregson2014} % [ ADMM and fluid simulation, inverse problems ]
+, simulation of deformable structures 
+ \cite{Overby2017} % [ ADMM projective dynamics hyperelastic models ]
+and contact dynamics 
+ \cite{Daviet2020} % [ ADMM thin objects frictional contact ]
+, to name a few.
+
+Recent developments addressed the possibility of accelerating the ADMM method.
+In 
+ \cite{Goldfarb2013} % [ ADMM with acceleration, both convex (non-strongly-convex require step skipping), one differentiable]
+an acceleration scheme has been proposed for special cases of symmetric ADMMs, and assuming that one of the two objectives is differentiable.
+%
+In 
+ \cite{Goldstein2014} % [ ADMM and AMA with Nesterov. Convergence ok if both problems are strongly convex (non-strongly-convex require restarting), one is quadratic ].
+the Nesterov acceleration method has been proposed for problems where at least one of the two objectives is quadratic, and under the assumption that both objectives are strongly convex; if the latter assumption is not guaranteed, restarting strategies are needed.
+%
+In 
+ \cite{Kadkhodaie2015} %  [ A2DM2 accelerated ADMM, strongly convex ]
+a similar acceleration method has been proposed, with $\mathcal{O}(\frac{1}{k^2})$ convergence rate, but without the assumption that one of the objectives must be quadratic.
+%
+Other types of improvements, based on the Anderson fixed point acceleration, have been put forward in 
+ \cite{Zhang2019} % [ ADMM + Anderson acceleration, computergraphics ].
+ \cite{Ouyang2020} % [ ADMM + Anderson acceleration, using Douglas-Rachford splitting, computergraphics ]
+, showing high performance in computer graphics applications.
+
+Recent developments also focused on the possibility of parallelizing the ADMM iterations on GPU and parallel architectures
+\cite{Schubiger2020} % [ GPU implementation of ADMM | no need to compute residuals each iteration | krylov solver each iteration ].
+.
 
@@
 a specialized ADMM method for quadratic problems with conic constraints has been presented: on the top of this method we develop a solver that can exploit the nature of non-smooth dynamical problems.
+An attractive property of such ADMM is that it requires the solution of a linear system per each iteration, but unlike IPMs, such linear system most often remains unchanged during the iterations, so a factorization can be reused multiple times with a great benefit in terms of speed.
 
-% TO ADD IN EXTENDED VERSION
-%In our embodiment, unknowns are primal variables (velocity measures) and dual variables (impulses in contact points and joints), cone constraints stem from the Coulomb friction law, and the system matrix includes terms from the mass matrices and from the tangent stiffness/damping matrices (hence it is sparse, but not necessarily diagonal).
-%This paper will present a model for the non-smooth dynamics, it will show how to cast it as a QP with convex conic constraint, it will present the ADMM method to solve it, it will discuss practical implementation details and then it will show some benchmarks.
+In our embodiment, unknowns are primal variables (velocity measures) and dual variables (impulses in contact points and joints), the convex cone constraints stem from the Coulomb friction law, and the system matrix includes terms from the mass matrices and from the tangent stiffness/damping matrices (hence it is sparse, but not necessarily diagonal).
+
+This paper will present a model for the non-smooth dynamics, it will show how to cast it as a QP with convex conic constraint, it will present the ADMM method to solve it, it will discuss practical implementation details and then it will show benchmarks.
 
@@
 We remark that the ADMM method just makes the assumption of $\Upsilon$ being convex, so the problem can be generalized to $\vect{\gamma}_\epsilon \in \mathcal{C}$ where $\mathcal{C}$ is a generic convex set and at the same time we assume an associated
-% TO ADD IN EXTENDED VERSION
-%\footnote{Not all dissipative constitutive models are associated, for example in some computational plasticity models the plastic flow might deviate respect to the normal to the yeld surface, and the Coulmb contact model itself is not associated -in fact the contact velocity $\vect{u}$ can range into a wider cone than the polar of the friction cone, but our DVI model compensates fur such effect.
+ \footnote{Not all dissipative constitutive models are associated, for example in some computational plasticity models the plastic flow might deviate respect to the normal to the yeld surface, and the Coulomb contact model itself is not associated -in fact the contact velocity $\vect{u}$ can range into a wider cone than the polar of the friction cone, but our DVI model compensates fur such effect.
 We will assume associated models heretofore, and we do not deal with non-associated models to avoid additional complexity for the moment.}
 flow $\vect{u}_\epsilon \in - \mathcal{N}_{C}(\vect{\gamma}_\epsilon)$.
-For instance, $\mathcal{C}$ could be a capped friction cone to represent plasticization of contacts, or a cone translated downward to represent possible adhesion in contact up to a threshold, or the Von Mises yield region if $\vect{\gamma}_\epsilon$ represent stresses in finite elments undergoing plasticization.
+For instance, $\mathcal{C}$ could be a capped friction cone to represent plasticization of contacts, or a cone translated downward to represent possible adhesion in contact up to a threshold, or the Von Mises yield region if $\vect{\gamma}_\epsilon$ represent stresses in finite elements undergoing plasticization.
 
@@
 \begin{align}
- \vect{x}^{k+1} &\in \text{argmin} \mathcal{L}_{\rho}(\vect{x},\vect{z}^k,\vect{y}^k) \\
- \vect{z}^{k+1} &\in \text{argmin} \mathcal{L}_{\rho}(\vect{x}^{k+1},\vect{z},\vect{y}^k) \\
+ \vect{x}^{k+1} &\in \text{argmin} \mathcal{L}_{\rho}(\vect{x},\vect{z}^k,\vect{y}^k) \label{eq:lag1}\\
+ \vect{z}^{k+1} &\in \text{argmin} \mathcal{L}_{\rho}(\vect{x}^{k+1},\vect{z},\vect{y}^k) \label{eq:lag2}\\
  \vect{y}^{k+1} &= \vect{y}^{k} + \rho ( A \vect{x}^{k+1} + B \vect{z}^{k+1} - \vect{b} )
@@
 Although the convergence is fast in the first iterations, it tends to deteriorate later as in fixed-point iterations; however in practical scenarios where loose tolerances can be accepted, the method proves to be very efficient and robust.
+
+The performance of ADMM depends on the efficiency of the updates in \eqref{eq:lag1} and \eqref{eq:lag2}.
+We will design the method such that the step \eqref{eq:lag2} will correspond to a very efficient projection onto a cone with separable structure, whereas the only bottleneck will be \eqref{eq:lag1}, corresponding to a quadratic optimization to be solved with a linear system.
 
@@
 \end{align}
-that is also, using the property $\vect{y}^T\vect{r} + \frac{\rho}{2}\norm{\vect{r}} = \frac{\rho}{2}\norm{\vect{r}+\frac{1}{\rho} \vect{y}} - \frac{1}{2\rho} \norm{y}$, the following
+that is also, using the property $\vect{y}^T\vect{r} + \frac{\rho}{2}\norm{\vect{r}} = \frac{\rho}{2}\norm{\vect{r}+\frac{1}{\rho} \vect{y}} - \frac{1}{2\rho} \norm{\vect{y}}$, the following
 \begin{align}
@@
 + \frac{\sigma}{2} \norm{\tilde{\vect{\gamma}}_\epsilon - \vect{\gamma}_\epsilon + \frac{1}{\sigma} \vect{w}} 
-&- \frac{1}{2\sigma} \norm{w}
+&- \frac{1}{2\sigma} \norm{\vect{w}}
 + \frac{\rho}{2} \norm{\tilde{\vect{z}} - \vect{z} + \frac{1}{\rho} \vect{y}}
-- \frac{1}{2\rho} \norm{y}
+- \frac{1}{2\rho} \norm{\vect{y}}
 \end{align}
@@
 \begin{align}
-    \left[ N + (\sigma+\rho) I \right] \tilde{\vect{\gamma}}_\epsilon = (\sigma+\rho) \vect{\gamma}_\epsilon^k - \vect{r} - \vect{y}^k
+    \left[ N + (\sigma+\rho) I \right] \tilde{\vect{\gamma}}_\epsilon = \sigma \vect{\gamma}_\epsilon^k + \rho \vect{z}^k - \vect{r} - \vect{y}^k
 		\label{eq:admm_1b}
@@
 		 H   & D_\epsilon \\
-		 D^T & (\sigma+\rho) I
+		 D^T & -(\sigma+\rho) I
 		\end{bmatrix}
@@
 		 \vect{k} \\
-		 \vect{b}_\epsilon + (\sigma+\rho) \vect{\gamma}_\epsilon^k-\vect{y}^k 
+		 -\vect{b}_\epsilon + \sigma \vect{\gamma}_\epsilon^k + \rho \vect{z}^k -\vect{y}^k 
 		\end{Bmatrix}
@@
 		 H   & D_\epsilon \\
-		 D^T & (\sigma+\rho) I
+		 D^T & -(\sigma+\rho) I
 		\end{bmatrix}
 		\begin{Bmatrix}
-		 \vect{v}    \\
+		  \tilde{\vect{v}}^{k+1}  \\
 		 -\tilde{\vect{\gamma}}_\epsilon^{k+1}
@@
 		 \vect{k} \\
-		 \vect{b}_\epsilon + (\sigma+\rho) \vect{\gamma}_\epsilon^k-\vect{y}^k 
+		 -\vect{b}_\epsilon + \sigma \vect{\gamma}_\epsilon^k +\rho \vect{z}^k - \vect{y}^k 
 		\end{Bmatrix}
@@
 		\vect{\gamma}^{k+1}_\epsilon &=\tilde{\vect{\gamma}}_\epsilon^{k+1} + (1-\alpha) \vect{\gamma}_\epsilon^k \\
-		\vect{z}^{k+1} &= \Pi_\Upsilon \left( \alpha \vect{\gamma}_\epsilon^{k+1} + (1-\alpha) \vect{z}^k  - \frac{1}{\rho} \vect{y}^k \right) \\
-		\vect{y}^{k+1} &= \vect{y}^k + \rho \left( \alpha \vect{\gamma}_\epsilon^{k+1} + (1-\alpha) \vect{z}^k - \vect{z}^{k+1} \right) \\
+		\vect{v}^{k+1} &=\tilde{\vect{v}}^{k+1} + (1-\alpha) \vect{v}^k \\
+		\vect{z}^{k+1} &= \Pi_\Upsilon \left( \alpha \tilde{\vect{\gamma}}_\epsilon^{k+1} + (1-\alpha) \vect{z}^k  + \frac{1}{\rho} \vect{y}^k \right) \\
+		\vect{y}^{k+1} &= \vect{y}^k + \rho \left( \alpha \tilde{\vect{\gamma}}_\epsilon^{k+1} + (1-\alpha) \vect{z}^k - \vect{z}^{k+1} \right) \\
 		\vect{r}_{\text{prim}} &= \vect{\gamma}_\epsilon^{k+1}-\vect{z}^{k+1} \\
@@
 Also, $\sigma$ acts similarly to a regularization term, and can be left to low values anyway.
+
+\section{Optimizations}
+
+An interesting property of this formulation is that one does not need to recover $\vect{v} = \amatr{H}^{-1}(\vect{k}+\amatr{D}^T \vect)\vect{\gamma}^{k+1}$ at the end because, for $\alpha=1$, this is also a byproduct of the linear system solve as $\vect{v}= \tilde{\vect{v}}^{k+1}$.
+In case of $\alpha\neq1$, one needs to keep it updated as 
+$\vect{v}^{k+1} =\tilde{\vect{v}}^{k+1} + (1-\alpha) \vect{v}^k$; 
+but this also means that an initial value of $\vect{v}^0$, consistent with the initial value $\vect{\gamma}^0$, must be computed, hence before the iteration one has to initialize 
+\[
+\vect{v}^0 = \amatr{H}^{-1}(\vect{k} + \amatr{D} \vect{\gamma}^0)
+\]
+
+We see that the dual residual $\vect{r}_{\text{dual}}$ is computed using the $\amatr{N}$ matrix, but can be inefficient as we do not want to compute it explicitly.
+One option is to remember the definition of the Schur complement and use
+\[
+\vect{r}_{\text{dual}} = \amatr{D}^T \vect{v}^{k+1} - \amatr{E} \vect{\gamma}^{k+1} + \vect{b}_\epsilon + \vect{y}
+\]
+Another even faster option (but only for $\alpha=1$) would be to compute the dual residual using
+\[
+\vect{r}_{\text{dual}} = \rho (\vect{z}^{k+1} - \vect{z}^{k} )
+\]
+
+TO COMPLETE
+
+\section{Step size selection}
+
+In general, small values of $\rho$ cause a fast reduction of the dual residual, but a slow reduction of the primal residual.
+This has a nice physical interpretation: truncating iterations based on too small $\rho$ most often lead to solutions where parts do not interpenetrate, but some contact forces do not satisfy the inclusion in the Coulomb cone.
+That is, some parts might stick like contacts were glued.
+
+On the other hand, large values of $\rho$ cause a fast reduction of the primal residual, but a slow reduction of the dual residual.
+This means that truncating iterations with too large $\rho$ most often lead to solutions where contact forces satisfy very well the inclusion in the Coulomb friction cones, but parts might interpenetrate.
+
+Clearly, this calls for the introduction of a scheme that adaptively changes the step size $\rho$ in order to achieve a balanced reduction of both residuals.
+
+TO DO
+
+Step size:
+
+In \cite{Wohlberg2017} [ a step size selection via residual balancing, scale residuals given problem measuring units]
+
+In \cite{Xu_adaptive_2017} [ Spectral stepsize selection as in Barzilai Borwein ]
+
+\section{Benchmarks}
+
+TO DO
 

### 2020-10-30 2021/ADMM_journal/author.tex
Update paper ADMM

@@
 
-\documentclass[1p]{elsarticle}
+\documentclass[final,3p]{elsarticle}
 
@@
 \usepackage{amsmath}
+\usepackage{amsthm}
 \usepackage{cleveref}
@@
 \usepackage{tabularx}
-%\usepackage{algorithm}
-%\usepackage{algpseudocode}
 \usepackage[ruled,vlined, linesnumbered]{algorithm2e}
@@
 
-\def\avect#1{{\boldsymbol{#1}}}
-\def\amatr#1{{\boldsymbol{#1}}}
+%\def\avect#1{{\boldsymbol{#1}}}
+\def\amatr#1{{#1}}
+\newcommand{\vect}[1]{\bm{#1}}
+%\newcommand{\mat}[1]{#1}
+\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
 
-\newcommand{\vect}[1]{\bm{#1}}
-\newcommand{\mat}[1]{#1}
-\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
+\DeclareMathOperator*{\argmin}{arg\,min} % thin space, limits underneath in displays
+
+\newtheorem*{definition*}{Definition}
+\newtheorem{definition}{Definition}
+\newtheorem*{remark*}{Remark}
+\newtheorem{remark}{Remark}
+\newtheorem*{theorem*}{Theorem}
+\newtheorem{theorem}{Theorem}[section]
+\newtheorem*{corollary*}{Corollary}
+\newtheorem{corollary}{Corollary}[theorem]
+\newtheorem{lemma}[theorem]{Lemma}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%5
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Definitions}
+
+In this section we list some basic definitions that we will use in the rest of the paper.
+
+\begin{definition}
+A \textit{second order cone}, also Lorentz cone, is a self-dual self-scaled symmetric cone defined as
+\begin{equation}
+\label{eq:lorentzcone}
+\mathcal{K} 
+= \left \{ (x_0,\vect{x}_1) \in \mathbb{R} \times \mathbb{R}^{p-1}: \norm{\vect{x}_1}_2 \leq x_0  \right \} 
+\; =   -\mathcal{K}^*.
+\end{equation}
+\end{definition}
+
+\begin{definition}
+The \textit{dual cone} $\mathcal{K}^*$ of a set $\mathcal{K}$ in a vector space equipped with an inner product, is always convex and is defined as
+\begin{equation}
+\label{eq:dualcone}
+\mathcal{K}^* 
+= \left \{ \vect{y} \in \mathbb{R}^n: \left\langle \vect{y}, \vect{x} \right\rangle \geq 0 \quad \forall \vect{x} \in \mathcal{K}  \right \}.
+\end{equation}
+\end{definition}
+
+\begin{definition}
+The \textit{polar cone} is defined as
+\begin{equation}
+\label{eq:polarcone}
+\mathcal{K}^\circ 
+= \left \{ \vect{y} \in \mathbb{R}^n: \left\langle \vect{y}, \vect{x} \right\rangle \leq 0 \quad \forall \vect{x} \in \mathcal{K}  \right \} 
+\; =   -\mathcal{K}^*.
+\end{equation}
+\end{definition}
+
+\begin{definition}
+The \textit{normal cone} to a closed convex set $\mathcal{K}$ at the point $\vect{x}\in\mathcal{K}$ is
+closed and convex and is defined as
+\begin{align}
+\label{eq:normalcone}
+\mathcal{N}_{\mathcal{K}}(\vect{x}) 
+&= \{ \vect{y} \in \mathbb{R}^n : \left\langle \vect{y}, \vect{x} - \vect{z} \right\rangle \geq 0, \forall \vect{z} \in \mathcal{K} \} 
+\end{align}
+\end{definition}
+
+\begin{definition}
+The \textit{indicator function} of a subset $\mathcal{A} \in \mathcal{E}$ is a scalar function $I: \mathcal{E} \mapsto \mathbb{R}$
+defined as:
+\begin{equation}
+\label{eq:indicator}
+{I}_{\mathcal{A}}(\vect{x}) = 
+\left\{ 
+\begin{array}[pos]{l}
+0     \;  \text{if} \; \vect{x} \in \mathcal{A} \\
+\infty   \; \text{if} \; \vect{x} \notin \mathcal{A}
+\end{array}
+\right.
+\end{equation}
+\end{definition}
+
+\begin{remark*}
+The indicator function of any closed set is lower semicontinuous.
+\end{remark*}
+
+\begin{definition}
+The \textit{subdifferential} $\partial f(\vect{x}_0)$ at $\vect{x}_0$ of a convex, possibly non differentiable scalar function $f: \mathbb{R}^n \mapsto \mathbb{R}$, is the closed convex set of all subgradients $\vect{g}$ at $\vect{x}_0$:
+\begin{equation}
+\label{eq:subdifferential}
+\partial f(\vect{x}_0) = \left\{ \vect{g} : 
+  f(\vect{x}) -  f(\vect{x}_0) \geq \left\langle \vect{g}, (\vect{x} - \vect{x}_0) \right\rangle \; \forall \vect{x} \in \mathcal{E} \right\}.
+\end{equation}
+\end{definition}
+
+\begin{remark*}
+The indicator function of any closed set is lower semicontinuous.
+\end{remark*}
+
+\begin{remark*}
+The subdifferential is a set valued function $\partial f(\vect{x}): \mathbb{R}^n \rightrightarrows \mathbb{R}^n$ in general, but as a special case, if $f(\vect{x})$ is differentiable, $\partial f(\vect{x}) = \left\{ \nabla f(\vect{x}) \right\}$.
+\end{remark*}
+
+\begin{remark*}
+The subdifferential of an indicator function of a convex set is also its normal cone:
+%
+\begin{equation}
+\label{eq:sub_ind}
+\partial {I}_{\mathcal{K}}(\vect{x}) = \mathcal{N}_{\mathcal{K}}(\vect{x}).
+\end{equation}
+\end{remark*}
+
+\begin{definition}
+The \textit{projection} of $\vect{y}$ on a nonempty closed convex set $\mathcal{E}$ is a $\mathbb{R}^n \rightarrow \mathbb{R}^n$ mapping defined as:
+%
+\begin{align}
+\label{eq:projector}
+\Pi_{\mathcal{E}}(\vect{y}) = \mathrm{argmin}_{\vect{x} \in \mathcal{E}} \norm{ \vect{x} - \vect{y} }^2_2
+\end{align}
+%
+If $f(\vect{y})$ is the indicator function $I_{\mathcal{E}}(\vect{y})$ to a nonempty closed convex set ${\mathcal{E}}$, defined as in \eqref{eq:indicator}, then $\mathrm{prox}_{f}(\vect{y})$ is the orthogonal projection onto that set:
+%
+\begin{align}
+\label{eq:projectorindicator}
+\mathrm{prox}_{I_{\mathcal{E}}}(\vect{y})  = \Pi_{\mathcal{E}}(\vect{y}) 
+\end{align}
+\label{def:projection}
+\end{definition}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 	\begin{empheq}[left=\empheqlbrace]{align}
-    M \vect{v} - \vect{k} - D_\epsilon \vect{\gamma}_\epsilon &= 0 \\
-    D_{\epsilon}^T \vect{v}  + \vect{b}_\epsilon &= \vect{u}_\epsilon\\
-    -\Upsilon^{\circ} \ni \vect{u}_\epsilon  \quad \bot &\quad  \vect{\gamma}_\epsilon \in \Upsilon
+    H \vect{v} - \vect{k} - D_\epsilon \vect{\gamma}_\epsilon &= 0   \label{eq:ChronoMCCP_a} \\
+    D_{\epsilon}^T \vect{v}  + \vect{b}_\epsilon &= \vect{u}_\epsilon   \label{eq:ChronoMCCP_b} \\
+    -\Upsilon^{\circ} \ni \vect{u}_\epsilon  \quad \bot &\quad  \vect{\gamma}_\epsilon \in \Upsilon  \label{eq:ChronoMCCP_c}
 	\end{empheq}
@@
 \begin{itemize}
-		\item unknown $\vect{v}$ is the speed at the end of the time step, 
-		\item unknown $\vect{\gamma}_\epsilon$ is the reaction in contacts and bilateral joints, a vector-signed measure with Lebesgue decomposition in atomic parts (impacts) and continuous parts (continuous reactions),
-    \item $M$ is a positive definite, block diagonal matrix containing masses and inertia tensors of the bodies
+		\item unknown $\vect{v}$ is the speed at the end of the time step $h$
+		\item unknown $\vect{\gamma}_\epsilon$ is the reaction in contacts and bilateral joints, a vector-signed measure with Lebesgue decomposition in atomic parts (impacts) and continuous parts (continuous reactions)
+    \item $H$ is a positive definite matrix containing $M$, the block-diagonal matrix of the masses and inertia tensors of the bodies; if stiff
+		loads are added too (ex. when finite elements come into play) it becomes a block-sparse matrix including the tangent stiffness and damping matrices,
+		for instance $H=M-{h^2}\nabla_q\vect{f}-{h}\nabla\vect{f}$ in case of the time stepper in \cite{TasoraAnitescuCMAME10}
     \item $D_\epsilon$ is a sparse matrix, the transpose jacobian of all constraints
-    \item $\vect{k}$ is a vector containing terms proportional to applied forces, among other things 
+    \item $\vect{k}$ is a vector containing terms proportional to applied forces $\vect{f}$
     \item $\vect{b}$ is a vector containing constraint stabilization terms, among other things
@@
 Introducing the Schur complement 
-\[
-N=D_{\epsilon}^T M^{-1} D_{\epsilon}
-\]
+\begin{align}
+N=D_{\epsilon}^T H^{-1} D_{\epsilon}
+\label{eq:schur_n}
+\end{align}
 and the vector
-\[
-\vect{r}_\epsilon = D_{\epsilon}^T M^{-1} \vect{k} + \vect{b}_\epsilon
-\]
+\begin{align}
+\vect{r}_\epsilon = D_{\epsilon}^T H^{-1} \vect{k} + \vect{b}_\epsilon
+\label{eq:schur_r}
+\end{align}
 such that
@@
 %
-and in fact, the CCP \eqref{eq:ChronoCCP} can be written in the more conventional language of the KKT optimality conditions on \textit{dual variables} $\vect{y}$ multipliers, for $\vect{y}=-\vect{u}_\epsilon$ and \textit{primal variables} $\vect{\gamma}_\epsilon$:
+and in fact, the CCP \eqref{eq:ChronoCCP} can be written in the more conventional language of the KKT optimality conditions on \textit{dual variables} $\vect{y}$ multipliers, for $\vect{y}=-\vect{u}_\epsilon$, and \textit{primal variables} $\vect{\gamma}_\epsilon$:
 \begin{subequations}
@@
 \end{subequations}
-After the convex program \eqref{eq:ChronoCCP_min} is solved by ADMM, one can compute $\vect{v} = M^{-1}( \vect{k} + \bar{D}_\epsilon \vect{\gamma}_\epsilon)$ with a quick postprocessing step.
+After the convex program \eqref{eq:ChronoCCP_min} is solved by ADMM, one can compute 
+\begin{align}
+\vect{v} = H^{-1}( \vect{k} + \bar{D}_\epsilon \vect{\gamma}_\epsilon)
+\label{eq:v_post}
+\end{align}
+with an immediate step.
 
@@
 Also, in sake of highest generality, in presence of finite elements one might need to use tangent stiffness matrices $K$ and damping matrices $D$ to accommodate an implicit integration scheme for stiff elements, and this means that the Schur complement would be computed as $N=D_{\epsilon}^T H^{-1} D_{\epsilon}$.
-Here $H$, taking the place of the original $M$ matrix, is rather a linear combination of $M$, $K$, $D$ (depending on the integration scheme) and in general is not diagonal anymore.
-
-% TO ADD IN EXTENDED VERSION
-%This means that one cannot easily compute the Schur complement $N$, even in factored form, and this has some consequences on the implementation of the ADMM, that in fact we will rewrite in an unrolled form.
+Here $H$ is a linear combination of $M$, $K$, $D$ (depending on the integration scheme) and in general is not diagonal anymore.
 
@@
 \end{subequations}
-by introducing an augmented Lagrangian
+where $f : \mathbb{R}^{n_x}\rightarrow\overline{\mathbb{R}}$ and  $g : \mathbb{R}^{n_z}\rightarrow\overline{\mathbb{R}}$ are proper convex lower-semicontinuous functions,
+and $A \in \mathbb{R}^{n_x \times n_y}$, $B \in \mathbb{R}^{n_z \times n_y}$ are linear operators.
+
+The optimality conditions require that primal feasibility
 \begin{align}
-\mathcal{L}_{\rho}(\vect{x},\vect{z},\vect{y})= f(\vect{x}) + g(\vect{z}) + \vect{y}^T (A \vect{x} + B \vect{z} - \vect{b}) + \frac{\rho}{2} \norm{A \vect{x} + B \vect{z} - \vect{b}}
+A \vect{x}^\star + B \vect{z}^\star - \vect{b} = \vect{0} \label{eq:primalfeasibility1}
 \end{align}
-and by iterating over two minimization problems as in the following loop:
+and dual feasibility
+\begin{subequations}
 	\begin{align}
- \vect{x}^{k+1} &\in \text{argmin} \mathcal{L}_{\rho}(\vect{x},\vect{z}^k,\vect{y}^k) \label{eq:lag1}\\
- \vect{z}^{k+1} &\in \text{argmin} \mathcal{L}_{\rho}(\vect{x}^{k+1},\vect{z},\vect{y}^k) \label{eq:lag2}\\
+     0 \in \partial	f(\vect{x}^\star) + A^T \vect{y}^\star \label{eq:dualfeasibility1}\\
+	   0 \in \partial	g(\vect{z}^\star) + B^T \vect{y}^\star \label{eq:dualfeasibility2}
+	\end{align}
+	\label{eq:admm}
+\end{subequations}
+must hold at the solution $(\vect{x}^\star,\vect{z}^\star,\vect{y}^\star)$, saddle point of the Lagrangian with dual variables $\vect{y} \in \mathbb{R}^{n_y}$.
+
+By introducing an augmented Lagrangian
+\begin{align}
+\mathcal{L}_{\rho}(\vect{x},\vect{z},\vect{y})= f(\vect{x}) + g(\vect{z}) + \vect{y}^T (A \vect{x} + B \vect{z} - \vect{b}) + \frac{\rho}{2} \norm{A \vect{x} + B \vect{z} - \vect{b}}_2^2
+\end{align}
+the ADMM method converges to the solution $(\vect{x}^\star,\vect{z}^\star,\vect{y}^\star)$ by iterating over two distinct minimization problems as in the following loop:
+\begin{align}
+ \vect{x}^{k+1} &\in \argmin_{\vect{x}} \mathcal{L}_{\rho}(\vect{x},\vect{z}^k,\vect{y}^k) \label{eq:lag1}\\
+ \vect{z}^{k+1} &\in \argmin_{\vect{z}} \mathcal{L}_{\rho}(\vect{x}^{k+1},\vect{z},\vect{y}^k) \label{eq:lag2}\\
  \vect{y}^{k+1} &= \vect{y}^{k} + \rho ( A \vect{x}^{k+1} + B \vect{z}^{k+1} - \vect{b} )
@@
 \end{align}
-Although the convergence is fast in the first iterations, it tends to deteriorate later as in fixed-point iterations; however in practical scenarios where loose tolerances can be accepted, the method proves to be very efficient and robust.
+The convergence of such method has rate $\mathcal{O}\left(\frac{1}{k}\right)$.
+Although the convergence is fast in the first iterations, it tends to deteriorate later as in most fixed-point iterations; however in practical scenarios where loose tolerances can be accepted, even this basic formulation of ADMM proves to be efficient and robust.
 
@@
 
-In order to rewrite \eqref{eq:ChronoCCP_min} as a sum of two functions as in \eqref{eq:admm}, we introduce the auxiliary variables 
-$(\tilde{\vect{\gamma}}_\epsilon,\tilde{\vect{z}})$, 
-we state the $\vect{\gamma}_\epsilon \in \Upsilon$ constraint by adding a non-smooth penalty function given by the indicator function 
-$\mathcal{I}_\Upsilon (\vect{\gamma}_\epsilon)$ 
-where $\mathcal{I}_\Upsilon(\vect{\gamma}_\epsilon)=0$ for $\vect{\gamma}_\epsilon \in \Upsilon$ and $=\infty$ otherwise, another indicator function 
-$\mathcal{I}_{\vect{\gamma}_\epsilon = \vect{z}}(\vect{\gamma}_\epsilon,\vect{z})$,
+In order to rewrite \eqref{eq:ChronoCCP_min} as a sum of two functions as in \eqref{eq:admm}, we introduce 
+$\vect{\gamma}_\epsilon \in \mathbb{R}^{n_c}$ as primal variables, 
+$\vect{z} \in \mathbb{R}^{n_c}$ as auxiliary variables, 
+we add $\vect{\gamma}_\epsilon - \vect{z} = \vect{0}$ as the linear constraint, 
+we reformulate the $\vect{\gamma}_\epsilon \in \Upsilon$ constraint by adding a non-smooth penalty function given by the indicator function 
+$\mathcal{I}_\Upsilon (\vect{z})$,
 and finally we have
@@
 	\begin{align}
-    \text{min} \quad & \frac{1}{2} \tilde{\vect{\gamma}}_\epsilon^T N \tilde{\vect{\gamma}}_\epsilon + \vect{r}^T_\epsilon \tilde{\vect{\gamma}}_\epsilon  +  \mathcal{I}_{\tilde{\vect{\gamma}}_\epsilon = \tilde{\vect{z}}}(\tilde{\vect{\gamma}}_\epsilon,\tilde{\vect{z}}) +  \mathcal{I}_\Upsilon(\vect{\gamma}_\epsilon)  \\
-	  \text{s.t.} \quad & (\tilde{\vect{\gamma}}_\epsilon,\tilde{\vect{z}})  =  (\vect{\gamma}_\epsilon,\vect{z})
+    \text{min} \quad & \frac{1}{2} \vect{\gamma}_\epsilon^T N \vect{\gamma}_\epsilon + \vect{r}^T_\epsilon \vect{\gamma}_\epsilon  
+		+  \mathcal{I}_\Upsilon(\vect{z})  \\
+	  \text{s.t.} \quad & \vect{\gamma}_\epsilon =  \vect{z}
 	\end{align}
@@
 
-We can write the augmented Lagrangian introducing two step size parameters $\sigma$ and $\rho$ and two vector of multipliers $\vect{w}$ and $\vect{y}$, obtaining:
+We can write the augmented Lagrangian introducing a step size parameter $\rho$ and a vector of dual variables $\vect{y}$, obtaining:
 \begin{align}
-\mathcal{L}_{\sigma,\rho} \left((\tilde{\vect{\gamma}}_\epsilon,\tilde{\vect{z}}),(\vect{\gamma}_\epsilon,\vect{z}),(\vect{y},\vect{w})\right) &= 
-\frac{1}{2} \tilde{\vect{\gamma}}_\epsilon^T N \tilde{\vect{\gamma}}_\epsilon + \vect{r}^T_\epsilon \tilde{\vect{\gamma}}_\epsilon  
-+  \mathcal{I}_{\tilde{\vect{\gamma}}_\epsilon = \tilde{\vect{z}}}(\tilde{\vect{\gamma}}_\epsilon,\tilde{\vect{z}}) 
-+  \mathcal{I}_\Upsilon(\vect{\gamma}_\epsilon) \nonumber \\
-+ \vect{w}^T (\tilde{\vect{\gamma}}_\epsilon - \vect{\gamma}_\epsilon) 
-&+ \frac{\sigma}{2} \norm{\tilde{\vect{\gamma}}_\epsilon - \vect{\gamma}_\epsilon} 
-+ \vect{y}^T (\tilde{\vect{z}} - \vect{z}) 
-+ \frac{\rho}{2} \norm{\tilde{\vect{z}} - \vect{z}}
+\mathcal{L}_{\rho} \left(\vect{\gamma}_\epsilon,\vect{z},\vect{y} \right) &= 
+\frac{1}{2} \vect{\gamma}_\epsilon^T N \vect{\gamma}_\epsilon + \vect{r}^T_\epsilon \vect{\gamma}_\epsilon  
++  \mathcal{I}_\Upsilon(\vect{z}) \nonumber \\
+&+ \vect{y}^T (\vect{\gamma}_\epsilon - \vect{z}) 
++ \frac{\rho}{2} \norm{\vect{\gamma}_\epsilon - \vect{z}}_2^2
 \end{align}
-that is also, using the property $\vect{y}^T\vect{r} + \frac{\rho}{2}\norm{\vect{r}} = \frac{\rho}{2}\norm{\vect{r}+\frac{1}{\rho} \vect{y}} - \frac{1}{2\rho} \norm{\vect{y}}$, the following
+that is also, using the property $\vect{y}^T\vect{r} + \frac{\rho}{2}\norm{\vect{r}}_2^2 = \frac{\rho}{2}\norm{\vect{r}+\frac{1}{\rho} \vect{y}}_2^2 - \frac{1}{2\rho} \norm{\vect{y}}_2^2$, the following
 \begin{align}
-\mathcal{L}_{\sigma,\rho} \left((\tilde{\vect{\gamma}}_\epsilon,\tilde{\vect{z}}),(\vect{\gamma}_\epsilon,\vect{z}),(\vect{y},\vect{w})\right) &= 
-\frac{1}{2} \tilde{\vect{\gamma}}_\epsilon^T N \tilde{\vect{\gamma}}_\epsilon + \vect{r}^T_\epsilon \tilde{\vect{\gamma}}_\epsilon  
-+  \mathcal{I}_{\tilde{\vect{\gamma}}_\epsilon = \tilde{\vect{z}}}(\tilde{\vect{\gamma}}_\epsilon,\tilde{\vect{z}}) 
-+  \mathcal{I}_\Upsilon(\vect{\gamma}_\epsilon) \nonumber \\
-+ \frac{\sigma}{2} \norm{\tilde{\vect{\gamma}}_\epsilon - \vect{\gamma}_\epsilon + \frac{1}{\sigma} \vect{w}} 
-&- \frac{1}{2\sigma} \norm{\vect{w}}
-+ \frac{\rho}{2} \norm{\tilde{\vect{z}} - \vect{z} + \frac{1}{\rho} \vect{y}}
-- \frac{1}{2\rho} \norm{\vect{y}}
+\mathcal{L}_{\rho} \left(\vect{\gamma}_\epsilon,\vect{z},\vect{y} \right) &= 
+\frac{1}{2} \vect{\gamma}_\epsilon^T N \vect{\gamma}_\epsilon + \vect{r}^T_\epsilon \vect{\gamma}_\epsilon  
++  \mathcal{I}_\Upsilon(\vect{z}) \nonumber \\
+&+ \frac{\rho}{2} \norm{\vect{\gamma}_\epsilon - \vect{z} + \frac{1}{\rho} \vect{y}}_2^2
+- \frac{1}{2\rho} \norm{\vect{y}}_2^2
+\label{eq:augmentedlagrangian}
 \end{align}
@@
 The first step of ADMM requires to compute 
-\[
-(\tilde{\vect{\gamma}}_\epsilon^{k+1},\tilde{\vect{z}}^{k+1}) = \text{argmin} \mathcal{L}_{\sigma,\rho} \left((\tilde{\vect{\gamma}}_\epsilon,\tilde{\vect{z}}),(\vect{\gamma}_\epsilon^k,\vect{z}^k),(\vect{y}^k,\vect{w}^k)\right)
-\] 
-This can be simplified remembering that $\vect{w}^k$ will always be zero, that some terms are constant and can be left out, and that the $\mathcal{I}_{\tilde{\vect{\gamma}}_\epsilon = \tilde{\vect{z}}}(\tilde{\vect{\gamma}}_\epsilon,\tilde{\vect{z}})$ term can be elided to turn finally the unconstrained $\text{argmin}\mathcal{L}_{\sigma,\rho}$ into the equality constrained optimization problem:
-\begin{subequations}
 \begin{align}
-    \text{min} \quad  &  \frac{1}{2} \tilde{\vect{\gamma}}_\epsilon^T N \tilde{\vect{\gamma}}_\epsilon + \vect{r}^T_\epsilon \tilde{\vect{\gamma}}_\epsilon  
-+ \frac{\sigma}{2} \norm{\tilde{\vect{\gamma}}_\epsilon - \vect{\gamma}_\epsilon^k}
-+ \frac{\rho}{2} \norm{\tilde{\vect{z}} - \vect{z}^k + \frac{1}{\rho} \vect{y}^k} \\
-	  \text{s.t.} \quad &  \tilde{\vect{\gamma}} = \tilde{\vect{z}}
+\vect{\gamma}_\epsilon^{k+1} &= \argmin_{\vect{\gamma}_\epsilon} \mathcal{L}_{\rho} \left( \vect{\gamma}_\epsilon,\vect{z}^k,\vect{y}^k \right) \\
+ &= \argmin_{\vect{\gamma}_\epsilon} \frac{1}{2} \vect{\gamma}_\epsilon^T N \vect{\gamma}_\epsilon + \vect{r}^T_\epsilon \vect{\gamma}_\epsilon  
++ \frac{\rho}{2} (\vect{\gamma}_\epsilon - \vect{z}^k + \frac{1}{\rho} \vect{y}^k)^T(\vect{\gamma}_\epsilon - \vect{z}^k + \frac{1}{\rho} \vect{y}^k)
+- \frac{1}{2\rho} \norm{\vect{y}^k}_2^2 \\
+ &= \argmin_{\vect{\gamma}_\epsilon} \frac{1}{2} \vect{\gamma}_\epsilon^T (N+ \rho I) \vect{\gamma}_\epsilon 
+   + (\vect{r}_\epsilon - \rho \vect{z}^k + \vect{y}^k)^T \vect{\gamma}_\epsilon + C^k 
 \end{align}
-	\label{eq:admm_1a}
-\end{subequations}
-%
-For such problem, the KKT conditions lead to the following linear problem:
+
+This is an unconstrained convex quadratic program whose KKT optimality conditions lead to the following linear problem:
 \begin{align}
-    \left[ N + (\sigma+\rho) I \right] \tilde{\vect{\gamma}}_\epsilon = \sigma \vect{\gamma}_\epsilon^k + \rho \vect{z}^k - \vect{r} - \vect{y}^k
+    \left[ N + \rho I \right] \vect{\gamma}_\epsilon^{k+1} =  \rho \vect{z}^k - \vect{r} - \vect{y}^k
 		\label{eq:admm_1b}
 \end{align}
-%
+
+%Optionally, following the idea of \cite{Stellato2020}, one could also introduce a small regularization parameter $\sigma$, that helps in case $N$ is near singular and $\rho$ is very small:
+%\begin{align}
+    %\left[ N + (\sigma+\rho) I \right] \tilde{\vect{\gamma}}_\epsilon = \sigma \vect{\gamma}_\epsilon^k + \rho \vect{z}^k - \vect{r} - \vect{y}^k
+		%\label{eq:admm_1breg}
+%\end{align}
+
 The linear system \eqref{eq:admm_1b} can be solved as it is, but in our case it would be better to exploit the fact that the Schur matrix $N$ is a product $D_\epsilon^T H^{-1} D_\epsilon$.
 We would like to avoid computing $H^{-1}$, even storing a precomputed inverse for all iterations would be unpractical because often dense (except when $H$ is a diagonal mass matrix $M$).
-So we propose to replace \eqref{eq:admm_1b} with the equivalent saddle point problem:
+So we propose to replace \eqref{eq:admm_1b} with an equivalent saddle point problem:
+
+\begin{theorem}
+\label{th:schurtokkt}
+Let $\vect{\gamma}_\epsilon^{k+1}$ be a solution to \eqref{eq:admm_1b}, then it is also a solution to
+%
 \begin{subequations}
@@
 		 H   & D_\epsilon \\
-		 D^T & -(\sigma+\rho) I
+		 D_\epsilon^T & - \rho I
 		\end{bmatrix}
 		\begin{Bmatrix}
-		 \vect{v}    \\
-		 -\tilde{\vect{\gamma}}_\epsilon^{k+1} 
+		 \vect{v}^{k+1}   \\
+		 -\vect{\gamma}_\epsilon^{k+1} 
 		\end{Bmatrix}
@@
 		 \vect{k} \\
-		 -\vect{b}_\epsilon + \sigma \vect{\gamma}_\epsilon^k + \rho \vect{z}^k -\vect{y}^k 
+		 -\vect{b}_\epsilon + \rho \vect{z}^k -\vect{y}^k 
 		\end{Bmatrix}
@@
 \end{subequations}
-that, even if it introduces an auxiliary variable $\vect{v}$, is very sparse and does not require computing any $H^{-1}$.
+%
+\end{theorem}
 
-A remarkable side effect of this approach is that its auxiliary variable $\vect{v}$ is also the velocity term in the original MDI 
-\eqref{eq:ChronoMCCP}, so one does not need to compute it as $\vect{v}=H^{-1}(\vect{k} + D_\epsilon \vect{\gamma}_\epsilon )$ after the iteration converged because it is a byproduct of the linear solver.
+\begin{proof}
+Performing the multiplication of the upper part of the saddle point matrix one has 
+$H \vect{v}^{k+1} - D_\epsilon \vect{\gamma}_\epsilon^{k+1} = \vect{k}$,
+pre-multiplying by $H^{-1}$ one gets 
+$\vect{v}^{k+1} = H^{-1} D_\epsilon \vect{\gamma}_\epsilon^{k+1} + H^{-1} \vect{k}$. 
+The multiplication of the lower part gives 
+$D_\epsilon^T \vect{v}^{k+1} + \rho I \vect{\gamma}_\epsilon^{k+1} = -\vect{b}_\epsilon + \rho \vect{z}^k -\vect{y}^k $, 
+hence after substitution of $\vect{v}^{k+1}$ one gets
+$D_\epsilon^T H^{-1} D_\epsilon \vect{\gamma}_\epsilon^{k+1}  + D_\epsilon^T H^{-1} \vect{k} + \rho I \vect{\gamma}_\epsilon^{k+1} = -\vect{b}_\epsilon + \rho \vect{z}^k -\vect{y}^k$. 
+Recalling the definition of $N$ in \eqref{eq:schur_n} and the definition of $\vect{r}_\epsilon$ in \eqref{eq:schur_r}, this can be written also as 
+$(N + \rho I) \vect{\gamma}_\epsilon^{k+1} =  \rho \vect{z}^k - \vect{r} - \vect{y}^k$. 
+\end{proof}
 
-The second step of ADMM is a minimization problem too, but it can be rephrased in terms of a projection on the separable set $\Upsilon$ of the friction cones, note that this requires an inexpensive and parallelizable operation.
+\begin{corollary}
+The linear system \eqref{eq:admm_1c} introduces an auxiliary variable $\vect{v} \in \mathbb{R}^{n_v}$, however the matrix is very sparse and does not require storing or computing any $H^{-1}$ matrix as we should do if using \eqref{eq:admm_1b}.
+\end{corollary}
 
-Using these results, and adding a Krasnoselskii-Mann relaxation factor $\alpha \in(0,2]$, one obtains the final algorithm:
+\begin{corollary}
+A positive side effect of this approach is that its auxiliary variable $\vect{v}$ is also the velocity term in the original mixed CCP 
+\eqref{eq:ChronoMCCP}, so one does not need to compute it as $\vect{v}=H^{-1}(\vect{k} + D_\epsilon \vect{\gamma}_\epsilon )$ after the iteration converged because it is already a byproduct of the linear solver in \eqref{eq:admm_1c}.
+\end{corollary}
+
+\begin{corollary}
+If constraint compliance is added to the MDI, a compliance matrix $C_\epsilon$ can be present in the CCP, thus leading to a modified version of \eqref{eq:ChronoMCCP_b}:
+\[
+D_{\epsilon}^T \vect{v}  +C_\epsilon \vect{\gamma}_\epsilon + \vect{b}_\epsilon = \vect{u}_\epsilon
+\]
+In such case one would have $N=D_{\epsilon}^T H^{-1} D_{\epsilon} + C_\epsilon$, and \eqref{eq:admm_1c} would change into: 
+\begin{subequations}
+	\begin{align}
+    \begin{bmatrix}
+		 H   & D_\epsilon \\
+		 D_\epsilon^T & - \rho I - C
+		\end{bmatrix}
+		\begin{Bmatrix}
+		 \vect{v}^{k+1}   \\
+		 -\vect{\gamma}_\epsilon^{k+1} 
+		\end{Bmatrix}
+		=
+		\begin{Bmatrix}
+		 \vect{k} \\
+		 -\vect{b}_\epsilon + \rho \vect{z}^k -\vect{y}^k 
+		\end{Bmatrix}
+	\end{align}
+	\label{eq:admm_1c_compl}
+\end{subequations}
+\end{corollary}
+
+The second step of ADMM requires to compute 
+\[
+\vect{z}^{k+1} = \argmin_{\vect{z}} 
+\mathcal{L}_{\rho} \left(\vect{\gamma}_\epsilon^{k+1},\vect{z},\vect{y}^k\right)
+\] 
+%
+This is a minimization problem too, but it can be rephrased in terms of a projection on the $\Upsilon$ set, which has a separable structure.
+In fact, recalling \eqref{eq:augmentedlagrangian} and removing constant terms: 
+\begin{align}
+\vect{z}^{k+1} &= \argmin_{\vect{z}}  
+\mathcal{I}_\Upsilon(\vect{z}) 
++ \frac{\rho}{2} \norm{ \vect{\gamma}_\epsilon^{k+1} - \vect{z} + \frac{1}{\rho} \vect{y}^k}_2^2
+\end{align}
+%
+Using the definition \ref{def:projection}, this leads to:
+\begin{align}
+\vect{z}^{k+1} &= \Pi_\Upsilon \left(  \vect{\gamma}_\epsilon^{k+1} +\frac{1}{\rho} \vect{y}^k \right )
+\end{align}
+
+Note that the projection $\Pi_\Upsilon$ can be performed efficiently in our case, because $\Upsilon$ is the Cartesian product of $n_c$ Coulomb second-order friction cones of lower dimension, assuming a set of $n_c$ contacts:
+\[
+\Upsilon = \Upsilon_1 \times \Upsilon_2 \times \ldots \times \Upsilon_{n_c}, \quad \Upsilon_i \subset \mathbb{R}^3
+\]
+The separable structure allows the projection to be performed as a tuple of simpler projections:
+\[
+\Pi_\Upsilon(\vect{p}) = \left\{ \Pi_{\Upsilon_1}(\vect{p}_1), \Pi_{\Upsilon_2}(\vect{p}_2), \ldots,  \Pi_{\Upsilon_{n_c}}(\vect{p}_{n_c}) \right\}, \quad \vect{p}_i \in \mathrm{R}^3
+\]
+%
+Note that this requires an inexpensive and parallelizable operation.
+
+A basic form of ADMM will then iterate over these updates:
+\begin{align}
+    \begin{bmatrix}
+		 H   & D_\epsilon \\
+		 D_\epsilon^T & - \rho I - C
+		\end{bmatrix}
+		\begin{Bmatrix}
+		 \vect{v}^{k+1}   \\
+		 -\vect{\gamma}_\epsilon^{k+1} 
+		\end{Bmatrix}
+		&=
+		\begin{Bmatrix}
+		 \vect{k} \\
+		 -\vect{b}_\epsilon + \rho \vect{z}^k -\vect{y}^k 
+		\end{Bmatrix} \label{eq:admm_step1} \\
+		\vect{z}^{k+1} &= \Pi_\Upsilon \left(  \vect{\gamma}_\epsilon^{k+1} +\frac{1}{\rho} \vect{y}^k \right ) \label{eq:admm_step2} \\
+		\vect{y}^{k+1} &= \vect{y}^k + \rho \left( \vect{\gamma}_\epsilon^{k+1} - \vect{z}^{k+1} \right) \label{eq:admm_step3}
+\end{align}
+%
+Before being able to apply this iteration in real problems, we should introduce some optimizations that will make a big difference for the performance of the method, and that will be discussed in the next section.
+
+\section{Implementation}
+
+\subsection{Residuals and termination}
+
+In order to monitor the convergence of the method, a measure of residual quantities must be introduced. 
+Given the definitions \eqref{eq:primalfeasibility1},\eqref{eq:primalfeasibility1},\eqref{eq:primalfeasibility1}, one can define a primal and a dual residual as:
+%
+\begin{align}
+ \vect{r}_{\text{primal}}^{k+1} &= \vect{\gamma}_\epsilon^{k+1}-\vect{z}^{k+1}   \\
+ \vect{r}_{\text{dual}}^{k+1} &= N \vect{\gamma}_\epsilon^{k+1}+\vect{r}+\vect{y}^{k+1} 
+\end{align}
+
+An alternative way to compute $\vect{r}_{\text{dual}}^{k+1}$, that does not need to use the $N$ matrix, is 
+\[
+   \vect{r}_{\text{dual}}^{k+1} = D_\epsilon^T \vect{v}^{k+1} +C \vect{\gamma}_\epsilon^{k+1} + \vect{r} + \vect{y}^{k+1}
+\]
+Another option (see \cite{Boyd2011}) is to compute it as
+\[
+   \vect{r}_{\text{dual}}^{k+1} = \rho (\vect{z}^{k+1} - \vect{z}^{k})
+\]
+which is even faster, at the expense of storing the previous value of $\vect{z}$.
+
+The iteration is terminated when both $\vect{r}_{\text{primal}}$ and $\vect{r}_{\text{dual}}$ fall under prescribed tolerances.
+
+Here primal and dual residuals have an interesting physical interpretation: a large primal residual means that the solution provides reaction forces that do not satisfy the set inclusions (e.g.
+the friction might be overestimated), whereas a large dual residual means that the speed (in the contact point metric) is wrong.
+That is, in the discussed time stepping scheme, the measuring units of $\vect{r}_{\text{primal}}$ is the same of reaction impulses, e.g.
+\si{N.s}, whereas the measuring units of $\vect{r}_{\text{primal}}$ can be considered the same of speeds, e.g.
+\si{m/s}.
+
+We remark that, depending on the measuring units, the two residual errors can have different importance: this means that it is useful to set two distinct tolerances $\epsilon_{\text{primal}}$ and $\epsilon_{\text{dual}}$.
+Different heuristics can be used to define relative tolerances too, as shown in \cite{Stellato2020}.
+
+\subsection{Step size selection}
+
+The basic method outlined in \eqref{eq:admm_step1}-\eqref{eq:admm_step3} uses a fixed step $\rho$.
+
+In general, small values of $\rho$ cause a fast reduction of the dual residual, but a slow reduction of the primal residual.
+This has a  physical interpretation: truncating iterations based on too small $\rho$ most often lead to solutions where parts do not interpenetrate, but some contact forces do not satisfy the inclusion in the Coulomb cone.
+That is, some parts might stick like contacts were glued.
+
+On the other hand, large values of $\rho$ cause a fast reduction of the primal residual, but a slow reduction of the dual residual.
+This means that truncating iterations with too large $\rho$ most often lead to solutions where contact forces satisfy very well the inclusion in the Coulomb friction cones, but parts might interpenetrate.
+
+Clearly, this calls for the introduction of a scheme that adaptively changes the step size $\rho$ in order to achieve a balanced reduction of both residuals.
+In most cases one can use heuristics and adjust $\rho$ only one or two times during the iterations, hence allowing to reuse the factorization of the linear system \eqref{eq:admm_step1} as much as possible.
+In fact, the most time-consuming step is the factorization of the matrix in \eqref{eq:admm_step1}, but this happens only at the first iteration and at all iterations where $\rho$ changes, otherwise one can reuse the last factorization because the matrix is the same, and perform only the back solve step with the different right hand side.
+On average, in our tests, the back solve operation took less than $1/10$ of the time needed for the factorization.
+
+We tested different policies for scaling the $\rho$ step in our algorithm. 
+All policies share the following concept: starting from an initial value $\rho_0$, the step is scaled once each $n_s$ iterations such that the primal and dual residuals are well balanced 
+\cite{Wohlberg2017}, % [ a step size selection via residual balancing, scale residuals given problem measuring units]:
+that is:
+\begin{align}
+	\rho^{k+1} = \rho^{k} \eta_\rho 
+\end{align}
+
+We define the \textit{balanced} scaling policy as
+\begin{align}
+	\eta_\rho = \frac{ \norm{\vect{r}_{\text{primal}}^{k}} }{ \norm{\vect{r}_{\text{dual}}^{k}} }  
+\end{align}
+We define the \textit{balanced normalized} scaling policy as
+\begin{align}
+	\eta_\rho = \left( \frac{\norm{\vect{r}_{\text{primal}}^{k}}}{ \norm{\vect{z}^k}+\norm{\vect{\gamma}_\epsilon^k} +10^{-11} } \right)
+															\left( \frac{\norm{\vect{r}_{\text{dual}}^{k}}}  { \norm{\vect{y}^k } +10^{-11} }  \right)^{-1}
+\end{align}
+
+Another idea is presented in 
+ \cite{Xu_adaptive_2017}, % [ Spectral stepsize selection as in Barzilai Borwein ]
+where the the step is changed using the Barzilai-Borwein spectral formula. This is referenced as the \textit{spectral} scaling policy in the following.
+
+In all the cases above, one needs safeguards to avoid excessive scaling, for example we limit the scaling factor $\eta_\rho$ by clamping it in the $\left[\frac{1}{50}, 50 \right]$ interval by default.
+Also, in order to avoid an excessive number of factorizations, one could force the scaling factor to $\eta_\rho=1$ (hence no change in $\rho$) if it is not too far from the unity anyway - for example by default we use a tolerance interval $\left[\frac{1}{2}, 2 \right]$, and if the scaling falls inside that interval, it is forced to $\eta_\rho=1$.
+
+\subsection{Custom treatment of bilateral joints}
+
+We experienced that it is possible to use a $\rho$ value that changes on a per-constraint basis.
+If one could know in advance which contact will be active at the solution, one could split $\rho_i$ values between very high or very low values for the fastest convergence, however this is not possible because the set of active constraints is known only at the solution.
+However, we still can use a
+simple heuristics consisting in forcing a small value $\rho_b = 1e-9$ for bilateral constraints, where one knows in advance that the corresponding dual variable $y_i$ would be null anyway.
+This produces better convergence results in systems containing both contacts and many bilateral joints, such as when simulating articulated robots that grasp some objects.
+
+This means that, instead of using a scalar $\rho^{k}$, in our code we use a diagonal matrix $\Theta^{k}$ defined as:
+\begin{align}
+ \Theta_{i,j}^{k} = 
+		\left\{
+			\begin{matrix}
+			0 & \mathrm{if} & i \neq j\\
+			\rho_b & \mathrm{if} & i = j  \;  \mathrm{and} \; i \in \mathcal{C}_B \\
+			\rho^{k} & \mathrm{if} & i = j  \;  \mathrm{and} \; i \notin \mathcal{C}_B \\
+			\end{matrix}
+		\right. 
+\end{align}
+
+\subsection{Preconditioning}
+
+We experienced that it
+
+\section{Final algorithm}
+
+Using the above results, one can finally obtain the ADDM algorithm for solving the CCP of the non-smooth dynamics timestepper:
+
+TO UPDATE!!!!
+
+\begin{algorithm}
+\DontPrintSemicolon
+\KwIn{ Initial approximations $\vect{\gamma}_\epsilon^0$, $\vect{y}^0$ }
+\KwOut{ Solution $\vect{\gamma}_\epsilon$, $\vect{v}$ }
+\SetKwFunction{StepAdjustPolicy}{StepAdjustPolicy}
+\While{$i \leq n_{max iters}$}{
+    $ \begin{bmatrix}
+			 H   & D_\epsilon \\
+			 D^T & - \Theta^k I - C
+			\end{bmatrix}
+			\begin{Bmatrix}
+				\vect{v}^{k+1}  \\
+			 -\vect{\gamma}_\epsilon^{k+1}
+			\end{Bmatrix}
+			=
+			\begin{Bmatrix}
+			 \vect{k} \\
+			 -\vect{b}_\epsilon + \Theta^k \vect{z}^k - \vect{y}^k 
+			\end{Bmatrix} $\;
+	  $\vect{z}^{k+1} = \Pi_\Upsilon \left(  \vect{\gamma}_\epsilon^{k+1} +(\Theta^k)^{-1} \vect{y}^k \right )$ \;
+		$\vect{y}^{k+1} = \vect{y}^k + \Theta^k \left( \vect{\gamma}_\epsilon^{k+1} - \vect{z}^{k+1} \right) $ \;
+		$\vect{r}_{\text{primal}}^{k+1} = \vect{\gamma}_\epsilon^{k+1}-\vect{z}^{k+1}$ \;
+		$\vect{r}_{\text{dual}}^{k+1}   = \Theta^k (\vect{z}^{k+1} - \vect{z}^{k})$ \;
+		\lIf{$\norm{\vect{r}_{\text{primal}}^{k+1}} < \epsilon_{\text{primal}}$ and $\norm{\vect{r}_{\text{dual}}^{k+1}} < \epsilon_{\text{dual}}$ }{
+			break \;
+		}
+		$\rho^{k+1} = $ \StepAdjustPolicy{$\rho^{k}$},  $\rho^{k+1} \rightarrow \Theta^{k+1}$ \;
+}
+%\Return{location}\;
+\caption{Simple form of the ADMM algorithm, with no $N$ Schur complement.}
+\label{algo:admm1}
+\end{algorithm}
+
+TO DO:
+
+TO UPDATE!!!!
+
+ALSO MAYBE THIS ALPHA VARIANT WILL BE REMOVED!!!
+
+** Using these results, and adding a Krasnoselskii-Mann relaxation factor $\alpha \in(0,2]$, one obtains the final algorithm:
 %
@@
 		\vect{r}_{\text{prim}} &= \vect{\gamma}_\epsilon^{k+1}-\vect{z}^{k+1} \\
-		\vect{r}_{\text{dual}} &= N \vect{\gamma}_\epsilon^{k+1}+\vect{r}-\vect{y}^{k+1} 
+		\vect{r}_{\text{dual}} &= N \vect{\gamma}_\epsilon^{k+1}+\vect{r}+\vect{y}^{k+1} 
 	\end{align}
@@
 
-The iteration is terminated when $\vect{r}_{\text{prim}}$ and $\vect{r}_{\text{dual}}$ fall under prescribed tolerances.
-We remark that setting proper values for $\rho$ is of fundamental importance in order to obtain a good speed of convergence.
-In most cases one can use heuristics and adjust $\rho$ only one or two times during the iterations, hence allowing to reuse the factorization of the linear system.
-Also, $\sigma$ acts similarly to a regularization term, and can be left to low values anyway.
+TO DO: CUT THIS "`OPTIMIZATIONS"' SECT, HERE ONLY FOR CUT PASTING
 
@@
 
-TO COMPLETE
+\section{Benchmarks}
 
-\section{Step size selection}
+\subsection{High stack with odd mass ratio}
 
-In general, small values of $\rho$ cause a fast reduction of the dual residual, but a slow reduction of the primal residual.
-This has a nice physical interpretation: truncating iterations based on too small $\rho$ most often lead to solutions where parts do not interpenetrate, but some contact forces do not satisfy the inclusion in the Coulomb cone.
-That is, some parts might stick like contacts were glued.
+This benchmark represents a typical worst case scenario in non-smooth dynamics, where
+a heavy object sits on a high stack of rigid bodies of much lower mass, as shown in Fig.\ref{fig:t6_snapshot}. The stable stacking
+is already a demanding case for contact dynamics, but the presence of extremely odd mass ratios add a
+further complication.
 
-On the other hand, large values of $\rho$ cause a fast reduction of the primal residual, but a slow reduction of the dual residual.
-This means that truncating iterations with too large $\rho$ most often lead to solutions where contact forces satisfy very well the inclusion in the Coulomb friction cones, but parts might interpenetrate.
+To this end we designed two sub-cases: \textsc{Test 1.a} and \textsc{Test 1.b}. Both feature
+a vertical stack of 20 spheres with mass $m=\SI{10}{kg}$ each, except the 10th sphere that has
+a mass of $m=\SI{10000}{kg}$. A further sphere with mass $m=\SI{1000}{kg}$ is placed on the floor.
+All rigid bodies are subject to a vertical gravitational field $g=\SI{9.8}{m/s^2}$. For \textsc{Test 1.a}
+one has the analytical solution (all contacts are active). The \textsc{Test 1.b}
+differs from \textsc{Test 1.a} in that the heavy sphere is pulled upward by a force twice its weight, hence
+causing a separation at the contact below it.
 
-Clearly, this calls for the introduction of a scheme that adaptively changes the step size $\rho$ in order to achieve a balanced reduction of both residuals.
+As expected, in both \textsc{Test 1.a} and \textsc{Test 1.b} the projected Jacobi iteration, one of the most common
+methods in non-smooth dynamics, has a very bad convergence rate \footnote{It is a known fact that the simulation 
+of high stacks of objects is a major difficulty in non-smooth dynamics, especially in real-time simulators
+and in video games where the projected fixed-point iterations are truncated to keep the computational
+time within a limited frame budget. In those cases, the limited precision of the solution will lead to underestimated
+contact forces, undesired slow interpenetration of the parts and, ultimately, to unnatural collapses of stacks.}.
+As shown in Fig.\ref{fig:t6_convergence_b}, 
+the ADMM method presented in this paper behave
+much better than the projected Jacobi iteration and, interesting enough, it also converges faster
+than the Preconditioned Spectral Projected Gradient with Fall-Back (P-SPG-FB), an
+efficient method for non-smooth problems presented in \cite{hammadTOG2015}.
 
-TO DO
+\begin{figure}[!tbp]
+  \centering
+  \begin{minipage}[t]{0.48\textwidth}
+    \includegraphics[width=\textwidth]{t6_snapshot.png}
+    \caption{Setup of \textsc{Test 1}.}
+		\label{fig:t6_snapshot}
+  \end{minipage}
+  \hfill
+  \begin{minipage}[t]{0.48\textwidth}
+    \includegraphics[width=\textwidth]{t6_convergence_b.pdf}
+    \caption{Convergence of ADMM methods compared to the Jacobi projected fixed point iteration, for \textsc{Test 1.b} 
+		The plot shows the violation of unilateral constraints in terms of penetrating residual velocity. }
+		\label{fig:t6_convergence_b}
+  \end{minipage}
+\end{figure}
 
-Step size:
+\begin{figure}[!tbp]
+  \centering
+  \begin{minipage}[t]{0.48\textwidth}
+    \includegraphics[width=\textwidth]{t6_primdual_a.pdf}
+    \caption{Convergence of primal and dual residuals in \textsc{Test 1.a}.}
+		\label{fig:t6_primdual_a}
+  \end{minipage}
+  \hfill
+  \begin{minipage}[t]{0.48\textwidth}
+    \includegraphics[width=\textwidth]{t6_primdual_b.pdf}
+    \caption{Convergence of primal and dual residuals in \textsc{Test 1.b}. }
+		\label{fig:t6_primdual_b}
+  \end{minipage}
+\end{figure}
 
-In \cite{Wohlberg2017} [ a step size selection via residual balancing, scale residuals given problem measuring units]
+For \textsc{Test 1.a}, Fig.\ref{fig:t6_primdual_a} shows that the ADMM method using the \textit{balanced} adaptive step size can converge to the exact solution in less than 20 iterations (the primal residual is not visible in the semilogarithmic plot as it is immediately zero from the first iteration).
 
-In \cite{Xu_adaptive_2017} [ Spectral stepsize selection as in Barzilai Borwein ]
+Note also that we force the update of the step every $n_s$ steps because each update of $\rho$ corresponds of a costly factorization of a large matrix.
+Fig.\ref{fig:t6_primdual_a_ns} shows that, in this benchmark, the convergence improves for denser updates, 
+meeting the tolerance in the same amount of updates: this would suggest using a small $n_s$ anyway. However, for more complex and randomized scenarios, 
+we found that a good tradeoff is $n_s=5$. This tradeoff value can change depending on the speed of solver used for the factorization: the higher the performance
+for the back solve respect to the factorization, the higher is the optimal $n_s$.
 
-\section{Benchmarks}
+The FADMM method with the default fixed step $\rho=0.05$, instead, converges slowly. 
+Repeating the FADMM test with different step sizes as shown in Fig.\ref{fig:t6_primdual_a_rho}, 
+however, the convergence is greatly improved, suggesting that the FADMM method is sensitive to the initial choice of the time step.
+
+\begin{figure}[!tbp]
+  \centering
+  \begin{minipage}[t]{0.48\textwidth}
+    \includegraphics[width=\textwidth]{t6_primdual_a_ns.pdf}
+    \caption{Convergence of primal and dual residuals in \textsc{Test 1.a} for varying frequency $n_s$ of automatic updates to the step size, in the ADMM algorithm.}
+		\label{fig:t6_primdual_a_ns}
+  \end{minipage}
+  \hfill
+	\begin{minipage}[t]{0.48\textwidth}
+    \includegraphics[width=\textwidth]{t6_primdual_a_rho.pdf}
+    \caption{Convergence of primal and dual residuals in \textsc{Test 1.a} for varying values of the step size $\rho$ in the FADMM algorithm.}
+		\label{fig:t6_primdual_a_rho}
+  \end{minipage}
+\end{figure}
+
+For \textsc{Test 1.b}, convergence plots in Fig.\ref{fig:t6_primdual_b} show that the FADMM is very efficient even if starting from 
+a fixed not optimal step size, although the performance deteriorates if no preconditioning is used. The ADMM method with balanced 
+step adaptivity $n_s=5$ converges quite well, especially with the help of a preconditioner.
+
+Showing that a solver is capable of handling this class of object-stacking problems is relevant, for instance,
+to the field of civil engineering, because it shares the same difficulties of simulating tall masonry structures 
+when assessing stability and seismic response via discrete elements \cite{Beatini2017}.
+
+\subsection{Bulldozer on stone slabs}
 

### 2020-10-30 2021/NODYCON_nonsmooth_ADMM/author.tex
correct typo with sign
correct other sign typo
correct wrong sign in formula - it was right in Matlab code though
fix typos
Update paper ADMM

@@
 \begin{align}
-\mathcal{L}_{\rho}(\vect{x},\vect{z},\vect{y})= f(\vect{x}) + g(\vect{z}) + \vect{y}^T (A \vect{x} + B \vect{z} - \vect{b}) + \frac{\rho}{2} \norm{A \vect{x} + B \vect{z} - \vect{b}}
+\mathcal{L}_{\rho}(\vect{x},\vect{z},\vect{y})= f(\vect{x}) + g(\vect{z}) + \vect{y}^T (A \vect{x} + B \vect{z} - \vect{b}) + \frac{\rho}{2} \norm{A \vect{x} + B \vect{z} - \vect{b}}^2
 \end{align}
@@
 + \vect{w}^T (\tilde{\vect{\gamma}}_\epsilon - \vect{\gamma}_\epsilon) 
-&+ \frac{\sigma}{2} \norm{\tilde{\vect{\gamma}}_\epsilon - \vect{\gamma}_\epsilon} 
+&+ \frac{\sigma}{2} \norm{\tilde{\vect{\gamma}}_\epsilon - \vect{\gamma}_\epsilon}^2
 + \vect{y}^T (\tilde{\vect{z}} - \vect{z}) 
-+ \frac{\rho}{2} \norm{\tilde{\vect{z}} - \vect{z}}
++ \frac{\rho}{2} \norm{\tilde{\vect{z}} - \vect{z}}^2
 \end{align}
-that is also, using the property $\vect{y}^T\vect{r} + \frac{\rho}{2}\norm{\vect{r}} = \frac{\rho}{2}\norm{\vect{r}+\frac{1}{\rho} \vect{y}} - \frac{1}{2\rho} \norm{y}$, the following
+that is also, using the property $\vect{y}^T\vect{r} + \frac{\rho}{2}\norm{\vect{r}}^2 = \frac{\rho}{2}\norm{\vect{r}+\frac{1}{\rho} \vect{y}}^2 - \frac{1}{2\rho} \norm{y}^2$, the following
 \begin{align}
@@
 +  \mathcal{I}_\Upsilon(\vect{\gamma}_\epsilon) \nonumber \\
-+ \frac{\sigma}{2} \norm{\tilde{\vect{\gamma}}_\epsilon - \vect{\gamma}_\epsilon + \frac{1}{\sigma} \vect{w}} 
-&- \frac{1}{2\sigma} \norm{w}
-+ \frac{\rho}{2} \norm{\tilde{\vect{z}} - \vect{z} + \frac{1}{\rho} \vect{y}}
-- \frac{1}{2\rho} \norm{y}
++ \frac{\sigma}{2} \norm{\tilde{\vect{\gamma}}_\epsilon - \vect{\gamma}_\epsilon + \frac{1}{\sigma} \vect{w}}^2 
+&- \frac{1}{2\sigma} \norm{w}^2
++ \frac{\rho}{2} \norm{\tilde{\vect{z}} - \vect{z} + \frac{1}{\rho} \vect{y}}^2
+- \frac{1}{2\rho} \norm{y}^2
 \end{align}
@@
     \text{min} \quad  &  \frac{1}{2} \tilde{\vect{\gamma}}_\epsilon^T N \tilde{\vect{\gamma}}_\epsilon + \vect{r}^T_\epsilon \tilde{\vect{\gamma}}_\epsilon  
-+ \frac{\sigma}{2} \norm{\tilde{\vect{\gamma}}_\epsilon - \vect{\gamma}_\epsilon^k}
-+ \frac{\rho}{2} \norm{\tilde{\vect{z}} - \vect{z}^k + \frac{1}{\rho} \vect{y}^k} \\
++ \frac{\sigma}{2} \norm{\tilde{\vect{\gamma}}_\epsilon - \vect{\gamma}_\epsilon^k}^2
++ \frac{\rho}{2} \norm{\tilde{\vect{z}} - \vect{z}^k + \frac{1}{\rho} \vect{y}^k}^2 \\
 	  \text{s.t.} \quad &  \tilde{\vect{\gamma}} = \tilde{\vect{z}}
@@
 		\vect{\gamma}^{k+1}_\epsilon &=\tilde{\vect{\gamma}}_\epsilon^{k+1} + (1-\alpha) \vect{\gamma}_\epsilon^k \\
-		\vect{z}^{k+1} &= \Pi_\Upsilon \left( \alpha \vect{\gamma}_\epsilon^{k+1} + (1-\alpha) \vect{z}^k  - \frac{1}{\rho} \vect{y}^k \right) \\
-		\vect{y}^{k+1} &= \vect{y}^k + \rho \left( \alpha \vect{\gamma}_\epsilon^{k+1} + (1-\alpha) \vect{z}^k - \vect{z}^{k+1} \right) \\
+		\vect{z}^{k+1} &= \Pi_\Upsilon \left( \alpha \tilde{\vect{\gamma}}_\epsilon^{k+1} + (1-\alpha) \vect{z}^k  + \frac{1}{\rho} \vect{y}^k \right) \\
+		\vect{y}^{k+1} &= \vect{y}^k + \rho \left( \alpha \tilde{\vect{\gamma}}_\epsilon^{k+1} + (1-\alpha) \vect{z}^k - \vect{z}^{k+1} \right) \\
 		\vect{r}_{\text{prim}} &= \vect{\gamma}_\epsilon^{k+1}-\vect{z}^{k+1} \\
-		\vect{r}_{\text{dual}} &= N \vect{\gamma}_\epsilon^{k+1}+\vect{r}-\vect{y}^{k+1} 
+		\vect{r}_{\text{dual}} &= N \vect{\gamma}_\epsilon^{k+1}+\vect{r}+\vect{y}^{k+1} 
 	\end{align}

### 2020-11-04 2021/ADMM_journal/author.tex
Update article

@@
 \usepackage{tabularx}
-\usepackage[ruled,vlined, linesnumbered]{algorithm2e}
+\usepackage[ruled,vlined]{algorithm2e}  % ruled,vlined,linesnumbered
 \usepackage{array}
@@
 
-\title{Solving non smooth dynamics using ADMM}
+\title{Solving Cone Complementarity Problems in Non-Smooth Dynamics using the Alternating Direction Method of Multipliers}
 
@@
 		loads are added too (ex. when finite elements come into play) it becomes a block-sparse matrix including the tangent stiffness and damping matrices,
-		for instance $H=M-{h^2}\nabla_q\vect{f}-{h}\nabla\vect{f}$ in case of the time stepper in \cite{TasoraAnitescuCMAME10}
+		for instance $H=M-{h^2}\nabla_q\vect{f}-{h}\nabla_v\vect{f}$ in case of the first-order time stepper in \cite{TasoraAnitescuCMAME10}
     \item $D_\epsilon$ is a sparse matrix, the transpose jacobian of all constraints
@@
 		 H   & D_\epsilon \\
-		 D_\epsilon^T & - \rho I - C
+		 D_\epsilon^T & - \rho I - C_\epsilon
 		\end{bmatrix}
@@
 \[
-   \vect{r}_{\text{dual}}^{k+1} = D_\epsilon^T \vect{v}^{k+1} +C \vect{\gamma}_\epsilon^{k+1} + \vect{r} + \vect{y}^{k+1}
+   \vect{r}_{\text{dual}}^{k+1} = D_\epsilon^T \vect{v}^{k+1} +C_\epsilon \vect{\gamma}_\epsilon^{k+1} + \vect{r} + \vect{y}^{k+1}
 \]
@@
 \begin{align}
-	\eta_\rho = \frac{ \norm{\vect{r}_{\text{primal}}^{k}} }{ \norm{\vect{r}_{\text{dual}}^{k}} }  
+	\eta_\rho = \frac{ \norm{\vect{r}_{\text{primal}}^{k}}_\infty }{ \norm{\vect{r}_{\text{dual}}^{k}}_\infty }  
 \end{align}
 We define the \textit{balanced normalized} scaling policy as
-\begin{align}
-	\eta_\rho = \left( \frac{\norm{\vect{r}_{\text{primal}}^{k}}}{ \norm{\vect{z}^k}+\norm{\vect{\gamma}_\epsilon^k} +10^{-11} } \right)
-															\left( \frac{\norm{\vect{r}_{\text{dual}}^{k}}}  { \norm{\vect{y}^k } +10^{-11} }  \right)^{-1}
+% this is the "balanced_fast" policy in our MATLAB code
+\begin{align}
+	\eta_\rho = \left( \frac{\norm{\vect{r}_{\text{primal}}^{k}}_\infty}{ \max{ \left( \norm{\vect{z}^k}_\infty,\norm{\vect{\gamma}_\epsilon^k}_\infty \right) } } \right)
+															\left( \frac{\norm{\vect{r}_{\text{dual}}^{k}}_\infty}  { \norm{\vect{y}^k }_\infty }  \right)^{-1}
 \end{align}
@@
 
-\subsection{Preconditioning}
-
-We experienced that it
-
 \section{Final algorithm}
@@
 Using the above results, one can finally obtain the ADDM algorithm for solving the CCP of the non-smooth dynamics timestepper:
-
-TO UPDATE!!!!
 
@@
 \DontPrintSemicolon
-\KwIn{ Initial approximations $\vect{\gamma}_\epsilon^0$, $\vect{y}^0$ }
-\KwOut{ Solution $\vect{\gamma}_\epsilon$, $\vect{v}$ }
+\KwIn{ Initial approximations $\vect{y}^0$, $\vect{z}^0$ }
+\KwOut{ Solution $\vect{\gamma}_\epsilon^\star$, $\vect{v}^\star$, optionally also $\vect{y}^\star$, $\vect{z}^\star$ }
 \SetKwFunction{StepAdjustPolicy}{StepAdjustPolicy}
@@
 			 H   & D_\epsilon \\
-			 D^T & - \Theta^k I - C
+			 D^T & - \Theta^k I - C_\epsilon
 			\end{bmatrix}
@@
 		$\vect{r}_{\text{dual}}^{k+1}   = \Theta^k (\vect{z}^{k+1} - \vect{z}^{k})$ \;
-		\lIf{$\norm{\vect{r}_{\text{primal}}^{k+1}} < \epsilon_{\text{primal}}$ and $\norm{\vect{r}_{\text{dual}}^{k+1}} < \epsilon_{\text{dual}}$ }{
+		\If{$\norm{\vect{r}_{\text{primal}}^{k+1}} < \epsilon_{\text{primal}}$ and $\norm{\vect{r}_{\text{dual}}^{k+1}} < \epsilon_{\text{dual}}$ }{
 			break \;
 		}
-		$\rho^{k+1} = $ \StepAdjustPolicy{$\rho^{k}$},  $\rho^{k+1} \rightarrow \Theta^{k+1}$ \;
+		$\rho^{k+1} = $ \StepAdjustPolicy{$\rho^{k}$} \; 
+		$\rho^{k+1} \rightarrow \Theta^{k+1}$ \;
 }
 %\Return{location}\;
-\caption{Simple form of the ADMM algorithm, with no $N$ Schur complement.}
+\caption{The optimized ADMM algorithm}
 \label{algo:admm1}
@@
 
-TO DO:
-
-TO UPDATE!!!!
-
-ALSO MAYBE THIS ALPHA VARIANT WILL BE REMOVED!!!
-
-** Using these results, and adding a Krasnoselskii-Mann relaxation factor $\alpha \in(0,2]$, one obtains the final algorithm:
-%
-\begin{subequations}
-	\begin{align}
-    \begin{bmatrix}
-		 H   & D_\epsilon \\
-		 D^T & -(\sigma+\rho) I
-		\end{bmatrix}
-		\begin{Bmatrix}
-		  \tilde{\vect{v}}^{k+1}  \\
-		 -\tilde{\vect{\gamma}}_\epsilon^{k+1}
-		\end{Bmatrix}
-		&=
-		\begin{Bmatrix}
-		 \vect{k} \\
-		 -\vect{b}_\epsilon + \sigma \vect{\gamma}_\epsilon^k +\rho \vect{z}^k - \vect{y}^k 
-		\end{Bmatrix}
-		\\
-		\vect{\gamma}^{k+1}_\epsilon &=\tilde{\vect{\gamma}}_\epsilon^{k+1} + (1-\alpha) \vect{\gamma}_\epsilon^k \\
-		\vect{v}^{k+1} &=\tilde{\vect{v}}^{k+1} + (1-\alpha) \vect{v}^k \\
-		\vect{z}^{k+1} &= \Pi_\Upsilon \left( \alpha \tilde{\vect{\gamma}}_\epsilon^{k+1} + (1-\alpha) \vect{z}^k  + \frac{1}{\rho} \vect{y}^k \right) \\
-		\vect{y}^{k+1} &= \vect{y}^k + \rho \left( \alpha \tilde{\vect{\gamma}}_\epsilon^{k+1} + (1-\alpha) \vect{z}^k - \vect{z}^{k+1} \right) \\
-		\vect{r}_{\text{prim}} &= \vect{\gamma}_\epsilon^{k+1}-\vect{z}^{k+1} \\
-		\vect{r}_{\text{dual}} &= N \vect{\gamma}_\epsilon^{k+1}+\vect{r}+\vect{y}^{k+1} 
-	\end{align}
-	\label{eq:admm_final}
-\end{subequations}
-
-TO DO: CUT THIS "`OPTIMIZATIONS"' SECT, HERE ONLY FOR CUT PASTING
-
-\section{Optimizations}
-
-An interesting property of this formulation is that one does not need to recover $\vect{v} = \amatr{H}^{-1}(\vect{k}+\amatr{D}^T \vect)\vect{\gamma}^{k+1}$ at the end because, for $\alpha=1$, this is also a byproduct of the linear system solve as $\vect{v}= \tilde{\vect{v}}^{k+1}$.
-In case of $\alpha\neq1$, one needs to keep it updated as 
-$\vect{v}^{k+1} =\tilde{\vect{v}}^{k+1} + (1-\alpha) \vect{v}^k$; 
-but this also means that an initial value of $\vect{v}^0$, consistent with the initial value $\vect{\gamma}^0$, must be computed, hence before the iteration one has to initialize 
-\[
-\vect{v}^0 = \amatr{H}^{-1}(\vect{k} + \amatr{D} \vect{\gamma}^0)
-\]
-
-We see that the dual residual $\vect{r}_{\text{dual}}$ is computed using the $\amatr{N}$ matrix, but can be inefficient as we do not want to compute it explicitly.
-One option is to remember the definition of the Schur complement and use
-\[
-\vect{r}_{\text{dual}} = \amatr{D}^T \vect{v}^{k+1} - \amatr{E} \vect{\gamma}^{k+1} + \vect{b}_\epsilon + \vect{y}
-\]
-Another even faster option (but only for $\alpha=1$) would be to compute the dual residual using
-\[
-\vect{r}_{\text{dual}} = \rho (\vect{z}^{k+1} - \vect{z}^{k} )
-\]
+The algorithm above contains a single computational bottleneck, that is the solution of the linear system.
+However, if a direct solver is used, a LU decomposition\footnote{In many cases the H matrix is hermitian, for example when just block-diagonal mass matrices are used, hence the decomposition of the symmetric indefinite saddle point matrix can be done via a LDL decomposition, which is faster than the LU decomposition.
+However there are cased where un-symmetric tangent stiffness matrices might be added to H, as happens in some finite element problems, precluding this optimization.} could be computed at the beginning and then recomputed only when the step size is adjusted, otherwise in all other iterations with $\rho^{k+1}=\rho^{k}$ the decomposition is unaltered and only a relatively inexpensive back solve is required.
+
+\subsection{Preconditioning}
+
+We experienced that the convergence of Algorithm \ref{algo:admm1} is negatively affected by the presence of odd mass ratios in the mechanical system.
+The immediate effect of uneven mass ratio is a broadening of the interval of singular values of the $N$ matrix.
+Note that this can be also a consequence of using different measuring units for different constraint multipliers $\gamma_i$, for instance even a system with odd mass ratios can have a badly conditioned $N$ matrix if some constraint reactions are assumed in newtons, other in kilo newtons - clearly, as the choice of measuring units is arbitrary, one should find a way to make the solver as much insensitive as possible to the scaling of the constraints.
+
+A possible remedy for this difficulty is to perform a simplified form of preconditioning by doing a diagonal scaling of $N$ and $\vect{r}_\epsilon$ before starting the iteration, thus operating on the scaled form of the problem:
+%
+\begin{align}
+	\text{min} \quad & \frac{1}{2} \breve{\vect{\gamma}}_\epsilon^T \breve{N} \breve{\vect{\gamma}}_\epsilon + \breve{\vect{r}}^T_\epsilon \breve{\vect{\gamma}}_\epsilon \\
+	\text{s.t.} \quad & \breve{\vect{\gamma}}_\epsilon \in \breve{\Upsilon} 
+	\label{eq:ChronoCCP_min_scaled}
+\end{align}
+%
+with $\breve{\vect{\gamma}}_\epsilon = S^{-1} \vect{\gamma}_\epsilon$, $\breve{N} = S N S$, 
+$\breve{\vect{r}}^T_\epsilon = S \vect{r}^T_\epsilon$. 
+We set $S_{ii} = \sqrt{\frac{1}{N_{ii}}}$. 
+%As usual we avoid using directly $N$ and we rather compute
+%\[
+%N_{ii} = D_{\epsilon,i}^T H^{-1} D_{\epsilon,i} + C_{\epsilon,ii}
+%\]
+Note that other more advanced (and CPU intensive) preconditioners could be used, for example in \cite{Stellato2020} a modified Ruitz equilibration is proposed.  
+After the solution is computed, one can quickly recover $\vect{\gamma}_\epsilon = S \breve{\vect{\gamma}}_\epsilon$.
+
+Note that in sake of high performance we never build explicitly $\breve{N}$ and $\breve{\vect{r}}^T_\epsilon$: what we need is just a scaling of the saddle point matrix involved in \eqref{eq:admm_step1}, that will simply include the scaled jacobians
+$\breve{D}_\epsilon = S D_\epsilon$, the scaled compliance
+$\breve{C}_\epsilon = S C_\epsilon S$,  the scaled term 
+$\breve{\vect{b}}_\epsilon = S \breve{\vect{b}}_\epsilon$.
+
+The preconditioned form of the ADMM algorithm, that we call P-ADMM heretofore, is not reported here for compactness: it is enough to replace the above scaled quantities in Agorithm \ref{algo:admm1}, where $\breve{\vect{\gamma}}_\epsilon,\breve{\vect{y}},\breve{\vect{z}}$ will replace $\vect{\gamma}_\epsilon, \vect{y}, \vect{z}$, then a final step is added after convergence, to recover the unscaled solution: $\vect{\gamma}_\epsilon = S \breve{\vect{\gamma}}_\epsilon$.
+Moreover, if one is interested in the residuals in the not-scaled original form, for instance to verify convergence, these can be computed as
+%
+\begin{align}
+\vect{r}_{\text{primal}}^{k+1} &= S \left( \breve{\vect{\gamma}}_\epsilon^{k+1}-\breve{\vect{z}}^{k+1} \right) \nonumber \\
+\vect{r}_{\text{dual}}^{k+1}   &= S^{-1} \Theta^k (\breve{\vect{z}}^{k+1} - \breve{\vect{z}}^{k}) \nonumber
+\end{align}
+
+\subsection{Warm starting}
+
+One of the nice properties of ADMM methods is that they can be easily warm started, something that for example is difficult to do with IPMs. 
+This is a relevant feature in our context, because we must solve the problem \eqref{eq:ChronoCCP} at each time step, where one can expect some degree of temporal coherency in the values of $\vect{\gamma}_\epsilon$ between each step.
+This is especially true for simulations involving stacked objects, like when simulating environments for robots, granular flows or masonry buildings, because the contact forces often show limited changes over time, and the amount of contacts that change state can be limited.
+If so, one can reuse the last computed values of $\vect{\gamma}_\epsilon$ in the previous time step to warm-start the ADMM solver at the next time step.
+
+This poses two difficulties.
+First of all, one must develop algorithms and data structures that are able to convey information from the contact at time $t_A$ to a time $t_B = t_A+h$.
+in fact, after ADMM computes the contact forces at step $t_A$, those could be saved into the contact data structures, however those contacts would be recomputed by the collision detection engine at the next time step $t_B$: since the objects move, also the contact manifold changes, thus it is not trivial to compute which contacts are persistent between the two steps.
+By using some heuristics and tolerances, one can detect if some contacts at time $t_B$ are the same contacts used at time $t_A$, if so, their last reaction forces $\vect{\gamma}_\epsilon$ can be copied to the new contacts and used for the warm starting, whereas completely new contacts are initialized with $\vect{\gamma}_\epsilon = 0$.
+
+Furthermore, one can see that the ADDM Algorithm \ref{algo:admm1} is a fixed point 
+$\{\vect{y}^{k+1},\vect{z}^{k+1}\} = \vect{g} \left(\{\vect{y}^k,\vect{z}^k\}\right)$, thus one needs $\{\vect{y}^0,\vect{z}^0\}$ to warm start it, rather than $\vect{\gamma}_\epsilon^0$.
+This leads to two options: store both $\vect{y}$ and $\vect{z}$ values in the contact data (wasting some memory in case of large scale simulations, and complicating both data structures and bookkeeping) otherwise assume that the last ADMM run converged exactly up to $\vect{r}_{\text{primal}}=0$, so one could just store $\vect{\gamma}_\epsilon$ values in contacts, and then compute
+%
+\begin{align}
+\vect{v}^0 &= H^{-1}( \vect{k} + D_\epsilon \vect{\gamma}_\epsilon^0) \nonumber \\
+\vect{y}^0 &= - (D_\epsilon^T \vect{v}^0 + C \vect{\gamma}_\epsilon^0 + \vect{b}_\epsilon)  \nonumber \\
+\vect{z}^0 &= \vect{\gamma}_\epsilon^0 \nonumber
+\end{align}
+This idea, implemented in Algorithm \ref{algo:admm2}, provides a speedup of 2x-10x in our tests.
+
+\begin{algorithm}
+\DontPrintSemicolon
+\KwIn{ Initial approximation $\vect{\gamma}_\epsilon^0$, optional: $\vect{v}^0$}
+\KwOut{ Solution $\vect{\gamma}_\epsilon^\star$, $\vect{v}^\star$, optionally also $\vect{y}^\star$, $\vect{z}^\star$ }
+\SetKwFunction{ADMM}{ADMM}
+\If{$\vect{v}^0$ is not provided}
+{
+	$\vect{v}^0 = H^{-1}( \vect{k} + D_\epsilon \vect{\gamma}_\epsilon^0)$ \;
+}
+$\vect{y}^0 = - (D_\epsilon^T \vect{v}^0 + C \vect{\gamma}_\epsilon^0 + \vect{b}_\epsilon)$ \;
+$\vect{z}^0 = \vect{\gamma}_\epsilon^0$ \;
+$\vect{\gamma}_\epsilon^\star, \vect{v}^\star = $ \ADMM{$\vect{y}^0$,$\vect{z}^0$} \; 
+\caption{Warm-started ADMM}
+\label{algo:admm2}
+\end{algorithm}
+
+% REMOVED - the overrelaxation form is not discussed
+%** Using these results, and adding a Krasnoselskii-Mann relaxation factor $\alpha \in(0,2]$, one obtains the final algorithm:
+%
+%\begin{subequations}
+	%\begin{align}
+    %\begin{bmatrix}
+		 %H   & D_\epsilon \\
+		 %D^T & -(\sigma+\rho) I
+		%\end{bmatrix}
+		%\begin{Bmatrix}
+		  %\tilde{\vect{v}}^{k+1}  \\
+		 %-\tilde{\vect{\gamma}}_\epsilon^{k+1}
+		%\end{Bmatrix}
+		%&=
+		%\begin{Bmatrix}
+		 %\vect{k} \\
+		 %-\vect{b}_\epsilon + \sigma \vect{\gamma}_\epsilon^k +\rho \vect{z}^k - \vect{y}^k 
+		%\end{Bmatrix}
+		%\\
+		%\vect{\gamma}^{k+1}_\epsilon &=\tilde{\vect{\gamma}}_\epsilon^{k+1} + (1-\alpha) \vect{\gamma}_\epsilon^k \\
+		%\vect{v}^{k+1} &=\tilde{\vect{v}}^{k+1} + (1-\alpha) \vect{v}^k \\
+		%\vect{z}^{k+1} &= \Pi_\Upsilon \left( \alpha \tilde{\vect{\gamma}}_\epsilon^{k+1} + (1-\alpha) \vect{z}^k  + \frac{1}{\rho} \vect{y}^k \right) \\
+		%\vect{y}^{k+1} &= \vect{y}^k + \rho \left( \alpha \tilde{\vect{\gamma}}_\epsilon^{k+1} + (1-\alpha) \vect{z}^k - \vect{z}^{k+1} \right) \\
+		%\vect{r}_{\text{prim}} &= \vect{\gamma}_\epsilon^{k+1}-\vect{z}^{k+1} \\
+		%\vect{r}_{\text{dual}} &= N \vect{\gamma}_\epsilon^{k+1}+\vect{r}+\vect{y}^{k+1} 
+	%\end{align}
+	%\label{eq:admm_final}
+%\end{subequations}
+%
+%An interesting property of this formulation is that one does not need to recover $\vect{v} = \amatr{H}^{-1}(\vect{k}+\amatr{D}^T \vect)\vect{\gamma}^{k+1}$ at the end because, for $\alpha=1$, this is also a byproduct of the linear system solve as $\vect{v}= \tilde{\vect{v}}^{k+1}$.
+%In case of $\alpha\neq1$, one needs to keep it updated as 
+%$\vect{v}^{k+1} =\tilde{\vect{v}}^{k+1} + (1-\alpha) \vect{v}^k$; 
+%but this also means that an initial value of $\vect{v}^0$, consistent with the initial value $\vect{\gamma}^0$, must be computed, hence before the iteration one has to initialize 
+%\[
+%\vect{v}^0 = \amatr{H}^{-1}(\vect{k} + \amatr{D} \vect{\gamma}^0)
+%\]
 
@@
 
-\begin{figure}[h!]
+\begin{figure}[!tbp]
   \centering
-\includegraphics[width=0.50\textwidth]{t8_convergence.pdf}
-\includegraphics[width=0.35\textwidth, trim=0cm -3cm 0 3cm]{t8_snapshot.png}
-\caption{Convergence of the ADMMM method within one time step of the wrecking ball benchmark (600 bricks in four walls): residual in frictional constraint violation compared to fixed point Jacobi iterations and to first-order SPG methods.}
-\label{fig:t8}
+  \begin{minipage}[t]{0.48\textwidth}
+    \includegraphics[width=0.8\textwidth, trim=0cm -3cm 0 3cm]{t8_snapshot.png}
+    \caption{Snapshot from \textsc{Test 2}, the wrecking ball benchmark (600 bricks in four walls).}
+		\label{fig:t6_primdual_a_ns}
+  \end{minipage}
+  \hfill
+	\begin{minipage}[t]{0.48\textwidth}
+    \includegraphics[width=\textwidth]{t8_convergence.pdf}
+    \caption{Constraint violation compared to fixed point Jacobi iterations and to SPG methods, in \textsc{Test 2}.}
+		\label{fig:t8_convergence}
+  \end{minipage}
 \end{figure}
 
+\begin{figure}[!tbp]
+  \centering
+  \begin{minipage}[t]{0.48\textwidth}
+    \includegraphics[width=\textwidth]{t8_primdual.pdf}
+    \caption{Primal-dual convergence of the algorithm in \textsc{Test 2}.}
+		\label{fig:t8_primdual}
+  \end{minipage}
+  \hfill
+	\begin{minipage}[t]{0.48\textwidth}
+    \includegraphics[width=\textwidth]{t8_compliance.pdf}
+    \caption{Effect of regularization or compliance in contacts in \textsc{Test 2}.}
+  \end{minipage}
+\end{figure}
+
 \section{Results and conclusion}
 
-We performed benchmarks involving multi-body systems with contacts between multiple parts, showing that the performance of the ADMM method is capable of handling problems that would converge too slowly using conventional projected fixed point methods (see Fig.\ref{fig:t8}) or even first-order SPG spectral methods \cite{heynIJNME2013}.
+We performed benchmarks involving multi-body systems with contacts between multiple parts, showing that the performance of the ADMM method is capable of handling problems that would converge too slowly using conventional projected fixed point methods or first-order SPG spectral methods \cite{heynIJNME2013}.
 
@@
 We noted that ADMM can be successfully applied to problems that exhibit temporal coherence because, unlike IPMs, it supports warm-starting. 
+Another optimization that allowed superior performance is the adoption of a diagonal preconditioning, with block scaling for the triplets of lagrangian multipliers relative to the conic constraints.
+
+Further research on this topic might address the acceleration of the ADMM method by means of Anderson acceleration.
 

### 2020-11-19 2021/ADMM_journal/author.tex
Add benchmarks in ADMM paper.

@@
 
+Note that also the last $\rho^k$ value from the previous time step can be used as the initial $\rho^0$ for the next step, adding a further optimization.
+Statistically, given some degree of temporal coherency in the mechanical system, such pre-tuned value would be better than a default value like $\rho^0=0.1$.
+
 \begin{algorithm}
@@
 
+All benchmarks have been computed on a Intel\textsuperscript{\textregistered} Core\textsuperscript{\texttrademark} i7-8750H CPU, clocked at \SI{2.20}{GHz}, with 6 physical cores and \SI{16}{GB} of RAM.
+The ADMM method has been implemented
+in C++ within the Chrono open-source library, using the Eigen library v.3.37 for dense and sparse linear algebra, and Intel\textsuperscript{\textregistered} MKL Pardiso as a direct solver for sparse linear systems.
+
 \subsection{High stack with odd mass ratio}
@@
   \begin{minipage}[t]{0.48\textwidth}
-    \includegraphics[width=\textwidth]{t6_snapshot.png}
+    \includegraphics[width=\textwidth]{pics/t6_snapshot.png}
     \caption{Setup of \textsc{Test 1}.}
@@
   \begin{minipage}[t]{0.48\textwidth}
-    \includegraphics[width=\textwidth]{t6_convergence_b.pdf}
+    \includegraphics[width=\textwidth]{pics/t6_convergence_b.pdf}
     \caption{Convergence of ADMM methods compared to the Jacobi projected fixed point iteration, for \textsc{Test 1.b} 
@@
   \begin{minipage}[t]{0.48\textwidth}
-    \includegraphics[width=\textwidth]{t6_primdual_a.pdf}
+    \includegraphics[width=\textwidth]{pics/t6_primdual_a.pdf}
     \caption{Convergence of primal and dual residuals in \textsc{Test 1.a}.}
@@
   \begin{minipage}[t]{0.48\textwidth}
-    \includegraphics[width=\textwidth]{t6_primdual_b.pdf}
+    \includegraphics[width=\textwidth]{pics/t6_primdual_b.pdf}
     \caption{Convergence of primal and dual residuals in \textsc{Test 1.b}. }
@@
   \begin{minipage}[t]{0.48\textwidth}
-    \includegraphics[width=\textwidth]{t6_primdual_a_ns.pdf}
+    \includegraphics[width=\textwidth]{pics/t6_primdual_a_ns.pdf}
     \caption{Convergence of primal and dual residuals in \textsc{Test 1.a} for varying frequency $n_s$ of automatic updates to the step size, in the ADMM algorithm.}
@@
 	\begin{minipage}[t]{0.48\textwidth}
-    \includegraphics[width=\textwidth]{t6_primdual_a_rho.pdf}
+    \includegraphics[width=\textwidth]{pics/t6_primdual_a_rho.pdf}
     \caption{Convergence of primal and dual residuals in \textsc{Test 1.a} for varying values of the step size $\rho$ in the FADMM algorithm.}
@@
 
-\subsection{Bulldozer on stone slabs}
-
-TO DO
+\subsection{Wrecking ball}
+
+We modeled four parallel walls made with 10 rows of 15 bricks each.
+Bricks length, height, width are respectively \SI{3.96}{m}, \SI{2}{m}, \SI{4}{m}, and their density is \SI{100}{kg/m^3}.
+A wrecking ball with diameter \SI{8}{m} and density \SI{8000}{kg/m^3} impacts horizontally with the 600 bricks.
+The density of the bricks respect to the ball is deliberately high in order to generate a badly conditioned problem with odd mass ratios.
+The friction coefficient is $\mu=0.4$ and the time step is $h=\SI{0.02}{s}$.
+Figure \ref{fig:t6_primdual_a_ns} shows a snapshot from the simulation.
+The convergence of the ADMM method, as shown in Fig.\ref{fig:t8_convergence} for a random time step, is noticeably superior to the convergence of fixed point iterations such as projected Jacobi.
+The method exhibit even better convergence when using the diagonal preconditioning \eqref{eq:ChronoCCP_min_scaled}, as shown Fig.\ref{fig:t8_primdual}, because of the odd scaling of the masses.
 
@@
   \begin{minipage}[t]{0.48\textwidth}
-    \includegraphics[width=0.8\textwidth, trim=0cm -3cm 0 3cm]{t8_snapshot.png}
+    \includegraphics[width=\textwidth, trim=0cm -0cm 0cm 0cm]{pics/t8_snapshot.png}
     \caption{Snapshot from \textsc{Test 2}, the wrecking ball benchmark (600 bricks in four walls).}
@@
 	\begin{minipage}[t]{0.48\textwidth}
-    \includegraphics[width=\textwidth]{t8_convergence.pdf}
+    \includegraphics[width=\textwidth]{pics/t8_convergence.pdf}
     \caption{Constraint violation compared to fixed point Jacobi iterations and to SPG methods, in \textsc{Test 2}.}
@@
   \begin{minipage}[t]{0.48\textwidth}
-    \includegraphics[width=\textwidth]{t8_primdual.pdf}
+    \includegraphics[width=\textwidth]{pics/t8_primdual.pdf}
     \caption{Primal-dual convergence of the algorithm in \textsc{Test 2}.}
@@
 	\begin{minipage}[t]{0.48\textwidth}
-    \includegraphics[width=\textwidth]{t8_compliance.pdf}
+    \includegraphics[width=\textwidth]{pics/t8_compliance.pdf}
     \caption{Effect of regularization or compliance in contacts in \textsc{Test 2}.}
@@
 
+\subsection{Robotic manipulator}
+
+\begin{figure}[!tbp]
+  \centering
+	\includegraphics[width=0.32\textwidth]{pics/robot_80.png}
+  \includegraphics[width=0.32\textwidth]{pics/robot_200.png}
+	\includegraphics[width=0.32\textwidth]{pics/robot_280.png}
+	\hfill
+  \caption{\textsc{Test 3}.
+Robot interacting with the environment via contacts between the gripper and few boxes.
+The robot features extremely odd mass ratios, rheonomic constraints and redundant constraints.}
+	\label{fig:tire}
+\end{figure}
+
+This benchmark features a 6-DOF industrial robot whose end effector interacts with 15 boxes with height \SI{0.3}{m} and width \SI{0.4}{m}, moving them in the working area via contacts and collisions, as shown in Fig.\ref{fig:robot_200}.
+The friction coefficient between the boxes is $\mu=0.4$, the time step is $h=\SI{0.01}{s}$.
+The difficulty highlighted by this benchmark is the simultaneous presence of non-smooth contacts and bilateral constraints in a critical articulated mechanism.
+In fact the simulation of a robot arm actuated via motorized joints is a trivial problem, but here we do the opposite: we drive the end effector via a rheonomic constraint defining an imposed trajectory respect to the base, hence the rest of the arm will naturally move to the configuration prescribed by joint constraints; however since the mass of the end effector is much lighter than the bicept, since some joints are redundant\footnote{The robot has a gravity compensator modeled via two revolute joints and one cylindrical joint, leading to an overconstrained loop.} and since the robot passes close to a singularity, the simulation of the robot alone is a challenging task.
+That is, even without contacts, most iterative solvers like SOR or Jacobi would converge too slowly, and even Krylov linear solvers like GMRES, CG, MINRES would require many iterations unless equipped by some preconditioner.
+In our test, convergence to high precision is achieved in 15 iterations at most, where just one or two of them require a refactorization of the saddle-point matrix.
+Moreover, when there are no contacts, the solution is provided in exactly one step, as the iteration boils down to a single factorization followed by a forward/back solve just like when simulating the robot via a direct solver in the conventional context of smooth dynamics.
+
+\subsection{Deformable bars}
+
+This benchmark aims at estimating the efficiency and robustness of the method in problems involving also finite elements.
+Few methods already exist in literature that can simulate deformable structures in the context of non-smooth dynamics, but most of them target the field of interactive computer graphics, where the stiffness of structures is often very low.
+In our case, being interested in engineering applications, we require that the method would be able to handle both very deformable structures (ex.
+rubber-like materials) as well as very stiff structures.
+In this benchmark, we simulate the fall of 15 deformable bars, each modeled with a mesh of tetrahedral finite elements, for a total of 9360 degrees of freedom.
+The outer skin of the bars can collide with each other and with rigid bodies - in this case there is just a single rigid body, namely the flat ground.
+At the contact points there is a friction coefficient $\mu=0.3$ and a null restitution coefficient, the time step is $h=\SI{0.05}{s}$.
+In all simulations the material of the bars has a density of \SI{1e3}{kg/m^3}, a 0.01 Rayleigh damping coefficient and a Poisson ratio $\nu=0.3$, but in the case of \textsc{Test 4.a} (see Fig.\ref{fig:ribbons_1}) the Young modulus is $E=\SI{1e6}{Pa}$, whereas in \textsc{Test 4.b} (see Fig.\ref{fig:ribbons_2}) the Young modulus is many orders of magnitude higher: $E=\SI{2e11}{Pa}$.
+Tests has shown that our ADMM method converges equally well even in the case of extreme stiffness, converging to the required  tolerances with comparable number of iterations.
+At each time step, both \textsc{Test 4.a} and \textsc{Test 4.b} required an average number of 3 factorizations for the adjustment of the $\rho^i$ step and a variable number of forward/back solves in the range of $4\div50$.
+When bars come into contact, an average number of $1650$ contacts is found, leading to saddle-point linear problems with about 15 thousands of unknowns that are factorized, on average, in \SI{0.116}{s} using the sparse direct solver.
+Each forward/back solve requires \SI{0.0069}{s} on average.
+Each time step then required about one second time of wall clock time.
+
+\begin{figure}[!tbp]
+  \centering
+	\includegraphics[width=0.24\textwidth]{pics/ribbons_3_20.png}
+  \includegraphics[width=0.24\textwidth]{pics/ribbons_3_40.png}
+	\includegraphics[width=0.24\textwidth]{pics/ribbons_3_60.png}
+	\includegraphics[width=0.24\textwidth]{pics/ribbons_3_80.png}
+	\hfill
+  \caption{\textsc{Test 4.a}.
+Frictional contact between deformable bars with low stiffness, at $t=\SI{0.1}{s},t=\SI{0.2}{s},t=\SI{0.3}{s},t=\SI{0.4}{s}$.
+False color of the mesh represents the instantaneous norm of the speed of the nodes.}
+	\label{fig:ribbons_1}
+\end{figure}
+
+%\begin{figure}[!tbp]
+  %\centering
+	%\includegraphics[width=0.24\textwidth]{pics/ribbons_1_20.png}
+  %\includegraphics[width=0.24\textwidth]{pics/ribbons_1_40.png}
+	%\includegraphics[width=0.24\textwidth]{pics/ribbons_1_60.png}
+	%\includegraphics[width=0.24\textwidth]{pics/ribbons_1_80.png}
+	%\hfill
+  %\caption{\textsc{Test 3.c}. Frictional contact between deformable bars with low stiffness, at $t=\SI{0.1}{s},t=\SI{0.2}{s},t=\SI{0.3}{s},t=\SI{0.4}{s}$.}
+	%\label{fig:ribbons_3}
+%\end{figure}
+
+\begin{figure}[!tbp]
+  \centering
+	\includegraphics[width=0.24\textwidth]{pics/ribbons_2_20.png}
+  \includegraphics[width=0.24\textwidth]{pics/ribbons_2_40.png}
+	\includegraphics[width=0.24\textwidth]{pics/ribbons_2_60.png}
+	\includegraphics[width=0.24\textwidth]{pics/ribbons_2_80.png}
+	\hfill
+  \caption{\textsc{Test 4.b}. Frictional contact between deformable bars with high stiffness ($E=\SI{200}{GPa}$) at $t=\SI{0.1}{s},t=\SI{0.2}{s},t=\SI{0.3}{s},t=\SI{0.4}{s}$.}
+	\label{fig:ribbons_2}
+\end{figure}
+
+\subsection{Deformable tire}
+
+\begin{figure}[!tbp]
+  \centering
+	\includegraphics[width=0.32\textwidth]{pics/tire_5.png}
+  \includegraphics[width=0.32\textwidth]{pics/tire_10.png}
+	\includegraphics[width=0.32\textwidth]{pics/tire_15.png}
+	%\includegraphics[width=0.24\textwidth]{pics/tire_20.png}
+	\hfill
+  \caption{\textsc{Test 5}.
+Large deformation of a finite element mesh colliding with a plane.
+Thanks to the non-smooth formulation, large time steps can be used (only ten steps have been used from start to end) and no penalty parameters must be used in contact points.
+Contacts between lugs and ground show zero interpenetration regardless of the stiffness of the mesh.}
+	\label{fig:tire}
+\end{figure}
+
 \section{Results and conclusion}
@@
 
-Our ADMM method requires few computational primitives: basically a projection of dual variables on conic sets, a backward solve of a linear system, and a forward solve.
+Our ADMM method requires few computational primitives: basically a projection of dual variables on conic sets, a forward/backward solve of a linear system, and its factorization.
 The latter is a computational bottleneck, but it can be performed only once per run, as the matrix does not change often during the iterations.

### 2020-11-20 2021/ADMM_journal/author.tex
Addded FADDM algorithm. Completed benchmark section.

@@
 			 H   & D_\epsilon \\
-			 D^T & - \Theta^k I - C_\epsilon
+			 D^T & - \Theta^k - C_\epsilon
 			\end{bmatrix}
@@
 %\Return{location}\;
-\caption{The optimized ADMM algorithm}
+\caption{ADMM for solving CCPs in nonsmooth dynamics}
 \label{algo:admm1}
@@
 
+\section{Accelerated ADMM}
+
+Following the idea presented in \cite{Goldstein2014}, one can improve the convergence of ADMM by introducing a predictor-corrector step based on the Nesterov acceleration, with minimal computational overhead.
+Heretofore we will call this variant as FADDM (Fast ADMM).
+The convergence of the original Nesterov-accelerated ADMM requires the assumption that both $f(\vect{x})$ and $g(\vect{z})$ in \eqref{eq:admm} are strongly convex.
+
+\begin{definition}
+A function $f: \mathbb{R}^n \rightarrow \mathbb{R}$, possibly non differentiable, is \textit{$\beta$-strongly convex} if for all points $\vect{x},\vect{y}$ in its domain it holds:
+\begin{align}
+\label{eq:strongconvex}
+f(\vect{x}) - f(\vect{y}) \geq \left\langle \vect{u}, \vect{x} - \vect{y}  \right\rangle + \frac{\beta}{2} \norm{ \vect{x} - \vect{y} }^2
+ \quad \forall \vect{u} \in \partial f(\vect{y})
+\end{align}
+\end{definition}
+
+In our case, $f(\vect{x})$ is a quadratic function, hence strongly convex, but $g(\vect{z})$ is an indicator function, hence convex but not strongly convex.
+In the case where just one of the two functions is strongly convex, a restart scheme can be applied as in \cite{Goldstein2014}, where the method reverts to the original ADMM if the combined residual $r_{\text{comb}}$ is not monotonically decreasing.
+The monotone decrease is assessed with a factor $\epsilon_r < 1$, as a default value we use
+$\epsilon_r =0.999$ to avoid too frequent restarts. 
+Using the same performance optimization strategies that we used to develop Algorithm \ref{algo:admm1}, we finally obtain the following algorithm:
+
+\begin{algorithm}
+\DontPrintSemicolon
+\KwIn{ Initial approximations $\vect{\gamma}_\epsilon^0$, $\vect{y}^0$ }
+\KwOut{ Solution $\vect{\gamma}_\epsilon^\star$, $\vect{v}^\star$, optionally also $\vect{y}^\star$, $\vect{z}^\star$ }
+\SetKwFunction{StepAdjustPolicy}{StepAdjustPolicy}
+$\alpha^{0} = 1, \quad \Theta=\Theta(\rho), \quad r_{\text{comb}}^{0} = \infty$ \;
+\While{$i \leq n_{max iters}$}{
+    $\vect{z}^{k+1} = \Pi_\Upsilon \left(  \vect{\gamma}_\epsilon^{k} +(\Theta^k)^{-1} \vect{y}^k \right )$ \;
+		$ \begin{bmatrix}
+			 H   & D_\epsilon \\
+			 D^T & - \Theta - C_\epsilon
+			\end{bmatrix}
+			\begin{Bmatrix}
+				\vect{v}^{k+1}  \\
+			 -\vect{\gamma}_\epsilon^{k+1}
+			\end{Bmatrix}
+			=
+			\begin{Bmatrix}
+			 \vect{k} \\
+			 -\vect{b}_\epsilon + \Theta \vect{z}^{k+1} - \vect{y}^k 
+			\end{Bmatrix} $\;
+		$\vect{y}^{k+1} = \vect{y}^k + \Theta \left( \vect{\gamma}_\epsilon^{k+1} - \vect{z}^{k+1} \right) $ \;
+		$\vect{r}_{\text{primal}}^{k+1} = \vect{\gamma}_\epsilon^{k+1}-\vect{z}^{k+1}$ \;
+		$\vect{r}_{\text{dual}}^{k+1}   = \Theta (\vect{z}^{k+1} - \vect{z}^{k})$ \;
+
+		\If{$\norm{\vect{r}_{\text{primal}}^{k+1}} < \epsilon_{\text{primal}}$ and $\norm{\vect{r}_{\text{dual}}^{k+1}} < \epsilon_{\text{dual}}$ }{
+			break \;
+		}
+
+		$r_{\text{comb}}^{k+1}= \Theta^{-1} \norm{\vect{y}^{k+1}-\vect{y}^{k}}_2 + \Theta \norm{\vect{\gamma}_\epsilon^{k+1}-\vect{\gamma}_\epsilon^{k} }_2$
+
+		\If {$r_{\text{comb}}^{k+1} < \epsilon_r r_{\text{comb}}^{k}$ } {
+            $\alpha^{k+1} = \frac{ 1 + \sqrt{1 + 4 (\alpha^{k})^2} }{ 2 } $ \;
+            $\vect{\gamma}_\epsilon^{k+1} = \vect{\gamma}_\epsilon^{k} + \frac{\alpha^{k}-1}{\alpha^{k+1}} (\vect{\gamma}_\epsilon^{k+1} - \vect{\gamma}_\epsilon^{k}) $ \;
+            $\vect{v}^{k+1} = \vect{v}^{k} + \frac{\alpha^{k}-1}{\alpha^{k+1}} (\vect{v}^{k+1} - \vect{v}^{k}) $  \;
+						$\vect{y}^{k+1} = \vect{y}^{k} + \frac{\alpha^{k}-1}{\alpha^{k+1}} (\vect{y}^{k+1} - \vect{y}^{k}) $  \;
+		}
+    \Else{
+            $\alpha^{k+1} = 1 $ \;
+            $r_{\text{comb}}^{k+1} = \frac{1}{\epsilon_r} r_{\text{comb}}^{k} $ \;
+    }
+}
+\caption{FADMM for solving CCPs in nonsmooth dynamics}
+\label{algo:fadmm1}
+\end{algorithm}
+
+We note in passing that the Algorithm \ref{algo:fadmm1} is based on a 
+$\vect{z} \rightarrow \vect{\gamma}_\epsilon \rightarrow \vect{y}$ 
+update ordering whereas the Algorithm \ref{algo:admm1} is based on a 
+$\vect{\gamma}_\epsilon \rightarrow  \vect{z} \rightarrow  \vect{y}$ ordering, leading
+to a slightly modified warm starting method.
+%Both can be used, but the second has the benefit that it is easier to warm 
+% start it from $\{\vect{\gamma}_\epsilon^{0}, \vect{y}^{0}\}$
+
+Also, a preconditioned version with diagonal scaling can be obtained with the already discussed transformation of \eqref{eq:ChronoCCP_min_scaled}, leading to a method that we call P-FADMM in the following.
+
 \section{Benchmarks}
@@
 The robot features extremely odd mass ratios, rheonomic constraints and redundant constraints.}
-	\label{fig:tire}
+	\label{fig:robot}
 \end{figure}
 
-This benchmark features a 6-DOF industrial robot whose end effector interacts with 15 boxes with height \SI{0.3}{m} and width \SI{0.4}{m}, moving them in the working area via contacts and collisions, as shown in Fig.\ref{fig:robot_200}.
+This benchmark features a 6-DOF industrial robot whose end effector interacts with 15 boxes with height \SI{0.3}{m} and width \SI{0.4}{m}, moving them in the working area via contacts and collisions, as shown in Fig.\ref{fig:robot}.
 The friction coefficient between the boxes is $\mu=0.4$, the time step is $h=\SI{0.01}{s}$.
@@
 
-\subsection{Deformable tire}
-
-\begin{figure}[!tbp]
-  \centering
-	\includegraphics[width=0.32\textwidth]{pics/tire_5.png}
-  \includegraphics[width=0.32\textwidth]{pics/tire_10.png}
-	\includegraphics[width=0.32\textwidth]{pics/tire_15.png}
+%\subsection{Deformable tire}
+%
+%\begin{figure}[!tbp]
+  %\centering
+	%\includegraphics[width=0.32\textwidth]{pics/tire_5.png}
+  %\includegraphics[width=0.32\textwidth]{pics/tire_10.png}
+	%\includegraphics[width=0.32\textwidth]{pics/tire_15.png}
 	%\includegraphics[width=0.24\textwidth]{pics/tire_20.png}
-	\hfill
-  \caption{\textsc{Test 5}.
+	%\hfill
+  %\caption{\textsc{Test 5}.
 Large deformation of a finite element mesh colliding with a plane.
@@
 Contacts between lugs and ground show zero interpenetration regardless of the stiffness of the mesh.}
-	\label{fig:tire}
-\end{figure}
+	%\label{fig:tire}
+%\end{figure}
 

### 2020-11-23 2020/GVSETS_2020/GVSETS.tex

@@
 \usepackage{lastpage}
+
+\usepackage{biblatex}
+\addbibresource{../../bibliography/refsAutonomousVehicles.bib,Bib/refsMachineLearning,../../bibliography/refsChronoSpecific.bib,../../bibliography/refsCompSci,../../bibliography/refsDEM.bib,../../bibliography/refsFSI.bib,Bib/refsGraphics.bib,../../bibliography/refsMBS.bib,../../bibliography/refsRobotics.bib,../../bibliography/refsSBELspecific.bib,../../bibliography/refsSensors.bib,../../bibliography/refsTerramech.bib,../../bibliography/refDRL.bib}
+\addbibresource{../../bibliography/refDRL.bib}
+\addbibresource{../../bibliography/refsMBS.bib}
 
@@
 %% ============================================================
-\bibliographystyle{unsrt}
-\bibliography{Bib/refsAutonomousVehicles,Bib/refsMachineLearning,Bib/refsChronoSpecific,Bib/refsCompSci,Bib/refsDEM,Bib/refsFSI,Bib/refsGraphics,Bib/refsMBS,Bib/refsRobotics,Bib/refsSBELspecific,Bib/refsSensors,Bib/refsTerramech,Bib/refsDRL}
+\printbibliography
 

### 2020-11-23 2021/NODYCON_PyChrono/author.tex
Added PyChrono NoDyCon Paper

%%%%%%%%%%%%%%%%%%%% author.tex %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% sample root file for your "contribution" to a proceedings volume
%
% Use this file as a template for your own input.
%
%%%%%%%%%%%%%%%% Springer %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass{svproc}
%
% RECOMMENDED %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%

% to typeset URLs, URIs, and DOIs
\usepackage{url}
\usepackage{graphicx}
\def\UrlFont{\rmfamily}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[ruled,vlined, linesnumbered]{algorithm2e}
\usepackage{algorithmic}

\begin{document}
\mainmatter              % start of a contribution
%
\title{PyChrono and gym-chrono: a Deep Reinforcement Learning framework leveraging Multibody Dynamics to \\ control Autonomous Vehicles and Robots.}
%
\titlerunning{PyChrono and gym-chrono}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Simone Benatti\inst{1} \and Aaron Young\inst{2} \and
Asher Elmquist \inst{2} \and Jay Taves \inst{2} \and Radu Serban \inst{2} \and
Dario Mangoni\inst{1} \and Alessandro Tasora\inst{1} \and
Dan Negrut \inst{2} }

%
\authorrunning{Simone Benatti et al.} % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
%\tocauthor{Ivar Ekeland, Roger Temam, Jeffrey Dean, David Grove,
%Craig Chambers, Kim B. Bruce, and Elisa Bertino}
%
\institute{Dipartimento di Ingegneria ed Architettura, University of Parma, Parma, Italy,\\
\email{simone.benatti@studenti.unipr.it}
\and
University of Wisconsin -- Madison, Madison, WI, USA}

\maketitle              % typeset the title of the contribution

\begin{abstract}
gym-chrono is a set of simulated environments for Deep Reinforcement Learning (DRL) extending OpenAI Gym \cite{Brockman16Gym} with robotics and autonomous driving tasks. The physics of these environments is simulated thanks to Project Chrono~\cite{chronoOverview2016}, an open-source multi physics simulation engine capable of simulating Multibody Dynamics with constraints and smooth or non-smooth contacts. The most used Deep Learning frameworks (such as PyTorch and Tensorflow ) have Python API, thus using Python to implement DRL algorithms is the most convenient option. For this reason a condition for the creation of these environments has been the development of PyChrono, a Python module consisting of the Python bindings to Project Chrono C++ API, to effectively interface the simulation capabilities of Project Chrono with Deep Learning frameworks.
\keywords{Multi-Body, Simulation, Deep Learning}
\end{abstract}
%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}

%containing also the motivation and the purpose of the study together with a brief summary of previous works on the topic
Reinforcement Learning (RL) \cite{SuttonBarto1998RL}  is a Machine Learning technique based on agent-environment interactions: at each interaction the agent performs an \textit{action} and collects the \textit{state} of the system and a \textit{reward} measuring its performance in solving some task; the goal of RL is, given the state, picking the action that maximizes the expected sum of reward, thus solving the task. In the last few years \cite{Mnih13} Deep Learning used in conjunction with RL (called Deep Reinforcement Learning, DRL) has demonstrated to be a viable approach to solve complex real-world robotics tasks \cite{openai2018HandManipulation}. DRL methods, like any other Deep Learning approach, require large dataset to optimize the Neural Networks, these dataset can by collected by sampling from real robots or through simulation, the latter being safer, cheaper and easier to set up and parallelize.

The results obtained in DRL-based controls has arisen interest in physics engine providing a Python API. The DRL community heavily relies MuJoCo \cite{todorovMujoco2012} and PyBullet \cite{bulletPhysicsEngine2020}
for robotics environments and on CARLA \cite{carlaAVsim2017} and AirSim \cite{shah2018airsim}for autonomous driving.

We provide in a single Python framework a set of reinforcement learning environments that feature:
\begin{itemize}
    \item Multibody Dynamics simulation with constraints and smooth or non-smooth contacts
    \item Deformable bodies simulation through Finite Element Analysis simulation
    \item Vehicle dynamics simulation tools
    \item Sensor simulation
\end{itemize}

\section{PyChrono}
\subsection{Interpreted Language}
Python is an interpreted language: This brings obvious upsides in terms of ease of use, but greatly limits the computational performance. Python has also limited parallelization capabilities, since each Python process is locked on a single thread (Global Interpreter Lock, also known as GIL).

The reason behind Python success in high performance computing is the interfacing with compiled libraries through \textit{Bindings}. In this scenario Python is used as high level interface.

\subsubsection{SWIG}
To create a Python \textit{Wrapper} for Chrono we rely on SWIG \cite{SWIG96} (Simplified Wrapper and Interface Generator). It generates bindings between C/C++ code and common scripting languages, such as Perl, C\# and Python, without need to modify the underlying code.

What the Python/C++ bindings do is creating hooks to call the compiled libraries, therefore while the functions are called from the Python API, the computation, under the hood, calls the same libraries as the C++ API. By doing this, the computational overhead of using Python is minimal, while getting the usage and compatibility benefits.

Besides allowing to solve the Python performance issues, Python bindings also allow to use multithreading when it is leveraged by the underlying code. This fits well into Python purpose, being an high-level programming language relying on binary libraries for computation.
In figure \ref{fig:OpenMP} we show how PyChrono can still use multiple CPU cores while being used from Python in a simulation with finite elements and contacts.

\begin{figure}[ht]
\includegraphics[width=0.75\textwidth]{Figures/OpenMPtest.png}
\caption{Speedup in computation when parallelizing the task on 2 cores.}
\label{fig:OpenMP}
\end{figure}

\subsection{Package Features}
\subsubsection{Multi Body Dynamics}
PyChrono wraps the Multi Body Dynamics simulation classes of Project Chrono, providing:
\begin{itemize}
    \item Multi Body constrained Dynamics
    \item Smooth and non-smooth contacts
    \item FEA based deformable bodies: beams, shells and 3D elements.
    \item Various solvers and timesteppers
    \item 1D shaft systems simulation
\end{itemize}
These features are provided by 3 sub-modules: \textit{core, fea }and \textit{mkl} (the latter only providing the interface to MKL PARDISO solver)
\subsubsection{Modeling and Visualization Tools}
The \textit{irrlicht} and \textit{postprocess} sub-module provide real-time 3D rendering and post processing tools (such as POV-Ray scripts for higher quality rendering).
\subsubsection{Vehicle Dynamics}
Wrapping all of Project Chrono vehicle simulation classes, PyChrono supports template-based tools for modeling wheeled and tracked vehicles. The templates include wheeled vehicle parts (steering, suspension, wheel etc.), tracked vehicle parts (track-shoe, sprocket, idler etc.), terrain, powertrain and driver. 
For some of these classes have several sub-classes that represent different mechanisms or physics detail. To mention a couple, there are 19 different suspension mechanisms (Double Wishbone, Multi Link etc) and 13 different tire models (rigid, Pacejka, deformable etc.). PyChrono vehicle also provides complete vehicle models that combined the aforementioned parts to give the user a pre-cooked vehicle model. This feature is especially welcome in contexts, such as the topic of this work, in which the focus is not on a specific vehicle model, but rather on vehicle dynamics simulation in general.
\subsubsection{Sensor Simulation}
The \textit{sensor} sub-module wraps Chrono::Sensor functionalities, being:
\begin{itemize}
    \item Support for Exteroceptive sensing, which is the sensing providing scene information. More specifically it uses NVIDIA OptiX library ray tracing to simulate camera and Lidar sensors, whose process runs on the GPU.
    \item Support for Interoceptive sensing, which is the sensing providing information from the simulation itself (although processed, such as converting positions to GPS coordinates). This family inludes inertial measurment units sensors (IMU) and GPS.
    \item Parametrization of sensors specifics (update rate, field of view, lag) and noise models for every sensor supported.
\end{itemize}
\paragraph{Sensor Data Pipeline in Python API}
Camera and Lidar sensors outputs (RGB images and point clouds respectively) memory footprint are large and in a vision-based control moving these data at each controller step might be a major bottleneck. To solve this, we cast map the data as NumPy arrays \cite{van2011numpy} using the NumPy SWIG interface. Doing so, the NumPy array class is wrapped around the raw array without instantiating new memory. This comes with no overhead, saving both time and memory.

\subsubsection{Deployment of the Python package}
Anaconda is a Python distribution that creates \textit{virtual environments} of Python to have various version of Python and Python packages on the same machine. It also manages the installation and dependencies management of Python packages through the Conda Package Manager. Through Anaconda toolchain it is possible to create packages using \textit{conda-build} and make it available to the public through the Anaconda Cloud platform.

For these reasons we identified Anaconda packages as the right tool to deploy and distribute PyChrono.
This deployment strategy makes the simulation tool more easily available for the user, and PyChrono had great benefited from this, having reached more than 3000 downloads at the time of writing this document.

\section{gym-chrono}
\subsection{Motivations and Features}
gym-chrono is a Python module that provides a set of Reinforcement Learning virtual environments that use PyChrono for physics simulation. This package is an extension of OpenAI gym package, and all its environment inherit from the OpenAI gyn environment class. Since every third-party algorithms and DRL framework is compatible with OpenAI gym environments, they can be used for gym-chrono environments as well. The same goes for environments utilities, such as the subprocess environments parallelization provided by OpenAI Baselines. Moreover, once the package is installed through Pytohn package manager, its environments can be called from anywhere in the machine by their unique ID.

\subsection{Environments}
This is a list of  the gym-chrono environments. We will refer to state as $s$ and the action as $a$
\subsubsection{Benchmarking Environments}
These 2 environments are the simplest of the set, and are used to benchmark algorithms or hyperparameters tuning, without the pretense of modeling real world robots. The first is a reversed pendulum to be balanced ($s \in \mathbb{R}^4$, $a \in \mathbb{R}$), the second a 4-legged walker called \textit{ant} ($s \in \mathbb{R}^{30}$, $a \in \mathbb{R}^8$)

\begin{figure}[ht]
 \begin{minipage}[b]{0.6\linewidth}
    \centering
    \includegraphics[width=0.75\textwidth]{ProcSci_TeX/Figures/pendulum.png}
    \caption{Pendulum}
    \label{fig:pendulum}
 \end{minipage}
 \hspace{0.4cm}
 \begin{minipage}[b]{0.6\linewidth}
    \centering
    \includegraphics[width=0.75\textwidth]{ProcSci_TeX/Figures/ant.png}
    \caption{Ant}
    \label{fig:ant}
 \end{minipage}
\end{figure}

\subsubsection{Robotics Environments}
These environments feature models of real robots created with our SolidWorks plugin. We provide a 6DOF robitc arm \ref{fig:Comau} that has to reach a random end-effector position without colliding with the floor or its own parts ($s \in \mathbb{R}^{18}$, $a \in \mathbb{R}^6$) and a hexapod \ref{fig:Hexa} that must learn to walk ($s \in \mathbb{R}^{53}$, $a \in \mathbb{R}^{18}$).

\begin{figure}[ht]
 \begin{minipage}[b]{0.6\linewidth}
    \centering
    \includegraphics[width=0.75\textwidth]{ProcSci_TeX/Figures/Comau.jpg}
    \caption{Robotic Arm Environment}
    \label{fig:Comau}
 \end{minipage}
 %\hspace{0.4cm}
 \begin{minipage}[b]{0.6\linewidth}
    \centering
    \includegraphics[width=0.75\textwidth]{ProcSci_TeX/Figures/obst_aovid.png}
    \caption{Obstacle Avoidance Environment}
    \label{fig:Obst_avoid}
 \end{minipage}
\end{figure}

\subsubsection{Autonomous Driving Environments}
These environments leverage the \textit{vehicle} and \textit{sensor} modules to reproduce autonomous driving scenarios, and are divided into 2 sub-categories:
\begin{itemize}
    \item Vision-only: \newline
    These environments observation is an RGB image, thus the cannot navigate if the direction is not encoded in the image (such as in the cone track). We provide an obstacle avoidance (figure\ref{fig:Obst_avoid}) and a cone-track environment (figure \ref{fig:Hallway})
    \item Sensor Fusion: \newline
    These environments observation is a set (tuple) of heterogeneous tensors, usually an image and a vector. Being able to encode the scene and the GPS information in the state they are capable of autonomous navigation. This family is composed of a convoy following and a off-road navigation environments.
\end{itemize}
\section{DRL algorithm implementation}
\subsection{PPO Algorithm}
To solve this environment we wrote a custom PyTorch \cite{paszke2017PyTorch} implementation of the Clipped Objective version of the Proximal Policy optimization algorithm \cite{Schulman2017PPO}.

Let the ratio $r_t(\theta) = \frac{\pi_\theta(a_t|s_t)}{\pi_{\theta_k}(a_t|s_t)} $. Then:
\begin{equation} \label{eq:PPO_clip}
    \mathcal{L}_{\theta k}^{CLIP}(\theta) = \underset{\tau\sim\pi_k}{E}\left[\sum_{t=0}^\infty\left[\min\left(r_t(\theta)\hat{A}_t^{\pi_k}, clip\left(r_t(\theta), 1-\epsilon, 1+\epsilon\right)\right)\hat{A}_t^{\pi_k}\right]\right] 
\end{equation}
With:
\begin{itemize}
    \item[$t$]{timestep. The sum to $\infty$ is a generalization, in finite horizon scenarios the sum ends a t the terminal step. The data might be collected over multiple simulation, in this case in the sum there will be several initial and terminal timesteps.}
    \item[$a_t$]{Action at timestep t}
    \item[$s_t$]{State at timestep t}
    \item[$\theta$]{Policy NN parameters}
    \item[$\pi_\theta$]{Policy, the probability distribution of actions given the state. $\pi_\theta(a_t|s_t)$ is the probability of taking the action $a_t$ in state $s_t$}
    \item[$\epsilon$]{Clipping ratio, $\sim 0.2$}
    \item[$k$]{The k-th parameter update. Before the update the parameters are $\theta_k$, and the Advantage Function is estimated on the Policy $\pi_{\theta_k}$}.
    \item[$\hat{A}_t^{\pi_k}$]{Estimate of the Advantage Function at timestep t w.r.t. the policy $\pi_{\theta_k}$}
\end{itemize}
We mentioned the Advantage Function. To define it, we first need to introduce the \textit{State Value Function} (or often only Value Function, also VF) as:
\begin{equation} \label{eq:stateVF}
    V^\pi(s_t) \quad := \quad \mathbb{ E} \left[ \sum_{l=0}^\infty \gamma^l r_{t+l} \;|\; s_t \right] 
\end{equation}
Where:
\begin{itemize}
    \item $r:{t+l}$ is the Reward at timestep $t+l$
    \item $\gamma$ is the discount factor
\end{itemize}
The VF, being the expected value of the (disconuted) sum of future reward given the current policy $\pi$, tells how much a state is \textit{good} when following the current policy.

The Advantege Function (AF) is defined as:
%\begin{equation} \label{eq:AdfFunct}
%    A^\pi(s_t, a_t) := r_t + \gamma V^\pi(s_{t+1}) - V^\pi(s_t)
%\end{equation}
\begin{equation}\label{eq:TDeq} A^{\pi,\gamma} = \mathbb{E}_{s^{t+1}}\left[r_t+\gamma V^{\pi,\gamma}(s_{t+1})- V^{\pi,\gamma}(s_t)\right] = E_{s^{t+1}}\left[\delta_t^{V^{\pi,\gamma}}\right]
\end{equation}
Being affected by a lower variance, it is more convenient to estimate the VF through a NN rather than the AF, using the unbiased estimator proposed by Schulman et al. \cite{Schulman15GAE}.

The generalized advantage estimator GAE($\gamma, \lambda)$ is defined as a $\lambda$ exponentially-weighted average:
\begin{equation}\label{eq:GAE}
\hat{A}_t^{GAE(\gamma,\lambda)} = \sum_{l=0}^\infty(\gamma\lambda)^l\delta_{t+l}^V
\end{equation}

\subsection{Continuous Control}
When using a DRL technique with a stochastic policy we must sample the action from a distribution determined by the NN. To do so, in continuous task, the typical solution is to sample the action from a multivariate gaussian distribution. The means of the distributions are the NN outputs, while the variances are an optimizable parameters. Variances are high during the early stages of the training to encourage exploration and narrows down as the policy converges.
%\begin{figure}[ht]
%\includegraphics[width=0.75\textwidth]{ProcSci_TeX/Figures/action_sampling.png}
%\caption{Action sampling from a NN output.}
%\label{fig:act_sampl}
%\end{figure}

%\subsection{NN architectures}

\subsubsection{Implementation Details and Features}
Our custom implementation of the PPO algorithm relies on PyTorch DL framework and uses Adam optimizer

Our algorithm is able to deal with vector, image and image+vector tuple inputs, given that the right NN model is called (we also provide several NN architecture to accommodate different input shapes).

\section{Results obtained}
Using the aforementioned algorithm, with different NN architecture according to the observation shape, we have been able to train the autonomous agents to solve all the task of gym-chrono, while we mention here just a couple of them for sake of brevity.

\paragraph{Hexapod}
The Hexapod walker is the gym-chrono environment featuring the highest number of actions, thus being the hardest of robotics environments. In fact, as reported in figure \ref{fig:Hexa_rew}, it took 6000 episodes to converge, being almost 3 times the number of policy updates required by the robotic arm.

\begin{figure}[ht]
 \begin{minipage}[b]{0.6\linewidth}
    \centering
    \includegraphics[width=0.75\textwidth]{ProcSci_TeX/Figures/Hexapod.jpg}
    \caption{Hexapod Environment}
    \label{fig:Hexa}
 \end{minipage}
 %\hspace{0.4cm}
 \begin{minipage}[b]{0.6\linewidth}
    \centering
    \includegraphics[width=0.75\textwidth]{ProcSci_TeX/Figures/hexa_rew.png}
    \caption{Hexapod Reward Progression}
    \label{fig:Hexa_rew}
 \end{minipage}
\end{figure}

\paragraph{Hallway cone track}
In this environment \ref{fig:Hallway} a model of a RC car has to drive in a reproduction of a real indoor hallway in a track delimited by red and green cones. The agent only controls 1 action (the steering) since the speed is feed-back controlled to keep a constant value, thus the convergence can be reached in a relatively short amount of iteration (as shown in figure \ref{fig:Hallway_rew}). This being said, this scenario proves the capability of PyChrono of dealing with large scenarios (the hallway 3D mesh) and large dataset (the 160x90 image observation is the largest of the set). In addition, this environment could be solved only after the starting position was placed in proximity of a turn. Starting on a straight, with a 0 steering angle, means starting in a local maximum that the algorithm is likely to overfit.

\begin{figure}[ht]
 \begin{minipage}[b]{0.6\linewidth}
    \centering
    \includegraphics[width=0.75\textwidth]{ProcSci_TeX/Figures/hallway.png}
    \caption{Hallway Cone Track Environment}
    \label{fig:Hallway}
 \end{minipage}
 %\hspace{0.4cm}
 \begin{minipage}[b]{0.6\linewidth}
    \centering
    \includegraphics[width=0.75\textwidth]{ProcSci_TeX/Figures/hallway_rewards.pdf}
    \caption{Hallway Cone Track Reward Progression}
    \label{fig:Hallway_rew}
 \end{minipage}
\end{figure}

\section{Conclusions}
%containing the core  findings and highlights  of the work. 
Together with a a concerted effort to improve the Python wrappers of Chrono, that lead to an Anaconda-distributed package with a good user base \cite{pyChronoCondaWebSite}, we built a set of increasingly challenging DRL environments, and used state-of-the-art continuous actions DRL algorithms to solve them. The first step has been building and solving environments such as the inverted pendulum and the 4-legged walker \cite{Benatti194Legs}. Then, we included models of real 6-DOF robots by leveraging the tools for 3D CAD parsing of Chrono \cite{Benatti19Solidworks}, and more complex, real-world robotic walkers. This feature proved to be useful by making changes in the model extremely easy to be passed to the training environment.

The latest development have been in the field of autonomous driving in off-road conditions, simulating vehicle dynamics and terrain deformation. In this context, we have been able to train autonomous agents to navigate in off-road unknown scenarios with obstacles and height irregularities to reach a target location.
These results have been made possible by the capabilities of Chrono in Vehicle and Sensor simulation, whose API allows to model vehicle and attach sensors to them, easing the creation of autonomous driving virtual environments, while providing detailed models of vehicles and terrain for accurate physical simulation.

\bibliographystyle{plain}
\bibliography{Bib/DRL, Bib/refsMBS, Bib/refsChronoSpecific, Bib/refsAutonomousVehicles}

\end{document}

### 2020-11-23 2021/ADMM_journal/author.tex
Typo corrected. Abstract fixed.

@@
 \ead{alessandro.tasora@unipr.it}
+\author{Dario Mangoni}
+\ead{dario.mangoni@unipr.it}
 \author{Simone Benatti}
 \ead{simone.benatti@studenti.unipr.it}
-\author{Dario Mangoni}
-\ead{dario.mangoni@unipr.it}
+\author{Rinaldo Garziera}
+\ead{rinaldo.garziera@unipr.it}
 \address{Universit\`a degli Studi di Parma, \\Dpt. of Engineering and Architecture\\Parco Area delle Scienze, 181/A, 43124 Parma, Italy }
@@
 \begin{abstract}
-This work discusses an efficient formulation of a geometrically exact three-dimensional beam which can be used in dynamical simulations involving large displacements, collisions and non-linear materials.
-To this end, we base our model on the shear-flexible Cosserat rod theory and we implement it in the context of Isogeometric Analysis (IGA).
-According to the IGA approach, the centerline of the beam is parameterized using splines; in our work the rotation of the section is parameterized by a spline interpolation of quaternions, and time integration of rotations is performed using the exponential map of quaternions.
-Aiming at an efficient and robust simulation of contacts, we propose the adoption of a non-smooth dynamics formulation based on differential-variational inequalities.
-The model has been implemented in an open-source physics simulation library that can simulate actuators, finite elements, rigid bodies, constraints, collisions and frictional contacts. 
-This beam model has been tested on various benchmarks in order to assess its validity in non-linear static and dynamic analysis; in all cases the model behaved consistently with theoretical results and experimental data.
-\end{abstract}
-
-\begin{keyword}
-ADMM \sep non-smooth dynamics \sep contact
-\end{keyword}
-
-\maketitle              % typeset the title of the contribution
-
-\begin{abstract}
 We propose to use the Alternating Direction Method of Multipliers (ADMM) for solving the variational inequality that arises in non-smooth contact problems in multibody simulations.
@@
 \end{keyword}
+
+\maketitle              % typeset the title of the contribution
 
@@
 \begin{align}
-\vect{v} = H^{-1}( \vect{k} + \bar{D}_\epsilon \vect{\gamma}_\epsilon)
+\vect{v} = H^{-1}( \vect{k} + D_\epsilon \vect{\gamma}_\epsilon)
 \label{eq:v_post}

### 2020-11-30 2021/ADMM_journal/author.tex
Added the supersimplified form of the preconditioned ADMM Corrected typos

@@
 \begin{itemize}
-		\item unknown $\vect{v}$ is the speed at the end of the time step $h$
-		\item unknown $\vect{\gamma}_\epsilon$ is the reaction in contacts and bilateral joints, a vector-signed measure with Lebesgue decomposition in atomic parts (impacts) and continuous parts (continuous reactions)
-    \item $H$ is a positive definite matrix containing $M$, the block-diagonal matrix of the masses and inertia tensors of the bodies; if stiff
+		\item unknown $\vect{v} \in \mathbb{R}^{n_v}$ is the speed at the end of the time step $h$
+		\item unknown $\vect{\gamma}_\epsilon \in \mathbb{R}^{n_c}$ is the reaction in contacts and bilateral joints, a vector-signed measure with Lebesgue decomposition in atomic parts (impacts) and continuous parts (continuous reactions)
+    \item $H \in \mathbb{R}^{n_v \times n_v}$ is a positive definite matrix containing $M$, 
+		the block-diagonal matrix of the masses and inertia tensors of the bodies; if stiff
 		loads are added too (ex. when finite elements come into play) it becomes a block-sparse matrix including the tangent stiffness and damping matrices,
 		for instance $H=M-{h^2}\nabla_q\vect{f}-{h}\nabla_v\vect{f}$ in case of the first-order time stepper in \cite{TasoraAnitescuCMAME10}
-    \item $D_\epsilon$ is a sparse matrix, the transpose jacobian of all constraints
-    \item $\vect{k}$ is a vector containing terms proportional to applied forces $\vect{f}$
-    \item $\vect{b}$ is a vector containing constraint stabilization terms, among other things
-    \item $\Upsilon$ is the Cartesian product of all cones of admissible constraint forces, $\Upsilon = \bigtimes_i \Upsilon_i$
-    \item if a frictional contact is added, $\Upsilon_i \subset \mathbb{R}^3$ is a second order Lorentz cone with aperture proportional to friction coefficient
-    \item if a bilateral constraint is added, $\Upsilon_i = \mathbb{R}$ and $\Upsilon^\circ_i = \{0\}$
-    \item if a unilateral constraint is added, $\Upsilon_i = \mathbb{R}^+$ and $\Upsilon^\circ_i = \mathbb{R}^-$
+    \item $D_\epsilon \in \mathbb{R}^{n_v \times n_c}$ is a sparse matrix, the transpose jacobian of all constraints
+    \item $\vect{k} \in \mathbb{R}^{n_v}$ is a vector containing terms proportional to applied forces $\vect{f}$ and to the last known speed,
+    \item $\vect{b} \in \mathbb{R}^{n_c}$ is a vector containing constraint stabilization terms and rheonomic terms,
+    \item $\Upsilon$ is the Cartesian product of all cones of admissible constraint forces, $\Upsilon = \bigtimes_j \Upsilon_j$
+    \item if a frictional contact is added, $\Upsilon_j \subset \mathbb{R}^3$ is a second order Lorentz cone with aperture proportional to friction coefficient
+    \item if a bilateral constraint is added, $\Upsilon_j = \mathbb{R}$ and $\Upsilon^\circ_j = \{0\}$
+    \item if a unilateral constraint is added, $\Upsilon_j = \mathbb{R}^+$ and $\Upsilon^\circ_j = \mathbb{R}^-$
     \item $\Upsilon^{\circ}$ is the polar cone, opposite of the dual cone, i.e. $\Upsilon^{*} = -\Upsilon^\circ$.
@@
 In order to monitor the convergence of the method, a measure of residual quantities must be introduced. 
-Given the definitions \eqref{eq:primalfeasibility1},\eqref{eq:primalfeasibility1},\eqref{eq:primalfeasibility1}, one can define a primal and a dual residual as:
+Given the definitions \eqref{eq:primalfeasibility1},\eqref{eq:dualfeasibility1},\eqref{eq:dualfeasibility2}, one can define a primal and a dual residual as:
 %
@@
 However, we still can use a
-simple heuristics consisting in forcing a small value $\rho_b = 1e-9$ for bilateral constraints, where one knows in advance that the corresponding dual variable $y_i$ would be null anyway.
+simple heuristics consisting in forcing a small value $\rho_b = \SI{1e-9}{}$ for bilateral constraints, where one knows in advance that the corresponding dual variable $y_i$ would be null anyway.
 This produces better convergence results in systems containing both contacts and many bilateral joints, such as when simulating articulated robots that grasp some objects.
@@
 Note that in sake of high performance we never build explicitly $\breve{N}$ and $\breve{\vect{r}}^T_\epsilon$: what we need is just a scaling of the saddle point matrix involved in \eqref{eq:admm_step1}, that will simply include the scaled jacobians
-$\breve{D}_\epsilon = S D_\epsilon$, the scaled compliance
+$\breve{D}_\epsilon = D_\epsilon S$, the scaled compliance
 $\breve{C}_\epsilon = S C_\epsilon S$,  the scaled term 
-$\breve{\vect{b}}_\epsilon = S \breve{\vect{b}}_\epsilon$.
+$\breve{\vect{b}}_\epsilon = S \vect{b}_\epsilon$, the scaled set
+$\breve{\Upsilon} = \left\{ \breve{\vect{\gamma}} : S \breve{\vect{\gamma}}_\epsilon \in \Upsilon  \right\}$.
+
+The scaled form of the problem would require a projection $\Pi_{\breve{\Upsilon}}$, however this can be simplified we use the same $S_{ii}$ values for each triplet corresponding to the $j$-th friction cone constraint, for uniform scaling of the $j$-th cone, thus $\Pi_{\breve{\Upsilon}}(\breve{\vect{z}}) = \Pi_{\Upsilon}(\breve{\vect{z}})$.
+More in general, for a convex set $\mathcal{C}$, again assuming a uniform scaling for each $\mathcal{C}_j$ set, one can use the simplification
+\[
+\Pi_{\breve{\mathcal{C}}}(\breve{\vect{z}}) 
+= \Pi_{\breve{\mathcal{C}}}(S^{-1}\vect{z}) 
+= S^{-1} \Pi_{\mathcal{C}}(\vect{z}) 
+= S^{-1} \Pi_{\mathcal{C}}( S \breve{\vect{z}})
+\]
 
@@
 \vect{r}_{\text{dual}}^{k+1}   &= S^{-1} \Theta^k (\breve{\vect{z}}^{k+1} - \breve{\vect{z}}^{k}) \nonumber
+\end{align}
+
+As an alternative, one can see that the preconditioned algorithm can be written with the original variables $\vect{\gamma}_\epsilon, \vect{y}, \vect{z}$ if one substitutes the scaled variables and performs some algebraic simplifications, obtaining the following steps:
+\begin{subequations}
+\begin{align}
+\begin{bmatrix}
+			 H   & D_\epsilon \\
+			 D^T & - \Theta^k S^{-2} - C_\epsilon
+			\end{bmatrix}
+			\begin{Bmatrix}
+				\vect{v}^{k+1}  \\
+			 -\vect{\gamma}_\epsilon^{k+1}
+			\end{Bmatrix}
+			=
+			\begin{Bmatrix}
+			 \vect{k} \\
+			 -\vect{b}_\epsilon + \Theta^k S^{-2} \vect{z}^k - \vect{y}^k 
+			\end{Bmatrix} \\
+	  \vect{z}^{k+1} = \Pi_\Upsilon \left(  \vect{\gamma}_\epsilon^{k+1} +(\Theta^k)^{-1} S^{2} \vect{y}^k \right ) \\
+		\vect{y}^{k+1} = \vect{y}^k + \Theta^k S^{-2} \left( \vect{\gamma}_\epsilon^{k+1} - \vect{z}^{k+1} \right)
+\end{align}
+\label{eq:admm_precond_simple}
+\end{subequations}
+%
+The method above corresponds to using the original ADMM algorithm with a non-uniform step $\Theta^k S^{-2}$, with $(S^{-2})_{ii} = {N_{ii}}$, hence it avoids the necessity of scaling all variables and matrices.
+However, when one needs to compute the residuals (for instance for termination criteria or for step scaling policies), in this simplified algorithm \eqref{eq:admm_precond_simple} those must be computed as:
+%
+\begin{align}
+\vect{r}_{\text{primal}}^{k+1} &= \left( \vect{\gamma}_\epsilon^{k+1}-\vect{z}^{k+1} \right) \nonumber \\
+\vect{r}_{\text{dual}}^{k+1}   &= \Theta^k S^{-2} (\vect{z}^{k+1} - \vect{z}^{k}) \nonumber \\
+\breve{\vect{r}}_{\text{primal}}^{k+1} &= S^{-1} \left( \vect{\gamma}_\epsilon^{k+1}-\vect{z}^{k+1} \right) \nonumber \\
+\breve{\vect{r}}_{\text{dual}}^{k+1}   &= \Theta^k S^{-1} (\vect{z}^{k+1} - \vect{z}^{k}) \nonumber
 \end{align}

### 2020-11-30 2021/NODYCON_PyChrono/author.tex
Fix path to figures

@@
     \centering
-    \includegraphics[width=0.75\textwidth]{ProcSci_TeX/Figures/pendulum.png}
+    \includegraphics[width=0.75\textwidth]{Figures/pendulum.png}
     \caption{Pendulum}
@@
     \centering
-    \includegraphics[width=0.75\textwidth]{ProcSci_TeX/Figures/ant.png}
+    \includegraphics[width=0.75\textwidth]{Figures/ant.png}
     \caption{Ant}
@@
     \centering
-    \includegraphics[width=0.75\textwidth]{ProcSci_TeX/Figures/Comau.jpg}
+    \includegraphics[width=0.75\textwidth]{Figures/Comau.jpg}
     \caption{Robotic Arm Environment}
@@
     \centering
-    \includegraphics[width=0.75\textwidth]{ProcSci_TeX/Figures/obst_aovid.png}
+    \includegraphics[width=0.75\textwidth]{Figures/obst_aovid.png}
     \caption{Obstacle Avoidance Environment}
@@
     \centering
-    \includegraphics[width=0.75\textwidth]{ProcSci_TeX/Figures/Hexapod.jpg}
+    \includegraphics[width=0.75\textwidth]{Figures/Hexapod.jpg}
     \caption{Hexapod Environment}
@@
     \centering
-    \includegraphics[width=0.75\textwidth]{ProcSci_TeX/Figures/hexa_rew.png}
+    \includegraphics[width=0.75\textwidth]{Figures/hexa_rew.png}
     \caption{Hexapod Reward Progression}
@@
     \centering
-    \includegraphics[width=0.75\textwidth]{ProcSci_TeX/Figures/hallway.png}
+    \includegraphics[width=0.75\textwidth]{Figures/hallway.png}
     \caption{Hallway Cone Track Environment}
@@
     \centering
-    \includegraphics[width=0.75\textwidth]{ProcSci_TeX/Figures/hallway_rewards.pdf}
+    \includegraphics[width=0.75\textwidth]{Figures/hallway_rewards.pdf}
     \caption{Hallway Cone Track Reward Progression}

### 2020-12-07 2021/NODYCON_nonsmooth_ADMM/author.tex
Update NODYCON paper to reflect requests from the reviewers

@@
 
+Simulation of complex systems involving multiple contacts between mechanical parts can lead to problems of non-smooth dynamics.
+In such framework, contacts and frictional effects are represented by set-valued functions, thus leading to a more complex mathematical framework respect to conventional smooth Ordinary Differential Equations (ODE) or Differential-Algebraic Equations (DAE).
+In the latter, in fact, frictional contacts are often approximated as smooth regularizations, hence introducing spring-like penalty forces that allow the adoption of typical ODE or DAE time integrators, but at the same time incurring in low stability unless very small time steps are used.
+
 Following the seminal works by Moreau 
@@
 These methods are based on a simple projection operator,  matrix-by-vector multiplications and inner products.
-Their convergence is better than for fixed points iterations, however they share with them the following issue: that they often require the building of a Delassus operator, a matrix that is easy to compute in efficient (factored and sparse) format if diagonal masses are used because these can be easily inverted, but hard to handle if finite elements are added to the problem, because they introduce stiffness and damping matrices whose inverse would be needed as well.
+Their convergence is better than for fixed points iterations, however they share with them the following issue: that they often require the building of a Delassus operator, that is a Schur complement matrix that is easy to compute in efficient (factored and sparse) format if diagonal masses are used because these can be easily inverted, but hard to handle if finite elements are added to the problem, because they introduce stiffness and damping matrices whose inverse would be needed as well.
 
@@
 
-\begin{figure}[h!]
+\begin{figure}[!tbp]
 \centering
@@
 
+\subsection{Deformable tire}
+
+\begin{figure}[!tbp]
+  \centering
+	\includegraphics[width=0.32\textwidth]{tire_5.png}
+  \includegraphics[width=0.32\textwidth]{tire_10.png}
+	%\includegraphics[width=0.32\textwidth]{tire_15.png}
+	\includegraphics[width=0.32\textwidth]{tire_20.png}
+	\hfill
+  \caption{\textsc{Test 5}.
+Large deformation of a finite element mesh colliding with a plane.
+Thanks to the non-smooth formulation, large time steps can be used (only ten steps have been used from start to end) and no penalty parameters must be used in contact points.
+Contacts between lugs and ground show zero interpenetration regardless of the stiffness of the mesh.}
+	\label{fig:tire}
+\end{figure}
+
 \section{Results and conclusion}
@@
 We performed benchmarks involving multi-body systems with contacts between multiple parts, showing that the performance of the ADMM method is capable of handling problems that would converge too slowly using conventional projected fixed point methods (see Fig.\ref{fig:t8}) or even first-order SPG spectral methods \cite{heynIJNME2013}.
+An additional benefit of ADMM, when compared with those methods, is that it does not require the computation of a Schur complement, hence it can be applied seamlessly to problems featuring a block-sparse matrix in place of the diagonal $M$ in \eqref{eq:ChronoMCCP}, such as when introducing stiffness and damping matrices for finite elements (see \ref{fig:tire}).
 

### 2021-02-16 2021/ADMM_journal/author.tex
Rephrase
Added changes requested by reviewers after revision 1. Notation section, better bibliography, corrected typos, added some sentences.

@@
 
-\documentclass[final,3p]{elsarticle}
+\documentclass[AMA,STIX1COL]{WileyNJD-v2}
+
+\articletype{Article Type}%
+
+\received{28 December 2020}
+\revised{- - -}
+\accepted{- - -}
+
+\raggedbottom
 
@@
 \usepackage{tabularx}
-\usepackage[ruled,vlined]{algorithm2e}  % ruled,vlined,linesnumbered
+%\usepackage[ruled,vlined]{algorithm2e}  % ruled,vlined,linesnumbered % NOT WORKIGN IN IJNME, they use algorithmicx
 \usepackage{array}
@@
 \usepackage{xcolor}
-\usepackage[title]{appendix}
+%\usepackage[title]{appendix}
 \usepackage{empheq}
+\usepackage{framed} % Framing content
+\usepackage{nomencl} % Nomenclature package
+\makenomenclature
+\setlength{\nomlabelwidth}{2cm}
+\setlength{\nomitemsep}{-0.1cm} % Baseline skip between items
 
@@
 \newtheorem*{definition*}{Definition}
-\newtheorem{definition}{Definition}
+%\newtheorem{definition}{Definition}
 \newtheorem*{remark*}{Remark}
-\newtheorem{remark}{Remark}
+%\newtheorem{remark}{Remark}
 \newtheorem*{theorem*}{Theorem}
-\newtheorem{theorem}{Theorem}[section]
+%\newtheorem{theorem}{Theorem}[section]
 \newtheorem*{corollary*}{Corollary}
-\newtheorem{corollary}{Corollary}[theorem]
-\newtheorem{lemma}[theorem]{Lemma}
+%\newtheorem{corollary}{Corollary}[theorem]
+%\newtheorem{lemma}[theorem]{Lemma}
 
@@
 
-\title{Solving Cone Complementarity Problems in Non-Smooth Dynamics using the Alternating Direction Method of Multipliers}
+\title{Solving Variational Inequalities and Cone Complementarity Problems in Non-Smooth Dynamics using the Alternating Direction Method of Multipliers}
 
 %\author{Alessandro Tasora \\ alessandro.tasora@unipr.it}
-\author{Alessandro Tasora}
-\ead{alessandro.tasora@unipr.it}
+\author{Alessandro Tasora*}
+
 \author{Dario Mangoni}
-\ead{dario.mangoni@unipr.it}
+
 \author{Simone Benatti}
-\ead{simone.benatti@studenti.unipr.it}
+
 \author{Rinaldo Garziera}
-\ead{rinaldo.garziera@unipr.it}
-\address{Universit\`a degli Studi di Parma, \\Dpt. of Engineering and Architecture\\Parco Area delle Scienze, 181/A, 43124 Parma, Italy }
+
+\address{\orgdiv{Dpt. of Engineering and Architecture}, \orgname{University of Parma}, \orgaddress{Parco Area delle Scienze, 181/A, 43124 Parma, \country{Italy}} }
+
+\corres{*Corresponding Author. \email{alessandro.tasora@unipr.it}}
 
@@
 
-\begin{abstract}
-We propose to use the Alternating Direction Method of Multipliers (ADMM) for solving the variational inequality that arises in non-smooth contact problems in multibody simulations.
-The ADMM method is based on simple computational primitives and offers good convergence properties especially for scenarios where loose tolerance on the precision can be accepted, thus making it attractive also for real-time applications.
-The method scales well with the problem size and require a small number of tuning parameters.
-Differently from other methods known in literature, such as fixed point iterations, it can easily handle problems that mix finite elements and rigid bodies, and it is only marginally affected by odd mass ratios or ill-posed problems.
-Another attractive feature is that it can be easily warm-started.
-\end{abstract}
-%
-\begin{keyword}
-ADMM \sep non-smooth dynamics \sep contact
-\end{keyword}
+\abstract[Summary]{
+This work presents a numerical method for the solution of variational inequalities arising in non-smooth flexible multi-body problems that involve set-valued forces.
+For the special case of hard frictional contacts, the method solves a second order cone complementarity problem.
+We ground our algorithm on the Alternating Direction Method of Multipliers (ADMM), an efficient and robust optimization method that draws on few computational primitives.
+In order to improve computational performance, we reformulated the original ADMM scheme in order to exploit the sparsity of constraint jacobians and we added optimizations such as warm starting and adaptive step scaling.
+The proposed method can be used in scenarios that pose major difficulties to other methods available in literature for complementarity in contact dynamics, namely when using very stiff finite elements and when simulating articulated mechanisms with odd mass ratios.
+The method can have applications in the fields of robotics, vehicle dynamics, virtual reality, and multi-physics simulation in general.
+}
+
+\keywords{
+Contact, friction, ADMM, non-smooth dynamics, simulation
+}
 
@@
 
-Following the seminal works by Moreau 
-\cite{mor88,Jean1992} 
-most formulations for non-smooth dynamical problems are based on Measure Differential Inclusions (MDI): just like differential inclusions they allow set-valued force laws (such as the Coulomb-Amontons dry friction model) but also generalize to the case where velocity is assumed to be a possibly discontinuous function of bounded variation in order to allow impulsive events.
-
-These problems can be solved by means of special time stepping methods that offer superior robustness and stability at the cost of solving a complementarity problem, or more in general a Variational Inequality (VI), per each time step
- \cite{acary2008numerical}. 
-In this context, unknowns to be solved are velocity measures and reaction impulses at contact points and at joints: in cases of many parts with lot of frictional contacts, the large dimension of the VI could lead to a bottleneck in the time stepping process.
-This stimulated lot of research on efficient numerical methods in the last three decades.
-
-One of the former approaches, presented in 
+Problems involving frictional contacts can be found in many engineering fields.
+Their simulation is a challenging task because the discontinuous nature of contact phenomena discourages the adoption of conventional time integration schemes, which assume continuous velocities and accelerations.
+In fact, if on the one hand it is possible to turn a discontinuous model into a smooth Ordinary Differential Equation (ODE) by approximating contact forces via spring-damper penalty forces, on the other this usually requires the adoption of extremely small time steps.
+
+In search of better numerical performance and stability, even in case of large time steps, we rather formulate the original non-smooth dynamical problem as a Measure Differential Inclusion (MDI).
+MDIs were pioneered by Moreau in the eighties, among others, aiming at a viable approach to the simulation of mechanical systems with hard contacts
+\cite{Moreau1983,moreau1988topics}. 
+Being differential inclusions, they can directly embed set-valued force laws, such as the Coulomb friction model, but they also generalize to the case where velocity is assumed to be a (possibly) discontinuous function of bounded variation, therefore they admit impulsive events.
+
+MDIs can be solved using special time stepping methods that offer high stability and robustness, however this comes at the cost of solving a Complementarity Problems (CP) per each time step \cite{acary2008numerical}.
+These CPs correspond to the broader class of Variational Inequalities (VI) \cite{Kinderleher1980,Harker1990}.
+
+In general, the solution of VIs or CPs represents the major numerical bottleneck in the entire time stepping process of the MDI, especially when dealing with many parts or many contacts.
+For instance, similar scenarios can happen when simulating packaging devices, ground-machine interaction, rock dynamics, masonry structures and granular materials, where the unknowns of the VIs (contact forces) can amount to thousands or millions.
+This challenge stimulated research on efficient numerical methods in the last three decades.
+
+Among the former strategies for solving this class of problems, we cite the idea of solving a Linear Complementarity Problem (LCP) per each time step
 \cite{StTr95}  % LCP solved via Lemke
-, was based on solving a Linear Complementarity Problem (LCP) per each time step.
-LCPs are sub cases of VIs, for whom a direct method (Lemke algorithm) does exist.
-However, direct LCP solvers offer exact solution at the cost of very expensive pivoting sub iterations that scale badly with increasing number of unknowns, and for these reason they are not much used nowadays.
-
-On the other hand, approximate but efficient methods based on fixed-point iterations became popular in the area of real-time simulators and robotics \cite{Bender2014}.
-In most cases they are based on stationary methods like Gauss-Seidell or Jacobi iterations interleaved with projections on friction cones.
-They can be used to solve the LCP or the Cone Complementarity Problem (CCP), another special case of the VI in which the multibody contact problem can be formulated.
-Attempts were made in order to parallelize them and to increase their convergence, such as over-relaxation, Krasnoselskii-Mann smoothing and warm starting, 
-\cite{massSplittingRichard2012,TasoraAnitescuCMAME10} % P-GS iteration
-but in general they perform poorly when there are odd mass ratios, and when articulated mechanisms such as robots are added to the scenarios - in those cases convergence often stalls, and if the iteration is prematurely truncated, reaction impulses are badly estimated and mechanisms fall apart or bend, and objects might interpenetrate.
-
-Another option is to solve the CCP as an optimization problem, using first-order optimization methods such as 
-the Nesterov accelerated projected gradient descend  
-\cite{hammadTOG2015}
-or the Barzilai-Borwein spectral projected gradient
-\cite{heynIJNME2013}.
-These methods are based on a simple projection operator,  matrix-by-vector multiplications and inner products.
-Their convergence is better than for fixed points iterations, however they share with them the following issue: that they often require the building of a Delassus operator, a matrix that is easy to compute in efficient (factored and sparse) format if diagonal masses are used because these can be easily inverted, but hard to handle if finite elements are added to the problem, because they introduce stiffness and damping matrices whose inverse would be needed as well.
-
-A further class of solvers is represented by non-smooth newton methods, such as the one presented in 
+.
+Once cast as a LCP, the VI problem can be solved by direct LCP solvers such as the Lemke algorithm.
+However, LCP direct solvers often lead to complex sub-iterations that do not scale well with increasing number of unknowns, hence they are not much used nowadays.
+Another drawback of casting the problem as a LCP is that they introduce a polyhedral approximation, for example the Coulomb friction cone would be approximated as a faceted pyramid, introducing a numerical source of anisotropy.
+
+At the other end of the spectrum are iterative methods that aim at high robustness and efficiency even at the expense of sacrificing accuracy, since this is often acceptable in fields like real time simulation, video games and virtual reality \cite{Bender2014}.
+In these fields, the most used approach nowadays consists in fixed point iterations, similarly to Gauss-Seidel, SOR or Jacobi stationary methods, interleaved with simple projections onto the constraint sets at each iteration
+\cite{anitescuTasora2008,massSplittingRichard2012,AleMihaiFriction2013} % P-GS iteration
+.
+The major issue of these methods is that they exhibit slow convergence when the system contains long kinematic chains or long sequences of contacts, such as when including high stacks of objects \cite{TasoraAnitescuCMAME10}.
+Their convergence is even more challenged if odd mass ratios are encountered: this often happens when simulating industrial robots, earth-moving machines, complex vehicle suspensions, car drivelines.
+In these cases the convergence might stall soon, hence if the iterations are truncated in order to meet a time budget in real-time applications, the resulting simulation can be affected by low accuracy - usually detectable in the form of objects that interpenetrate at the point of contact and mechanisms that fall apart.
+Optimizations like warm starting and sub-stepping can alleviate the drawbacks of these methods \cite{Macklin2019a}.
+
+A more advanced class of solvers is represented by non-smooth Newton methods, such as the one presented in 
 \cite{Macklin2019} % Non-smooth Newton Method
-, that assume a generic Nonlinear Complemetarity Problem (NCP), again a sub case of a VI, and represent the NCP complementarity constraints using non-smooth functions like the Fisher-Burmeister function.
-A generalized non-smooth Newton method can be used to find the zero of the functions, at the cost of solving a linear system per each iteration.
-
-Under mild assumptions, the VI can be stated as a convex CCP, that can be also cast as an optimization problem, namely a Quadratic Program (QP) with convex constraints \cite{anitescuTasora2008}.
-For such a class of optimization problems, one can use Interior Point Methods (IPMs), a class of solvers that share some similarities with the above mentioned non-smooth Newton methods - for example the solution of a saddle-point linear system is required at each iteration.
-IPMs offer the best theoretical convergence.
-However, their implementation is quite intricate, and despite the encouraging theoretical properties, in practice they do not scale well for large problems.
-Also, there are no reliable ways to warm-start them
-\cite{Mangoni2018}
-.
+, that assume a generic Nonlinear Complemetarity Problem (NCP), again a sub case of a VI, and that enforce the NCP complementarity constraints using non-smooth functions like the Fisher-Burmeister function.
+In these methods, a generalized non-smooth Newton method can be used to find the zero of the functions, but this comes at the cost of solving a linear system per each iteration.
+
+Recent efforts pushed forward the idea of casting the VI of non-smooth dynamics as a convex optimization problem, namely a Quadratic Program (QP) with convex constraints.
+This requires a simplifying assumption about the Coulomb friction being associative instead of non-associative.
+This artificial associativity of the friction model means that the relative sliding velocity in a contact point will be restricted to the dual cone of the Coulomb friction cone, whereas the original model has a more relaxed condition, assuming that such velocity could be even parallel to the surface: the practical consequence is that grazing motion cause an artificial lift-off - however this artifact can be attenuated thanks to the introduction of a stabilization term \cite{ani04}.
+In this setting, the problem is also a convex Cone Complementarity Problem (CCP).
+The convexity assumption allows the adoption of numerical methods already available for large scale optimization problems.
+In \cite{heynIJNME2013}, the Barzilai-Borwein spectral projected gradient is used to this end, and in \cite{hammadTOG2015} we used the Nesterov accelerated projected gradient descend to simulate large scale contact problems.
+These are first-order optimization methods, in the sense that they do not require the computation of the Hessian matrix: in fact they rely only on  matrix-by-vector multiplications and inner products (similarly to Krylov subspace methods in the field of linear systems), plus a fast projection operator.
+Although they offer superior convergence with respect to fixed point iterations, they are not easily applicable to problems involving finite elements.
+In fact, this class of solvers often perform the matrix-by-vector primitive over a Schur complement matrix, a product of constraint jacobians and inverse of the mass matrix.
+Since the mass matrix is often (block) diagonal, the Schur complement can be quickly factored at the beginning of the iteration, but if finite elements are used, the Schur matrix would also involve to some extent the inverse of element stiffness matrices and damping matrices, a costly and inefficient operation unless special developments are considered \cite{Francu2017}.
+Extending complementarity-based approaches in order to encompass finite elements is important because recent research has shown that complementarity-based formulations for large deformation dynamics with contacts often show superior performance when compared to penalty-based approaches \cite{Yu2020,Bozorgmehri2021}.
+
+Another class of solvers that can draw on the optimization-based approach are Interior Point Methods (IPMs).
+These solvers share some similarities with the previously mentioned non-smooth Newton methods, in that they require the solution of a saddle-point linear system at each iteration.
+IPMs offer the best theoretical convergence for the problem at hand \cite{Potra2000}.
+However, their implementation is complex, and despite the encouraging theoretical properties, in practical scenarios they do not scale well for large problems.
+One of their drawbacks, also, is that at the present state of knowledge there are no efficient ways to warm-start them \cite{Mangoni2018}.
+This led us to investigate other methods for constrained optimization problems.
 
@@
 as a practical way to solve constrained optimization problems by iterating over the minimization of two (not necessarily differentiable) objectives.
-ADMM introduces auxiliary variables and cast the original problem as the minimization of a separable objective subject to a linear constraint between the original variables and the auxiliary variables.
+ADMM introduces auxiliary variables and casts the original problem as the minimization of a separable objective subject to a linear constraint between the original variables and the auxiliary variables.
 With a proper choice of auxiliary variables, the minimization of the separable function is obtained iterating on two distinct minimization steps of lower complexity.
-The method iterates updating primal variables, auxiliary variables and dual variables similarly to a fixed point iteration: as such, they compares less favorably to IPMs if one is interested in convergence to high precision, however recent developments showed that in practical scenarios they offer superior speed, scalability and robustness, moreover their structure map well into parallel computing architectures
+The method iterates updating primal variables, auxiliary variables and dual variables similarly to a fixed point iteration: as such, they compare less favourably to IPMs if one is interested in convergence to high precision, however recent developments showed that in practical scenarios they offer superior speed, scalability and robustness
  \cite{Boyd2011}. % [ ADMM general theory and development ].
@@
 
-ADMM methods have been used in many fields, for instance computer graphics 
- \cite{Zhang2019}
-, computational fluid dynamics
- \cite{Gregson2014} % [ ADMM and fluid simulation, inverse problems ]
-, simulation of deformable structures 
- \cite{Overby2017} % [ ADMM projective dynamics hyperelastic models ]
+ADMM methods have been used in many fields, with recent and notable examples in computer graphics 
+ \cite{Zhang2019},
+computational fluid dynamics
+ \cite{Gregson2014}, % [ ADMM and fluid simulation, inverse problems ]
+simulation of deformable structures 
+ \cite{Overby2017}, % [ ADMM projective dynamics hyperelastic models ]
 and contact dynamics 
- \cite{Daviet2020} % [ ADMM thin objects frictional contact ]
-, to name a few.
+ \cite{Daviet2020,LeLidec2020}. % [ ADMM thin objects frictional contact ] % [ ADMM in appendix for differentiable dynamics ]
 
@@
  \cite{Kadkhodaie2015} %  [ A2DM2 accelerated ADMM, strongly convex ]
-a similar acceleration method has been proposed, with $\mathcal{O}(\frac{1}{k^2})$ convergence rate, but without the assumption that one of the objectives must be quadratic.
+a similar acceleration method has been proposed, with $\mathcal{O}\left(\frac{1}{k^2}\right)$ convergence rate, but without the assumption that one of the objectives must be quadratic.
 %
@@
 Recent developments also focused on the possibility of parallelizing the ADMM iterations on GPU and parallel architectures
-\cite{Schubiger2020} % [ GPU implementation of ADMM | no need to compute residuals each iteration | krylov solver each iteration ].
-.
+\cite{Schubiger2020}. % [ GPU implementation of ADMM | no need to compute residuals each iteration | krylov solver each iteration ].
 
@@
 \cite{Stellato2020} % [OSQP: an operator splitting solver for quadratic programs].
-a specialized ADMM method for quadratic problems with conic constraints has been presented: on the top of this method we develop a solver that can exploit the nature of non-smooth dynamical problems.
-An attractive property of such ADMM is that it requires the solution of a linear system per each iteration, but unlike IPMs, such linear system most often remains unchanged during the iterations, so a factorization can be reused multiple times with a great benefit in terms of speed.
-
-In our embodiment, unknowns are primal variables (velocity measures) and dual variables (impulses in contact points and joints), the convex cone constraints stem from the Coulomb friction law, and the system matrix includes terms from the mass matrices and from the tangent stiffness/damping matrices (hence it is sparse, but not necessarily diagonal).
+a specialized ADMM method for quadratic problems (OSQP) has been presented: building off this method, we develop a solver that can exploit the nature of non-smooth dynamical problems.
+An attractive property of the approach is that it requires the solution of a linear system per each iteration, but unlike IPMs, the linear system most often remains unchanged during the iterations, so a factorization can be reused multiple times with a great benefit in terms of speed.
+
+In the approach proposed herein, unknowns are primal variables (velocity measures) and dual variables (impulses in contact points and joints), the convex cone constraints stem from the Coulomb friction law, and the system matrix includes terms from the mass matrices and from the tangent stiffness/damping matrices (hence it is sparse, but not necessarily diagonal).
 
 This paper will present a model for the non-smooth dynamics, it will show how to cast it as a QP with convex conic constraint, it will present the ADMM method to solve it, it will discuss practical implementation details and then it will show benchmarks.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%% NOTE FOR THE TYPESETTING DEPARTMENT: IF THE NOMENCLATURE FRAME
+%% CONSUMES TOO MUCH SPACE IN THE ARTICLE PAGES, YOU CAN ALSO REMOVE IT. WE ARE
+%% PROPOSING IT, BUT WE CAN ALSO DO WITHOUT IT, IF WE ARE GOING OUT OF THE PAGE BUDGET.
+
+    \nomenclature[10]{ADMM}{Alternating Direction Method of Multipliers}
+    \nomenclature[10]{FADMM}{Fast Alternating Direction Method of Multipliers}
+    \nomenclature[20]{VI}{Variational Inequality}
+    \nomenclature[30]{MDI}{Measure Differential Inclusion}
+    \nomenclature[40]{CP}{Complementarity Problem}
+    \nomenclature[40]{NCP}{Nonlinear Complementarity Problem}
+    \nomenclature[50]{LCP}{Linear Complementarity Problem}
+    \nomenclature[60]{CCP}{Cone Complementarity Problem}
+    \nomenclature[70]{QP}{Quadratic Program}
+    %\nomenclature[80]{IPM}{Interior Point Method}
+
+    \begin{table*}[!t]
+       \begin{framed}
+         \printnomenclature
+       \end{framed}
+    \end{table*}
 
@@
 \begin{remark*}
-The indicator function of any closed set is lower semicontinuous.
-\end{remark*}
-
-\begin{remark*}
 The subdifferential is a set valued function $\partial f(\vect{x}): \mathbb{R}^n \rightrightarrows \mathbb{R}^n$ in general, but as a special case, if $f(\vect{x})$ is differentiable, $\partial f(\vect{x}) = \left\{ \nabla f(\vect{x}) \right\}$.
@@
 \end{align}
-%
+\end{definition}
+
+\begin{remark*}
 If $f(\vect{y})$ is the indicator function $I_{\mathcal{E}}(\vect{y})$ to a nonempty closed convex set ${\mathcal{E}}$, defined as in \eqref{eq:indicator}, then $\mathrm{prox}_{f}(\vect{y})$ is the orthogonal projection onto that set:
@@
 \label{def:projection}
-\end{definition}
+\end{remark*}
 
@@
 \begin{itemize}
-		\item unknown $\vect{v} \in \mathbb{R}^{n_v}$ is the speed at the end of the time step $h$
-		\item unknown $\vect{\gamma}_\epsilon \in \mathbb{R}^{n_c}$ is the reaction in contacts and bilateral joints, a vector-signed measure with Lebesgue decomposition in atomic parts (impacts) and continuous parts (continuous reactions)
+		\item unknown $\vect{v} \in \mathbb{R}^{n_v}$ is the speed at the end of the time step $h$,
+		\item unknown $\vect{\gamma}_\epsilon \in \mathbb{R}^{n_c}$ is the reaction in contacts and bilateral joints, a vector-signed measure with Lebesgue decomposition in atomic parts (impacts) and continuous parts (continuous reactions),
     \item $H \in \mathbb{R}^{n_v \times n_v}$ is a positive definite matrix containing $M$, 
@@
 		loads are added too (ex. when finite elements come into play) it becomes a block-sparse matrix including the tangent stiffness and damping matrices,
-		for instance $H=M-{h^2}\nabla_q\vect{f}-{h}\nabla_v\vect{f}$ in case of the first-order time stepper in \cite{TasoraAnitescuCMAME10}
-    \item $D_\epsilon \in \mathbb{R}^{n_v \times n_c}$ is a sparse matrix, the transpose jacobian of all constraints
+		for instance $H=M-{h^2}\nabla_q\vect{f}-{h}\nabla_v\vect{f}$ in case of the first-order time steppers in \cite{TasoraAnitescuCMAME10,Francu2017},
+    \item $D_\epsilon \in \mathbb{R}^{n_v \times n_c}$ is a sparse matrix, the transpose jacobian of all constraints,
     \item $\vect{k} \in \mathbb{R}^{n_v}$ is a vector containing terms proportional to applied forces $\vect{f}$ and to the last known speed,
     \item $\vect{b} \in \mathbb{R}^{n_c}$ is a vector containing constraint stabilization terms and rheonomic terms,
-    \item $\Upsilon$ is the Cartesian product of all cones of admissible constraint forces, $\Upsilon = \bigtimes_j \Upsilon_j$
-    \item if a frictional contact is added, $\Upsilon_j \subset \mathbb{R}^3$ is a second order Lorentz cone with aperture proportional to friction coefficient
-    \item if a bilateral constraint is added, $\Upsilon_j = \mathbb{R}$ and $\Upsilon^\circ_j = \{0\}$
-    \item if a unilateral constraint is added, $\Upsilon_j = \mathbb{R}^+$ and $\Upsilon^\circ_j = \mathbb{R}^-$
+    \item $\Upsilon$ is the Cartesian product of all cones of admissible constraint forces, $\Upsilon = \bigtimes_j \Upsilon_j$, in detail:
+    \item if a frictional contact is added, $\Upsilon_j \subset \mathbb{R}^3$ is a second order Lorentz cone with aperture proportional to friction coefficient,
+    \item if a bilateral constraint is added, $\Upsilon_j = \mathbb{R}$ and $\Upsilon^\circ_j = \{0\}$,
+    \item if a unilateral constraint is added, $\Upsilon_j = \mathbb{R}^+$ and $\Upsilon^\circ_j = \mathbb{R}^-$,
     \item $\Upsilon^{\circ}$ is the polar cone, opposite of the dual cone, i.e. $\Upsilon^{*} = -\Upsilon^\circ$.
@@
 %
-and in fact, the CCP \eqref{eq:ChronoCCP} can be written in the more conventional language of the KKT optimality conditions on \textit{dual variables} $\vect{y}$ multipliers, for $\vect{y}=-\vect{u}_\epsilon$, and \textit{primal variables} $\vect{\gamma}_\epsilon$:
+and in fact, the CCP \eqref{eq:ChronoCCP} can be written in the more conventional language of the Karush–Kuhn–Tucker optimality conditions on \textit{dual variables} $\vect{y}$ multipliers, for $\vect{y}=-\vect{u}_\epsilon$, and \textit{primal variables} $\vect{\gamma}_\epsilon$:
 \begin{subequations}
@@
 
-We remark that the ADMM method just makes the assumption of $\Upsilon$ being convex, so the problem can be generalized to $\vect{\gamma}_\epsilon \in \mathcal{C}$ where $\mathcal{C}$ is a generic convex set and at the same time we assume an associated
- \footnote{Not all dissipative constitutive models are associated, for example in some computational plasticity models the plastic flow might deviate respect to the normal to the yeld surface, and the Coulomb contact model itself is not associated -in fact the contact velocity $\vect{u}$ can range into a wider cone than the polar of the friction cone, but our DVI model compensates fur such effect.
+We remark that the ADMM method just makes the assumption of $\Upsilon$ being convex, so the problem can be generalized to $\vect{\gamma}_\epsilon \in \mathcal{C}$ where $\mathcal{C}$ is a generic convex set, and at the same time we assume an associated
+ \footnote{Not all dissipative constitutive models are associated.
+For example, in some computational plasticity models, the plastic flow might deviate with respect to the normal to the yield surface.
+Moreover, the Coulomb contact model itself is not associated -in fact the contact velocity $\vect{u}$ can range into a wider cone than the polar of the friction cone, but our MDI model compensates for such effect via a stabilization term.
 We will assume associated models heretofore, and we do not deal with non-associated models to avoid additional complexity for the moment.}
@@
 
-Also, in sake of highest generality, in presence of finite elements one might need to use tangent stiffness matrices $K$ and damping matrices $D$ to accommodate an implicit integration scheme for stiff elements, and this means that the Schur complement would be computed as $N=D_{\epsilon}^T H^{-1} D_{\epsilon}$.
-Here $H$ is a linear combination of $M$, $K$, $D$ (depending on the integration scheme) and in general is not diagonal anymore.
+Also, aiming at highest generality, in presence of finite elements one might need to use tangent stiffness matrices $K$ and damping matrices $D$ to accommodate an implicit integration scheme for stiff elements, and this means that the Schur complement would be computed as $N=D_{\epsilon}^T H^{-1} D_{\epsilon}$.
+Here $H$ is a linear combination of $M$, $K$, $D$ (depending on the integration scheme) and in general is not block-diagonal anymore.
 
@@
 
-This is an unconstrained convex quadratic program whose KKT optimality conditions lead to the following linear problem:
+This is an unconstrained convex quadratic program whose optimality conditions lead to the following linear problem:
 \begin{align}
@@
 %
-Note that this requires an inexpensive and parallelizable operation.
+In a more general setting, where $\vect{\gamma}_\epsilon$ contains both contact reactions and reactions in bilateral and unilateral joints, the projection operator for the $i$-th bilateral reaction is simply $\Pi_{\mathbb{R}}(p_i) = p_i$ and the projection operator for the $i$-th unilateral reaction is $\Pi_{\mathbb{R}^+}(p_i) = \max{(0,p_i)}$.
+
+Note that the entire $\Pi_\Upsilon$ projection requires an inexpensive and parallelizable operation.
 
@@
 
-We experienced that it is possible to use a $\rho$ value that changes on a per-constraint basis.
+We experimented with the idea of using a $\rho$ value that changes on a per-constraint basis.
 If one could know in advance which contact will be active at the solution, one could split $\rho_i$ values between very high or very low values for the fastest convergence, however this is not possible because the set of active constraints is known only at the solution.
@@
 \begin{align}
+\Theta^k = \Theta(\rho^k), \quad 
  \Theta_{i,j}^{k} = 
@@
 \begin{algorithm}
-\DontPrintSemicolon
-\KwIn{ Initial approximations $\vect{y}^0$, $\vect{z}^0$ }
-\KwOut{ Solution $\vect{\gamma}_\epsilon^\star$, $\vect{v}^\star$, optionally also $\vect{y}^\star$, $\vect{z}^\star$ }
-\SetKwFunction{StepAdjustPolicy}{StepAdjustPolicy}
-\While{$i \leq n_{max iters}$}{
-    $ \begin{bmatrix}
+\caption{ADMM for solving CCPs in nonsmooth dynamics}
+\label{algo:admm1}
+\hspace*{\algorithmicindent} \textbf{Input:}  { Initial approximations $\vect{y}^0$, $\vect{z}^0$ } \\
+\hspace*{\algorithmicindent} \textbf{Outputs:}  { Solution $\vect{\gamma}_\epsilon^\star$, $\vect{v}^\star$, optionally also $\vect{y}^\star$, $\vect{z}^\star$ }
+\begin{algorithmic}[1]
+\State $\Theta^0=\Theta(\rho^0)$
+\While{$k \leq n_{max iters}$}
+    \State $ \begin{bmatrix}
 			 H   & D_\epsilon \\
@@
 			 -\vect{b}_\epsilon + \Theta^k \vect{z}^k - \vect{y}^k 
-			\end{Bmatrix} $\;
-	  $\vect{z}^{k+1} = \Pi_\Upsilon \left(  \vect{\gamma}_\epsilon^{k+1} +(\Theta^k)^{-1} \vect{y}^k \right )$ \;
-		$\vect{y}^{k+1} = \vect{y}^k + \Theta^k \left( \vect{\gamma}_\epsilon^{k+1} - \vect{z}^{k+1} \right) $ \;
-		$\vect{r}_{\text{primal}}^{k+1} = \vect{\gamma}_\epsilon^{k+1}-\vect{z}^{k+1}$ \;
-		$\vect{r}_{\text{dual}}^{k+1}   = \Theta^k (\vect{z}^{k+1} - \vect{z}^{k})$ \;
-		\If{$\norm{\vect{r}_{\text{primal}}^{k+1}} < \epsilon_{\text{primal}}$ and $\norm{\vect{r}_{\text{dual}}^{k+1}} < \epsilon_{\text{dual}}$ }{
-			break \;
-		}
-		$\rho^{k+1} = $ \StepAdjustPolicy{$\rho^{k}$} \; 
-		$\rho^{k+1} \rightarrow \Theta^{k+1}$ \;
-}
+			\end{Bmatrix} $ \Comment{Solve for $\vect{v}^{k+1}$, $\vect{\gamma}_\epsilon^{k+1}$}
+	  \State $\vect{z}^{k+1} = \Pi_\Upsilon \left(  \vect{\gamma}_\epsilon^{k+1} +(\Theta^k)^{-1} \vect{y}^k \right )$  \Comment{Project onto $\Upsilon$}
+		\State $\vect{y}^{k+1} = \vect{y}^k + \Theta^k \left( \vect{\gamma}_\epsilon^{k+1} - \vect{z}^{k+1} \right) $  
+		\State $\vect{r}_{\text{primal}}^{k+1} = \vect{\gamma}_\epsilon^{k+1}-\vect{z}^{k+1}$
+		\State $\vect{r}_{\text{dual}}^{k+1}   = \Theta^k (\vect{z}^{k+1} - \vect{z}^{k})$  
+		\If{$\norm{\vect{r}_{\text{primal}}^{k+1}} < \epsilon_{\text{primal}}$ and $\norm{\vect{r}_{\text{dual}}^{k+1}} < \epsilon_{\text{dual}}$ }
+			\State \textbf{break}
+		\EndIf
+		\State $\rho^{k+1} = \Call{StepAdjustPolicy}{\rho^{k}}$
+		\State $\Theta^{k+1} = \Theta(\rho^{k+1})$
+\EndWhile
 %\Return{location}\;
-\caption{ADMM for solving CCPs in nonsmooth dynamics}
-\label{algo:admm1}
+\end{algorithmic}
 \end{algorithm}
@@
 We experienced that the convergence of Algorithm \ref{algo:admm1} is negatively affected by the presence of odd mass ratios in the mechanical system.
-The immediate effect of uneven mass ratio is a broadening of the interval of singular values of the $N$ matrix.
+The immediate effect of uneven mass ratio is a broadening of the spectrum of the $N$ matrix.
 Note that this can be also a consequence of using different measuring units for different constraint multipliers $\gamma_i$, for instance even a system with odd mass ratios can have a badly conditioned $N$ matrix if some constraint reactions are assumed in newtons, other in kilo newtons - clearly, as the choice of measuring units is arbitrary, one should find a way to make the solver as much insensitive as possible to the scaling of the constraints.
@@
 
-Note that in sake of high performance we never build explicitly $\breve{N}$ and $\breve{\vect{r}}^T_\epsilon$: what we need is just a scaling of the saddle point matrix involved in \eqref{eq:admm_step1}, that will simply include the scaled jacobians
+Note that in search of high performance we never build explicitly $\breve{N}$ and $\breve{\vect{r}}^T_\epsilon$: what we need is just a scaling of the saddle point matrix involved in \eqref{eq:admm_step1}, that will simply include the scaled jacobians
 $\breve{D}_\epsilon = D_\epsilon S$, the scaled compliance
@@
 First of all, one must develop algorithms and data structures that are able to convey information from the contact at time $t_A$ to a time $t_B = t_A+h$.
-in fact, after ADMM computes the contact forces at step $t_A$, those could be saved into the contact data structures, however those contacts would be recomputed by the collision detection engine at the next time step $t_B$: since the objects move, also the contact manifold changes, thus it is not trivial to compute which contacts are persistent between the two steps.
+In fact, after ADMM computes the contact forces at step $t_A$, those could be saved into the contact data structures, however those contacts would be recomputed by the collision detection engine at the next time step $t_B$: since the objects move, also the contact manifold changes, thus it is not trivial to compute which contacts are persistent between the two steps.
 By using some heuristics and tolerances, one can detect if some contacts at time $t_B$ are the same contacts used at time $t_A$, if so, their last reaction forces $\vect{\gamma}_\epsilon$ can be copied to the new contacts and used for the warm starting, whereas completely new contacts are initialized with $\vect{\gamma}_\epsilon = 0$.
@@
 \begin{algorithm}
-\DontPrintSemicolon
-\KwIn{ Initial approximation $\vect{\gamma}_\epsilon^0$, optional: $\vect{v}^0$}
-\KwOut{ Solution $\vect{\gamma}_\epsilon^\star$, $\vect{v}^\star$, optionally also $\vect{y}^\star$, $\vect{z}^\star$ }
-\SetKwFunction{ADMM}{ADMM}
-\If{$\vect{v}^0$ is not provided}
-{
-	$\vect{v}^0 = H^{-1}( \vect{k} + D_\epsilon \vect{\gamma}_\epsilon^0)$ \;
-}
-$\vect{y}^0 = - (D_\epsilon^T \vect{v}^0 + C \vect{\gamma}_\epsilon^0 + \vect{b}_\epsilon)$ \;
-$\vect{z}^0 = \vect{\gamma}_\epsilon^0$ \;
-$\vect{\gamma}_\epsilon^\star, \vect{v}^\star = $ \ADMM{$\vect{y}^0$,$\vect{z}^0$} \; 
 \caption{Warm-started ADMM}
 \label{algo:admm2}
+\hspace*{\algorithmicindent} \textbf{Input:} { Initial approximation $\vect{\gamma}_\epsilon^0$, optional: $\vect{v}^0$} \\
+\hspace*{\algorithmicindent} \textbf{Output:} { Solution $\vect{\gamma}_\epsilon^\star$, $\vect{v}^\star$, optionally also $\vect{y}^\star$, $\vect{z}^\star$ }
+\begin{algorithmic}[1]
+\If{$\vect{v}^0$ is not provided}
+	\State $\vect{v}^0 = H^{-1}( \vect{k} + D_\epsilon \vect{\gamma}_\epsilon^0)$ 
+\EndIf
+\State $\vect{y}^0 = - (D_\epsilon^T \vect{v}^0 + C \vect{\gamma}_\epsilon^0 + \vect{b}_\epsilon)$
+\State $\vect{z}^0 = \vect{\gamma}_\epsilon^0$
+\State $\vect{\gamma}_\epsilon^\star, \vect{v}^\star = $ \Call{ADMM}{$\vect{y}^0$,$\vect{z}^0$}
+\end{algorithmic}
 \end{algorithm}
@@
 \begin{algorithm}
-\DontPrintSemicolon
-\KwIn{ Initial approximations $\vect{\gamma}_\epsilon^0$, $\vect{y}^0$ }
-\KwOut{ Solution $\vect{\gamma}_\epsilon^\star$, $\vect{v}^\star$, optionally also $\vect{y}^\star$, $\vect{z}^\star$ }
-\SetKwFunction{StepAdjustPolicy}{StepAdjustPolicy}
-$\alpha^{0} = 1, \quad \Theta=\Theta(\rho), \quad r_{\text{comb}}^{0} = \infty$ \;
-\While{$i \leq n_{max iters}$}{
-    $\vect{z}^{k+1} = \Pi_\Upsilon \left(  \vect{\gamma}_\epsilon^{k} +(\Theta^k)^{-1} \vect{y}^k \right )$ \;
-		$ \begin{bmatrix}
+\caption{FADMM for solving CCPs in nonsmooth dynamics}
+\label{algo:fadmm1}
+\hspace*{\algorithmicindent} \textbf{Input:} { Initial approximations $\vect{\gamma}_\epsilon^0$, $\vect{y}^0$ } \\
+\hspace*{\algorithmicindent} \textbf{Output:} { Solution $\vect{\gamma}_\epsilon^\star$, $\vect{v}^\star$, optionally also $\vect{y}^\star$, $\vect{z}^\star$ }
+\begin{algorithmic}[1]
+\State $\alpha^{0} = 1, \quad \Theta=\Theta(\rho), \quad r_{\text{comb}}^{0} = \infty$
+\While{$k \leq n_{max iters}$}
+    \State $\vect{z}^{k+1} = \Pi_\Upsilon \left(  \vect{\gamma}_\epsilon^{k} +(\Theta^k)^{-1} \vect{y}^k \right )$ \Comment{Project onto $\Upsilon$}
+		\State $ \begin{bmatrix}
 			 H   & D_\epsilon \\
@@
 			 -\vect{b}_\epsilon + \Theta \vect{z}^{k+1} - \vect{y}^k 
-			\end{Bmatrix} $\;
-		$\vect{y}^{k+1} = \vect{y}^k + \Theta \left( \vect{\gamma}_\epsilon^{k+1} - \vect{z}^{k+1} \right) $ \;
-		$\vect{r}_{\text{primal}}^{k+1} = \vect{\gamma}_\epsilon^{k+1}-\vect{z}^{k+1}$ \;
-		$\vect{r}_{\text{dual}}^{k+1}   = \Theta (\vect{z}^{k+1} - \vect{z}^{k})$ \;
-
-		\If{$\norm{\vect{r}_{\text{primal}}^{k+1}} < \epsilon_{\text{primal}}$ and $\norm{\vect{r}_{\text{dual}}^{k+1}} < \epsilon_{\text{dual}}$ }{
-			break \;
-		}
-
-		$r_{\text{comb}}^{k+1}= \Theta^{-1} \norm{\vect{y}^{k+1}-\vect{y}^{k}}_2 + \Theta \norm{\vect{\gamma}_\epsilon^{k+1}-\vect{\gamma}_\epsilon^{k} }_2$
-
-		\If {$r_{\text{comb}}^{k+1} < \epsilon_r r_{\text{comb}}^{k}$ } {
-            $\alpha^{k+1} = \frac{ 1 + \sqrt{1 + 4 (\alpha^{k})^2} }{ 2 } $ \;
-            $\vect{\gamma}_\epsilon^{k+1} = \vect{\gamma}_\epsilon^{k} + \frac{\alpha^{k}-1}{\alpha^{k+1}} (\vect{\gamma}_\epsilon^{k+1} - \vect{\gamma}_\epsilon^{k}) $ \;
-            $\vect{v}^{k+1} = \vect{v}^{k} + \frac{\alpha^{k}-1}{\alpha^{k+1}} (\vect{v}^{k+1} - \vect{v}^{k}) $  \;
-						$\vect{y}^{k+1} = \vect{y}^{k} + \frac{\alpha^{k}-1}{\alpha^{k+1}} (\vect{y}^{k+1} - \vect{y}^{k}) $  \;
-		}
-    \Else{
-            $\alpha^{k+1} = 1 $ \;
-            $r_{\text{comb}}^{k+1} = \frac{1}{\epsilon_r} r_{\text{comb}}^{k} $ \;
-    }
-}
-\caption{FADMM for solving CCPs in nonsmooth dynamics}
-\label{algo:fadmm1}
+			\end{Bmatrix} $ \Comment{Solve for $\vect{v}^{k+1}$, $\vect{\gamma}_\epsilon^{k+1}$}
+		\State $\vect{y}^{k+1} = \vect{y}^k + \Theta \left( \vect{\gamma}_\epsilon^{k+1} - \vect{z}^{k+1} \right) $ 
+		\State $\vect{r}_{\text{primal}}^{k+1} = \vect{\gamma}_\epsilon^{k+1}-\vect{z}^{k+1}$ 
+		\State $\vect{r}_{\text{dual}}^{k+1}   = \Theta (\vect{z}^{k+1} - \vect{z}^{k})$ 
+		\If{$\norm{\vect{r}_{\text{primal}}^{k+1}} < \epsilon_{\text{primal}}$ and $\norm{\vect{r}_{\text{dual}}^{k+1}} < \epsilon_{\text{dual}}$ }
+			\State \textbf{break}
+		\EndIf
+
+		\State $r_{\text{comb}}^{k+1}= \Theta^{-1} \norm{\vect{y}^{k+1}-\vect{y}^{k}}_2 + \Theta \norm{\vect{\gamma}_\epsilon^{k+1}-\vect{\gamma}_\epsilon^{k} }_2$
+
+		\If {$r_{\text{comb}}^{k+1} < \epsilon_r r_{\text{comb}}^{k}$ } 
+            \State $\alpha^{k+1} = \frac{ 1 + \sqrt{1 + 4 (\alpha^{k})^2} }{ 2 } $ 
+            \State $\vect{\gamma}_\epsilon^{k+1} = \vect{\gamma}_\epsilon^{k} + \frac{\alpha^{k}-1}{\alpha^{k+1}} (\vect{\gamma}_\epsilon^{k+1} - \vect{\gamma}_\epsilon^{k}) $ 
+            \State $\vect{v}^{k+1} = \vect{v}^{k} + \frac{\alpha^{k}-1}{\alpha^{k+1}} (\vect{v}^{k+1} - \vect{v}^{k}) $ 
+						\State $\vect{y}^{k+1} = \vect{y}^{k} + \frac{\alpha^{k}-1}{\alpha^{k+1}} (\vect{y}^{k+1} - \vect{y}^{k}) $ 
+    \Else
+            \State $\alpha^{k+1} = 1 $ 
+            \State $r_{\text{comb}}^{k+1} = \frac{1}{\epsilon_r} r_{\text{comb}}^{k} $ 
+    \EndIf
+\EndWhile
+\end{algorithmic}
 \end{algorithm}
@@
 
+In literature there are many examples of iterative solvers for non-smooth dynamics, but many of them are targeting mostly computer-graphics applications, where efficiency and robustness are more important than fidelity and precision.
+For instance, many benchmarks in computer graphics involve objects with low stiffness (for example tissues, ropes, plastics), whereas we need to test our solver with structures of engineering interest, such as steel structures.
+Also, the proposed method allows an accurate retrieval of the contact stresses, something that is particularly important for mechanics-related applications and that can be neglected in the computer graphics field.
+On the basis of these requirements we designed the benchmarks that follow.
+
 All benchmarks have been computed on a Intel\textsuperscript{\textregistered} Core\textsuperscript{\texttrademark} i7-8750H CPU, clocked at \SI{2.20}{GHz}, with 6 physical cores and \SI{16}{GB} of RAM.
 The ADMM method has been implemented
-in C++ within the Chrono open-source library, using the Eigen library v.3.37 for dense and sparse linear algebra, and Intel\textsuperscript{\textregistered} MKL Pardiso as a direct solver for sparse linear systems.
+in C++ within the Chrono open-source library \cite{chronoOverview2016}, using the Eigen library v.3.37 for dense and sparse linear algebra, and Intel\textsuperscript{\textregistered} MKL Pardiso \cite{Schenk2004} as a direct solver for sparse linear systems.
 
@@
 
-We performed benchmarks involving multi-body systems with contacts between multiple parts, showing that the performance of the ADMM method is capable of handling problems that would converge too slowly using conventional projected fixed point methods or first-order SPG spectral methods \cite{heynIJNME2013}.
+We performed benchmarks involving multi-body systems with contacts between multiple parts, showing that the performance of the ADMM method is capable of handling problems that would converge too slowly using conventional projected fixed point methods or first-order spectral methods.
 
@@
 
-\bibliographystyle{spmpsci}
 \bibliography{../../bibliography/refsMBS,../../bibliography/refsOPT}

### 2021-02-16 2021/ADMM_journal/responses_to_reviewers.tex
Added responses to reviewers (revision n.1)

\documentclass[final,12pt]{article}

% For journal layout:
% \documentclass[final,1p,times]{elsarticle}
% \documentclass[final,1p,times,twocolumn]{elsarticle}
% \documentclass[final,3p,times]{elsarticle}
% \documentclass[final,3p,times,twocolumn]{elsarticle}
% \documentclass[final,5p,times]{elsarticle}
% \documentclass[final,5p,times,twocolumn]{elsarticle}

%% Use packages ---------------------------------------------
\usepackage[english]{babel}
\usepackage{amssymb} 
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{color}
\usepackage{subcaption}

%% Customization --------------------------------------------
% New commands Tasora 
\def\vect#1{{\boldsymbol{#1}}}
\def\quat#1{{\boldsymbol{#1}}}
\def\conj{\hskip0.1em^*} % the matrix as [R(a)] , with letter between two (autoresizing) square brackets
\def\amatr#1{{#1}}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}

% New commands Fusai 
\newcommand{\hl}[1]{\textcolor{red}{#1}}

% comandi per risposte a reviewers  etc.

% METTERE LA SEGUENTE LINEA COME COMMENTO, O TOGLIERLA, PER VEDERE LA LETTERA 
% DI RISPOSTA SENZA LE NOSTRE ANNOTAZIONI Giovanni:... Alessandro:.. Dario...  
% E LASCIARE SOLO COMMENTI E RISPOSTE

\def\showcomments{1}

\definecolor{mycommentcolor}{rgb}{0.23, 0.45, 0.6}
\definecolor{alecolor}{rgb}{0.0, 0.6, 0.0}

\def\reviewercomment#1{{\vskip4mm \color{mycommentcolor} \textit{#1} \vskip2mm}}

\def\alessandro#1{
\ifx\showcomments\undefined
\else
    {\color{alecolor} \small \begingroup \leftskip=1.5cm \noindent Alessandro: #1 \par \endgroup}
\fi
}

% Title
\title{ \textbf{Responses to Reviewer's  on:} \\
Solving Variational Inequalities and Cone Complementarity Problems in Non-Smooth Dynamics using the Alternating Direction Method of Multipliers}

\author{A. Tasora, D. Mangoni, S. Benatti, R. Garziera}

%% Document ----------------------------------------------
\begin{document}

\maketitle

% Sections

We would like to thank the reviewers for their thoughtful comments and
efforts towards the improvement of our manuscript. In the following document we include
our replies to their comments, where the light text color (in italic font) is used
for the original comment from the reviewers.

\section{Reviewer 1}

We thank the Reviewer for the useful comments. 
In the following we will address his requests and suggestions.

\reviewercomment{
1. The references cited for the Measure Differential Inclusion (MDI) include a CISM course and a
conference. These are not resources that are easily accessible to everyone. I would strongly
recommend that an accessible resource be cited in their place. Same holds for references 3,
10, 18, 20, 21, 35.
}

We replaced references [1] and [2] with journal articles, as suggested by the Reviewer. We kept reference [3] as is, because it is not a conference but a quite popular book. We kept also reference [10] because the ACM SIGGRAPH conferences are important and their proceedings are always accessible. The article [18] is a journal paper, available online nowadays (some difficulty to access it depends on the fact that Morrocco, the name of the author, is spelled Morroco in some databases). We left other references to conferences when proceedings are accessible and there are no alternative versions on journals.

\reviewercomment{
2. DOI’s are probably not useful for older papers that are already published. I would suggest
uniformity by mentioning DOI for all or none.
}

I suppose unnecessary DOIs will be managed later, by the typesetting department at Wiley, because in the "`Author guidelines"' web page of NME they say \textit{Where possible the DOI for the reference should be included at the end of the reference}.

\reviewercomment{
3. The authors cite Ref [3] as a source to discussion complementarity problems. But this is not a
resource again available easily and I would recommend citing a more resourceful reference.
}

We added more references to VIs and complementarity problems in that paragraph, mentioning two other popular references.

\reviewercomment{
4. I would recommend a nomenclature and abbreviation listing at the start considering that there
are several abbreviations employed in the text.
}

We added a listing of the abbreviations at the beginning of the paper.

\reviewercomment{
5. In page 3, the authors specify that the CCP and NCP approaches are not suitable for
applicability with finite elements. This is a very generic statement. I would like to highlight two
recent works where CCP has been implemented with ANCF elements and in a Finite Element
framework. These works also compare the performance with standard methods (like penalty)
that are often employed. \\
...
}

We thank the Reviewer for pointing out these recent papers. We added them to the list of references and we improved the discussion (see revised interoduction). 
We note, in passing, that the CCP methods outlined in the two references above use a Schur matrix in the form $N=D^TM^{-1}D$, hence without the introduction of tangent stiffness and damping matrices as jacobians of internal forces (something that could be needed when the stiffness of the elements is too large to be handled by the time stepping method without running into instability). This was the rationale of our paragraph: once the Schur matrix becomes too dense to be handled efficiently, one should avoid assembling it at all - and in the current paper we avoid building it, in fact. This is also discussed later, between Formula 31 and Theorem 1.

\reviewercomment{
6. From the introduction and literature review, the aim of the authors is unclear. Is this paper
oriented towards computer graphics or mechanics-related applications? Most of the
discussions pertaining to the available literature is aimed at computer graphics applications.
While non-penetration is an important condition, accurate retrieval of the contact stresses is
also particularly important for mechanics-related applications. This is not discussed in the
literature review nor in the results.
}

The paper is oriented more towards mechanical application, indeed. The fact that literature is mostly targeting computer graphics depends on the fact that complementarity-based approaches are very robust - a quality that triggered lot of interest in the computer graphics community in the last decade. At the other hand, the CG community is often not interested in results that are precise at an engineering level (as they can accept visually-plausible results with low fidelity), so this is stimulating our research on methods that could satisfy the needs of both words: the need of stability and robustness, from the CG field, and the need of precision from the engineering field. Another point that facilitates the problem in the CG area is the fact that CG problems with contacts and deformable parts often involve jelly-like objects with low stiffness (rods, hairs, tissues, flesh), whereas in the engineering area we often deal with steel, aluminium and other structures with high stiffness. We added a discussion at the beginning of the Benchmark section.

\reviewercomment{
7. The “Remark” before and after Definition 6 is the same. Is there a particular reason for the
repetition?
}

Thank you for noticing this. We removed the repetition.

\section{Reviewer 2}

\reviewercomment{
The manuscript focuses on accelerating the solution of the frictional contact problem for rigid and flexible body dynamics. The authors work with an established way of modeling the frictional contact problem, and subsequently carrying out its time discretization. The resulting CCP is solved by means of a new technique, which draws on work done in optimization in relation to the Alternating Direction Method of Multipliers method.
}

We thank the Reviewer for his notes, that are helping us to increase the quality of the paper. Also we thank the Reviewer because of the many stylistic improvements that he listed, about grammar, syntax and style.

 We address his requests and comments in the following paragraphs:

\reviewercomment{
- without adding $C_\epsilon$ in Eq. 33(a), would this approach lead to a unique set of frictional contact forces? Or is there a possibility of finding an infinite number of compatible solutions? Is adding $C_\epsilon$ eliminating this possibility?
}

Correct, the $C_\epsilon$ has an effect toward the regularization of the system, therefore a nonzero $C_\epsilon$ guarantees that there is a unique set of solutions (the mechanical equivalence is that there are both normal and tangential springs at each contact, thus eliminating the redundancy of forces that would arise even in very simple problems like a three leg table over a flat plane if there is static friction). For zero $C_\epsilon$, one cannot rule out the case that $\rho$ becomes close to zero because of the $\rho$ automatic adjustment during the iteration, and in that case one could approach singularity (note that this would also correspond to the pivot-breakdown of the solver if a direct solver is used).

\reviewercomment{
- the CCP approach advocated here has some numerical artifacts, e.g., if a brick moves really fast on ground, there will be a lifting force that reduces the normal force at the brick-ground interface. Is the ADMM approach advocated here good enough to deal with other frictional contact models that lead to optimization problems whose solution eschews the said lift-off artifact?
}

This is an interesting point. The CCP approach we refer to, is based on a convexified model where one can experience the "lifting" effect. \footnote{We note, in passing, that this lift off is balanced by a stabilization coefficient that makes the sliding object to float with a gap of height $h\approxeq v \mu h$, where $v$ is tangential speed, $\mu$ is friction coeff. and $h$ is time step, so it is negligible for small time steps or small speeds.} The proper way to get rid of the lifting effect would be to use the original non-convexified complementarity problem. Using the De Saxce bipotential formulation of the generic non-associated Coulomb friction, CCP of Eq.13 looses the friendly affine structure of the 
$\vect{u}_\epsilon =  N \vect{\gamma}_\epsilon + \vect{r}_\epsilon$ 
term, and one would use rather 
$\vect{u}_\epsilon =   N \vect{\gamma}_\epsilon + \vect{r}_\epsilon + \tilde{\vect{u}}_\epsilon(\vect{\gamma}_\epsilon)$ where $\tilde{\vect{u}}_\epsilon(\vect{\gamma}_\epsilon)$ is a term that depends quadratically on the velocity of tangential sliding. However, this done, the resulting CCP 
$-\Upsilon^{\circ} \ni \vect{u}_\epsilon \bot \vect{\gamma}_\epsilon \in \Upsilon$ would not lead any more to a convex optimization problem. Although the paper [31] (Anderson Acceleration etc., Ouyang et al. 2020) speculates that ADMM could solve non-convex problems, this has been proved only empirically and for limited cases, so we prefer not to discuss the non-convex case in our current manuscript: it would be an interesting extension though, and we thank the Reviewer for the suggestion.

\reviewercomment{
Suggested list of corrections, small improvements:
- Abstract: “multi-flexible-body problems” – perhaps “flexible multi-body problems”?
}

$\rightarrow$ Corrected.

\reviewercomment{
- Abstract: “In sake of computational performance” – “In order to improve computational performance”?
}

$\rightarrow$ Corrected.

\reviewercomment{
- P.1.: “In sake of better numerical performance and stability” – “In search of better numerical performance and stability”. NOTE: the authors should revisit the use of “in sake” since it’s not a proper language construct. It shows up in several places in the manuscript
}

$\rightarrow$ Corrected, also in other places. Thank you for helping us to improve the language of the manuscript.

\reviewercomment{
- P2: “stimulated many researches on” – “stimulated research on” ?
}

$\rightarrow$ Corrected.

\reviewercomment{
- P2: “algorithm, however” – maybe start a new sentence “algorithm. However,…”
}

$\rightarrow$ Corrected.

\reviewercomment{
- P2: “sub iterations” – “sub-iterations”
}

$\rightarrow$ Corrected.

\reviewercomment{
- P2: “is even more critical if odd mass” – “is even more challenged if odd mass”
}

$\rightarrow$ Corrected.

\reviewercomment{
- P2: “happen when simulating” – “happens when simulating”
}

$\rightarrow$ Corrected.

\reviewercomment{
- P2: “in form” – “in the form”
}

$\rightarrow$ Corrected.

\reviewercomment{
- P2: “newton methods” – “Newton methods”
}

$\rightarrow$ Corrected.

\reviewercomment{
- P2: “This requires a simplifying assumption about the Coulomb friction being associative instead of non-associative” – can you please explain what this statement means, perhaps even in one sentence in less, in the manuscript?
}

Thank you for the suggestion. We added two or three lines explaining this point:  
\textit{This artificial associativity of the friction model means that the relative sliding velocity in a contact point will be restricted to the dual cone of the Coulomb friction cone, whereas the original model has a more relaxed condition, assuming that such velocity could be even parallel to the surface: the practical consequence is that grazing motion cause an artificial lift-off - however this artifact can be attenuated thanks to the introduction of a stabilization term [cite{ani04}]}

\reviewercomment{
- P2: “offer a superior convergence respect” – “offer superior convergence with respect”
}

$\rightarrow$ Corrected.

\reviewercomment{
- P2: “in fact they require” – “in that they require”
}

$\rightarrow$ Corrected.

\reviewercomment{
- P2: “and cast the original” – “and casts the original”
}

$\rightarrow$ Corrected.

\reviewercomment{
- P3: “they compares”
}

$\rightarrow$ Corrected.

\reviewercomment{
- P3: “on the top of this method we” – “building off this method, we”
}

$\rightarrow$ Corrected.

\reviewercomment{
- P3: “of such ADMM is that” – “of the approach is that”
}

$\rightarrow$ Corrected.

\reviewercomment{
- P3: “such linear system most often” – “the linear system most often”
}

$\rightarrow$ Corrected.

\reviewercomment{
- P3: “In our embodiment” – “In the approach proposed herein”
}

$\rightarrow$ Corrected.\\

\reviewercomment{
- P4: “Remark. The indicator function of any closed set is lower semicontinuous…” – this shows up twice
}

$\rightarrow$ Corrected.

\reviewercomment{
- P4 (footnote): “the plastic flow might deviate respect to the” – “the plastic flow might deviate with respect to the”
}

$\rightarrow$ Corrected.

\reviewercomment{
- P9: “We experienced that it is possible to use a” – “We experimented with the idea of using a”
}

$\rightarrow$ Corrected.

\reviewercomment{
- P11: “in fact, after ADMM computes” – “In fact, after ADMM computes”
}

$\rightarrow$ Corrected.

\reviewercomment{
- P12: “Chrono open-source library” – provide reference to this library. Also to Eigen and Pardiso
}

Two references to papers about the mentioned libraries have been added.

\vskip15mm

Best regards,

\vskip10mm

the Authors

\end{document}
\endinput
%% End document -----------------------------------------

### 2021-02-17 2021/ADMM_journal/responses_to_reviewers.tex
Last typos and changes to V2. Also, added 3 additional papers, one is about a similar method from 2007 that I discovered today.

@@
 \definecolor{alecolor}{rgb}{0.0, 0.6, 0.0}
+\definecolor{blucolor}{rgb}{0.0, 0.0, 0.78}
 
@@
 efforts towards the improvement of our manuscript. In the following document we include
-our replies to their comments, where the light text color (in italic font) is used
+our replies to their comments, where the {\color{mycommentcolor} light text color} (in italic font) is used
 for the original comment from the reviewers.
+
+In the PDF of the revised manuscript, changed text is typed in {\color{blucolor} blue color} for easier comparison with the first version.
 

### 2021-02-17 2021/ADMM_journal/author_v1.tex
old revision for reference

--- 2021-02-16 2021/ADMM_journal/author.tex
@@
 \usepackage{empheq}
-\usepackage{framed} % Framing content
-\usepackage{nomencl} % Nomenclature package
-\makenomenclature
-\setlength{\nomlabelwidth}{2cm}
-\setlength{\nomitemsep}{-0.1cm} % Baseline skip between items
 
@@
 \abstract[Summary]{
-This work presents a numerical method for the solution of variational inequalities arising in non-smooth flexible multi-body problems that involve set-valued forces.
+This work presents a numerical method for the solution of variational inequalities arising in non-smooth multi-flexible-body problems that involve set-valued forces.
 For the special case of hard frictional contacts, the method solves a second order cone complementarity problem.
 We ground our algorithm on the Alternating Direction Method of Multipliers (ADMM), an efficient and robust optimization method that draws on few computational primitives.
-In order to improve computational performance, we reformulated the original ADMM scheme in order to exploit the sparsity of constraint jacobians and we added optimizations such as warm starting and adaptive step scaling.
+In sake of computational performance, we reformulated the original ADMM scheme in order to exploit the sparsity of constraint jacobians and we added optimizations such as warm starting and adaptive step scaling.
 The proposed method can be used in scenarios that pose major difficulties to other methods available in literature for complementarity in contact dynamics, namely when using very stiff finite elements and when simulating articulated mechanisms with odd mass ratios.
@@
 
-In search of better numerical performance and stability, even in case of large time steps, we rather formulate the original non-smooth dynamical problem as a Measure Differential Inclusion (MDI).
-MDIs were pioneered by Moreau in the eighties, among others, aiming at a viable approach to the simulation of mechanical systems with hard contacts
-\cite{Moreau1983,moreau1988topics}. 
+In sake of better numerical performance and stability, even in case of large time steps, we rather formulate the original non-smooth dynamical problem as a Measure Differential Inclusion (MDI).
+MDIs were pioneered by Moreau in the eighties, among others, in sake of a viable approach to the simulation of mechanical systems with hard contacts
+\cite{mor88,Jean1992}. 
 Being differential inclusions, they can directly embed set-valued force laws, such as the Coulomb friction model, but they also generalize to the case where velocity is assumed to be a (possibly) discontinuous function of bounded variation, therefore they admit impulsive events.
 
-MDIs can be solved using special time stepping methods that offer high stability and robustness, however this comes at the cost of solving a Complementarity Problems (CP) per each time step \cite{acary2008numerical}.
-These CPs correspond to the broader class of Variational Inequalities (VI) \cite{Kinderleher1980,Harker1990}.
+MDIs can be solved using special time stepping methods that offer high stability and robustness, however this comes at the cost of solving a  Variational Inequality (VI) per each time step.
+In many cases, VIs correspond to Complementarity Problems (CP)
+ \cite{acary2008numerical}.
 
@@
 For instance, similar scenarios can happen when simulating packaging devices, ground-machine interaction, rock dynamics, masonry structures and granular materials, where the unknowns of the VIs (contact forces) can amount to thousands or millions.
-This challenge stimulated research on efficient numerical methods in the last three decades.
+This challenge stimulated many researches on efficient numerical methods in the last three decades.
 
@@
 .
-Once cast as a LCP, the VI problem can be solved by direct LCP solvers such as the Lemke algorithm.
-However, LCP direct solvers often lead to complex sub-iterations that do not scale well with increasing number of unknowns, hence they are not much used nowadays.
+Once cast as a LCP, the VI problem can be solved by direct LCP solvers such as the Lemke algorithm, however LCP direct solvers often lead to complex sub iterations that do not scale well with increasing number of unknowns, hence they are not much used nowadays.
 Another drawback of casting the problem as a LCP is that they introduce a polyhedral approximation, for example the Coulomb friction cone would be approximated as a faceted pyramid, introducing a numerical source of anisotropy.
@@
 The major issue of these methods is that they exhibit slow convergence when the system contains long kinematic chains or long sequences of contacts, such as when including high stacks of objects \cite{TasoraAnitescuCMAME10}.
-Their convergence is even more challenged if odd mass ratios are encountered: this often happens when simulating industrial robots, earth-moving machines, complex vehicle suspensions, car drivelines.
-In these cases the convergence might stall soon, hence if the iterations are truncated in order to meet a time budget in real-time applications, the resulting simulation can be affected by low accuracy - usually detectable in the form of objects that interpenetrate at the point of contact and mechanisms that fall apart.
+Their convergence is even more critical if odd mass ratios are encountered: this often happen when simulating industrial robots, earth-moving machines, complex vehicle suspensions, car drivelines.
+In these cases the convergence might stall soon, hence if the iterations are truncated in order to meet a time budget in real-time applications, the resulting simulation can be affected by low accuracy - usually detectable in form of objects that interpenetrate at the point of contact and mechanisms that fall apart.
 Optimizations like warm starting and sub-stepping can alleviate the drawbacks of these methods \cite{Macklin2019a}.
 
-A more advanced class of solvers is represented by non-smooth Newton methods, such as the one presented in 
+A more advanced class of solvers is represented by non-smooth newton methods, such as the one presented in 
 \cite{Macklin2019} % Non-smooth Newton Method
@@
 Recent efforts pushed forward the idea of casting the VI of non-smooth dynamics as a convex optimization problem, namely a Quadratic Program (QP) with convex constraints.
-This requires a simplifying assumption about the Coulomb friction being associative instead of non-associative.
-This artificial associativity of the friction model means that the relative sliding velocity in a contact point will be restricted to the dual cone of the Coulomb friction cone, whereas the original model has a more relaxed condition, assuming that such velocity could be even parallel to the surface: the practical consequence is that grazing motion cause an artificial lift-off - however this artifact can be attenuated thanks to the introduction of a stabilization term \cite{ani04}.
+This requires a simplifying assumption about the Coulomb friction being associative instead of non-associative, a simplification whose artifacts can be attenuated thanks to the introduction of a stabilization term \cite{ani04}.
 In this setting, the problem is also a convex Cone Complementarity Problem (CCP).
@@
 These are first-order optimization methods, in the sense that they do not require the computation of the Hessian matrix: in fact they rely only on  matrix-by-vector multiplications and inner products (similarly to Krylov subspace methods in the field of linear systems), plus a fast projection operator.
-Although they offer superior convergence with respect to fixed point iterations, they are not easily applicable to problems involving finite elements.
+Although they offer a superior convergence respect to fixed point iterations, they are not easily applicable to problems involving finite elements.
 In fact, this class of solvers often perform the matrix-by-vector primitive over a Schur complement matrix, a product of constraint jacobians and inverse of the mass matrix.
 Since the mass matrix is often (block) diagonal, the Schur complement can be quickly factored at the beginning of the iteration, but if finite elements are used, the Schur matrix would also involve to some extent the inverse of element stiffness matrices and damping matrices, a costly and inefficient operation unless special developments are considered \cite{Francu2017}.
-Extending complementarity-based approaches in order to encompass finite elements is important because recent research has shown that complementarity-based formulations for large deformation dynamics with contacts often show superior performance when compared to penalty-based approaches \cite{Yu2020,Bozorgmehri2021}.
 
 Another class of solvers that can draw on the optimization-based approach are Interior Point Methods (IPMs).
-These solvers share some similarities with the previously mentioned non-smooth Newton methods, in that they require the solution of a saddle-point linear system at each iteration.
+These solvers share some similarities with the previously mentioned non-smooth Newton methods, in fact they require the solution of a saddle-point linear system at each iteration.
 IPMs offer the best theoretical convergence for the problem at hand \cite{Potra2000}.
@@
 as a practical way to solve constrained optimization problems by iterating over the minimization of two (not necessarily differentiable) objectives.
-ADMM introduces auxiliary variables and casts the original problem as the minimization of a separable objective subject to a linear constraint between the original variables and the auxiliary variables.
+ADMM introduces auxiliary variables and cast the original problem as the minimization of a separable objective subject to a linear constraint between the original variables and the auxiliary variables.
 With a proper choice of auxiliary variables, the minimization of the separable function is obtained iterating on two distinct minimization steps of lower complexity.
-The method iterates updating primal variables, auxiliary variables and dual variables similarly to a fixed point iteration: as such, they compare less favourably to IPMs if one is interested in convergence to high precision, however recent developments showed that in practical scenarios they offer superior speed, scalability and robustness
+The method iterates updating primal variables, auxiliary variables and dual variables similarly to a fixed point iteration: as such, they compares less favourably to IPMs if one is interested in convergence to high precision, however recent developments showed that in practical scenarios they offer superior speed, scalability and robustness
  \cite{Boyd2011}. % [ ADMM general theory and development ].
@@
  \cite{Kadkhodaie2015} %  [ A2DM2 accelerated ADMM, strongly convex ]
-a similar acceleration method has been proposed, with $\mathcal{O}\left(\frac{1}{k^2}\right)$ convergence rate, but without the assumption that one of the objectives must be quadratic.
+a similar acceleration method has been proposed, with $\mathcal{O}(\frac{1}{k^2})$ convergence rate, but without the assumption that one of the objectives must be quadratic.
 %
@@
 \cite{Stellato2020} % [OSQP: an operator splitting solver for quadratic programs].
-a specialized ADMM method for quadratic problems (OSQP) has been presented: building off this method, we develop a solver that can exploit the nature of non-smooth dynamical problems.
-An attractive property of the approach is that it requires the solution of a linear system per each iteration, but unlike IPMs, the linear system most often remains unchanged during the iterations, so a factorization can be reused multiple times with a great benefit in terms of speed.
-
-In the approach proposed herein, unknowns are primal variables (velocity measures) and dual variables (impulses in contact points and joints), the convex cone constraints stem from the Coulomb friction law, and the system matrix includes terms from the mass matrices and from the tangent stiffness/damping matrices (hence it is sparse, but not necessarily diagonal).
+a specialized ADMM method for quadratic problems (OSQP) has been presented: on the top of this method we develop a solver that can exploit the nature of non-smooth dynamical problems.
+An attractive property of such ADMM is that it requires the solution of a linear system per each iteration, but unlike IPMs, such linear system most often remains unchanged during the iterations, so a factorization can be reused multiple times with a great benefit in terms of speed.
+
+In our embodiment, unknowns are primal variables (velocity measures) and dual variables (impulses in contact points and joints), the convex cone constraints stem from the Coulomb friction law, and the system matrix includes terms from the mass matrices and from the tangent stiffness/damping matrices (hence it is sparse, but not necessarily diagonal).
 
 This paper will present a model for the non-smooth dynamics, it will show how to cast it as a QP with convex conic constraint, it will present the ADMM method to solve it, it will discuss practical implementation details and then it will show benchmarks.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%% NOTE FOR THE TYPESETTING DEPARTMENT: IF THE NOMENCLATURE FRAME
-%% CONSUMES TOO MUCH SPACE IN THE ARTICLE PAGES, YOU CAN ALSO REMOVE IT. WE ARE
-%% PROPOSING IT, BUT WE CAN ALSO DO WITHOUT IT, IF WE ARE GOING OUT OF THE PAGE BUDGET.
-
-    \nomenclature[10]{ADMM}{Alternating Direction Method of Multipliers}
-    \nomenclature[10]{FADMM}{Fast Alternating Direction Method of Multipliers}
-    \nomenclature[20]{VI}{Variational Inequality}
-    \nomenclature[30]{MDI}{Measure Differential Inclusion}
-    \nomenclature[40]{CP}{Complementarity Problem}
-    \nomenclature[40]{NCP}{Nonlinear Complementarity Problem}
-    \nomenclature[50]{LCP}{Linear Complementarity Problem}
-    \nomenclature[60]{CCP}{Cone Complementarity Problem}
-    \nomenclature[70]{QP}{Quadratic Program}
-    %\nomenclature[80]{IPM}{Interior Point Method}
-
-    \begin{table*}[!t]
-       \begin{framed}
-         \printnomenclature
-       \end{framed}
-    \end{table*}
 
@@
 \end{definition}
+
+\begin{remark*}
+The indicator function of any closed set is lower semicontinuous.
+\end{remark*}
 
@@
  \footnote{Not all dissipative constitutive models are associated.
-For example, in some computational plasticity models, the plastic flow might deviate with respect to the normal to the yield surface.
+For example, in some computational plasticity models, the plastic flow might deviate respect to the normal to the yield surface.
 Moreover, the Coulomb contact model itself is not associated -in fact the contact velocity $\vect{u}$ can range into a wider cone than the polar of the friction cone, but our MDI model compensates for such effect via a stabilization term.
@@
 
-Also, aiming at highest generality, in presence of finite elements one might need to use tangent stiffness matrices $K$ and damping matrices $D$ to accommodate an implicit integration scheme for stiff elements, and this means that the Schur complement would be computed as $N=D_{\epsilon}^T H^{-1} D_{\epsilon}$.
+Also, in sake of highest generality, in presence of finite elements one might need to use tangent stiffness matrices $K$ and damping matrices $D$ to accommodate an implicit integration scheme for stiff elements, and this means that the Schur complement would be computed as $N=D_{\epsilon}^T H^{-1} D_{\epsilon}$.
 Here $H$ is a linear combination of $M$, $K$, $D$ (depending on the integration scheme) and in general is not block-diagonal anymore.
@@
 
-We experimented with the idea of using a $\rho$ value that changes on a per-constraint basis.
+We experienced that it is possible to use a $\rho$ value that changes on a per-constraint basis.
 If one could know in advance which contact will be active at the solution, one could split $\rho_i$ values between very high or very low values for the fastest convergence, however this is not possible because the set of active constraints is known only at the solution.
@@
 
-Note that in search of high performance we never build explicitly $\breve{N}$ and $\breve{\vect{r}}^T_\epsilon$: what we need is just a scaling of the saddle point matrix involved in \eqref{eq:admm_step1}, that will simply include the scaled jacobians
+Note that in sake of high performance we never build explicitly $\breve{N}$ and $\breve{\vect{r}}^T_\epsilon$: what we need is just a scaling of the saddle point matrix involved in \eqref{eq:admm_step1}, that will simply include the scaled jacobians
 $\breve{D}_\epsilon = D_\epsilon S$, the scaled compliance
@@
 First of all, one must develop algorithms and data structures that are able to convey information from the contact at time $t_A$ to a time $t_B = t_A+h$.
-In fact, after ADMM computes the contact forces at step $t_A$, those could be saved into the contact data structures, however those contacts would be recomputed by the collision detection engine at the next time step $t_B$: since the objects move, also the contact manifold changes, thus it is not trivial to compute which contacts are persistent between the two steps.
+in fact, after ADMM computes the contact forces at step $t_A$, those could be saved into the contact data structures, however those contacts would be recomputed by the collision detection engine at the next time step $t_B$: since the objects move, also the contact manifold changes, thus it is not trivial to compute which contacts are persistent between the two steps.
 By using some heuristics and tolerances, one can detect if some contacts at time $t_B$ are the same contacts used at time $t_A$, if so, their last reaction forces $\vect{\gamma}_\epsilon$ can be copied to the new contacts and used for the warm starting, whereas completely new contacts are initialized with $\vect{\gamma}_\epsilon = 0$.
@@
 
-In literature there are many examples of iterative solvers for non-smooth dynamics, but many of them are targeting mostly computer-graphics applications, where efficiency and robustness are more important than fidelity and precision.
-For instance, many benchmarks in computer graphics involve objects with low stiffness (for example tissues, ropes, plastics), whereas we need to test our solver with structures of engineering interest, such as steel structures.
-Also, the proposed method allows an accurate retrieval of the contact stresses, something that is particularly important for mechanics-related applications and that can be neglected in the computer graphics field.
-On the basis of these requirements we designed the benchmarks that follow.
-
 All benchmarks have been computed on a Intel\textsuperscript{\textregistered} Core\textsuperscript{\texttrademark} i7-8750H CPU, clocked at \SI{2.20}{GHz}, with 6 physical cores and \SI{16}{GB} of RAM.
 The ADMM method has been implemented
-in C++ within the Chrono open-source library \cite{chronoOverview2016}, using the Eigen library v.3.37 for dense and sparse linear algebra, and Intel\textsuperscript{\textregistered} MKL Pardiso \cite{Schenk2004} as a direct solver for sparse linear systems.
+in C++ within the Chrono open-source library, using the Eigen library v.3.37 for dense and sparse linear algebra, and Intel\textsuperscript{\textregistered} MKL Pardiso as a direct solver for sparse linear systems.
 
@@
 
+%\bibliographystyle{spmpsci}
 \bibliography{../../bibliography/refsMBS,../../bibliography/refsOPT}

### 2021-03-02 2021/MMT_interpolation/PaperQuaternionInterpolation.tex
Add the interpolation Legnai paper for MMT

\section{Introduction}
\label{sec:introduction}
In many theoretical and application fields there is the need to interpolate between a given set of configurations in 3D space, in order to obtain a smooth motion of a rigid frame. This is a typical problem of industrial robotics (for manipulation, welding, or inspection) \cite{li2009research} \cite{neubauer2015smooth}, aerospace (for path planning) \cite{yongguo2010trajectory} and computer graphics areas (for camera motions) \cite{shoemake1985animating}, but it also has applications in the field of computational solid mechanics (see, for example, \cite{romero2004interpolation}, \cite{ghosh2008consistent} and \cite{bauchau2014interpolation}). It is known that the motion of a rigid body in 3D space is representable by the combination of a translation and a rotation relative to a reference frame: motion interpolation is therefore the act of producing two continuous curves (one in $\mathbb{R}^{3}$ and the other in $\mathsf{SO}(3)$) from a set of \emph{key frames}. The problem of spatial interpolation in Euclidean $\mathbb{R}^{3}$ space has been thoroughly studied and many operative algorithms are available, spacing from simple linear interpolation to smooth B\'ezier, B-Spline and NURBS curves. On the other hand, the interpolation of orientations still poses some difficulties, due to the peculiar characteristics of rotations in 3D space.

As a matter of fact, the orientation of solid objects is mathematically representable in many ways: rotation matrices, axis-angle formalism, Rodrigues vectors and quaternions are only part of a numerous list of possibilities. Each of these representation offers both advantages and drawbacks, making them individually suitable for a specific application and less for another. For example, if we parametrize a rotation matrix through Cardan angles (both in the intrinsic Euler $Z-X^{'}-Z^{''}$ form or in the extrinsic Tait-Bryan $X-Y-Z$ one), we obtain a straightforward and intuitive method that allows us to combine successive rotations with simple linear algebra operations (see \ref{app:RotatMatrix} for more information). However, a major problem of this approach is that any adopted set of angles will have at least one configuration leading to singularity, making such choice unfit for generic applications.

A better strategy is to represent rotations by the means of unit quaternions \cite{hamilton1844ii}. These mathematical objects possess four components and are usually expressed as hyper-complex numbers in the form $\quat{q} = q_0 + q_{1} i + q_{2} j + q_{3} k$ (refer to \ref{app:Quaternion} for a deeper description). Although being less intuitive and still affected by some drawbacks (e.g. the same orientation may be expressed by two different and opposite quaternions), this formulation provides consistent advantages in the field of rotation interpolation. In fact, in order to represent a valid rotation, it is sufficient for a quaternion to have a unit norm, thus lying on the surface of a unitary 3-sphere (hypersphere). On the other hand, a rotation matrix must be orthogonal with determinant equal to $+1$: from a computational point of view, the normalization of an array is much faster than the one of a 3x3 matrix (with Gram-Schmidt or similar algorithms). In addition, quaternions do not suffer of singularity problems and they only deal with four numbers instead of nine, saving valuable computer memory.

It is therefore apparent that the quaternion representation of orientations is preferable for the production of rotation curves. However, a mere linear interpolation of quaternion components is not suitable for the purpose, since it leads to non-valid rotations by cutting through the hypersphere. To overcome this problem, a certain number of proper quaternion interpolation schemes have been presented through the years. A first method for quaternion interpolation, known as SLERP (Spherical Linear intERPolation), has been put forward by Ken Shoemake in 1985 for computer animation purposes \cite{shoemake1985animating}: instead of a linear interpolation (LERP) between two consecutive orientations, the algorithm produces a segment of geodesic on the surface of the quaternion unitary sphere. In this way, one obtains the shortest angular path between two poses performed at constant velocity \cite{dam1998quaternions}. This method works nicely between two orientations, but the mere replication on multiple frames produces a segmented curve, characterized by abrupt velocity spikes at junction points. Therefore, many other efforts were made in order to improve the smoothness of the curve. In the same paper \cite{shoemake1985animating} Shoemake proposed to generate the equivalent of a cubic B\'ezier curve in $\mathsf{S}^3$ space, achieving $C^1$ continuity. In another work the same author presented an alternative approach, which has become very popular in computer animation and is commonly referred to as SQUAD (Spherical QUADrangle interpolation) \cite{Shoemake1987QuaternionCA}. This is basically a cubic interpolation method which uses a nested SLERP algorithm between four control points; although its $C^{1}$ continuity and differentiability where not properly demonstrated in the original source \cite{dam1998quaternions}, new proofs were given in following works (such as in \cite{dam1998quaternions} and \cite{kim1996compact}). Schlag \cite{schlag1991using} extended the de Casteljau construction scheme of B\'ezier curves to obtain B-Spline quaternion curves but not did not achieve $C^{2}$ continuity in $\mathsf{SO}(3)$. On the contrary, Pletinckx \cite{pletinckx1989quaternion} produced an iterative algorithm which converged to an infinite-degree curve that, although being extremely smooth, did not have closed form formulation and required many in-between points \cite{kim1995ac}). Barr \cite{barr1992smooth} presented a method to obtain Splines (simplifying a previous algorithm proposed by Gabriel et Kajiya \cite{gabriel1985spline}) that minimize a measure of the acceleration on curved spaces. However, this approach requires an optimization algorithm to perform the highly computational expensive minimization. 
In addition, the method formulates the derivatives as finite differences resulting in a discrete sequence of points which required an additional SLERP in-between. We can mention that later Ramamoorthi \cite{ramamoorthi1997fast} improved the Barr algorithm \cite{barr1992smooth} by producing a faster cubic quaternion B-Spline algorithm through a new optimization criterion based on Euler-Lagrange functional. Ravani \cite{ravani1995bezier} investigated the possibility to extend the concept of B\'ezier curve to Riemannian manifolds and particularly to Lie groups. He obtained a generalized de Casteljau construction scheme which, however, still suffers of high computational burden: it is required to recursively solve multiple nonlinear differential equations and a two-point boundary value problem. Kim et al. focused on a method to generate quaternion cumulative basis B-Splines in $\mathsf{SO}(3)$ which preserved $C^{k}$ continuity \cite{kim1995ac} \cite{kim1995general}. This method produces very smooth curves but needs an iterative procedure to approximate the control points; in addition, the convergence is not assured when two consecutive orientations have large angular or axis displacements. More recently, many other techniques were investigated to tackle the problem. For example, Basarab \cite{basarab2010interpolating} introduced a new class of quaternion interpolation curves reformulating the Kim et al. cumulative basis approach through the use of atomic functions: the curves obtained are infinitely differentiable and their curvature can be regulated by an appropriate choice of the atomic function parameters. Bolotnikov \cite{bolotnikov2015polynomial} provided a procedure to perform quaternion interpolation by solving the polynomial Lagrange interpolation problem from a mathematical perspective. Tan et al. \cite{tan2018smooth} proposed a class of parametric quintic polynomials for quaternion interpolation, looking for a smooth, locally modifiable and efficiently-computed spline curve; the result obtained is second-order continuous and the adjustment of some tension parameters ensure the passage through given keyframes and local shape controllability. Lastly, Pu et al. \cite{pu2020c2} proposed a new type of B-Spline based on logarithmic quaternions; the problem is resolved in $\mathbb{R}^{3}$ and then remapped into $\mathsf{S}^3$.

The method proposed in this paper aims at a different goal. We provide a simple yet effective algorithm to perform a smooth and exact interpolation of an arbitrary set of rotations without the need for iterative optimization procedures or complex mathematical formulation: thus, we address those applications where a simple implementation and a small computational overhead are high priorities. In fact, we propose a new practical interpolation scheme which preserves the desirable characteristics of SLERP (easy algorithm and constant-velocity output trajectory) while overcoming its lack of $C^{1}$ continuity at curve joints. The target is achieved with the repeated combination of SLERP interpolation segments, smoothly connected by fillet tracts generated by simple polynomial motion laws. The duration of these fillets and the angular velocity profiles can be easily and arbitrarily modified by the user to obtain the desired trajectory. With respect of other mentioned interpolation algorithms (namely quaternion B-Splines and SQUAD), the formulation is therefore easier and the computational burden is lower.

The paper is organized as follows. Section \ref{sec:interpolation} formulates the algorithm, starting from the approximate interpolation of three poses (i.e. only passing through the extremes), then exposing a method to exactly interpolate all the three given poses and finally extending the exact method to any number of given key frames orientations. Section \ref{sec:numerical_examples} presents some numerical examples of the method, comparing it with the some of the principal interpolation schemes presented in literature. Finally, examples in the field of 3D animation and in the simulation of industrial robots are also provided. %Section \ref{sec:conclusions} summarizes the conclusions of the work. Finally, appendix sections recall the main principles of rotation matrices and quaternions (\ref{app:RotatMatrix}, \ref{app:Quaternion}), describe a possible implementation of some literature interpolation algorithms (\ref{app:otherinterpolations}) and present the polynomial motion laws adopted for the proposed interpolation method (\ref{app:poly5}).

%-------------------------------------------------------------------------------
% Nomenclature table
\nomenclature[10]{$\vect{v}$}{Vector, $\vect{v} \in \mathbb{R}^3$. All vectors are in bold.}
\nomenclature[20]{$\vect{v}_{(1)}$}{Vector expressed in the basis of frame (1)}
\nomenclature[30]{$\vect{u}$}{Unit vector representing the axis of rotation}
\nomenclature[40]{$R_{12}$}{Rotation matrix of frame (2) with respect to frame (1)}
\nomenclature[50]{$\underline{\vect{v}}$}{Skew symmetric 3x3 matrix built from vector $\vect{v}$}
\nomenclature[60]{$t$}{Time (total duration of the interpolation is $T$)}
\nomenclature[70]{$\quat{q}$}{Quaternion, $\quat{q} \in \mathbb{H}$}
\nomenclature[80]{$\quat{q}^*$}{Quaternion conjugate}
\nomenclature[90]{$\vect{\omega}$}{Angular velocity vector}
\nomenclature[100]{$\norm{\vect{.}}$}{Euclidean norm of vector}
\nomenclature[110]{SLERP}{Spherical Linear intERPolation method}
\nomenclature[120]{SQUAD}{Spherical QUADrangle interpolation method}
\nomenclature[130]{$\mathsf{SO}(3)$}{Special Orthogonal Group of 3D rotations}

\begin{table*}[!t]
   \begin{framed}
     \printnomenclature
   \end{framed}
\end{table*}

%-------------------------------------------------------------------------------
\section{Interpolation between angular poses}
\label{sec:interpolation}
The description of the algorithm is based on the notation and the basic properties of rotations recalled in \ref{app:RotatMatrix} \cite{legnani1996a,legnani1996b}. Rotation matrix $R_{ij}$ denotes the angular position of frame $j$ with respect to $i$. Subscript (0) will be used to denote the absolute reference frame, while digits 1, 2, \ldots will be used to denote further reference frames. For example $\vect{u}_{(i)}$ is the representation in frame $i$ of the unit vector $\vect{u}$.

\subsection{Interpolation between 2 poses}
\label{subsec:connetti2}
An angular trajectory that modifies an initial orientation $R_{01}$ into a final orientation $R_{02}$ in ``smooth'' way can be obtained by rotating around a suitable unit vector $\vect{u}$ of the suitable angular value $\overline{\varphi}$ (see \ref{app:RotatMatrix}). We can represent graphically this concept in Fig. \ref{fig:connetti2}. By operating in the absolute frame $(0)$ we can write
\begin{equation}
\overline{R}(\vect{u}_{(0)},\overline{\varphi})R_{01}=R_{02}
\qquad\qquad
    \overline{R}(\vect{u}_{(0)},\overline{\varphi}) =R_{02}R_{10}
    \label{eq:rotat_u0}
\end{equation}
being $R_{ji}$ the inverse of $R_{ij}$.\\
While operating in the local frame $(1)$ from Eq. \ref{eq:RRR12} we get
\begin{equation}
R_{01}\overline{R}(\vect{u}_{(1)},\overline{\varphi})=R_{02}
\qquad\qquad
    \overline{R}(\vect{u}_{(1)},\overline{\varphi})=R_{12} =R_{10}R_{02}
    \label{eq:rotat_u1}
\end{equation}
the rotation axis $\vect{u}$ and angle $\overline{\varphi}$ can be extracted by the rotation matrix using  eq. \ref{eq:extract}.\\ % (see \ref{app:RotatMatrix}).\\
A generic intermediate orientation $R$ between the two poses can be obtained by one of the two equivalent equations
\begin{equation}
    \begin{array}{l}
        R = R_{01} \overline{R}(\vect{u}_{(1)},\varphi)\\[2ex]
        R = \overline{R}(\vect{u}_{(0)},\varphi) R_{01}
    \end{array}
    \qquad\qquad
    % \varphi=0,...,\overline{\varphi}
    \varphi \in \left[ 0, \overline{\varphi} \right]
    \label{eq:rotat_int}
\end{equation}
If the rotation is applied at constant velocity for a time $T$, the angular velocity vector $\vect{\omega}$ expressed in frame $(0)$ and $(1)$ will be
\begin{equation}
    \vect{\omega}_{(0)} = \vect{u}_{(0)} \cfrac{\overline{\varphi}}{T}
    \qquad\qquad
    \vect{\omega}_{(1)} = \vect{u}_{(1)} \cfrac{\overline{\varphi}}{T}
    \label{eq:omega}
\end{equation}
If the angle $\varphi$ changes with a specific law of motion assuming suitable velocity, Eq. (\ref{eq:omega}) will define the average velocity $\norm{\vect{\omega}}$ whose unit vector will be $\vect{u}$ in any case.

%Assuming  $\norm{\vect{\omega}}=1$, the motion time $T$ coincide with the rotation angle $T=\overline{\varphi}$.

\subsection{Interpolation between 3 poses}
\label{subsec:connetti3}
If we want to rotate frame (1) into frame (3) passing through frame (2), it is possible to adopt twice the strategy of section \ref{subsec:connetti2} (see Fig. \ref{fig:connetti3}). Since in general the unit vectors $\vect{u}_1$ and $\vect{u}_2$ are different from each other, to preserve the continuity of angular velocity it is necessary to smoothly stop in (2) and then to start again. In the next sections we will present two methodologies to generate the desired motion without the necessity to stop but preserving the velocity and acceleration continuity.

% Direct angular motion (2-3 poses)
\begin{figure}[tbp]
    \begin{minipage}[t]{0.49\textwidth}
        \centering
        \includegraphics{figures/connetti2}
        \caption{Conventional representation of an angular motion between two frames, consisting in a rotation of angle $\varphi$ about a fixed rotation axis $\vect{u}$. The strategy adopted is described in Section \ref{subsec:connetti2}.}
        \label{fig:connetti2}
    \end{minipage}
    \hfill
    \begin{minipage}[t]{0.49\textwidth}
        \centering
        \includegraphics{figures/connetti3}
        \caption{Conventional representation of an angular motion between three frames, starting from frame (1) and ending to frame (3) while passing through frame (2). The strategy adopted is described in Section \ref{subsec:connetti3}.}
        \label{fig:connetti3}
    \end{minipage}
\end{figure}

\subsection{Approximate method}
\label{subsec:connetti3appr}
In this section we present a methodology to generate a trajectory from frame (1) to frame (3) passing near frame (2) (Fig. \ref{fig:connetti3a}). This is an approximate solution to our problem.

We initially observe that the frames (2) and (3) can be reached from the previous frames by suitable rotations $\overline{\alpha}$ and $\overline{\beta}$ around suitable axes $\vect{u}_1$ and $\vect{u}_2$
\begin{equation}
    \begin{array}{l}
        R_{02} = R_{01}R_{12} 
        \qquad\qquad 
        R_{12} = \overline{R}(\vect{u}_1,\overline{\alpha})\\[2ex]
        R_{03} = R_{02}R_{23} 
        \qquad\qquad  
        R_{23} = \overline{R}(\vect{u}_2,\overline{\beta})
    \end{array}
    \label{eq:rotA,B}
\end{equation}

% Approx interpolation of 3 poses
\begin{figure}[tbp]
    \begin{minipage}[t]{0.49\textwidth}
        \centering
        \includegraphics{figures/connetti3a}
        \caption{Approximate angular trajectory between three frames (first definition). The strategy adopted is described in Section \ref{subsec:connetti3appr}.}
        \label{fig:connetti3a}
    \end{minipage}
    \hfill
    \begin{minipage}[t]{0.49\textwidth}
        \centering
        \includegraphics{figures/connetti3b}
        \caption{Approximate angular trajectory between three frames (variant to pass closer to intermediate frame). The strategy adopted is described in Section \ref{subsec:connetti3appr}. See also Fig. \ref{fig:sfera1}}
        \label{fig:connetti3b}
    \end{minipage}
\end{figure}

% Motion laws
\begin{figure}[tbp]
    \begin{minipage}[t]{0.49\textwidth}
        \centering
        \includegraphics{figures/Legge1}
        \caption{Unitary motion laws for the angles $\alpha$ and $\beta$, to be used in the approximate connection of the three frames with Eq. \ref{eq:cambiaA,B}.}
        \label{fig:legge1}
    \end{minipage}
    \hfill
    \begin{minipage}[t]{0.49\textwidth}
        \centering
        \includegraphics{figures/legge5sva-j}
        \caption{Position, velocity ed acceleration of the motion laws $s_1$ and $s_2$ presented in Fig. \ref{fig:legge1}.}
        \label{fig:legge1sva}
    \end{minipage}
\end{figure}

So a trajectory from pose (1) to pose (3) can be obtained by the following equation
\begin{equation}
    R = R_{01}\overline{R}(\vect{u}_{1(1)},\alpha)\overline{R}(\vect{u}_{2(2)},\beta) 
    \qquad\qquad 
    % \alpha=0,...,\overline{\alpha}, \beta=0,...,\overline{\beta}
    \alpha \in \left[ 0, \overline{\alpha} \right], \beta \in \left[ 0, \overline{\beta} \right]
    \label{eq:raccordo}
\end{equation}
where for $\alpha=\beta=0$ we get $R=R_{01}$, while for $\alpha=\overline{\alpha}$ and $\beta=\overline{\beta}$ we get  $R=R_{03}$.

To pass near frame (2) it is necessary to assign suitable initial and final conditions to the variation law of  $\alpha$ and $\beta$. Initially it is necessary that $\alpha$ varies quickly and slowly in the final part. Similarly initially $\beta$ should vary slowly and quickly in the final part. As an example consider the following law (Fig. \ref{fig:legge1} and Fig. \ref{fig:legge1sva})
\begin{equation}
    \begin{array}{c}
        \alpha(t) = \overline{\alpha} s_{1}(t)
        \qquad\qquad
        \beta(t)=\overline{\beta} s_{2}(t)
        \qquad\qquad
        % t=0,...,T\\[2ex]
        t \in \left[ 0, T \right] \\[2ex]
        \dot{s}_1(0)>0 \qquad \dot{s}_1(T)=0
        \qquad\qquad
        \ddot{s}_1(0)=\ddot{s}_1(T)=0\\[2ex]
        \dot{s}_2(0)=0 \qquad \dot{s}_2(T)>0
        \qquad\qquad
        \ddot{s}_2(0)=\ddot{s}_2(T)=0
    \end{array}
    \label{eq:cambiaA,B}
\end{equation}

To pass closer to pose (2) it is possible to adopt the following strategy (Figure \ref{fig:connetti3b})
\begin{itemize}
    \item perform an initial rotation  $\alpha_1$ around a $\vect{u}_1$ and reach frame $(a)$,
    \item then move from frame  $(a)$ to $(b)$ using the previously described algorithm shown in Fig. \ref{fig:connetti3a} and Eq. \ref{eq:raccordo} with rotations $\alpha_2$ and $\beta_1$
    \item finally move from frame $(b)$ to $(3)$ rotating around $\vect{u}_2$ by an angle $\beta_2$.
\end{itemize}
The mentioned operations are performed assuming $\overline{\alpha}=\alpha_1+\alpha_2$, $\overline{\beta}=\beta_1+\beta_2$.
So the trajectory is generated by the following three steps

\begin{equation}
    \begin{array}{lll}
        step\ 1: & R=R_{01}\overline{R}(\vect{u}_1,\alpha) & \alpha \in [0, \alpha_1]\\

        step\ 2: & R=R_{01}\overline{R}(\vect{u}_1,\alpha_1)\overline{R}(\vect{u}_1,\alpha)\overline{R}(\vect{u}_2,\beta) & \alpha \in [0, \alpha_2], \beta \in [0,\beta_1]\\

        step\ 3: &
        R=R_{01}\overline{R}(\vect{u}_1,\overline{\alpha})\overline{R}(\vect{u}_2,\beta_1)\overline{R}(\vect{u}_2,\beta) & \beta \in [0, \beta_2]
    \end{array}
    \label{eq:3step}
\end{equation}

In practice steps 1 and 3 are classical SLERP interpolations segments smoothly connected by step 2. The adoption of the law of motions similar to those of Figure  \ref{fig:legge1sva} guarantees continuity of velocity and acceleration in the connection points (a) and (b) if the following  conditions are assigned
\begin{equation}
    \begin{array}{lcl}
        s_1(0)= 0 && s_2(0)=0\\
        \dot{s}_1(0)=\dot{\overline{\alpha}} && \dot{s}_2(0)=0 \\
        \ddot{s}_1(0)=0 && \ddot{s}_2(0)=0\\
        s_1(T)= \alpha_2 && s_2(T)=\beta_1\\
        \dot{s}_1(T)=0 && \dot{s}_2(T)=\dot{\overline{\beta}} \\
        \ddot{s}_1(T)=0 && \ddot{s}_2(T)=0
    \end{array}
    \label{eq:limits}
\end{equation}
where $\dot{\overline{\alpha}}$ and $\dot{\overline{\beta}}$ are, respectively, the angular velocity during the first and third step of the motion. These conditions can be easily obtained by the polynomial law described in section \ref{app:poly5}.
By assuming $s_1+s_2=k t$ and so $s_2=kt-s_1$, $\dot{s}_2=k-\dot{s}_1$, $\ddot{s}_2=-\ddot{s}_1$ for the particular law adopted $s_2(t)=s_1(T-t)$. Moreover $s_1(0)=0$, $\dot{s}_1(0)=1$, $\ddot{s}_1(0)=0$, $s_1(T)=1$,  $\dot{s}_1(T)=0$, $\ddot{s}_1(T)=0$ (see \ref{app:poly5}).

A numerical example of the exposed procedure is presented in Fig. \ref{fig:legnani_approx_compo} and Fig. \ref{fig:legnani_approx_vel}. The law of motion can be adjusted to obtain constant angular velocity as shown in Fig. \ref{fig:raccordo3-qNorm} and Fig. \ref{fig:raccordo3-wNorm}.

To generate the motion with constant velocity it is possible to adopt the following numerical approach. The motion description of Figure \ref{fig:legnani_approx_compo} is sampled at predefined time intervals $i=0, 1, 2,\ldots n$ then, for any time interval, the absolute value of the angular rotation is evaluated. Finally the absolute time $t_i$ for the $i-th$ sampled orientation is determined as
\begin{equation}
    \begin{array}{l}
        t_0=0\\
        t_i=t_{i-1}+\cfrac{\Delta\varphi_i}{\norm{\vect{\omega}}}
    \end{array}
    \label{eq:deltaT}
\end{equation}
where $\Delta\varphi_i$ is the angular rotation in the $i-th$ interval (between the time instant $i$ and $i-1$) and $\norm{\vect{\omega}}$ is the desired angular velocity ($\norm{\vect{\omega}}$=1 for the example of Fig. \ref{fig:raccordo3-qNorm} and Fig. \ref{fig:raccordo3-wNorm}). The proposed methodology can be simply adapted to any velocity profile by assigning the required value of $\norm{\vect{\omega}}$ as a function of time. For example it is possible to start the motion with null velocity, increase it until a predefined value is reached and slowing down to smoothly stop at the end of the motion.

% Legnani approx between 3 poses
\begin{figure}[tbp]
    \begin{minipage}[t]{0.49\textwidth}
        \centering
        \includegraphics[width=6.5cm]{figures/legnani_approx_compo.png}
        \caption{Numerical example of a trajectory connecting two poses and passing near to an intermediate pose (quaternion components versus time). The strategy adopted is described in \ref{subsec:connetti3appr}.}
        \label{fig:legnani_approx_compo}
    \end{minipage}
    \hfill
    \begin{minipage}[t]{0.49\textwidth}
        \centering
        \includegraphics[width=6.5cm]{figures/legnani_approx_vel.png}
        \caption{Angular velocity of interpolation method shown in Fig. \ref{fig:legnani_approx_compo}.}
        \label{fig:legnani_approx_vel}
    \end{minipage}
\end{figure}

% Legnani approx constant vel between 3 poses
\begin{figure}[tbp]
    \begin{minipage}[t]{0.49\textwidth}
        \centering
        \includegraphics[width=6.5cm]{figures/legnani_constvel_compo.png}
        \caption{Numerical example of a trajectory connecting two poses and passing near to an intermediate pose, with constant angular velocity (quaternion components versus time). The strategy adopted is described in \ref{subsec:connetti3appr}.}
        \label{fig:raccordo3-qNorm}
    \end{minipage}
    \hfill
    \begin{minipage}[t]{0.49\textwidth}
        \centering
        \includegraphics[width=6.5cm]{figures/legnani_constvel_vel.png}
        \caption{Angular velocity of interpolation method shown in Fig. \ref{fig:raccordo3-qNorm}.}
        \label{fig:raccordo3-wNorm}
    \end{minipage}
\end{figure}

\subsection{Exact method}
\label{subsec:connetti3esatto}
In order to generate a trajectory that rotates from frame (1) to (3) and passes exactly through frame (2), it is possible to apply twice the algorithm  seen in Section \ref{subsec:connetti3appr}.

At first (see Fig. \ref{fig:connetti3esatto}), a suitable rotation axis $\vect{u}_r$ is chosen to be between $\vect{u}_1$ and $\vect{u}_2$ by using positive arbitrary weights $k_1$ and $k_2$
\begin{equation}
    \vect{u}_r= \cfrac{k_1 \vect{u}_1 + k_2 \vect{u}_2}{\norm{k_1 \vect{u}_1 + k_2 \vect{u}_2}}
    %\label{eq:ur}
    \label{eq:Ur}
\end{equation}
and, chosen suitable rotation value  $\alpha'$ e $\beta'$ it is possible to generate the frames $(a')$ e $(b')$ by rotation around the axis $\vect{u}_r$. The intermediate auxiliary poses are determined as
\begin{equation}
    R_{0a'}=R(\vect{u}_r,-\alpha')R_{02} 
    \qquad\qquad   
    R_{0b'}=R(\vect{u}_r,\beta')R_{02}
    \label{eq:auxiliary}
\end{equation}
Then it is possible to apply twice the described algorithm: the first time along the poses $(1)-(a')-(2)$ and then along $(2)-(b')-(3)$. A comparison between the approximate and the exact algorithm is presented on the unitary sphere in Fig. \ref{fig:sfera1} and Fig. \ref{fig:sfera2}, while a numerical example is reported in Fig. \ref{fig:q3_legnani_exact} and Fig. \ref{fig:q3_legnani_exact_velocity}.

\subsection{Exact method for arbitrary number of rotations}
\label{subsec:connetti_esatto_general}
The presented exact algorithm can be extended to interpolate an arbitrary number of rotations. For each intermediate frame $(i)$, i.e. from $(2)$ to $(N-1)$, the fillet rotation axes are computed by weighted average of the axes connecting $(i)$ with $(i-1)$ and $(i+1)$. Next, the auxiliary poses $(A)$, $(B)$, $(C)$, \ldots, $(X)$ (see Fig. \ref{fig:legnani_general_method}) are chosen on the fillet axes by rotating of suitable desired angles. Finally, it is possible to compute the axes $\vect{u}_{1A}$, $\vect{u}_{BC}$, $\vect{u}_{D4}$, \ldots, $\vect{u}_{X-1, X}$ and introduce the additional knots $(a)$, $(b)$, $(c)$, \ldots, $(x)$ at desired angular position. For example, assuming 4 input rotations to be interpolated and referring to Fig. \ref{fig:legnani_general_method}, we can create the desired trajectory through the following steps:
\begin{equation}
    \begin{array}{ll}
        step\ 1: & R = R_{01} R(\vect{u}_{1A(1)}, \varphi_{1a})\\

        step\ 2: & R = R_{0a} R(\vect{u}_{1A(1)}, \varphi_{aA}) R(\vect{u}_{AB(2)}, \varphi_{A2})\\

        step\ 3: & R = R_{02} R(\vect{u}_{AB(2)}, \varphi_{2B}) R(\vect{u}_{BC(B)}, \varphi_{Bb})\\

        step\ 4: & R = R_{0b} R(\vect{u}_{BC(B)}, \varphi_{bc})\\

        step\ 5: & R = R_{0c} R(\vect{u}_{BC(B)}, \varphi_{cC}) R(\vect{u}_{CD(3)}, \varphi_{C3})\\

        step\ 6: & R = R_{03} R(\vect{u}_{CD(3)}, \varphi_{3D}) R(\vect{u}_{D4(D)}, \varphi_{Dd})\\

        step\ 7: & R = R_{0d} R(\vect{u}_{D4(D)}, \varphi_{d4})\\
    \end{array}
\end{equation}
This process can be repeated to interpolate any number of given rotations.

% Schematic exact/extended exact method concept
\begin{figure}[tbp]
    \begin{minipage}[t]{0.49\textwidth}
        \centering
        \includegraphics{figures/connetti3esatto}
        \caption{Exact trajectory connecting three different frames by using the intermediate poses $(a')$ and $(b')$. The strategy adopted is described in \ref{subsec:connetti3esatto}.}
        \label{fig:connetti3esatto}
    \end{minipage}
    \hfill
    \begin{minipage}[t]{0.49\textwidth}
        \centering
        \includegraphics[width=6cm]{figures/connetti4esatto.eps}
        \caption{Exact trajectory connecting four different frames by using the intermediate poses $(A)$, $(B)$, $(C)$ and $(D)$. The strategy adopted is described in \ref{subsec:connetti_esatto_general}}
        \label{fig:legnani_general_method}
    \end{minipage}
\end{figure}

% Unit sphere representation of approx/exact Legnani
\begin{figure}[tbp]
    \begin{minipage}[t]{0.49\textwidth}
        \centering
        \includegraphics[width=0.95\textwidth]{figures/sfera1a}
        \caption{Representation on the unitary sphere of the approximate algorithm (see also Fig. \ref{fig:connetti3b}).}
        \label{fig:sfera1}
    \end{minipage}
    \hfill
    \begin{minipage}[t]{0.49\textwidth}
        \centering
        \includegraphics[width=0.95\textwidth]{figures/sfera2a}
        \caption{Representation on the unitary sphere of the exact algorithm (see also Fig. \ref{fig:connetti3esatto}).}
        \label{fig:sfera2}
    \end{minipage}  
\end{figure}

% Numerical example of 3 quat interp: Legnani exact (q and w)
\begin{figure}[tbp]
    \begin{minipage}[t]{0.49\textwidth}
        \centering
        \includegraphics[width=6.5cm]{figures/legnani_exact_compo.png}
        \caption{Numerical example of a trajectory connecting two poses and passing exactly through an intermediate pose (quaternion components versus time). The strategy adopted is described in \ref{subsec:connetti3esatto}.}
        \label{fig:q3_legnani_exact}
    \end{minipage}
    \hfill
    \begin{minipage}[t]{0.49\textwidth}
        \centering
        \includegraphics[width=6.5cm]{figures/legnani_exact_vel.png}
        \caption{Angular velocity of interpolation method shown in Fig. \ref{fig:q3_legnani_exact}.}
        \label{fig:q3_legnani_exact_velocity}
    \end{minipage}
\end{figure}

\subsection{Assigning predefined velocity at the assigned poses}
The proposed algorithm generates 3D orientation trajectories connecting the assigned poses. The angular velocity profile can be easily adjusted using the following guidelines which permit to assign exact angular velocity values (modulus and direction) in the initial, intermediate and final points.\\
To assign the direction of the angular velocity it is possible to freely assign the orientation vectors used to generate the trajectory near the points (e.g. $\vect{u}_r$ in Figure \ref{fig:connetti3esatto}, and $\vect{u}_{AB}$ and $\vect{u}_{CD}$ in Figure \ref{fig:legnani_general_method}), in fact the adoption of Eq. \ref{eq:Ur} is a suggestion that generally smooth the trajectory but it is not mandatory.\\
To assign the desired modulus of the angular velocity it sufficient to apply eq. \ref{eq:deltaT} using a suitable variable value of $||\omega||$ along the trajectory. 
At first the whole trajectory is considered to evaluate the total rotation around the trajectory $\overline{\varphi}=\sum \varphi_i$. Then the desired law of variation of $\varphi$ with respect to the time is chosen ($0\leq \varphi \leq \overline{\varphi}$) respecting the desired velocity in the points of interest. This procedure permits to assign the desired value of $||\omega_i||$ for each time interval.

%-------------------------------------------------------------------------------
\section{Numerical examples}
\label{sec:numerical_examples}
In this section we present some numerical examples of rotation interpolation, followed by some illustrated applications.

First of all, we can quantitatively compare the presented interpolation method with other common algorithms available in literature (refer to \ref{app:otherinterpolations} for further information). To this purpose, we assign some rotation key frames and observe the four quaternion component trends together with the corresponding angular velocity components, both of which with respect to time $t \in [0, 1]$. In particular, we compare the SLERP \cite{shoemake1985animating} method, a 2\textsuperscript{nd}-order quaternion B-Spline method (based on \cite{kim1995ac}), the SQUAD method (based on \cite{Shoemake1987QuaternionCA} implementation) and the proposed method. We observe that:

\begin{itemize}
    \item SLERP passes though all the imposed frames without overshoots and minimizes the angular path. Velocity is piecewise constant but manifests abrupt discontinuities at curve joints (step variations); consequently, angular acceleration locally presents infinite values. See Fig. \ref{fig:chrono_spline1_compo} and Fig. \ref{fig:chrono_spline1_vel}.
    \item 2\textsuperscript{nd}-order quaternion B-Spline is very smooth, but does not exactly reach the intermediate imposed frames. Velocity is continuous and mainly constituted by linear tracts, but presents drastically rapid changes at curve joints; consequently, angular acceleration will locally be very high. See Fig. \ref{fig:chrono_spline2_compo} and Fig. \ref{fig:chrono_spline2_vel}.
    \item SQUAD exactly interpolates all the given poses at the cost of a slight increase of the angular path. Velocity is continuous and smooth but it manifests a continuously variable and bouncy trend. See Fig. \ref{fig:chrono_squad_compo} and Fig. \ref{fig:chrono_squad_vel}.
    \item The proposed method produces a smooth trajectory, exactly passing through all the given poses with modest deviation from the minimal SLERP path and slight overshoots at curve joints. As for SLERP, velocity is preserved constant through the in-between interpolation tracts, but with the advantage of smooth connections at curve joints. The transitions can be arbitrarily adapted to fulfil specific local constraints. See Fig. \ref{fig:legnani_eqspaced_compo} and Fig. \ref{fig:legnani_eqspaced_vel}.
\end{itemize}

To better visualize the achieved results in rotation interpolation, we present some snapshots of a 3D simulation as well. The task was performed by leveraging an open-source multibody simulation engine (Chrono::Engine \cite{tasora2015chrono}), in which the discussed algorithms were implemented.

Fig. \ref{fig:strobo_trail} shows a free body subject to the spatial and rotational interpolation of four key poses (blue). The red/green/blue curves are integral with the X/Y/Z local axes of the moving body, while in gray we shown the stroboscopic trail of the configurations. All the poses are smoothly and exactly connected. Fig. \ref{fig:axes_trail} shows a comparison between the proposed method and SLERP; the four imposed rotation frames are the same as the previous ones (Fig. \ref{fig:strobo_trail}), but there is no spatial motion. One can observe the undesired spikes produced by SLERP by looking at the red/green/blue axes curves. Finally, Fig. \ref{fig:chrono_robot} shows a potential application of the proposed algorithm: the smooth interpolation of poses performed by the end-effector of an industrial robot (namely a 6-DOF robot arm equipped with painting tools).

% Numerical example of 6 quat interp: SLERP (q and w)
\begin{figure}[tbp]
    \begin{minipage}[t]{0.49\textwidth}
        \centering
        \includegraphics[width=6cm]{figures/chrono_spline1_compo.png}
        \caption{Numerical example of a trajectory connecting six poses using the SLERP interpolation method (quaternion components versus time). All the given frames are exactly reached by a non-smooth trajectory, having minimum angular length.}
        \label{fig:chrono_spline1_compo}
    \end{minipage}
    \hfill
    \begin{minipage}[t]{0.49\textwidth}
        \centering
        \includegraphics[width=6.5cm]{figures/chrono_spline1_vel.png}
        \caption{Angular velocity of SLERP interpolation method shown in Fig. \ref{fig:chrono_spline1_compo}. Severe discontinuities at curve joints are notable.}
        \label{fig:chrono_spline1_vel}
    \end{minipage}
\end{figure}

% Numerical example of 6 quat interp: BSpline2 (q and w)
\begin{figure}[tbp]
    \begin{minipage}[t]{0.49\textwidth}
        \centering
        \includegraphics[width=6.5cm]{figures/chrono_spline2_compo.png}
        \caption{Numerical example of a trajectory connecting six poses using a 2\textsuperscript{nd}-order quaternion B-Spline (quaternion components versus time). Angular trajectory is very smooth, but does not exactly pass through all the given frames.}
        \label{fig:chrono_spline2_compo}
    \end{minipage}
    \hfill
    \begin{minipage}[t]{0.49\textwidth}
        \centering
        \includegraphics[width=6.5cm]{figures/chrono_spline2_vel.png}
        \caption{Angular velocity of 2\textsuperscript{nd}-order quaternion B-Spline interpolation method shown in Fig. \ref{fig:chrono_spline2_compo}. The trend is continuous, but with abrupt spikes at curve joints.}
        \label{fig:chrono_spline2_vel}
    \end{minipage}
\end{figure}

% Numerical example of 6 quat interp: SQUAD (q and w)
\begin{figure}[tbp]
    \begin{minipage}[t]{0.49\textwidth}
        \centering
        \includegraphics[width=6.5cm]{figures/chrono_squad_compo.png}
        \caption{Numerical example of a trajectory connecting six poses using the SQUAD interpolation method (quaternion components versus time). All the given frames are exactly reached by a smooth trajectory.}
        \label{fig:chrono_squad_compo}
    \end{minipage}
    \hfill
    \begin{minipage}[t]{0.49\textwidth}
        \centering
        \includegraphics[width=6.5cm]{figures/chrono_squad_vel.png}
        \caption{Angular velocity of SQUAD interpolation method shown in Fig. \ref{fig:chrono_squad_compo}. The trend is continuous but highly oscillating.}
        \label{fig:chrono_squad_vel}
    \end{minipage}
\end{figure}

% Numerical example of 6 quat interp: Legnani exact (q and w)
\begin{figure}[tbp]
    \begin{minipage}[t]{0.49\textwidth}
        \centering
        \includegraphics[width=6.5cm]{figures/legnani_eqspaced_compo.png}
        \caption{Numerical example of a trajectory connecting six poses using the proposed extended exact interpolation method (quaternion components versus time). All the given frames are exactly reached by a smooth trajectory.}
        \label{fig:legnani_eqspaced_compo}
    \end{minipage}
    \hfill
    \begin{minipage}[t]{0.49\textwidth}
        \centering
        \includegraphics[width=6.5cm]{figures/legnani_eqspaced_vel.png}
        \caption{Angular velocity of the proposed extended exact interpolation method shown in Fig. \ref{fig:legnani_eqspaced_compo}. The trend remains constant over the intermediate phases and smoothly transitions from one tract to the other.}
        \label{fig:legnani_eqspaced_vel}
    \end{minipage}
\end{figure}

% Chrono simulation: stroboscopic trail
\begin{figure}[tbp]
    \centering
    \includegraphics[trim=0 1cm 0 3cm,clip,width=0.99\textwidth]{figures/strobo_trail.PNG}
    \caption{Simulated interpolation of four poses using the proposed method in association with a spatial motion \cite{tasora2015chrono}. The assigned key frames are shown in blue, while the moving body axes are represented by RGB trails (key frame passages are highlighted). It is notable that the generated trajectory is smooth and exactly passes though all the given rotations.}
    \label{fig:strobo_trail}
\end{figure}

% Chrono simulation: axes trail  
\begin{figure}[tbp]
    \begin{minipage}[t]{0.49\textwidth}
        \centering
        \includegraphics[width=6.5cm]{figures/axes_trail_slerp.PNG}
    \end{minipage}
    \hfill
    \begin{minipage}[t]{0.49\textwidth}
        \centering
        \includegraphics[width=6.5cm]{figures/axes_trail_legnani.png}
    \end{minipage}
    \caption{Simulated comparison \cite{tasora2015chrono} between the proposed method (left) and SLERP (right) interpolation of the same four poses of Fig.\ref{fig:strobo_trail}, without spatial motion. The rotating body axes are represented by RGB trails and the passage through the assigned key frames is highlighted as white and black dots. It may be noted from the axes trails that the SLERP trajectory manifests spikes at curve joints, therefore discontinuities in angular velocity; on the contrary, the proposed algorithm connects the given orientations with smooth curves.}
    \label{fig:axes_trail}
\end{figure}

% Chrono simulation: robot application
\begin{figure*}[tbp]
	\centering
    \subfloat[]{\includegraphics[width=0.45\textwidth]{figures/robot1.jpg}}
    % \hfill
    % \subfloat[]{\includegraphics[width=0.25\textwidth]{figures/robot2.jpg}}
    \hfill
    \subfloat[]{\includegraphics[width=0.45\textwidth]{figures/robot3.jpg}}
    % \hfill
    % \subfloat[]{\includegraphics[width=0.25\textwidth]{figures/robot4.jpg}}
    \\
    \subfloat[]{\includegraphics[width=0.45\textwidth]{figures/robot5.jpg}}
    % \hfill
    % \subfloat[]{\includegraphics[width=0.25\textwidth]{figures/robot6.jpg}}
    \hfill
    \subfloat[]{\includegraphics[width=0.45\textwidth]{figures/robot7.jpg}}
    % \hfill
    % \subfloat[]{\includegraphics[width=0.25\textwidth]{figures/robot8.jpg}}
    \caption{Simulation \cite{tasora2015chrono} of an industrial 6-DOF robot whose end-effector, equipped with a painting tool, interpolates between various key poses using the proposed algorithm in association with a spatial motion. The provided example illustrates one of the many possible applications suitable for the algorithm.}
    \label{fig:chrono_robot}
\end{figure*}

%-------------------------------------------------------------------------------
\section{Benchmarks}
\label{sec:benchmarks}
In order to show that the computational effort of the proposed method is competitive with respect to other interpolation methods, we implemented a dedicated benchmark. In particular, we tested the interpolation of different number of key frames, evaluating 1E6 of intermediate samples for each. The same test has been repeated 100 times for each scenario, and the average values of computational time have been reported in Table \ref{tab:benchmark_time}. One can observe that the proposed method is indeed competitive with SLERP for a reasonably low number of given poses and, despite the progressive increase of time in function of the key frames number, its CPU time is still shorter than the 2\textsuperscript{nd}-order quaternion B-Spline and the SQUAD method.

The benchmarks have been computed on a Intel\textsuperscript{\textregistered} Core\textsuperscript{\texttrademark} i7-10510U CPU, clocked at \SI{1.80}{GHz}, with 4 physical cores and \SI{16}{GB} of RAM. The linear algebra formulas for the benchmark have been implemented in C++ in the Chrono::Engine open-source library \cite{tasora2015chrono}, enabling the /O2 compiler optimization flag on the MSVC compiler, and using a single thread. In search of a fair comparison, no special optimization techniques have been put in place: in fact, the performance of the methods can be further improved using advanced programming techniques such as ASM assembly directives and AVX2 vectorization, that we have not considered at the moment.

\begin{table}[tp]
	\centering
	\caption{Computation time required for the interpolation of different numbers of given frames by the discussed methods, expressed in $\SI{}{ms}$. The test involves the evaluation of 1E6 intermediate samples and has been repeated 100 times for each of them (average values reported).}
	\setlength{\tabcolsep}{3pt}
	\begin{tabular}{c|ccccc}
		\toprule
		& SLERP & B-Spline2 & SQUAD & Proposed method \\
		\midrule
        4 key frames & 48.2 & 479.1 & 494.5 & 66.1 \\
        50 key frames & 49.6 & 483.6 & 524.1 & 142.0 \\
        100 key frames & 50.6 & 491.9 & 543.0 & 194.1 \\
		\bottomrule
	\end{tabular}
	\label{tab:benchmark_time}
\end{table}

Table \ref{tab:benchmark_dist} compares a measure of the total angular distance required by each of the methods discussed to interpolate the same 100 random key frames. It is observable that, as expected, the proposed algorithm travels a longer path with respect to SLERP but, in exchange, smoothly interpolates all the given poses. On the other hand, the angular distance travelled is slightly less if compared to SQUAD. At last, the 2\textsuperscript{nd}-order quaternion B-Spline presents the shortest angular path of all methods, but sacrifices the exact passage among all the given key frames.

\begin{table}[tp]
	\centering
	\caption{Example of angular distance travelled to interpolate the same 100 random key frames by the discussed methods, expressed in $\SI{}{rad}$.}
	\setlength{\tabcolsep}{3pt}
	\begin{tabular}{c|cccc}
		\toprule
	    & SLERP & B-Spline2 & SQUAD & Proposed method \\
		\midrule
        Angle & 112.6 & 84.2 & 124.5 & 117.1 \\
		\bottomrule
	\end{tabular}
	\label{tab:benchmark_dist}
\end{table}

%-------------------------------------------------------------------------------
\section{Conclusions}
\label{sec:conclusions}
We proposed a practical method, based on the composition of rotations and simple polynomials, that provides a smooth interpolation between an arbitrary number of given angular poses. The algorithm is rooted on repeated SLERP interpolation segments smoothly connected by suitable fillet tracts. The duration of the different segments can be arbitrarily modified by the user to optimize the trajectory. The methodology intrinsically generates transitions in orientation which preserve the continuity of angular position, velocity and acceleration. The mathematical formulation is simple because it involves simple combination of rotations and the computational burden is limited, as showed by the provided benchmark. It is easy to generate constant angular velocity trajectories as well as motions with predefined velocity profiles. Specific angular velocity vectors can be imposed in the control points. The comparison with other known methodologies shows that:
1) with respect to SLERP, it does not present velocity discontinuities but slightly increases angular distance covered and computation time;
2) with respect to a quaternion B-Spline, the computational burden is lower and it is easier to impose the exact crossing of the assigned angular positions;
3) with respect to SQUAD, the angular distance covered is smaller, the computation time is lower and there are no oscillations in velocities.

The proposed methodology can be easily used to generate 3D motions of mechanical parts and it can be useful in those cases where a simple formula for smooth interpolation of key poses is desired: we can mention, for instance, biomechanics (when one must interpolate between sample poses obtained via motion capture experiments) as well as computer graphics, video games and robotics. Finally, a possible application related to industrial robots has been presented to this end.

%-------------------------------------------------------------------------------

\clearpage
\appendix

%-------------------------------------------------------------------------------
\section{Rotation matrices}
\label{app:RotatMatrix}
The present appendix section recalls, without proof, some fundamental notions on rotation matrices \cite{joe-angoli,joe-prj} and defines the notation used throughout this paper.

\subsection{Conversion between different reference frames}
\label{app:RotatMatrix1}
The conversion of vectors between two different reference frames $(i)$ and $(j)$ can be obtained by means of the rotation matrix $R_{ji}$ representing their relative pose. Naming $\vect{v}_{(i)}$ the original three-components vector and $\underline{\vect{v}}_{(i)}$ its skew-symmetric matrix representation as in
\begin{equation}
    \vect{v}_{(i)} = \left[
    \begin{array}{c}
            v_{xi} \\
            v_{yi} \\
            v_{zi} \\
    \end{array}
    \right] \qquad {\underline{\vect{v}}_{(i)}} = \left[
    \begin{array}{ccc}
            0 & -v_{zi} & v_{yi} \\
            v_{zi} & 0 & -v_{xi} \\
            -v_{yi} & v_{xi} & 0 \\
    \end{array}
    \right]
    \label{eq:VV_}
\end{equation}
we obtain the following notable relations:
\begin{equation}
    \vect{v}_{(j)} = R_{ji} \vect{v}_{(i)} \qquad\qquad \underline{\vect{v}}_{(j)} = R_{ji} \underline{\vect{v}}_{(i)} R_{ij}
    \label{eq:cambioRif}
\end{equation}
Each column of matrix $R_{ji}$ contains the unit vectors of the axes of frame  $(i)$ as seen by frame $(j)$
\begin{equation}
    R_{ji} = \left[
    \begin{array}{c|c|c}
        x_{x} & y_{x} & z_{x} \\
        x_{y} & y_{y} & z_{y} \\
        x_{z} & y_{z} & z_{z} \\
    \end{array}
    \right]
     = \left[
    \begin{array}{c|c|c}
         &  &  \\
        \vect{u}_x & \vect{u}_y & \vect{u}_z \\
         &  & \\
    \end{array}
    \right]
    \label{eq:column}
\end{equation}
This matrix also represent the orientation (or attitude) of frame $(i)$ with respect to frame $(j)$.

Since the presented properties hold for any frames, Eq. \ref{eq:cambioRif} can also be written as $\vect{v}_{(i)} = R_{ij} \vect{v}_{(j)}$; thus, we deduce  that
\begin{equation}
    R_{ij}=R_{ji}^{-1}
    \label{eq:inverseR}
\end{equation}
Considering that the columns of matrix $R$ in Eq. \ref{eq:column} are orthogonal to each other, we conclude that matrix $R$ itself is orthogonal and so the following properties holds
\begin{equation}
    R^{-1}=R^T \qquad\qquad \det(R)=\pm 1
    \label{eq:RiRT}
\end{equation}
The determinant is positive if both frames are right or left, and negative if one is right and the other is left.

In case of multiple frame change it is possible to use the following composition rule
\begin{equation}
    R_{ik} = R_{ij} R_{jk}
    \label{eq:MultiR}
\end{equation}

\subsection{Rotation of vectors and frames}
\label{app:RotatMatrix2}
Let us consider a vector $\vect{v} = [v_x\ v_y\ v_z]^T$ to be rotated from a starting position $\vect{v}_s$ to reach a final position $\vect{v}_f$ of an angle $\varphi$ around axis $\vect{u}$. This transformation can be represented by matrix $R$ with:
\begin{equation}
    \vect{v}_f = R \vect{v}_s 
    \qquad\qquad 
    R(\vect{u},\varphi) = 1 + \underline{\vect{u}}\sin\varphi + \underline{\vect{u}}^2(1-\cos\varphi)
    \label{eq:rotat_phi_u}
\end{equation}
Of course the vectors and the unit vector $\vect{u}$ must be represented by their projections in the same reference frame that we can indicate (for instance) as $(r)$. If a rotation is expressed in frame $(r)$, the same rotation in a different frame $(s)$ will be (see Eq. \ref{eq:cambioRif})
\begin{equation}
    R_{(s)}= R_{sr} R_{(r)} R_{rs}
    \label{eq:changeRif}
\end{equation}
Let's indicate by subscript $(0)$ the absolute reference system in which we describe the angular position of two frames (1) and (2).
Remembering Eq. \ref{eq:column}, if we rotate the frame whose attitude is $R_{01}$ of an angle $\varphi$ around an axis $\vect{u}$ to reach the orientation $R_{02}$, we can write (each column of $R$ is transformed according to Eq. \ref{eq:cambioRif}):
\begin{equation}
    R_{02}=\overline{R}(\vect{u}_{(0)},\varphi) R_{01}
    \label{eq:rotR01_02}
\end{equation}
So, if the initial angular position $R_{01}$ and the final one $R_{02}$ are known, the rotation which transform the former to the latter is determined in frame $(0)$ as
\begin{equation}
    \overline{R}(\vect{u}_{(0)},\varphi) = R_{02} R_{01}^{-1} = R_{02} R_{10}
    \label{eq:RotFrame0}
\end{equation}
Remembering Eqs. \ref{eq:cambioRif}, \ref{eq:MultiR} and \ref{eq:changeRif}, if the result is sought in frame $(1)$ or $(2)$, the rotation becomes simply
\begin{equation}
   \overline{R}(\vect{u}_{(1)}, \varphi) = \overline{R}(\vect{u}_{(2)},\varphi) = R_{12}
    \label{eq:RRR12}
\end{equation}
Once the matrix $\overline{R}$ is known, the rotation angle and axis may be extracted as
\begin{equation}
    \begin{array}{l}
        c=\cos\varphi=\frac{x_{x}+y_{y}+z_{z}-1}{2}\\[2ex] s=\sin\varphi=\frac{\sqrt{(y_{z}-z_{y})^2+(x_{z}-z_{x})^2+(x_{y}-y_{x})^2}}{2}\\[2ex]
        \varphi=\operatorname{atan2}(s,c); \hspace{4ex} \pi < \varphi \leq \pi\\
        \ \\
        \underline{\vect{u}}=\cfrac{1}{2\sin\varphi} \left(\overline{R}-\overline{R}^T\right)
        \qquad
        \begin{array}{l}
          u_x=\frac{(y_{z}-z_{y})}{2\sin \varphi} \\[2ex]
          u_y=\frac{(z_{x}-x_{z})}{2\sin \varphi} \\[2ex]
          u_z=\frac{(x_{y}-y_{x})}{2\sin \varphi}
        \end{array}
        \qquad\qquad
        \sin\varphi\ne 0
    \end{array}
    \label{eq:extract}
\end{equation}

%-------------------------------------------------------------------------------
\section{Quaternions, rotations and exponentials}
\label{app:Quaternion}

We list some properties of quaternions that are used in interpolation schemes discussed in this work.

\subsection{Main properties}
Rotations in 3D space can be represented by four scalars $q_0,q_1,q_2,q_3$
%$\quat{q} = \{ q_0,q_1,q_2,q_3 \}$, 
called Euler parameters, related to the rotation angle $\varphi$ and rotation axis $\vect{u}$ as \cite{tasora}
%
\begin{equation}
    \begin{array}{c}
    	{q}_0 = \cos \left(\frac{\varphi}{2} \right) \\[2ex]
     	{q}_1 = u_x \sin \left(\frac{\varphi}{2}\right)  \quad
     	{q}_2 = u_y \sin \left(\frac{\varphi}{2}\right)  \quad
     	{q}_3 = u_z \sin \left(\frac{\varphi}{2}\right) 	
    \end{array}
    \label{eq:defQ}
\end{equation}
By the definitions (Eq. \ref{eq:defQ}), the following constrain holds: $q_0^2+q_1^2+q_2^2+q_3^2=1$.
Euler parameters can be interpreted as unit-length \textit{quaternions} $\quat{q} \in \mathbb{H}_1$, $||\quat{q}||=1$, that is a subset of quaternion hyper-complex numbers $\mathbb{H}$ written as
\begin{equation}
    \quat{q}= q_0 + q_1\: i + q_2\: j + q_3\: k  
    \label{eq:quaternion}
\end{equation}
In this representation \cite{hamilton1844ii}, the following properties hold
\begin{equation}
    \begin{array}{c}
        i^2=j^2=k^2=-1\\[2ex]
        ij=-ji=k;\qquad jk=-kj=i; \qquad ki=-ik=j
    \end{array}
    \label{eq:quaternionProp}
\end{equation}

Quaternions can be written succinctly with the notation $\quat{q}=[q_{s},\vect{q}_{v}]$, showing the \textit{scalar part} $q_s=q_0$ and the \textit{vector part} $\vect{q}_v = \{q_1, q_2, q_3\}$; this simplifies the expression of quaternion multiplication. In fact, with $\quat{\tau}\in\mathbb{H}$, $\quat{\eta}\in\mathbb{H}$, by means of the properties \eqref{eq:quaternionProp} one obtains the product formula:
\begin{align} 
\quat{\tau} \: \quat{\eta} = [ \tau_{s}\eta{s} - \vect{\tau}_{v}\cdot\vect{\eta}_{v}, \;
\tau_{s}\vect{\eta}_{v} + \rho_{s}\vect{\eta}_{v} + \vect{\eta}_{v} \times \vect{\eta}_{v} ]
	%\label{eq:quatproduct}
\end{align}

A rotation matrix can be constructed from a quaternion as
\begin{equation}
\begin{array}{rl}
R(\quat{q})& =\left[
	\begin{matrix}
	 q_0^2 + q_1^2 - q_2^2 - q_3^2  &  2(q_1 q_2 - q_3 q_0)   &  2(q_1 q_3 + q_2 q_0)  \cr
	 2(q_1 q_2 + q_3 q_0)  &  q_0^2 - q_1^2 + q_2^2 - q_3^2  &   2(-q_1 q_0 + q_2 q_3) \cr
     2(q_1 q_3 - q_2 q_0)  &  2(q_1 q_0 + q_2 q_3)  & q_0 ^2 -q_1^2 -q_2^2 + q_3^2   
	\end{matrix}
	\right] =\\
&=\left[
\begin{array}{ccc}
        x_{x}            &       y_{x}           &       z_{x}  \\
        x_{y}            &       y_{y}           &       z_{y}  \\
        x_{z}            &       y_{z}           &       z_{z}
\end{array}\right]
\end{array}
\label{new_eq:fromQtoR}
\end{equation}
Vice versa, if the matrix $R$ is known, the elements of the quaternion are extracted as
\begin{equation}
\begin{array}{l}
q_0=\frac{1}{2} \sqrt{t+1},\qquad t=x_{x}+y_{y}+z_{z};\\[2ex]
  q_1=\frac{(y_{z}-z_{y})}{4\: q_0};\qquad \qquad
  q_2=\frac{(z_{x}-x_{z})}{4\: q_0};\qquad \qquad
  q_3=\frac{(z_{y}-y_{x})}{4\: q_0}
\label{eq:fromRtoQ}
\end{array}
\end{equation}
which are valid for $q_0 \neq 0$; while for $q_0=0$ the following relations hold
\begin{equation}
\begin{array}{l}
        \left\{
        \begin{array}{c}
                4 q_2 q_3 =y_{z}+z_{y} \\
                4 q_1 q_3 =z_{x}+x_{z} \\
                4 q_1 q_2 =y_{x}+x_{y}
        \end{array}
        \right.
\ \\
{\small
                q_1=\pm \sqrt{\cfrac{x_{x}+1}{2}} \qquad
                q_2=\pm \sqrt{\cfrac{y_{y}+1}{2}} \qquad
                q_3=\pm \sqrt{\cfrac{z_{z}+1}{2}}
}
\end{array}
\label{eq:fromRtoQ2}
\end{equation}

A sequence of rotations can be represented by the product of two quaternions:
similarly to $R_t=R(\quat{q}_{2}) R(\quat{q}_{1})$, the quaternion $\quat{q}_{t}$, expressing the final rotation obtained by the sequence of a rotation $\quat{q}_{1}$ followed by a rotation $\quat{q}_{2}$, is obtained as 
\begin{equation}
    \quat{q}_{t}=\quat{q}_{2}\: \quat{q}_{1}
\end{equation}

We denote the quaternion conjugate of $\quat{q}$ as $\quat{q}^*$, with $\quat{q}^*= q_0 - q_1 i  - q_2 j -  q_3 k$. This corresponds to an inverse rotation and in fact one has:
\begin{equation}
    \amatr{R}(\quat{q}^*) = \amatr{R}(\quat{q})^T = \amatr{R}(\quat{q})^{-1} 
\end{equation}

Quaternions, as well as rotation matrices define an algebraic structure over a differential manifold with properties of closure, associativity, presence of identity element and inverse element, hence they belong to a \textit{Lie group}. The Lie group of rotation matrices is called $\mathsf{SO}(3)$, the special orthogonal group of orthogonal matrices with $\text{det}=1$. The Lie group of unimodular quaternions $\mathbb{H}_1$ is topologically isomorphic to the $\mathsf{S}^3$ sphere, as such it is a double cover of $\mathsf{SO}(3)$, hence two distinct opposite quaternions represent the same rotation matrix $\amatr{R} \in \mathsf{SO}(3)$.

Given a $\gamma(t) : \mathbb{R} \rightarrow \mathsf{SO}(3)$ be a one parameter sub-group of $\mathsf{SO}(3)$ for which $\gamma(0) = I$, an \emph{exponential map} is defined as $\exp(\amatr{\Theta})=\gamma(1)$, where $\amatr{\Theta} \in \mathfrak{so}(3)$ is an element of the underlying Lie algebra.

Rotation matrices can be built as exponential maps from elements $\amatr{\Theta} \in \mathfrak{so}(3)$ and vice versa:  
%
\begin{align}
\amatr{R}      &= \mathrm{exp}(\amatr{\Theta})=R(\vect{u},\varphi) \label{Rexp}\\
\amatr{\Theta} &= \mathrm{log}(\amatr{R})=\underline{\vect{u}}\varphi \label{Rlog}
\end{align}
%
The \emph{rotation vector} $\vect{\theta} \in \mathbb{R}^3$, defined for a finite rotation of angle $\varphi$ about a unit vector $\vect{u}$, is related to the 3x3 skew-symmetric matrix $\amatr{\Theta} \in \mathfrak{so}(3)$ via 
%
\begin{align}
    \vect{\theta} &= \text{axis}(\amatr{\Theta})=\vect{u}\varphi \\ 
    \amatr{\Theta} &=  \mathrm{skew}(\vect{\theta}) = \underline{\vect{\theta}}=\underline{\vect{u}}\varphi
\end{align}

Similarly, the exponential map of unit quaternions links $\mathbb{H}_1$ to its Lie algebra $\mathsf{Im}(\mathbb{H})$ of \emph{pure quaternions} $\vect{\rho} =  [0, \vect{\nu}]$:
%
\begin{align}
\quat{q}      &= \mathrm{exp}(\vect{\rho}) \label{eq:rexp}\\
\vect{\rho} &= \mathrm{log}(\vect{q})   \label{eq:rlog}
\end{align}
%
The exponential map \eqref{eq:rexp} can be explicitly computed using the closed-form expression 
$\exp([s,\vect{v}]) =  e^s \left[\cos(||\vect{v}||),\frac{\vect{v}}{||\vect{v}||}\sin(||\vect{v}||)\right]$, hence:
%
\begin{align}
\label{eq:at:quatexpa}
  \exp([0, \vect{\nu}]) &=  \left[\cos(||\vect{\nu}||),\frac{\vect{\nu}}{||\vect{\nu}||}\sin(||\vect{\nu}||)\right]   \\
\label{eq:at:quatexpb}
  \exp([0,\vect{u}\xi]) &=  [\cos(\xi),\vect{u}\sin(\xi)]  \quad  \text{for} \; ||\vect{u}||=1
\end{align}
%
We can use the  \textrm{pure}() and \textrm{imag}() operators to convert pure quaternions  $[0,\vect{\nu}]$ from and to rotation vectors $\vect{\theta}$ by observing that $\xi =  \frac{1}{2} \varphi$, $\vect{\nu} = \frac{1}{2} \vect{\theta}$: 
\begin{align}
    \vect{\theta}  &= 2 \: \text{imag}(\vect{\rho})  \\
    \vect{\rho}    &= \tfrac{1}{2} \text{pure}(\vect{\theta})  
    \end{align}

To pass directly from rotation vector $\vect{\theta}$ to quaternion $\vect{q}=[s,\vect{v}]$, and vice versa, one can use the compact expressions:
\begin{align}
\vect{q} &= 
%\mathrm{exp}(\vect{\rho}) =
\mathrm{exp} \left( \tfrac{1}{2} \text{pure}(\vect{\theta}) \right) = 
\left[ 
\cos\left(\frac{||\vect{\theta}||}{2}\right), 
\frac{\vect{\theta} }{ ||\vect{\theta}|| } \sin\left( \frac{||\vect{\theta}||}{2}\right) 
\right] 
\label{eq:qexponential}  \\
\vect{\theta} &= 
%\mathrm{axis}(\vect{\rho}) =
2 \: \mathrm{imag} ( \text{log}(\vect{q}) ) = 
2\frac{\vect{v}}{||\vect{v}||}\tan^{-1}\left(\frac{||\vect{v}||}{s}\right) 
\label{eq:qlogarithm}
\end{align} 
Note that the last expression is singular for zero rotations, so when $||\vect{v}|| < \epsilon$ one can compute it as the simplified expression $\vect{\theta} = 2 \vect{v}$. Also note that rather than using $\tan^{-1}()$ it is advisable to use $\mathrm{atan2}()$.

As a special case, the scalar exponential of a quaternion boils down to:
\begin{align}
    \nonumber
        \quat{q}^{\:t} &= \left[ \cos(\xi),\vect{u}\sin(\xi) \right]^t \\ 
    \label{eq:at:quatpowd}
      \quat{q}^{\:t}  &= \left[ \cos(\xi t),\vect{u}\sin(\xi t) \right]
\end{align}

Finally, one can see that the interpolation method of Eq.\ref{eq:raccordo} can be expressed also via a double product of quaternions:
\begin{align}
        \quat{q} = \quat{q}_{01}
        \mathrm{exp} \left( \tfrac{1}{2} \text{pure}(\vect{u}_1 \alpha) \right)
        \mathrm{exp} \left( \tfrac{1}{2} \text{pure}(\vect{u}_2 \beta) \right)
         \label{eq:at:raccordo_quaternioni}
\end{align}
or equivalently as 
%$\quat{q} = \quat{q}_{01}  \quat{q}_{12}^{\alpha/\overline{\alpha}} \quat{q}_{23}^{\beta/\overline{\beta}}$
$\quat{q} = \quat{q}_{01}
\mathrm{exp} \left( \alpha \, \mathrm{log}(\quat{q}_{01}^* \quat{q}_{02})  \right)
\mathrm{exp} \left( \beta \, \mathrm{log}(\quat{q}_{02}^* \quat{q}_{03})  \right)
$
.

\subsection{Velocity and acceleration} 
%From the definition of the exponential map, it holds 
%$\gamma(t)=\exp(t \amatr{\Omega})$, with $\gamma(0)=I$,
%therefore
%$\dot{\gamma}(0)=\amatr{\Omega}$.
%For a more generic case with $R(t)=\gamma(t) R_0$, $R(0)=R_0$, one multiplies both terms of the %first equation by $R_0$ to obtain 
%$\gamma(t) R_0 = \exp(t\amatr{\Omega}) R_0$, \hl{***per omega costante?***}
%then differentiates to obtain
%$\dot{R}(t) = \amatr{\Omega} R_0$, where $\amatr{\vect{\omega}} = \underline{\vect{\omega}}_e$ is the skew-symmetric matrix that contains the angular velocity $\vect{\omega}_e$ expressed in absolute (extrinsic) coordinates. This is written succinctly as:
Introducing the vector of angular velocity $\vect{\omega}_e$ expressed in absolute (extrinsic) coordinates, one has
\begin{align}
\dot{R} = \underline{\vect{\omega}}_e R\qquad
\label{eq:Adot}
\end{align}
Similarly, by a second differentiation, and introducing the (extrinsic) angular acceleration $\vect{\dot{\omega}}_e$, one has
\begin{align}
\ddot{R}=\underline{\vect{\dot{\omega}}}_e R+\underline{\vect\omega}^2_e R\qquad
\label{eq:Adotdot}
\end{align}

Given quaternion derivatives, it is possible to compute extrinsic angular velocity and extrinsic angular acceleration, or vice versa, using the following relations:
\begin{align}
\vect{\omega}_e &= 2 \; {F}(q\conj) \dot{\quat{q}} \\
\vect{\dot{\omega}}_e &= 2 \; {F}(q\conj) \ddot{\quat{q}} \\
\dot{\quat{q}} &= \tfrac{1}{2} {F}(q\conj)^T  \vect{\omega}_e  \\
\ddot{\quat{q}} &= \tfrac{1}{2} {F}(\dot{q}\conj)^T  \vect{\omega}_e + \tfrac{1}{2} {F}(q\conj)^T  \vect{\dot{\omega}}_e 
\end{align}
with the introduction of the matrix 
\begin{align}
\label{eq:at:rotFm} 
	\amatr{F(q)}&= 
	\left[
	\begin{matrix}
	 +q_1 & +q_0  & +q_3  & -q_2  \cr
	 +q_2 & -q_3  & +q_0  & +q_1  \cr
	 +q_3 & +q_2  & -q_1  & +q_0  
	\end{matrix}
	\right]
\end{align}

%-------------------------------------------------------------------------------
\section{Other interpolation schemes}
\label{app:otherinterpolations}
In the following, for completeness, we briefly describe other interpolation methods that are popular in literature and which we used in our benchmarks for comparisons.

\subsection{SLERP}
The spherical linear interpolation (SLERP \cite{shoemake1985animating}), with domain $t \in [0,1]$, interpolates two quaternions
$\quat{q}_a$ and $\quat{q}_b$ using the exponential formula \eqref{eq:at:quatpowd} to compute:
%
\begin{align}
\label{eq:at:quatrotSLERPa}
 \quat{q}(t) &= f_{\text{SLERP}}(\quat{q}_a,\quat{q}_b,t)  \nonumber \\
 &= \left( \quat{q}_b \quat{q}_a^{*} \right)^t \quat{q}_a
\end{align}

The norm of the quaternion is guaranteed to be unit length. The rotation matrix $R(t)$, if needed, can be computed from $\quat{q}(t)$ using \eqref{new_eq:fromQtoR}.

\subsection{SQUAD}
This method, also presented by Shoemake \cite{Shoemake1987QuaternionCA}, performs a cubic interpolation with domain $t \in [0,1]$ passing exactly at two extreme attitudes expressed by rotation quaternions $\quat{q}_1$ and $\quat{q}_2$, and approaching two intermediate control quaternions $\quat{p}_1$ and $\quat{p}_2$ similarly to cubic Bezier curves. It does this by using nested calls to the SLERP function, that is:
\begin{align}
\label{eq:at:quatrotSQUAD}
 \quat{q}(t) &=  f_{\text{SQUAD}}(\quat{q}_1,\quat{p}_1,\quat{p}_2,\quat{q}_2,t) \nonumber \\
  &= f_{\text{SLERP}}( f_{\text{SLERP}}(\quat{q}_1,\quat{q}_2 ,t),  f_{\text{SLERP}}(\quat{p}_1,\quat{p}_2 ,t), 2 t (1-t) )
\end{align}
In our implementation, an arbitrary sequence of $\quat{q}_i$ quaternions can be provided, to be interpolated exactly: for each span, the two intermediate $\quat{p}_1$ and $\quat{p}_2$ quaternions are computed to ensure $C^2$ continuity across span using the following formula: assuming the $i$-th span with quaternions 
$\{ \quat{q}_i, \quat{p}_i, \quat{p}_{i+1}, \quat{q}_{i+1}\}$, we compute 
\begin{align}
\label{eq:at:quatrotSQUADinner}
 \quat{p}_i &= \quat{q}_i \mathrm{exp} \left( - \frac{1}{4} \left( \mathrm{log}(\quat{q}_i^* \quat{q}_{i+1}) + \mathrm{log}(\quat{q}_i^* \quat{q}_{i-1})  \right)  \right)  
\end{align}

\subsection{Quaternion B-splines}
For completeness in the following we report the formulation of quaternion B-splines, that was put forward by Kim et al. in the nineties \cite{kim1995ac}, and that is used in our benchmarks for comparison with our method. 
We assumes a sequence of $n$ given quaternion control points $\quat{q}_i$, and a knot vector with $t_i$ knots. We assume that conventional B-spline basis functions $B_{i,k}(t)$ can be computed as $(k-1)$-order piecewise polynomials with non-zero support in $[t_i,t_{i+k}]$. Defining the cumulative form of the basis:
\begin{equation}
    \tilde{B}_{i,k}(t) = \sum_{j=1}^{n} B_{j,k}(t)  
\end{equation}
one can compute the interpolated quaternion with the expression
\begin{equation}
    \quat{q}(t) = \quat{q}_0^{\tilde{B}_{0,k}(t)} \prod_{i=1}^{n} \exp \left( \log(\quat{q}_{i-1}^* \quat{q}_i) \tilde{B}_{i,k}(t) \right)
\end{equation}
This method produces an interpolation that is $C^{k-2}$ continuous. For $k=2$ it boils down to the simple SLERP interpolation, but for higher orders, differently from the other interpolation schemes presented in this paper, it does not pass through the $\quat{q}_i$ values (except if knot multiplicity is enforced, for instance at the beginning and at the end of the curve).

\subsection{Triplets of angle functions}
Assuming that one computes a rotation matrix as a sequence of three rotations about three axis, one has
\begin{equation}
    R = R(\alpha,\beta,\gamma) = R(\gamma) R(\beta) R(\alpha)
\end{equation}
where $R(\alpha)$, $R(\beta)$ and $R(\gamma)$ are matrices of rotation about the X, Y, or Z orthogonal axes, in whatever user predefined order. 
If one provides three scalar functions for the three angles $\alpha(t)$, $\beta(t)$, $\gamma(t)$, one can obtain 
\begin{equation}
    R(t) = R(\gamma(t)) R(\beta(t)) R(\alpha(t))
\end{equation}
Similarly, if using quaternions, one would compute the double product
$\quat{q}(t)=\quat{q}(\gamma(t))\quat{q}(\beta(t))\quat{q}(\alpha(t))$. For instance, $\alpha(t)$ could be a B-spline or Bezier interpolation between $\alpha_i$ corresponding to $n$ keyframe rotations, and similarly for $\beta(t)$, $\gamma(t)$.
In any case, whatever the choice of ordering of the sequence of the three axis, there is always a singular attitude for whom, passing close to it, curves may degenerate in sharp and unpredictable oscillations. For this reason, interpolations based on triplets of angle functions are seldom used.

%-------------------------------------------------------------------------------
\section{Polynomial motion law}
\label{app:poly5}
In this section we summarize the quantities needed to define a five-degree polynomial motion law with assigned initial and final conditions of position, velocity and acceleration. If we name
\begin{itemize}
    \item[$s_i$, $s_f$]: initial and final position
    \item[$v_i$, $v_f$]: initial and final velocity
    \item[$a_i$, $a_f$]: initial and final acceleration
    \item[$\Delta s$]$= s_f - s_i$
\end{itemize}
and we impose the following relations
\begin{equation}
    \begin{array}{l}
        A = s_i\\
        B = v_i\\
        C = \cfrac{a_i}{2}\\
        D = \cfrac{a_f - 3a_i}{2T} - \cfrac{6v_i + 4v_f}{T^2} + \cfrac{10 \Delta s}{T^3}\\
        E = \cfrac{3a_i - 2a_f}{2T^2} + \cfrac{8v_i + 7v_f}{T^3} - \cfrac{15 \Delta s}{T^4}\\
        F = \cfrac{a_f - a_i}{2T^3} - \cfrac{3(v_i + v_f)}{T^4} + \cfrac{6 \Delta s}{T^5}\\
    \end{array}
\end{equation}
we may formulate the desired motion law and its derivatives as follows:
\begin{equation}
    \begin{array}{l}
        s(t) = Ft^5 + Et^4 + Dt^3 + Ct^2 + Bt + A\\
        \dot{s}(t) = 5Ft^4 + 4Et^3 + 3Dt^2 + 2Ct + B\\
        \ddot{s}(t) = 20Ft^3 + 12Et^2 + 6Dt + 2C\\
    \end{array}
    \label{eq:poly5}
\end{equation}

### 2021-03-02 2021/MMT_interpolation/main.tex
Add the interpolation Legnai paper for MMT

\documentclass[preprint, 12pt]{elsarticle}

% For journal layout:
% \documentclass[final,1p,times]{elsarticle}
% \documentclass[final,1p,times,twocolumn]{elsarticle}
% \documentclass[final,3p,times]{elsarticle}
% \documentclass[final,3p,times,twocolumn]{elsarticle}
% \documentclass[final,5p,times]{elsarticle}
% \documentclass[final,5p,times,twocolumn]{elsarticle}

%% Use packages ---------------------------------------------
\usepackage[english]{babel}
\usepackage{amssymb} 
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{color}
\usepackage{subcaption}
\usepackage{framed} % Framing content
\usepackage{nomencl} % Nomenclature package
\usepackage{booktabs}
\usepackage{siunitx} % SI units
\makenomenclature
\setlength{\nomitemsep}{-\parskip} % Baseline skip between items

%% Customization --------------------------------------------
% New commands Tasora 
\def\vect#1{{\boldsymbol{#1}}}
\def\quat#1{{\boldsymbol{#1}}}
\def\conj{\hskip0.1em^*} % the matrix as [R(a)] , with letter between two (autoresizing) square brackets
\def\amatr#1{{#1}}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}

% New commands Fusai 
\newcommand{\hl}[1]{\textcolor{red}{#1}}

%% Document ----------------------------------------------
\begin{document}

\journal{Mechanism and Machine Theory}

\begin{frontmatter}

% Title
\title{A practical algorithm for smooth interpolation between different angular positions}

% Authors and affiliations
\author[1,2,*]{Giovanni Legnani}
\author[2]{Irene Fassi}
\author[3]{Alessandro Tasora}
\author[3]{Dario Fusai}
\address[*]{corresponding author, giovanni.legnani@unibs.it}
\address[1]{DIMI-UniBS Dipartimento Ingegneria Meccanica e Industriale, Università di Brescia, Via Branze 38, 25123 Brescia, Italy}
\address[2]{STIIMA-CNR Istituto di Sistemi e Tecnologie Industriali per il Manifatturiero Avanzato, National Research Council, Via Alfonso Corti 12, 20133 Milano, Italy}
\address[3]{Dipartimento di Ingegneria e Architettura, Università di Parma, Parco Area delle Scienze, 181/A - 43124 PARMA, Italy}

% Abstract
\begin{abstract}
This paper proposes a new methodology for the interpolation of a given set of 3D rotation poses that have to be reached in successive times by preserving continuity in orientation, angular velocity and angular acceleration. The discussed algorithm ensures the generation of smooth angular trajectories without singularities. The distinctive features of the proposed approach are the straightforward formulation, the reduced computational burden and the lack of iterative procedures. The presented methodology has applications in the generation of spatial motion of mechanical systems (e.g. robotics, flying devices) or in 3D computer graphics. After a theoretical introduction, the proposed algorithm is compared with other methods available in literature and some possible applications are presented.
\end{abstract}

% Keywords
\begin{keyword}
Smooth rotation interpolation \sep Angular trajectory \sep Rotations \sep Quaternions
\end{keyword}

\end{frontmatter}

% Sections
\include{PaperQuaternionInterpolation}

% Bibliography
\bibliographystyle{plain}
% \bibliographystyle{elsart-num}
% \bibliographystyle{alpha}
\bibliography{bib/bibliografiaRobot}

\end{document}
\endinput
%% End document -----------------------------------------
